### 2015-11-23 Baxter_followUps_2016.Rmd
initial commit

---
title: "Follow Up Samples"
author: "Niel"
date: "November 20, 2015"
output: 
  html_document:
    keep_md: true
---

```{r startup, echo=F, message=F, warning=F}
deps = c("pROC","randomForest","AUCRF","knitr","rmarkdown");
for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
#setwd('~/Desktop/Baxter_followUps_2016/')

tax <- read.delim('data/followUps.final.an.unique_list.0.03.cons.taxonomy', sep='\t', header=T, row.names=1)
shared <- read.delim('data/followUps.final.an.unique_list.0.03.subsample.0.03.filter.shared', header=T, sep='\t')

metaF <- read.delim('data/followUps_metadata.txt', header=T, sep='\t')
metaF$lesion <- factor(NA, levels=c(0,1))
metaF$lesion[metaF$dx!='normal'] <-1
metaI <- read.delim('data/initials_metadata.tsv', header=T, sep='\t')
metaI$lesion <- factor(NA, levels=c(0,1))
metaI$lesion[metaI$dx=='normal'] <-0
metaI$lesion[metaI$dx!='normal'] <-1
```

```{r dists, echo=FALSE, fig.height=6, fig.width=6}
source('code/read.dist.R')

dist <- read.dist('data/followUps.final.an.unique_list.thetayc.0.03.lt.ave.dist')

intra <- c()
for(i in 1:nrow(metaF)){
  intra[i] <- as.numeric(dist[as.character(metaF$initial[i]),as.character(metaF$followUp[i])])
}

intra_ade <- intra[metaF$dx=='adenoma']
intra_canc <- intra[metaF$dx=='cancer']

inter_ade <- dist[as.character(metaF[metaF$dx=='adenoma','initial']),as.character(metaF[metaF$dx=='adenoma','initial'])]
inter_ade <- as.numeric(as.vector(unlist(inter_ade)))

inter_canc <- dist[as.character(metaF[metaF$dx=='cancer','initial']),as.character(metaF[metaF$dx=='cancer','initial'])]
inter_canc <- as.numeric(as.vector(unlist(inter_canc)))

inter <- dist[as.character(metaF$initial),as.character(metaF$followUp)]
inter <- as.numeric(as.vector(unlist(inter)))

plot(density(inter), type='l',lwd=3, col='royalblue', ylim=c(0,4), xlab='thetaYC distance', main='', las=1, tck=-0.02)
#points(density(inter_ade), type='l', lwd=3, col='dark orange')
#points(density(inter_canc), type='l', lwd=3, col='dark red')
points(density(intra_ade), type='l', lwd=3, col='orange')
points(density(intra_canc), type='l', lwd=3, col='red')
legend('topleft', legend=c('Between individuals','Within individuals (adenoma)','Within individuals (cancer)'), lwd=3, col=c('blue','orange','red'))
```

**Figure 1. Inter- vs Intra-personal Beta Diversity**  
The first thing I tested was whether the microbiomes of patients were more similar to themselves over time than to others. Surprisingly, there was no difference between intra- and interindividual thetaYC distances. Distances within cancer patients seems to be slightly less than within adenomas, but the difference isn't significant.  There's no difference when splitting interindividual distances by diagnosis. There's also no difference when comparing distances between initial samples to distances between followups.

```{r combined_model, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
# Make training set with samples withouth follow ups
train <- merge(metaI, shared, by.x='sample', by.y='Group')
train <- train[-match(metaF$initial, train$sample),]
train <- train[,c('lesion','fit_result',colnames(train)[grep('Otu[0123456789]', colnames(train))])]

# Get OTU abundances for initial samples that have followups
initial <- merge(metaF, shared, by.x='initial', by.y='Group')
initial <- initial[,c('lesion','fit_result',colnames(initial)[grep('Otu[0123456789]', colnames(initial))])]

# Get OTU abundances for follow ups
followups <- merge(metaF, shared, by.x='followUp', by.y='Group')
followups <- followups[,c('lesion','fit_result',colnames(followups)[grep('Otu[0123456789]', colnames(followups))])]

set.seed(112015)
rf <- AUCRF(lesion~., data=train, pdel=0.1, ntree=1000, ranking='MDA')
rf_opt <- rf$RFopt
#plot(rf)

train_probs <- predict(rf_opt, type='prob')[,2]
initial_probs <- predict(rf_opt, initial, type='prob')[,2]
follow_probs <- predict(rf_opt, followups, type='prob')[,2]

train_roc <- roc(train$lesion~train_probs)
cutoff <- coords(train_roc, x='best', ret='thr')

par(mar=c(4,4,1,1), xpd=F)
plot(initial_probs, follow_probs, ylim=c(0,1), xlim=c(0,1), pch=21, bg=c('orange','red')[metaF$dx], ylab='Probability of lesion (follow up sample)', xlab='Probability of lesion (initial sample)', main='OTUs + FIT')
abline(h=cutoff, v=cutoff, lty=2)
legend('bottomleft', legend=c('Cutoff','Initial Diagnosis:', 'Adenoma','Cancer'), pch=c(NA,NA,21,21), pt.bg=c(NA,NA,'orange','red'), lty=c(2,NA,NA,NA))
text(0.9,0.9, labels=sprintf('%.1f%%', 100*sum(initial_probs>cutoff & follow_probs>cutoff)/length(initial_probs)))
text(0.9,0.1, labels=sprintf('%.1f%%', 100*sum(initial_probs>cutoff & follow_probs<cutoff)/length(initial_probs)))
text(0.1,0.4, labels=sprintf('%.1f%%', 100*sum(initial_probs<cutoff & follow_probs<cutoff)/length(initial_probs)))
text(0.1,0.9, labels=sprintf('%.1f%%', 100*sum(initial_probs<cutoff & follow_probs>cutoff)/length(initial_probs)))
```

**Figure 2A. Combined Model (FIT + OTUs) applied to initial and follow up samples.**  
Next, I generated a random forest model for distinguishing normal from lesion using all of the initial samples that didn't have follow ups (n=422). The optimal model used `r rf$Kopt-1` OTUs and FIT. I then applied the model to the other initial samples and their follow ups. The scatter plot below shows how the model classified each individual's initial and follow up samples (n=67). The model detected `r sum(initial_probs[metaF$dx=='cancer']>cutoff)` of `r length(initial_probs[metaF$dx=='cancer'])` cancers and `r sum(initial_probs[metaF$dx=='adenoma']>cutoff)` of `r length(initial_probs[metaF$dx=='adenoma'])` adenomas in the initial samples. Of the `r sum(initial_probs>cutoff)` individuals who tested positive with the initial sample, `r sum(initial_probs>cutoff & follow_probs>cutoff)` remained positive with the follow up sample. 10.4% of individuls were negative for both the initial and followups (all adenomas). Surprisingly, 13.4% of individuals were negative at the initial sampling, but became positive at the follow up.

```{r fit_dif, echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0.05,5000), log='y', ylab='FIT Results (ng/ml)', xaxt='n', xlab='', yaxt='n')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
axis(2, at=c(0.1,1.1,10.1,100.1,1000.1), labels=c(0,1,10,100,1000))

points(rep(1,41), metaF[metaF$dx=='adenoma','fit_result']+0.1, pch=21, bg='orange', cex=0.8, lwd=0.5)
points(rep(2,41), metaF[metaF$dx=='adenoma','fit_followUp']+0.1, pch=21, bg='orange', cex=0.8, lwd=0.5)
segments(1, metaF[metaF$dx=='adenoma','fit_result']+0.1, 2, metaF[metaF$dx=='adenoma','fit_followUp']+0.1, col='orange')
ade_p<-t.test(metaF[metaF$dx=='adenoma','fit_result'], metaF[metaF$dx=='adenoma','fit_followUp'], paired=T, alternative='greater')$p.value
text(1.5, 3000, sprintf('p=%.1g',ade_p))

points(rep(3,26), metaF[metaF$dx=='cancer','fit_result']+0.1, pch=21, bg='red', cex=0.8, lwd=0.5)
points(rep(4,26), metaF[metaF$dx=='cancer','fit_followUp']+0.1, pch=21, bg='red', cex=0.8, lwd=0.5)
segments(3, metaF[metaF$dx=='cancer','fit_result']+0.1, 4, metaF[metaF$dx=='cancer','fit_followUp']+0.1, col='red')
canc_p <- t.test(metaF[metaF$dx=='cancer','fit_result'], metaF[metaF$dx=='cancer','fit_followUp'], paired=T, alternative='greater')$p.value
text(3.5, 3000, sprintf('p=%.1g',canc_p))

```

**Figure 2B.** This next figure shows the change in FIT result from the initial to the follow up samples. For the adenomas, it's fairly random. Some go from positive to negative, but just as many go form negative to positive. For the cancers, however, a large portion of them go from high FIT results to zero, while none of them go from zero to higher results. As a result, there is a significant reduction in FIT results for cancers, but not for adenomas.

```{r comb_dif, echo=FALSE, fig.height=6, fig.width=6}
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (Combined model)', xaxt='n', xlab='')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))

points(rep(1,41), initial_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=0.8, lwd=0.5)
points(rep(2,41), follow_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=0.8, lwd=0.5)
segments(1, initial_probs[metaF$dx=='adenoma'], 2, follow_probs[metaF$dx=='adenoma'], col='orange')
ade_p<-t.test(initial_probs[metaF$dx=='adenoma'], follow_probs[metaF$dx=='adenoma'], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
canc_p <- t.test(initial_probs[metaF$dx=='cancer'], follow_probs[metaF$dx=='cancer'], paired=T, alternative='greater')$p.value
text(3.5, 1, sprintf('p=%.1g',canc_p))
```

**Figure 2C.** This next figure shows how the model results changed over time. Based on the previous figure, you'd expect outcome of the model to change more for the cancer samples than for the adenomas. Indeed, there is a significant reduction in the probability of lesion for the cancers but not the adenomas. This is most likely due to the reduction in FIT results indiviuduals that had cancers.

```{r otu_model, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
# Make training set with samples withouth follow ups
train <- merge(metaI, shared, by.x='sample', by.y='Group')
train <- train[-match(metaF$initial, train$sample),]
train <- train[,c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]

# Get OTU abundances for initial samples that have followups
initial <- merge(metaF, shared, by.x='initial', by.y='Group')
initial <- initial[,c('lesion',colnames(initial)[grep('Otu[0123456789]', colnames(initial))])]

# Get OTU abundances for follow ups
followups <- merge(metaF, shared, by.x='followUp', by.y='Group')
followups <- followups[,c('lesion',colnames(followups)[grep('Otu[0123456789]', colnames(followups))])]

set.seed(112015)
rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=1000, ranking='MDA')
rf_opt <- rf_otu$RFopt
#plot(rf_otu)

train_probs <- predict(rf_opt, type='prob')[,2]
initial_probs <- predict(rf_opt, initial, type='prob')[,2]
follow_probs <- predict(rf_opt, followups, type='prob')[,2]

train_roc <- roc(train$lesion~train_probs)
cutoff <- coords(train_roc, x='best', ret='thr')

par(mar=c(4,4,1,1), xpd=F)
plot(initial_probs, follow_probs, ylim=c(0,1), xlim=c(0,1), pch=21, bg=c('orange','red')[metaF$dx], ylab='Probability of lesion (follow up sample)', xlab='Probability of lesion (initial sample)', main='OTUs only')
abline(h=cutoff, v=cutoff, lty=2)
legend('bottomleft', legend=c('Cutoff','Initial Diagnosis:', 'Adenoma','Cancer'), pch=c(NA,NA,21,21), pt.bg=c(NA,NA,'orange','red'), lty=c(2,NA,NA,NA), cex=0.8)
text(0.9,0.9, labels=sprintf('%.1f%%', 100*sum(initial_probs>cutoff & follow_probs>cutoff)/length(initial_probs)))
text(0.9,0.1, labels=sprintf('%.1f%%', 100*sum(initial_probs>cutoff & follow_probs<cutoff)/length(initial_probs)))
text(0.1,0.4, labels=sprintf('%.1f%%', 100*sum(initial_probs<cutoff & follow_probs<cutoff)/length(initial_probs)))
text(0.1,0.9, labels=sprintf('%.1f%%', 100*sum(initial_probs<cutoff & follow_probs>cutoff)/length(initial_probs)))

```

**Figure 3A. Otucome of Microbiome model (no FIT) on the initial and followup samples.** I did the same thing as figure 2, but left out the FIT results from the model. In this case the optimal used only `r rf_otu$Kopt-1` OTUs, yet it detected lesions nearly as well as the combined model. It detected detected `r sum(initial_probs[metaF$dx=='cancer']>cutoff)` of `r length(initial_probs[metaF$dx=='cancer'])` cancers (worse than the combined model) and `r sum(initial_probs[metaF$dx=='adenoma']>cutoff)` of `r length(initial_probs[metaF$dx=='adenoma'])` adenomas (better than the combined model) in the initial samples. Of the `r sum(initial_probs>cutoff)` initial samples that tested positive, `r sum(initial_probs>cutoff & follow_probs>cutoff)` remained positive. In summary, compared to the combined model, the microbiome only model was better at detected adenomas, worse at detecting cancers, and individuals were more likely to appear to "normal" at the follow up.

```{r dif_otu, echo=FALSE, fig.height=6, fig.width=6}
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))

points(rep(1,41), initial_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=0.8, lwd=0.5)
points(rep(2,41), follow_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=0.8, lwd=0.5)
segments(1, initial_probs[metaF$dx=='adenoma'], 2, follow_probs[metaF$dx=='adenoma'], col='orange')
ade_p<-t.test(initial_probs[metaF$dx=='adenoma'], follow_probs[metaF$dx=='adenoma'], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
canc_p <- t.test(initial_probs[metaF$dx=='cancer'], follow_probs[metaF$dx=='cancer'], paired=T, alternative='greater')$p.value
text(3.5, 1, sprintf('p=%.1g',canc_p))

```

**Figure 3B.** Unlike the combined model, there was significant reducation in probability of lesion for both patients with adenomas and those with cancer when using the microbiome only model.

```{r time, echo=FALSE, fig.height=5, fig.width=5}
plot(metaF$time, follow_probs-initial_probs, ylim=c(-0.8,0.5), xlab='Time between samples (days)',ylab='Change in model outcome')
```

**Figure 4.** I thought perhaps a longer time between sampling would increase the likelihood that an individual would appear "normal". However, there was no correlation between the change in model outcome and the time between samples.

```{r, echo=FALSE, warning=FALSE, fig.height=6, fig.width=9}
#OTU Scatterplots
i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
f_abunds <- followups[,-c(1,2)]/10530 + 0.0001
taxa <- c('Porphyromonas (OTU126)','Peptostreptococcus (OTU566)','Fusobacterium (OTU205)', 'Odoribacter (OTU164)', 'Prevotella (OTU147)','Parvimonas (OTU397)')

otus <- c("Otu000126", "Otu000566", "Otu000205", "Otu000164", "Otu000147", "Otu000397")

layout(matrix(c(1,2,3,4,5,6),nrow=2, byrow=T))
par(mar=c(1.5,1.5,2.5,1),oma=c(3,3,1,1))
for(i in 1:6){
  plot(log10(jitter(i_abunds[,otus[i]], amount=0.00001)), log10(jitter(f_abunds[,otus[i]], amount=0.00001)), xaxt='n', yaxt='n', ylab='', xlab='', pch=21, bg=c('orange','red')[metaF$dx], main=taxa[i], xlim=c(-4,max(log10(i_abunds[,otus[i]]))), ylim=c(-4,max(log10(i_abunds[,otus[i]]))), cex=1.5)
  axis(1, at=c(-4,-3,-2,-1), labels=c(0,0.1,1,10) )
  axis(2, at=c(-4,-3,-2,-1), labels=c(0,0.1,1,10) )
  abline(a=0,b=1, lty=2)
  legend('top', legend=sprintf('p=%.2g',wilcox.test(i_abunds[,otus[i]],f_abunds[,otus[i]], alternative='greater', paired=T)$p.value), bty='n')
}

layout(matrix(c(1,1),nrow=1))
mtext('% Relative abundance (initial)', side=1, line=3)
mtext('% Relative abundance (follow up)', side=2, line=3)
```

**Figure 6. The usual suspects.** Finally I wanted to see whether the OTUs most commonly associated with cancer were less abundant in the follow up samples. The figure above shows 6 of the 14 OTUs used in the microbiota only model. Most of them should look very familiar. You can see that for most of them there is a significant reduction in relative abundance from the initial sample to the follow up (p-values based on one-sided paired wilcoxon test). Interestngly, for Fusobacterium there is not a significant reduction unless you only compare cancer samples (`r sprintf('p=%.2g', wilcox.test(i_abunds[metaF$dx=='cancer',otus[3]],f_abunds[metaF$dx=='cancer',otus[3]], alternative='greater', paired=T)$p.value)`).

**Wild Speculation:**
To me, these finding are consistent with the idea that these oral pathogens are enriched on tumors, because they like the environment. When the tumors are removed their abundance goes down. They may accelerate tumorigenesis when they colonize the tumor, but first it's the tumor and/or inflammation that enables them to colonize in the first place. It's also possible, as we've  hypothesized in lab meetings, that Fuso binds first during the earliest stages (or even prior to) tumorigenesis.  Its presence then enables the other pathogens to adhere and exacerbate disease (similar to the progression of periodontitis). That would explain why people often find Fuso associated with adenomas, while these other oral pathogens are more indicative of cancers.

### 2015-11-23 code/read.dist.R

@@ -1,0 +1,30 @@
+read.dist <- function(file, input='lt', make.square=T, diag=0){
+  if(input=='lt'){
+    stuff <- scan(file, what='character') #gets all of the elements of the file
+    n <- as.numeric(stuff[1]) # n = number of groups/samples in file
+    stuff <- stuff[-1] # removes number of groups from list of stuff
+    m <- data.frame(matrix(NA, nrow=n, ncol=n) ) #makes empty matrix based on number of groups
+    diag(m) <- diag #fills in diagonal with specified value
+
+    c <- 1 # c keeps track of postion in stuff vector
+    for(i in 1:n){
+      group <- stuff[c] #get group name
+      colnames(m)[i] <- group
+      rownames(m)[i] <- group
+  
+      if(i > 1){
+      m[i,1:(i-1)] <- stuff[(c+1):(c+i-1)] #fills in matrix with values from stuff
+      }
+      c<-c+i #this because math
+    }
+    if(make.square){
+      m[upper.tri(m)] <- m[lower.tri(m)] #fills in upper triangle
+    }
+  }
+   
+  if(input=='square'){ #reads in square matrix
+    m<-read.table(file, skip=1, row.names=1)
+    colnames(m) <- rownames(m)
+  }
+  return(m)
+}

### 2016-04-28 Baxter_followUps_2016.Rmd
added change in oral bugs abunds

@@
 }
-#setwd('~/Desktop/Baxter_followUps_2016/')
+setwd('~/Desktop/Baxter_followUps_2016/')
 
@@
 
-```{r dists, echo=FALSE, fig.height=6, fig.width=6}
+```{r pval, echo=F}
+pval <- function(p){
+  if(p < 0.001){p <- 'p<0.001'}
+  else{p <- sprintf('p=%.1g', p)}
+  return(p)
+}
+
+```
+
+```{r dists, echo=FALSE, fig.height=5, fig.width=10}
 source('code/read.dist.R')
@@
 dist <- read.dist('data/followUps.final.an.unique_list.thetayc.0.03.lt.ave.dist')
-
 intra <- c()
@@
 }
-
 intra_ade <- intra[metaF$dx=='adenoma']
 intra_canc <- intra[metaF$dx=='cancer']
-
-inter_ade <- dist[as.character(metaF[metaF$dx=='adenoma','initial']),as.character(metaF[metaF$dx=='adenoma','initial'])]
-inter_ade <- as.numeric(as.vector(unlist(inter_ade)))
-
-inter_canc <- dist[as.character(metaF[metaF$dx=='cancer','initial']),as.character(metaF[metaF$dx=='cancer','initial'])]
-inter_canc <- as.numeric(as.vector(unlist(inter_canc)))
-
 inter <- dist[as.character(metaF$initial),as.character(metaF$followUp)]
@@
 
-plot(density(inter), type='l',lwd=3, col='royalblue', ylim=c(0,4), xlab='thetaYC distance', main='', las=1, tck=-0.02)
-#points(density(inter_ade), type='l', lwd=3, col='dark orange')
-#points(density(inter_canc), type='l', lwd=3, col='dark red')
+#pdf('figures/betaDiv.pdf', height=4.5, width=10)
+layout(matrix(c(1,2), nrow=1))
+par(mar=c(3,4,1,1), mgp=c(2,0.5,0))
+plot(density(inter), type='l',lwd=3, col='royalblue', ylim=c(0,4), xlab='thetaYC Index', main='', las=1, tck=-0.02)
 points(density(intra_ade), type='l', lwd=3, col='orange')
@@
 legend('topleft', legend=c('Between individuals','Within individuals (adenoma)','Within individuals (cancer)'), lwd=3, col=c('blue','orange','red'))
+
+dist <- read.dist('data/followUps.final.an.unique_list.sharedsobs.0.03.lt.ave.dist')
+intra <- c()
+for(i in 1:nrow(metaF)){
+  intra[i] <- as.numeric(dist[as.character(metaF$initial[i]),as.character(metaF$followUp[i])])
+}
+intra_ade <- intra[metaF$dx=='adenoma']
+intra_canc <- intra[metaF$dx=='cancer']
+inter <- dist[as.character(metaF$initial),as.character(metaF$followUp)]
+inter <- as.numeric(as.vector(unlist(inter)))
+
+par(mar=c(3,4.5,1,1), mgp=c(2,0.5,0))
+plot(density(inter), type='l',lwd=3, col='royalblue', ylim=c(0,0.03), , ylab='', xlab='Shared OTUs', main='', las=1, tck=-0.02)
+axis(2, at=0.015, line=2.5, tick=F, labels='Density')
+points(density(intra_ade), type='l', lwd=3, col='orange')
+points(density(intra_canc), type='l', lwd=3, col='red')
+legend('topright', legend=c('Between individuals','Within individuals (adenoma)','Within individuals (cancer)'), lwd=3, col=c('blue','orange','red'))
+
+#dev.off()
 ```
@@
 
+abline(h=100, lty=2)
 ```
@@
 text(3.5, 1, sprintf('p=%.1g',canc_p))
+
+abline(h=cutoff, lty=2)
 ```
@@
 
+abline(h=cutoff, lty=2)
 ```
@@
   abline(a=0,b=1, lty=2)
-  legend('top', legend=sprintf('p=%.2g',wilcox.test(i_abunds[,otus[i]],f_abunds[,otus[i]], alternative='greater', paired=T)$p.value), bty='n')
+  #legend('top', legend=sprintf('p=%.2g',wilcox.test(i_abunds[,otus[i]],f_abunds[,otus[i]], alternative='greater', paired=T)$p.value), bty='n')
 }
@@
 That would explain why people often find Fuso associated with adenomas, while these other oral pathogens are more indicative of cancers.
+
+```{r beforeVSafter, echo=FALSE, warning=FALSE, fig.height=6, fig.width=9, eval=FALSE}
+dim(shared)
+
+adeVcanc <- rbind(followups,initial)
+adeVcanc <- adeVcanc[,-1]
+adeVcanc$type <- c(rep(1),rep(0))
+adeVcanc$type <- as.factor(adeVcanc$type)
+
+set.seed(112015)
+rf_otu <- AUCRF(type~., data=adeVcanc, pdel=0.1, ntree=1000, ranking='MDA')
+rf_opt <- rf_otu$RFopt
+
+```
+
+```{r otu_abunds, echo=FALSE, fig.height=6, fig.width=6}
+i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
+f_abunds <- followups[,-c(1,2)]/10530 + 0.0001
+taxa <- c('P. asaccharolytica','P. stomatis','F. nucleatum', 'P. micra')
+
+otus <- c("Otu000126", "Otu000566", "Otu000205", "Otu000397")
+
+pdf('figures/otu_abunds.pdf', height=4, width=8)
+layout(matrix(c(1:4), nrow=1))
+par(mar=c(4,0.5,1,0.5), oma=c(0,4,0,0))
+for(i in 1:length(otus)){
+  plot(0, type='n', log='y', xlim=c(0.75,2.25), ylim=c(0.0001,0.2), ylab='', xaxt='n', xlab='', yaxt='n', las=2)
+  points(rep(1,nrow(i_abunds)), i_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
+  points(rep(2,nrow(f_abunds)), f_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
+  segments(1, i_abunds[,otus[i]], 2, f_abunds[,otus[i]], col=c('orange','red')[metaF$dx])
+  axis(1, at=c(1,2), c('Initial','Follow Up'))
+  mtext(taxa[i], 1, line=2.5, cex=0.8)
+  p_c<-wilcox.test(initial[metaF$dx=='cancer',otus[i]], followups[metaF$dx=='cancer',otus[i]], paired=T, alternative='greater')$p.value
+  text(1.5,0.19, labels=sprintf('Carcinomas: p=%.2g', p_c))
+  p_a<-wilcox.test(initial[metaF$dx=='adenoma',otus[i]], followups[metaF$dx=='adenoma',otus[i]], paired=T, alternative='greater')$p.value
+  text(1.5,0.13, labels=sprintf('Adenomas: p=%.2g', p_a))
+  if(i==1){axis(2, at=c(0.0001,0.001,0.01,0.1,1), c(0,0.1,1,10,100))
+      mtext('% Relative Abundace', 2, line=2.5, cex=0.8)}
+  #if(i==4){legend('top', c('Adenoma','Carcinoma'), pch=21, pt.bg=c('orange','red'))}
+}
+dev.off()
+```

### 2016-05-02 Baxter_followUps_2016.Rmd
added change in model by tumor stage

@@
 train <- train[-match(metaF$initial, train$sample),]
-train <- train[,c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]
+train <- train[metaI$dx!='adenoma',c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]
 
@@
 ```{r dif_otu, echo=FALSE, fig.height=6, fig.width=6}
+
 plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
@@
 
-points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
-points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
+points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=c(21,22,23,24,25)[metaF$stage], bg='red', cex=0.8, lwd=0.5)
+points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=c(21,22,23,24,25)[metaF$stage], bg='red', cex=0.8, lwd=0.5)
 segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
@@
 **Figure 3B.** Unlike the combined model, there was significant reducation in probability of lesion for both patients with adenomas and those with cancer when using the microbiome only model.
+
+```{r diff_by_stage, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
+# Make training set with samples withouth follow ups
+train <- merge(metaI, shared, by.x='sample', by.y='Group')
+train <- train[-match(metaF$initial, train$sample),]
+train <- train[, c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]
+
+# Get OTU abundances for initial samples that have followups
+initial <- merge(metaF, shared, by.x='initial', by.y='Group')
+initial <- initial[,c('lesion',colnames(initial)[grep('Otu[0123456789]', colnames(initial))])]
+
+# Get OTU abundances for follow ups
+followups <- merge(metaF, shared, by.x='followUp', by.y='Group')
+followups <- followups[,c('lesion',colnames(followups)[grep('Otu[0123456789]', colnames(followups))])]
+
+set.seed(050216)
+rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=500, ranking='MDA')
+rf_opt <- rf_otu$RFopt
+#plot(rf_otu)
+
+train_probs <- predict(rf_opt, type='prob')[,2]
+initial_probs <- predict(rf_opt, initial, type='prob')[,2]
+follow_probs <- predict(rf_opt, followups, type='prob')[,2]
+
+layout(matrix(c(1,2,3,4), nrow=1))
+par(mar=c(5,3,1,0.5))
+#non advanced
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Nonadvanced\nAdenoma', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+stripchart(at=1, initial_probs[metaF$Dx=='Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$Dx=='Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$Dx=='Adenoma'], 2, follow_probs[metaF$Dx=='Adenoma'], col='orange')
+ade_p<-t.test(initial_probs[metaF$Dx=='Adenoma'], follow_probs[metaF$Dx=='Adenoma'], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+#advanced
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Advanced\nAdenoma', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+stripchart(at=1, initial_probs[metaF$Dx=='adv Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$Dx=='adv Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$Dx=='adv Adenoma'], 2, follow_probs[metaF$Dx=='adv Adenoma'], col='orange')
+ade_p<-t.test(initial_probs[metaF$Dx=='adv Adenoma'], follow_probs[metaF$Dx=='adv Adenoma'], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+#Early (stage 1 or 2)
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Cancer\nStage I or II', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+stripchart(at=1, initial_probs[metaF$stage==1 | metaF$stage==2], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$stage==1 | metaF$stage==2], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$stage==1 | metaF$stage==2], 2, follow_probs[metaF$stage==1 | metaF$stage==2], col='red')
+ade_p<-t.test(initial_probs[metaF$stage==1 | metaF$stage==2], follow_probs[metaF$stage==1 | metaF$stage==2], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+#Early (stage 3 or 4)
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Cancer\nStage III or IV', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+stripchart(at=1, initial_probs[metaF$stage==3 | metaF$stage==4], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$stage==3 | metaF$stage==4], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$stage==3 | metaF$stage==4], 2, follow_probs[metaF$stage==3 | metaF$stage==4], col='red')
+ade_p<-t.test(initial_probs[metaF$stage==3 | metaF$stage==4], follow_probs[metaF$stage==3 | metaF$stage==4], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+```
 

### 2016-05-02 Baxter_followUps_2016.Rmd
removed otu scatter plots

@@
 
-```{r, echo=FALSE, warning=FALSE, fig.height=6, fig.width=9}
-#OTU Scatterplots
-i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
-f_abunds <- followups[,-c(1,2)]/10530 + 0.0001
-taxa <- c('Porphyromonas (OTU126)','Peptostreptococcus (OTU566)','Fusobacterium (OTU205)', 'Odoribacter (OTU164)', 'Prevotella (OTU147)','Parvimonas (OTU397)')
-
-otus <- c("Otu000126", "Otu000566", "Otu000205", "Otu000164", "Otu000147", "Otu000397")
-
-layout(matrix(c(1,2,3,4,5,6),nrow=2, byrow=T))
-par(mar=c(1.5,1.5,2.5,1),oma=c(3,3,1,1))
-for(i in 1:6){
-  plot(log10(jitter(i_abunds[,otus[i]], amount=0.00001)), log10(jitter(f_abunds[,otus[i]], amount=0.00001)), xaxt='n', yaxt='n', ylab='', xlab='', pch=21, bg=c('orange','red')[metaF$dx], main=taxa[i], xlim=c(-4,max(log10(i_abunds[,otus[i]]))), ylim=c(-4,max(log10(i_abunds[,otus[i]]))), cex=1.5)
-  axis(1, at=c(-4,-3,-2,-1), labels=c(0,0.1,1,10) )
-  axis(2, at=c(-4,-3,-2,-1), labels=c(0,0.1,1,10) )
-  abline(a=0,b=1, lty=2)
-  #legend('top', legend=sprintf('p=%.2g',wilcox.test(i_abunds[,otus[i]],f_abunds[,otus[i]], alternative='greater', paired=T)$p.value), bty='n')
-}
-
-layout(matrix(c(1,1),nrow=1))
-mtext('% Relative abundance (initial)', side=1, line=3)
-mtext('% Relative abundance (follow up)', side=2, line=3)
-```
-
-**Figure 6.
-The usual suspects.** Finally I wanted to see whether the OTUs most commonly associated with cancer were less abundant in the follow up samples.
-The figure above shows 6 of the 14 OTUs used in the microbiota only model.
-Most of them should look very familiar.
-You can see that for most of them there is a significant reduction in relative abundance from the initial sample to the follow up (p-values based on one-sided paired wilcoxon test).
-Interestngly, for Fusobacterium there is not a significant reduction unless you only compare cancer samples (`r sprintf('p=%.2g', wilcox.test(i_abunds[metaF$dx=='cancer',otus[3]],f_abunds[metaF$dx=='cancer',otus[3]], alternative='greater', paired=T)$p.value)`).
-
-**Wild Speculation:**
-To me, these finding are consistent with the idea that these oral pathogens are enriched on tumors, because they like the environment.
-When the tumors are removed their abundance goes down.
-They may accelerate tumorigenesis when they colonize the tumor, but first it's the tumor and/or inflammation that enables them to colonize in the first place.
-It's also possible, as we've  hypothesized in lab meetings, that Fuso binds first during the earliest stages (or even prior to) tumorigenesis.
-Its presence then enables the other pathogens to adhere and exacerbate disease (similar to the progression of periodontitis).
-That would explain why people often find Fuso associated with adenomas, while these other oral pathogens are more indicative of cancers.
-
 ```{r beforeVSafter, echo=FALSE, warning=FALSE, fig.height=6, fig.width=9, eval=FALSE}
@@
 ```
+
+```{r in_vs_fu, echo=FALSE, fig.height=6, fig.width=6}
+init_vs_follow <- rbind(initial,followups)[,-1]
+init_vs_follow$time <- factor(NA, levels=c(0,1))
+init_vs_follow$time[1:67]<-0
+init_vs_follow$time[68:134]<-1
+
+pvals <- c()
+for(i in 1:(ncol(init_vs_follow)-1)){
+  pvals[i]<-aov(time~init_vs_follow[,i], data=init_vs_follow)$p.value
+}
+
+#rf_time <- AUCRF(time~., data=init_vs_follow, ntree=500, pdel=0.1, ranking='MDA')
+
+```

### 2016-08-02 Baxter_followUps_2016.Rmd
removed unused chunk

--- 2016-04-28 Baxter_followUps_2016.Rmd
@@
 ---
-title: "Follow Up Samples"
+title: "Follow-up Samples"
 author: "Niel"
@@
 
-```{r dists, echo=FALSE, fig.height=5, fig.width=10}
+```{r dists, echo=FALSE, fig.height=5, fig.width=10, cache=T}
 source('code/read.dist.R')
@@
 
-set.seed(112015)
-rf <- AUCRF(lesion~., data=train, pdel=0.1, ntree=1000, ranking='MDA')
+set.seed(050416)
+rf <- AUCRF(lesion~., data=train, pdel=0.05, ntree=500, ranking='MDA')
 rf_opt <- rf$RFopt
@@
 ```{r fit_dif, echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
-plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0.05,5000), log='y', ylab='FIT Results (ng/ml)', xaxt='n', xlab='', yaxt='n')
+plot(1, type='n', xlim=c(0.5,4.5), ylim=c(0.05,5000), log='y', ylab='FIT Results (ng/ml)', xaxt='n', xlab='', yaxt='n')
 mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
-axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
 axis(2, at=c(0.1,1.1,10.1,100.1,1000.1), labels=c(0,1,10,100,1000))
@@
 ```{r comb_dif, echo=FALSE, fig.height=6, fig.width=6}
-plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (Combined model)', xaxt='n', xlab='')
+pdf('figures/mmt_dif.pdf', height=6, width=8)
+layout(1)
+par(mgp=c(2.5,0.5,0))
+plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1), ylab='Probability of lesion (Combined model)', xaxt='n', xlab='', las=2)
 mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
-axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+axis(1, at=c(1,2,3,4), labels=c('Initial','Follow-up','Initial','Follow-up'))
 
-points(rep(1,41), initial_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=0.8, lwd=0.5)
-points(rep(2,41), follow_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=0.8, lwd=0.5)
+jit <- jitter(rep(0,41), amount=0.05)
+points(rep(1,41)+jit, initial_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=1, lwd=0.5)
+points(rep(2,41)+jit, follow_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=1, lwd=0.5)
 segments(1, initial_probs[metaF$dx=='adenoma'], 2, follow_probs[metaF$dx=='adenoma'], col='orange')
@@
 
-points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
-points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
+jit <- jitter(rep(0,26), amount=0.05)
+points(rep(3,26)+jit, initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=1, lwd=0.5)
+points(rep(4,26)+jit, follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=1, lwd=0.5)
 segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
@@
 abline(h=cutoff, lty=2)
+dev.off()
 ```
@@
 
-```{r otu_model, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
+```{r otu_model, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6, eval=T}
 # Make training set with samples withouth follow ups
 train <- merge(metaI, shared, by.x='sample', by.y='Group')
-train <- train[-match(metaF$initial, train$sample),]
+train <- train[-which(train$sample %in% metaF$initial),]
 train <- train[,c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]
@@
 set.seed(112015)
-rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=1000, ranking='MDA')
+rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=500, ranking='MDA')
 rf_opt <- rf_otu$RFopt
-#plot(rf_otu)
 
@@
 ```{r dif_otu, echo=FALSE, fig.height=6, fig.width=6}
+
 plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
 mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
-axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
 
@@
 
+```{r diff_by_stage, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
+# Make training set with samples without follow ups
+train <- merge(metaI, shared, by.x='sample', by.y='Group')
+train <- train[-which(train$sample %in% metaF$initial),]
+train <- train[, c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]
+
+# Get OTU abundances for initial samples that have followups
+initial <- merge(metaF, shared, by.x='initial', by.y='Group')
+initial <- initial[,c('lesion',colnames(initial)[grep('Otu[0123456789]', colnames(initial))])]
+
+# Get OTU abundances for follow ups
+followups <- merge(metaF, shared, by.x='followUp', by.y='Group')
+followups <- followups[,c('lesion',colnames(followups)[grep('Otu[0123456789]', colnames(followups))])]
+
+set.seed(050216)
+rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=500, ranking='MDA')
+rf_opt <- rf_otu$RFopt
+#plot(rf_otu)
+
+train_probs <- predict(rf_opt, type='prob')[,2]
+initial_probs <- predict(rf_opt, initial, type='prob')[,2]
+follow_probs <- predict(rf_opt, followups, type='prob')[,2]
+
+layout(matrix(c(1,2,3,4), nrow=1))
+par(mar=c(5,3,1,0.5))
+#non advanced
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Nonadvanced\nAdenoma', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
+stripchart(at=1, initial_probs[metaF$Dx=='Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$Dx=='Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$Dx=='Adenoma'], 2, follow_probs[metaF$Dx=='Adenoma'], col='orange')
+ade_p<-t.test(initial_probs[metaF$Dx=='Adenoma'], follow_probs[metaF$Dx=='Adenoma'], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+#advanced
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Advanced\nAdenoma', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
+stripchart(at=1, initial_probs[metaF$Dx=='adv Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$Dx=='adv Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$Dx=='adv Adenoma'], 2, follow_probs[metaF$Dx=='adv Adenoma'], col='orange')
+ade_p<-t.test(initial_probs[metaF$Dx=='adv Adenoma'], follow_probs[metaF$Dx=='adv Adenoma'], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+#Early (stage 1 or 2)
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Cancer\nStage I or II', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
+stripchart(at=1, initial_probs[metaF$stage==1 | metaF$stage==2], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$stage==1 | metaF$stage==2], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$stage==1 | metaF$stage==2], 2, follow_probs[metaF$stage==1 | metaF$stage==2], col='red')
+ade_p<-t.test(initial_probs[metaF$stage==1 | metaF$stage==2], follow_probs[metaF$stage==1 | metaF$stage==2], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+#Early (stage 3 or 4)
+plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
+mtext(text='Cancer\nStage III or IV', at=c(1.5), side=1, line=4, font=2)
+axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
+stripchart(at=1, initial_probs[metaF$stage==3 | metaF$stage==4], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+stripchart(at=2, follow_probs[metaF$stage==3 | metaF$stage==4], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
+segments(1, initial_probs[metaF$stage==3 | metaF$stage==4], 2, follow_probs[metaF$stage==3 | metaF$stage==4], col='red')
+ade_p<-t.test(initial_probs[metaF$stage==3 | metaF$stage==4], follow_probs[metaF$stage==3 | metaF$stage==4], paired=T, alternative='greater')$p.value
+text(1.5, 1, sprintf('p=%.1g',ade_p))
+
+```
+
 ```{r time, echo=FALSE, fig.height=5, fig.width=5}
@@
 However, there was no correlation between the change in model outcome and the time between samples.
-
-```{r, echo=FALSE, warning=FALSE, fig.height=6, fig.width=9}
-#OTU Scatterplots
-i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
-f_abunds <- followups[,-c(1,2)]/10530 + 0.0001
-taxa <- c('Porphyromonas (OTU126)','Peptostreptococcus (OTU566)','Fusobacterium (OTU205)', 'Odoribacter (OTU164)', 'Prevotella (OTU147)','Parvimonas (OTU397)')
-
-otus <- c("Otu000126", "Otu000566", "Otu000205", "Otu000164", "Otu000147", "Otu000397")
-
-layout(matrix(c(1,2,3,4,5,6),nrow=2, byrow=T))
-par(mar=c(1.5,1.5,2.5,1),oma=c(3,3,1,1))
-for(i in 1:6){
-  plot(log10(jitter(i_abunds[,otus[i]], amount=0.00001)), log10(jitter(f_abunds[,otus[i]], amount=0.00001)), xaxt='n', yaxt='n', ylab='', xlab='', pch=21, bg=c('orange','red')[metaF$dx], main=taxa[i], xlim=c(-4,max(log10(i_abunds[,otus[i]]))), ylim=c(-4,max(log10(i_abunds[,otus[i]]))), cex=1.5)
-  axis(1, at=c(-4,-3,-2,-1), labels=c(0,0.1,1,10) )
-  axis(2, at=c(-4,-3,-2,-1), labels=c(0,0.1,1,10) )
-  abline(a=0,b=1, lty=2)
-  #legend('top', legend=sprintf('p=%.2g',wilcox.test(i_abunds[,otus[i]],f_abunds[,otus[i]], alternative='greater', paired=T)$p.value), bty='n')
-}
-
-layout(matrix(c(1,1),nrow=1))
-mtext('% Relative abundance (initial)', side=1, line=3)
-mtext('% Relative abundance (follow up)', side=2, line=3)
-```
-
-**Figure 6.
-The usual suspects.** Finally I wanted to see whether the OTUs most commonly associated with cancer were less abundant in the follow up samples.
-The figure above shows 6 of the 14 OTUs used in the microbiota only model.
-Most of them should look very familiar.
-You can see that for most of them there is a significant reduction in relative abundance from the initial sample to the follow up (p-values based on one-sided paired wilcoxon test).
-Interestngly, for Fusobacterium there is not a significant reduction unless you only compare cancer samples (`r sprintf('p=%.2g', wilcox.test(i_abunds[metaF$dx=='cancer',otus[3]],f_abunds[metaF$dx=='cancer',otus[3]], alternative='greater', paired=T)$p.value)`).
-
-**Wild Speculation:**
-To me, these finding are consistent with the idea that these oral pathogens are enriched on tumors, because they like the environment.
-When the tumors are removed their abundance goes down.
-They may accelerate tumorigenesis when they colonize the tumor, but first it's the tumor and/or inflammation that enables them to colonize in the first place.
-It's also possible, as we've  hypothesized in lab meetings, that Fuso binds first during the earliest stages (or even prior to) tumorigenesis.
-Its presence then enables the other pathogens to adhere and exacerbate disease (similar to the progression of periodontitis).
-That would explain why people often find Fuso associated with adenomas, while these other oral pathogens are more indicative of cancers.
 
@@
 set.seed(112015)
-rf_otu <- AUCRF(type~., data=adeVcanc, pdel=0.1, ntree=1000, ranking='MDA')
+rf_otu <- AUCRF(type~., data=adeVcanc, pdel=0.1, ntree=500, ranking='MDA')
 rf_opt <- rf_otu$RFopt
@@
 
-```{r otu_abunds, echo=FALSE, fig.height=6, fig.width=6}
+```{r otu_abunds, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=7}
 i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
@@
 
-pdf('figures/otu_abunds.pdf', height=4, width=8)
 layout(matrix(c(1:4), nrow=1))
 par(mar=c(4,0.5,1,0.5), oma=c(0,4,0,0))
+
+jity<-jitter(rep(0,67), amount=0.00001)
+jitx<-jitter(rep(0,67), amount=0.05)
+
 for(i in 1:length(otus)){
-  plot(0, type='n', log='y', xlim=c(0.75,2.25), ylim=c(0.0001,0.2), ylab='', xaxt='n', xlab='', yaxt='n', las=2)
-  points(rep(1,nrow(i_abunds)), i_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
-  points(rep(2,nrow(f_abunds)), f_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
-  segments(1, i_abunds[,otus[i]], 2, f_abunds[,otus[i]], col=c('orange','red')[metaF$dx])
-  axis(1, at=c(1,2), c('Initial','Follow Up'))
+  plot(0, type='n', log='y', xlim=c(0.75,2.25), ylim=c(0.00009,0.2), ylab='', xaxt='n', xlab='', yaxt='n', las=2)
+  points(rep(1,nrow(i_abunds))+jitx, jity+i_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
+  points(rep(2,nrow(f_abunds))+jitx, jity+f_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
+  segments(1+jitx, i_abunds[,otus[i]], 2+jitx, f_abunds[,otus[i]], col=c('orange','red')[metaF$dx])
+  axis(1, at=c(1,2), c('Initial','Follow-up'))
   mtext(taxa[i], 1, line=2.5, cex=0.8)
@@
 }
-dev.off()
 ```
+
+**Figure 5.** I compared the abundance of the four main cancer-associated OTUs over time.
+There was a significant decrease in the abundance of these OTUs for patients with carcinomas, but for patients with adenomas.

### 2016-08-02 Baxter_followUps_2016.Rmd
added legend for stage figure

@@
 
-```{r diff_by_stage, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
+```{r diff_by_stage, echo=FALSE, cache=TRUE, fig.height=4, fig.width=7}
+
 # Make training set with samples without follow ups
@@
 layout(matrix(c(1,2,3,4), nrow=1))
-par(mar=c(5,3,1,0.5))
+par(mar=c(5,3,1,0.5), mgp=c(1.5,0.5,0))
 #non advanced
@@
 ```
+
+There doesn't seem to be a difference in the outcome of the model by stage.
 

### 2016-09-29 exploratory/notebook/progress_report1_23092016.Rmd
code and analysis update

---
title: 'CRC Follow up: Progress Report 1'
author: "Marc Sze"
date: "September 23, 2016"
output: html_document
---

```{r read_relevent_libs, echo=FALSE, message=FALSE, warning=FALSE}

source('code/functions.R')
source('code/graphFunctions.R')

loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
           "gridExtra", "scales", "wesanderson", "vegan"))

```

Based on Niel's previous findings in his [Genome Medicine Paper](https://genomemedicine.biomedcentral.com/articles/10.1186/s13073-016-0290-3) I had a good idea of why combining Fit with the bacterial microbiome was beneficial.

This follow up study focuses in on 67 individuals who had both a before surgery stool sample and a 1 year post-surgery stool sample.  The first thing that I wanted to do was rebuild the models that Niel used specifically for his preliminary analysis of this data.

Training utilized those without follow up samples (n=490) and the testing utilized samples from those with follow up (n=67).  Initially looked at 4 models (normal vs. cancer, normal + adenoma vs. SRN + cancer, normal vs. lesion (adenoma + cancer), and three groups (normal vs. adenoma, vs. cancer)).  One of the things that I wanted to do was reduce the number of variables that were part of the model and investigate if I can do this without a corresponding loss in accuracy.  I used a specialized program known as [Boruta](https://cran.r-project.org/web/packages/Boruta/Boruta.pdf).  The number of variables fluctuate between 40-120 and Boruta can trim this list down to 10-30 variables.  Talking with Niel he suspects using cross-validation can also similarily shorten the list down as well.  The downside of using the `AUCRFcv` function is that it takes forever.

```{r AUC_graph1, echo=FALSE, message=FALSE, warning=FALSE}

# With Fit
sens_specific_table <- read.csv("results/tables/ROCCurve_sens_spec.csv", header = T, row.names = 1)
corr_pvalue_ROC_table <- read.csv("results/tables/ROCCurve_corrected_pvalues.csv", header = T, row.names = 1)

# Without Fit
sens_specific_table_noFit <- read.csv("results/tables/ROCCurve_sens_spec_NOFIT.csv", header = T, row.names = 1)
corr_pvalue_ROC_table_noFit <-read.csv("results/tables/ROCCurve_NOFIT_corrected_pvalues.csv", header = T, row.names = 1)

grid.arrange(
  # With Fit graph
  ggplot(sens_specific_table, aes(sensitivities, specificities)) + 
  geom_line(aes(group = model, color = model), size = 1.5) + scale_x_continuous(trans = "reverse") + 
  theme_bw() + xlab("Sensitivity") + ylab("Specificity") + ggtitle("With Fit") + 
  theme(axis.title = element_text(face = "bold")),

  # Without Fit graph
  ggplot(sens_specific_table_noFit, aes(sensitivities, specificities)) + 
  geom_line(aes(group = model, color = model), size = 1.5) + scale_x_continuous(trans = "reverse") + 
  theme_bw() + xlab("Sensitivity") + ylab("Specificity") + ggtitle("Without Fit") + 
  theme(axis.title = element_text(face = "bold"))

)

```

So one thing is that without Fit the models overall do worse but it seems that the Boruta selection is a quick way to remove data points without a loss in accuracy. For this part I am trying to wrap this up by comparing to the cross validated OTUs present in more than 50% of the runs.

We can next look at the accuracy of prediction in the follow up samples.  For this I split them based on whether the initial sample was diagnosed with adenoma or cancer.  I also only want to focus in on the lesion and SRNlesion models since those are the most important.  If we look at the adenoma's first:

```{r Accuracy_Adenoma_graph2, echo=FALSE, message=FALSE, warning=FALSE}

cutoffs_withFit <- read.csv("results/tables/withFIT.cutoffs.csv")
df_InitFollow_ALL_withFit <- read.csv("results/tables/withFIT.models.datatable.csv", header = T)

cutoffs_noFit <- read.csv("results/tables/noFIT.cutoffs.csv")
df_InitFollow_ALL_noFit <- read.csv("results/tables/noFIT.models.datatable.csv", header = T)

Names_facet <- c('SRNlesion' = "Classify SRN + Cancer", 'lesion' = "Classify Lesion")

grid.arrange(

  # With Fit in Model accuracy
  filter(df_InitFollow_ALL_withFit, 
         diagnosis == "adenoma" & dataset == "All" & model != "threeGroups" & model != "cancer") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive)) + 
    geom_jitter(aes(color=detailed_diagnosis), width = 0.3) + 
    scale_color_manual(name = "Polyp Type", values = c("cyan", "blue"), 
                       breaks = c("Adenoma", "adv Adenoma"), labels = c("Adenoma", "SRN")) + 
    facet_wrap(~model, labeller = as_labeller(Names_facet)) + coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(cutoffs_withFit, dataset == "All" & model != "threeGroups" & model != "cancer"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("Adenomas Prediction - Fit in Model") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold")),

  # Without Fit in Model accuracy
   filter(df_InitFollow_ALL_noFit, diagnosis == "adenoma" & dataset == "All" & model != "threeGroups") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive)) + 
    geom_jitter(aes(color=detailed_diagnosis), width = 0.3) + 
    scale_color_manual(name = "Polyp Type", values = c("cyan", "blue"), 
                       breaks = c("Adenoma", "adv Adenoma"), labels = c("Adenoma", "SRN")) + 
    facet_wrap(~model, labeller = as_labeller(Names_facet)) + coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(cutoffs_noFit, dataset == "All" & model != "threeGroups"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("Adenomas Prediction - No Fit in Model") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold"))
)

```

We can also look at the cancer ones as well.....

```{r Accuracy_cancer_graph3, echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(

  # With Fit model Cancer samples
  filter(df_InitFollow_ALL_withFit, 
         diagnosis != "adenoma" & dataset == "All" & model != "threeGroups" & model != "cancer") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive)) + 
    geom_jitter(aes(color=factor(diseaseFree, levels = c("n", "y", "unknown"))), width = 0.3) + 
    scale_color_manual(name = "Cancer\nFree", values = wes_palette("GrandBudapest")) + 
    facet_wrap(~model, labeller = as_labeller(Names_facet)) + coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(cutoffs_withFit, dataset == "All" & model != "threeGroups" & model != "cancer"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("CRC Prediction - Fit in Model") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold")),

   # Withot Fit model Cancer samples
  filter(df_InitFollow_ALL_noFit, 
         diagnosis != "adenoma" & dataset == "All" & model != "threeGroups") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive)) + 
    geom_jitter(aes(color=factor(diseaseFree, levels = c("n", "y", "unknown"))), width = 0.3) + 
    scale_color_manual(name = "Cancer\nFree", values = wes_palette("GrandBudapest")) + 
    facet_wrap(~model, labeller = as_labeller(Names_facet)) + coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(cutoffs_noFit, dataset == "All" & model != "threeGroups"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("CRC Prediction - Fit in Model") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold"))
)

```

So what I take from this overall is that the microbiome only model tends to do better with matching up with the inital and followup samples for the adenomas.  It doesn't do a good job of classifying cancers.  The question is why does having Fit in the model not make it align as well with the follow up samples?

```{r Sig_Imp_OTUs_lesionNoFit_graph4, echo=FALSE, message=FALSE, warning=FALSE}

# Load needed data
lesion_confirmed_vars_NOFIT <- read.csv('results/tables/lesion_confirmed_vars_NOFIT.csv', 
                                  header = TRUE, stringsAsFactors = F) %>% filter(otus != "fit_result")
SRNlesion_confirmed_vars_NOFIT <- read.csv('results/tables/SRNlesion_confirmed_vars_NOFIT.csv', 
                                     header = TRUE, stringsAsFactors = F) %>% filter(otus != "fit_result")
threeGroups_confirmed_vars_NOFIT <- read.csv('results/tables/threeGroups_confirmed_vars_NOFIT.csv', 
                                       header = TRUE, stringsAsFactors = F) %>% filter(otus != "fit_result")

tax <- read.delim('data/process/followUps.final.an.unique_list.0.03.cons.taxonomy', 
                  sep='\t', header=T, row.names=1)
tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
rownames(tax_df) <- rownames(tax)
colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
# This uses the R regex to remove numerals and the brackets the \\ is used to escape specific function components
rm(tax)

lesion_selected_taxa <- tax_df[lesion_confirmed_vars_NOFIT[, 'otus'], ]
lesion_selected_labs <- createTaxaLabeller(lesion_selected_taxa)

SRNlesion_selected_taxa <- tax_df[SRNlesion_confirmed_vars_NOFIT[, 'otus'], ]
SRNlesion_selected_labs <- createTaxaLabeller(SRNlesion_selected_taxa)

pvalue_summary_table_lesion <- read.csv("results/tables/lesion_Imp_vars_NOFIT_pvalue_summary.csv", header = T, row.names = 1)

pvalue_summary_table_SRNlesion <- read.csv("results/tables/SRNlesion_Imp_vars_NOFIT_pvalue_summary.csv", 
                                           header = T, row.names = 1)

specific_data_lesion_init <- read.csv("results/tables/lesion_imp_OTUs_abundance_test_init.csv")

specific_data_SRNlesion_init <- read.csv("results/tables/SRNlesion_imp_OTUs_abundance_test_init.csv")

specific_data_lesion_follow <- read.csv("results/tables/lesion_imp_OTUs_abundance_test_follow.csv")

specific_data_SRNlesion_follow <- read.csv("results/tables/SRNlesion_imp_OTUs_abundance_test_follow.csv")

grid.arrange(
  # Initial select microbiome sequences compared to normal
  filter(specific_data_lesion_init, 
         variable %in% rownames(pvalue_summary_table_lesion)[pvalue_summary_table_lesion$initial <= 0.05]) %>% 
    ggplot(aes(factor(sampleType, levels = c("normal", "lesion")), log10(value+1), group = 1)) + 
    geom_jitter(aes(color=factor(sampleType, levels = c("normal", "lesion"))), width = 0.5) + 
    scale_color_discrete(name = "Sample Type", breaks = c("normal", "lesion"), labels = c("normal", "initial")) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + ggtitle("Lesion Model") + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black") + coord_cartesian(ylim = c(0, 3.5)) + 
    facet_wrap(~variable, labeller = as_labeller(lesion_selected_labs)) + ylab("Log10 Subsampled Sequence Reads") + 
    xlab("") + theme_bw() + theme(strip.text.x = element_text(size = 8), axis.text.x = element_blank(), 
                                  axis.ticks.x = element_blank(), plot.title = element_text(face = "bold")) + 
    annotate("text", label = paste("P-value = ", format(round(pvalue_summary_table_lesion$initial[
      pvalue_summary_table_lesion$initial <= 0.05], digits = 7))), x = 1.5, y = 3.5),

  # Follow up microbiome sequences compared to normal
  filter(specific_data_lesion_follow, 
         variable %in% rownames(pvalue_summary_table_lesion)[pvalue_summary_table_lesion$initial <= 0.05]) %>% 
    ggplot(aes(factor(sampleType, levels = c("normal", "lesion")), log10(value+1), group = 1)) + 
    geom_jitter(aes(color=factor(sampleType, levels = c("normal", "lesion"))), width = 0.5) + 
    scale_color_discrete(name = "Sample Type", breaks = c("normal", "lesion"), labels = c("normal", "follow up")) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black") + coord_cartesian(ylim = c(0, 3.5)) + 
    facet_wrap(~variable, labeller = as_labeller(lesion_selected_labs)) + ylab("Log10 Subsampled Sequence Reads") + 
    xlab("") + theme_bw() + theme(strip.text.x = element_text(size = 8), axis.text.x = element_blank(), 
                                  axis.ticks.x = element_blank()) + 
    annotate("text", label = paste("P-value = ", format(round(pvalue_summary_table_lesion$follow_up[
      pvalue_summary_table_lesion$initial <= 0.05], digits = 7))), x = 1.5, y = 3.5)
)

```

If we look at this in more detail we can see that the OTUs that are used in the no FIT model (specifically Porphymonas since it is not in the Fit model) look more like the normals after follow up.

```{r Sig_Imp_OTUs_SRNlesionNoFit_graph5, echo=FALSE, message=FALSE, warning=FALSE}

# Graph for SRN Lesion model only
grid.arrange(
  # Initial select microbiome sequences compared to normal
  filter(specific_data_SRNlesion_init, 
         variable %in% rownames(pvalue_summary_table_SRNlesion)[pvalue_summary_table_SRNlesion$initial <= 0.05]) %>% 
    ggplot(aes(factor(sampleType, levels = c("normalorAdenoma", "SRNlesionORCancer")), log10(value+1), group = 1)) + 
    geom_jitter(aes(color=factor(sampleType, levels = c("normalorAdenoma", "SRNlesionORCancer"))), width = 0.5) + 
    scale_color_discrete(name = "Sample Type", breaks = c("normalorAdenoma", "SRNlesionORCancer"), 
                         labels = c("Normal or Adenoma", "initial")) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + ggtitle("SRN + Cancer Model") + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black") + coord_cartesian(ylim = c(0, 3.5)) + 
    facet_wrap(~variable, labeller = as_labeller(SRNlesion_selected_labs)) + ylab("Log10 Subsampled Sequence Reads") + 
    xlab("") + theme_bw() + theme(strip.text.x = element_text(size = 8), axis.text.x = element_blank(), 
                                  axis.ticks.x = element_blank(), plot.title = element_text(face = "bold")) + 
    annotate("text", label = paste("P-value = ", format(round(pvalue_summary_table_SRNlesion$initial[
      pvalue_summary_table_SRNlesion$initial <= 0.05], digits = 7))), x = 1.5, y = 3.5),

  # Follow up microbiome sequences compared to normal
  filter(specific_data_SRNlesion_follow, 
         variable %in% rownames(pvalue_summary_table_SRNlesion)[pvalue_summary_table_SRNlesion$initial <= 0.05]) %>% 
    ggplot(aes(factor(sampleType, levels = c("normalorAdenoma", "SRNlesionORCancer")), log10(value+1), group = 1)) + 
    geom_jitter(aes(color=factor(sampleType, levels = c("normalorAdenoma", "SRNlesionORCancer"))), width = 0.5) + 
    scale_color_discrete(name = "Sample Type", breaks = c("normalorAdenoma", "SRNlesionORCancer"), 
                         labels = c("Normal or Adenoma", "follow up")) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black") + coord_cartesian(ylim = c(0, 3.5)) + 
    facet_wrap(~variable, labeller = as_labeller(SRNlesion_selected_labs)) + 
    ylab("Log10 Subsampled Sequence Reads") + 
    xlab("") + theme_bw() + theme(strip.text.x = element_text(size = 8), axis.text.x = element_blank(), 
                                  axis.ticks.x = element_blank()) + 
    annotate("text", label = paste("P-value = ", format(round(pvalue_summary_table_SRNlesion$follow_up[
      pvalue_summary_table_SRNlesion$initial <= 0.05], digits = 7))), x = 1.5, y = 3.5)
)

```

This seems to hold true regardless of the model used.  So whether if it is lesion or SRN lesion the OTUs that are significantly different in the initials as a whole look more like normals on follow up.

```{r Fit_graph6, echo=FALSE, message=FALSE, warning=FALSE}

## Need to look at how FIT measurements compare to normal individuals

metaI <- read.csv("data/process/metaI_modified.csv", header = T, row.names = 1)
metaF <- read.csv("data/process/metaF_modified.csv", header = T, row.names = 1)

normalFit <- filter(metaI, dx == "normal") %>% select(fit_result)

initial_Fit_Pvalue <- wilcox.test(normalFit$fit_result, metaF$initial)$p.value
followUp_Fit_Pvalue <- wilcox.test(normalFit$fit_result, metaF$fit_followUp)$p.value

grid.arrange(
  # Initial Fit compared to normal Fit
  c(normalFit$fit_result, metaF$fit_result) %>% as.data.frame() %>% 
    mutate(sampleType = c(rep("normal", length(rownames(normalFit))), 
                                                           rep("initial", length(rownames(metaF))))) %>% 
    ggplot(aes(factor(sampleType, levels = c("normal", "initial")), log10(.+1), group = 1)) + 
    geom_jitter(aes(color=factor(sampleType, levels = c("normal", "initial"))), width = 0.5) + 
    scale_color_discrete(name = "Sample Type") + coord_cartesian(ylim = c(0, 4)) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + ggtitle("Normal versus Initial") + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black")  + 
    ylab("Fit Result (ng/mL)") + xlab("") + theme_bw() + 
    theme(strip.text.x = element_text(size = 8), axis.text.x = element_blank(), 
                                  axis.ticks.x = element_blank(), plot.title = element_text(face = "bold")) + 
    annotate("text", label = paste("P-value = ", format(round(initial_Fit_Pvalue, digits = 42), scientific = TRUE)), 
             x = 1.5, y = 3),

  # Follow up microbiome sequences compared to normal
  c(normalFit$fit_result, metaF$fit_followUp) %>% as.data.frame() %>% 
    mutate(sampleType = c(rep("normal", length(rownames(normalFit))), 
                          rep("follow_up", length(rownames(metaF))))) %>% 
    ggplot(aes(factor(sampleType, levels = c("normal", "follow_up")), log10(.+1), group = 1)) + 
    geom_jitter(aes(color=factor(sampleType, levels = c("normal", "follow_up"))), width = 0.5) + 
    scale_color_discrete(name = "Sample Type") + coord_cartesian(ylim = c(0, 4)) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + ggtitle("Normal versus Follow Up") + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black") + 
    ylab("Fit Result (ng/mL)") + xlab("") + theme_bw() + 
    theme(strip.text.x = element_text(size = 8), axis.text.x = element_blank(), plot.title = element_text(face = "bold"), 
                                  axis.ticks.x = element_blank()) + 
    annotate("text", label = paste("P-value = ", format(round(followUp_Fit_Pvalue, digits = 7))), x = 1.5, y = 3)
)

```

In contrast the Fit values remain significantly elevated in the follow up samples versus those in the normal group.  This is probably the main driver of the difference between the with Fit and without FIt models and why the adenoma seems to do better with the no Fit model.  Another reason why it might do better is below:

```{r amountChange_graph7, echo=FALSE, message=FALSE, warning=FALSE}

thetaCompTotal <- dissplit('data/process/followUps.final.an.unique_list.thetayc.0.03.lt.ave.dist',split=F, meta = F)

difference_table_treatment <- pickDistanceValues(
  as.character(metaF$initial), as.character(metaF$followUp), thetaCompTotal, metaF, 
  c("fit_result", "fit_followUp", "Dx_Bin", "dx")) %>% 
  mutate(fit_difference = fit_result - fit_followUp)

thetaDiff_follow_pvalue <- wilcox.test(distance ~ dx, data = difference_table_treatment)$p.value

fitDiff_follow_pvalue <- wilcox.test(fit_difference ~ dx, data = difference_table_treatment)$p.value

grid.arrange(
  # Difference from bacterial community structure between adenoma and cancer
  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), distance, group = 1)) + 
    geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
    scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
                       breaks = c("Adenoma", "adv Adenoma", "Cancer"), labels = c("Adenoma", "SRN", "Cancer")) + 
    coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold")) + 
    annotate("text", label = paste("P-value = ", round(thetaDiff_follow_pvalue, digits = 2)), x = 1.5, y = 0.75),

  # Difference from fit between adenoma and cancer
  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), fit_difference, group = 1)) + 
    geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
    scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
                       breaks = c("Adenoma", "adv Adenoma", "Cancer"), labels = c("Adenoma", "SRN", "Cancer")) + 
    ylab("Change in Fit from Initial to Follow Up") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold")) + 
    annotate("text", label = paste("P-value = ", format(round(fitDiff_follow_pvalue, digits = 7))), x = 1.5, y = 0)
)

```

SO what this shows here is that there is a lot of change versus initial going on with cancer but very little with adenoma in fit.  Whereas a similar amount of change is going on in thetayc for both adenoma and cancer.

This leads into the final part where the thetayc of those with follow ups though still look different from normal.  So even though there are individual members changing with and without polyp/tumor the overall community still seems different than normal individuals.

```{r NMDS_followSamples_vs_Norms_graph8, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}

breakDown_samples <- c(rep("initial", length(rownames(metaF))), rep("follow_up", length(rownames(metaF))), 
                       rep("normal", length(metaI$sample[metaI$dx == "normal"])))

theta_init_follow_dist <- pickDistanceValues(
  as.character(c(metaF$initial, metaF$followUp, metaI$sample[metaI$dx == "normal"])), 
  as.character(c(metaF$initial, metaF$followUp, metaI$sample[metaI$dx == "normal"])), 
  thetaCompTotal, withMeta = FALSE)

set.seed(050416)
thetayc.mds <- metaMDS(as.dist(theta_init_follow_dist), trace = 0) %>% scores() %>% as.data.frame() %>% 
  mutate(samples = factor(breakDown_samples))

ggplot(data = thetayc.mds, aes(x=NMDS1, y=NMDS2)) + geom_point(aes(color=samples)) + theme_bw() + coord_equal() + 
  stat_ellipse(aes(group = samples, color = samples, fill = samples), alpha = 0.25, geom = "polygon")

```

However, overall the samples start off looking pretty similar to normals regardless.  There is a significant difference though between the groups as measured by PERMANOVA.

### 2016-09-29 code/functions.R

@@ -16,47 +16,72 @@ loadLibs <- function(deps){
 
 # This function creates a square distance matrix and splits it based 
 # on specified criteria for init and follow. 
-dissplit <- function(file, metafile, init='initial', follow='followUp'){
+dissplit <- function(file, metafile, init='initial', follow='followUp', split=T, meta = T){
   source('code/read.dist.R')
   dist <- read.dist(file)
   intra <- c()
-  for(i in 1:nrow(metafile)){
-    intra[i] <- as.numeric(dist[
-      as.character(metafile[i, init]),as.character(metafile[i, follow])])
+  # Complete only if split is wanted
+  if (meta == T){
+    
+    for(i in 1:nrow(metafile)){
+      intra[i] <- as.numeric(dist[
+        as.character(metafile[i, init]),as.character(metafile[i, follow])])
+    }
+    intra_ade <- intra[metaF$dx=='adenoma']
+    intra_canc <- intra[metaF$dx=='cancer']
+    
+    ###Pull out only interindividual distances and convert to vector
+    inter <- dist[as.character(metaF$initial),as.character(metaF$followUp)]
+    
+    #Remove those that are in the intra and replace with NA
+    diag(inter) <- NA
+    #Check that it only took out 67 measurements
+    length(which(is.na(inter)))
+    
+    inter <- as.numeric(as.vector(unlist(inter)))
+    inter <- inter[!is.na(inter)]
+    
+    combined <- list(intra_ade, intra_canc, inter)
+    names(combined) <- c("intra_ade", "intra_canc", "inter")
+    
   }
-  intra_ade <- intra[metaF$dx=='adenoma']
-  intra_canc <- intra[metaF$dx=='cancer']
   
-  ###Pull out only interindividual distances and convert to vector
-  inter <- dist[as.character(metaF$initial),as.character(metaF$followUp)]
+ if (split == T){
+    return(combined)
+  } else {
+    return(dist)
+  }
   
-  #Remove those that are in the intra and replace with NA
-  diag(inter) <- NA
-  #Check that it only took out 67 measurements
-  length(which(is.na(inter)))
   
-  inter <- as.numeric(as.vector(unlist(inter)))
-  inter <- inter[!is.na(inter)]
-  
-  combined <- list(intra_ade, intra_canc, inter)
-  names(combined) <- c("intra_ade", "intra_canc", "inter")
-  
-  return(combined)
 }
 
 
[…]

### 2016-09-29 code/graphFunctions.R

@@ -35,7 +35,7 @@ follow_Abundance_Fit_Graph <- function(taxaFile, labelFile, followUpFile, initia
     ggplot(aes(factor(sampleType, levels = c("initial", "follow_up")), 
                log10(jitteredValues + 1.1), group = factor(EDRN))) + 
     geom_line(aes(color = factor(Disease_Free))) + geom_point(aes(color = factor(Disease_Free))) +  
-    facet_wrap(~variable, labeller = as_labeller(labelFile)) + scale_y_continuous(limits = c(0, 3)) + 
+    facet_wrap(~variable, labeller = as_labeller(labelFile)) + coord_cartesian(ylim = c(0, 3)) + 
     theme_bw() + ylab("Log Total Sequences") + xlab("") + ggtitle("Adenoma") + 
     scale_colour_manual(values = "blue") +  
     theme(legend.position="none", plot.title = element_text(size=20, face="bold"), 
@@ -47,7 +47,7 @@ follow_Abundance_Fit_Graph <- function(taxaFile, labelFile, followUpFile, initia
                log10(jitteredValues+1.1), group = factor(EDRN))) + 
     geom_line(aes(color = factor(Disease_Free, levels = c("n", "y", "unknown")))) + 
     geom_point(aes(color = factor(Disease_Free, levels = c("n", "y", "unknown")))) + 
-    facet_wrap(~variable, labeller = as_labeller(labelFile)) + scale_y_continuous(limits = c(0, 3)) + 
+    facet_wrap(~variable, labeller = as_labeller(labelFile)) + coord_cartesian(ylim = c(0, 3)) + 
     theme_bw() + ylab("Log Total Sequences") + xlab("") + ggtitle(paste(positiveTitle)) + 
     scale_colour_manual(values = c("darkred", "pink", "red")) + 
     theme(legend.position="none", plot.title = element_text(size=20, face="bold"), 
@@ -58,7 +58,7 @@ follow_Abundance_Fit_Graph <- function(taxaFile, labelFile, followUpFile, initia
     ggplot(aes(factor(variable, levels = c("initial", "follow_up")), 
                log10(jitteredValues+1.1), group = factor(EDRN))) + 
     geom_line(aes(color = factor(Disease_Free))) + geom_point(aes(color = factor(Disease_Free))) + 
[…]

### 2016-09-29 exploratory/20160915_RandomForestModelingPt2_NOFIT.R

@@ -350,18 +350,64 @@ grid.arrange(
 )
 
 
+### Create a new group with normals and follow ups without cancer: find OTUs that are still different
+followUp_Select <- rbind(filter(metaI, dx == "normal") %>% select(sample, dx) %>% mutate(dx = as.character(dx)) %>% 
+                           mutate(disease = rep(0, length(rownames(.)))), 
+                         select(metaF, followUp) %>% mutate(dx = rep("follow_up", 
+                                                                     length(rownames(metaF)))) %>% rename(sample = followUp) %>% 
+                           mutate(disease = rep(1, length(rownames(metaF))))) %>% 
+  inner_join(shared, by = c("sample" = "Group")) %>% select(disease, contains("Otu0")) %>% mutate(disease = factor(disease))
 
 
+set.seed(050416)
+follow_v_norm_diffs <- Boruta(disease~., data=followUp_Select, mcAdj=TRUE, maxRuns=1000)
+follow_v_norm_diffs_rf_opt <- AUCRF(disease~., data=followUp_Select, pdel=0.05, ntree=500, ranking='MDA')$RFopt
+
+# Get the confirmed important variables
+follow_v_norm_diffs_confirmed_vars_OTUs <- as.data.frame(follow_v_norm_diffs['finalDecision'])  %>% 
+  mutate(otus = rownames(.))  %>% filter(finalDecision == "Confirmed")  %>% select(otus)
 
 
+# create table for p-value test
+pValue_test_data <- as.data.frame(cbind(c(metaI$sample[metaI$dx == "normal"], metaF$initial, metaF$followUp), 
+                                                          c(rep("normal", length(rownames(metaI[metaI$dx == "normal", ]))), 
+                                                            rep("initial", length(rownames(metaF))), 
+                                                            rep("follow_up", length(rownames(metaF)))))) %>% 
+  rename(Group = V1, sampleType = V2) %>% mutate(Group = as.integer(as.character((Group)))) %>% 
+  inner_join(shared, by = "Group") %>% select(Group, sampleType, one_of(follow_v_norm_diffs_confirmed_vars_OTUs$otus)) %>% 
+  mutate(Otutotal8OTUs = rowSums(.[, follow_v_norm_diffs_confirmed_vars_OTUs$otus]))
 
 
[…]

### 2016-10-14 code/20160816_RandomForestModelingPt1.R
update code and add needed components for results

@@ -1,0 +1,1270 @@
+## Run Previous Analysis align follow up data
+## Code used and modified from Niel .Rmd file
+## Marc Sze
+
+###Load needed Libraries and functions
+source('code/functions.R')
+source('code/graphFunctions.R')
+
+loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson"))
+
+### Read in necessary data 
+
+tax <- read.delim('data/process/followUps.final.an.unique_list.0.03.cons.taxonomy', 
+                  sep='\t', header=T, row.names=1)
+shared <- read.delim('data/process/followUps.final.an.unique_list.0.03.subsample.0.03.filter.shared', 
+                     header=T, sep='\t')
+
+### Organize tables for train and test sets (first focus is to look at non-cancer versus cancer)
+metaF <- read.delim('data/process/followUps_metadata.txt', header=T, sep='\t') %>% mutate(lesion = factor(NA, levels=c(0,1)))
+metaF$cancer[metaF$dx =='normal' | metaF$dx =='adenoma'] <- 0
+metaF$cancer[metaF$dx =='cancer'] <- 1
+metaF$cancer <- factor(metaF$cancer)
+
+metaI <- read.delim('data/process/initials_metadata.tsv', header=T, sep='\t') %>% mutate(lesion = factor(NA, levels=c(0,1)))
+metaI$cancer[metaI$dx == 'normal' | metaI$dx == 'adenoma'] <- 0
+metaI$cancer[metaI$dx == 'cancer'] <- 1
+metaI$cancer <- factor(metaI$cancer)
+
+### Run Second organization to sort lesion vs. non-lesion
+metaF$lesion[metaF$dx =='normal'] <- 0
+metaF$lesion[metaF$dx =='adenoma' | metaF$dx == 'cancer'] <- 1
+metaF$lesion <- factor(metaF$lesion)
+
+metaI$lesion[metaI$dx == 'normal'] <- 0
+metaI$lesion[metaI$dx == 'adenoma' | metaI$dx == 'cancer'] <- 1
+metaI$lesion <- factor(metaI$lesion)
+
+### Run third organization to sort SRN (Adv Adenoma) to cancer group and Adenoma to non-cancer group
+metaF$SRNlesion[metaF$Dx_Bin =='normal' | metaF$Dx_Bin == 'Adenoma'] <- 0
[…]

### 2016-10-14 code/20160919_ToDoDataAnalysis_pt1.R

@@ -1,0 +1,371 @@
+## A To Do List to follow up on stemming from original exploratory analysis
+## Focus strictly on Lesion, SRNLesion, and three groups (Normal, Adenoma, Cancer) classifications
+## Marc Sze
+
+# Load required dependencies and libraries
+source('code/functions.R')
+source('code/graphFunctions.R')
+
+loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson", "vegan"))
+
+
+## Need to look at differences between initial and follow up between microbiome (thetayc) and fit.
+# Does the microbiome show a difference in community versus the fit?
+
+metaI <- read.delim('data/process/initials_metadata.tsv', header=T, sep='\t')
+metaF <- read.delim('data/process/followUps_metadata.txt', header=T, sep='\t') %>% mutate(lesion = factor(NA, levels=c(0,1)))
+thetaCompTotal <- dissplit('data/process/followUps.final.an.unique_list.thetayc.0.03.lt.ave.dist',split=F, meta = F)
+
+difference_table_treatment <- pickDistanceValues(
+  as.character(metaF$initial), as.character(metaF$followUp), thetaCompTotal, metaF, 
+  c("fit_result", "fit_followUp", "Dx_Bin", "dx")) %>% 
+  mutate(fit_difference = fit_result - fit_followUp)
+
+thetaDiff_follow_pvalue <- wilcox.test(distance ~ dx, data = difference_table_treatment)$p.value
+
+fitDiff_follow_pvalue <- wilcox.test(fit_difference ~ dx, data = difference_table_treatment)$p.value
+
+grid.arrange(
+  # Difference from bacterial community structure between adenoma and cancer
+  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), distance, group = 1)) + 
+    geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
+    stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
+    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
+    scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
[…]

### 2016-10-14 code/20161003_Run_base_RF_wFit.R

@@ -51,6 +51,7 @@ for(i in 1:length(train)){
 
 }  
 
+
 #Set up Boruta picked data list
 
 impfactor_Data_List <- list(cancer = c(), lesion = c(), SRNlesion = c(), threeway = c())

### 2016-10-14 code/20161004_Run_base_testModels_wFit.R

@@ -58,6 +58,14 @@ initial_predictions <- lapply(rf_opt_NameList, function(x) predict(x, initial, t
 followup_predictions <- lapply(rf_opt_NameList, function(x) predict(x, followups, type='prob'))
 
 
+#Create ROC data for lesion model
+
+lesion_combined_ROC <- roc(c(initial$lesion, followups$lesion) ~ 
+                             c(initial_predictions$lesion_rf_opt[, 2], followup_predictions$lesion_rf_opt[, 2]))
+
+write.csv(matrix(c("withFit", lesion_combined_ROC$auc), ncol = 2), "results/tables/testSetAUC.csv", row.names = F)
+
+
 # Join good_metaf table with metaF table
 combined_metaData <- inner_join(good_metaf, metaF, by = "EDRN")
 
@@ -187,7 +195,8 @@ crc_graph <- grid.arrange(
 ggsave(file = "results/figures/crc_predictions_withFit.tiff", crc_graph, width=9, height = 17, dpi = 300)
 
 
-
+rm(list=setdiff(ls(), "lesion_combined_ROC"))
+save.image("exploratory/wFit_Test_ROC.RData")
 
 
 

### 2016-10-14 code/20161005_Run_base_testModels_woFit.R

@@ -51,6 +51,16 @@ initial_predictions <- lapply(rf_opt_NameList, function(x) predict(x, initial, t
 followup_predictions <- lapply(rf_opt_NameList, function(x) predict(x, followups, type='prob'))
 
 
+#Create ROC data for lesion model
+
+AUCData <- read.csv("results/tables/testSetAUC.csv", header = T, stringsAsFactors = F)
+
+lesion_combined_ROC <- roc(c(initial$lesion, followups$lesion) ~ 
+                             c(initial_predictions$lesion_rf_opt[, 2], followup_predictions$lesion_rf_opt[, 2]))
+
+write.csv(rbind(AUCData, c("woFit", lesion_combined_ROC$auc)), "results/tables/testSetAUC.csv", row.names = F)
+
+
 # Join good_metaf table with metaF table
 combined_metaData <- inner_join(good_metaf, metaF, by = "EDRN")
 
@@ -175,7 +185,8 @@ crc_graph <- grid.arrange(
 ggsave(file = "results/figures/crc_predictions_noFit.tiff", crc_graph, width=8, height = 8, dpi = 300)
 
 
-
+rm(list=setdiff(ls(), "lesion_combined_ROC"))
+save.image("exploratory/woFit_Test_ROC.RData")
 
 
 

### 2016-10-14 code/20161006_Run_Beta_tests.R

@@ -215,11 +215,13 @@ set.seed(050416)
 pValueList[["bdiver_Test_adn_IF"]] <- adonis(as.dist(polyp_only_theta_init_follow_dist) ~ factor(breakDown_samples))
 
 set.seed(050416)
-thetayc.mds.List[["bdiver_Test_adn_IF"]] <- metaMDS(as.dist(polyp_only_theta_init_follow_dist)) %>% 
+thetayc.mds.List[["bdiver_Test_adn_IF"]] <- metaMDS(as.dist(polyp_only_theta_init_follow_dist), trace = 0) %>% 
   scores() %>% as.data.frame() %>%  mutate(samples = factor(breakDown_samples))
 
 ggplot(data = thetayc.mds.List[["bdiver_Test_adn_IF"]], aes(x=NMDS1, y=NMDS2)) + geom_point(aes(color=samples)) + theme_bw() + coord_equal() + 
-  stat_ellipse(aes(group = samples, color = samples, fill = samples), alpha = 0.25, geom = "polygon")
+  stat_ellipse(aes(group = samples, color = samples, fill = samples), alpha = 0.25, geom = "polygon") + 
+  annotate("text", label = paste("PERMANOVA = ", round(pValueList[["bdiver_Test_adn_IF"]]$aov.tab$`Pr(>F)`[1], 
+                                                       digits = 3)), x = 0.25, y = -0.50)
 
 
 ## Compare adenoma init, follow versus normal

### 2016-10-14 code/20161007_Imp_OTU_proportion_tests.R

@@ -112,9 +112,10 @@ for(i in 1:length(rownames(lesion_selected_taxa_WF))){
 }
 
 OTU_WF_pvalues_summary <- as.data.frame(cbind(
-  OTU_WF_pvalues, p.adjust(OTU_WF_pvalues, method = "bonferroni"), lesion_selected_labs_WF))
+  OTU_WF_pvalues, p.adjust(OTU_WF_pvalues, method = "bonferroni"), 
+  lesion_selected_labs_WF, rownames(lesion_selected_taxa_WF)))
 
-colnames(OTU_WF_pvalues_summary) <- c("OTU_WF_pvalues", "OTU_WF_adjust_pvalues", "Tax_ID") 
+colnames(OTU_WF_pvalues_summary) <- c("OTU_WF_pvalues", "OTU_WF_adjust_pvalues", "Tax_ID", "otus") 
 
 
 OTU_WoF_pvalues <- c()
@@ -126,9 +127,10 @@ for(i in 1:length(rownames(lesion_selected_taxa_WoF))){
 }
 
 OTU_WoF_pvalues_summary <- as.data.frame(cbind(
-  OTU_WoF_pvalues, p.adjust(OTU_WoF_pvalues, method = "bonferroni"), lesion_selected_labs_WoF))
+  OTU_WoF_pvalues, p.adjust(OTU_WoF_pvalues, method = "bonferroni"), 
+  lesion_selected_labs_WoF, rownames(lesion_selected_taxa_WoF)))
 
-colnames(OTU_WoF_pvalues_summary) <- c("OTU_WoF_pvalues", "OTU_WoF_adjust_pvalues", "Tax_ID") 
+colnames(OTU_WoF_pvalues_summary) <- c("OTU_WoF_pvalues", "OTU_WoF_adjust_pvalues", "Tax_ID", "otus") 
 
 
 # figure out if having any of the OTUs in more represented in those with chemo/rads versus none
@@ -144,9 +146,10 @@ for(i in 1:length(rownames(lesion_selected_taxa_WF))){
 }
 
 chemo_OTU_WF_pvalues_summary <- as.data.frame(cbind(
-  chemo_OTU_WF_pvalues, p.adjust(chemo_OTU_WF_pvalues, method = "bonferroni"), lesion_selected_labs_WF))
+  chemo_OTU_WF_pvalues, p.adjust(chemo_OTU_WF_pvalues, method = "bonferroni"), 
[…]

### 2016-10-14 code/20161011_Supplemental_work.R

@@ -1,0 +1,53 @@
+### Supplemental data Figures and add ons
+### Add aspects that may or may not be used in the supplement
+## Marc Sze
+
+
+###Load needed Libraries and functions
+source('code/functions.R')
+source('code/graphFunctions.R')
+
+loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson"))
+
+
+# Load needed data
+thetaCompTotal <- dissplit('data/process/followUps.final.an.unique_list.thetayc.0.03.lt.ave.dist',split=F, meta = F)
+metaI <- read.csv("results/tables/mod_metadata/metaI_final.csv", stringsAsFactors = F, header = T)
+metaF <- read.csv("results/tables/mod_metadata/metaF_final.csv", stringsAsFactors = F, header = T)
+good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", stringsAsFactors = F, header = T)
+
+#Generate distance table to be used
+difference_table_treatment <- pickDistanceValues(
+  as.character(metaF$initial), as.character(metaF$followUp), thetaCompTotal, metaF, 
+  c("fit_result", "fit_followUp", "Dx_Bin", "dx", "time", "EDRN"))
+
+# Plot the change in distance versus time
+time_graph <- ggplot(difference_table_treatment, aes(x = time, y = distance)) + 
+  geom_point(aes(color = dx), size = 4) + theme_bw() + 
+  scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
+                     breaks = c("adenoma", "cancer"), labels = c("Adenoma", "Cancer")) + 
+  coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("Time (Days)") + 
+  theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"))
+
+# Create a summary table 
+summary_statistics <- rbind(
+  c("Adenoma", mean(filter(difference_table_treatment, dx == "adenoma")[, "distance"]), 
+    sd(filter(difference_table_treatment, dx == "adenoma")[, "distance"]), 
+    mean(filter(difference_table_treatment, dx == "adenoma")[, "time"]), 
+    sd(filter(difference_table_treatment, dx == "adenoma")[, "time"])), 
[…]

### 2016-10-14 code/graphFunctions.R

@@ -6,7 +6,8 @@
 # This function creates a visual representation of initial and follow up samples important variables for a chosen model
 # These are at initial and follow up sampling.
 follow_Abundance_Fit_Graph <- function(taxaFile, labelFile, followUpFile, initialData, followData, metadata, 
-                                       jitterAmount = 0.08, n, modelOFinterest, rowOFinterest = "Disease_Free", positiveTitle = "Cancer"){
+                                       jitterAmount = 0.08, n, modelOFinterest, rowOFinterest = "Disease_Free", 
+                                       positiveTitle = "Cancer", fit = TRUE){
   
   loadLibs(c("dplyr", "ggplot2", "reshape2", "wesanderson"))
   
@@ -24,7 +25,118 @@ follow_Abundance_Fit_Graph <- function(taxaFile, labelFile, followUpFile, initia
   
   
   fit_results <- select(metadata, starts_with(modelOFinterest), EDRN, Diagnosis, fit_result, fit_followUp) %>% 
-    rename(initial = fit_result, follow_up = fit_followUp) %>% melt(id=c(paste(modelOFinterest), "EDRN", "Diagnosis")) %>% 
+    rename(initial = fit_result, follow_up = fit_followUp) %>% 
+    melt(id=c(paste(modelOFinterest), "EDRN", "Diagnosis")) %>% 
+    mutate(jitteredValues = as.numeric(value) + rep(jitter_values, 2)) %>% 
+    mutate(Disease_Free = rep(metadata[, rowOFinterest], 2))
+  
+  # Create graphs to be used
+  
+  if(fit == TRUE){
+    
+    # Adenoma OTUs graph
+    adenomaOTUs <- filter(select_otu_data, Diagnosis == "adenoma") %>% 
+      ggplot(aes(factor(sampleType, levels = c("initial", "follow_up")), 
+                 log10(jitteredValues + 1.1), group = factor(EDRN))) + 
+      geom_line(aes(color = factor(Disease_Free))) + geom_point(aes(color = factor(Disease_Free))) +  
+      facet_wrap(~variable, labeller = as_labeller(labelFile)) + coord_cartesian(ylim = c(0, 4)) + 
+      theme_bw() + ylab("Log Total Sequences") + xlab("") + ggtitle("Adenoma") + 
+      scale_colour_manual(values = "blue") +  
[…]

### 2016-10-14 exploratory/notebook/result_outline_20161011.Rmd

---
title: 'CRC Follow up: Progress Report 2 - Result Outline'
author: "Marc Sze"
date: "October 11, 2016"
output: html_document
---

```{r Load_Needed_Libraries, echo=FALSE, message=FALSE, warning=FALSE}

source('code/functions.R')
source('code/graphFunctions.R')

loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))

```

### Notes Figure 1 and 2

For the first part of the results I want to focus on how the overall community doesn't really change between initial and follow ups.  So I am undecided on having alpha diversity (but this is a simple addition - possibly to the supplement).  Main focus on this part is how thetayc change is similar in adenoma and cancer (could show Fit changes - but this might be better in the supplement to keep focus on microbiome component) and how the overall community doesn't change between initial and follow up for adenoma but does for cancer.

```{r Figure_1, echo=FALSE, message=FALSE, warning=FALSE}

# Load needed data
thetaCompTotal <- dissplit('data/process/followUps.final.an.unique_list.thetayc.0.03.lt.ave.dist',split=F, meta = F)
metaI <- read.csv("results/tables/mod_metadata/metaI_final.csv", stringsAsFactors = F, header = T)
metaF <- read.csv("results/tables/mod_metadata/metaF_final.csv", stringsAsFactors = F, header = T)
good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", stringsAsFactors = F, header = T)

difference_table_treatment <- pickDistanceValues(
  as.character(metaF$initial), as.character(metaF$followUp), thetaCompTotal, metaF, 
  c("fit_result", "fit_followUp", "Dx_Bin", "dx")) %>% 
  mutate(fit_difference = fit_result - fit_followUp)

pValueList <- list(thetaDiff = wilcox.test(distance ~ dx, data = difference_table_treatment)$p.value, 
                   fitDiff = wilcox.test(fit_difference ~ dx, data = difference_table_treatment)$p.value)

#Difference between initial and follow up for thetayc and fit broken down by adenoma and cancer
diff_adn_v_crc <- grid.arrange(
  # Difference from bacterial community structure between adenoma and cancer
  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), distance, group = 1)) + 
    geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
    scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
                       breaks = c("Adenoma", "adv Adenoma", "Cancer"), labels = c("Adenoma", "SRN", "Cancer")) + 
    coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold")) + 
    annotate("text", label = paste("P-value = ", round(pValueList[["thetaDiff"]], digits = 2)), x = 1.5, y = 0.75),

  # Difference from fit between adenoma and cancer
  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), fit_difference, group = 1)) + 
    geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
    stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
    scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
                       breaks = c("Adenoma", "adv Adenoma", "Cancer"), labels = c("Adenoma", "SRN", "Cancer")) + 
    ylab("Change in Fit from Initial to Follow Up") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          title = element_text(face="bold")) + 
    annotate("text", label = paste("P-value = ", format(round(pValueList[["fitDiff"]], digits = 7))), x = 1.5, y = 0)
)

```

**Figure 1:** Shows Thetayc difference between samples from the same individual in adenoma or cancer as well as change in Fit between the same indivdiual in adenoma or cancer.

```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE, fig.width=6, fig.height=6}

# Load in needed Adenoma Data
polyp_only_theta_init_follow_dist <- pickDistanceValues(
  as.character(c(metaF$initial[metaF$dx == "adenoma"], metaF$followUp[metaF$dx == "adenoma"])), 
  as.character(c(metaF$initial[metaF$dx == "adenoma"], metaF$followUp[metaF$dx == "adenoma"])), 
  thetaCompTotal, withMeta = FALSE)

breakDown_samples <- c(rep("initial", length(metaF$dx[metaF$dx == "adenoma"])), 
                       rep("follow_up", length(metaF$dx[metaF$dx == "adenoma"])))

set.seed(050416)
pValueList[["bdiver_Test_adn_IF"]] <- adonis(as.dist(polyp_only_theta_init_follow_dist) ~ factor(breakDown_samples))

set.seed(050416)
thetayc.mds.List <- list(bdiver_Test_adn_IF = metaMDS(as.dist(polyp_only_theta_init_follow_dist), trace = 0) %>% 
    scores() %>% as.data.frame() %>%  mutate(samples = factor(breakDown_samples)))

# Load in needed CRC data
crc_only_theta_init_follow_dist <- pickDistanceValues(
  as.character(c(metaF$initial[metaF$dx != "adenoma"], metaF$followUp[metaF$dx != "adenoma"])), 
  as.character(c(metaF$initial[metaF$dx != "adenoma"], metaF$followUp[metaF$dx != "adenoma"])), 
  thetaCompTotal, withMeta = FALSE)

breakDown_samples <- c(rep("initial", length(metaF$dx[metaF$dx != "adenoma"])), 
                       rep("follow_up", length(metaF$dx[metaF$dx != "adenoma"])))

set.seed(050416)
pValueList[["bdiver_Test_crc_IF"]] <- adonis(as.dist(crc_only_theta_init_follow_dist) ~ factor(breakDown_samples))

set.seed(050416)
thetayc.mds.List[["bdiver_Test_crc_IF"]] <- metaMDS(as.dist(crc_only_theta_init_follow_dist), trace = 0) %>% 
  scores() %>% as.data.frame() %>% 
  mutate(samples = factor(breakDown_samples))

# Plot the data
grid.arrange(

  ggplot(data = thetayc.mds.List[["bdiver_Test_adn_IF"]], aes(x=NMDS1, y=NMDS2)) + geom_point(aes(color=samples)) + theme_bw() + coord_equal() + ggtitle("Adenoma") + 
  stat_ellipse(aes(group = samples, color = samples, fill = samples), alpha = 0.25, geom = "polygon") + 
  annotate("text", label = paste("PERMANOVA = ", round(pValueList[["bdiver_Test_adn_IF"]]$aov.tab$`Pr(>F)`[1], 
                                                       digits = 3)), x = 0.20, y = -0.50, size = 2.5), 
  ggplot(data = thetayc.mds.List[["bdiver_Test_crc_IF"]], aes(x=NMDS1, y=NMDS2)) + 
  geom_point(aes(color=samples)) + theme_bw() + coord_equal() + ggtitle("Cancer") + 
  stat_ellipse(aes(group = samples, color = samples, fill = samples), alpha = 0.25, geom = "polygon") + 
    annotate("text", label = paste("PERMANOVA = ", round(pValueList[["bdiver_Test_crc_IF"]]$aov.tab$`Pr(>F)`[1], 
                                                       digits = 3)), x = 0.25, y = -0.50, size = 2.5), ncol = 1, nrow = 2 
)

```

**Figure 2:** Shows overall bacterial community composition before and after polyp or Cancer removal.  THere is a significant change in the community for cancer but not for adenoma

### Notes for Figure 3 and Figure 4

Moving on from here I want to then talk about the previous Lesion model (Adenoma + CRC) that used Fit and the microbiome.  I think Figure 3 can be the ROC curve of the training model with and without Fit. It could possibly have in the legend the AUC on the test set.  For Figure 4 I would want to show the breakdown by Adenoma and CRC with respect to initial and follow up samples.

```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE}

# Load in necessary Data
testSetAUC <- read.csv("results/tables/testSetAUC.csv", stringsAsFactors = F, header = T)
rownames(testSetAUC) <- testSetAUC$V1
colnames(testSetAUC) <- c("Model", "AUC")
wFit_ROC_table <- read.csv("results/tables/ROCCurve_sens_spec.csv")
woFit_ROC_table <- read.csv("results/tables/ROCCurve_noFit_sens_spec.csv")

load("exploratory/wFit_Test_ROC.RData")
wFit_lesion <- lesion_combined_ROC
rm(lesion_combined_ROC)

load("exploratory/woFit_Test_ROC.RData")
woFit_lesion <- lesion_combined_ROC
rm(lesion_combined_ROC)

#Create data table to be graphed

graphData <- rbind(
  mutate(wFit_ROC_table, FIT = rep("wFit", length(rownames(wFit_ROC_table)))) %>% 
    filter(model == "lesionALL"), 
  mutate(woFit_ROC_table, FIT = rep("woFit", length(rownames(woFit_ROC_table)))) %>% 
    filter(model == "lesionALL")
)

# Get Pvalue difference between two ROCs

ggplot(graphData, aes(sensitivities, specificities)) + 
  geom_line(aes(group = FIT, color = FIT), size = 1.5) + scale_x_continuous(trans = "reverse") + 
  theme_bw() + xlab("Sensitivity") + ylab("Specificity") + geom_abline(intercept = 1, linetype = 2) + 
  scale_color_discrete(name = "Lesion Model", breaks = c("wFit", "woFit"), 
                       labels = c(paste("with Fit\n", "(Test Set AUC = ", 
                                        round(testSetAUC['withFit', 'AUC'], digits = 3), ")", sep=""), 
                                  paste("without Fit\n", "(Test Set AUC = ", 
                                        round(testSetAUC['woFit', 'AUC'], digits = 3), ")", sep=""))) + 
  theme(axis.title = element_text(face = "bold"), 
        legend.position = c(0.75, 0.4), legend.title = element_text(face="bold")) + 
  annotate("text", label = paste("P-value = ", round(roc.test(wFit_lesion, woFit_lesion)$p.value, 
                                                     digits = 3)), x = 0.9, y = 0.9, size = 4)

```

**Figure 3:** ROC curve of the lesion model with and without Fit.  Overall there is no significant difference between the two models.  This was built using a total of ```r length(rownames(metaI))``` individuals which was then tested on ```r length(rownames(metaF))``` which had initial and follow up data available.

```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=12}

# Read in needed data
withFit_cutoffs <- read.csv("results/tables/withFIT.cutoffs.csv", header = T, stringsAsFactors = F)
woFit_cutoffs <- read.csv("results/tables/noFIT.cutoffs.csv", header = T, stringsAsFactors = F)

withFit_data <- read.csv("results/tables/withFIT.models.datatable.csv", header = T, stringsAsFactors = F)
woFit_data <- read.csv("results/tables/noFIT.models.datatable.csv", header = T, stringsAsFactors = F)

pvalue_data <- read.csv("results/tables/lesion_WF_IF_pvalue_summary.csv", header = T, stringsAsFactors = F)
rownames(pvalue_data) <- pvalue_data$Disease

# Graph the Adenoma data only

grid.arrange(
  # Graph the adenoma ALL data only
  filter(withFit_data, diagnosis == "adenoma" & dataset == "All" & model == "lesion") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive, group = factor(EDRN))) + 
    geom_point(aes(color=detailed_diagnosis)) + geom_line(aes(color = detailed_diagnosis)) + 
    scale_color_manual(name = "Polyp Type", values = c("cyan", "blue"), 
                       breaks = c("Adenoma", "adv Adenoma"), labels = c("Adenoma", "SRN")) + 
    coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(withFit_cutoffs, model == "lesion", dataset == "All"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("Adenomas (With Fit)") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          legend.position = "none", title = element_text(face="bold")),

  # Graph adenoma select data only
  filter(woFit_data, diagnosis == "adenoma" & dataset == "All" & model == "lesion") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive, group = factor(EDRN))) + 
    geom_point(aes(color=detailed_diagnosis)) + geom_line(aes(color = detailed_diagnosis)) + 
    scale_color_manual(name = "Polyp Type", values = c("cyan", "blue"), 
                       breaks = c("Adenoma", "adv Adenoma"), labels = c("Adenoma", "SRN")) + 
    coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(woFit_cutoffs, model == "lesion", dataset == "All"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("Adenomas (Without Fit)") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          legend.position = c(-1, 0.15), title = element_text(face="bold")),

  # Graph the cancer ALL data only
  filter(withFit_data, diagnosis != "adenoma" & dataset == "All" & model == "lesion") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive, group = factor(EDRN))) + 
    geom_point(aes(color=factor(diseaseFree, levels = c("n", "y", "unknown")))) + 
    geom_line(linetype = 2, alpha = 0.3) + 
    scale_color_manual(name = "Cancer\nFree", values = wes_palette("GrandBudapest")) + 
    coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(withFit_cutoffs, model == "lesion", dataset == "All"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("Cancer (With Fit)") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          legend.position = "none", title = element_text(face="bold")),

  # Graph cancer select data only
  filter(woFit_data,  diagnosis != "adenoma" & dataset == "All" & model == "lesion") %>%
    ggplot(aes(factor(time_point, levels = c("initial", "followup")), positive, group = factor(EDRN))) + 
    geom_point(aes(color=factor(diseaseFree, levels = c("n", "y", "unknown")))) + 
    geom_line(linetype = 2, alpha = 0.3) + 
    scale_color_manual(name = "Cancer Free", label = c("No", "Yes", "Unknown"), values = wes_palette("GrandBudapest")) + 
    coord_cartesian(ylim = c(0, 1)) + 
    geom_hline(data = filter(woFit_cutoffs, model == "lesion", dataset == "All"), 
               aes(yintercept = cutoff), linetype = 2) + 
    ggtitle("Cancer (Without Fit)") + ylab("Postive Probability") + xlab("") + theme_bw() + 
    theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
          legend.position = c(-1, 0.15), title = element_text(face="bold"))
)

```

**Figure 4:** Statistics utilized a Fisher exact Test with bonferroni correction to test if there was a signifcant drop in positive calls for adenomas, advanced adenomas (SRN), or cancer.  For the Fit model the adjusted pvalues were ```r round(pvalue_data['Adenoma', 'WF_padjust'], digits = 3)```, ```r round(pvalue_data['advAdenoma', 'WF_padjust'], digits = 3)```, and ```r format(pvalue_data['Cancer', 'WF_padjust'], scientific = TRUE, digits = 3)``` for adenoma, SRN, and cancer. Conversely, the adjusted pvalues for the model without Fit was ```r round(pvalue_data['Adenoma', 'WoF_padjust'], digits = 3)```, ```r round(pvalue_data['advAdenoma', 'WoF_padjust'], digits = 3)```, and ```r format(pvalue_data['Cancer', 'WoF_padjust'], scientific = TRUE, digits = 3)``` for adenoma, SRN, and cancer.

### Notes for Figure 5 and Supplement tables

Here I want to show that there are one, two, and two significant OTUs that have a significant decrease in proportion between intiial and follow up in lesion w/fit, lesion w/o fit, and the 4 common CRC bugs.  Figure 5 can focus on the two specific OTUs that were significant in either all three (Porphymonas) or only two of the three models (Parvimonas).  I can have the statistic test tables for lesion with fit, lesion without fit, and the four common literature OTUs in the supplemental as well I also want to show that it is not due to anything but the surgery itself (supplement table).  So there is no diference in proportion of those with and without Chemo with respect to these OTUs.

```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=12}

# Read in needed data

good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", stringsAsFactors = F, header = T)
shared <- read.delim('data/process/followUps.final.an.unique_list.0.03.subsample.0.03.filter.shared', 
                     header=T, sep='\t')

tax <- read.delim('data/process/followUps.final.an.unique_list.0.03.cons.taxonomy', 
                  sep='\t', header=T, row.names=1)

wfit_pvalue_summary <- read.csv("results/tables/lesion_OTU_WF_Pvalue_summary.csv", header = T, stringsAsFactors = F)
wofit_pvalue_summary <- read.csv("results/tables/lesion_OTU_WoF_Pvalue_summary.csv", header = T, stringsAsFactors = F)
crcSp_pvalue_summary <- read.csv("results/tables/lesion_OTU_crcSp_Pvalue_summary.csv", header = T, stringsAsFactors = F)

# Create needed initial and follow up data tables
initial <- inner_join(good_metaf, shared, by = c("initial" = "Group")) %>% 
  select(SRNlesion, lesion, fit_result, contains("Otu0"))

followups <- inner_join(good_metaf, shared, by = c("followUp" = "Group")) %>% 
  select(lesionf, fit_followUp, contains("Otu0")) %>% rename(SRNlesion = lesionf) %>% rename(fit_result = fit_followUp)

followups <- cbind(good_metaf$lesionf, followups)
colnames(followups)[1] <- c("lesion")

# Convert taxa table to a data frame with columns for each taxonomic division
tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
rownames(tax_df) <- rownames(tax)
colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
rm(tax)
# This uses the R regex to remove numerals and the brackets the \\ is used to escape specific function components

# create needed labels for Boruta picked important variables for each model

lesion_selected_taxa <- tax_df[filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05)[, 'otus'], ]
lesion_selected_labs <- createTaxaLabeller(lesion_selected_taxa)

cancer_positive <- as.data.frame(cbind(rep(as.character(good_metaf$Disease_Free), 2), rep(good_metaf$EDRN, 2))) %>% 
  rename(Disease_Free = V1, EDRN = V2)

lesion_model_impf_graphs <- follow_Abundance_Fit_Graph(lesion_selected_taxa, lesion_selected_labs, 
                                                       cancer_positive, initial, followups, select(good_metaf, -lesionf), 
                                                       0.08, 67, "lesion", fit = FALSE)

grid.arrange(
  lesion_model_impf_graphs$adenoma_OTUs, lesion_model_impf_graphs$cancer_OTUs
)

```

**Figure 5:** Porphyromonas (```r paste(filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05 & Tax_ID == "Porphyromonas")[, 'otus'])```) and Parvimonas (```r paste(filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05 & Tax_ID == "Parvimonas")[, 'otus'])```) Log 10 total subsampled sequences for initial and follow split by whether the individual had an adenoma or cancer.  There was a significant decrease in the proportion of individuals in the cancer group who had either Porphyromonas (```r round(as.numeric(filter(wfit_pvalue_summary, OTU_WF_adjust_pvalues < 0.05 & Tax_ID == "Porphyromonas")[, 'OTU_WF_adjust_pvalues']), digits = 3)```) in the with Fit model.  In the without Fit model both Porphyromonas (```r round(as.numeric(filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05 & Tax_ID == "Porphyromonas")[, 'OTU_WoF_adjust_pvalues']), digits = 3)```) and Parvimonas (```r round(as.numeric(filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05 & Tax_ID == "Parvimonas")[, 'OTU_WoF_adjust_pvalues']), digits = 3)```) had a significant decrease in proportion for the cancer group.

**Supplemental Table 1:** List of Important OTU variables in the with Fit model

```{r Supplemental_Table1, echo=FALSE, message=FALSE, warning=FALSE}

supplemental_table1 <- wfit_pvalue_summary

colnames(supplemental_table1) <- c("P-values", "Adjusted P-values", "Lowest Taxonomic ID", "OTU")

kable(supplemental_table1, format = "markdown", align = 'c', padding = 2)
```

**Supplemental Table 2:** List of Important OTU variables in the without Fit model

```{r Supplemental_Table2, echo=FALSE, message=FALSE, warning=FALSE}

supplemental_table2 <- wofit_pvalue_summary

colnames(supplemental_table2) <- c("P-values", "Adjusted P-values", "Lowest Taxonomic ID", "OTU")

kable(supplemental_table2, format = "markdown", align = 'c', padding = 2)

```

**Supplemental Table 3:** List of Important OTU variables in the with Fit model and whether there is an increased proportion based on surgery alone versus surgery and chemo with or without radiation in those with cancer.

```{r Supplemental_Table3, echo=FALSE, message=FALSE, warning=FALSE}

supplemental_table3 <- read.csv("results/tables/chemo_lesion_OTU_WF_Pvalue_summary.csv", stringsAsFactors = F, header = T)

colnames(supplemental_table3) <- c("P-values", "Adjusted P-values", "Lowest Taxonomic ID", "OTU")

kable(supplemental_table3, format = "markdown", align = 'c', padding = 2)

```

**Supplemental Table 4:** List of Important OTU variables in the without Fit model and whether there is an increased proportion based on surgery alone versus surgery and chemo with or without radiation in those with cancer.

```{r Supplemental_Table4, echo=FALSE, message=FALSE, warning=FALSE}

supplemental_table4 <- read.csv("results/tables/chemo_lesion_OTU_WoF_Pvalue_summary.csv", stringsAsFactors = F, header = T)

colnames(supplemental_table4) <- c("P-values", "Adjusted P-values", "Lowest Taxonomic ID", "OTU")

kable(supplemental_table4, format = "markdown", align = 'c', padding = 2)

```

### Final notes and Summary Statistics

```{r create_count_data, echo=FALSE, message=FALSE, warning=FALSE}

selectData <- inner_join(shared, good_metaf, by = c("Group" = "initial")) %>% select(Diagnosis, one_of(filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05)[, 'otus']))

testTable <- matrix(nrow = 2, ncol = 2, dimnames = list(rows = c("Cancer", "Adenoma"), cols = c("Yes", "No")))

testTable["Cancer", "Yes"] <- length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 > 0 & Diagnosis != "adenoma"))) + length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 == 0 & Diagnosis != "adenoma"))) + length(rownames(filter(selectData, Otu000126 == 0 & Otu000397 > 0 & Diagnosis != "adenoma")))

testTable["Cancer", "No"] <- length(rownames(filter(selectData, Diagnosis != "adenoma"))) - testTable["Cancer", "Yes"]

testTable["Adenoma", "Yes"] <- length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 > 0 & Diagnosis == "adenoma"))) + length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 == 0 & Diagnosis == "adenoma"))) + length(rownames(filter(selectData, Otu000126 == 0 & Otu000397 > 0 & Diagnosis == "adenoma")))

testTable["Adenoma", "No"] <- length(rownames(filter(selectData, Diagnosis == "adenoma"))) - testTable["Adenoma", "Yes"]

crc2bugs_pvalue <- fisher.test(testTable)$p.value

```

A total of ```r round(((length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 > 0 & Diagnosis != "adenoma"))) + length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 == 0 & Diagnosis != "adenoma"))) + length(rownames(filter(selectData, Otu000126 == 0 & Otu000397 > 0 & Diagnosis != "adenoma")))) / length(rownames(filter(selectData, Diagnosis != "adenoma")))) * 100, digits = 2)```% of all cancer cases in the test set had either one of Porphyromonas or Parvimonas or both detected in their intial stool sample.  In contrast for all the adenoma cases this was```r round(((length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 > 0 & Diagnosis == "adenoma"))) + length(rownames(filter(selectData, Otu000126 > 0 & Otu000397 == 0 & Diagnosis == "adenoma"))) + length(rownames(filter(selectData, Otu000126 == 0 & Otu000397 > 0 & Diagnosis == "adenoma")))) / length(rownames(filter(selectData, Diagnosis == "adenoma")))) * 100, digits = 2)```%.  This increase in proportion is statistically significant at p-value = ```r format(crc2bugs_pvalue, scientific = TRUE, digits = 3)```.

Finally, there does not seem to be a difference based on length of time of sampling, with respect to initial versus follow up thetayc.

```{r supplemental_Figure, echo=FALSE, message=FALSE, warning=FALSE}

source('20161011_Supplemental_work.R')
rownames(pvalues_summary) <- pvalues_summary[, "test_values"]

time_graph

```
**Figure S1:** Thetayc distance between initial and follow up versus time to follow up sample. There is a significant difference in time of second sampling between the adenoma and cancer group p-value = ```r format(as.numeric(pvalues_summary["time", "uncorrected_p_value"]), scientific = TRUE, digits = 3)```.

### 2016-10-20 code/20161004_Run_base_testModels_wFit.R
add and update code to include abund stat analysis

@@ -53,9 +53,9 @@ rf_opt_NameList <- list(SRNlesion_rf_opt = orig_rf_opt[["SRNlesion"]],
                         SRNlesion_selected_rf_opt = selected_rf_opt[["SRNlesion"]], 
                         lesion_rf_opt= orig_rf_opt[["lesion"]], lesion_selected_rf_opt = selected_rf_opt[["lesion"]])
 
-initial_predictions <- lapply(rf_opt_NameList, function(x) predict(x, initial, type='prob'))
+initial_predictions <- lapply(rf_opt_NameList, function(x) predict(x, newdata = initial, type='prob'))
 
-followup_predictions <- lapply(rf_opt_NameList, function(x) predict(x, followups, type='prob'))
+followup_predictions <- lapply(rf_opt_NameList, function(x) predict(x, newdata = followups, type='prob'))
 
 
 #Create ROC data for lesion model
@@ -65,6 +65,10 @@ lesion_combined_ROC <- roc(c(initial$lesion, followups$lesion) ~
 
 write.csv(matrix(c("withFit", lesion_combined_ROC$auc), ncol = 2), "results/tables/testSetAUC.csv", row.names = F)
 
+variableList <- c("sensitivities", "specificities")
+sens_specif_table <- makeSensSpecTable(list(lesion = lesion_combined_ROC), variableList, modelList = "withFitTest")
+
+write.csv(sens_specif_table, "results/tables/wFit_test_ROCCurve_sens_spec.csv")
 
 # Join good_metaf table with metaF table
 combined_metaData <- inner_join(good_metaf, metaF, by = "EDRN")

### 2016-10-20 code/20161018_Model_OTU_Abundance_Tests.R

@@ -1,0 +1,72 @@
+### Test amount increase/decrease of each model OTU
+### Does the overall amount increase/decrease even if the proportion positive does not
+## Marc Sze
+
+
+# Load in needed functions and libraries
+
+source('code/functions.R')
+source('code/graphFunctions.R')
+
+loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))
+
+# Read in needed data
+
+good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", stringsAsFactors = F, header = T)
+shared <- read.delim('data/process/followUps.final.an.unique_list.0.03.subsample.0.03.filter.shared', 
+                     header=T, sep='\t')
+
+tax <- read.delim('data/process/followUps.final.an.unique_list.0.03.cons.taxonomy', 
+                  sep='\t', header=T, row.names=1)
+
+
+wfit_RFOpt_vars <- read.csv("results/tables/lesion_RFOpt_Imp_Vars.csv", stringsAsFactors = F, header = T) %>% 
+  rename(vars = x)
+
+wofit_RFOpt_vars <- read.csv("results/tables/lesion_RFOpt_NOFIT_Imp_Vars.csv", stringsAsFactors = F, header = T) %>% 
+  rename(vars = x)
+
+# Convert taxa table to a data frame with columns for each taxonomic division
+tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
+rownames(tax_df) <- rownames(tax)
+colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
+tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
+rm(tax)
+# This uses the R regex to remove numerals and the brackets the \\ is used to escape specific function components
+
+# Create needed initial and follow up data tables for with fit model and test statistically
+initial_lesion_wfit <- inner_join(good_metaf, shared, by = c("initial" = "Group")) %>% 
+  select(lesion, one_of(filter(wfit_RFOpt_vars, vars != "fit_result")[, "vars"]))
+
[…]

### 2016-10-20 exploratory/notebook/result_outline_20161011.Rmd

@@
           title = element_text(face="bold")) + 
-    annotate("text", label = paste("P-value = ", round(pValueList[["thetaDiff"]], digits = 2)), x = 1.5, y = 0.75),
+    annotate("text", label = paste("P-value = ", round(pValueList[["thetaDiff"]], digits = 2)), x = 1.5, y = 0.7),
 
   # Difference from fit between adenoma and cancer
-  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), fit_difference, group = 1)) + 
+  ggplot(difference_table_treatment, aes(factor(dx, levels = c("adenoma", "cancer")), fit_difference*-1, group = 1)) + 
     geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
@@
                        breaks = c("Adenoma", "adv Adenoma", "Cancer"), labels = c("Adenoma", "SRN", "Cancer")) + 
-    ylab("Change in Fit from Initial to Follow Up") + xlab("") + theme_bw() + 
+    ylab("Change in Fit from Follow up to Initial") + xlab("") + theme_bw() + 
     theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"), 
           title = element_text(face="bold")) + 
-    annotate("text", label = paste("P-value = ", format(round(pValueList[["fitDiff"]], digits = 7))), x = 1.5, y = 0)
+    annotate("text", label = paste("P-value = ", format(round(pValueList[["fitDiff"]], digits = 7))), x = 1.5, y = 100)
 )
@@
 woFit_ROC_table <- read.csv("results/tables/ROCCurve_noFit_sens_spec.csv")
+wFit_test_ROC_table <- read.csv("results/tables/wFit_test_ROCCurve_sens_spec.csv")
+woFit_test_ROC_table <- read.csv("results/tables/woFit_test_ROCCurve_sens_spec.csv")
 
@@
 
+load("exploratory/wFit_Test_ROC.RData")
+test_wFit_lesion <- lesion_combined_ROC
+rm(lesion_combined_ROC)
+
+load("exploratory/woFit_Test_ROC.RData")
+test_woFit_lesion <- lesion_combined_ROC
+rm(lesion_combined_ROC)
+
 #Create data table to be graphed
@@
   mutate(woFit_ROC_table, FIT = rep("woFit", length(rownames(woFit_ROC_table)))) %>% 
-    filter(model == "lesionALL")
+    filter(model == "lesionALL"), 
+  mutate(wFit_test_ROC_table, FIT = rep("wFitTest", length(rownames(wFit_test_ROC_table)))), 
+  mutate(woFit_test_ROC_table, FIT = rep("woFitTest", length(rownames(woFit_test_ROC_table))))
 )
 
-# Get Pvalue difference between two ROCs
+#Get the pvalues
+
+corr_pvalue_ROC_table <- getROCPvalue(rocNameList = list(
+  wfit = wFit_lesion, wofit = woFit_lesion, test_wfit = test_wFit_lesion, test_wofit = test_woFit_lesion), 
+  modelList = c("wfit", "wofit", "test_wfit", "test_wofit"), 4, multi = F)
+
+# Get Pvalue difference between two ROCs (Plot the test set data as well)
 
@@
   theme_bw() + xlab("Sensitivity") + ylab("Specificity") + geom_abline(intercept = 1, linetype = 2) + 
-  scale_color_discrete(name = "Lesion Model", breaks = c("wFit", "woFit"), 
-                       labels = c(paste("with Fit\n", "(Test Set AUC = ", 
-                                        round(testSetAUC['withFit', 'AUC'], digits = 3), ")", sep=""), 
-                                  paste("without Fit\n", "(Test Set AUC = ", 
-                                        round(testSetAUC['woFit', 'AUC'], digits = 3), ")", sep=""))) + 
+  scale_color_discrete(name = "Lesion Model", breaks = c("wFit", "woFit", "wFitTest", "woFitTest"), 
+                       labels = c(paste("with Fit (Train Set)\n", sep = ""), 
+                                  paste("without Fit (Train Set)\n", sep=""), 
+                                  paste("with Fit (Test Set)\n", sep=""), 
+                                  paste("without Fit (Test Set)\n", sep=""))) + 
   theme(axis.title = element_text(face = "bold"), 
-        legend.position = c(0.75, 0.4), legend.title = element_text(face="bold")) + 
-  annotate("text", label = paste("P-value = ", round(roc.test(wFit_lesion, woFit_lesion)$p.value, 
-                                                     digits = 3)), x = 0.9, y = 0.9, size = 4)
+        legend.position = c(0.75, 0.4), legend.title = element_text(face="bold"))
 
@@
 
-**Figure 3:** ROC curve of the lesion model with and without Fit.
-Overall there is no significant difference between the two models.
+**Figure 3:** ROC curve of the lesion model with and without Fit for the training and test set.
+Overall there was no significant difference between with and without Fit models for the training set (```r paste("P-value = ", round(roc.test(wFit_lesion, woFit_lesion)$p.value, digits = 3))```).
 This was built using a total of ```r length(rownames(metaI))``` individuals which was then tested on ```r length(rownames(metaF))``` which had initial and follow up data available.
@@
 
-lesion_selected_taxa <- tax_df[filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05)[, 'otus'], ]
+lesion_selected_taxa <- tax_df[filter(wofit_pvalue_summary, OTU_WoF_adjust_pvalues < 0.05 | otus == "Otu000205")[, 'otus'], ]
+lesion_selected_taxa <- lesion_selected_taxa[c("Otu000126", "Otu000397", "Otu000205"), ]
 lesion_selected_labs <- createTaxaLabeller(lesion_selected_taxa)
 
-cancer_positive <- as.data.frame(cbind(rep(as.character(good_metaf$Disease_Free), 2), rep(good_metaf$EDRN, 2))) %>% 
+cancer_positive <- as.data.frame(cbind(rep(as.character(good_metaf$Disease_Free), 3), rep(good_metaf$EDRN, 3))) %>% 
   rename(Disease_Free = V1, EDRN = V2)

### 2016-12-02 code/Run_Get_Imp_OTUs_noFIT.R
add code scripts that will be used

@@ -1,0 +1,252 @@
+### Try to identify the important OTUs 
+### Specifically try and identify which OTUs are involved
+### with the decrease in positive probability between intial and follow up
+## Marc Sze
+
+
+###Load needed Libraries and functions
+source('code/functions.R')
+source('code/graphFunctions.R')
+
+loadLibs(c("randomForest", "dplyr", "ggplot2", "reshape2", 
+  "gridExtra", "scales", "wesanderson", "caret", "doMC"))
+
+# Read in data tables
+good_metaf <- read.csv(
+  "results/tables/mod_metadata/good_metaf_final.csv", 
+  stringsAsFactors = F, header = T) %>% 
+  mutate(lesion_follow = ifelse(Disease_Free == "n", 1, 0))
+
+shared <- read.delim('data/process/final.0.03.subsample.shared', 
+  header=T, sep='\t') %>% select(-label, -numOtus)
+
+
+#################################################################################
+#                                                                               #
+#                                                                               #
+#               Data Clean up for better and faster modeling                    #
+#                                                                               #
+#################################################################################
+
+#Create an insync version to merge with filtered_shared
+good_metaf_select <- data_frame(
+  sample = c(good_metaf$initial, good_metaf$followUp)) %>% 
+  mutate(lesion = c(good_metaf$lesion, good_metaf$lesion_follow), 
+         Hx_Prev = rep(good_metaf$Hx_Prev, 2), 
+         Hx_Fam_CRC = rep(good_metaf$Hx_Fam_CRC, 2), 
+         White = rep(good_metaf$White, 2), 
+         BMI = rep(good_metaf$BMI, 2), 
+         Age = rep(good_metaf$Age, 2), 
+         Gender = rep(good_metaf$Gender, 2))
+
+test_data <- inner_join(good_metaf_select, shared, 
+  by = c("sample" = "Group"))
+
+rm(shared)
+
+# Filter and use only specific data
+
+test_data <- test_data %>% 
+  select(sample, lesion, Hx_Prev, Hx_Fam_CRC, White, 
+    BMI, Age, Gender, contains("Otu0")) %>% 
+  mutate(lesion = factor(lesion), Gender = factor(Gender)) %>% 
+  mutate(Hx_Prev = factor(Hx_Prev), Hx_Fam_CRC = factor(Hx_Fam_CRC), 
+    White = factor(White))
+
+#Filter out rows that are not all complete
+
+test_data <- test_data[complete.cases(test_data), ]
+
+# Need to create dummy variables for those that are factor classes
[…]

### 2016-12-02 code/qsubmission.sh

@@ -39,6 +39,6 @@ run_96_RF.pbs run_97_RF.pbs run_98_RF.pbs run_99_RF.pbs run_9_RF.pbs"
 for s in ${submission_twelve}
 do
 	#qsub $s
-	qsub -W depend=afterok:1194181 $s
+	qsub -W depend=afterok:1195789 $s
 	sleep 1
 done

### 2016-12-02 submission/manuscript_outline_20161024.Rmd
first draft manuscript

---
title: ''
csl: mbio.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  html_document: default
geometry: margin=1.0in
bibliography: null
---

# Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('code/functions.R')
source('code/graphFunctions.R')

loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)
opts_chunk$set("message" = FALSE)

```

```{r rdata_tables}
# Extra data and meta data needed
good_metaf <- read.csv('results/tables/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
metaI <- read.csv('results/tables/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
total_sub_Seqs <- read.delim('data/process/final.0.03.subsample.shared', header = T) %>% 
  select(contains("Otu0")) %>% rowSums()

# initial microbiome and FIT description data 
change_summary <- read.csv('results/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
alpha_summary <- read.csv('results/tables/alpha_table_summary.csv', header = T, row.names = 1)
beta_summary <- read.csv('results/tables/beta_diver_summary.csv', header = T, row.names = 1)
time_pvalues <- read.csv("results/tables/time_pvalues.csv", header = T, row.names = 1)

# training of model data
model_train_summary <- read.csv('results/tables/ROC_model_summary.csv', header = T)
model_test_AUC_summary <- read.csv('results/tables/auc_summary.csv', header = T, row.names = 1)
imp_model_vars <- read.csv('results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
rf_otu_tax <- read.csv('results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
rf_lowest_ID <- createTaxaLabeller(rf_otu_tax)

# model prediction and adjustment performance
IF_model_wilcox_paired_summary <- read.csv('results/tables/IF_model_wicox_paired_pvalue_summary.csv', header = T, row.names = 1)
confusion_summary <- read.csv('results/tables/confusion_summary.csv', header = T, row.names = 1)
follow_up_prob_summary <- read.csv('results/tables/follow_up_probability_summary.csv', header = T)
followup_cm <- read.csv('results/tables/followup_confusion_matrix.csv', header = T, row.names = 1)
initial_cm <- read.csv('results/tables/initial_confusion_matrix.csv', header = T, row.names = 1)
roc_pvalues <- read.csv('results/tables/roc_pvalue_summary.csv', header = T, row.names = 1)

# Predictive of both initial and follow up and in lesion model
pvalue_common_imp_vars <- read.csv('results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)

```

\vspace{25mm}

\begin{center}
Running Title: Human Microbiome and Colorectal Cancer

\vspace{10mm}

Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss${^\dagger}$

\vspace{10mm}

Contributions: All authors were involved in the conception and design of the study. MAS analyzed the data. NTB processed samples and analyzed the data. All authors interpreted the data. MAS and PDS wrote the manuscript. All authors reviewed and revised the manuscript. All authors read and approved the final manuscript.

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage
\linenumbers

## Abstract

Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.  Previous research suggests that the bacterial microbiome can be used as a biomarker for colorectal cancer (CRC).  These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.  In this current study we attempt to combine specific physical characteristics, patient history, Fecal Immunochemical Test (FIT), and the bacterial microbiome to build the best model for classifying those with and without lesion (adenoma or carcinoma).  From this improved model we applied it to a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment. Carcinoma dependent differences in initial and follow up samples were identified for the overall bacterial community composition (P-value = ```r paste(beta_summary['Carcinoma', 'PERMANOVA'])```).  The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.  There was a significant decrease in probabliity of lesion from the model after treatment for both the adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```) group.  The top 5 most important operational taxonomic units for this model were ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.  For prediction on samples with follow up the initial sensitivity was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) and follow up sensitivity was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).  Overall, this model can accurately predict and subsequently respond to treatment by lowering the probability of lesion.

\newpage

### Importance

This is one of the first studies to investigate within humans what happens to the bacterial microbiome before and after colorectal cancer treatment.  Specifically, it aims to assess how the random forest machine learning algorithm built models respond to treatment and adjust their positive probability calls of whether the individual has an adenoma or carcinoma.

\newpage

## Introduction

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age (**insert citation**).  Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening (**insert citation**).  However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year (**insert citation**).  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance (**insert citation**). The vast majority of CRCs are not inherited and the exact etiology to disease is not well worked out (**insert citation**).

Recent reports have found that the microbiome between those with and without CRC are different and that bacteia associated with CRC tissue is different that that found on the normal mucosa (**Insert citations**).  Further, there is evidence to suggest that more inflammatory mouth-associated bacteria are more prevalent in the guts of those with CRC (**insert citation**).  This has led to the hypothesis that these bacteria replace traditional residents that produce metabolites such as butyrate (**insert citation**) and lead to a more inflammatory state (**insert citation**).  This inflammatory state allows for more of these types of bacteria to populate the gut and leads to neoplasia and eventually CRC (**insert citation**).

Building upon these findings a number of groups, including our own, have investigated using the bacterial microbiome as a potential biomarker for adenoma and CRC detection (**insert citations**).  These collectively show that a reasonable area under the curve (AUC) can be obtained when using the stool bacterial microbiome to classify disease state (**insert citation**).  Further, a number of groups have extended this observation to show that using either specific bacterial species or the bacterial microbiome as a whole paired with the Fecal Immunochemical Test (FIT) can either increase the AUC of the model (**insert citations**).

In this study we further refine the random forest model by including more background information on the indivdiuals within it. We use a rigourous and accepted approach for machine learning training and testing to validate this model (**insert citation**).  We then apply it to individuals in which we have samples before and after treatment to assess how well the model adjusts their probability of having an adenoma or carcinoma.  Finally, we investigate which specific Operational Taxonomic Units (OTUs) are most affected by treatment.  Within this study lesion refers to both adenoma and carcinoma.

\newpage

## Results

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics comparing the initial to the follow up samples there was no difference between the adenoma and carcinoma group (P-value = ```r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)```) [Figure 1a]. There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = ```r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)```) [Figure 1b].  Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = ```r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)```) but not for the adenoma group (P-value = ```r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)```) [Figure 2]. When all follow up samples were compared to each other there was no significant overall difference between them (P-value = ```r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)```).  There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, and evenness after correction for multiple comparisons [Table S1].  Time of follow up sample from initial sampling did not have an effect on our prediction data set (uncorrected P-value = ```r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)```).

***Outcome of Model Training***
The range of the AUC for model training ranged from a minimum of ```r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)``` to a maximum of ```r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)```.  To be conservative the model chosen for prediction of disease in the follow up samples had an AUC in the middle of all the ```r length(rownames(model_train_summary))``` runs which was ```r format(model_test_AUC_summary["middle", "ROC_cv"], 
digits = 3)```.  Interestingly, the worst AUC model from training performed the best on it's respective 80/20 split test data [Figure 3].  In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = ```r format(model_test_AUC_summary["best", "AUC"], digits = 3)```, middle training model test set AUC = ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```, worse training model test set AUC = ```r format(model_test_AUC_summary["worse", "AUC"], digits = 3)```).  That is to say it had the smallest change in AUC in comparison to the other minimum and maximum AUC trained models.  There was no significant difference between the AUC of the best and middle training models (P-value = ```r format(roc_pvalues["BH_adj", "best_vs_middle"], digits = 3)```).  There was a significant difference between the worse training model and both the best training model (P-value = ```r format(roc_pvalues["BH_adj", "best_vs_worse"], digits = 3)```) and middle training model (P-value = ```r format(roc_pvalues["BH_adj", "middle_vs_worse"], digits = 3)```).

***Most Important Variables to the Model***
Overall, there were a total of ```r format(length(rownames(imp_model_vars)))``` variables identified as being present in more than 50% of the training models [Table S2]. The top 5 most important bacterial OTUs were ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.  These 5 OTUs were present in at least ```r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])``` out of the total ```r length(rownames(model_train_summary))``` different 80/20 runs.

***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
A total of ```r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))``` sample was omitted from the original ```r length(rownames(good_metaf))``` sample set since it was missing a complete set of follow up data.  This left a total of ```r sum(initial_cm$ref_yes)``` samples for test predictions.  After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4].  This decrease was significant for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)``` [Figure 4a] and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4b] alone.  This also held specifically for those with screen relative neoplasias (SRN) (P-value = ```r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)```).  A total of ```r initial_cm["pred_yes", "ref_yes"]``` or ```r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)```% of all samples were correctly predicted to have an adenoma or carcinoma.  Although there was a decrease in positive probability only ```r followup_cm["pred_no", "ref_no"]``` of the total ```r sum(initial_cm$ref_yes)``` individuals were classified as adenoma or carcinoma free on follow up (successful classification of ```r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)```%). There was no significant difference between the predictions and actual diagnosis as measured by the Mcnemar test (P-value = ```r confusion_summary["McnemarPValue", "initial"]```).  However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = ```r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)```).  Although there were discordent results the respective sensitivity for the initial group was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% and for follow up was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```%, respectively.

There was ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual who still clearly had CRC on follow up as well as ```r length(rownames(filter(good_metaf, Disease_Free == "unknown")))``` individuals whose status on follow up was unknown.  Although the ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = ```r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)```).  Interestingly, ```r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))``` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].

The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)```) or  radiation therapy (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)```).  There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)```) or radiation therapy (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)```).

***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples***
Overall, there were a total of ```r length(rownames(pvalue_common_imp_vars))``` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].  A total of ```r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))``` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```.  In general, ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]``` (```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```) was decreased from initial to follow up [Figure 5].  The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].

\newpage

## Discussion
In this study we show that the overall community structure shows very little change between controls and those with either adenoma or carcinoma [Table S1] in our larger training set.  With respect to the follow up sample group there was very little difference in thetayc distance metric between those with adenoma or carcinoma[Figure 1a].  FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].  When investigating the bacterial community, specifically between these individuals, there was very little change in overall community in the adenoma group but there was a change in the carcinoma group [Figure 2].  This information showed that the removal and treatment of carcinoma for these cursory metrics had the largest response.  Yet these metrics alone would not be able to identify differences in adenomas.

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  We chose the middle training model, based on AUC, from 100 80/20 (train/test) splits.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using this model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample as a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)``` and for follow ups was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```.  Although there was a decrease in overall probability of an adenoma or carcinoma only ```r followup_cm["pred_no", "ref_no"]``` were below the 0.5 threshold out of the total ```r sum(followup_cm$ref_no)``` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]``` was to ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  This research provides evidence that it is possible to create a highily sensitive model for detection of adenoma or carcinoma with bacterial microbiome data.  It accomplishes this by using a unique sample set in which before and after stool samples are available for assessment.  By using these types of samples we are able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool (**insert citation**).  Based on previous work by Jobin, et al. (**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for the training models.  Similarily, Porphyromonas has also been implicated in colorectal cancer (**insert citation**).  Interestingly, the other OTUs had taxonomic identification for resident gut microbes.  This could suggest that changes to the resident microbiome are important to the initiation of adenoma or carcinoma formation (**insert citation**) and provide support for the hypothesis that an initial change in the bacterial microbiome could pave the way for more inflammatory species: whether by creation of a new niche for oral microbes (**insert citation**) or allowing for a bloom of existing pro-inflammatory residents (**insert citation**).

Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.  One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.  This has been suggested to be the case in a previous study (**insert Baxter Study**).  Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the best model does not utilize these specific bacteria for prediction purposes.

One limitation is that we do not know whether these individuals eventually had a CRC remission.  This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having CRC or adenomas.  Another limitation is that we do not know if adding modern tests such as the stool DNA test (**insert citation**) could help improve our overall AUC.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing that it can successfully predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.  Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations.

Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.  The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures associated with these lesions.  It also shows that these algorithms successfully react to successful treatment regimens and may be able to one day diagnose CRC with a high level of accuracy.

\newpage

## Methods

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cacner, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).  Kept apart from this training set were a total of ```r length(rownames(good_metaf))``` individuals that not only had a sample as described previoulsy but also a follow up sample between ```r min(good_metaf$time)``` - ```r max(good_metaf$time)``` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) (**insert citation**). 16S rRNA gene sequencing was completed as previously described by Kozich, et al. (**insert citation**).  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Which of each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described (**insert citations**).  The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to ```r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))``` in an attempt to minimize uneven sampling.

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the ```r length(rownames(good_metaf))``` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v```r paste(packageVersion("caret"))```).  The method of choosing the best model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  This was repeated 100 times with the best model after all the repeats chosen to be used on the ```r length(rownames(good_metaf))``` samples with follow up.  The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as having or not having a lesion.

***Selection of Important OTUs***
In order to assess which variables were most central to all the models the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) was recorded and those varialbes that were present more than 50% of the time were considered important.

A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples and specifically lesion.  This new model OTUs were then compared to the overall lesion model and common OTUs were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.

***Statistical Analysis***
The R software package (v```r paste(version$major, ".", version$minor, sep = "")```) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v```r paste(packageVersion("vegan"))```) while comparisons between ROC curves utilized the method by DeLong et al. (**insert citation**) executed by the pROC (v```r paste(packageVersion("pROC"))```) package.  Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied (**insert citation**) and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of the BH corrected ones.

***Reproducible methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Baxter_followUps_2016.

\newpage

## Acknowledgements

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. This study was supported by funding from the National Institutes of Health to P. Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).

\newpage

**Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = ```r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)```).  B) A significant difference was found between the adenoma and carcinoma group for Fit (P-value = ```r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)```).

**Figure 2: NMDS of the Overall Bacterial Community Changes.**  A) NMDS of the intial and follow up samples for the Adenoma group.  B) NMDS of the initial and follow up samples for the Carcinoma group.

**Figure 3: Graph of the Receiver Operating Characteristic Curve on Test Set Performance of the Best, Middle, and Worse Training Models.** For each of the 100 training cohort sets used had ```r round(length(rownames(metaI))*0.8)``` individuals and the testing cohort sets had ```r round(length(rownames(good_metaf))*0.2)``` individuals.  The AUC on the test sets for the best, middle, and worse models from training were ```r format(model_test_AUC_summary["best", "AUC"], digits = 3)```, ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```, and ```r format(model_test_AUC_summary["worse", "AUC"], digits = 3)```, respectively.  cvAUC is the 10-fold cross-validated AUC from training.

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Initial and Follow Up***  A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.  THe dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.

**Figure 5: Lesion Model OTU with a Significant Decrease in Relative Abundance that is also Predictive of Initial and Follow Up.**  After multiple comparison correction ```r rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05))``` (```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```) was the only one with a P-value < 0.05.  The dotted line represents the average relative abundance in the control training group.

\newpage

**Figure S1: Thetayc Graphed Against Time of Follow up Sample from Initial**

\newpage

## References

### 2016-12-02 submission/supplemental_outline_20161024.Rmd
first draft manuscript

---
title: ""
bibliography:
csl: mbio.csl
output: 
  pdf_document:
    keep_tex: true
fontsize: 12pt
geometry: margin=1.0in
---

# Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment

\vspace{10mm}

\begin{center}
Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss

\vspace{10mm}

Supplemental
\end{center}

\newpage

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('code/functions.R')
source('code/graphFunctions.R')

loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))

```

**Table S1: P-values of Differences between Initial and Follow Up Samples for Richness, Shannon Diversity, and Evenness.**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

alpha_table <- read.csv("results/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
supplemental_table1 <- separate(alpha_table, X, c("comparison", "alpha_metric"), sep="_")

colnames(supplemental_table1) <- c("Comparison", "Alpha Metric", "P-value", "BH Adjusted P-value")

kable(supplemental_table1, format = "markdown", align = 'c', padding = 2)
```

\newpage

**Table S2: Summary of Most Important Variables For the Model**

```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}

imp_model_vars <- read.csv('results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
rf_otu_tax <- read.csv('results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
rf_otu_tax <- cbind(otu = rownames(rf_otu_tax), rf_otu_tax)

all_data <- full_join(imp_model_vars, rf_otu_tax, by = c("Variable" = "otu"))

imp_model_vars <- cbind(Variable = imp_model_vars$Variable, Tax = createTaxaLabeller(all_data), 
                        total_appearance = imp_model_vars$total_appearance)

colnames(imp_model_vars) <- c("Variable", "Lowest Taxonomic ID", "Total Model Appearances")

kable(imp_model_vars, format = "markdown", align = 'c', padding = 2)

```

\newpage

**Table S3: Summary of Probability Assignment of Follow Up Individuals**

```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}

follow_up_prob_summary <- read.csv('results/tables/follow_up_probability_summary.csv', stringsAsFactors = F, header = T)

st3 <- as.data.frame(filter(follow_up_prob_summary, sampleType == "initial")[, "Yes"]) %>% 
  mutate(follow_yes = filter(follow_up_prob_summary, sampleType == "followup")[, "Yes"]) %>% 
  mutate(disease_free = filter(follow_up_prob_summary, sampleType == "initial")[, "disease_free"]) %>% 
  mutate(diagnosis = filter(follow_up_prob_summary, sampleType == "initial")[, "Dx_Bin"])

st3$disease_free[st3$disease_free == "n"] <- "No"
st3$disease_free[st3$disease_free == "unknown"] <- "Unknown"
st3$disease_free[st3$disease_free == "y"] <- "Yes"

colnames(st3) <- c("Initial Positive Probability", "Follow Up Positive Probability", "Disease Free on Follow Up", "Initial Diagnosis")

kable(st3, format = "markdown", align = 'c', padding = 2)

```

\newpage

**Table S4: Summary of OTUs in both Lesion and Initial/Follow Up Models**

```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}

pvalue_common_imp_vars <- read.csv('results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)

st4 <- select(pvalue_common_imp_vars, otu) %>% 
  mutate(Tax = createTaxaLabeller(select(pvalue_common_imp_vars, -Pvalue, -BH_corr))) %>% 
  mutate(Pvalue = pvalue_common_imp_vars$Pvalue) %>% 
  mutate(BH_corr = pvalue_common_imp_vars$BH_corr)

colnames(st4) <- c("OTU", "Lowest Taxonomic ID", "P-value", "BH Corrected P-value")

kable(st4, format = "markdown", align = 'c', padding = 2)

```

```{r supplemental_figure1, echo=FALSE, message=FALSE, warning=FALSE}

diff_table_time <- read.csv("results/tables/time_datatable.csv", header = T)

# Plot the change in distance versus time
time_graph <- ggplot(diff_table_time, aes(x = time, y = distance)) + 
  geom_point(aes(color = dx), size = 4) + theme_bw() + 
  scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
                     breaks = c("adenoma", "cancer"), labels = c("Adenoma", "Cancer")) + 
  coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("Time from Initial (Days)") + 
  theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"))

ggsave(file = "results/figures/FigureS1.pdf", time_graph, 
       width=6, height = 8, dpi = 300)
```

### 2016-12-05 submission/manuscript_outline_20161024.Rmd
abstract changes

@@
 In this current study we attempt to combine specific physical characteristics, patient history, Fecal Immunochemical Test (FIT), and the bacterial microbiome to build the best model for classifying those with and without lesion (adenoma or carcinoma).
-From this improved model we applied it to a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
-Carcinoma dependent differences in initial and follow up samples were identified for the overall bacterial community composition (P-value = ```r paste(beta_summary['Carcinoma', 'PERMANOVA'])```).
+This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
+Using NMDS, the thetayc distance between initial and follow up showed a significant difference (P-value = ```r paste(beta_summary['Carcinoma', 'PERMANOVA'])```) for the carcinoma group only.
 The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.

### 2016-12-05 submission/manuscript_outline_20161024.tex
abstract changes

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}

\begin{document}

\section{Differences in the Stool Microbiome Before and After Colorectal
Cancer
Treatment}\label{differences-in-the-stool-microbiome-before-and-after-colorectal-cancer-treatment}

\vspace{25mm}

\begin{center}
Running Title: Human Microbiome and Colorectal Cancer

\vspace{10mm}

Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss${^\dagger}$

\vspace{10mm}

Contributions: All authors were involved in the conception and design of the study. MAS analyzed the data. NTB processed samples and analyzed the data. All authors interpreted the data. MAS and PDS wrote the manuscript. All authors reviewed and revised the manuscript. All authors read and approved the final manuscript.

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this current study
we attempt to combine specific physical characteristics, patient
history, Fecal Immunochemical Test (FIT), and the bacterial microbiome
to build the best model for classifying those with and without lesion
(adenoma or carcinoma). This model was tested on a \texttt{66} person
group that included samples before and after treatment to allow for the
assessment of how the model adjusts risk after treatment. Using NMDS,
the thetayc distance between initial and follow up showed a significant
difference (P-value = \texttt{0.001}) for the carcinoma group only. The
model chosen had a 10-fold cross validated AUC of \texttt{0.819} and a
test set AUC of \texttt{0.815}. There was a significant decrease in
probabliity of lesion from the model after treatment for both the
adenoma (P-value = \texttt{8.48e-07}) and carcinoma (P-value =
\texttt{5.96e-08}) group. The top 5 most important operational taxonomic
units for this model were \texttt{Lachnospiraceae\ (Otu000013)},
\texttt{Escherichia/Shigella\ (Otu000018)},
\texttt{Ruminococcaceae\ (Otu000020)},
\texttt{Ruminococcus\ (Otu000017)}, and
\texttt{Porphyromonas\ (Otu000153)}. For prediction on samples with
follow up the initial sensitivity was \texttt{98.5}\% (\texttt{65/66})
and follow up sensitivity was \texttt{100}\% (\texttt{1\ /\ 1}).
Overall, this model can accurately predict and subsequently respond to
treatment by lowering the probability of lesion.

\newpage

\subsubsection{Importance}\label{importance}

This is one of the first studies to investigate within humans what
happens to the bacterial microbiome before and after colorectal cancer
treatment. Specifically, it aims to assess how the random forest machine
learning algorithm built models respond to treatment and adjust their
positive probability calls of whether the individual has an adenoma or
carcinoma.

\newpage

\subsection{Introduction}\label{introduction}

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age (\textbf{insert citation}). Over the last few years
death due to the disease has seen a significant decrease thanks mainly
to improvements in screening (\textbf{insert citation}). However,
despite this giant improvement there are still approximately 50,000
deaths from the disease a year (\textbf{insert citation}). It is
estimated that around 5-10\% of all CRCs can be explained by autosomal
dominant inheritance (\textbf{insert citation}). The vast majority of
CRCs are not inherited and the exact etiology to disease is not well
worked out (\textbf{insert citation}).

Recent reports have found that the microbiome between those with and
without CRC are different and that bacteia associated with CRC tissue is
different that that found on the normal mucosa (\textbf{Insert
citations}). Further, there is evidence to suggest that more
inflammatory mouth-associated bacteria are more prevalent in the guts of
those with CRC (\textbf{insert citation}). This has led to the
hypothesis that these bacteria replace traditional residents that
produce metabolites such as butyrate (\textbf{insert citation}) and lead
to a more inflammatory state (\textbf{insert citation}). This
inflammatory state allows for more of these types of bacteria to
populate the gut and leads to neoplasia and eventually CRC
(\textbf{insert citation}).

Building upon these findings a number of groups, including our own, have
investigated using the bacterial microbiome as a potential biomarker for
adenoma and CRC detection (\textbf{insert citations}). These
collectively show that a reasonable area under the curve (AUC) can be
obtained when using the stool bacterial microbiome to classify disease
state (\textbf{insert citation}). Further, a number of groups have
extended this observation to show that using either specific bacterial
species or the bacterial microbiome as a whole paired with the Fecal
Immunochemical Test (FIT) can either increase the AUC of the model
(\textbf{insert citations}).

In this study we further refine the random forest model by including
more background information on the indivdiuals within it. We use a
rigourous and accepted approach for machine learning training and
testing to validate this model (\textbf{insert citation}). We then apply
it to individuals in which we have samples before and after treatment to
assess how well the model adjusts their probability of having an adenoma
or carcinoma. Finally, we investigate which specific Operational
Taxonomic Units (OTUs) are most affected by treatment. Within this study
lesion refers to both adenoma and carcinoma.

\newpage

\subsection{Results}\label{results}

\textbf{\emph{Bacterial Community and Fit Changes before and after
Treatment}} Based on thetayc distance metrics comparing the initial to
the follow up samples there was no difference between the adenoma and
carcinoma group (P-value = \texttt{0.697}) {[}Figure 1a{]}. There was a
difference in FIT between initial and follow up samples with the
carcinoma group having a significant decrease in FIT versus the adenoma
group (P-value = \texttt{2.15e-05}) {[}Figure 1b{]}. Although the
thetayc distance metric change was similar between adenoma and carcinoma
the directionality of the change was significant in the cancer group
between initial and follow up (P-value = \texttt{0.001}) but not for the
adenoma group (P-value = \texttt{1}) {[}Figure 2{]}. When all follow up
samples were compared to each other there was no significant overall
difference between them (P-value = \texttt{0.643}). There was no
significant difference between initial and follow up samples for
observed OTUs, Shannon diversity, and evenness after correction for
multiple comparisons {[}Table S1{]}. Time of follow up sample from
initial sampling did not have an effect on our prediction data set
(uncorrected P-value = \texttt{0.784}).

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of \texttt{0.795} to a maximum of
\texttt{0.854}. To be conservative the model chosen for prediction of
disease in the follow up samples had an AUC in the middle of all the
\texttt{100} runs which was \texttt{0.819}. Interestingly, the worst AUC
model from training performed the best on it's respective 80/20 split
test data {[}Figure 3{]}. In fact the 80/20 test performance showed that
the AUC for the middle model chosen was the most stable (best training
model test set AUC = \texttt{0.682}, middle training model test set AUC
= \texttt{0.815}, worse training model test set AUC = \texttt{0.931}).
That is to say it had the smallest change in AUC in comparison to the
other minimum and maximum AUC trained models. There was no significant
difference between the AUC of the best and middle training models
(P-value = \texttt{0.0698}). There was a significant difference between
the worse training model and both the best training model (P-value =
\texttt{0.000397}) and middle training model (P-value =
\texttt{0.0444}).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of \texttt{32} variables identified as being present in
more than 50\% of the training models {[}Table S2{]}. The top 5 most
important bacterial OTUs were \texttt{Lachnospiraceae\ (Otu000013)},
\texttt{Escherichia/Shigella\ (Otu000018)},
\texttt{Ruminococcaceae\ (Otu000020)},
\texttt{Ruminococcus\ (Otu000017)}, and
\texttt{Porphyromonas\ (Otu000153)}. These 5 OTUs were present in at
least \texttt{99} out of the total \texttt{100} different 80/20 runs.

\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
Decrease in Positive Probability Prediction}} A total of \texttt{1}
sample was omitted from the original \texttt{67} sample set since it was
missing a complete set of follow up data. This left a total of
\texttt{66} samples for test predictions. After multiple comparison
correction there was a significant overall decrease in positive
probability of a carcinoma and adenoma (P-value = \texttt{8.27e-11})
{[}Figure 4{]}. This decrease was significant for both adenoma (P-value
= \texttt{8.48e-07} {[}Figure 4a{]} and carcinoma (P-value =
\texttt{5.96e-08}) {[}Figure 4b{]} alone. This also held specifically
for those with screen relative neoplasias (SRN) (P-value =
\texttt{1.45e-04}). A total of \texttt{65} or \texttt{98.5}\% of all
samples were correctly predicted to have an adenoma or carcinoma.
Although there was a decrease in positive probability only \texttt{26}
of the total \texttt{66} individuals were classified as adenoma or
carcinoma free on follow up (successful classification of
\texttt{40.9}\%). There was no significant difference between the
predictions and actual diagnosis as measured by the Mcnemar test
(P-value = \texttt{1}). However, the predictions were significantly
discordent with the diagnosis for the follow up samples (P-value =
\texttt{1.17e-09}). Although there were discordent results the
respective sensitivity for the initial group was \texttt{98.5}\% and for
follow up was \texttt{100}\%, respectively.

There was \texttt{1} individual who still clearly had CRC on follow up
as well as \texttt{5} individuals whose status on follow up was unknown.
Although the \texttt{1} individual had a decrease in positive
probability their follow up sample was still higher than the cutoff
threshold of 0.5 (positive probability = \texttt{0.902}). Interestingly,
\texttt{2} individuals who were unknown on follow up still were over the
threshold cutoff of 0.5 even though, like the \texttt{1} individual with
clear CRC on follow up, the probability of an adenoma or carcinoma
decreased {[}Table S3{]}.

The follow up positive probabilities were not affected by either
chemotherapy treatment (uncorrected P-value = \texttt{0.583}) or
radiation therapy (uncorrected P-value = \texttt{0.809}). There was also
no difference in the amount of change in the positive probability based
on whether individuals received chemotherapy (unccorected P-value =
\texttt{0.224}) or radiation therapy (uncorrected P-value =
\texttt{0.846}).

\textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
Follow Up Versus Initial Samples}} Overall, there were a total of
\texttt{8} OTUs that were common between the main lesion model and the
model for classifying initial and follow up samples specifically
{[}Table S4{]}. A total of \texttt{1} OTU was still significant after
multiple comparison correction and it's lowest taxonomic identification
was to \texttt{Blautia}. In general, \texttt{Otu000012}
(\texttt{Blautia}) was decreased from initial to follow up {[}Figure
5{]}. The relative abundance was not drastically different then the mean
of the values observed in the control training set {[}Figure 5{]}.

\newpage

\subsection{Discussion}\label{discussion}

In this study we show that the overall community structure shows very
little change between controls and those with either adenoma or
carcinoma {[}Table S1{]} in our larger training set. With respect to the
follow up sample group there was very little difference in thetayc
distance metric between those with adenoma or carcinoma{[}Figure 1a{]}.
FIT had a large change in the initial and follow up samples in the
carcinoma group versus the adenoma {[}Figure 1b{]}. When investigating
the bacterial community, specifically between these individuals, there
was very little change in overall community in the adenoma group but
there was a change in the carcinoma group {[}Figure 2{]}. This
information showed that the removal and treatment of carcinoma for these
cursory metrics had the largest response. Yet these metrics alone would
not be able to identify differences in adenomas.

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). We chose the middle training model, based on AUC, from 100
80/20 (train/test) splits. It's 10-fold cross validated AUC was similar
to it's test set AUC which was not the case for both the best and worse
training model {[}Figure 3{]}. Using this model we predicted the
probability of a lesion in the inital and follow up samples {[}Figure
4{]}. There was a signficant decrease in positive probability regardless
of whether the sample as a carcinoma or adenoma. The overall sensitivity
for lesion detection in the intial samples was \texttt{98.5} and for
follow ups was \texttt{100}. Although there was a decrease in overall
probability of an adenoma or carcinoma only \texttt{26} were below the
0.5 threshold out of the total \texttt{65} individuals who were
diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S4{]}. Only a single OTU was significant
after multiple comparison correction and the lowest taxonomic
identification of \texttt{Otu000012} was to \texttt{Blautia}. Although
there was a difference in the relative abundance at initial and follow
up these values were not drastically different from the relative
abundance values observed in the control individuals of the training set
{[}Figure 5{]}. This research provides evidence that it is possible to
create a highily sensitive model for detection of adenoma or carcinoma
with bacterial microbiome data. It accomplishes this by using a unique
sample set in which before and after stool samples are available for
assessment. By using these types of samples we are able to show that
this model is reactive. That is to say that after surgery for removal of
the adenoma or carcinoma it decreases the positive probability to
reflect a lower likelihood of the individual having an adenoma or
carcinoma.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
(\textbf{insert citation}). Based on previous work by Jobin, et al.
(\textbf{insert citation}) it may not be surprising to see E.coli in the
top 5 OTUs for the training models. Similarily, Porphyromonas has also
been implicated in colorectal cancer (\textbf{insert citation}).
Interestingly, the other OTUs had taxonomic identification for resident
gut microbes. This could suggest that changes to the resident microbiome
are important to the initiation of adenoma or carcinoma formation
(\textbf{insert citation}) and provide support for the hypothesis that
an initial change in the bacterial microbiome could pave the way for
more inflammatory species: whether by creation of a new niche for oral
microbes (\textbf{insert citation}) or allowing for a bloom of existing
pro-inflammatory residents (\textbf{insert citation}).

Naturally, it is curious that normal staples of many screening studies
such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
present in the majority of the training models. One potential
explanation for this is that FIT provides the same information to the
model as these three organisms and so the model uses FIT preferentially
over them. This has been suggested to be the case in a previous study
(\textbf{insert Baxter Study}). Regardless, our study does not argue
against the importance of these bacterium in CRC initiation or
pathogenesis but rather that the best model does not utilize these
specific bacteria for prediction purposes.

One limitation is that we do not know whether these individuals
eventually had a CRC remission. This information would help to
strengthen the case for our Random Forest based model keeping a number
of individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having CRC or adenomas. Another limitation
is that we do not know if adding modern tests such as the stool DNA test
(\textbf{insert citation}) could help improve our overall AUC. Finally,
although our training and test set are relatively large we still run the
risk of overfitting or having a model that may not be immediately
extrapolateable to other populations. We've done our best to safeguard
against this by not only running 10-fold cross validation but also
having over 100 different 80/20 splits to try and mimic the type of
variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing
that it can successfully predict both carcinoma and adenoma our study
provides further data that these patient factors in conjunction with the
bacterial microbiome could potentially influence CRC and perhaps have a
role in formation of adenomas. Further studies need to be carried out to
verify our findings since not only are we dealing with stool, which
could be very different than the communities present on the actual
tissue, but also are dealing with correlations.

Despite these limitations we think that these findings significantly add
to the existing scientific knowledge on CRC and the bacterial
microbiome. The ability for machine learning algorithms to take
bacterial microbiome data and successfully lower positive probability
after either adenoma or carcinoma removal provides evidence that there
are specific signatures associated with these lesions. It also shows
that these algorithms successfully react to successful treatment
regimens and may be able to one day diagnose CRC with a high level of
accuracy.

\newpage

\subsection{Methods}\label{methods}

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
(\textbf{insert citation}). In brief, study exclusion involved those who
had already undergone surgery, radiation, or chemotherapy, had
colorectal cancer before a baseline stool sample could be obtained, had
IBD, a known hereditary non-polyposis colorectal cacner, or Familial
adenomatous polyposis. Samples used to build the model used for
prediction were collected either prior to a colonoscopy or between 1 - 2
weeks after. The bacterial microbiome has been shown to nomralize within
this time period (\textbf{insert citation}). Kept apart from this
training set were a total of \texttt{67} individuals that not only had a
sample as described previoulsy but also a follow up sample between
\texttt{188} - \texttt{546} days after surgery and treatment had been
completed. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
citation}). 16S rRNA gene sequencing was completed as previously
described by Kozich, et al. (\textbf{insert citation}). In brief, DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with control, adenoma, and carcinoma
evenly represented on each run. Which of each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described (\textbf{insert citations}). The general
processing workflow using mothur is as follows: Paired-end reads were
first merged into contigs, quality filtered, aligned to the SILVA
database, screening for chimeras, classified with a naive Bayesian
classifier using the Ribosomal Database Project (RDP), and clustered
into Operational Taxonomic Units (OTUs) using a 97\% similarity cutoff
with an average neighbor clustering algorithm. The numer of sequences
for each sample was rarified to \texttt{10521} in an attempt to minimize
uneven sampling.

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the \texttt{67} individuals with
follow up samples. The model included data on FIT, the bacterial
microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was
caucasion or not, history of cancer, and family history of cancer.
Non-binary data was checked for near zero variance and auto correlation.
Data columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v\texttt{6.0.73}). The method of
choosing the best model involved taking the \texttt{490} samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
twenty times through 10-fold cross validation to optimzie the model's
Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
This resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. This was repeated 100
times with the best model after all the repeats chosen to be used on the
\texttt{67} samples with follow up. The default cutoff of a probability
of 0.5 was used as the threshold to classify individuals as having or
not having a lesion.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models the number of times a
variable was present in the top 10\% of mean decrease in accuracy (MDA)
was recorded and those varialbes that were present more than 50\% of the
time were considered important.

A second model using the same types of conditions was used to identify
OTUs that could predict initial and follow up samples and specifically
lesion. This new model OTUs were then compared to the overall lesion
model and common OTUs were selected to test for changes between initial
and follow up using a paired wilcoxson rank sum test.

\textbf{\emph{Statistical Analysis}} The R software package
(v\texttt{3.3.0}) was used for all statisitical analysis. Comparisons
between bacterial community structure utilized PERMANOVA (\textbf{insert
citation}) in the vegan package (v\texttt{2.4.1}) while comparisons
between ROC curves utilized the method by DeLong et al. (\textbf{insert
citation}) executed by the pROC (v\texttt{1.8}) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied (\textbf{insert citation}) and a corrected
P-value of less than 0.05 was considered significant. Unless otherwise
stated the P-values reported are those of the BH corrected ones.

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Baxter_followUps_2016}.

\newpage

\subsection{Acknowledgements}\label{acknowledgements}

The authors thank the Great Lakes-New England Early Detection Research
Network for providing the fecal samples that were used in this study.
This study was supported by funding from the National Institutes of
Health to P. Schloss (R01GM099514, P30DK034933) and to the Early
Detection Research Network (U01CA86400).

\newpage

\textbf{Figure 1: Change in Thetayc and Fit between initial and follow
up in adenoma or carcinoma group.} A) No significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
\texttt{0.697}). B) A significant difference was found between the
adenoma and carcinoma group for Fit (P-value = \texttt{2.15e-05}).

\textbf{Figure 2: NMDS of the Overall Bacterial Community Changes.} A)
NMDS of the intial and follow up samples for the Adenoma group. B) NMDS
of the initial and follow up samples for the Carcinoma group.

\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
on Test Set Performance of the Best, Middle, and Worse Training Models.}
For each of the 100 training cohort sets used had \texttt{392}
individuals and the testing cohort sets had \texttt{13} individuals. The
AUC on the test sets for the best, middle, and worse models from
training were \texttt{0.682}, \texttt{0.815}, and \texttt{0.931},
respectively. cvAUC is the 10-fold cross-validated AUC from training.

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Initial and Follow Up}* A) Positive probability adjustment
of those with carcinoma from intial to follow up sample B) Positive
probability adjustment of those with adenoma as well as those with SRN
and the probability adjustment from initial to follow up sample. THe
dotted line represents the threshold used to make the decision of
whether a sample was lesion positive or not.

\textbf{Figure 5: Lesion Model OTU with a Significant Decrease in
Relative Abundance that is also Predictive of Initial and Follow Up.}
After multiple comparison correction \texttt{1} (\texttt{Blautia}) was
the only one with a P-value \textless{} 0.05. The dotted line represents
the average relative abundance in the control training group.

\newpage

\textbf{Figure S1: Thetayc Graphed Against Time of Follow up Sample from
Initial}

\newpage

\subsection{References}\label{references}

\end{document}

### 2016-12-05 submission/manuscript_outline_20161024.Rmd
abstract update v2

@@
 These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.
-In this current study we attempt to combine specific physical characteristics, patient history, Fecal Immunochemical Test (FIT), and the bacterial microbiome to build the best model for classifying those with and without lesion (adenoma or carcinoma).
+In this study we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for either, after surgical treatment.
 This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
-Using NMDS, the thetayc distance between initial and follow up showed a significant difference (P-value = ```r paste(beta_summary['Carcinoma', 'PERMANOVA'])```) for the carcinoma group only.
 The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.
 There was a significant decrease in probabliity of lesion from the model after treatment for both the adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```) group.
-The top 5 most important operational taxonomic units for this model were ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.
-For prediction on samples with follow up the initial sensitivity was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) and follow up sensitivity was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
-Overall, this model can accurately predict and subsequently respond to treatment by lowering the probability of lesion.
+The top 5 most important operational taxonomic units (OTUs) for prediction of lesion classified to ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.
+The prediction of lesion on initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
+Overall, this model could accurately predict those with an adenoma or carcinoma and lower probability of having a lesion based on follow up samples after treatment.
 

### 2016-12-05 submission/manuscript_outline_20161024.tex
abstract update v2

@@
 have mostly focused on investigating how the bacterial microbiome is
-used at a single point in time to predict disease. In this current study
-we attempt to combine specific physical characteristics, patient
-history, Fecal Immunochemical Test (FIT), and the bacterial microbiome
-to build the best model for classifying those with and without lesion
-(adenoma or carcinoma). This model was tested on a \texttt{66} person
-group that included samples before and after treatment to allow for the
-assessment of how the model adjusts risk after treatment. Using NMDS,
-the thetayc distance between initial and follow up showed a significant
-difference (P-value = \texttt{0.001}) for the carcinoma group only. The
-model chosen had a 10-fold cross validated AUC of \texttt{0.819} and a
-test set AUC of \texttt{0.815}. There was a significant decrease in
-probabliity of lesion from the model after treatment for both the
-adenoma (P-value = \texttt{8.48e-07}) and carcinoma (P-value =
-\texttt{5.96e-08}) group. The top 5 most important operational taxonomic
-units for this model were \texttt{Lachnospiraceae\ (Otu000013)},
+used at a single point in time to predict disease. In this study we
+assessed whether a model built with bacterial microbiome data could
+accurately predict adenoma or carcinoma and adjust positive probability
+for either, after surgical treatment. This model was tested on a
+\texttt{66} person group that included samples before and after
+treatment to allow for the assessment of how the model adjusts risk
+after treatment. The model chosen had a 10-fold cross validated AUC of
+\texttt{0.819} and a test set AUC of \texttt{0.815}. There was a
+significant decrease in probabliity of lesion from the model after
+treatment for both the adenoma (P-value = \texttt{8.48e-07}) and
+carcinoma (P-value = \texttt{5.96e-08}) group. The top 5 most important
+operational taxonomic units (OTUs) for prediction of lesion classified
+to \texttt{Lachnospiraceae\ (Otu000013)},
 \texttt{Escherichia/Shigella\ (Otu000018)},
@@
 \texttt{Ruminococcus\ (Otu000017)}, and
-\texttt{Porphyromonas\ (Otu000153)}. For prediction on samples with
-follow up the initial sensitivity was \texttt{98.5}\% (\texttt{65/66})
-and follow up sensitivity was \texttt{100}\% (\texttt{1\ /\ 1}).
-Overall, this model can accurately predict and subsequently respond to
-treatment by lowering the probability of lesion.
+\texttt{Porphyromonas\ (Otu000153)}. The prediction of lesion on initial
+samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66}) while
+follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Overall, this model
+could accurately predict those with an adenoma or carcinoma and lower
+probability of having a lesion based on follow up samples after
+treatment.
 

### 2016-12-05 submission/manuscript_outline_20161024.Rmd
abstract update v 3

@@
 These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.
-In this study we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for either, after surgical treatment.
+In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.
 This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
 The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.
-There was a significant decrease in probabliity of lesion from the model after treatment for both the adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```) group.
+: For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```).
 The top 5 most important operational taxonomic units (OTUs) for prediction of lesion classified to ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.
-The prediction of lesion on initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
+In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
 Overall, this model could accurately predict those with an adenoma or carcinoma and lower probability of having a lesion based on follow up samples after treatment.

### 2016-12-05 submission/manuscript_outline_20161024.tex
abstract update v 3

--- 2016-12-05 submission/manuscript_outline_20161024.tex
@@
 have mostly focused on investigating how the bacterial microbiome is
-used at a single point in time to predict disease. In this current study
-we attempt to combine specific physical characteristics, patient
-history, Fecal Immunochemical Test (FIT), and the bacterial microbiome
-to build the best model for classifying those with and without lesion
-(adenoma or carcinoma). This model was tested on a \texttt{66} person
-group that included samples before and after treatment to allow for the
-assessment of how the model adjusts risk after treatment. Using NMDS,
-the thetayc distance between initial and follow up showed a significant
-difference (P-value = \texttt{0.001}) for the carcinoma group only. The
-model chosen had a 10-fold cross validated AUC of \texttt{0.819} and a
-test set AUC of \texttt{0.815}. There was a significant decrease in
-probabliity of lesion from the model after treatment for both the
-adenoma (P-value = \texttt{8.48e-07}) and carcinoma (P-value =
-\texttt{5.96e-08}) group. The top 5 most important operational taxonomic
-units for this model were \texttt{Lachnospiraceae\ (Otu000013)},
+used at a single point in time to predict disease. In this study, we
+assessed whether a model built with bacterial microbiome data could
+accurately predict adenoma or carcinoma and adjust positive probability
+for lesion (adenoma or carcinoma) after surgical treatment. This model
+was tested on a \texttt{66} person group that included samples before
+and after treatment to allow for the assessment of how the model adjusts
+risk after treatment. The model chosen had a 10-fold cross validated AUC
+of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
+up samples our Random Forest model significantly decreased the positive
+probability of lesion compared to the initial samples for both adenoma
+(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
+\texttt{5.96e-08}). The top 5 most important operational taxonomic units
+(OTUs) for prediction of lesion classified to
+\texttt{Lachnospiraceae\ (Otu000013)},
 \texttt{Escherichia/Shigella\ (Otu000018)},
@@
 \texttt{Ruminococcus\ (Otu000017)}, and
-\texttt{Porphyromonas\ (Otu000153)}. For prediction on samples with
-follow up the initial sensitivity was \texttt{98.5}\% (\texttt{65/66})
-and follow up sensitivity was \texttt{100}\% (\texttt{1\ /\ 1}).
-Overall, this model can accurately predict and subsequently respond to
-treatment by lowering the probability of lesion.
+\texttt{Porphyromonas\ (Otu000153)}. In the test set, the prediction of
+lesion for the initial samples had a sensitivity of \texttt{98.5}\%
+(\texttt{65/66}) while follow up it was \texttt{100}\%
+(\texttt{1\ /\ 1}). Overall, this model could accurately predict those
+with an adenoma or carcinoma and lower probability of having a lesion
+based on follow up samples after treatment.
 

### 2016-12-05 code/Run_Compare_models.R
update code to use

@@ -8,8 +8,8 @@
 source('code/functions.R')
 source('code/graphFunctions.R')
 
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", 
-           "wesanderson", "caret", "randomForest"))
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
+  "gridExtra", "scales", "wesanderson", "caret", "randomForest"))
 
 load("exploratory/RF_model_Imp_OTU.RData")
 
@@ -18,66 +18,88 @@ imp_vars_list <- list()
 for(i in 1:length(test_tune_list)){
   
   imp_vars_list[[paste("run_", i, sep = "")]] <- 
-    varImp(test_tune_list[[paste("data_split", i, sep = "")]], scale = FALSE)$importance %>% 
+    varImp(test_tune_list[[paste("data_split", i, sep = "")]], 
+      scale = FALSE)$importance %>% 
     mutate(Variable = rownames(.)) %>% arrange(desc(Overall))
 }
 
 
 # Calculate number of times an OTU is in the top 10% of overall importance
 
-OTU_appearance_table <- as.data.frame(data_frame(Variable = imp_vars_list[["run_1"]]$Variable) %>% 
-                                        mutate(total_appearance = 0))
+OTU_appearance_table <- as.data.frame(data_frame(
+  Variable = imp_vars_list[["run_1"]]$Variable) %>% 
+mutate(total_appearance = 0))
+
 rownames(OTU_appearance_table) <- OTU_appearance_table$Variable
 
 for(j in 1:length(imp_vars_list)){
   
-  tempVars <- imp_vars_list[[j]][c(1:round(length(rownames(imp_vars_list[[j]]))*0.10)), ][, "Variable"]
+  tempVars <- imp_vars_list[[j]][c(
+    1:round(length(rownames(imp_vars_list[[j]]))*0.10)), ][, "Variable"]
   
   for(i in tempVars){
     
-    OTU_appearance_table[i, "total_appearance"] <- OTU_appearance_table[i, "total_appearance"] + 1
+    OTU_appearance_table[i, "total_appearance"] <- 
+    OTU_appearance_table[i, "total_appearance"] + 1
   }
 }
 
-OTU_appearance_table <- arrange(OTU_appearance_table, desc(total_appearance))
+OTU_appearance_table <- arrange(OTU_appearance_table, 
[…]

### 2016-12-05 code/qsubmission.sh

@@ -1,52 +1,39 @@
 #!/bin/sh
 
-submission="run_100_RF.pbs run_10_RF.pbs run_11_RF.pbs run_12_RF.pbs run_13_RF.pbs 
-run_14_RF.pbs run_15_RF.pbs run_16_RF.pbs" 
-
-submission_two="run_17_RF.pbs run_18_RF.pbs run_19_RF.pbs 
-run_1_RF.pbs run_20_RF.pbs run_21_RF.pbs run_22_RF.pbs run_23_RF.pbs" 
-
-submission_three="run_24_RF.pbs run_25_RF.pbs run_26_RF.pbs run_27_RF.pbs run_28_RF.pbs 
-run_29_RF.pbs run_2_RF.pbs run_30_RF.pbs" 
-
-submission_four="run_31_RF.pbs run_32_RF.pbs run_33_RF.pbs run_34_RF.pbs run_35_RF.pbs 
-run_36_RF.pbs run_37_RF.pbs run_38_RF.pbs" 
-
-submission_five="run_39_RF.pbs run_3_RF.pbs run_40_RF.pbs run_41_RF.pbs 
-run_42_RF.pbs run_43_RF.pbs run_44_RF.pbs run_45_RF.pbs" 
-
-submission_six="run_46_RF.pbs run_47_RF.pbs run_48_RF.pbs run_49_RF.pbs 
-run_4_RF.pbs run_50_RF.pbs run_51_RF.pbs run_52_RF.pbs run_53_RF.pbs" 
-
-submission_seven="run_54_RF.pbs run_55_RF.pbs run_56_RF.pbs run_57_RF.pbs 
-run_58_RF.pbs run_59_RF.pbs" 
-
-submission_eight="run_61_RF.pbs run_62_RF.pbs run_63_RF.pbs 
-run_64_RF.pbs run_65_RF.pbs run_66_RF.pbs run_67_RF.pbs run_68_RF.pbs"
-
-submission_nine="run_69_RF.pbs run_6_RF.pbs run_70_RF.pbs run_71_RF.pbs 
-run_72_RF.pbs run_73_RF.pbs run_74_RF.pbs run_75_RF.pbs run_5_RF.pbs" 
-
-submission_ten="run_76_RF.pbs run_77_RF.pbs run_78_RF.pbs run_79_RF.pbs 
[…]

### 2016-12-05 submission/manuscript_outline_20161024.Rmd
update abstract again

@@
 In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
-Overall, this model could accurately predict those with an adenoma or carcinoma and lower probability of having a lesion based on follow up samples after treatment.
+Overall, our model accurately predicted those with an adenoma or carcinoma and following treatment it also decreased the positive probability of having a lesion.
 

### 2016-12-05 submission/manuscript_outline_20161024.tex
update abstract again

@@
 (\texttt{65/66}) while follow up it was \texttt{100}\%
-(\texttt{1\ /\ 1}). Overall, this model could accurately predict those
-with an adenoma or carcinoma and lower probability of having a lesion
-based on follow up samples after treatment.
+(\texttt{1\ /\ 1}). Overall, our model accurately predicted those with
+an adenoma or carcinoma and following treatment it also decreased the
+positive probability of having a lesion.
 

### 2016-12-05 submission/manuscript_outline_20161024.Rmd
updated manuscript w/ newer abstract

@@
 
-source('code/functions.R')
-source('code/graphFunctions.R')
+source('../code/functions.R')
+source('../code/graphFunctions.R')
 
-loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
-           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))
+loadLibs(c("dplyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
@@
 # Extra data and meta data needed
-good_metaf <- read.csv('results/tables/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
-metaI <- read.csv('results/tables/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
-total_sub_Seqs <- read.delim('data/process/final.0.03.subsample.shared', header = T) %>% 
+good_metaf <- read.csv('../results/tables/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
+metaI <- read.csv('../results/tables/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
+total_sub_Seqs <- read.delim('../data/process/final.0.03.subsample.shared', header = T) %>% 
   select(contains("Otu0")) %>% rowSums()
@@
 # initial microbiome and FIT description data 
-change_summary <- read.csv('results/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
-alpha_summary <- read.csv('results/tables/alpha_table_summary.csv', header = T, row.names = 1)
-beta_summary <- read.csv('results/tables/beta_diver_summary.csv', header = T, row.names = 1)
-time_pvalues <- read.csv("results/tables/time_pvalues.csv", header = T, row.names = 1)
+change_summary <- read.csv('../results/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
+alpha_summary <- read.csv('../results/tables/alpha_table_summary.csv', header = T, row.names = 1)
+beta_summary <- read.csv('../results/tables/beta_diver_summary.csv', header = T, row.names = 1)
+time_pvalues <- read.csv("../results/tables/time_pvalues.csv", header = T, row.names = 1)
 
 # training of model data
-model_train_summary <- read.csv('results/tables/ROC_model_summary.csv', header = T)
-model_test_AUC_summary <- read.csv('results/tables/auc_summary.csv', header = T, row.names = 1)
-imp_model_vars <- read.csv('results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
-rf_otu_tax <- read.csv('results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
+model_train_summary <- read.csv('../results/tables/ROC_model_summary.csv', header = T)
+model_test_AUC_summary <- read.csv('../results/tables/auc_summary.csv', header = T, row.names = 1)
+imp_model_vars <- read.csv('../results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
+rf_otu_tax <- read.csv('../results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
 rf_lowest_ID <- createTaxaLabeller(rf_otu_tax)
@@
 # model prediction and adjustment performance
-IF_model_wilcox_paired_summary <- read.csv('results/tables/IF_model_wicox_paired_pvalue_summary.csv', header = T, row.names = 1)
-confusion_summary <- read.csv('results/tables/confusion_summary.csv', header = T, row.names = 1)
-follow_up_prob_summary <- read.csv('results/tables/follow_up_probability_summary.csv', header = T)
-followup_cm <- read.csv('results/tables/followup_confusion_matrix.csv', header = T, row.names = 1)
-initial_cm <- read.csv('results/tables/initial_confusion_matrix.csv', header = T, row.names = 1)
-roc_pvalues <- read.csv('results/tables/roc_pvalue_summary.csv', header = T, row.names = 1)
+IF_model_wilcox_paired_summary <- read.csv('../results/tables/IF_model_wicox_paired_pvalue_summary.csv', header = T, row.names = 1)
+confusion_summary <- read.csv('../results/tables/confusion_summary.csv', header = T, row.names = 1)
+follow_up_prob_summary <- read.csv('../results/tables/follow_up_probability_summary.csv', header = T)
+followup_cm <- read.csv('../results/tables/followup_confusion_matrix.csv', header = T, row.names = 1)
+initial_cm <- read.csv('../results/tables/initial_confusion_matrix.csv', header = T, row.names = 1)
+roc_pvalues <- read.csv('../results/tables/roc_pvalue_summary.csv', header = T, row.names = 1)
 
 # Predictive of both initial and follow up and in lesion model
-pvalue_common_imp_vars <- read.csv('results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
+pvalue_common_imp_vars <- read.csv('../results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
 
@@
 : For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```).
-The top 5 most important operational taxonomic units (OTUs) for prediction of lesion classified to ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.
 In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
-Overall, our model accurately predicted those with an adenoma or carcinoma and following treatment it also decreased the positive probability of having a lesion.
+Our model predicted that ```r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% of the 67-person cohort had normal colons and ```r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% had a lesion.
+Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.
+Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.
 
@@
 ## Discussion
-In this study we show that the overall community structure shows very little change between controls and those with either adenoma or carcinoma [Table S1] in our larger training set.
-With respect to the follow up sample group there was very little difference in thetayc distance metric between those with adenoma or carcinoma[Figure 1a].
+In this study we show that the overall community structure as measured by different alpha diversity metrics, shows very little change between controls and those with either adenoma or carcinoma [Table S1] in our larger training set.
+With respect to the follow up sample group there was very little difference in magnitude of change in the thetayc distance metric between those with adenoma or carcinoma[Figure 1a].
 FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].
-When investigating the bacterial community, specifically between these individuals, there was very little change in overall community in the adenoma group but there was a change in the carcinoma group [Figure 2].
+An NMDS showed that there was very little change between initial and follow up for the adenoma group but there was a change in the carcinoma group [Figure 2].
 This information showed that the removal and treatment of carcinoma for these cursory metrics had the largest response.

### 2016-12-05 submission/manuscript_outline_20161024.tex
updated manuscript w/ newer abstract

@@
 (P-value = \texttt{8.48e-07}) and carcinoma (P-value =
-\texttt{5.96e-08}). The top 5 most important operational taxonomic units
-(OTUs) for prediction of lesion classified to
-\texttt{Lachnospiraceae\ (Otu000013)},
-\texttt{Escherichia/Shigella\ (Otu000018)},
-\texttt{Ruminococcaceae\ (Otu000020)},
-\texttt{Ruminococcus\ (Otu000017)}, and
-\texttt{Porphyromonas\ (Otu000153)}. In the test set, the prediction of
-lesion for the initial samples had a sensitivity of \texttt{98.5}\%
-(\texttt{65/66}) while follow up it was \texttt{100}\%
-(\texttt{1\ /\ 1}). Overall, our model accurately predicted those with
-an adenoma or carcinoma and following treatment it also decreased the
-positive probability of having a lesion.
+\texttt{5.96e-08}). In the test set, the prediction of lesion for the
+initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
+while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
+predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
+and \texttt{60.6}\% had a lesion. Our model suggests that treatment does
+significantly reduce the probability of having a colonic lesion. Further
+surveillance of these individuals will enable us to determine whether
+models such as the one we present here can also be used to predict
+recurrence of colorectal cancer.
 
@@
 
-In this study we show that the overall community structure shows very
-little change between controls and those with either adenoma or
-carcinoma {[}Table S1{]} in our larger training set. With respect to the
-follow up sample group there was very little difference in thetayc
+In this study we show that the overall community structure as measured
+by different alpha diversity metrics, shows very little change between
+controls and those with either adenoma or carcinoma {[}Table S1{]} in
+our larger training set. With respect to the follow up sample group
+there was very little difference in magnitude of change in the thetayc
 distance metric between those with adenoma or carcinoma{[}Figure 1a{]}.
 FIT had a large change in the initial and follow up samples in the
-carcinoma group versus the adenoma {[}Figure 1b{]}. When investigating
-the bacterial community, specifically between these individuals, there
-was very little change in overall community in the adenoma group but
-there was a change in the carcinoma group {[}Figure 2{]}. This
-information showed that the removal and treatment of carcinoma for these
-cursory metrics had the largest response. Yet these metrics alone would
-not be able to identify differences in adenomas.
+carcinoma group versus the adenoma {[}Figure 1b{]}. An NMDS showed that
+there was very little change between initial and follow up for the
+adenoma group but there was a change in the carcinoma group {[}Figure
+2{]}. This information showed that the removal and treatment of
+carcinoma for these cursory metrics had the largest response. Yet these
+metrics alone would not be able to identify differences in adenomas.
 

### 2016-12-05 submission/supplemental_outline_20161024.Rmd

@@
 
-source('code/functions.R')
-source('code/graphFunctions.R')
+source('../code/functions.R')
+source('../code/graphFunctions.R')
 
-loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
-           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
@@
 
-alpha_table <- read.csv("results/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
+alpha_table <- read.csv("../results/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
 supplemental_table1 <- separate(alpha_table, X, c("comparison", "alpha_metric"), sep="_")
@@
 
-imp_model_vars <- read.csv('results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
-rf_otu_tax <- read.csv('results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
+imp_model_vars <- read.csv('../results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
+rf_otu_tax <- read.csv('../results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
 rf_otu_tax <- cbind(otu = rownames(rf_otu_tax), rf_otu_tax)
@@
 
-follow_up_prob_summary <- read.csv('results/tables/follow_up_probability_summary.csv', stringsAsFactors = F, header = T)
+follow_up_prob_summary <- read.csv('../results/tables/follow_up_probability_summary.csv', stringsAsFactors = F, header = T)
 
@@
 
-pvalue_common_imp_vars <- read.csv('results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
+pvalue_common_imp_vars <- read.csv('../results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
 
@@
 
-diff_table_time <- read.csv("results/tables/time_datatable.csv", header = T)
+diff_table_time <- read.csv("../results/tables/time_datatable.csv", header = T)
 
@@
 
-ggsave(file = "results/figures/FigureS1.pdf", time_graph, 
+ggsave(file = "../results/figures/FigureS1.pdf", time_graph, 
        width=6, height = 8, dpi = 300)

### 2016-12-05 code/Run_render_paper.R
update render paper script

@@ -9,6 +9,6 @@ source('code/graphFunctions.R')
 loadLibs(c("knitr", "rmarkdown"))
 
 # Render the final pdfs
-render('submission/manuscript_outline_20161024.Rmd', run_pandoc = FALSE, clean = FALSE)
+render('submission/manuscript_outline_20161024.Rmd', clean = FALSE)
 
-render('submission/supplemental_outline_20161024.Rmd', run_pandoc = FALSE, clean=FALSE)
+render('submission/supplemental_outline_20161024.Rmd', clean=FALSE)

### 2016-12-06 submission/manuscript_outline_20161024.tex
temp changes

@@
     \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
 \fi
@@
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
@@
 
+\maketitle
+
 \section{Differences in the Stool Microbiome Before and After Colorectal
@@
 \newpage
-
 \linenumbers
@@
 initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
-while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
+while follow up it was \texttt{100}\% (\texttt{1 / 1}). Our model
 predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
@@
 difference between the AUC of the best and middle training models
-(P-value = \texttt{0.0698}). There was a significant difference between
+(P-value = \texttt{NULL}). There was a significant difference between
 the worse training model and both the best training model (P-value =
-\texttt{0.000397}) and middle training model (P-value =
-\texttt{0.0444}).
+\texttt{NULL}) and middle training model (P-value = \texttt{NULL}).
 
@@
 more than 50\% of the training models {[}Table S2{]}. The top 5 most
-important bacterial OTUs were \texttt{Lachnospiraceae\ (Otu000013)},
-\texttt{Escherichia/Shigella\ (Otu000018)},
-\texttt{Ruminococcaceae\ (Otu000020)},
-\texttt{Ruminococcus\ (Otu000017)}, and
-\texttt{Porphyromonas\ (Otu000153)}. These 5 OTUs were present in at
+important bacterial OTUs were \texttt{Bacteria (Otu000013)},
+\texttt{Escherichia/Shigella (Otu000018)},
+\texttt{Bacteria (Otu000020)}, \texttt{Ruminococcus (Otu000017)}, and
+\texttt{Porphyromonas (Otu000153)}. These 5 OTUs were present in at
 least \texttt{99} out of the total \texttt{100} different 80/20 runs.
@@
 probability their follow up sample was still higher than the cutoff
-threshold of 0.5 (positive probability = \texttt{0.902}). Interestingly,
+threshold of 0.5 (positive probability = \texttt{0.927}). Interestingly,
 \texttt{2} individuals who were unknown on follow up still were over the
@@
 The follow up positive probabilities were not affected by either
-chemotherapy treatment (uncorrected P-value = \texttt{0.583}) or
-radiation therapy (uncorrected P-value = \texttt{0.809}). There was also
-no difference in the amount of change in the positive probability based
-on whether individuals received chemotherapy (unccorected P-value =
-\texttt{0.224}) or radiation therapy (uncorrected P-value =
-\texttt{0.846}).
+chemotherapy treatment (uncorrected P-value = \texttt{0.919}) or
+radiation therapy (uncorrected P-value = \texttt{1}). There was also no
+difference in the amount of change in the positive probability based on
+whether individuals received chemotherapy (unccorected P-value =
+\texttt{0.578}) or radiation therapy (uncorrected P-value =
+\texttt{0.904}).
 
@@
 pathogenesis but rather that the best model does not utilize these
-specific bacteria for prediction purposes.
+specific bacteria for prediction purposes. Another potential reason why
+we did not identify the usual suspects is that they may not change much
+between our initial and follow up samples. That is to say that they are
+consistently present even after removal of the lesion but surgery.
 
@@
 0.75 had one of the two columns removed. This pre-processing was
-performed with the R package caret (v\texttt{6.0.73}). The method of
+performed with the R package caret (v\texttt{6.0.72}). The method of
 choosing the best model involved taking the \texttt{490} samples and
@@
 \textbf{\emph{Statistical Analysis}} The R software package
-(v\texttt{3.3.0}) was used for all statisitical analysis. Comparisons
+(v\texttt{3.3.2}) was used for all statisitical analysis. Comparisons
 between bacterial community structure utilized PERMANOVA (\textbf{insert

### 2016-12-06 submission/manuscript_outline_20161024.Rmd

@@
 That is to say it had the smallest change in AUC in comparison to the other minimum and maximum AUC trained models.
-There was no significant difference between the AUC of the best and middle training models (P-value = ```r format(roc_pvalues["BH_adj", "best_vs_middle"], digits = 3)```).
-There was a significant difference between the worse training model and both the best training model (P-value = ```r format(roc_pvalues["BH_adj", "best_vs_worse"], digits = 3)```) and middle training model (P-value = ```r format(roc_pvalues["BH_adj", "middle_vs_worse"], digits = 3)```).
+There was no significant difference between the AUC of the best and middle training models (P-value = ```r format(roc_pvalues["best", "middle"], digits = 3)```).
+There was also no difference in the middle model versus worse (P-value = ```r format(roc_pvalues["middle", "worse"], digits = 3)```) or full data model (P-value = ```r format(roc_pvalues["middle", "full"], digits = 3)```).
+The two comparisons with a significant difference were between the worse training model and the best training model (P-value = ```r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)```) and the worse training model and full data model (P-value = ```r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)```).
 
@@
 Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the best model does not utilize these specific bacteria for prediction purposes.
+Another potential reason why we did not identify the usual suspects is that they may not change much between our initial and follow up samples.
+That is to say that they are consistently present even after removal of the lesion but surgery.
 

### 2016-12-06 submission/manuscript_outline_20161024.tex

--- 2016-12-05 submission/manuscript_outline_20161024.tex
@@
 difference between the AUC of the best and middle training models
-(P-value = \texttt{0.0698}). There was a significant difference between
-the worse training model and both the best training model (P-value =
-\texttt{0.000397}) and middle training model (P-value =
-\texttt{0.0444}).
+(P-value = \texttt{0.419}). There was also no difference in the middle
+model versus worse (P-value = \texttt{0.178}) or full data model
+(P-value = \texttt{1}). The two comparisons with a significant
+difference were between the worse training model and the best training
+model (P-value = \texttt{7.94e-04}) and the worse training model and
+full data model (P-value = \texttt{3.52e-03}).
 
@@
 more than 50\% of the training models {[}Table S2{]}. The top 5 most
-important bacterial OTUs were \texttt{Lachnospiraceae\ (Otu000013)},
+important bacterial OTUs were \texttt{Bacteria\ (Otu000013)},
 \texttt{Escherichia/Shigella\ (Otu000018)},
-\texttt{Ruminococcaceae\ (Otu000020)},
-\texttt{Ruminococcus\ (Otu000017)}, and
+\texttt{Bacteria\ (Otu000020)}, \texttt{Ruminococcus\ (Otu000017)}, and
 \texttt{Porphyromonas\ (Otu000153)}. These 5 OTUs were present in at

### 2016-12-06 submission/manuscript_outline_20161024.Rmd
manuscript method model update and mcnemar initial

@@
 Although there was a decrease in positive probability only ```r followup_cm["pred_no", "ref_no"]``` of the total ```r sum(initial_cm$ref_yes)``` individuals were classified as adenoma or carcinoma free on follow up (successful classification of ```r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)```%).
-There was no significant difference between the predictions and actual diagnosis as measured by the Mcnemar test (P-value = ```r confusion_summary["McnemarPValue", "initial"]```).
+There was no significant difference between the predictions and actual diagnosis for the initial samples in the 67-sample cohort test set.
 However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = ```r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)```).
@@
 This pre-processing was performed with the R package caret (v```r paste(packageVersion("caret"))```).
-The method of choosing the best model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
+Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
 This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
@@
 This was repeated 100 times with the best model after all the repeats chosen to be used on the ```r length(rownames(good_metaf))``` samples with follow up.
-The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as having or not having a lesion.
+Once the ideal mtry was found the entire ```r length(rownames(metaI))``` sample set was used to create the final Random Forest model.
+The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.
 
 ***Selection of Important OTUs***
-In order to assess which variables were most central to all the models the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) was recorded and those varialbes that were present more than 50% of the time were considered important.
+In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.
+This final collated list of variables was what was considered the most important.
 
-A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples and specifically lesion.
-This new model OTUs were then compared to the overall lesion model and common OTUs were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.
+A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.
+This new model OTUs were then compared to the overall lesion model and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.
 

### 2016-12-06 submission/manuscript_outline_20161024.tex
manuscript method model update and mcnemar initial

@@
 probability of lesion compared to the initial samples for both adenoma
-(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
-\texttt{5.96e-08}). In the test set, the prediction of lesion for the
-initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
+(P-value = \texttt{7.95e-08}) and carcinoma (P-value =
+\texttt{7.95e-08}). In the test set, the prediction of lesion for the
+initial samples had a sensitivity of \texttt{100}\% (\texttt{66/66})
 while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
-predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
-and \texttt{60.6}\% had a lesion. Our model suggests that treatment does
+predicted that \texttt{36.4}\% of the 67-person cohort had normal colons
+and \texttt{63.6}\% had a lesion. Our model suggests that treatment does
 significantly reduce the probability of having a colonic lesion. Further
@@
 correction there was a significant overall decrease in positive
-probability of a carcinoma and adenoma (P-value = \texttt{8.27e-11})
+probability of a carcinoma and adenoma (P-value = \texttt{1.14e-11})
 {[}Figure 4{]}. This decrease was significant for both adenoma (P-value
-= \texttt{8.48e-07} {[}Figure 4a{]} and carcinoma (P-value =
-\texttt{5.96e-08}) {[}Figure 4b{]} alone. This also held specifically
+= \texttt{7.95e-08} {[}Figure 4a{]} and carcinoma (P-value =
+\texttt{7.95e-08}) {[}Figure 4b{]} alone. This also held specifically
 for those with screen relative neoplasias (SRN) (P-value =
-\texttt{1.45e-04}). A total of \texttt{65} or \texttt{98.5}\% of all
+\texttt{5.35e-04}). A total of \texttt{66} or \texttt{100}\% of all
 samples were correctly predicted to have an adenoma or carcinoma.
-Although there was a decrease in positive probability only \texttt{26}
+Although there was a decrease in positive probability only \texttt{24}
 of the total \texttt{66} individuals were classified as adenoma or
 carcinoma free on follow up (successful classification of
-\texttt{40.9}\%). There was no significant difference between the
-predictions and actual diagnosis as measured by the Mcnemar test
-(P-value = \texttt{1}). However, the predictions were significantly
+\texttt{37.9}\%). There was no significant difference between the
+predictions and actual diagnosis for the initial samples in the
+67-sample cohort test set. However, the predictions were significantly
 discordent with the diagnosis for the follow up samples (P-value =
-\texttt{1.17e-09}). Although there were discordent results the
-respective sensitivity for the initial group was \texttt{98.5}\% and for
+\texttt{4.19e-10}). Although there were discordent results the
+respective sensitivity for the initial group was \texttt{100}\% and for
 follow up was \texttt{100}\%, respectively.
@@
 of whether the sample as a carcinoma or adenoma. The overall sensitivity
-for lesion detection in the intial samples was \texttt{98.5} and for
+for lesion detection in the intial samples was \texttt{100} and for
 follow ups was \texttt{100}. Although there was a decrease in overall
-probability of an adenoma or carcinoma only \texttt{26} were below the
+probability of an adenoma or carcinoma only \texttt{24} were below the
 0.5 threshold out of the total \texttt{65} individuals who were
@@
 0.75 had one of the two columns removed. This pre-processing was
-performed with the R package caret (v\texttt{6.0.73}). The method of
-choosing the best model involved taking the \texttt{490} samples and
+performed with the R package caret (v\texttt{6.0.73}). Optimiation of
+the mtry hyperparameter as well as data on the best and worse
+performance of the model involved taking the \texttt{490} samples and
 making 100 80/20 (train/test) splits in the data where control and
@@
 times with the best model after all the repeats chosen to be used on the
-\texttt{67} samples with follow up. The default cutoff of a probability
-of 0.5 was used as the threshold to classify individuals as having or
-not having a lesion.
+\texttt{67} samples with follow up. Once the ideal mtry was found the
+entire \texttt{490} sample set was used to create the final Random
+Forest model. The default cutoff of a probability of 0.5 was used as the
+threshold to classify individuals as positive or negative of lesion.
 
 \textbf{\emph{Selection of Important OTUs}} In order to assess which
-variables were most central to all the models the number of times a
-variable was present in the top 10\% of mean decrease in accuracy (MDA)
-was recorded and those varialbes that were present more than 50\% of the
-time were considered important.
+variables were most central to all the models we counted the number of
+times a variable was present in the top 10\% of mean decrease in
+accuracy (MDA) for each different 80/20 split model and then filtered
+this list to variables that were only present more than 50\% of the
+time. This final collated list of variables was what was considered the
+most important.
 
 A second model using the same types of conditions was used to identify
-OTUs that could predict initial and follow up samples and specifically
-lesion. This new model OTUs were then compared to the overall lesion
-model and common OTUs were selected to test for changes between initial
-and follow up using a paired wilcoxson rank sum test.
+OTUs that could predict initial and follow up samples. This new model
+OTUs were then compared to the overall lesion model and the resulting
+common OTUs to both models were selected to test for changes between
+initial and follow up using a paired wilcoxson rank sum test.
 

### 2016-12-06 submission/manuscript_outline_20161024.Rmd
manuscript update discssion

@@
 ## Discussion
-In this study we show that the overall community structure as measured by different alpha diversity metrics, shows very little change between controls and those with either adenoma or carcinoma [Table S1] in our larger training set.
-With respect to the follow up sample group there was very little difference in magnitude of change in the thetayc distance metric between those with adenoma or carcinoma[Figure 1a].
-FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].
-An NMDS showed that there was very little change between initial and follow up for the adenoma group but there was a change in the carcinoma group [Figure 2].
-This information showed that the removal and treatment of carcinoma for these cursory metrics had the largest response.
-Yet these metrics alone would not be able to identify differences in adenomas.
+In our training set we show that the overall community structure as measured by different alpha diversity metrics, shows very little change between controls and those with either adenoma or carcinoma [Table S1].
+With respect to our test set there was very little difference in magnitude of change in the thetayc distance metric between those with adenoma or carcinoma [Figure 1a].
+In contrast, FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].
+An NMDS showed that there was very little observable change between initial and follow up for the adenoma group but there was one for the carcinoma group [Figure 2].
+This cursory information is suggestive that treatment of carcinoma, had the largest response.
 

### 2016-12-06 submission/manuscript_outline_20161024.tex
manuscript update discssion

@@
 
-In this study we show that the overall community structure as measured
-by different alpha diversity metrics, shows very little change between
-controls and those with either adenoma or carcinoma {[}Table S1{]} in
-our larger training set. With respect to the follow up sample group
-there was very little difference in magnitude of change in the thetayc
-distance metric between those with adenoma or carcinoma{[}Figure 1a{]}.
-FIT had a large change in the initial and follow up samples in the
-carcinoma group versus the adenoma {[}Figure 1b{]}. An NMDS showed that
-there was very little change between initial and follow up for the
-adenoma group but there was a change in the carcinoma group {[}Figure
-2{]}. This information showed that the removal and treatment of
-carcinoma for these cursory metrics had the largest response. Yet these
-metrics alone would not be able to identify differences in adenomas.
+In our training set we show that the overall community structure as
+measured by different alpha diversity metrics, shows very little change
+between controls and those with either adenoma or carcinoma {[}Table
+S1{]}. With respect to our test set there was very little difference in
+magnitude of change in the thetayc distance metric between those with
+adenoma or carcinoma {[}Figure 1a{]}. In contrast, FIT had a large
+change in the initial and follow up samples in the carcinoma group
+versus the adenoma {[}Figure 1b{]}. An NMDS showed that there was very
+little observable change between initial and follow up for the adenoma
+group but there was one for the carcinoma group {[}Figure 2{]}. This
+cursory information is suggestive that treatment of carcinoma, had the
+largest response.
 

### 2016-12-07 submission/manuscript_outline_20161024.tex
update beta diversity errors

@@
 the directionality of the change was significant in the cancer group
-between initial and follow up (P-value = \texttt{0.001}) but not for the
-adenoma group (P-value = \texttt{1}) {[}Figure 2{]}. When all follow up
-samples were compared to each other there was no significant overall
-difference between them (P-value = \texttt{0.643}). There was no
+between initial and follow up (P-value = \texttt{0.002}) but not for the
+adenoma group (P-value = \texttt{0.997}) {[}Figure 2{]}. When all follow
+up samples were compared to each other there was no significant overall
+difference between them (P-value = \texttt{0.085}). There was no
 significant difference between initial and follow up samples for

### 2016-12-07 code/Run_Beta_Diversity_tests.R

@@ -85,9 +85,8 @@ beta_diver_summary['Adenoma', ] <- c(
 # Create NMDS axis of Adenoma only from thetayc distances
 set.seed(050416)
 thetayc.mds.List <- list(
-  bdiver_Test_adn_IF = metaMDS(as.dist(polyp_only_theta_init_follow_dist), 
-    trace = 0) %>% scores() %>% as.data.frame() %>%  
-  mutate(samples = factor(breakDown_samples)))
+  bdiver_Test_adn_IF = metaMDS(as.dist(polyp_only_theta_init_follow_dist), trymax = 3000) %>% 
+    scores() %>% as.data.frame() %>% mutate(samples = factor(breakDown_samples)))
 
 # Load in needed Carcinoma data
 crc_only_theta_init_follow_dist <- pickDistanceValues(
@@ -115,7 +114,7 @@ beta_diver_summary['Carcinoma', ] <- c(
 # Create NMDS axis of Carcinoma only from thetayc distances
 set.seed(050416)
 thetayc.mds.List[["bdiver_Test_crc_IF"]] <- metaMDS(
-  as.dist(crc_only_theta_init_follow_dist), trace = 0) %>% 
+  as.dist(crc_only_theta_init_follow_dist), trymax = 1000) %>% 
   scores() %>% as.data.frame() %>% 
   mutate(samples = factor(breakDown_samples))
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd

@@
 In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.
-This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
-The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.
-: For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```).
-In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).
-Our model predicted that ```r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% of the 67-person cohort had normal colons and ```r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% had a lesion.
+This model was tested on a `r sum(initial_cm$ref_yes)` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
+The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
+For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)`) and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)`).
+Our model predicted that `r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)`% of the 67-person cohort had normal colons and `r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)`% had a lesion.
+Some OTUs that changed the most before and after treatment included OTUs that were affiliated with members of `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][1:2]` and `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][3]`.
 Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.
@@
 ***Bacterial Community and Fit Changes before and after Treatment***
-Based on thetayc distance metrics comparing the initial to the follow up samples there was no difference between the adenoma and carcinoma group (P-value = ```r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)```) [Figure 1a].
-There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = ```r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)```) [Figure 1b].
-Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = ```r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)```) but not for the adenoma group (P-value = ```r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)```) [Figure 2].
-When all follow up samples were compared to each other there was no significant overall difference between them (P-value = ```r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)```).
+Based on thetayc distance metrics comparing the initial to the follow up samples there was no difference between the adenoma and carcinoma group (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
+There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
+Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) [Figure 2].
+When all follow up samples were compared to each other there was no significant overall difference between them (P-value = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
 There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, and evenness after correction for multiple comparisons [Table S1].
-Time of follow up sample from initial sampling did not have an effect on our prediction data set (uncorrected P-value = ```r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)```).
+Time of follow up sample from initial sampling did not have an effect on our prediction data set (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
 
 ***Outcome of Model Training***
-The range of the AUC for model training ranged from a minimum of ```r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)``` to a maximum of ```r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)```.
-To be conservative the model chosen for prediction of disease in the follow up samples had an AUC in the middle of all the ```r length(rownames(model_train_summary))``` runs which was ```r format(model_test_AUC_summary["middle", "ROC_cv"],
-digits = 3)```.
+The range of the AUC for model training ranged from a minimum of `r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)` to a maximum of `r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)`.
+To be conservative the model chosen for prediction of disease in the follow up samples had an AUC in the middle of all the `r length(rownames(model_train_summary))` runs which was `r format(model_test_AUC_summary["middle", "ROC_cv"],
+digits = 3)`.
 Interestingly, the worst AUC model from training performed the best on it's respective 80/20 split test data [Figure 3].
-In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = ```r format(model_test_AUC_summary["best", "AUC"], digits = 3)```, middle training model test set AUC = ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```, worse training model test set AUC = ```r format(model_test_AUC_summary["worse", "AUC"], digits = 3)```).
+In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, middle training model test set AUC = `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, worse training model test set AUC = `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`).
 That is to say it had the smallest change in AUC in comparison to the other minimum and maximum AUC trained models.
-There was no significant difference between the AUC of the best and middle training models (P-value = ```r format(roc_pvalues["best", "middle"], digits = 3)```).
-There was also no difference in the middle model versus worse (P-value = ```r format(roc_pvalues["middle", "worse"], digits = 3)```) or full data model (P-value = ```r format(roc_pvalues["middle", "full"], digits = 3)```).
-The two comparisons with a significant difference were between the worse training model and the best training model (P-value = ```r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)```) and the worse training model and full data model (P-value = ```r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)```).
+There was no significant difference between the AUC of the best and middle training models (P-value = `r format(roc_pvalues["best", "middle"], digits = 3)`).
+There was also no difference in the middle model versus worse (P-value = `r format(roc_pvalues["middle", "worse"], digits = 3)`) or full data model (P-value = `r format(roc_pvalues["middle", "full"], digits = 3)`).
+The two comparisons with a significant difference were between the worse training model and the best training model (P-value = `r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)`) and the worse training model and full data model (P-value = `r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)`).
 
 ***Most Important Variables to the Model***
-Overall, there were a total of ```r format(length(rownames(imp_model_vars)))``` variables identified as being present in more than 50% of the training models [Table S2].
-The top 5 most important bacterial OTUs were ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.
-These 5 OTUs were present in at least ```r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])``` out of the total ```r length(rownames(model_train_summary))``` different 80/20 runs.
+Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Table S2].
+The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.
+These 5 OTUs were present in at least `r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])` out of the total `r length(rownames(model_train_summary))` different 80/20 runs.
 
 ***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
-A total of ```r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))``` sample was omitted from the original ```r length(rownames(good_metaf))``` sample set since it was missing a complete set of follow up data.
-This left a total of ```r sum(initial_cm$ref_yes)``` samples for test predictions.
-After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4].
-This decrease was significant for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)``` [Figure 4a] and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4b] alone.
-This also held specifically for those with screen relative neoplasias (SRN) (P-value = ```r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)```).
-A total of ```r initial_cm["pred_yes", "ref_yes"]``` or ```r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)```% of all samples were correctly predicted to have an adenoma or carcinoma.
-Although there was a decrease in positive probability only ```r followup_cm["pred_no", "ref_no"]``` of the total ```r sum(initial_cm$ref_yes)``` individuals were classified as adenoma or carcinoma free on follow up (successful classification of ```r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)```%).
+A total of `r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))` sample was omitted from the original `r length(rownames(good_metaf))` sample set since it was missing a complete set of follow up data.
+This left a total of `r sum(initial_cm$ref_yes)` samples for test predictions.
+After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = `r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4].
+This decrease was significant for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)` [Figure 4a] and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4b] alone.
+This also held specifically for those with screen relative neoplasias (SRN) (P-value = `r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)`).
+A total of `r initial_cm["pred_yes", "ref_yes"]` or `r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)`% of all samples were correctly predicted to have an adenoma or carcinoma.
+Although there was a decrease in positive probability only `r followup_cm["pred_no", "ref_no"]` of the total `r sum(initial_cm$ref_yes)` individuals were classified as adenoma or carcinoma free on follow up (successful classification of `r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)`%).
 There was no significant difference between the predictions and actual diagnosis for the initial samples in the 67-sample cohort test set.
-However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = ```r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)```).
-Although there were discordent results the respective sensitivity for the initial group was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% and for follow up was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```%, respectively.
+However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = `r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)`).
+Although there were discordent results the respective sensitivity for the initial group was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)`% and for follow up was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`%, respectively.
 
-There was ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual who still clearly had CRC on follow up as well as ```r length(rownames(filter(good_metaf, Disease_Free == "unknown")))``` individuals whose status on follow up was unknown.
-Although the ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = ```r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)```).
-Interestingly, ```r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))``` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].
+There was `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual who still clearly had CRC on follow up as well as `r length(rownames(filter(good_metaf, Disease_Free == "unknown")))` individuals whose status on follow up was unknown.
+Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).
+Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].
 
-The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)```) or  radiation therapy (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)```).
-There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)```) or radiation therapy (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)```).
+The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)`) or  radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).
+There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).
 
 ***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
-Overall, there were a total of ```r length(rownames(pvalue_common_imp_vars))``` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].
-A total of ```r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))``` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```.
-In general, ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]``` (```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```) was decreased from initial to follow up [Figure 5].
+Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].
+A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
+In general, `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was decreased from initial to follow up [Figure 5].
 The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].
@@
 There was a signficant decrease in positive probability regardless of whether the sample as a carcinoma or adenoma.
-The overall sensitivity for lesion detection in the intial samples was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)``` and for follow ups was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```.
-Although there was a decrease in overall probability of an adenoma or carcinoma only ```r followup_cm["pred_no", "ref_no"]``` were below the 0.5 threshold out of the total ```r sum(followup_cm$ref_no)``` individuals who were diagnosed as not having a carcinoma on follow up.
+The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.
+Although there was a decrease in overall probability of an adenoma or carcinoma only `r followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.
 
 We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].
-Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]``` was to ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```.
+Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
 Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].
@@
 The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).
-Kept apart from this training set were a total of ```r length(rownames(good_metaf))``` individuals that not only had a sample as described previoulsy but also a follow up sample between ```r min(good_metaf$time)``` - ```r max(good_metaf$time)``` days after surgery and treatment had been completed.
+Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
 This study was approved by the University of Michigan Institutional Review Board.
@@
 The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
-The numer of sequences for each sample was rarified to ```r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))``` in an attempt to minimize uneven sampling.
+The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.
 
 ***Model Creation***
-The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the ```r length(rownames(good_metaf))``` individuals with follow up samples.
+The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.
 The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer.
@@
 Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.
-This pre-processing was performed with the R package caret (v```r paste(packageVersion("caret"))```).
-Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
+This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
+Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
 This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
 This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
-This was repeated 100 times with the best model after all the repeats chosen to be used on the ```r length(rownames(good_metaf))``` samples with follow up.
-Once the ideal mtry was found the entire ```r length(rownames(metaI))``` sample set was used to create the final Random Forest model.
+This was repeated 100 times with the best model after all the repeats chosen to be used on the `r length(rownames(good_metaf))` samples with follow up.
+Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model.
 The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.
@@
 ***Statistical Analysis***
-The R software package (v```r paste(version$major, ".", version$minor, sep = "")```) was used for all statisitical analysis.
-Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v```r paste(packageVersion("vegan"))```) while comparisons between ROC curves utilized the method by DeLong et al.
-(**insert citation**) executed by the pROC (v```r paste(packageVersion("pROC"))```) package.
+The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.
+Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al.
+(**insert citation**) executed by the pROC (v`r paste(packageVersion("pROC"))`) package.
 Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.
@@
 
-**Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = ```r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)```).
-B) A significant difference was found between the adenoma and carcinoma group for Fit (P-value = ```r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)```).
+**Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+B) A significant difference was found between the adenoma and carcinoma group for Fit (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 
@@
 
-**Figure 3: Graph of the Receiver Operating Characteristic Curve on Test Set Performance of the Best, Middle, and Worse Training Models.** For each of the 100 training cohort sets used had ```r round(length(rownames(metaI))*0.8)``` individuals and the testing cohort sets had ```r round(length(rownames(good_metaf))*0.2)``` individuals.
-The AUC on the test sets for the best, middle, and worse models from training were ```r format(model_test_AUC_summary["best", "AUC"], digits = 3)```, ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```, and ```r format(model_test_AUC_summary["worse", "AUC"], digits = 3)```, respectively.
+**Figure 3: Graph of the Receiver Operating Characteristic Curve on Test Set Performance of the Best, Middle, and Worse Training Models.** For each of the 100 training cohort sets used had `r round(length(rownames(metaI))*0.8)` individuals and the testing cohort sets had `r round(length(rownames(good_metaf))*0.2)` individuals.
+The AUC on the test sets for the best, middle, and worse models from training were `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, and `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`, respectively.
 cvAUC is the 10-fold cross-validated AUC from training.
@@
 
-**Figure 5: Lesion Model OTU with a Significant Decrease in Relative Abundance that is also Predictive of Initial and Follow Up.**  After multiple comparison correction ```r rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05))``` (```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```) was the only one with a P-value < 0.05.
+**Figure 5: Lesion Model OTU with a Significant Decrease in Relative Abundance that is also Predictive of Initial and Follow Up.**  After multiple comparison correction `r rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05))` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was the only one with a P-value < 0.05.
 The dotted line represents the average relative abundance in the control training group.

### 2016-12-07 submission/manuscript_outline_20161024.tex

--- 2016-12-06 submission/manuscript_outline_20161024.tex
@@
 for lesion (adenoma or carcinoma) after surgical treatment. This model
-was tested on a \texttt{66} person group that included samples before
-and after treatment to allow for the assessment of how the model adjusts
-risk after treatment. The model chosen had a 10-fold cross validated AUC
-of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
-up samples our Random Forest model significantly decreased the positive
-probability of lesion compared to the initial samples for both adenoma
-(P-value = \texttt{7.95e-08}) and carcinoma (P-value =
-\texttt{7.95e-08}). In the test set, the prediction of lesion for the
-initial samples had a sensitivity of \texttt{100}\% (\texttt{66/66})
-while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
-predicted that \texttt{36.4}\% of the 67-person cohort had normal colons
-and \texttt{63.6}\% had a lesion. Our model suggests that treatment does
-significantly reduce the probability of having a colonic lesion. Further
-surveillance of these individuals will enable us to determine whether
-models such as the one we present here can also be used to predict
-recurrence of colorectal cancer.
+was tested on a 66 person group that included samples before and after
+treatment to allow for the assessment of how the model adjusts risk
+after treatment. The model used for prediction had an AUC of 0.813. For
+the follow up samples our Random Forest model significantly decreased
+the positive probability of lesion compared to the initial samples for
+both adenoma (P-value = 7.95e-08) and carcinoma (P-value = 7.95e-08).
+Our model predicted that 36.4\% of the 67-person cohort had normal
+colons and 63.6\% had a lesion. Some OTUs that changed the most before
+and after treatment included OTUs that were affiliated with members of
+Blautia, Clostridium\_XlVa and Escherichia/Shigella. Our model suggests
+that treatment does significantly reduce the probability of having a
+colonic lesion. Further surveillance of these individuals will enable us
+to determine whether models such as the one we present here can also be
+used to predict recurrence of colorectal cancer.
 
@@
 the follow up samples there was no difference between the adenoma and
-carcinoma group (P-value = \texttt{0.697}) {[}Figure 1a{]}. There was a
+carcinoma group (P-value = 0.697) {[}Figure 1a{]}. There was a
 difference in FIT between initial and follow up samples with the
 carcinoma group having a significant decrease in FIT versus the adenoma
-group (P-value = \texttt{2.15e-05}) {[}Figure 1b{]}. Although the
-thetayc distance metric change was similar between adenoma and carcinoma
-the directionality of the change was significant in the cancer group
-between initial and follow up (P-value = \texttt{0.001}) but not for the
-adenoma group (P-value = \texttt{1}) {[}Figure 2{]}. When all follow up
-samples were compared to each other there was no significant overall
-difference between them (P-value = \texttt{0.643}). There was no
-significant difference between initial and follow up samples for
-observed OTUs, Shannon diversity, and evenness after correction for
-multiple comparisons {[}Table S1{]}. Time of follow up sample from
-initial sampling did not have an effect on our prediction data set
-(uncorrected P-value = \texttt{0.784}).
+group (P-value = 2.15e-05) {[}Figure 1b{]}. Although the thetayc
+distance metric change was similar between adenoma and carcinoma the
+directionality of the change was significant in the cancer group between
+initial and follow up (P-value = 0.002) but not for the adenoma group
+(P-value = 0.997) {[}Figure 2{]}. When all follow up samples were
+compared to each other there was no significant overall difference
+between them (P-value = 0.085). There was no significant difference
+between initial and follow up samples for observed OTUs, Shannon
+diversity, and evenness after correction for multiple comparisons
+{[}Table S1{]}. Time of follow up sample from initial sampling did not
+have an effect on our prediction data set (uncorrected P-value = 0.784).
 
 \textbf{\emph{Outcome of Model Training}} The range of the AUC for model
-training ranged from a minimum of \texttt{0.795} to a maximum of
-\texttt{0.854}. To be conservative the model chosen for prediction of
-disease in the follow up samples had an AUC in the middle of all the
-\texttt{100} runs which was \texttt{0.819}. Interestingly, the worst AUC
-model from training performed the best on it's respective 80/20 split
-test data {[}Figure 3{]}. In fact the 80/20 test performance showed that
-the AUC for the middle model chosen was the most stable (best training
-model test set AUC = \texttt{0.682}, middle training model test set AUC
-= \texttt{0.815}, worse training model test set AUC = \texttt{0.931}).
+training ranged from a minimum of 0.795 to a maximum of 0.854. To be
+conservative the model chosen for prediction of disease in the follow up
+samples had an AUC in the middle of all the 100 runs which was 0.819.
+Interestingly, the worst AUC model from training performed the best on
+it's respective 80/20 split test data {[}Figure 3{]}. In fact the 80/20
+test performance showed that the AUC for the middle model chosen was the
+most stable (best training model test set AUC = 0.682, middle training
+model test set AUC = 0.815, worse training model test set AUC = 0.931).
 That is to say it had the smallest change in AUC in comparison to the
@@
 difference between the AUC of the best and middle training models
-(P-value = \texttt{0.419}). There was also no difference in the middle
-model versus worse (P-value = \texttt{0.178}) or full data model
-(P-value = \texttt{1}). The two comparisons with a significant
-difference were between the worse training model and the best training
-model (P-value = \texttt{7.94e-04}) and the worse training model and
-full data model (P-value = \texttt{3.52e-03}).
+(P-value = 0.419). There was also no difference in the middle model
+versus worse (P-value = 0.178) or full data model (P-value = 1). The two
+comparisons with a significant difference were between the worse
+training model and the best training model (P-value = 7.94e-04) and the
+worse training model and full data model (P-value = 3.52e-03).
 
 \textbf{\emph{Most Important Variables to the Model}} Overall, there
-were a total of \texttt{32} variables identified as being present in
-more than 50\% of the training models {[}Table S2{]}. The top 5 most
-important bacterial OTUs were \texttt{Bacteria\ (Otu000013)},
-\texttt{Escherichia/Shigella\ (Otu000018)},
-\texttt{Bacteria\ (Otu000020)}, \texttt{Ruminococcus\ (Otu000017)}, and
-\texttt{Porphyromonas\ (Otu000153)}. These 5 OTUs were present in at
-least \texttt{99} out of the total \texttt{100} different 80/20 runs.
+were a total of 32 variables identified as being present in more than
+50\% of the training models {[}Table S2{]}. The top 5 most important
+bacterial OTUs were Bacteria (Otu000013), Escherichia/Shigella
+(Otu000018), Bacteria (Otu000020), Ruminococcus (Otu000017), and
+Porphyromonas (Otu000153). These 5 OTUs were present in at least 99 out
+of the total 100 different 80/20 runs.
 
 \textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
-Decrease in Positive Probability Prediction}} A total of \texttt{1}
-sample was omitted from the original \texttt{67} sample set since it was
-missing a complete set of follow up data. This left a total of
-\texttt{66} samples for test predictions. After multiple comparison
-correction there was a significant overall decrease in positive
-probability of a carcinoma and adenoma (P-value = \texttt{1.14e-11})
-{[}Figure 4{]}. This decrease was significant for both adenoma (P-value
-= \texttt{7.95e-08} {[}Figure 4a{]} and carcinoma (P-value =
-\texttt{7.95e-08}) {[}Figure 4b{]} alone. This also held specifically
-for those with screen relative neoplasias (SRN) (P-value =
-\texttt{5.35e-04}). A total of \texttt{66} or \texttt{100}\% of all
-samples were correctly predicted to have an adenoma or carcinoma.
-Although there was a decrease in positive probability only \texttt{24}
-of the total \texttt{66} individuals were classified as adenoma or
-carcinoma free on follow up (successful classification of
-\texttt{37.9}\%). There was no significant difference between the
-predictions and actual diagnosis for the initial samples in the
+Decrease in Positive Probability Prediction}} A total of 1 sample was
+omitted from the original 67 sample set since it was missing a complete
+set of follow up data. This left a total of 66 samples for test
+predictions. After multiple comparison correction there was a
+significant overall decrease in positive probability of a carcinoma and
+adenoma (P-value = 1.14e-11) {[}Figure 4{]}. This decrease was
+significant for both adenoma (P-value = 7.95e-08 {[}Figure 4a{]} and
+carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This also held
+specifically for those with screen relative neoplasias (SRN) (P-value =
+5.35e-04). A total of 66 or 100\% of all samples were correctly
+predicted to have an adenoma or carcinoma. Although there was a decrease
+in positive probability only 24 of the total 66 individuals were
+classified as adenoma or carcinoma free on follow up (successful
+classification of 37.9\%). There was no significant difference between
+the predictions and actual diagnosis for the initial samples in the
 67-sample cohort test set. However, the predictions were significantly
 discordent with the diagnosis for the follow up samples (P-value =
-\texttt{4.19e-10}). Although there were discordent results the
-respective sensitivity for the initial group was \texttt{100}\% and for
-follow up was \texttt{100}\%, respectively.
+4.19e-10). Although there were discordent results the respective
+sensitivity for the initial group was 100\% and for follow up was 100\%,
+respectively.
 
-There was \texttt{1} individual who still clearly had CRC on follow up
-as well as \texttt{5} individuals whose status on follow up was unknown.
-Although the \texttt{1} individual had a decrease in positive
-probability their follow up sample was still higher than the cutoff
-threshold of 0.5 (positive probability = \texttt{0.927}). Interestingly,
-\texttt{2} individuals who were unknown on follow up still were over the
-threshold cutoff of 0.5 even though, like the \texttt{1} individual with
-clear CRC on follow up, the probability of an adenoma or carcinoma
-decreased {[}Table S3{]}.
+There was 1 individual who still clearly had CRC on follow up as well as
+5 individuals whose status on follow up was unknown. Although the 1
+individual had a decrease in positive probability their follow up sample
+was still higher than the cutoff threshold of 0.5 (positive probability
+= 0.927). Interestingly, 2 individuals who were unknown on follow up
+still were over the threshold cutoff of 0.5 even though, like the 1
+individual with clear CRC on follow up, the probability of an adenoma or
+carcinoma decreased {[}Table S3{]}.
 
 The follow up positive probabilities were not affected by either
-chemotherapy treatment (uncorrected P-value = \texttt{0.919}) or
-radiation therapy (uncorrected P-value = \texttt{1}). There was also no
-difference in the amount of change in the positive probability based on
-whether individuals received chemotherapy (unccorected P-value =
-\texttt{0.578}) or radiation therapy (uncorrected P-value =
-\texttt{0.904}).
+chemotherapy treatment (uncorrected P-value = 0.919) or radiation
+therapy (uncorrected P-value = 1). There was also no difference in the
+amount of change in the positive probability based on whether
+individuals received chemotherapy (unccorected P-value = 0.578) or
+radiation therapy (uncorrected P-value = 0.904).
 
 \textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
-Follow Up Versus Initial Samples}} Overall, there were a total of
-\texttt{8} OTUs that were common between the main lesion model and the
-model for classifying initial and follow up samples specifically
-{[}Table S4{]}. A total of \texttt{1} OTU was still significant after
-multiple comparison correction and it's lowest taxonomic identification
-was to \texttt{Blautia}. In general, \texttt{Otu000012}
-(\texttt{Blautia}) was decreased from initial to follow up {[}Figure
-5{]}. The relative abundance was not drastically different then the mean
-of the values observed in the control training set {[}Figure 5{]}.
+Follow Up Versus Initial Samples}} Overall, there were a total of 8 OTUs
+that were common between the main lesion model and the model for
+classifying initial and follow up samples specifically {[}Table S4{]}. A
+total of 1 OTU was still significant after multiple comparison
+correction and it's lowest taxonomic identification was to Blautia. In
+general, Otu000012 (Blautia) was decreased from initial to follow up
+{[}Figure 5{]}. The relative abundance was not drastically different
+then the mean of the values observed in the control training set
+{[}Figure 5{]}.
 
@@
 of whether the sample as a carcinoma or adenoma. The overall sensitivity
-for lesion detection in the intial samples was \texttt{100} and for
-follow ups was \texttt{100}. Although there was a decrease in overall
-probability of an adenoma or carcinoma only \texttt{24} were below the
-0.5 threshold out of the total \texttt{65} individuals who were
-diagnosed as not having a carcinoma on follow up.
+for lesion detection in the intial samples was 100 and for follow ups
+was 100. Although there was a decrease in overall probability of an
+adenoma or carcinoma only 24 were below the 0.5 threshold out of the
+total 65 individuals who were diagnosed as not having a carcinoma on
+follow up.
 
@@
 after multiple comparison correction and the lowest taxonomic
-identification of \texttt{Otu000012} was to \texttt{Blautia}. Although
-there was a difference in the relative abundance at initial and follow
-up these values were not drastically different from the relative
-abundance values observed in the control individuals of the training set
-{[}Figure 5{]}. This research provides evidence that it is possible to
-create a highily sensitive model for detection of adenoma or carcinoma
-with bacterial microbiome data. It accomplishes this by using a unique
-sample set in which before and after stool samples are available for
-assessment. By using these types of samples we are able to show that
-this model is reactive. That is to say that after surgery for removal of
-the adenoma or carcinoma it decreases the positive probability to
-reflect a lower likelihood of the individual having an adenoma or
-carcinoma.
+identification of Otu000012 was to Blautia. Although there was a
+difference in the relative abundance at initial and follow up these
+values were not drastically different from the relative abundance values
+observed in the control individuals of the training set {[}Figure 5{]}.
+This research provides evidence that it is possible to create a highily
+sensitive model for detection of adenoma or carcinoma with bacterial
+microbiome data. It accomplishes this by using a unique sample set in
+which before and after stool samples are available for assessment. By
+using these types of samples we are able to show that this model is
+reactive. That is to say that after surgery for removal of the adenoma
+or carcinoma it decreases the positive probability to reflect a lower
+likelihood of the individual having an adenoma or carcinoma.
 
@@
 this time period (\textbf{insert citation}). Kept apart from this
-training set were a total of \texttt{67} individuals that not only had a
-sample as described previoulsy but also a follow up sample between
-\texttt{188} - \texttt{546} days after surgery and treatment had been
-completed. This study was approved by the University of Michigan
-Institutional Review Board. All study participants provided informed
-consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+training set were a total of 67 individuals that not only had a sample
+as described previoulsy but also a follow up sample between 188 - 546
+days after surgery and treatment had been completed. This study was
+approved by the University of Michigan Institutional Review Board. All
+study participants provided informed consent and the study itself
+conformed to the guidelines set out by the Helsinki Declaration.
 
@@
 with an average neighbor clustering algorithm. The numer of sequences
-for each sample was rarified to \texttt{10521} in an attempt to minimize
-uneven sampling.
+for each sample was rarified to 10521 in an attempt to minimize uneven
+sampling.
 
@@
 citation}) algorithm was used to create the model used for prediction of
-lesion (adenoma or carcinoma) for the \texttt{67} individuals with
-follow up samples. The model included data on FIT, the bacterial
-microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was
-caucasion or not, history of cancer, and family history of cancer.
-Non-binary data was checked for near zero variance and auto correlation.
-Data columns that had near zero variance were removed. Columns that were
-correlated with each other over a Spearman correlation coefficient of
-0.75 had one of the two columns removed. This pre-processing was
-performed with the R package caret (v\texttt{6.0.73}). Optimiation of
-the mtry hyperparameter as well as data on the best and worse
-performance of the model involved taking the \texttt{490} samples and
-making 100 80/20 (train/test) splits in the data where control and
-lesion were equally represented in the 80 and 20 split, respectively.
-This 80\% portion was then split again into an 80/20 split, and run
-twenty times through 10-fold cross validation to optimzie the model's
-Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
-This resulting model was then tested on the 20\% of the data that was
-originally held out from this overall process. This was repeated 100
-times with the best model after all the repeats chosen to be used on the
-\texttt{67} samples with follow up. Once the ideal mtry was found the
-entire \texttt{490} sample set was used to create the final Random
+lesion (adenoma or carcinoma) for the 67 individuals with follow up
+samples. The model included data on FIT, the bacterial microbiome, sex,
+age, Body Mass Index (BMI), whether the indivdiual was caucasion or not,
+history of cancer, and family history of cancer. Non-binary data was
+checked for near zero variance and auto correlation. Data columns that
+had near zero variance were removed. Columns that were correlated with
+each other over a Spearman correlation coefficient of 0.75 had one of
+the two columns removed. This pre-processing was performed with the R
+package caret (v6.0.73). Optimiation of the mtry hyperparameter as well
+as data on the best and worse performance of the model involved taking
+the 490 samples and making 100 80/20 (train/test) splits in the data
+where control and lesion were equally represented in the 80 and 20
+split, respectively. This 80\% portion was then split again into an
+80/20 split, and run twenty times through 10-fold cross validation to
+optimzie the model's Receiver Operator Characteristic (ROC) and tuneing
+both mtry and ntree. This resulting model was then tested on the 20\% of
+the data that was originally held out from this overall process. This
+was repeated 100 times with the best model after all the repeats chosen
+to be used on the 67 samples with follow up. Once the ideal mtry was
+found the entire 490 sample set was used to create the final Random
 Forest model. The default cutoff of a probability of 0.5 was used as the
@@
 
-\textbf{\emph{Statistical Analysis}} The R software package
-(v\texttt{3.3.0}) was used for all statisitical analysis. Comparisons
-between bacterial community structure utilized PERMANOVA (\textbf{insert
-citation}) in the vegan package (v\texttt{2.4.1}) while comparisons
-between ROC curves utilized the method by DeLong et al. (\textbf{insert
-citation}) executed by the pROC (v\texttt{1.8}) package. Comparisons
-between probabilities as well as overall amount of OTU between initial
-and follow up samples utilized a paired wilcoxson ranked sum test. Where
-multiple comparison testing was needed a Benjamini-Hochberg (BH)
-correction was applied (\textbf{insert citation}) and a corrected
-P-value of less than 0.05 was considered significant. Unless otherwise
-stated the P-values reported are those of the BH corrected ones.
+\textbf{\emph{Statistical Analysis}} The R software package (v3.3.0) was
+used for all statisitical analysis. Comparisons between bacterial
+community structure utilized PERMANOVA (\textbf{insert citation}) in the
+vegan package (v2.4.1) while comparisons between ROC curves utilized the
+method by DeLong et al. (\textbf{insert citation}) executed by the pROC
+(v1.8) package. Comparisons between probabilities as well as overall
+amount of OTU between initial and follow up samples utilized a paired
+wilcoxson ranked sum test. Where multiple comparison testing was needed
+a Benjamini-Hochberg (BH) correction was applied (\textbf{insert
+citation}) and a corrected P-value of less than 0.05 was considered
+significant. Unless otherwise stated the P-values reported are those of
+the BH corrected ones.
 
@@
 found between the adenoma and carcinoma group for thetayc (P-value =
-\texttt{0.697}). B) A significant difference was found between the
-adenoma and carcinoma group for Fit (P-value = \texttt{2.15e-05}).
+0.697). B) A significant difference was found between the adenoma and
+carcinoma group for Fit (P-value = 2.15e-05).
 
@@
 on Test Set Performance of the Best, Middle, and Worse Training Models.}
-For each of the 100 training cohort sets used had \texttt{392}
-individuals and the testing cohort sets had \texttt{13} individuals. The
-AUC on the test sets for the best, middle, and worse models from
-training were \texttt{0.682}, \texttt{0.815}, and \texttt{0.931},
-respectively. cvAUC is the 10-fold cross-validated AUC from training.
+For each of the 100 training cohort sets used had 392 individuals and
+the testing cohort sets had 13 individuals. The AUC on the test sets for
+the best, middle, and worse models from training were 0.682, 0.815, and
+0.931, respectively. cvAUC is the 10-fold cross-validated AUC from
+training.
 
@@
 Relative Abundance that is also Predictive of Initial and Follow Up.}
-After multiple comparison correction \texttt{1} (\texttt{Blautia}) was
-the only one with a P-value \textless{} 0.05. The dotted line represents
-the average relative abundance in the control training group.
+After multiple comparison correction 1 (Blautia) was the only one with a
+P-value \textless{} 0.05. The dotted line represents the average
+relative abundance in the control training group.
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
add discussion changes

@@
 We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).
-We chose the middle training model, based on AUC, from 100 80/20 (train/test) splits.
+Our middle training model, based on AUC, from 100 80/20 (train/test) splits was similar to the full training data model.
 It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].
-Using this model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].
-There was a signficant decrease in positive probability regardless of whether the sample as a carcinoma or adenoma.
+Using the full training data model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].
+There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.
 The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.
@@
 We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].
-Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
+Many of the OTUs identified classified to normal flora bacterium [Table S4].
+Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
 Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].
-This research provides evidence that it is possible to create a highily sensitive model for detection of adenoma or carcinoma with bacterial microbiome data.
-It accomplishes this by using a unique sample set in which before and after stool samples are available for assessment.
-By using these types of samples we are able to show that this model is reactive.
+This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.
+It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.
+By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.
 That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.
@@
 Based on previous work by Jobin, et al.
-(**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for the training models.
+(**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for this model.
 Similarily, Porphyromonas has also been implicated in colorectal cancer (**insert citation**).
-Interestingly, the other OTUs had taxonomic identification for resident gut microbes.
+Interestingly, many of the other OTUs had taxonomic identification for resident gut microbes.
 This could suggest that changes to the resident microbiome are important to the initiation of adenoma or carcinoma formation (**insert citation**) and provide support for the hypothesis that an initial change in the bacterial microbiome could pave the way for more inflammatory species: whether by creation of a new niche for oral microbes (**insert citation**) or allowing for a bloom of existing pro-inflammatory residents (**insert citation**).
@@
 This has been suggested to be the case in a previous study (**insert Baxter Study**).
-Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the best model does not utilize these specific bacteria for prediction purposes.
-Another potential reason why we did not identify the usual suspects is that they may not change much between our initial and follow up samples.
-That is to say that they are consistently present even after removal of the lesion but surgery.
+Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.
+Another potential reason why we did not identify the "usual suspects"" is that these bacteria may not change much between initial and follow up samples in those with an identified lesion.
+That is to say that the bacteria are consistently present even after removal of the lesion by surgery.
+Finally, it is likely that within our test set there was not enough indviduals in which detection  was made or relative abundance high enough for these bacteria to be significant using a paired wilcoxson test.
 
-One limitation is that we do not know whether these individuals eventually had a CRC remission.
-This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having CRC or adenomas.
+One limitation in this study is that we do not know whether individuals in our test set eventually had a subsequent CRC diagnosis.
+This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
 Another limitation is that we do not know if adding modern tests such as the stool DNA test (**insert citation**) could help improve our overall AUC.
+Another limitation is that this study drew heavily from those with caucasian ancestry.
+The results may not be immediately representative of those with either Asian or African ancestry.
 Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.
@@
 By adding patient data such as age, BMI, etc.
-to the model and showing that it can successfully predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.
-Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations.
+to the model and showing that it can successfully help to predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.
+Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
 
@@
 The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures associated with these lesions.
-It also shows that these algorithms successfully react to successful treatment regimens and may be able to one day diagnose CRC with a high level of accuracy.
+It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day diagnose CRC with a high level of accuracy.
 

### 2016-12-07 submission/manuscript_outline_20161024.tex
add discussion changes

@@
 and the bacterial microbiome to be able to predict lesions (adenoma or
-carcinoma). We chose the middle training model, based on AUC, from 100
-80/20 (train/test) splits. It's 10-fold cross validated AUC was similar
-to it's test set AUC which was not the case for both the best and worse
-training model {[}Figure 3{]}. Using this model we predicted the
-probability of a lesion in the inital and follow up samples {[}Figure
-4{]}. There was a signficant decrease in positive probability regardless
-of whether the sample as a carcinoma or adenoma. The overall sensitivity
-for lesion detection in the intial samples was 100 and for follow ups
-was 100. Although there was a decrease in overall probability of an
-adenoma or carcinoma only 24 were below the 0.5 threshold out of the
-total 65 individuals who were diagnosed as not having a carcinoma on
-follow up.
+carcinoma). Our middle training model, based on AUC, from 100 80/20
+(train/test) splits was similar to the full training data model. It's
+10-fold cross validated AUC was similar to it's test set AUC which was
+not the case for both the best and worse training model {[}Figure 3{]}.
+Using the full training data model we predicted the probability of a
+lesion in the inital and follow up samples {[}Figure 4{]}. There was a
+signficant decrease in positive probability regardless of whether the
+sample was a carcinoma or adenoma. The overall sensitivity for lesion
+detection in the intial samples was 100 and for follow ups was 100.
+Although there was a decrease in overall probability of an adenoma or
+carcinoma only 24 were below the 0.5 threshold out of the total 65
+individuals who were diagnosed as not having a carcinoma on follow up.
 
 We then investigated which OTUs could potentially be more important in
-our model {[}Figure 5 \& Table S4{]}. Only a single OTU was significant
-after multiple comparison correction and the lowest taxonomic
-identification of Otu000012 was to Blautia. Although there was a
-difference in the relative abundance at initial and follow up these
-values were not drastically different from the relative abundance values
-observed in the control individuals of the training set {[}Figure 5{]}.
-This research provides evidence that it is possible to create a highily
-sensitive model for detection of adenoma or carcinoma with bacterial
-microbiome data. It accomplishes this by using a unique sample set in
-which before and after stool samples are available for assessment. By
-using these types of samples we are able to show that this model is
-reactive. That is to say that after surgery for removal of the adenoma
-or carcinoma it decreases the positive probability to reflect a lower
-likelihood of the individual having an adenoma or carcinoma.
+our model {[}Figure 5 \& Table S4{]}. Many of the OTUs identified
+classified to normal flora bacterium {[}Table S4{]}. Only a single OTU
+though was significant after multiple comparison correction and the
+lowest taxonomic identification of Otu000012 was to Blautia. Although
+there was a difference in the relative abundance at initial and follow
+up these values were not drastically different from the relative
+abundance values observed in the control individuals of the training set
+{[}Figure 5{]}. This research provides evidence that it is possible to
+use bacterial microbiome data to create a highily sensitive model, that
+is reactive to therapy, for detection of adenoma or carcinoma. It
+accomplishes this by using a unique sample set in which before and after
+surgery stool samples are available for assessment. By using these types
+of samples we are not only able to show sensitivity of lesion prediction
+but also able to show that this model is reactive. That is to say that
+after surgery for removal of the adenoma or carcinoma it decreases the
+positive probability to reflect a lower likelihood of the individual
+having an adenoma or carcinoma.
 
@@
 (\textbf{insert citation}) it may not be surprising to see E.coli in the
-top 5 OTUs for the training models. Similarily, Porphyromonas has also
-been implicated in colorectal cancer (\textbf{insert citation}).
-Interestingly, the other OTUs had taxonomic identification for resident
-gut microbes. This could suggest that changes to the resident microbiome
-are important to the initiation of adenoma or carcinoma formation
-(\textbf{insert citation}) and provide support for the hypothesis that
-an initial change in the bacterial microbiome could pave the way for
-more inflammatory species: whether by creation of a new niche for oral
-microbes (\textbf{insert citation}) or allowing for a bloom of existing
-pro-inflammatory residents (\textbf{insert citation}).
+top 5 OTUs for this model. Similarily, Porphyromonas has also been
+implicated in colorectal cancer (\textbf{insert citation}).
+Interestingly, many of the other OTUs had taxonomic identification for
+resident gut microbes. This could suggest that changes to the resident
+microbiome are important to the initiation of adenoma or carcinoma
+formation (\textbf{insert citation}) and provide support for the
+hypothesis that an initial change in the bacterial microbiome could pave
+the way for more inflammatory species: whether by creation of a new
+niche for oral microbes (\textbf{insert citation}) or allowing for a
+bloom of existing pro-inflammatory residents (\textbf{insert citation}).
 
@@
 against the importance of these bacterium in CRC initiation or
-pathogenesis but rather that the best model does not utilize these
-specific bacteria for prediction purposes. Another potential reason why
-we did not identify the usual suspects is that they may not change much
-between our initial and follow up samples. That is to say that they are
-consistently present even after removal of the lesion but surgery.
+pathogenesis but rather that the model does not utilize these specific
+bacteria for prediction purposes. Another potential reason why we did
+not identify the ``usual suspects''" is that these bacteria may not
+change much between initial and follow up samples in those with an
+identified lesion. That is to say that the bacteria are consistently
+present even after removal of the lesion by surgery. Finally, it is
+likely that within our test set there was not enough indviduals in which
+detection was made or relative abundance high enough for these bacteria
+to be significant using a paired wilcoxson test.
 
-One limitation is that we do not know whether these individuals
-eventually had a CRC remission. This information would help to
-strengthen the case for our Random Forest based model keeping a number
-of individuals above the cutoff threshold even though at follow up they
-were diagnosed as no longer having CRC or adenomas. Another limitation
-is that we do not know if adding modern tests such as the stool DNA test
-(\textbf{insert citation}) could help improve our overall AUC. Finally,
+One limitation in this study is that we do not know whether individuals
+in our test set eventually had a subsequent CRC diagnosis. This
+information would help to strengthen the case for our Random Forest
+based model keeping a number of individuals above the cutoff threshold
+even though at follow up they were diagnosed as no longer having a
+lesion. Another limitation is that we do not know if adding modern tests
+such as the stool DNA test (\textbf{insert citation}) could help improve
+our overall AUC. Another limitation is that this study drew heavily from
+those with caucasian ancestry. The results may not be immediately
+representative of those with either Asian or African ancestry. Finally,
 although our training and test set are relatively large we still run the
@@
 By adding patient data such as age, BMI, etc. to the model and showing
-that it can successfully predict both carcinoma and adenoma our study
-provides further data that these patient factors in conjunction with the
-bacterial microbiome could potentially influence CRC and perhaps have a
-role in formation of adenomas. Further studies need to be carried out to
-verify our findings since not only are we dealing with stool, which
-could be very different than the communities present on the actual
-tissue, but also are dealing with correlations.
+that it can successfully help to predict both carcinoma and adenoma our
+study provides further data that these patient factors in conjunction
+with the bacterial microbiome could potentially influence CRC and
+perhaps have a role in formation of adenomas. Further studies need to be
+carried out to verify our findings since not only are we dealing with
+stool, which could be very different than the communities present on the
+actual tissue, but also are dealing with correlations that may not be
+representative of the true pathogensis of disease.
 
@@
 are specific signatures associated with these lesions. It also shows
-that these algorithms successfully react to successful treatment
-regimens and may be able to one day diagnose CRC with a high level of
-accuracy.
+that these algorithms can not only successfully react to successful
+treatment regimens but also may be able to one day diagnose CRC with a
+high level of accuracy.
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
update methods section

@@
 The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).
-In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cacner, or Familial adenomatous polyposis.
+In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
 Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
@@
 The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.
-Which of each group was randomly assigned to avoid biases based on sample collection location.
+Each group was randomly assigned to avoid biases based on sample collection location.
 
@@
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
-Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
-This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
+Optimiation of the mtry hyperparameter as well as data on the best, middle, and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
+This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).
 This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
-This was repeated 100 times with the best model after all the repeats chosen to be used on the `r length(rownames(good_metaf))` samples with follow up.
-Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model.
-The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.
+Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which testing on the `r length(rownames(good_metaf))`-person cohort was completed.
+The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
 
@@
 In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.
-This final collated list of variables was what was considered the most important.
+This final collated list of variables was what was considered the most important for the lesion prediction model.
 
 A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.
-This new model OTUs were then compared to the overall lesion model and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.
+These OTUs were considered the most important for predicting initial and follow up samples.
+This new model important OTUs were then compared to the overall lesion model important OTUs and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.
 

### 2016-12-07 submission/manuscript_outline_20161024.tex
update methods section

@@
 colorectal cancer before a baseline stool sample could be obtained, had
-IBD, a known hereditary non-polyposis colorectal cacner, or Familial
+IBD, a known hereditary non-polyposis colorectal cancer, or Familial
 adenomatous polyposis. Samples used to build the model used for
@@
 between three sequencing runs with control, adenoma, and carcinoma
-evenly represented on each run. Which of each group was randomly
-assigned to avoid biases based on sample collection location.
+evenly represented on each run. Each group was randomly assigned to
+avoid biases based on sample collection location.
 
@@
 package caret (v6.0.73). Optimiation of the mtry hyperparameter as well
-as data on the best and worse performance of the model involved taking
-the 490 samples and making 100 80/20 (train/test) splits in the data
-where control and lesion were equally represented in the 80 and 20
+as data on the best, middle, and worse performance of the model involved
+taking the 490 samples and making 100 80/20 (train/test) splits in the
+data where control and lesion were equally represented in the 80 and 20
 split, respectively. This 80\% portion was then split again into an
-80/20 split, and run twenty times through 10-fold cross validation to
-optimzie the model's Receiver Operator Characteristic (ROC) and tuneing
-both mtry and ntree. This resulting model was then tested on the 20\% of
-the data that was originally held out from this overall process. This
-was repeated 100 times with the best model after all the repeats chosen
-to be used on the 67 samples with follow up. Once the ideal mtry was
-found the entire 490 sample set was used to create the final Random
-Forest model. The default cutoff of a probability of 0.5 was used as the
-threshold to classify individuals as positive or negative of lesion.
+80/20 split, and run through 20 repeated 10-fold cross validations to
+optimzie the model's AUC (Area Under the Curve of the Receiver Operator
+Characteristic). This resulting model was then tested on the 20\% of the
+data that was originally held out from this overall process. Once the
+ideal mtry was found the entire 490 sample set was used to create the
+final Random Forest model on which testing on the 67-person cohort was
+completed. The default cutoff of 0.5 was used as the threshold to
+classify individuals as positive or negative for lesion.
 
@@
 time. This final collated list of variables was what was considered the
-most important.
+most important for the lesion prediction model.
 
 A second model using the same types of conditions was used to identify
-OTUs that could predict initial and follow up samples. This new model
-OTUs were then compared to the overall lesion model and the resulting
-common OTUs to both models were selected to test for changes between
-initial and follow up using a paired wilcoxson rank sum test.
+OTUs that could predict initial and follow up samples. These OTUs were
+considered the most important for predicting initial and follow up
+samples. This new model important OTUs were then compared to the overall
+lesion model important OTUs and the resulting common OTUs to both models
+were selected to test for changes between initial and follow up using a
+paired wilcoxson rank sum test.
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
Figure legend updates

@@
 **Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
-B) A significant difference was found between the adenoma and carcinoma group for Fit (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+B) A significant difference was found between the adenoma and carcinoma group for FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 
@@
 
-**Figure 3: Graph of the Receiver Operating Characteristic Curve on Test Set Performance of the Best, Middle, and Worse Training Models.** For each of the 100 training cohort sets used had `r round(length(rownames(metaI))*0.8)` individuals and the testing cohort sets had `r round(length(rownames(good_metaf))*0.2)` individuals.
+**Figure 3: Graph of the Receiver Operating Characteristic Curve on Test Set Performance of the Best, Middle, Worse, and Full Training Models.** For each of the 100 training cohort sets used had `r round(length(rownames(metaI))*0.8)` individuals and the testing cohort sets had `r round(length(rownames(good_metaf))*0.2)` individuals.
 The AUC on the test sets for the best, middle, and worse models from training were `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, and `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`, respectively.
-cvAUC is the 10-fold cross-validated AUC from training.
+cvAUC is the 20 times repeated 10-fold cross-validated AUC from training.
 
 **Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Initial and Follow Up***  A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
-THe dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.
+The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.
 

### 2016-12-07 submission/manuscript_outline_20161024.tex
Figure legend updates

@@
 0.697). B) A significant difference was found between the adenoma and
-carcinoma group for Fit (P-value = 2.15e-05).
+carcinoma group for FIT (P-value = 2.15e-05).
 
@@
 \textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
-on Test Set Performance of the Best, Middle, and Worse Training Models.}
-For each of the 100 training cohort sets used had 392 individuals and
-the testing cohort sets had 13 individuals. The AUC on the test sets for
-the best, middle, and worse models from training were 0.682, 0.815, and
-0.931, respectively. cvAUC is the 10-fold cross-validated AUC from
-training.
+on Test Set Performance of the Best, Middle, Worse, and Full Training
+Models.} For each of the 100 training cohort sets used had 392
+individuals and the testing cohort sets had 13 individuals. The AUC on
+the test sets for the best, middle, and worse models from training were
+0.682, 0.815, and 0.931, respectively. cvAUC is the 20 times repeated
+10-fold cross-validated AUC from training.
 
@@
 probability adjustment of those with adenoma as well as those with SRN
-and the probability adjustment from initial to follow up sample. THe
+and the probability adjustment from initial to follow up sample. The
 dotted line represents the threshold used to make the decision of

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
results update of text

@@
 ***Bacterial Community and Fit Changes before and after Treatment***
-Based on thetayc distance metrics comparing the initial to the follow up samples there was no difference between the adenoma and carcinoma group (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
+Based on thetayc distance metrics, comparing the initial to the follow up samples, there was no difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
 There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) [Figure 2].
+Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+This change can be visualized using an NMDS [Figure 2].
 When all follow up samples were compared to each other there was no significant overall difference between them (P-value = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, and evenness after correction for multiple comparisons [Table S1].
-Time of follow up sample from initial sampling did not have an effect on our prediction data set (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
+There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
+Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
 
 ***Outcome of Model Training***
-The range of the AUC for model training ranged from a minimum of `r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)` to a maximum of `r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)`.
-To be conservative the model chosen for prediction of disease in the follow up samples had an AUC in the middle of all the `r length(rownames(model_train_summary))` runs which was `r format(model_test_AUC_summary["middle", "ROC_cv"],
+The range of the AUC for model training ranged from a minimum of `r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)` to a maximum of `r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)` with the middle of all `r length(rownames(model_train_summary))` runs having an AUC of `r format(model_test_AUC_summary["middle", "ROC_cv"],
 digits = 3)`.
@@
 In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, middle training model test set AUC = `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, worse training model test set AUC = `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`).
-That is to say it had the smallest change in AUC in comparison to the other minimum and maximum AUC trained models.
+That is to say it had the smallest change in AUC in comparison to the minimum and maximum AUC trained models.
+The middle model was close to the full training data AUC `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
 There was no significant difference between the AUC of the best and middle training models (P-value = `r format(roc_pvalues["best", "middle"], digits = 3)`).
 There was also no difference in the middle model versus worse (P-value = `r format(roc_pvalues["middle", "worse"], digits = 3)`) or full data model (P-value = `r format(roc_pvalues["middle", "full"], digits = 3)`).
-The two comparisons with a significant difference were between the worse training model and the best training model (P-value = `r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)`) and the worse training model and full data model (P-value = `r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)`).
+The two comparisons with a significant difference were between the worse training model and the best training model (P-value = `r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)`) and the worse training model and full training model (P-value = `r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)`).
 
@@
 ***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
-A total of `r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))` sample was omitted from the original `r length(rownames(good_metaf))` sample set since it was missing a complete set of follow up data.
+A total of `r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))` sample was omitted from the original `r length(rownames(good_metaf))` test sample set since it was missing a complete set of follow up data.
 This left a total of `r sum(initial_cm$ref_yes)` samples for test predictions.
@@
 This decrease was significant for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)` [Figure 4a] and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4b] alone.
-This also held specifically for those with screen relative neoplasias (SRN) (P-value = `r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)`).
-A total of `r initial_cm["pred_yes", "ref_yes"]` or `r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)`% of all samples were correctly predicted to have an adenoma or carcinoma.
+This decrease in probability of lesion also held specifically for those with screen relative neoplasias (SRN) (P-value = `r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)`).
+A total of `r initial_cm["pred_yes", "ref_yes"]` or `r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)`% of all samples were correctly predicted to have a lesion.
 Although there was a decrease in positive probability only `r followup_cm["pred_no", "ref_no"]` of the total `r sum(initial_cm$ref_yes)` individuals were classified as adenoma or carcinoma free on follow up (successful classification of `r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)`%).
-There was no significant difference between the predictions and actual diagnosis for the initial samples in the 67-sample cohort test set.
+There was no significant difference between the predictions and actual diagnosis for the initial samples in the `r length(rownames(good_metaf))`-sample cohort test set.
 However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = `r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)`).

### 2016-12-07 submission/manuscript_outline_20161024.tex
results update of text

@@
 \textbf{\emph{Bacterial Community and Fit Changes before and after
-Treatment}} Based on thetayc distance metrics comparing the initial to
-the follow up samples there was no difference between the adenoma and
-carcinoma group (P-value = 0.697) {[}Figure 1a{]}. There was a
+Treatment}} Based on thetayc distance metrics, comparing the initial to
+the follow up samples, there was no difference between the adenoma and
+carcinoma groups (P-value = 0.697) {[}Figure 1a{]}. There was a
 difference in FIT between initial and follow up samples with the
@@
 initial and follow up (P-value = 0.002) but not for the adenoma group
-(P-value = 0.997) {[}Figure 2{]}. When all follow up samples were
-compared to each other there was no significant overall difference
-between them (P-value = 0.085). There was no significant difference
-between initial and follow up samples for observed OTUs, Shannon
-diversity, and evenness after correction for multiple comparisons
-{[}Table S1{]}. Time of follow up sample from initial sampling did not
-have an effect on our prediction data set (uncorrected P-value = 0.784).
+(P-value = 0.997). This change can be visualized using an NMDS {[}Figure
+2{]}. When all follow up samples were compared to each other there was
+no significant overall difference between them (P-value = 0.085). There
+was no significant difference between initial and follow up samples for
+observed OTUs, Shannon diversity, or evenness after correction for
+multiple comparisons {[}Table S1{]}. Time of follow up sample from
+initial sampling, did not have a significant difference between adenoma
+and carcinoma (uncorrected P-value = 0.784).
 
 \textbf{\emph{Outcome of Model Training}} The range of the AUC for model
-training ranged from a minimum of 0.795 to a maximum of 0.854. To be
-conservative the model chosen for prediction of disease in the follow up
-samples had an AUC in the middle of all the 100 runs which was 0.819.
-Interestingly, the worst AUC model from training performed the best on
-it's respective 80/20 split test data {[}Figure 3{]}. In fact the 80/20
-test performance showed that the AUC for the middle model chosen was the
-most stable (best training model test set AUC = 0.682, middle training
-model test set AUC = 0.815, worse training model test set AUC = 0.931).
-That is to say it had the smallest change in AUC in comparison to the
-other minimum and maximum AUC trained models. There was no significant
-difference between the AUC of the best and middle training models
-(P-value = 0.419). There was also no difference in the middle model
-versus worse (P-value = 0.178) or full data model (P-value = 1). The two
-comparisons with a significant difference were between the worse
-training model and the best training model (P-value = 7.94e-04) and the
-worse training model and full data model (P-value = 3.52e-03).
+training ranged from a minimum of 0.795 to a maximum of 0.854 with the
+middle of all 100 runs having an AUC of 0.819. Interestingly, the worst
+AUC model from training performed the best on it's respective 80/20
+split test data {[}Figure 3{]}. In fact the 80/20 test performance
+showed that the AUC for the middle model chosen was the most stable
+(best training model test set AUC = 0.682, middle training model test
+set AUC = 0.815, worse training model test set AUC = 0.931). That is to
+say it had the smallest change in AUC in comparison to the minimum and
+maximum AUC trained models. The middle model was close to the full
+training data AUC 0.813. There was no significant difference between the
+AUC of the best and middle training models (P-value = 0.419). There was
+also no difference in the middle model versus worse (P-value = 0.178) or
+full data model (P-value = 1). The two comparisons with a significant
+difference were between the worse training model and the best training
+model (P-value = 7.94e-04) and the worse training model and full
+training model (P-value = 3.52e-03).
 
@@
 Decrease in Positive Probability Prediction}} A total of 1 sample was
-omitted from the original 67 sample set since it was missing a complete
-set of follow up data. This left a total of 66 samples for test
+omitted from the original 67 test sample set since it was missing a
+complete set of follow up data. This left a total of 66 samples for test
 predictions. After multiple comparison correction there was a
@@
 significant for both adenoma (P-value = 7.95e-08 {[}Figure 4a{]} and
-carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This also held
-specifically for those with screen relative neoplasias (SRN) (P-value =
-5.35e-04). A total of 66 or 100\% of all samples were correctly
-predicted to have an adenoma or carcinoma. Although there was a decrease
-in positive probability only 24 of the total 66 individuals were
-classified as adenoma or carcinoma free on follow up (successful
-classification of 37.9\%). There was no significant difference between
-the predictions and actual diagnosis for the initial samples in the
-67-sample cohort test set. However, the predictions were significantly
-discordent with the diagnosis for the follow up samples (P-value =
-4.19e-10). Although there were discordent results the respective
-sensitivity for the initial group was 100\% and for follow up was 100\%,
-respectively.
+carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This decrease in
+probability of lesion also held specifically for those with screen
+relative neoplasias (SRN) (P-value = 5.35e-04). A total of 66 or 100\%
+of all samples were correctly predicted to have a lesion. Although there
+was a decrease in positive probability only 24 of the total 66
+individuals were classified as adenoma or carcinoma free on follow up
+(successful classification of 37.9\%). There was no significant
+difference between the predictions and actual diagnosis for the initial
+samples in the 67-sample cohort test set. However, the predictions were
+significantly discordent with the diagnosis for the follow up samples
+(P-value = 4.19e-10). Although there were discordent results the
+respective sensitivity for the initial group was 100\% and for follow up
+was 100\%, respectively.
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
update introduction text

@@
 It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance (**insert citation**).
-The vast majority of CRCs are not inherited and the exact etiology to disease is not well worked out (**insert citation**).
+The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out (**insert citation**).
 
-Recent reports have found that the microbiome between those with and without CRC are different and that bacteia associated with CRC tissue is different that that found on the normal mucosa (**Insert citations**).
+Recent reports have found that the microbiome between those with and without CRC are different and that bacteia associated with CRC tissue is different than that found on the normal mucosa (**Insert citations**).
 Further, there is evidence to suggest that more inflammatory mouth-associated bacteria are more prevalent in the guts of those with CRC (**insert citation**).
@@
 These collectively show that a reasonable area under the curve (AUC) can be obtained when using the stool bacterial microbiome to classify disease state (**insert citation**).
-Further, a number of groups have extended this observation to show that using either specific bacterial species or the bacterial microbiome as a whole paired with the Fecal Immunochemical Test (FIT) can either increase the AUC of the model (**insert citations**).
+Further, a number of groups have extended this observation to show that using either specific bacterial species or the bacterial microbiome as a whole paired with the Fecal Immunochemical Test (FIT) can increase the AUC of the model (**insert citations**).
 
-In this study we further refine the random forest model by including more background information on the indivdiuals within it.
-We use a rigourous and accepted approach for machine learning training and testing to validate this model (**insert citation**).
-We then apply it to individuals in which we have samples before and after treatment to assess how well the model adjusts their probability of having an adenoma or carcinoma.
-Finally, we investigate which specific Operational Taxonomic Units (OTUs) are most affected by treatment.
+In this study we further refine the random forest model by including more background information on the individuals within it.
+We used a rigourous and accepted approach for machine learning training and testing to validate this model (**insert citation**).
+We then applied it to individuals in which we have samples before and after treatment to assess how well the model adjusts their probability of having an adenoma or carcinoma.
+Finally, we investigated which specific Operational Taxonomic Units (OTUs) are most affected by treatment.
 Within this study lesion refers to both adenoma and carcinoma.

### 2016-12-07 submission/manuscript_outline_20161024.tex
update introduction text

@@
 dominant inheritance (\textbf{insert citation}). The vast majority of
-CRCs are not inherited and the exact etiology to disease is not well
-worked out (\textbf{insert citation}).
+CRCs are not inherited and the exact etiology to disease has not been
+well worked out (\textbf{insert citation}).
 
@@
 without CRC are different and that bacteia associated with CRC tissue is
-different that that found on the normal mucosa (\textbf{Insert
+different than that found on the normal mucosa (\textbf{Insert
 citations}). Further, there is evidence to suggest that more
@@
 species or the bacterial microbiome as a whole paired with the Fecal
-Immunochemical Test (FIT) can either increase the AUC of the model
+Immunochemical Test (FIT) can increase the AUC of the model
 (\textbf{insert citations}).
@@
 In this study we further refine the random forest model by including
-more background information on the indivdiuals within it. We use a
+more background information on the individuals within it. We used a
 rigourous and accepted approach for machine learning training and
-testing to validate this model (\textbf{insert citation}). We then apply
-it to individuals in which we have samples before and after treatment to
-assess how well the model adjusts their probability of having an adenoma
-or carcinoma. Finally, we investigate which specific Operational
-Taxonomic Units (OTUs) are most affected by treatment. Within this study
-lesion refers to both adenoma and carcinoma.
+testing to validate this model (\textbf{insert citation}). We then
+applied it to individuals in which we have samples before and after
+treatment to assess how well the model adjusts their probability of
+having an adenoma or carcinoma. Finally, we investigated which specific
+Operational Taxonomic Units (OTUs) are most affected by treatment.
+Within this study lesion refers to both adenoma and carcinoma.
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
update abstract text

@@
 Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.
-Previous research suggests that the bacterial microbiome can be used as a biomarker for colorectal cancer (CRC).
-These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.
-In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.
+Previous research suggests that the bacterial microbiome can be used as a biomarker for CRC.
+In this study, we used a microbiome-based model that classified individuals as having an adenoma or carcinoma to assess whether the probability of having one of these lesions changed after surgical treatment.
+We also characterized the change in the gut microbiota before and after the surgical treatment.
 This model was tested on a `r sum(initial_cm$ref_yes)` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.

### 2016-12-07 submission/manuscript_outline_20161024.tex
update abstract text

@@
 to the disease. Previous research suggests that the bacterial microbiome
-can be used as a biomarker for colorectal cancer (CRC). These reports
-have mostly focused on investigating how the bacterial microbiome is
-used at a single point in time to predict disease. In this study, we
-assessed whether a model built with bacterial microbiome data could
-accurately predict adenoma or carcinoma and adjust positive probability
-for lesion (adenoma or carcinoma) after surgical treatment. This model
-was tested on a 66 person group that included samples before and after
-treatment to allow for the assessment of how the model adjusts risk
-after treatment. The model used for prediction had an AUC of 0.813. For
-the follow up samples our Random Forest model significantly decreased
-the positive probability of lesion compared to the initial samples for
-both adenoma (P-value = 7.95e-08) and carcinoma (P-value = 7.95e-08).
-Our model predicted that 36.4\% of the 67-person cohort had normal
-colons and 63.6\% had a lesion. Some OTUs that changed the most before
-and after treatment included OTUs that were affiliated with members of
-Blautia, Clostridium\_XlVa and Escherichia/Shigella. Our model suggests
-that treatment does significantly reduce the probability of having a
-colonic lesion. Further surveillance of these individuals will enable us
-to determine whether models such as the one we present here can also be
-used to predict recurrence of colorectal cancer.
+can be used as a biomarker for CRC. In this study, we used a
+microbiome-based model that classified individuals as having an adenoma
+or carcinoma to assess whether the probability of having one of these
+lesions changed after surgical treatment. We also characterized the
+change in the gut microbiota before and after the surgical treatment.
+This model was tested on a 66 person group that included samples before
+and after treatment to allow for the assessment of how the model adjusts
+risk after treatment. The model used for prediction had an AUC of 0.813.
+For the follow up samples our Random Forest model significantly
+decreased the positive probability of lesion compared to the initial
+samples for both adenoma (P-value = 7.95e-08) and carcinoma (P-value =
+7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
+normal colons and 63.6\% had a lesion. Some OTUs that changed the most
+before and after treatment included OTUs that were affiliated with
+members of Blautia, Clostridium\_XlVa and Escherichia/Shigella. Our
+model suggests that treatment does significantly reduce the probability
+of having a colonic lesion. Further surveillance of these individuals
+will enable us to determine whether models such as the one we present
+here can also be used to predict recurrence of colorectal cancer.
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
update code  for pvalues and methods section

@@
 
+***Investigation of Important CRC Bacteria***
+In order to investigate how specific bacteria that have been previously associated with CRC with selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.
+Specifically we wanted to observe if there were any differences based on whether the individual had an adenoma or carcinoma that may have been missed with our lesion versus normal model.
+
 ***Statistical Analysis***

### 2016-12-07 code/Run_potential_cancer_specific_OTUs.R

@@ -84,6 +84,42 @@ crc_select_data <- as.data.frame(cbind(value = c(shared[, good_counts_init[1]],
 
 write.csv(crc_select_data, "results/tables/adn_crc_maybe_diff.csv", row.names = F)
 
+# Run statistics testing
+
+pvalue_summary <- matrix(nrow = 4, ncol = 4, dimnames = list(
+  nrow = c("fn", "parv", "pept", "porp"), ncol = c("crc_pvalue", "crc_BH", "adn_pvalue", "adn_BH")))
+
+for(i in 1:length(good_counts_init)){
+  
+  pvalue_summary[i, "crc_pvalue"] <- wilcox.test(
+    filter(crc_select_data, otu == good_counts_init[i], Dx_Bin == "cancer", 
+           sampleType == "initial")[, "value"], 
+    filter(crc_select_data, otu == good_counts_init[i], Dx_Bin == "cancer", 
+           sampleType == "followup")[, "value"], paired = TRUE)$p.value
+  
+  pvalue_summary[i, "crc_BH"] <- p.adjust(pvalue_summary[i, "crc_pvalue"], method = "BH", n = 4)
+  
+}
+
+
+for(i in 1:length(good_counts_init)){
+  
+  pvalue_summary[i, "adn_pvalue"] <- wilcox.test(
+    filter(crc_select_data, otu == good_counts_init[i], Dx_Bin != "cancer", 
+           sampleType == "initial")[, "value"], 
+    filter(crc_select_data, otu == good_counts_init[i], Dx_Bin != "cancer", 
+           sampleType == "followup")[, "value"], paired = TRUE)$p.value
+  
+  pvalue_summary[i, "adn_BH"] <- p.adjust(pvalue_summary[i, "adn_pvalue"], method = "BH", n = 4)
+  
+}
+
+# Write out the pvalue table for future use
+
+write.csv(pvalue_summary, "results/tables/adn_crc_maybe_pvalue_summary.csv", row.names = F)
+
+
+
 
 
 

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
incorporate Fig 6 into manuscript and tab S5

@@
 pvalue_common_imp_vars <- read.csv('../results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
+
+# Results of CRC Bacteria
+pvalue_adn_crc_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
 
@@
 
+***Differences in Adenoma and Carcinoma in Previously Associated Cancer Bacteria***
+First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].
+The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus assacharolytica (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas stomatis (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S5].
+
 \newpage
@@
 Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].
+Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.
+We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].
 This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.
@@
 This has been suggested to be the case in a previous study (**insert Baxter Study**).
+It is also possible that these specific bacteria play a major role in the progression to carcinoma but may not be as important in the initiation of an adenoma, which would be supported by our data [Figure 5].
 Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.
@@
 
+**Figure 6: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
+There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus assacharolytica (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas stomatis (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).
+B) Adenoma initial and follow up samples.
+There were no significant differences between initial and follow up.
+
 \newpage

### 2016-12-07 code/Run_potential_cancer_specific_OTUs.R

@@ -116,7 +116,7 @@ for(i in 1:length(good_counts_init)){
 
 # Write out the pvalue table for future use
 
-write.csv(pvalue_summary, "results/tables/adn_crc_maybe_pvalue_summary.csv", row.names = F)
+write.csv(pvalue_summary, "results/tables/adn_crc_maybe_pvalue_summary.csv")
 
 
 

### 2016-12-07 submission/manuscript_outline_20161024.tex

@@
 
+\textbf{\emph{Differences in Adenoma and Carcinoma in Previously
+Associated Cancer Bacteria}} First, there was a clear magnitude
+difference in these specific OTUs based on whether they were from
+adenoma or carcinoma individuals {[}Figure 6{]}. The carcinoma samples
+showed a significant difference between initial and follow up samples
+for Peptostreptococcus assacharolytica (P-value = 0.0183) and
+Porphyromonas stomatis (P-value = 0.0154) whereas there were no
+signficant differences in any of these OTUs in the adenoma samples
+{[}Table S5{]}.
+
 \newpage
@@
 abundance values observed in the control individuals of the training set
-{[}Figure 5{]}. This research provides evidence that it is possible to
-use bacterial microbiome data to create a highily sensitive model, that
-is reactive to therapy, for detection of adenoma or carcinoma. It
-accomplishes this by using a unique sample set in which before and after
-surgery stool samples are available for assessment. By using these types
-of samples we are not only able to show sensitivity of lesion prediction
-but also able to show that this model is reactive. That is to say that
-after surgery for removal of the adenoma or carcinoma it decreases the
-positive probability to reflect a lower likelihood of the individual
-having an adenoma or carcinoma.
+{[}Figure 5{]}. Although we were interested in what we could use to
+classify those with either adenoma or carcinoma versus normal. We found
+that the traditional bacteria associated with CRC were higher in
+magnitude in the carcinoma group and there were significant differences
+in some of these OTUs between the initial and follow up samples
+{[}Figure 5 and Tabl S5{]}. This research provides evidence that it is
+possible to use bacterial microbiome data to create a highily sensitive
+model, that is reactive to therapy, for detection of adenoma or
+carcinoma. It accomplishes this by using a unique sample set in which
+before and after surgery stool samples are available for assessment. By
+using these types of samples we are not only able to show sensitivity of
+lesion prediction but also able to show that this model is reactive.
+That is to say that after surgery for removal of the adenoma or
+carcinoma it decreases the positive probability to reflect a lower
+likelihood of the individual having an adenoma or carcinoma.
 
@@
 over them. This has been suggested to be the case in a previous study
-(\textbf{insert Baxter Study}). Regardless, our study does not argue
-against the importance of these bacterium in CRC initiation or
+(\textbf{insert Baxter Study}). It is also possible that these specific
+bacteria play a major role in the progression to carcinoma but may not
+be as important in the initiation of an adenoma, which would be
+supported by our data {[}Figure 5{]}. Regardless, our study does not
+argue against the importance of these bacterium in CRC initiation or
 pathogenesis but rather that the model does not utilize these specific
@@
 
+\textbf{\emph{Investigation of Important CRC Bacteria}} In order to
+investigate how specific bacteria that have been previously associated
+with CRC with selected OTUs that taxonomically classified to
+Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
+Assacharolytica, and Porphyromonas Stomatis. Specifically we wanted to
+observe if there were any differences based on whether the individual
+had an adenoma or carcinoma that may have been missed with our lesion
+versus normal model.
+
 \textbf{\emph{Statistical Analysis}} The R software package (v3.3.0) was
@@
 
+\textbf{Figure 6: Previously Associated CRC Bacteria in Initial and
+Follow up Samples.} A) Carcinoma initial and follow up samples. There
+was a significant difference in initial and follow up sample for the
+OTUs classfied as Peptostreptococcus assacharolytica (P-value = 0.0183)
+and Porphyromonas stomatis (P-value = 0.0154). B) Adenoma initial and
+follow up samples. There were no significant differences between initial
+and follow up.
+
 \newpage

### 2016-12-07 submission/supplemental_outline_20161024.Rmd

@@
 
+\newpage
+
+**Table S5: Summary of P-values for Specific CRC OTUs between Initial and Follow Ups**
+
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+
+pvalue_adn_crc_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
+
+st5 <- cbind(tax_id = c("Fusobacterium Nucleatum", "Parvimonas Micra", 
+                        "Peptostreptococcus Assacharolytica", "Porphyromonas Stomatis"), 
+             crc_pvalue = format(pvalue_adn_crc_summary$crc_pvalue, digits = 3), 
+             crc_bh = format(pvalue_adn_crc_summary$crc_BH, digits = 3), 
+             adn_pvalue = format(pvalue_adn_crc_summary$adn_pvalue, digits = 3), 
+             adn_bh = format(pvalue_adn_crc_summary$adn_BH, digits = 3))
+
+colnames(st5) <- c("Taxonomic ID", "CRC P-value", "CRC BH Corrected P-value", 
+                   "Adenoma P-value", "Adenoma BH Corrected P-value")
+
+kable(st5, format = "markdown", align = 'c', padding = 2)
+
+```
+
 ```{r supplemental_figure1, echo=FALSE, message=FALSE, warning=FALSE}

### 2016-12-08 code/Run_Figure6.R
fix spelling mistake and wrong species

@@ -15,10 +15,10 @@ graph_data <- read.csv("results/tables/adn_crc_maybe_diff.csv", header = T, stri
 graph_data <- mutate(graph_data, j_values = jitter(value, 3))
 
 labs <- c(
-  paste("Fusobacterium Nucleatum ", "(", unique(graph_data$otu)[1], ")", sep = ""), 
-  paste("Parvimonas Micra ", "(", unique(graph_data$otu)[2], ")", sep = ""), 
-  paste("Peptostreptococcus Assacharolytica ", "(", unique(graph_data$otu)[3], ")", sep = ""), 
-  paste("Porphyromonas Stomatis ", "(", unique(graph_data$otu)[4], ")", sep = ""))
+  paste("Fusobacterium nucleatum ", "(", unique(graph_data$otu)[1], ")", sep = ""), 
+  paste("Parvimonas micra ", "(", unique(graph_data$otu)[2], ")", sep = ""), 
+  paste("Peptostreptococcus stomatis ", "(", unique(graph_data$otu)[3], ")", sep = ""), 
+  paste("Porphyromonas asaccharolytica ", "(", unique(graph_data$otu)[4], ")", sep = ""))
 
 names(labs) <- unique(graph_data$otu)
 

### 2016-12-08 submission/manuscript_outline_20161024.Rmd

@@
 First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].
-The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus assacharolytica (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas stomatis (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S5].
+The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S5].
 
@@
 **Figure 6: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
-There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus assacharolytica (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas stomatis (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).
+There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).
 B) Adenoma initial and follow up samples.

### 2016-12-08 submission/manuscript_outline_20161024.tex

@@
 showed a significant difference between initial and follow up samples
-for Peptostreptococcus assacharolytica (P-value = 0.0183) and
-Porphyromonas stomatis (P-value = 0.0154) whereas there were no
-signficant differences in any of these OTUs in the adenoma samples
-{[}Table S5{]}.
+for Peptostreptococcus stomatis (P-value = 0.0183) and Porphyromonas
+asaccharolytica (P-value = 0.0154) whereas there were no signficant
+differences in any of these OTUs in the adenoma samples {[}Table S5{]}.
 
@@
 was a significant difference in initial and follow up sample for the
-OTUs classfied as Peptostreptococcus assacharolytica (P-value = 0.0183)
-and Porphyromonas stomatis (P-value = 0.0154). B) Adenoma initial and
+OTUs classfied as Peptostreptococcus stomatis (P-value = 0.0183) and
+Porphyromonas asaccharolytica (P-value = 0.0154). B) Adenoma initial and
 follow up samples. There were no significant differences between initial

### 2016-12-09 submission/manuscript_outline_20161024.tex
update abstract and Importance blurb

@@
 early detection being used as a key component in mitigating deaths due
-to the disease. Previous research suggests that the bacterial microbiome
-can be used as a biomarker for CRC. In this study, we used a
-microbiome-based model that classified individuals as having an adenoma
-or carcinoma to assess whether the probability of having one of these
-lesions changed after surgical treatment. We also characterized the
-change in the gut microbiota before and after the surgical treatment.
+to the disease. Previous research suggests a link between stool
+bacterial microbiome and CRC. In this study, we used a model based on
+the microbiome, demographics, and prior medical history to classify
+individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
+this model to characterize the change in the gut microbiota before and
+after surgery. The overall objective was to investigate the changes in
+the microbiome after surgery in patients with adenomas or carcinomas.
 This model was tested on a 66 person group that included samples before
@@
 This is one of the first studies to investigate within humans what
-happens to the bacterial microbiome before and after colorectal cancer
-treatment. Specifically, it aims to assess how the random forest machine
-learning algorithm built models respond to treatment and adjust their
-positive probability calls of whether the individual has an adenoma or
-carcinoma.
+happens to the bacterial microbiome before and after adenoma and
+carcinoma treatment. Specifically, it aims to assess how a random forest
+machine learning algorithm built model respond to treatment and adjust
+it's positive probability calls of whether the individual has an adenoma
+or carcinoma due to surgical removal of the lesion.
 

### 2016-12-09 submission/manuscript_outline_20161024.Rmd

@@
 Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.
-Previous research suggests that the bacterial microbiome can be used as a biomarker for CRC.
-In this study, we used a microbiome-based model that classified individuals as having an adenoma or carcinoma to assess whether the probability of having one of these lesions changed after surgical treatment.
-We also characterized the change in the gut microbiota before and after the surgical treatment.
+Previous research suggests a link between stool bacterial microbiome and CRC.
+In this study, we used a model based on the microbiome, demographics, and prior medical history to classify individuals as having a lesion (i.e.
+adenoma or carcinoma).
+We then used this model to characterize the change in the gut microbiota before and after surgery.
+The overall objective was to investigate the changes in the microbiome after surgery in patients with adenomas or carcinomas.
 This model was tested on a `r sum(initial_cm$ref_yes)` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
@@
 
-This is one of the first studies to investigate within humans what happens to the bacterial microbiome before and after colorectal cancer treatment.
-Specifically, it aims to assess how the random forest machine learning algorithm built models respond to treatment and adjust their positive probability calls of whether the individual has an adenoma or carcinoma.
+This is one of the first studies to investigate within humans what happens to the bacterial microbiome before and after adenoma and carcinoma treatment.
+Specifically, it aims to assess how a random forest machine learning algorithm built model respond to treatment and adjust it's positive probability calls of whether the individual has an adenoma or carcinoma due to surgical removal of the lesion.
 

### 2016-12-15 submission/manuscript_outline_20161024.tex
update manuscript with new fig 4 and p-values

@@
 and after treatment to allow for the assessment of how the model adjusts
-risk after treatment. The model used for prediction had an AUC of 0.813.
+risk after treatment. The model used for prediction had an AUC of 0.763.
 For the follow up samples our Random Forest model significantly
 decreased the positive probability of lesion compared to the initial
-samples for both adenoma (P-value = 7.95e-08) and carcinoma (P-value =
+samples for both adenoma (P-value = 3.64e-11) and carcinoma (P-value =
 7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
@@
 \textbf{\emph{Outcome of Model Training}} The range of the AUC for model
-training ranged from a minimum of 0.795 to a maximum of 0.854 with the
-middle of all 100 runs having an AUC of 0.819. Interestingly, the worst
+training ranged from a minimum of 0.723 to a maximum of 0.795 with the
+middle of all 100 runs having an AUC of 0.761. Interestingly, the worst
 AUC model from training performed the best on it's respective 80/20
@@
 showed that the AUC for the middle model chosen was the most stable
-(best training model test set AUC = 0.682, middle training model test
-set AUC = 0.815, worse training model test set AUC = 0.931). That is to
+(best training model test set AUC = 0.646, middle training model test
+set AUC = 0.744, worse training model test set AUC = 0.904). That is to
 say it had the smallest change in AUC in comparison to the minimum and
 maximum AUC trained models. The middle model was close to the full
-training data AUC 0.813. There was no significant difference between the
-AUC of the best and middle training models (P-value = 0.419). There was
-also no difference in the middle model versus worse (P-value = 0.178) or
+training data AUC 0.763. There was no significant difference between the
+AUC of the best and middle training models (P-value = 1). There was also
+no difference in the middle model versus worse (P-value = 0.0431) or
 full data model (P-value = 1). The two comparisons with a significant
 difference were between the worse training model and the best training
-model (P-value = 7.94e-04) and the worse training model and full
-training model (P-value = 3.52e-03).
+model (P-value = 6.83e-04) and the worse training model and full
+training model (P-value = 1.2e-03).
 
 \textbf{\emph{Most Important Variables to the Model}} Overall, there
-were a total of 32 variables identified as being present in more than
+were a total of 37 variables identified as being present in more than
 50\% of the training models {[}Table S2{]}. The top 5 most important
@@
 (Otu000018), Bacteria (Otu000020), Ruminococcus (Otu000017), and
-Porphyromonas (Otu000153). These 5 OTUs were present in at least 99 out
+Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
 of the total 100 different 80/20 runs.
@@
 significant overall decrease in positive probability of a carcinoma and
-adenoma (P-value = 1.14e-11) {[}Figure 4{]}. This decrease was
-significant for both adenoma (P-value = 7.95e-08 {[}Figure 4a{]} and
+adenoma (P-value = 1.11e-11) {[}Figure 4{]}. This decrease was
+significant for both adenoma (P-value = 3.64e-11 {[}Figure 4a{]} and
 carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This decrease in
 probability of lesion also held specifically for those with screen
-relative neoplasias (SRN) (P-value = 5.35e-04). A total of 66 or 100\%
+relative neoplasias (SRN) (P-value = 7.63e-06). A total of 66 or 100\%
 of all samples were correctly predicted to have a lesion. Although there
@@
 was still higher than the cutoff threshold of 0.5 (positive probability
-= 0.927). Interestingly, 2 individuals who were unknown on follow up
+= 0.903). Interestingly, 1 individuals who were unknown on follow up
 still were over the threshold cutoff of 0.5 even though, like the 1
@@
 The follow up positive probabilities were not affected by either
-chemotherapy treatment (uncorrected P-value = 0.919) or radiation
-therapy (uncorrected P-value = 1). There was also no difference in the
-amount of change in the positive probability based on whether
-individuals received chemotherapy (unccorected P-value = 0.578) or
-radiation therapy (uncorrected P-value = 0.904).
+chemotherapy treatment (uncorrected P-value = 0.621) or radiation
+therapy (uncorrected P-value = 0.255). There was also no difference in
+the amount of change in the positive probability based on whether
+individuals received chemotherapy (unccorected P-value = 0.718) or
+radiation therapy (uncorrected P-value = 0.431).
 
@@
 the test sets for the best, middle, and worse models from training were
-0.682, 0.815, and 0.931, respectively. cvAUC is the 20 times repeated
+0.646, 0.744, and 0.904, respectively. cvAUC is the 20 times repeated
 10-fold cross-validated AUC from training.

### 2016-12-15 submission/manuscript_outline_20161024.Rmd

@@
 geometry: margin=1.0in
-bibliography: null
+bibliography: reference.bib
 ---

### 2016-12-16 submission/manuscript_outline_20161024.Rmd
update manuscript refs 1st paragraph

@@
 
-Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age (**insert citation**).
-Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening (**insert citation**).
-However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year (**insert citation**).
-It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance (**insert citation**).
-The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out (**insert citation**).
+Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].
+Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].
+However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].
+It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007].
+The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].
 

### 2016-12-16 submission/manuscript_outline_20161024.tex
update manuscript refs 1st paragraph

@@
 related deaths and is the second most common cancer death among men aged
-40-79 years of age (\textbf{insert citation}). Over the last few years
-death due to the disease has seen a significant decrease thanks mainly
-to improvements in screening (\textbf{insert citation}). However,
-despite this giant improvement there are still approximately 50,000
-deaths from the disease a year (\textbf{insert citation}). It is
-estimated that around 5-10\% of all CRCs can be explained by autosomal
-dominant inheritance (\textbf{insert citation}). The vast majority of
-CRCs are not inherited and the exact etiology to disease has not been
-well worked out (\textbf{insert citation}).
+40-79 years of age (1, 2). Over the last few years death due to the
+disease has seen a significant decrease thanks mainly to improvements in
+screening (1). However, despite this giant improvement there are still
+approximately 50,000 deaths from the disease a year (2). It is estimated
+that around 5-10\% of all CRCs can be explained by autosomal dominant
+inheritance (3). The vast majority of CRCs are not inherited and the
+exact etiology to disease has not been well worked out (2).
 
@@
 
-\subsection{References}\label{references}
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
+\hypertarget{refs}{}
+\hypertarget{ref-jemal_cancer_2010}{}
+1. \textbf{Jemal A}, \textbf{Siegel R}, \textbf{Xu J}, \textbf{Ward E}.
+2010. Cancer statistics, 2010. CA: a cancer journal for clinicians
+\textbf{60}:277--300.
+doi:\href{https://doi.org/10.3322/caac.20073}{10.3322/caac.20073}.
+
+\hypertarget{ref-haggar_colorectal_2009}{}
+2. \textbf{Haggar FA}, \textbf{Boushey RP}. 2009. Colorectal cancer
+epidemiology: Incidence, mortality, survival, and risk factors. Clinics
+in Colon and Rectal Surgery \textbf{22}:191--197.
+doi:\href{https://doi.org/10.1055/s-0029-1242458}{10.1055/s-0029-1242458}.
+
+\hypertarget{ref-green_very_2007}{}
+3. \textbf{Green RC}, \textbf{Green JS}, \textbf{Buehler SK},
+\textbf{Robb JD}, \textbf{Daftary D}, \textbf{Gallinger S},
+\textbf{McLaughlin JR}, \textbf{Parfrey PS}, \textbf{Younghusband HB}.
+2007. Very high incidence of familial colorectal cancer in Newfoundland:
+A comparison with Ontario and 13 other population-based studies.
+Familial Cancer \textbf{6}:53--62.
+doi:\href{https://doi.org/10.1007/s10689-006-9104-x}{10.1007/s10689-006-9104-x}.
 

### 2016-12-16 submission/manuscript_outline_20161024.tex
update introduction Pat's suggestions

@@
 inheritance (3). The vast majority of CRCs are not inherited and the
-exact etiology to disease has not been well worked out (2).
+exact etiology to disease has not been well worked out (2). Although
+many risk factors have been identified ({\textbf{???}}) and non-invasive
+screening techniques have started to be put into consistent use (4, 5)
+there has been a consistent increase in the incidence of CRC in the
+younger population.
 
-Recent reports have found that the microbiome between those with and
-without CRC are different and that bacteia associated with CRC tissue is
-different than that found on the normal mucosa (\textbf{Insert
-citations}). Further, there is evidence to suggest that more
-inflammatory mouth-associated bacteria are more prevalent in the guts of
-those with CRC (\textbf{insert citation}). This has led to the
-hypothesis that these bacteria replace traditional residents that
-produce metabolites such as butyrate (\textbf{insert citation}) and lead
-to a more inflammatory state (\textbf{insert citation}). This
-inflammatory state allows for more of these types of bacteria to
-populate the gut and leads to neoplasia and eventually CRC
-(\textbf{insert citation}).
+This increased incidence of CRC in the younger population is concerncing
+since having either an adenoma or carcinoma increases ones risk for
+future adenomas or carcinomas (6--8). This increased risk can also carry
+with it an increased risk of mortality due to this recurrence (9, 10).
+Therefore there has been a great amount of interest in early risk
+stratification tools (11, 12) that can help identify those they may be
+at risk of reccurence. Concurrently with this there has been a lot of
+interest in new areas such as the gut bacterial microbiome for insight
+into potential disease pathology.
 
-Building upon these findings a number of groups, including our own, have
-investigated using the bacterial microbiome as a potential biomarker for
-adenoma and CRC detection (\textbf{insert citations}). These
-collectively show that a reasonable area under the curve (AUC) can be
-obtained when using the stool bacterial microbiome to classify disease
-state (\textbf{insert citation}). Further, a number of groups have
-extended this observation to show that using either specific bacterial
-species or the bacterial microbiome as a whole paired with the Fecal
-Immunochemical Test (FIT) can increase the AUC of the model
-(\textbf{insert citations}).
+There has been promising work on the bacterial microbiome and it's
+ability to be able to complement existing screening methods such as
+Fecal Immunoglobulin Test (FIT) or act alone as a screening tool (13,
+14). There has also been research into how this microbiome could be
+altered directly on tumor tissue itself (15). There have also been a few
+studies that have shown how this microbiome (16) or specific members
+within it (17) could be directly involved ith the pathogenesis of CRC.
+These studies have helped to provide a tantilzing link between the
+bacterial microbiome and CRC. However, at this present time there
+remains limited information on the bacterial microbiome before and after
+successful surgery for removal of the adenoma or carcinoma.
 
-In this study we further refine the random forest model by including
-more background information on the individuals within it. We used a
-rigourous and accepted approach for machine learning training and
-testing to validate this model (\textbf{insert citation}). We then
-applied it to individuals in which we have samples before and after
-treatment to assess how well the model adjusts their probability of
-having an adenoma or carcinoma. Finally, we investigated which specific
-Operational Taxonomic Units (OTUs) are most affected by treatment.
-Within this study lesion refers to both adenoma and carcinoma.
+In this study we investigated what happened to the bacterial microbiome
+before and after surgery for both adenoma and carcinoma individuals. Our
+anlaysis includes both alpha and beta diversity analysis along with
+investigating indvidual operational taxonomic units (OTUs). We next
+utilized a Random Forest model that was trained on a completely separate
+data set and observed how this model as well as specific OTUs within
+this model performed on the initial and follow up samples. We then
+created a new model to classify intial and follow up samples and
+investigated how the model performs and what specific OUTs within the
+model change. Finally, we observed how specific bacteria that have been
+previously implicated in CRC changed before and after treatment in both
+the adenoma and carcinoma individuals.
 
@@
 
+\hypertarget{ref-liao_application_2013}{}
+4. \textbf{Liao C-S}, \textbf{Lin Y-M}, \textbf{Chang H-C}, \textbf{Chen
+Y-H}, \textbf{Chong L-W}, \textbf{Chen C-H}, \textbf{Lin Y-S},
+\textbf{Yang K-C}, \textbf{Shih C-H}. 2013. Application of quantitative
+estimates of fecal hemoglobin concentration for risk prediction of
+colorectal neoplasia. World Journal of Gastroenterology
+\textbf{19}:8366--8372.
+doi:\href{https://doi.org/10.3748/wjg.v19.i45.8366}{10.3748/wjg.v19.i45.8366}.
+
+\hypertarget{ref-johnson_multi-target_2016}{}
+5. \textbf{Johnson DH}, \textbf{Kisiel JB}, \textbf{Burger KN},
+\textbf{Mahoney DW}, \textbf{Devens ME}, \textbf{Ahlquist DA},
+\textbf{Sweetser S}. 2016. Multi-target stool DNA test: Clinical
+performance and impact on yield and quality of colonoscopy for
+colorectal cancer screening. Gastrointestinal Endoscopy.
+doi:\href{https://doi.org/10.1016/j.gie.2016.11.012}{10.1016/j.gie.2016.11.012}.
+
+\hypertarget{ref-laiyemo_short-_2013}{}
+6. \textbf{Laiyemo AO}, \textbf{Doubeni C}, \textbf{Brim H},
+\textbf{Ashktorab H}, \textbf{Schoen RE}, \textbf{Gupta S},
+\textbf{Charabaty A}, \textbf{Lanza E}, \textbf{Smoot DT}, \textbf{Platz
+E}, \textbf{Cross AJ}. 2013. Short- and long-term risk of colorectal
+adenoma recurrence among whites and blacks. Gastrointestinal Endoscopy
+\textbf{77}:447--454.
+doi:\href{https://doi.org/10.1016/j.gie.2012.11.027}{10.1016/j.gie.2012.11.027}.
+
+\hypertarget{ref-matsuda_five-year_2009}{}
+7. \textbf{Matsuda T}, \textbf{Fujii T}, \textbf{Sano Y}, \textbf{Kudo
+S-e}, \textbf{Oda Y}, \textbf{Igarashi M}, \textbf{Iishi H},
+\textbf{Murakami Y}, \textbf{Ishikawa H}, \textbf{Shimoda T},
+\textbf{Kaneko K}, \textbf{Yoshida S}. 2009. Five-year incidence of
+advanced neoplasia after initial colonoscopy in Japan: A multicenter
+retrospective cohort study. Japanese Journal of Clinical Oncology
+\textbf{39}:435--442.
+doi:\href{https://doi.org/10.1093/jjco/hyp047}{10.1093/jjco/hyp047}.
+
+\hypertarget{ref-ren_long-term_2016}{}
+8. \textbf{Ren J}, \textbf{Kirkness CS}, \textbf{Kim M}, \textbf{Asche
+CV}, \textbf{Puli S}. 2016. Long-term risk of colorectal cancer by
+gender after positive colonoscopy: Population-based cohort study.
+Current Medical Research and Opinion \textbf{32}:1367--1374.
+doi:\href{https://doi.org/10.1080/03007995.2016.1174840}{10.1080/03007995.2016.1174840}.
+
+\hypertarget{ref-loberg_long-term_2014}{}
+9. \textbf{Løberg M}, \textbf{Kalager M}, \textbf{Holme Ø}, \textbf{Hoff
+G}, \textbf{Adami H-O}, \textbf{Bretthauer M}. 2014. Long-term
+colorectal-cancer mortality after adenoma removal. The New England
+Journal of Medicine \textbf{371}:799--807.
+doi:\href{https://doi.org/10.1056/NEJMoa1315870}{10.1056/NEJMoa1315870}.
+
+\hypertarget{ref-freeman_natural_2013}{}
+10. \textbf{Freeman HJ}. 2013. Natural history and long-term outcomes of
+patients treated for early stage colorectal cancer. Canadian Journal of
+Gastroenterology = Journal Canadien De Gastroenterologie
+\textbf{27}:409--413.
+
+\hypertarget{ref-lee_identification_2016}{}
+11. \textbf{Lee JH}, \textbf{Lee JL}, \textbf{Park IJ}, \textbf{Lim
+S-B}, \textbf{Yu CS}, \textbf{Kim JC}. 2016. Identification of
+Recurrence-Predictive Indicators in Stage I Colorectal Cancer. World
+Journal of Surgery.
+doi:\href{https://doi.org/10.1007/s00268-016-3833-2}{10.1007/s00268-016-3833-2}.
+
+\hypertarget{ref-richards_evidence-based_2016}{}
+12. \textbf{Richards CH}, \textbf{Ventham NT}, \textbf{Mansouri D},
+\textbf{Wilson M}, \textbf{Ramsay G}, \textbf{Mackay CD},
+\textbf{Parnaby CN}, \textbf{Smith D}, \textbf{On J}, \textbf{Speake D},
+\textbf{McFarlane G}, \textbf{Neo YN}, \textbf{Aitken E},
+\textbf{Forrest C}, \textbf{Knight K}, \textbf{McKay A}, \textbf{Nair
+H}, \textbf{Mulholland C}, \textbf{Robertson JH}, \textbf{Carey FA},
+\textbf{Steele R}, \textbf{Scottish Surgical Research Group}. 2016. An
+evidence-based treatment algorithm for colorectal polyp cancers: Results
+from the Scottish Screen-detected Polyp Cancer Study (SSPoCS). Gut.
+doi:\href{https://doi.org/10.1136/gutjnl-2016-312201}{10.1136/gutjnl-2016-312201}.
+
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-zeller_potential_2014}{}
+14. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
+\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
+\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
+\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
+\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
+J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
+\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
+\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
+for early-stage detection of colorectal cancer. Molecular Systems
+Biology \textbf{10}:766.
+
+\hypertarget{ref-dejea_microbiota_2014}{}
+15. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
+\textbf{White JR}, \textbf{Mark Welch JL}, \textbf{Rossetti BJ},
+\textbf{Peterson SN}, \textbf{Snesrud EC}, \textbf{Borisy GG},
+\textbf{Lazarev M}, \textbf{Stein E}, \textbf{Vadivelu J},
+\textbf{Roslani AC}, \textbf{Malik AA}, \textbf{Wanyiri JW}, \textbf{Goh
+KL}, \textbf{Thevambiga I}, \textbf{Fu K}, \textbf{Wan F}, \textbf{Llosa
+N}, \textbf{Housseau F}, \textbf{Romans K}, \textbf{Wu X},
+\textbf{McAllister FM}, \textbf{Wu S}, \textbf{Vogelstein B},
+\textbf{Kinzler KW}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014.
+Microbiota organization is a distinct feature of proximal colorectal
+cancers. Proceedings of the National Academy of Sciences of the United
+States of America \textbf{111}:18321--18326.
+doi:\href{https://doi.org/10.1073/pnas.1406199111}{10.1073/pnas.1406199111}.
+
+\hypertarget{ref-zackular_manipulation_2016}{}
+16. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2016. Manipulation of the Gut Microbiota Reveals
+Role in Colon Tumorigenesis. mSphere \textbf{1}.
+doi:\href{https://doi.org/10.1128/mSphere.00001-15}{10.1128/mSphere.00001-15}.
+
+\hypertarget{ref-arthur_microbial_2014}{}
+17. \textbf{Arthur JC}, \textbf{Gharaibeh RZ}, \textbf{Mühlbauer M},
+\textbf{Perez-Chanona E}, \textbf{Uronis JM}, \textbf{McCafferty J},
+\textbf{Fodor AA}, \textbf{Jobin C}. 2014. Microbial genomic analysis
+reveals the essential role of inflammation in bacteria-induced
+colorectal cancer. Nature Communications \textbf{5}:4724.
+doi:\href{https://doi.org/10.1038/ncomms5724}{10.1038/ncomms5724}.
+
 \end{document}

### 2016-12-16 submission/manuscript_outline_20161024.Rmd

@@
 The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].
+Although many risk factors have been identified [@hagger_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.
 
-Recent reports have found that the microbiome between those with and without CRC are different and that bacteia associated with CRC tissue is different than that found on the normal mucosa (**Insert citations**).
-Further, there is evidence to suggest that more inflammatory mouth-associated bacteria are more prevalent in the guts of those with CRC (**insert citation**).
-This has led to the hypothesis that these bacteria replace traditional residents that produce metabolites such as butyrate (**insert citation**) and lead to a more inflammatory state (**insert citation**).
-This inflammatory state allows for more of these types of bacteria to populate the gut and leads to neoplasia and eventually CRC (**insert citation**).
+This increased incidence of CRC in the younger population is concerncing since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].
+This increased risk can also carry with it an increased risk of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].
+Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at risk of reccurence.
+Concurrently with this there has been a lot of interest in new areas such as the gut bacterial microbiome for insight into potential disease pathology.
 
-Building upon these findings a number of groups, including our own, have investigated using the bacterial microbiome as a potential biomarker for adenoma and CRC detection (**insert citations**).
-These collectively show that a reasonable area under the curve (AUC) can be obtained when using the stool bacterial microbiome to classify disease state (**insert citation**).
-Further, a number of groups have extended this observation to show that using either specific bacterial species or the bacterial microbiome as a whole paired with the Fecal Immunochemical Test (FIT) can increase the AUC of the model (**insert citations**).
+There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
+There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
+There have also been a few studies that have shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved ith the pathogenesis of CRC.
+These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC.
+However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma.
 
-In this study we further refine the random forest model by including more background information on the individuals within it.
-We used a rigourous and accepted approach for machine learning training and testing to validate this model (**insert citation**).
-We then applied it to individuals in which we have samples before and after treatment to assess how well the model adjusts their probability of having an adenoma or carcinoma.
-Finally, we investigated which specific Operational Taxonomic Units (OTUs) are most affected by treatment.
-Within this study lesion refers to both adenoma and carcinoma.
+In this study we investigated what happened to the bacterial microbiome before and after surgery for both adenoma and carcinoma individuals.
+Our anlaysis includes both alpha and beta diversity analysis along with investigating indvidual operational taxonomic units (OTUs).
+We next utilized a Random Forest model that was trained on a completely separate data set and observed how this model as well as specific OTUs within this model performed on the initial and follow up samples.
+We then created a new model to classify intial and follow up samples and investigated how the model performs and what specific OUTs within the model change.
+Finally, we observed how specific bacteria that have been previously implicated in CRC changed before and after treatment in both the adenoma and carcinoma individuals.
 

### 2016-12-22 submission/manuscript_outline_20161024.Rmd
update methods

@@
 ***Study Design and Patient Sampling***
-The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).
+The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].
 In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
@@
 16S rRNA gene sequencing was completed as previously described by Kozich, et al.
-(**insert citation**).
+[@kozich_development_2013].
 In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
@@
 ***Model Creation***
-The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.
-The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer.
+The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.
+The model included data on FIT and the bacterial microbiome.
 Non-binary data was checked for near zero variance and auto correlation.
@@
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
+
+***Analysis Overview***
 

### 2016-12-22 submission/manuscript_outline_20161024.Rmd
add a split for the two different models (methods)

@@
 
-***Model Creation***
+***Lesion Model Creation***
 The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.
@@
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
-Optimiation of the mtry hyperparameter as well as data on the best, middle, and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
+Optimization of the mtry hyperparameter involved taking the samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
 This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).
@@
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
+The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created.
+
+***Initial Follow Up Model Creation***
+We also investigated whether a model could be created that could identify before and after surgery samples.
+The training set utilized the `r length(rownames(good_metaf))`-person cohort that was previously used for testing of the lesion modeld.
+The creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.
 
 ***Analysis Overview***
+This
 

### 2016-12-22 submission/manuscript_outline_20161024.Rmd
rewrite methods order include an analysis overview section

@@
 
-***Analysis Overview***
-This
-
 ***Selection of Important OTUs***
 In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.
-This final collated list of variables was what was considered the most important for the lesion prediction model.
-
-A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.
-These OTUs were considered the most important for predicting initial and follow up samples.
-This new model important OTUs were then compared to the overall lesion model important OTUs and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.
-
-***Investigation of Important CRC Bacteria***
-In order to investigate how specific bacteria that have been previously associated with CRC with selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.
-Specifically we wanted to observe if there were any differences based on whether the individual had an adenoma or carcinoma that may have been missed with our lesion versus normal model.
+This final collated list of variables was what was considered the most important for the lesion or initial follow up models.
 
@@
 
+***Analysis Overview***
+Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigate.
+Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.
+All OTUs used in the lesion model were also analyzed using a paired wilcoxson test.
+The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
+The most important OTUs for this were used to build an updated model and this reduced feature model was assessed for it's similarity to the original model.
+We then used the initial follow up model to assess whether this model could classify samples better then the lesion model.
+The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.
+This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.
+Finally, in order to investigate the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.
+Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
+
 ***Reproducible methods.***
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Baxter_followUps_2016.
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
 

### 2016-12-22 submission/manuscript_outline_20161024.tex
rewrite methods order include an analysis overview section

@@
 \textbf{\emph{Study Design and Patient Sampling}} The sampling and
-design of the study was similar to that reported in Baxter, et al
-(\textbf{insert citation}). In brief, study exclusion involved those who
-had already undergone surgery, radiation, or chemotherapy, had
-colorectal cancer before a baseline stool sample could be obtained, had
-IBD, a known hereditary non-polyposis colorectal cancer, or Familial
-adenomatous polyposis. Samples used to build the model used for
-prediction were collected either prior to a colonoscopy or between 1 - 2
-weeks after. The bacterial microbiome has been shown to nomralize within
-this time period (\textbf{insert citation}). Kept apart from this
-training set were a total of 67 individuals that not only had a sample
-as described previoulsy but also a follow up sample between 188 - 546
-days after surgery and treatment had been completed. This study was
-approved by the University of Michigan Institutional Review Board. All
-study participants provided informed consent and the study itself
-conformed to the guidelines set out by the Helsinki Declaration.
+design of the study was similar to that reported in Baxter, et al (13).
+In brief, study exclusion involved those who had already undergone
+surgery, radiation, or chemotherapy, had colorectal cancer before a
+baseline stool sample could be obtained, had IBD, a known hereditary
+non-polyposis colorectal cancer, or Familial adenomatous polyposis.
+Samples used to build the model used for prediction were collected
+either prior to a colonoscopy or between 1 - 2 weeks after. The
+bacterial microbiome has been shown to nomralize within this time period
+(\textbf{insert citation}). Kept apart from this training set were a
+total of 67 individuals that not only had a sample as described
+previoulsy but also a follow up sample between 188 - 546 days after
+surgery and treatment had been completed. This study was approved by the
+University of Michigan Institutional Review Board. All study
+participants provided informed consent and the study itself conformed to
+the guidelines set out by the Helsinki Declaration.
 
@@
 citation}). 16S rRNA gene sequencing was completed as previously
-described by Kozich, et al. (\textbf{insert citation}). In brief, DNA
-extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
-and an epMotion 5075 automated pipetting system (Eppendorf). The V4
-variable region was amplified and the resulting product was split
-between three sequencing runs with control, adenoma, and carcinoma
-evenly represented on each run. Each group was randomly assigned to
-avoid biases based on sample collection location.
+described by Kozich, et al. (18). In brief, DNA extraction used the 96
+well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
+automated pipetting system (Eppendorf). The V4 variable region was
+amplified and the resulting product was split between three sequencing
+runs with control, adenoma, and carcinoma evenly represented on each
+run. Each group was randomly assigned to avoid biases based on sample
+collection location.
 
@@
 
-\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
-citation}) algorithm was used to create the model used for prediction of
-lesion (adenoma or carcinoma) for the 67 individuals with follow up
-samples. The model included data on FIT, the bacterial microbiome, sex,
-age, Body Mass Index (BMI), whether the indivdiual was caucasion or not,
-history of cancer, and family history of cancer. Non-binary data was
+\textbf{\emph{Lesion Model Creation}} The Random Forest (19) algorithm
+was used to create the model used for prediction of lesion (adenoma or
+carcinoma) for the 67 individuals with follow up samples. The model
+included data on FIT and the bacterial microbiome. Non-binary data was
 checked for near zero variance and auto correlation. Data columns that
@@
 the two columns removed. This pre-processing was performed with the R
-package caret (v6.0.73). Optimiation of the mtry hyperparameter as well
-as data on the best, middle, and worse performance of the model involved
-taking the 490 samples and making 100 80/20 (train/test) splits in the
-data where control and lesion were equally represented in the 80 and 20
-split, respectively. This 80\% portion was then split again into an
+package caret (v6.0.73). Optimization of the mtry hyperparameter
+involved taking the samples and making 100 80/20 (train/test) splits in
+the data where control and lesion were equally represented in the 80 and
+20 split, respectively. This 80\% portion was then split again into an
 80/20 split, and run through 20 repeated 10-fold cross validations to
@@
 completed. The default cutoff of 0.5 was used as the threshold to
-classify individuals as positive or negative for lesion.
+classify individuals as positive or negative for lesion. The
+hyperparameter, mtry, defines the number of variables to investigate at
+each split before a new division of the data is created.
+
+\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
+whether a model could be created that could identify before and after
+surgery samples. The training set utilized the 67-person cohort that was
+previously used for testing of the lesion modeld. The creation of this
+model and optimization of the mtry hyperparameter was completed using
+the same procedure that was used to create the lesion model.
 
@@
 time. This final collated list of variables was what was considered the
-most important for the lesion prediction model.
-
-A second model using the same types of conditions was used to identify
-OTUs that could predict initial and follow up samples. These OTUs were
-considered the most important for predicting initial and follow up
-samples. This new model important OTUs were then compared to the overall
-lesion model important OTUs and the resulting common OTUs to both models
-were selected to test for changes between initial and follow up using a
-paired wilcoxson rank sum test.
-
-\textbf{\emph{Investigation of Important CRC Bacteria}} In order to
-investigate how specific bacteria that have been previously associated
-with CRC with selected OTUs that taxonomically classified to
-Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
-Assacharolytica, and Porphyromonas Stomatis. Specifically we wanted to
-observe if there were any differences based on whether the individual
-had an adenoma or carcinoma that may have been missed with our lesion
-versus normal model.
+most important for the lesion or initial follow up models.
 
@@
 
+\textbf{\emph{Analysis Overview}} Differences in FIT between initial and
+follow ups for either adenoma or carcinoma were investigate. Next,
+initial and follow up samples were analyzed for differences in alpha and
+beta diversity. All OTUs used in the lesion model were also analyzed
+using a paired wilcoxson test. The lesion model was then tested for
+accuracy in prediction and whether it reduced the positive probability
+of lesion after surgery. The most important OTUs for this were used to
+build an updated model and this reduced feature model was assessed for
+it's similarity to the original model. We then used the initial follow
+up model to assess whether this model could classify samples better then
+the lesion model. The most important OTUs were then identified from this
+model and used to create a reduced feature initial follow up model. This
+reduced feature model, as was done with the lesion model, was compared
+to the full model for loss of accuracy. Finally, in order to investigate
+the relative abundance of specific bacteria, that have been previously
+associated with CRC, we selected OTUs that taxonomically classified to
+Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
+Assacharolytica, and Porphyromonas Stomatis. Specifically, we wanted to
+test if there were any differences based on whether the individual had
+an adenoma or carcinoma.
+
 \textbf{\emph{Reproducible methods.}} A detailed and reproducible
 description of how the data were processed and analyzed can be found at
-\url{https://github.com/SchlossLab/Baxter_followUps_2016}.
+\url{https://github.com/SchlossLab/Sze_followUps_2017}.
 
@@
 
+\hypertarget{ref-kozich_development_2013}{}
+18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/AEM.01043-13}{10.1128/AEM.01043-13}.
+
+\hypertarget{ref-breiman_random_2001}{}
+19. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
+\textbf{45}:5--32.
+doi:\href{https://doi.org/10.1023/a:1010933404324}{10.1023/a:1010933404324}.
+
 \end{document}

### 2017-01-03 code/Run_wilcoxson_all.R
add all wilcoxson script

@@ -1,0 +1,24 @@
+### All OTU Paired Wilcoxson Test
+### Are there differences in any OTU in initial and follow up
+### Uses the full lesion model data
+## Marc Sze
+
+
+#Load needed libraries
+source('code/functions.R')
+source('code/graphFunctions.R')
+
+loadLibs(c("dplyr", "scales", "wesanderson", "ggplot2"))
+
+#Load needed data sets
+lesion_data <- read.csv("results/tables/full_test_data.csv", header = T, row.names = 1)
+
+
+
+
+
+
+
+
+
+

### 2017-01-03 submission/supplemental_outline_20161024.Rmd
update supplemental figures to include paired wilcoxson

@@
 
+paired_table <- read.csv("../results/tables/OTU_paired_wilcoxson_test.csv", header = T, stringsAsFactors = F)
+
+label_names <- c("Full Data Set", "Adenoma Only", "CRC Only")
+names(label_names) <- c("ALL", "adn", "crc")
+
+pvalue_distribution_paired <- ggplot(paired_table, aes(x=BH_corrected)) + geom_histogram() + 
+  geom_vline(aes(xintercept = 0.05, color = "red"), linetype = 2, size = 1) + 
+  facet_wrap(~factor(analysis, levels = c("ALL", "adn", "crc")), 
+             labeller = as_labeller(label_names)) + theme_bw() + 
+  xlab("Benjamini-Hochberg Corrected P-values") + ylab("Counts") + 
+  theme(legend.position = "none")
+
+ggsave(file = "../results/figures/FigureS1.pdf", pvalue_distribution_paired, 
+       width=12, height = 6, dpi = 300)
+
+```
+
+```{r supplemental_figure2, echo=FALSE, message=FALSE, warning=FALSE}
+
 diff_table_time <- read.csv("../results/tables/time_datatable.csv", header = T)
@@
 
-ggsave(file = "../results/figures/FigureS1.pdf", time_graph, 
+ggsave(file = "../results/figures/FigureS2.pdf", time_graph, 
        width=6, height = 8, dpi = 300)

### 2017-01-03 submission/manuscript_outline_20161024.tex
update to get correct taxonomies

@@
 before and after treatment included OTUs that were affiliated with
-members of Blautia, Clostridium\_XlVa and Escherichia/Shigella. Our
-model suggests that treatment does significantly reduce the probability
-of having a colonic lesion. Further surveillance of these individuals
-will enable us to determine whether models such as the one we present
-here can also be used to predict recurrence of colorectal cancer.
+members of Blautia, Streptococcus and Escherichia/Shigella. Our model
+suggests that treatment does significantly reduce the probability of
+having a colonic lesion. Further surveillance of these individuals will
+enable us to determine whether models such as the one we present here
+can also be used to predict recurrence of colorectal cancer.
 
@@
 50\% of the training models {[}Table S2{]}. The top 5 most important
-bacterial OTUs were Bacteria (Otu000013), Escherichia/Shigella
-(Otu000018), Bacteria (Otu000020), Ruminococcus (Otu000017), and
+bacterial OTUs were Lachnospiraceae (Otu000013), Escherichia/Shigella
+(Otu000018), Ruminococcaceae (Otu000020), Ruminococcus (Otu000017), and
 Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
@@
 \textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
-Follow Up Versus Initial Samples}} Overall, there were a total of 8 OTUs
-that were common between the main lesion model and the model for
+Follow Up Versus Initial Samples}} Overall, there were a total of 11
+OTUs that were common between the main lesion model and the model for
 classifying initial and follow up samples specifically {[}Table S4{]}. A
@@
 \textbf{\emph{Analysis Overview}} Differences in FIT between initial and
-follow ups for either adenoma or carcinoma were investigate. Next,
+follow ups for either adenoma or carcinoma were investigated. Next,
 initial and follow up samples were analyzed for differences in alpha and

### 2017-01-03 submission/supplemental_outline_20161024.Rmd

--- 2016-12-07 submission/supplemental_outline_20161024.Rmd
@@
 ```
-
-```{r supplemental_figure1, echo=FALSE, message=FALSE, warning=FALSE}
-
-diff_table_time <- read.csv("../results/tables/time_datatable.csv", header = T)
-
-# Plot the change in distance versus time
-time_graph <- ggplot(diff_table_time, aes(x = time, y = distance)) + 
-  geom_point(aes(color = dx), size = 4) + theme_bw() + 
-  scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
-                     breaks = c("adenoma", "cancer"), labels = c("Adenoma", "Cancer")) + 
-  coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("Time from Initial (Days)") + 
-  theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"))
-
-ggsave(file = "../results/figures/FigureS1.pdf", time_graph, 
-       width=6, height = 8, dpi = 300)
-```

### 2017-01-05 submission/manuscript_outline_20161024.tex
update bib, rename supplementa tables, and add Supplemental Figs component

@@
 and after treatment to allow for the assessment of how the model adjusts
-risk after treatment. The model used for prediction had an AUC of 0.763.
+risk after treatment. The model used for prediction had an AUC of NA.
 For the follow up samples our Random Forest model significantly
@@
 exact etiology to disease has not been well worked out (2). Although
-many risk factors have been identified ({\textbf{???}}) and non-invasive
-screening techniques have started to be put into consistent use (4, 5)
-there has been a consistent increase in the incidence of CRC in the
-younger population.
+many risk factors have been identified (2) and non-invasive screening
+techniques have started to be put into consistent use (4, 5) there has
+been a consistent increase in the incidence of CRC in the younger
+population.
 
@@
 maximum AUC trained models. The middle model was close to the full
-training data AUC 0.763. There was no significant difference between the
+training data AUC NA. There was no significant difference between the
 AUC of the best and middle training models (P-value = 1). There was also
@@
 were a total of 37 variables identified as being present in more than
-50\% of the training models {[}Table S2{]}. The top 5 most important
+50\% of the training models {[}Figure S2{]}. The top 5 most important
 bacterial OTUs were Lachnospiraceae (Otu000013), Escherichia/Shigella
@@
 individual with clear CRC on follow up, the probability of an adenoma or
-carcinoma decreased {[}Table S3{]}.
+carcinoma decreased {[}Table S2{]}.
 
@@
 OTUs that were common between the main lesion model and the model for
-classifying initial and follow up samples specifically {[}Table S4{]}. A
+classifying initial and follow up samples specifically {[}Table S3{]}. A
 total of 1 OTU was still significant after multiple comparison
@@
 asaccharolytica (P-value = 0.0154) whereas there were no signficant
-differences in any of these OTUs in the adenoma samples {[}Table S5{]}.
+differences in any of these OTUs in the adenoma samples {[}Table S4{]}.
 
@@
 We then investigated which OTUs could potentially be more important in
-our model {[}Figure 5 \& Table S4{]}. Many of the OTUs identified
-classified to normal flora bacterium {[}Table S4{]}. Only a single OTU
+our model {[}Figure 5 \& Table S3{]}. Many of the OTUs identified
+classified to normal flora bacterium {[}Table S3{]}. Only a single OTU
 though was significant after multiple comparison correction and the
@@
 bacterial microbiome has been shown to nomralize within this time period
-(\textbf{insert citation}). Kept apart from this training set were a
-total of 67 individuals that not only had a sample as described
-previoulsy but also a follow up sample between 188 - 546 days after
-surgery and treatment had been completed. This study was approved by the
-University of Michigan Institutional Review Board. All study
-participants provided informed consent and the study itself conformed to
-the guidelines set out by the Helsinki Declaration.
+(18). Kept apart from this training set were a total of 67 individuals
+that not only had a sample as described previoulsy but also a follow up
+sample between 188 - 546 days after surgery and treatment had been
+completed. This study was approved by the University of Michigan
+Institutional Review Board. All study participants provided informed
+consent and the study itself conformed to the guidelines set out by the
+Helsinki Declaration.
 
@@
 FIT was analyzed as previously published using both OC FIT-CHEK and
-OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
-citation}). 16S rRNA gene sequencing was completed as previously
-described by Kozich, et al. (18). In brief, DNA extraction used the 96
-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
-automated pipetting system (Eppendorf). The V4 variable region was
-amplified and the resulting product was split between three sequencing
-runs with control, adenoma, and carcinoma evenly represented on each
-run. Each group was randomly assigned to avoid biases based on sample
-collection location.
+OC-Auto Micro 80 automated system (Polymedco Inc.) (19). 16S rRNA gene
+sequencing was completed as previously described by Kozich, et al. (20).
+In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
+Laboratories) and an epMotion 5075 automated pipetting system
+(Eppendorf). The V4 variable region was amplified and the resulting
+product was split between three sequencing runs with control, adenoma,
+and carcinoma evenly represented on each run. Each group was randomly
+assigned to avoid biases based on sample collection location.
 
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described (\textbf{insert citations}). The general
-processing workflow using mothur is as follows: Paired-end reads were
-first merged into contigs, quality filtered, aligned to the SILVA
-database, screening for chimeras, classified with a naive Bayesian
-classifier using the Ribosomal Database Project (RDP), and clustered
-into Operational Taxonomic Units (OTUs) using a 97\% similarity cutoff
-with an average neighbor clustering algorithm. The numer of sequences
-for each sample was rarified to 10521 in an attempt to minimize uneven
-sampling.
+has been previously described (20). The general processing workflow
+using mothur is as follows: Paired-end reads were first merged into
+contigs, quality filtered, aligned to the SILVA database, screening for
+chimeras, classified with a naive Bayesian classifier using the
+Ribosomal Database Project (RDP), and clustered into Operational
+Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
+neighbor clustering algorithm. The numer of sequences for each sample
+was rarified to 10521 in an attempt to minimize uneven sampling.
 
-\textbf{\emph{Lesion Model Creation}} The Random Forest (19) algorithm
+\textbf{\emph{Lesion Model Creation}} The Random Forest (21) algorithm
 was used to create the model used for prediction of lesion (adenoma or
@@
 used for all statisitical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA (\textbf{insert citation}) in the
-vegan package (v2.4.1) while comparisons between ROC curves utilized the
-method by DeLong et al. (\textbf{insert citation}) executed by the pROC
-(v1.8) package. Comparisons between probabilities as well as overall
-amount of OTU between initial and follow up samples utilized a paired
-wilcoxson ranked sum test. Where multiple comparison testing was needed
-a Benjamini-Hochberg (BH) correction was applied (\textbf{insert
-citation}) and a corrected P-value of less than 0.05 was considered
-significant. Unless otherwise stated the P-values reported are those of
-the BH corrected ones.
+community structure utilized PERMANOVA (22) in the vegan package
+(v2.4.1) while comparisons between ROC curves utilized the method by
+DeLong et al. (23) executed by the pROC (v1.8) package. Comparisons
+between probabilities as well as overall amount of OTU between initial
+and follow up samples utilized a paired wilcoxson ranked sum test. Where
+multiple comparison testing was needed a Benjamini-Hochberg (BH)
+correction was applied (24) and a corrected P-value of less than 0.05
+was considered significant. Unless otherwise stated the P-values
+reported are those of the BH corrected ones.
 
@@
 
-\textbf{Figure S1: Thetayc Graphed Against Time of Follow up Sample from
+\textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
+Analysis of OTUs in Initial versus Follow Up}
+
+\textbf{Figure S2: Summary of Important Variables in both Lesion and
+Initial/Follow Up models} A) MDA of the most important variables in the
+lesion model. The black point represents the median and the different
+colors are the different runs up to 100. B) The total number of
+appearances of each variable in the 100 different lesion models. The
+cutoff of 50\% was used to assess importance.
+
+\textbf{Figure S3: Thetayc Graphed Against Time of Follow up Sample from
 Initial}
@@
 
+\hypertarget{ref-obrien_impact_2013}{}
+18. \textbf{O'Brien CL}, \textbf{Allison GE}, \textbf{Grimpen F},
+\textbf{Pavli P}. 2013. Impact of colonoscopy bowel preparation on
+intestinal microbiota. PloS One \textbf{8}:e62815.
+doi:\href{https://doi.org/10.1371/journal.pone.0062815}{10.1371/journal.pone.0062815}.
+
+\hypertarget{ref-daly_evaluation_2013}{}
+19. \textbf{Daly JM}, \textbf{Bay CP}, \textbf{Levy BT}. 2013.
+Evaluation of fecal immunochemical tests for colorectal cancer
+screening. Journal of Primary Care \& Community Health
+\textbf{4}:245--250.
+doi:\href{https://doi.org/10.1177/2150131913487561}{10.1177/2150131913487561}.
+
 \hypertarget{ref-kozich_development_2013}{}
-18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+20. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \hypertarget{ref-breiman_random_2001}{}
-19. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
+21. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
 \textbf{45}:5--32.
@@
 
+\hypertarget{ref-anderson_permanova_2013}{}
+22. \textbf{Anderson MJ}, \textbf{Walsh DCI}. 2013. PERMANOVA, ANOSIM,
+and the Mantel test in the face of heterogeneous dispersions: What null
+hypothesis are you testing? Ecological Monographs \textbf{83}:557--574.
+doi:\href{https://doi.org/10.1890/12-2010.1}{10.1890/12-2010.1}.
+
+\hypertarget{ref-delong_comparing_1988}{}
+23. \textbf{DeLong ER}, \textbf{DeLong DM}, \textbf{Clarke-Pearson DL}.
+1988. Comparing the areas under two or more correlated receiver
+operating characteristic curves: A nonparametric approach. Biometrics
+\textbf{44}:837--845.
+
+\hypertarget{ref-benjamini_controlling_1995}{}
+24. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+false discovery rate: A practical and powerful approach to multiple
+testing. Journal of the Royal Statistical Society Series B
+(Methodological) \textbf{57}:289--300.
+
 \end{document}

### 2017-01-06 submission/manuscript_outline_20161024.Rmd
fix typo

@@
 The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].
-Although many risk factors have been identified [@hagger_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.
+Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.
 
@@
 ***Most Important Variables to the Model***
-Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Table S2].
+Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Figure S2].
 The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.
@@
 Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).
-Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].
+Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S2].
 
@@
 ***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
-Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].
+Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S3].
 A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
@@
 First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].
-The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S5].
+The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
 
@@
 
-We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].
-Many of the OTUs identified classified to normal flora bacterium [Table S4].
+We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S3].
+Many of the OTUs identified classified to normal flora bacterium [Table S3].
 Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
@@
 Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).
+The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].
 Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
@@
 ***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
-FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) (**insert citation**).
+FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013].
 16S rRNA gene sequencing was completed as previously described by Kozich, et al.
@@
 The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.
-This process has been previously described (**insert citations**).
+This process has been previously described [@kozich_development_2013].
 The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
@@
 The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.
-Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al.
-(**insert citation**) executed by the pROC (v`r paste(packageVersion("pROC"))`) package.
+Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al.
+[@delong_comparing_1988] executed by the pROC (v`r paste(packageVersion("pROC"))`) package.
 Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.
-Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied (**insert citation**) and a corrected P-value of less than 0.05 was considered significant.
+Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
 Unless otherwise stated the P-values reported are those of the BH corrected ones.
@@
 ***Analysis Overview***
-Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigate.
+Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.
 Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.
@@
 
-**Figure S1: Thetayc Graphed Against Time of Follow up Sample from Initial**
+**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of OTUs in Initial versus Follow Up**
+
+**Figure S2: Summary of Important Variables in both Lesion and Initial/Follow Up models** A) MDA of the most important variables in the lesion model.
+The black point represents the median and the different colors are the different runs up to 100.
+B) The total number of appearances of each variable in the 100 different lesion models.
+The cutoff of 50% was used to assess importance.
+
+**Figure S3: Thetayc Graphed Against Time of Follow up Sample from Initial**
 

### 2017-01-06 submission/supplemental_outline_20161024.Rmd

@@
 
-**Table S2: Summary of Most Important Variables For the Model**
-
-```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
-
-imp_model_vars <- read.csv('../results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
-rf_otu_tax <- read.csv('../results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
-rf_otu_tax <- cbind(otu = rownames(rf_otu_tax), rf_otu_tax)
-
-all_data <- full_join(imp_model_vars, rf_otu_tax, by = c("Variable" = "otu"))
-
-imp_model_vars <- cbind(Variable = imp_model_vars$Variable, Tax = createTaxaLabeller(all_data), 
-                        total_appearance = imp_model_vars$total_appearance)
-
-colnames(imp_model_vars) <- c("Variable", "Lowest Taxonomic ID", "Total Model Appearances")
-
-kable(imp_model_vars, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S3: Summary of Probability Assignment of Follow Up Individuals**
+**Table S2: Summary of Probability Assignment of Follow Up Individuals**
 
@@
 
-**Table S4: Summary of OTUs in both Lesion and Initial/Follow Up Models**
+**Table S3: Summary of OTUs in both Lesion and Initial/Follow Up Models**
 
@@
 
-**Table S5: Summary of P-values for Specific CRC OTUs between Initial and Follow Ups**
+**Table S4: Summary of P-values for Specific CRC OTUs between Initial and Follow Ups**
 

### 2017-01-09 code/Run_Beta_Diversity_tests.R
merge fig 1 and 2

@@ -118,8 +118,10 @@ thetayc.mds.List[["bdiver_Test_crc_IF"]] <- metaMDS(
   scores() %>% as.data.frame() %>% 
   mutate(samples = factor(breakDown_samples))
 
-# Save table for later use
+# Save tables for later use
 write.csv(beta_diver_summary, "results/tables/beta_diver_summary.csv")
+write.csv(thetayc.mds.List[["bdiver_Test_adn_IF"]], "results/tables/thetayc_adn_IF.csv", row.names = F)
+write.csv(thetayc.mds.List[["bdiver_Test_crc_IF"]], "results/tables/thetayc_crc_IF.csv", row.names = F)
 
 
 # Plot the NMDS for Adenoma Only and Carcinoma Only

### 2017-01-09 submission/manuscript_outline_20161024.tex
update text with merge

@@
 the follow up samples, there was no difference between the adenoma and
-carcinoma groups (P-value = 0.697) {[}Figure 1a{]}. There was a
+carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There was a
 difference in FIT between initial and follow up samples with the
@@
 (P-value = 0.997). This change can be visualized using an NMDS {[}Figure
-2{]}. When all follow up samples were compared to each other there was
-no significant overall difference between them (P-value = 0.085). There
-was no significant difference between initial and follow up samples for
-observed OTUs, Shannon diversity, or evenness after correction for
-multiple comparisons {[}Table S1{]}. Time of follow up sample from
-initial sampling, did not have a significant difference between adenoma
-and carcinoma (uncorrected P-value = 0.784).
+1c \& 1d{]}. When all follow up samples were compared to each other
+there was no significant overall difference between them (P-value =
+0.085). There was no significant difference between initial and follow
+up samples for observed OTUs, Shannon diversity, or evenness after
+correction for multiple comparisons {[}Table S1{]}. Time of follow up
+sample from initial sampling, did not have a significant difference
+between adenoma and carcinoma (uncorrected P-value = 0.784).
 
@@
 
-\textbf{\emph{Statistical Analysis}} The R software package (v3.3.0) was
+\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
 used for all statisitical analysis. Comparisons between bacterial
@@
 found between the adenoma and carcinoma group for thetayc (P-value =
-0.697). B) A significant difference was found between the adenoma and
+0.00198). B) A significant difference was found between the adenoma and
 carcinoma group for FIT (P-value = 2.15e-05).

### 2017-01-09 code/Run_Figure2.R
convert fig 6 to 2

@@ -42,7 +42,6 @@ crc_specific <- grid.arrange(
     theme(legend.title = element_text(face="bold", size = 8), 
           legend.text = element_text(size = 6), 
           legend.position = c(0.38, 0.82), 
-          legend.margin = margin(0, 0, 0, 0), 
           plot.margin = unit(c(1, 1, 1, 1), "lines"), 
           plot.title = element_text(size=20, face="bold"), 
           strip.text.x = element_text(size = 8)), 
@@ -63,7 +62,6 @@ crc_specific <- grid.arrange(
     theme(legend.title = element_text(face="bold", size = 8), 
           legend.text = element_text(size = 6), 
           legend.position = c(0.35, 0.85), 
-          legend.margin = margin(0, 0, 0, 0), 
           plot.margin = unit(c(1, 1, 1, 1), "lines"), 
           plot.title = element_text(size=20, face="bold"), 
           strip.text.x = element_text(size = 8))
@@ -71,7 +69,7 @@ crc_specific <- grid.arrange(
   )
 
 
-ggsave(file = "results/figures/Figure6.pdf", crc_specific, 
+ggsave(file = "results/figures/Figure2.pdf", crc_specific, 
        width=8.5, height = 11, dpi = 300)
 
 

### 2017-01-09 submission/manuscript_outline_20161024.Rmd
update text with merge
update with fig 2 to 6 swtich

@@
 Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-This change can be visualized using an NMDS [Figure 2].
+This change can be visualized using an NMDS [Figure 1c & 1d].
 When all follow up samples were compared to each other there was no significant overall difference between them (P-value = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
@@
 In contrast, FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].
-An NMDS showed that there was very little observable change between initial and follow up for the adenoma group but there was one for the carcinoma group [Figure 2].
+An NMDS showed that there was very little observable change between initial and follow up for the adenoma group but there was one for the carcinoma group [Figure 1c & 1d].
 This cursory information is suggestive that treatment of carcinoma, had the largest response.
@@
 
-**Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+**Figure 1: Changes and differences between the adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
 B) A significant difference was found between the adenoma and carcinoma group for FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+C) NMDS of the intial and follow up samples for the Adenoma group.
+D) NMDS of the initial and follow up samples for the Carcinoma group.
+For C) and D) the teal represents initial samples and the pink represents follow up samples.
 
-**Figure 2: NMDS of the Overall Bacterial Community Changes.**  A) NMDS of the intial and follow up samples for the Adenoma group.
-B) NMDS of the initial and follow up samples for the Carcinoma group.
+**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
+There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).
+B) Adenoma initial and follow up samples.
+There were no significant differences between initial and follow up.
 
@@
 The dotted line represents the average relative abundance in the control training group.
-
-**Figure 6: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
-There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).
-B) Adenoma initial and follow up samples.
-There were no significant differences between initial and follow up.
 

### 2017-01-09 code/Run_Figure3.R
incorporate updated Fig 3

@@ -21,14 +21,13 @@ lesion_auc_info <- read.csv("results/tables/auc_summary.csv",
 IF_model_roc <- read.csv("results/tables/IF_test_data_roc.csv", header = T)
 reduced_IF_model_roc <- read.csv("results/tables/reduced_IF_test_data_roc.csv", header = T)
 
-grid.arrange(
+roc_curve_graph <- grid.arrange(
   # Graph showing all variable and reduced variable lesion model
   filter(lesion_model_roc, run != "middle_roc") %>% 
     ggplot(aes(x = sensitivities, y = specificities)) + 
     geom_polygon(alpha = 0.25) + 
     geom_line(aes(group = run), size = 1.25, color = "black", alpha = 0.5) + 
-    geom_polygon(data = filter(reduced_lesion_model_roc, run != "full_roc"), 
-                 aes(y = specificities), alpha = 0.25, fill = "red") + 
+    geom_polygon(data = filter(reduced_lesion_model_roc, run != "full_roc"), alpha = 0.25, fill = "red") + 
     geom_line(data = filter(reduced_lesion_model_roc, run != "full_roc"), 
               aes(group = run), size = 1.25, color = "red", alpha = 0.5) + 
     geom_line(data = filter(reduced_lesion_model_roc, run == "full_roc"), 
@@ -40,8 +39,7 @@ grid.arrange(
   ggplot(IF_model_roc, aes(x = sensitivities, y = specificities)) + 
     geom_polygon(alpha = 0.25) + 
     geom_line(aes(group = run), size = 1.25, color = "black", alpha = 0.5) + 
-    geom_polygon(data = filter(reduced_IF_model_roc, run != "full_roc"), 
-                 aes(y = specificities), alpha = 0.25, fill = "red") + 
+    geom_polygon(data = filter(reduced_IF_model_roc, run != "full_roc"), alpha = 0.25, fill = "red") + 
     geom_line(data = filter(reduced_IF_model_roc, run != "full_roc"), 
               aes(group = run), size = 1.25, color = "red", alpha = 0.5) + 
     geom_line(data = filter(reduced_IF_model_roc, run == "full_roc"), 
@@ -52,7 +50,7 @@ grid.arrange(
  
 
 # Save graph image as a pdf
-ggsave(file = "results/figures/Figure3.pdf", roc_curve_graph, 
+ggsave(file = "results/figures/Figure3.pdf", roc_curve_graph, device = "pdf", 
        width=8, height = 8, dpi = 300)
 
 

### 2017-01-09 code/Run_Figure3.R
update code for pdf write out

@@ -23,7 +23,7 @@ reduced_IF_model_roc <- read.csv("results/tables/reduced_IF_test_data_roc.csv", 
 
 roc_curve_graph <- grid.arrange(
   # Graph showing all variable and reduced variable lesion model
-  filter(lesion_model_roc, run != "middle_roc") %>% 
+  test <- filter(lesion_model_roc, run != "middle_roc") %>% 
     ggplot(aes(x = sensitivities, y = specificities)) + 
     geom_polygon(alpha = 0.25) + 
     geom_line(aes(group = run), size = 1.25, color = "black", alpha = 0.5) + 
@@ -50,7 +50,7 @@ roc_curve_graph <- grid.arrange(
  
 
 # Save graph image as a pdf
-ggsave(file = "results/figures/Figure3.pdf", roc_curve_graph, device = "pdf", 
+ggsave(file = "results/figures/Figure3.tiff", roc_curve_graph, device = "tiff", 
        width=8, height = 8, dpi = 300)
 
 

### 2017-01-09 submission/manuscript_outline_20161024.Rmd

@@
 
-**Figure 3: Graph of the Receiver Operating Characteristic Curve on Test Set Performance of the Best, Middle, Worse, and Full Training Models.** For each of the 100 training cohort sets used had `r round(length(rownames(metaI))*0.8)` individuals and the testing cohort sets had `r round(length(rownames(good_metaf))*0.2)` individuals.
-The AUC on the test sets for the best, middle, and worse models from training were `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, and `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`, respectively.
-cvAUC is the 20 times repeated 10-fold cross-validated AUC from training.
+**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Initial/Follow up models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
+The blue line represents the reduced variable model using 100% of the data set.
+A) Lesion model.
+B) Initial/Follow up model
 

### 2017-01-09 submission/manuscript_outline_20161024.tex

@@
 little observable change between initial and follow up for the adenoma
-group but there was one for the carcinoma group {[}Figure 2{]}. This
-cursory information is suggestive that treatment of carcinoma, had the
-largest response.
+group but there was one for the carcinoma group {[}Figure 1c \& 1d{]}.
+This cursory information is suggestive that treatment of carcinoma, had
+the largest response.
 
@@
 
-\textbf{Figure 1: Change in Thetayc and Fit between initial and follow
-up in adenoma or carcinoma group.} A) No significant difference was
-found between the adenoma and carcinoma group for thetayc (P-value =
-0.00198). B) A significant difference was found between the adenoma and
-carcinoma group for FIT (P-value = 2.15e-05).
+\textbf{Figure 1: Changes and differences between the adenoma or
+carcinoma group.} A) No significant difference was found between the
+adenoma and carcinoma group for thetayc (P-value = 0.00198). B) A
+significant difference was found between the adenoma and carcinoma group
+for FIT (P-value = 2.15e-05). C) NMDS of the intial and follow up
+samples for the Adenoma group. D) NMDS of the initial and follow up
+samples for the Carcinoma group. For C) and D) the teal represents
+initial samples and the pink represents follow up samples.
 
-\textbf{Figure 2: NMDS of the Overall Bacterial Community Changes.} A)
-NMDS of the intial and follow up samples for the Adenoma group. B) NMDS
-of the initial and follow up samples for the Carcinoma group.
+\textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
+Follow up Samples.} A) Carcinoma initial and follow up samples. There
+was a significant difference in initial and follow up sample for the
+OTUs classfied as Peptostreptococcus stomatis (P-value = 0.0183) and
+Porphyromonas asaccharolytica (P-value = 0.0154). B) Adenoma initial and
+follow up samples. There were no significant differences between initial
+and follow up.
 
 \textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
-on Test Set Performance of the Best, Middle, Worse, and Full Training
-Models.} For each of the 100 training cohort sets used had 392
-individuals and the testing cohort sets had 13 individuals. The AUC on
-the test sets for the best, middle, and worse models from training were
-0.646, 0.744, and 0.904, respectively. cvAUC is the 20 times repeated
-10-fold cross-validated AUC from training.
+for lesion and Initial/Follow up models.} The shaded areas represents
+the range of values of a 100 different 80/20 splits of the test set data
+using either all variables (grey) or reduced variable (red) models. The
+blue line represents the reduced variable model using 100\% of the data
+set. A) Lesion model. B) Initial/Follow up model
 
@@
 relative abundance in the control training group.
-
-\textbf{Figure 6: Previously Associated CRC Bacteria in Initial and
-Follow up Samples.} A) Carcinoma initial and follow up samples. There
-was a significant difference in initial and follow up sample for the
-OTUs classfied as Peptostreptococcus stomatis (P-value = 0.0183) and
-Porphyromonas asaccharolytica (P-value = 0.0154). B) Adenoma initial and
-follow up samples. There were no significant differences between initial
-and follow up.
 

### 2017-01-11 submission/manuscript_outline_20161024.Rmd
formatting changes

@@
 title: ''
-csl: mbio.csl
+csl: microbiome.csl
 fontsize: 12pt
@@
 # model prediction and adjustment performance
-IF_model_wilcox_paired_summary <- read.csv('../results/tables/IF_model_wicox_paired_pvalue_summary.csv', header = T, row.names = 1)
+IF_model_wilcox_paired_summary <- read.csv('../results/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T)
 confusion_summary <- read.csv('../results/tables/confusion_summary.csv', header = T, row.names = 1)
@@
 
-\vspace{10mm}
-
-Contributions: All authors were involved in the conception and design of the study.
-MAS analyzed the data.
-NTB processed samples and analyzed the data.
-All authors interpreted the data.
-MAS and PDS wrote the manuscript.
-All authors reviewed and revised the manuscript.
-All authors read and approved the final manuscript.
-
 \vspace{20mm}
@@
 
-Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.
+**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.
 Previous research suggests a link between stool bacterial microbiome and CRC.
@@
 This model was tested on a `r sum(initial_cm$ref_yes)` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
-The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
+**Results:** The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
 For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)`) and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)`).
@@
 Some OTUs that changed the most before and after treatment included OTUs that were affiliated with members of `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][1:2]` and `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][3]`.
-Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.
+**Conclusions:** Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.
 Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.
@@
 
-### Importance
+### Keywords
 
-This is one of the first studies to investigate within humans what happens to the bacterial microbiome before and after adenoma and carcinoma treatment.
-Specifically, it aims to assess how a random forest machine learning algorithm built model respond to treatment and adjust it's positive probability calls of whether the individual has an adenoma or carcinoma due to surgical removal of the lesion.
+bacterial microbiome; colorectal cancer; polyps
 
@@
 
-## Introduction
+## Background
 
@@
 
-## Acknowledgements
-
-The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-This study was supported by funding from the National Institutes of Health to P.
-Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).
-
-\newpage
-
 **Figure 1: Changes and differences between the adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
@@
 
-**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Initial and Follow Up***  A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up for Reduced Variable Models**  A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
 The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.
-
-**Figure 5: Lesion Model OTU with a Significant Decrease in Relative Abundance that is also Predictive of Initial and Follow Up.**  After multiple comparison correction `r rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05))` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was the only one with a P-value < 0.05.
-The dotted line represents the average relative abundance in the control training group.
 
@@
 
-**Figure S2: Summary of Important Variables in both Lesion and Initial/Follow Up models** A) MDA of the most important variables in the lesion model.
+**Figure S2: Summary of Important Variables in the Lesion Model** A) MDA of the most important variables in the lesion model.
 The black point represents the median and the different colors are the different runs up to 100.
@@
 
-**Figure S3: Thetayc Graphed Against Time of Follow up Sample from Initial**
+**Figure S3: Summary of Important Variables in Initial Follow Up Model** A) MDA of the most important variables in the lesion model.
+The black point represents the median and the different colors are the different runs up to 100.
+B) The total number of appearances of each variable in the 100 different lesion models.
+The cutoff of 50% was used to assess importance.
+
+**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up for Full Variable Model** A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.
+
+**Figure S5: Thetayc Graphed Against Time of Follow up Sample from Initial**
+
+\newpage
+
+## Declarations
+
+### Ethics approval and consent to participate
+
+### Consent for publication
+
+### Availability of data and material
+
+### Competing Interests
+
+All authors declare that they do not have any relevent competing interests to report.
+
+### Funding
+
+This study was supported by funding from the National Institutes of Health to P. Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).
+
+### Authors' contributions
+
+All authors were involved in the conception and design of the study.
+MAS analyzed the data.
+NTB processed samples and analyzed the data.
+All authors interpreted the data.
+MAS and PDS wrote the manuscript.
+All authors reviewed and revised the manuscript.
+All authors read and approved the final manuscript.
+
+### Acknowledgements
+
+The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
 

### 2017-01-11 submission/manuscript_outline_20161024.Rmd
update tables to be loaded

@@
 time_pvalues <- read.csv("../results/tables/time_pvalues.csv", header = T, row.names = 1)
+all_OTU_pair_wilcox <- read.csv("../results/tables/OTU_paired_wilcoxson_test.csv", header = T)
 
-# training of model data
-model_train_summary <- read.csv('../results/tables/ROC_model_summary.csv', header = T)
-model_test_AUC_summary <- read.csv('../results/tables/auc_summary.csv', header = T, row.names = 1)
-imp_model_vars <- read.csv('../results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
-rf_otu_tax <- read.csv('../results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
-rf_lowest_ID <- createTaxaLabeller(rf_otu_tax)
+# Lesion Model (Full)
+lesion_top10_varsMDA <- read.csv('../results/tables/lesion_model_top_vars_MDA_Summary.csv', header = T)
+lesion_imp_vars <- read.csv('../results/tables/rf_otu_tax.csv', header = T, row.names = 1)
+lesion_AUC_run_summary <- read.csv('../results/tables/ROC_model_summary.csv', header = T)
 
-# model prediction and adjustment performance
-IF_model_wilcox_paired_summary <- read.csv('../results/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T)
-confusion_summary <- read.csv('../results/tables/confusion_summary.csv', header = T, row.names = 1)
-follow_up_prob_summary <- read.csv('../results/tables/follow_up_probability_summary.csv', header = T)
-followup_cm <- read.csv('../results/tables/followup_confusion_matrix.csv', header = T, row.names = 1)
-initial_cm <- read.csv('../results/tables/initial_confusion_matrix.csv', header = T, row.names = 1)
-roc_pvalues <- read.csv('../results/tables/roc_pvalue_summary.csv', header = T, row.names = 1)
+# Lesion Model (Reduced)
+reduced_lesion_AUC_run_summary <- read.csv('../results/tables/reduced_IF_ROC_model_summary.csv', header = T)
 
-# Predictive of both initial and follow up and in lesion model
-pvalue_common_imp_vars <- read.csv('../results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
+# IF Model (Full)
+IF_top10_varsMDA <- read.csv('../results/tables/IF_model_top_vars_MDA_Summary.csv', header = T)
+IF_imp_vars <- read.csv('../results/tables/if_rf_otu_tax.csv', header = T, row.names = 1)
+IF_AUC_run_summary <- read.csv('../results/tables/IF_ROC_model_summary.csv', header = T)
 
-# Results of CRC Bacteria
-pvalue_adn_crc_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
+# IF Model (Reduced)
+reduced_IF_AUC_run_summary <- read.csv('../results/tables/reduced_IF_ROC_model_summary.csv', header = T)
+
+# All model information
+all_probs_paired_wilcox <- read.csv('../results/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
+all_pred_stat_summary <- read.csv('../results/tables/all_models_summary_info.csv', header = T, stringsAsFactors = F)
+all_confu_matrix <- read.csv('../results/tables/all_models_confusion_summary.csv', header = T, stringsAsFactors = F)
+
+# Common Variable Info
+common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1)
 
@@
 
-bacterial microbiome; colorectal cancer; polyps
+bacterial microbiome; colorectal cancer; polyps; FIT; detection; risk factors
 

### 2017-01-11 submission/manuscript_outline_20161024.Rmd
add affiliations

@@
 
-Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss${^\dagger}$
+Marc A Sze^1^, Nielson T Baxter^2^, Mack T Ruffin IV^3^, Mary AM Rogers^4^, and Patrick D Schloss^1^${^\dagger}$
 
@@
 
-Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+^1^Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+^2^Department of Internal Medicine, Division of Infectious Disease, University of Michigan, Ann Arbor, MI
+^3^Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+^4^Department of Internal Medicine, University of Michigan, Ann Arbor, MI
 \end{center}

### 2017-01-11 submission/manuscript_outline_20161024.Rmd
update background of abstract

@@
 We then used this model to characterize the change in the gut microbiota before and after surgery.
-The overall objective was to investigate the changes in the microbiome after surgery in patients with adenomas or carcinomas.
-This model was tested on a `r sum(initial_cm$ref_yes)` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment.
+The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion.
+Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.
 **Results:** The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.

### 2017-01-11 submission/manuscript_outline_20161024.Rmd
update background abstract

@@
 Previous research suggests a link between stool bacterial microbiome and CRC.
-In this study, we used a model based on the microbiome, demographics, and prior medical history to classify individuals as having a lesion (i.e.
-adenoma or carcinoma).
-We then used this model to characterize the change in the gut microbiota before and after surgery.
+In this study, we used the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion (i.e.
+adenoma or carcinoma) or whether the sample was before or after surgery.
 The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion.

### 2017-01-12 code/Run_Figure1.R
add proper label names

@@ -25,14 +25,15 @@ thetayc_summary_table <- read.csv("results/tables/beta_diver_summary.csv", heade
 differences_graph <- grid.arrange(
   # Difference from bacterial community structure between adenoma and cancer
   ggplot(difference_table_treatment, 
-         aes(factor(dx, levels = c("adenoma", "cancer")), distance, group = 1)) + 
+         aes(factor(dx, levels = c("adenoma", "cancer"), labels = c("Adenoma", "Carcinoma")), 
+             distance, group = 1)) + 
     geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
     stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
     stat_summary(fun.y = mean, colour = "black", geom = "line") + 
     scale_color_manual(name = "Lesion Type", 
                        values = wes_palette("GrandBudapest"), 
                        breaks = c("adenoma", "adv_adenoma", "cancer"), 
-                       labels = c("Adenoma", "SRN", "Cancer")) + 
+                       labels = c("Adenoma", "SRN", "Carcinoma")) + 
     coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + 
     xlab("") + theme_bw() + ggtitle("A") + 
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
@@ -46,7 +47,7 @@ differences_graph <- grid.arrange(
   
   # Difference from fit between adenoma and cancer
   ggplot(difference_table_treatment, 
-         aes(factor(dx, levels = c("adenoma", "cancer")), 
+         aes(factor(dx, levels = c("adenoma", "cancer"), labels = c("Adenoma", "Carcinoma")), 
              fit_difference*-1, group = 1)) + 
     geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
     stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
@@ -54,7 +55,7 @@ differences_graph <- grid.arrange(
     scale_color_manual(name = "Lesion Type", 
                        values = wes_palette("GrandBudapest"), 
                        breaks = c("adenoma", "adv_adenoma", "cancer"), 
-                       labels = c("Adenoma", "SRN", "Cancer")) + 
+                       labels = c("Adenoma", "SRN", "Carcinoma")) + 
[…]

### 2017-01-12 submission/manuscript_outline_20161024.Rmd
modify abstract partway through results part

--- 2017-01-11 submission/manuscript_outline_20161024.Rmd
@@
 Previous research suggests a link between stool bacterial microbiome and CRC.
-In this study, we used a model based on the microbiome, demographics, and prior medical history to classify individuals as having a lesion (i.e.
+The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e.
 adenoma or carcinoma).
-We then used this model to characterize the change in the gut microbiota before and after surgery.
-The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion.
 Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.
-**Results:** The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
+We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.
+**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+
+fo The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
 For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)`) and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)`).

### 2017-01-12 submission/manuscript_outline_20161024.Rmd
3/4 results completed abstract

@@
 
+# Main CRC bacteria differences between Adenoma and Carcinoma
+main_crc_bugs_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
+
 # Lesion Model (Full)
@@
 # Lesion Model (Reduced)
-reduced_lesion_AUC_run_summary <- read.csv('../results/tables/reduced_IF_ROC_model_summary.csv', header = T)
+reduced_lesion_AUC_run_summary <- read.csv('../results/tables/Reduced_ROC_model_summary.csv', header = T)
 
@@
 # Common Variable Info
-common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1)
+common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
 
@@
 **Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+There was no significant difference in any indivdiual OTU between samples before and after surgery (P > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from the genera belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`
 

### 2017-01-12 submission/manuscript_outline_20161024.Rmd
abstract update completeion

@@
 A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from the genera belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`
-
-fo The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
-For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)`) and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)`).
-Our model predicted that `r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)`% of the 67-person cohort had normal colons and `r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)`% had a lesion.
-Some OTUs that changed the most before and after treatment included OTUs that were affiliated with members of `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][1:2]` and `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][3]`.
-**Conclusions:** Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.
+Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and 'r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`
+**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.
 Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

### 2017-01-12 submission/manuscript_outline_20161024.Rmd
complete intro edit

@@
 Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].
-Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].
-However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].
+Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
+However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].
 It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007].
-The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].
-Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.
+The vast majority of CRCs are not inherited and the exact etiology of the disease has not been well worked out [@haggar_colorectal_2009].
+Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been an consistent additional increase in the incidence of CRC in the younger population.
 
-This increased incidence of CRC in the younger population is concerncing since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].
-This increased risk can also carry with it an increased risk of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].
-Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at risk of reccurence.
-Concurrently with this there has been a lot of interest in new areas such as the gut bacterial microbiome for insight into potential disease pathology.
+This increased incidence of CRC in the younger population is concerning since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].
+This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].
+Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at most susceptibility to reccurence.
+Concurrently, there has also been a lot of interest in new areas that could have a role in disease pathogenesis, such as the gut bacterial microbiome.
 
@@
 There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
-There have also been a few studies that have shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved ith the pathogenesis of CRC.
+A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
 These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC.
-However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma.
+However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.
 
 In this study we investigated what happened to the bacterial microbiome before and after surgery for both adenoma and carcinoma individuals.
-Our anlaysis includes both alpha and beta diversity analysis along with investigating indvidual operational taxonomic units (OTUs).
-We next utilized a Random Forest model that was trained on a completely separate data set and observed how this model as well as specific OTUs within this model performed on the initial and follow up samples.
-We then created a new model to classify intial and follow up samples and investigated how the model performs and what specific OUTs within the model change.
-Finally, we observed how specific bacteria that have been previously implicated in CRC changed before and after treatment in both the adenoma and carcinoma individuals.
+Our anlaysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
+We also utilized Random Forest models and observed how these models well as specific OTUs within this model performed on before (initial) and after (follow up) surgery samples.
+We used these models to look for similar important OTUs to identify the most important ones for not only classifying initial and follow up samples but also lesion or normal.
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
75% completed results modification

@@
 main_crc_bugs_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
+crc_bug_info <- read.csv("../results/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)
 
@@
 ***Bacterial Community and Fit Changes before and after Treatment***
-Based on thetayc distance metrics, comparing the initial to the follow up samples, there was no difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
-There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-This change can be visualized using an NMDS [Figure 1c & 1d].
-When all follow up samples were compared to each other there was no significant overall difference between them (P-value = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
+Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
+There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
+The whole community structure before and after surgery are visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
 Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
 
-***Outcome of Model Training***
-The range of the AUC for model training ranged from a minimum of `r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)` to a maximum of `r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)` with the middle of all `r length(rownames(model_train_summary))` runs having an AUC of `r format(model_test_AUC_summary["middle", "ROC_cv"],
-digits = 3)`.
-Interestingly, the worst AUC model from training performed the best on it's respective 80/20 split test data [Figure 3].
-In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, middle training model test set AUC = `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, worse training model test set AUC = `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`).
-That is to say it had the smallest change in AUC in comparison to the minimum and maximum AUC trained models.
-The middle model was close to the full training data AUC `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`.
-There was no significant difference between the AUC of the best and middle training models (P-value = `r format(roc_pvalues["best", "middle"], digits = 3)`).
-There was also no difference in the middle model versus worse (P-value = `r format(roc_pvalues["middle", "worse"], digits = 3)`) or full data model (P-value = `r format(roc_pvalues["middle", "full"], digits = 3)`).
-The two comparisons with a significant difference were between the worse training model and the best training model (P-value = `r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)`) and the worse training model and full training model (P-value = `r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)`).
+***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
+We next examined whether there were and differences in previously well described carcinoma associated OTUs.
+These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
+First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
+Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].
+However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.
+
+***Full and Reduced Model Results***
+Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.
+The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.
+Interestingly, identification of the most important variables and reducing the variables considered to only these increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+
+The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].
+The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].
 
@@
 There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).
-
-***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
-Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S3].
-A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
-In general, `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was decreased from initial to follow up [Figure 5].
-The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].
-
-***Differences in Adenoma and Carcinoma in Previously Associated Cancer Bacteria***
-First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].
-The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add important variable component to results

@@
 There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
-Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
+There was no significant difference between initial and follow up samples for any single OTU [Figure S1].
 
@@
 The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].
+Interestingly, the test set AUC range for the initial and follow up performed much better then the training set AUCs.
+Both the full and reduced initial and follow up models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.
 
-***Most Important Variables to the Model***
-Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Figure S2].
-The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.
-These 5 OTUs were present in at least `r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])` out of the total `r length(rownames(model_train_summary))` different 80/20 runs.
+***Most Important Variables to the Models***
+The reduced models were built based on the most important variables to the respective full model.
+For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial and follow up model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
+For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].
 
-***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
-A total of `r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))` sample was omitted from the original `r length(rownames(good_metaf))` test sample set since it was missing a complete set of follow up data.
-This left a total of `r sum(initial_cm$ref_yes)` samples for test predictions.
-After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = `r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4].
-This decrease was significant for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)` [Figure 4a] and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4b] alone.
-This decrease in probability of lesion also held specifically for those with screen relative neoplasias (SRN) (P-value = `r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)`).
-A total of `r initial_cm["pred_yes", "ref_yes"]` or `r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)`% of all samples were correctly predicted to have a lesion.
-Although there was a decrease in positive probability only `r followup_cm["pred_no", "ref_no"]` of the total `r sum(initial_cm$ref_yes)` individuals were classified as adenoma or carcinoma free on follow up (successful classification of `r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)`%).
-There was no significant difference between the predictions and actual diagnosis for the initial samples in the `r length(rownames(good_metaf))`-sample cohort test set.
-However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = `r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)`).
-Although there were discordent results the respective sensitivity for the initial group was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)`% and for follow up was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`%, respectively.
+***Positive Probability Prediction after Surgical Removal of Adenoma or Carcinoma***
 
-There was `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual who still clearly had CRC on follow up as well as `r length(rownames(filter(good_metaf, Disease_Free == "unknown")))` individuals whose status on follow up was unknown.
-Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).
-Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S2].
+**Treatment and Time Differences**
 
-The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)`) or  radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).
 There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).
+Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add Fig 4 and S4 to manuscript results

@@
 ***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
-We next examined whether there were and differences in previously well described carcinoma associated OTUs.
+We next examined whether there were differences in previously well described carcinoma associated OTUs.
 These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
@@
 ***Positive Probability Prediction after Surgical Removal of Adenoma or Carcinoma***
+Regardless of model used there was a significant decrease in the positive probability of either the sample being lesion or an initial sample on follow up (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+The lesion model correctly kept the one individual who still had a carcinoma on follow up above the cut off point [Figure 4a & S4a] for a positive call while the initial and follow up model did not [Figure 4b & S4b].
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
added common OTUs and model data to results

--- 2017-01-13 submission/manuscript_outline_20161024.Rmd
@@
 ***Positive Probability Prediction after Surgical Removal of Adenoma or Carcinoma***
+Regardless of model used there was a significant decrease in the positive probability of either the sample being lesion or an initial sample on follow up [Figure 4 & S4] (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+
+For the full and reduced lesion model there was a significant difference in the classification for the lesion model between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)`, respectively) but not for the initial follow up model (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)`).
+However, the lesion model correctly kept the one individual who still had a carcinoma on follow up above the cut off point [Figure 4a & S4a] for a positive call while the initial and follow up models did not [Figure 4b & S4b].
+
+***Common OTUs to both Lesion and Initial and Follow Up Models***
+There were a total of `r length(common_vars_summary$otu)` OTUs that were common to both models.
+Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
complete result addition of common OTUs

@@
 Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.
+The majority of these OTUs had taxonomic identification to bacteria typically thought of as commensal [Table S2].
 
@@
 **Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
-There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).
+There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
 B) Adenoma initial and follow up samples.

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add updated supplement

@@
 common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
+
+#Therapy Info
+chemo_rad_summary <- read.csv('../results/tables/probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
 
@@
 
-There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).
-Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
+There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).
+Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add pdf with completed results

@@
 ---
-title: ''
-csl: microbiome.csl
-fontsize: 12pt
+title: ""
+bibliography: reference.bib
 output:
@@
   html_document: default
+csl: microbiome.csl
+fontsize: 12pt
 geometry: margin=1.0in
-bibliography: reference.bib
 ---
@@
 ```
-
+\begin{center}  
 \vspace{25mm}
 
-\begin{center}
 Running Title: Human Microbiome and Colorectal Cancer
@@
 **Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-There was no significant difference in any indivdiual OTU between samples before and after surgery (P > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.
-Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and 'r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`
+There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.
+Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
 **Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.
@@
 These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
+First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
 Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].
@@
 There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.
-The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.
-Although there was a decrease in overall probability of an adenoma or carcinoma only `r followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.
+The overall sensitivity for lesion detection in the intial samples was `r #format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r #format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.
+Although there was a decrease in overall probability of an adenoma or carcinoma only `r #followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r #sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.
 
@@
 Many of the OTUs identified classified to normal flora bacterium [Table S3].
-Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
+Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
 Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].

### 2017-01-13 submission/manuscript_outline_20161024.tex
add pdf with completed results

@@
 
+\textbackslash{}begin\{center\}\\
 \vspace{25mm}
 
-\begin{center}
 Running Title: Human Microbiome and Colorectal Cancer
@@
 
-Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss${^\dagger}$
-
-\vspace{10mm}
-
-Contributions: All authors were involved in the conception and design of the study.
-MAS analyzed the data.
-NTB processed samples and analyzed the data.
-All authors interpreted the data.
-MAS and PDS wrote the manuscript.
-All authors reviewed and revised the manuscript.
-All authors read and approved the final manuscript.
+Marc A Sze\textsuperscript{1}, Nielson T Baxter\textsuperscript{2}, Mack
+T Ruffin IV\textsuperscript{3}, Mary AM Rogers\textsuperscript{4}, and
+Patrick D Schloss\textsuperscript{1}\({^\dagger}\)
 
@@
 
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
+\(\dagger\) To whom correspondence should be addressed:
+\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
 
-Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
-\end{center}
+\textsuperscript{1}Department of Microbiology and Immunology, University
+of Michigan, Ann Arbor, MI\\
+\textsuperscript{2}Department of Internal Medicine, Division of
+Infectious Disease, University of Michigan, Ann Arbor, MI\\
+\textsuperscript{3}Department of Family Medicine and Community Medicine,
+Penn State Hershey Medical Center, Hershey, PA\\
+\textsuperscript{4}Department of Internal Medicine, University of
+Michigan, Ann Arbor, MI \textbackslash{}end\{center\}
 
@@
 
-Colorectal cancer (CRC) continues to be a worldwide health problem with
-early detection being used as a key component in mitigating deaths due
-to the disease. Previous research suggests a link between stool
-bacterial microbiome and CRC. In this study, we used a model based on
-the microbiome, demographics, and prior medical history to classify
-individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
-this model to characterize the change in the gut microbiota before and
-after surgery. The overall objective was to investigate the changes in
-the microbiome after surgery in patients with adenomas or carcinomas.
-This model was tested on a 66 person group that included samples before
-and after treatment to allow for the assessment of how the model adjusts
-risk after treatment. The model used for prediction had an AUC of NA.
-For the follow up samples our Random Forest model significantly
-decreased the positive probability of lesion compared to the initial
-samples for both adenoma (P-value = 3.64e-11) and carcinoma (P-value =
-7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
-normal colons and 63.6\% had a lesion. Some OTUs that changed the most
-before and after treatment included OTUs that were affiliated with
-members of Blautia, Streptococcus and Escherichia/Shigella. Our model
-suggests that treatment does significantly reduce the probability of
-having a colonic lesion. Further surveillance of these individuals will
-enable us to determine whether models such as the one we present here
-can also be used to predict recurrence of colorectal cancer.
+\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
+health problem with early detection being used as a key component in
+mitigating deaths due to the disease. Previous research suggests a link
+between stool bacterial microbiome and CRC. The overall objective was to
+investigate the changes in the bacterial microbiome after surgery in
+patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
+wanted to identify what within the community was different within those
+undergoing surgical removal of lesion. We also investigated the use of
+the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
+models to either classify individuals as having a lesion or whether,
+based on the bacterial microbiome, the sample could be classified
+correctly as before or after surgery.\\
+\textbf{Results:} Adenoma individual's bacterial microbiome were more
+similar to their pre-surgery sample then those with carcinoma (P-value =
+0.00198) and this was also reflected in FIT as well (P-value =
+2.15e-05). There was no significant difference in any indivdiual OTU
+between samples before and after surgery (P-value \textgreater{} 0.125.
+A model with a total of 37 variables was able to classify lesion with an
+AUC range of 0.847 to 0.791 while the model to classify samples as
+before and after had 33 with an AUC range of 0.79 to 0.651 for 100 20
+repeated 10-fold cross-validated runs. Both models had a significant
+decrease in the positive probability of a lesion between individual's
+before versus after surgery samples (P-value = 1.91e-11 and 6.72e-12).
+In total there were 14 OTUs that were common to both models and were
+mostly commensals with largest representation from OTUs belonging to
+Bacteroides, Blautia, Streptococcus, and Clostridiales.\\
+\textbf{Conclusions:} Our data suggests that treatment not only
+significantly reduces the probability of having a colonic lesion but
+also causes detectable changes in the bacterial microbiome. Further
+surveillance of these individuals will enable us to determine whether
+models such as the one we present here can also be used to predict
+recurrence of colorectal cancer.
 
@@
 
-\subsubsection{Importance}\label{importance}
+\subsubsection{Keywords}\label{keywords}
 
-This is one of the first studies to investigate within humans what
-happens to the bacterial microbiome before and after adenoma and
-carcinoma treatment. Specifically, it aims to assess how a random forest
-machine learning algorithm built model respond to treatment and adjust
-it's positive probability calls of whether the individual has an adenoma
-or carcinoma due to surgical removal of the lesion.
+bacterial microbiome; colorectal cancer; polyps; FIT; detection; risk
+factors
 
@@
 
-\subsection{Introduction}\label{introduction}
+\subsection{Background}\label{background}
 
@@
 related deaths and is the second most common cancer death among men aged
-40-79 years of age (1, 2). Over the last few years death due to the
-disease has seen a significant decrease thanks mainly to improvements in
-screening (1). However, despite this giant improvement there are still
-approximately 50,000 deaths from the disease a year (2). It is estimated
-that around 5-10\% of all CRCs can be explained by autosomal dominant
-inheritance (3). The vast majority of CRCs are not inherited and the
-exact etiology to disease has not been well worked out (2). Although
-many risk factors have been identified (2) and non-invasive screening
-techniques have started to be put into consistent use (4, 5) there has
-been a consistent increase in the incidence of CRC in the younger
-population.
+40-79 years of age {[}1,2{]}. Over the last few years death due to the
+disease has seen a significant decrease, thanks mainly to improvements
+in screening {[}1{]}. However, despite this improvement there are still
+approximately 50,000 deaths from the disease a year {[}2{]}. It is
+estimated that around 5-10\% of all CRCs can be explained by autosomal
+dominant inheritance {[}3{]}. The vast majority of CRCs are not
+inherited and the exact etiology of the disease has not been well worked
+out {[}2{]}. Although many risk factors have been identified {[}2{]} and
+non-invasive screening techniques have started to be put into consistent
+use {[}4,5{]} there has been an consistent additional increase in the
+incidence of CRC in the younger population.
 
-This increased incidence of CRC in the younger population is concerncing
+This increased incidence of CRC in the younger population is concerning
 since having either an adenoma or carcinoma increases ones risk for
-future adenomas or carcinomas (6--8). This increased risk can also carry
-with it an increased risk of mortality due to this recurrence (9, 10).
-Therefore there has been a great amount of interest in early risk
-stratification tools (11, 12) that can help identify those they may be
-at risk of reccurence. Concurrently with this there has been a lot of
-interest in new areas such as the gut bacterial microbiome for insight
-into potential disease pathology.
+future adenomas or carcinomas {[}6--8{]}. This increased risk can also
+carry with it an increased chance of mortality due to this recurrence
+{[}9,10{]}. Therefore there has been a great amount of interest in early
+risk stratification tools {[}11,12{]} that can help identify those they
+may be at most susceptibility to reccurence. Concurrently, there has
+also been a lot of interest in new areas that could have a role in
+disease pathogenesis, such as the gut bacterial microbiome.
 
@@
 ability to be able to complement existing screening methods such as
-Fecal Immunoglobulin Test (FIT) or act alone as a screening tool (13,
-14). There has also been research into how this microbiome could be
-altered directly on tumor tissue itself (15). There have also been a few
-studies that have shown how this microbiome (16) or specific members
-within it (17) could be directly involved ith the pathogenesis of CRC.
+Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
+{[}13,14{]}. There has also been research into how this microbiome could
+be altered directly on tumor tissue itself {[}15{]}. A few studies have
+now even shown how this microbiome {[}16{]} or specific members within
+it {[}17{]} could be directly involved with the pathogenesis of CRC.
 These studies have helped to provide a tantilzing link between the
@@
 remains limited information on the bacterial microbiome before and after
-successful surgery for removal of the adenoma or carcinoma.
+successful surgery for removal of the adenoma or carcinoma and whether
+it changes at all.
 
@@
 anlaysis includes both alpha and beta diversity analysis along with
-investigating indvidual operational taxonomic units (OTUs). We next
-utilized a Random Forest model that was trained on a completely separate
-data set and observed how this model as well as specific OTUs within
-this model performed on the initial and follow up samples. We then
-created a new model to classify intial and follow up samples and
-investigated how the model performs and what specific OUTs within the
-model change. Finally, we observed how specific bacteria that have been
-previously implicated in CRC changed before and after treatment in both
-the adenoma and carcinoma individuals.
+investigation of individual operational taxonomic units (OTUs). We also
+utilized Random Forest models and observed how these models well as
+specific OTUs within this model performed on before (initial) and after
+(follow up) surgery samples. We used these models to look for similar
+important OTUs to identify the most important ones for not only
+classifying initial and follow up samples but also lesion or normal.
 
@@
 Treatment}} Based on thetayc distance metrics, comparing the initial to
-the follow up samples, there was no difference between the adenoma and
-carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There was a
-difference in FIT between initial and follow up samples with the
-carcinoma group having a significant decrease in FIT versus the adenoma
-group (P-value = 2.15e-05) {[}Figure 1b{]}. Although the thetayc
-distance metric change was similar between adenoma and carcinoma the
-directionality of the change was significant in the cancer group between
-initial and follow up (P-value = 0.002) but not for the adenoma group
-(P-value = 0.997). This change can be visualized using an NMDS {[}Figure
-1c \& 1d{]}. When all follow up samples were compared to each other
-there was no significant overall difference between them (P-value =
-0.085). There was no significant difference between initial and follow
-up samples for observed OTUs, Shannon diversity, or evenness after
-correction for multiple comparisons {[}Table S1{]}. Time of follow up
-sample from initial sampling, did not have a significant difference
-between adenoma and carcinoma (uncorrected P-value = 0.784).
+the follow up samples, there was a significant difference between the
+adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
+was also a significant difference in FIT between initial and follow up
+samples with the carcinoma group having a significant decrease in FIT
+versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
+community structure before and after surgery are visualized on NMDS
+graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
+{[}Figure 1d{]} (PERMANOVA = 0.997). When all initial and follow up
+samples were compared to each other there was no significant overall
+difference between them (PERMANOVA = 0.085). There was also no
+significant difference between initial and follow up samples for
+observed OTUs, Shannon diversity, or evenness after correction for
+multiple comparisons {[}Table S1{]}. There was no significant difference
+between initial and follow up samples for any single OTU {[}Figure
+S1{]}.
 
-\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
-training ranged from a minimum of 0.723 to a maximum of 0.795 with the
-middle of all 100 runs having an AUC of 0.761. Interestingly, the worst
-AUC model from training performed the best on it's respective 80/20
-split test data {[}Figure 3{]}. In fact the 80/20 test performance
-showed that the AUC for the middle model chosen was the most stable
-(best training model test set AUC = 0.646, middle training model test
-set AUC = 0.744, worse training model test set AUC = 0.904). That is to
-say it had the smallest change in AUC in comparison to the minimum and
-maximum AUC trained models. The middle model was close to the full
-training data AUC NA. There was no significant difference between the
-AUC of the best and middle training models (P-value = 1). There was also
-no difference in the middle model versus worse (P-value = 0.0431) or
-full data model (P-value = 1). The two comparisons with a significant
-difference were between the worse training model and the best training
-model (P-value = 6.83e-04) and the worse training model and full
-training model (P-value = 1.2e-03).
+\textbf{\emph{Differences in Adenoma and Carcinoma of Previously
+Associated Cancer Bacteria}} We next examined whether there were
+differences in previously well described carcinoma associated OTUs.
+These included the OTUs that aligned with Porphyromonas asaccharolytica
+(Otu000153), Fusobacterium nucleatum (Otu000226), Parvimonas micra
+(Otu000460), and Peptostreptococcus stomatis (Otu000653). First, the
+carcinoma samples showed a significant difference between initial and
+follow up samples for Peptostreptococcus stomatis (P-value = 0.0183) and
+Porphyromonas asaccharolytica (P-value = 0.0154) whereas there were no
+signficant differences in any of these OTUs in the adenoma samples
+{[}Table S4{]}. Second, when these OTUs were present, there was a clear
+magnitude difference in these specific OTUs based on whether they were
+from adenoma or carcinoma individuals {[}Figure 2{]}. However, only a
+small percentage of those with adenoma or carcinoma were positive for
+any of these OTUs.
 
-\textbf{\emph{Most Important Variables to the Model}} Overall, there
-were a total of 37 variables identified as being present in more than
-50\% of the training models {[}Figure S2{]}. The top 5 most important
-bacterial OTUs were Lachnospiraceae (Otu000013), Escherichia/Shigella
-(Otu000018), Ruminococcaceae (Otu000020), Ruminococcus (Otu000017), and
-Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
-of the total 100 different 80/20 runs.
+\textbf{\emph{Full and Reduced Model Results}} Since differences were
+observed between initial and follow up samples and only a small number
+of individuals were positive for previously associated CRC bacteria we
+next investigated if we could create a model that could adequately
+classify and adjust lesion probability based on the bacterial community
+and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
+initial follow up model which had and AUC range of 0.451 to 0.67 after
+100 20 repeated 10-fold cross validations. Interestingly, identification
+of the most important variables and reducing the variables considered to
+only these increased the AUC in the lesion model (0.791 - 0.847) and
+initial and follow up model (0.651 - 0.79).
 
-\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
-Decrease in Positive Probability Prediction}} A total of 1 sample was
-omitted from the original 67 test sample set since it was missing a
-complete set of follow up data. This left a total of 66 samples for test
-predictions. After multiple comparison correction there was a
-significant overall decrease in positive probability of a carcinoma and
-adenoma (P-value = 1.11e-11) {[}Figure 4{]}. This decrease was
-significant for both adenoma (P-value = 3.64e-11 {[}Figure 4a{]} and
-carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This decrease in
-probability of lesion also held specifically for those with screen
-relative neoplasias (SRN) (P-value = 7.63e-06). A total of 66 or 100\%
-of all samples were correctly predicted to have a lesion. Although there
-was a decrease in positive probability only 24 of the total 66
-individuals were classified as adenoma or carcinoma free on follow up
-(successful classification of 37.9\%). There was no significant
-difference between the predictions and actual diagnosis for the initial
-samples in the 67-sample cohort test set. However, the predictions were
-significantly discordent with the diagnosis for the follow up samples
-(P-value = 4.19e-10). Although there were discordent results the
-respective sensitivity for the initial group was 100\% and for follow up
-was 100\%, respectively.
+The test set AUC range for the full and reduced lesion model were
+similar to that reported for the training set AUC ranges and the ROC
+curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
+final lesion model used falls within the range of both the full and
+reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
+range for the initial and follow up performed much better then the
+training set AUCs. Both the full and reduced initial and follow up
+models overlapped with each other {[}Figure 3b{]} there was a marked
+decrease in the ROC curve for the final model used.
 
-There was 1 individual who still clearly had CRC on follow up as well as
-5 individuals whose status on follow up was unknown. Although the 1
-individual had a decrease in positive probability their follow up sample
-was still higher than the cutoff threshold of 0.5 (positive probability
-= 0.903). Interestingly, 1 individuals who were unknown on follow up
-still were over the threshold cutoff of 0.5 even though, like the 1
-individual with clear CRC on follow up, the probability of an adenoma or
-carcinoma decreased {[}Table S2{]}.
+\textbf{\emph{Most Important Variables to the Models}} The reduced
+models were built based on the most important variables to the
+respective full model. For the lesion model there were a total of 37
+variables {[}Figure S2{]} whereas for the initial and follow up model
+there were a total of 33 variables {[}Figure S3{]}. For both models FIT
+resulted in the largest decrease in MDA {[}Figure S2a \& S3a{]}.
 
-The follow up positive probabilities were not affected by either
-chemotherapy treatment (uncorrected P-value = 0.621) or radiation
-therapy (uncorrected P-value = 0.255). There was also no difference in
-the amount of change in the positive probability based on whether
-individuals received chemotherapy (unccorected P-value = 0.718) or
-radiation therapy (uncorrected P-value = 0.431).
+\textbf{\emph{Positive Probability Prediction after Surgical Removal of
+Adenoma or Carcinoma}} Regardless of model used there was a significant
+decrease in the positive probability of either the sample being lesion
+or an initial sample on follow up {[}Figure 4 \& S4{]} (full lesion
+P-value = 1.11e-11, reduced lesion P-value = 1.91e-11, initial and
+follow up P-value = 6.71e-12, and reduced initial and follow up P-value
+= 6.72e-12).
 
-\textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
-Follow Up Versus Initial Samples}} Overall, there were a total of 11
-OTUs that were common between the main lesion model and the model for
-classifying initial and follow up samples specifically {[}Table S3{]}. A
-total of 1 OTU was still significant after multiple comparison
-correction and it's lowest taxonomic identification was to Blautia. In
-general, Otu000012 (Blautia) was decreased from initial to follow up
-{[}Figure 5{]}. The relative abundance was not drastically different
-then the mean of the values observed in the control training set
-{[}Figure 5{]}.
+For the full and reduced lesion model there was a significant difference
+in the classification for the lesion model between predicted and actual
+(P-value = 4.19e-10 and 6.98e-10, respectively) but not for the initial
+follow up model (P-value = 1.00 and 1.00). However, the lesion model
+correctly kept the one individual who still had a carcinoma on follow up
+above the cut off point {[}Figure 4a \& S4a{]} for a positive call while
+the initial and follow up models did not {[}Figure 4b \& S4b{]}.
 
-\textbf{\emph{Differences in Adenoma and Carcinoma in Previously
-Associated Cancer Bacteria}} First, there was a clear magnitude
-difference in these specific OTUs based on whether they were from
-adenoma or carcinoma individuals {[}Figure 6{]}. The carcinoma samples
-showed a significant difference between initial and follow up samples
-for Peptostreptococcus stomatis (P-value = 0.0183) and Porphyromonas
-asaccharolytica (P-value = 0.0154) whereas there were no signficant
-differences in any of these OTUs in the adenoma samples {[}Table S4{]}.
+\textbf{\emph{Common OTUs to both Lesion and Initial and Follow Up
+Models}} There were a total of 14 OTUs that were common to both models.
+Of these OTUs the most common taxonomic identifications were to Blautia,
+Bacteroides, Streptococcus, and Clostridiales. The majority of these
+OTUs had taxonomic identification to bacteria typically thought of as
+commensal {[}Table S2{]}.
+
+\textbf{Treatment and Time Differences}
+
+There was no difference in the amount of change in positive probability
+for either the full or reduced lesion model for either chemotherapy
+(P-value = 0.821 and 0.821) or radiation therapy (P-value = 0.69 and
+0.981). Although the initial follow up model was similar there was a
+significant decrease in positive probability for those treated with
+chemotherapy (P-value = 7.04e-04 and 5.07e-03Time of follow up sample
+from initial sampling, did not have a significant difference between
+adenoma and carcinoma (uncorrected P-value = 0.784).
 
@@
 sample was a carcinoma or adenoma. The overall sensitivity for lesion
-detection in the intial samples was 100 and for follow ups was 100.
-Although there was a decrease in overall probability of an adenoma or
-carcinoma only 24 were below the 0.5 threshold out of the total 65
-individuals who were diagnosed as not having a carcinoma on follow up.
+detection in the intial samples was and for follow ups was . Although
+there was a decrease in overall probability of an adenoma or carcinoma
+only were below the 0.5 threshold out of the total individuals who were
+diagnosed as not having a carcinoma on follow up.
 
@@
 though was significant after multiple comparison correction and the
-lowest taxonomic identification of Otu000012 was to Blautia. Although
-there was a difference in the relative abundance at initial and follow
-up these values were not drastically different from the relative
-abundance values observed in the control individuals of the training set
-{[}Figure 5{]}. Although we were interested in what we could use to
-classify those with either adenoma or carcinoma versus normal. We found
-that the traditional bacteria associated with CRC were higher in
-magnitude in the carcinoma group and there were significant differences
-in some of these OTUs between the initial and follow up samples
-{[}Figure 5 and Tabl S5{]}. This research provides evidence that it is
-possible to use bacterial microbiome data to create a highily sensitive
-model, that is reactive to therapy, for detection of adenoma or
-carcinoma. It accomplishes this by using a unique sample set in which
-before and after surgery stool samples are available for assessment. By
-using these types of samples we are not only able to show sensitivity of
-lesion prediction but also able to show that this model is reactive.
-That is to say that after surgery for removal of the adenoma or
-carcinoma it decreases the positive probability to reflect a lower
-likelihood of the individual having an adenoma or carcinoma.
+lowest taxonomic identification of was to . Although there was a
+difference in the relative abundance at initial and follow up these
+values were not drastically different from the relative abundance values
+observed in the control individuals of the training set {[}Figure 5{]}.
+Although we were interested in what we could use to classify those with
+either adenoma or carcinoma versus normal. We found that the traditional
+bacteria associated with CRC were higher in magnitude in the carcinoma
+group and there were significant differences in some of these OTUs
+between the initial and follow up samples {[}Figure 5 and Tabl S5{]}.
+This research provides evidence that it is possible to use bacterial
+microbiome data to create a highily sensitive model, that is reactive to
+therapy, for detection of adenoma or carcinoma. It accomplishes this by
+using a unique sample set in which before and after surgery stool
+samples are available for assessment. By using these types of samples we
+are not only able to show sensitivity of lesion prediction but also able
+to show that this model is reactive. That is to say that after surgery
+for removal of the adenoma or carcinoma it decreases the positive
+probability to reflect a lower likelihood of the individual having an
+adenoma or carcinoma.
 
@@
 \textbf{\emph{Study Design and Patient Sampling}} The sampling and
-design of the study was similar to that reported in Baxter, et al (13).
-In brief, study exclusion involved those who had already undergone
-surgery, radiation, or chemotherapy, had colorectal cancer before a
-baseline stool sample could be obtained, had IBD, a known hereditary
-non-polyposis colorectal cancer, or Familial adenomatous polyposis.
-Samples used to build the model used for prediction were collected
-either prior to a colonoscopy or between 1 - 2 weeks after. The
-bacterial microbiome has been shown to nomralize within this time period
-(18). Kept apart from this training set were a total of 67 individuals
-that not only had a sample as described previoulsy but also a follow up
-sample between 188 - 546 days after surgery and treatment had been
-completed. This study was approved by the University of Michigan
-Institutional Review Board. All study participants provided informed
-consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+design of the study was similar to that reported in Baxter, et al
+{[}13{]}. In brief, study exclusion involved those who had already
+undergone surgery, radiation, or chemotherapy, had colorectal cancer
+before a baseline stool sample could be obtained, had IBD, a known
+hereditary non-polyposis colorectal cancer, or Familial adenomatous
+polyposis. Samples used to build the model used for prediction were
+collected either prior to a colonoscopy or between 1 - 2 weeks after.
+The bacterial microbiome has been shown to nomralize within this time
+period {[}18{]}. Kept apart from this training set were a total of 67
+individuals that not only had a sample as described previoulsy but also
+a follow up sample between 188 - 546 days after surgery and treatment
+had been completed. This study was approved by the University of
+Michigan Institutional Review Board. All study participants provided
+informed consent and the study itself conformed to the guidelines set
+out by the Helsinki Declaration.
 
@@
 FIT was analyzed as previously published using both OC FIT-CHEK and
-OC-Auto Micro 80 automated system (Polymedco Inc.) (19). 16S rRNA gene
-sequencing was completed as previously described by Kozich, et al. (20).
-In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
-Laboratories) and an epMotion 5075 automated pipetting system
-(Eppendorf). The V4 variable region was amplified and the resulting
-product was split between three sequencing runs with control, adenoma,
-and carcinoma evenly represented on each run. Each group was randomly
-assigned to avoid biases based on sample collection location.
+OC-Auto Micro 80 automated system (Polymedco Inc.) {[}19{]}. 16S rRNA
+gene sequencing was completed as previously described by Kozich, et al.
+{[}20{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
+kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
+system (Eppendorf). The V4 variable region was amplified and the
+resulting product was split between three sequencing runs with control,
+adenoma, and carcinoma evenly represented on each run. Each group was
+randomly assigned to avoid biases based on sample collection location.
 
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described (20). The general processing workflow
+has been previously described {[}20{]}. The general processing workflow
 using mothur is as follows: Paired-end reads were first merged into
@@
 
-\textbf{\emph{Lesion Model Creation}} The Random Forest (21) algorithm
-was used to create the model used for prediction of lesion (adenoma or
-carcinoma) for the 67 individuals with follow up samples. The model
-included data on FIT and the bacterial microbiome. Non-binary data was
-checked for near zero variance and auto correlation. Data columns that
-had near zero variance were removed. Columns that were correlated with
-each other over a Spearman correlation coefficient of 0.75 had one of
-the two columns removed. This pre-processing was performed with the R
-package caret (v6.0.73). Optimization of the mtry hyperparameter
-involved taking the samples and making 100 80/20 (train/test) splits in
-the data where control and lesion were equally represented in the 80 and
-20 split, respectively. This 80\% portion was then split again into an
-80/20 split, and run through 20 repeated 10-fold cross validations to
-optimzie the model's AUC (Area Under the Curve of the Receiver Operator
-Characteristic). This resulting model was then tested on the 20\% of the
-data that was originally held out from this overall process. Once the
-ideal mtry was found the entire 490 sample set was used to create the
-final Random Forest model on which testing on the 67-person cohort was
-completed. The default cutoff of 0.5 was used as the threshold to
-classify individuals as positive or negative for lesion. The
-hyperparameter, mtry, defines the number of variables to investigate at
-each split before a new division of the data is created.
+\textbf{\emph{Lesion Model Creation}} The Random Forest {[}21{]}
+algorithm was used to create the model used for prediction of lesion
+(adenoma or carcinoma) for the 67 individuals with follow up samples.
+The model included data on FIT and the bacterial microbiome. Non-binary
+data was checked for near zero variance and auto correlation. Data
+columns that had near zero variance were removed. Columns that were
+correlated with each other over a Spearman correlation coefficient of
+0.75 had one of the two columns removed. This pre-processing was
+performed with the R package caret (v6.0.73). Optimization of the mtry
+hyperparameter involved taking the samples and making 100 80/20
+(train/test) splits in the data where control and lesion were equally
+represented in the 80 and 20 split, respectively. This 80\% portion was
+then split again into an 80/20 split, and run through 20 repeated
+10-fold cross validations to optimzie the model's AUC (Area Under the
+Curve of the Receiver Operator Characteristic). This resulting model was
+then tested on the 20\% of the data that was originally held out from
+this overall process. Once the ideal mtry was found the entire 490
+sample set was used to create the final Random Forest model on which
+testing on the 67-person cohort was completed. The default cutoff of 0.5
+was used as the threshold to classify individuals as positive or
+negative for lesion. The hyperparameter, mtry, defines the number of
+variables to investigate at each split before a new division of the data
+is created.
 
@@
 used for all statisitical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA (22) in the vegan package
+community structure utilized PERMANOVA {[}22{]} in the vegan package
 (v2.4.1) while comparisons between ROC curves utilized the method by
-DeLong et al. (23) executed by the pROC (v1.8) package. Comparisons
+DeLong et al. {[}23{]} executed by the pROC (v1.8) package. Comparisons
 between probabilities as well as overall amount of OTU between initial
@@
 multiple comparison testing was needed a Benjamini-Hochberg (BH)
-correction was applied (24) and a corrected P-value of less than 0.05
-was considered significant. Unless otherwise stated the P-values
+correction was applied {[}24{]} and a corrected P-value of less than
+0.05 was considered significant. Unless otherwise stated the P-values
 reported are those of the BH corrected ones.
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
-
-The authors thank the Great Lakes-New England Early Detection Research
-Network for providing the fecal samples that were used in this study.
-This study was supported by funding from the National Institutes of
-Health to P. Schloss (R01GM099514, P30DK034933) and to the Early
-Detection Research Network (U01CA86400).
-
-\newpage
-
 \textbf{Figure 1: Changes and differences between the adenoma or
@@
 \textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Initial and Follow Up}* A) Positive probability adjustment
-of those with carcinoma from intial to follow up sample B) Positive
-probability adjustment of those with adenoma as well as those with SRN
-and the probability adjustment from initial to follow up sample. The
-dotted line represents the threshold used to make the decision of
-whether a sample was lesion positive or not.
-
-\textbf{Figure 5: Lesion Model OTU with a Significant Decrease in
-Relative Abundance that is also Predictive of Initial and Follow Up.}
-After multiple comparison correction 1 (Blautia) was the only one with a
-P-value \textless{} 0.05. The dotted line represents the average
-relative abundance in the control training group.
+Results for Lesion and Initial and Follow Up for Reduced Variable
+Models} A) Positive probability adjustment of those with carcinoma from
+intial to follow up sample B) Positive probability adjustment of those
+with adenoma as well as those with SRN and the probability adjustment
+from initial to follow up sample. The dotted line represents the
+threshold used to make the decision of whether a sample was lesion
+positive or not.
 
@@
 
-\textbf{Figure S2: Summary of Important Variables in both Lesion and
-Initial/Follow Up models} A) MDA of the most important variables in the
-lesion model. The black point represents the median and the different
-colors are the different runs up to 100. B) The total number of
-appearances of each variable in the 100 different lesion models. The
-cutoff of 50\% was used to assess importance.
+\textbf{Figure S2: Summary of Important Variables in the Lesion Model}
+A) MDA of the most important variables in the lesion model. The black
+point represents the median and the different colors are the different
+runs up to 100. B) The total number of appearances of each variable in
+the 100 different lesion models. The cutoff of 50\% was used to assess
+importance.
 
-\textbf{Figure S3: Thetayc Graphed Against Time of Follow up Sample from
+\textbf{Figure S3: Summary of Important Variables in Initial Follow Up
+Model} A) MDA of the most important variables in the lesion model. The
+black point represents the median and the different colors are the
+different runs up to 100. B) The total number of appearances of each
+variable in the 100 different lesion models. The cutoff of 50\% was used
+to assess importance.
+
+\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
+Results for Lesion and Initial and Follow Up for Full Variable Model} A)
+Positive probability adjustment of those with carcinoma from intial to
+follow up sample B) Positive probability adjustment of those with
+adenoma as well as those with SRN and the probability adjustment from
+initial to follow up sample. The dotted line represents the threshold
+used to make the decision of whether a sample was lesion positive or
+not.
+
+\textbf{Figure S5: Thetayc Graphed Against Time of Follow up Sample from
 Initial}
+
+\newpage
+
+\subsection{Declarations}\label{declarations}
+
+\subsubsection{Ethics approval and consent to
+participate}\label{ethics-approval-and-consent-to-participate}
+
+\subsubsection{Consent for publication}\label{consent-for-publication}
+
+\subsubsection{Availability of data and
+material}\label{availability-of-data-and-material}
+
+\subsubsection{Competing Interests}\label{competing-interests}
+
+All authors declare that they do not have any relevent competing
+interests to report.
+
+\subsubsection{Funding}\label{funding}
+
+This study was supported by funding from the National Institutes of
+Health to P. Schloss (R01GM099514, P30DK034933) and to the Early
+Detection Research Network (U01CA86400).
+
+\subsubsection{Authors' contributions}\label{authors-contributions}
+
+All authors were involved in the conception and design of the study. MAS
+analyzed the data. NTB processed samples and analyzed the data. All
+authors interpreted the data. MAS and PDS wrote the manuscript. All
+authors reviewed and revised the manuscript. All authors read and
+approved the final manuscript.
+
+\subsubsection{Acknowledgements}\label{acknowledgements}
+
+The authors thank the Great Lakes-New England Early Detection Research
+Network for providing the fecal samples that were used in this study.
 
@@
 \hypertarget{ref-jemal_cancer_2010}{}
-1. \textbf{Jemal A}, \textbf{Siegel R}, \textbf{Xu J}, \textbf{Ward E}.
-2010. Cancer statistics, 2010. CA: a cancer journal for clinicians
-\textbf{60}:277--300.
-doi:\href{https://doi.org/10.3322/caac.20073}{10.3322/caac.20073}.
+1. Jemal A, Siegel R, Xu J, Ward E. Cancer statistics, 2010. CA: a
+cancer journal for clinicians. 2010;60:277--300.
 
 \hypertarget{ref-haggar_colorectal_2009}{}
-2. \textbf{Haggar FA}, \textbf{Boushey RP}. 2009. Colorectal cancer
-epidemiology: Incidence, mortality, survival, and risk factors. Clinics
-in Colon and Rectal Surgery \textbf{22}:191--197.
-doi:\href{https://doi.org/10.1055/s-0029-1242458}{10.1055/s-0029-1242458}.
+2. Haggar FA, Boushey RP. Colorectal cancer epidemiology: Incidence,
+mortality, survival, and risk factors. Clinics in Colon and Rectal
+Surgery. 2009;22:191--7.
 
 \hypertarget{ref-green_very_2007}{}
-3. \textbf{Green RC}, \textbf{Green JS}, \textbf{Buehler SK},
-\textbf{Robb JD}, \textbf{Daftary D}, \textbf{Gallinger S},
-\textbf{McLaughlin JR}, \textbf{Parfrey PS}, \textbf{Younghusband HB}.
-2007. Very high incidence of familial colorectal cancer in Newfoundland:
-A comparison with Ontario and 13 other population-based studies.
-Familial Cancer \textbf{6}:53--62.
-doi:\href{https://doi.org/10.1007/s10689-006-9104-x}{10.1007/s10689-006-9104-x}.
+3. Green RC, Green JS, Buehler SK, Robb JD, Daftary D, Gallinger S, et
+al. Very high incidence of familial colorectal cancer in Newfoundland: A
+comparison with Ontario and 13 other population-based studies. Familial
+Cancer. 2007;6:53--62.
 
 \hypertarget{ref-liao_application_2013}{}
-4. \textbf{Liao C-S}, \textbf{Lin Y-M}, \textbf{Chang H-C}, \textbf{Chen
-Y-H}, \textbf{Chong L-W}, \textbf{Chen C-H}, \textbf{Lin Y-S},
-\textbf{Yang K-C}, \textbf{Shih C-H}. 2013. Application of quantitative
-estimates of fecal hemoglobin concentration for risk prediction of
-colorectal neoplasia. World Journal of Gastroenterology
-\textbf{19}:8366--8372.
-doi:\href{https://doi.org/10.3748/wjg.v19.i45.8366}{10.3748/wjg.v19.i45.8366}.
+4. Liao C-S, Lin Y-M, Chang H-C, Chen Y-H, Chong L-W, Chen C-H, et al.
+Application of quantitative estimates of fecal hemoglobin concentration
+for risk prediction of colorectal neoplasia. World Journal of
+Gastroenterology. 2013;19:8366--72.
 
 \hypertarget{ref-johnson_multi-target_2016}{}
-5. \textbf{Johnson DH}, \textbf{Kisiel JB}, \textbf{Burger KN},
-\textbf{Mahoney DW}, \textbf{Devens ME}, \textbf{Ahlquist DA},
-\textbf{Sweetser S}. 2016. Multi-target stool DNA test: Clinical
-performance and impact on yield and quality of colonoscopy for
-colorectal cancer screening. Gastrointestinal Endoscopy.
-doi:\href{https://doi.org/10.1016/j.gie.2016.11.012}{10.1016/j.gie.2016.11.012}.
+5. Johnson DH, Kisiel JB, Burger KN, Mahoney DW, Devens ME, Ahlquist DA,
+et al. Multi-target stool DNA test: Clinical performance and impact on
+yield and quality of colonoscopy for colorectal cancer screening.
+Gastrointestinal Endoscopy. 2016;
 
 \hypertarget{ref-laiyemo_short-_2013}{}
-6. \textbf{Laiyemo AO}, \textbf{Doubeni C}, \textbf{Brim H},
-\textbf{Ashktorab H}, \textbf{Schoen RE}, \textbf{Gupta S},
-\textbf{Charabaty A}, \textbf{Lanza E}, \textbf{Smoot DT}, \textbf{Platz
-E}, \textbf{Cross AJ}. 2013. Short- and long-term risk of colorectal
-adenoma recurrence among whites and blacks. Gastrointestinal Endoscopy
-\textbf{77}:447--454.
-doi:\href{https://doi.org/10.1016/j.gie.2012.11.027}{10.1016/j.gie.2012.11.027}.
+6. Laiyemo AO, Doubeni C, Brim H, Ashktorab H, Schoen RE, Gupta S, et
+al. Short- and long-term risk of colorectal adenoma recurrence among
+whites and blacks. Gastrointestinal Endoscopy. 2013;77:447--54.
 
 \hypertarget{ref-matsuda_five-year_2009}{}
-7. \textbf{Matsuda T}, \textbf{Fujii T}, \textbf{Sano Y}, \textbf{Kudo
-S-e}, \textbf{Oda Y}, \textbf{Igarashi M}, \textbf{Iishi H},
-\textbf{Murakami Y}, \textbf{Ishikawa H}, \textbf{Shimoda T},
-\textbf{Kaneko K}, \textbf{Yoshida S}. 2009. Five-year incidence of
-advanced neoplasia after initial colonoscopy in Japan: A multicenter
-retrospective cohort study. Japanese Journal of Clinical Oncology
-\textbf{39}:435--442.
-doi:\href{https://doi.org/10.1093/jjco/hyp047}{10.1093/jjco/hyp047}.
+7. Matsuda T, Fujii T, Sano Y, Kudo S-e, Oda Y, Igarashi M, et al.
+Five-year incidence of advanced neoplasia after initial colonoscopy in
+Japan: A multicenter retrospective cohort study. Japanese Journal of
+Clinical Oncology. 2009;39:435--42.
 
 \hypertarget{ref-ren_long-term_2016}{}
-8. \textbf{Ren J}, \textbf{Kirkness CS}, \textbf{Kim M}, \textbf{Asche
-CV}, \textbf{Puli S}. 2016. Long-term risk of colorectal cancer by
-gender after positive colonoscopy: Population-based cohort study.
-Current Medical Research and Opinion \textbf{32}:1367--1374.
-doi:\href{https://doi.org/10.1080/03007995.2016.1174840}{10.1080/03007995.2016.1174840}.
+8. Ren J, Kirkness CS, Kim M, Asche CV, Puli S. Long-term risk of
+colorectal cancer by gender after positive colonoscopy: Population-based
+cohort study. Current Medical Research and Opinion. 2016;32:1367--74.
 
 \hypertarget{ref-loberg_long-term_2014}{}
-9. \textbf{Løberg M}, \textbf{Kalager M}, \textbf{Holme Ø}, \textbf{Hoff
-G}, \textbf{Adami H-O}, \textbf{Bretthauer M}. 2014. Long-term
-colorectal-cancer mortality after adenoma removal. The New England
-Journal of Medicine \textbf{371}:799--807.
-doi:\href{https://doi.org/10.1056/NEJMoa1315870}{10.1056/NEJMoa1315870}.
+9. Løberg M, Kalager M, Holme Ø, Hoff G, Adami H-O, Bretthauer M.
+Long-term colorectal-cancer mortality after adenoma removal. The New
+England Journal of Medicine. 2014;371:799--807.
 
 \hypertarget{ref-freeman_natural_2013}{}
-10. \textbf{Freeman HJ}. 2013. Natural history and long-term outcomes of
-patients treated for early stage colorectal cancer. Canadian Journal of
-Gastroenterology = Journal Canadien De Gastroenterologie
-\textbf{27}:409--413.
+10. Freeman HJ. Natural history and long-term outcomes of patients
+treated for early stage colorectal cancer. Canadian Journal of
+Gastroenterology = Journal Canadien De Gastroenterologie.
+2013;27:409--13.
 
 \hypertarget{ref-lee_identification_2016}{}
-11. \textbf{Lee JH}, \textbf{Lee JL}, \textbf{Park IJ}, \textbf{Lim
-S-B}, \textbf{Yu CS}, \textbf{Kim JC}. 2016. Identification of
+11. Lee JH, Lee JL, Park IJ, Lim S-B, Yu CS, Kim JC. Identification of
 Recurrence-Predictive Indicators in Stage I Colorectal Cancer. World
-Journal of Surgery.
-doi:\href{https://doi.org/10.1007/s00268-016-3833-2}{10.1007/s00268-016-3833-2}.
+Journal of Surgery. 2016;
 
 \hypertarget{ref-richards_evidence-based_2016}{}
-12. \textbf{Richards CH}, \textbf{Ventham NT}, \textbf{Mansouri D},
-\textbf{Wilson M}, \textbf{Ramsay G}, \textbf{Mackay CD},
-\textbf{Parnaby CN}, \textbf{Smith D}, \textbf{On J}, \textbf{Speake D},
-\textbf{McFarlane G}, \textbf{Neo YN}, \textbf{Aitken E},
-\textbf{Forrest C}, \textbf{Knight K}, \textbf{McKay A}, \textbf{Nair
-H}, \textbf{Mulholland C}, \textbf{Robertson JH}, \textbf{Carey FA},
-\textbf{Steele R}, \textbf{Scottish Surgical Research Group}. 2016. An
-evidence-based treatment algorithm for colorectal polyp cancers: Results
-from the Scottish Screen-detected Polyp Cancer Study (SSPoCS). Gut.
-doi:\href{https://doi.org/10.1136/gutjnl-2016-312201}{10.1136/gutjnl-2016-312201}.
+12. Richards CH, Ventham NT, Mansouri D, Wilson M, Ramsay G, Mackay CD,
+et al. An evidence-based treatment algorithm for colorectal polyp
+cancers: Results from the Scottish Screen-detected Polyp Cancer Study
+(SSPoCS). Gut. 2016;
 
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+13. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. 2016;8:37.
 
 \hypertarget{ref-zeller_potential_2014}{}
-14. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
-\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
-\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
-\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
-\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
-\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
-J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
-\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
-\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
-for early-stage detection of colorectal cancer. Molecular Systems
-Biology \textbf{10}:766.
+14. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+Potential of fecal microbiota for early-stage detection of colorectal
+cancer. Molecular Systems Biology. 2014;10:766.
 
 \hypertarget{ref-dejea_microbiota_2014}{}
-15. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
-\textbf{White JR}, \textbf{Mark Welch JL}, \textbf{Rossetti BJ},
-\textbf{Peterson SN}, \textbf{Snesrud EC}, \textbf{Borisy GG},
-\textbf{Lazarev M}, \textbf{Stein E}, \textbf{Vadivelu J},
-\textbf{Roslani AC}, \textbf{Malik AA}, \textbf{Wanyiri JW}, \textbf{Goh
-KL}, \textbf{Thevambiga I}, \textbf{Fu K}, \textbf{Wan F}, \textbf{Llosa
-N}, \textbf{Housseau F}, \textbf{Romans K}, \textbf{Wu X},
-\textbf{McAllister FM}, \textbf{Wu S}, \textbf{Vogelstein B},
-\textbf{Kinzler KW}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014.
-Microbiota organization is a distinct feature of proximal colorectal
-cancers. Proceedings of the National Academy of Sciences of the United
-States of America \textbf{111}:18321--18326.
-doi:\href{https://doi.org/10.1073/pnas.1406199111}{10.1073/pnas.1406199111}.
+15. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+Rossetti BJ, et al. Microbiota organization is a distinct feature of
+proximal colorectal cancers. Proceedings of the National Academy of
+Sciences of the United States of America. 2014;111:18321--6.
 
 \hypertarget{ref-zackular_manipulation_2016}{}
-16. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
-\textbf{Schloss PD}. 2016. Manipulation of the Gut Microbiota Reveals
-Role in Colon Tumorigenesis. mSphere \textbf{1}.
-doi:\href{https://doi.org/10.1128/mSphere.00001-15}{10.1128/mSphere.00001-15}.
+16. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
 
 \hypertarget{ref-arthur_microbial_2014}{}
-17. \textbf{Arthur JC}, \textbf{Gharaibeh RZ}, \textbf{Mühlbauer M},
-\textbf{Perez-Chanona E}, \textbf{Uronis JM}, \textbf{McCafferty J},
-\textbf{Fodor AA}, \textbf{Jobin C}. 2014. Microbial genomic analysis
-reveals the essential role of inflammation in bacteria-induced
-colorectal cancer. Nature Communications \textbf{5}:4724.
-doi:\href{https://doi.org/10.1038/ncomms5724}{10.1038/ncomms5724}.
+17. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+McCafferty J, et al. Microbial genomic analysis reveals the essential
+role of inflammation in bacteria-induced colorectal cancer. Nature
+Communications. 2014;5:4724.
 
 \hypertarget{ref-obrien_impact_2013}{}
-18. \textbf{O'Brien CL}, \textbf{Allison GE}, \textbf{Grimpen F},
-\textbf{Pavli P}. 2013. Impact of colonoscopy bowel preparation on
-intestinal microbiota. PloS One \textbf{8}:e62815.
-doi:\href{https://doi.org/10.1371/journal.pone.0062815}{10.1371/journal.pone.0062815}.
+18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
 
 \hypertarget{ref-daly_evaluation_2013}{}
-19. \textbf{Daly JM}, \textbf{Bay CP}, \textbf{Levy BT}. 2013.
-Evaluation of fecal immunochemical tests for colorectal cancer
-screening. Journal of Primary Care \& Community Health
-\textbf{4}:245--250.
-doi:\href{https://doi.org/10.1177/2150131913487561}{10.1177/2150131913487561}.
+19. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
+for colorectal cancer screening. Journal of Primary Care \& Community
+Health. 2013;4:245--50.
 
 \hypertarget{ref-kozich_development_2013}{}
-20. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq Illumina sequencing platform.
-Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/AEM.01043-13}{10.1128/AEM.01043-13}.
+20. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq Illumina sequencing
+platform. Applied and Environmental Microbiology. 2013;79:5112--20.
 
 \hypertarget{ref-breiman_random_2001}{}
-21. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
-\textbf{45}:5--32.
-doi:\href{https://doi.org/10.1023/a:1010933404324}{10.1023/a:1010933404324}.
+21. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+{[}cited 2013 Feb 7{]};45:5--32. Available from:
+\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}
 
 \hypertarget{ref-anderson_permanova_2013}{}
-22. \textbf{Anderson MJ}, \textbf{Walsh DCI}. 2013. PERMANOVA, ANOSIM,
-and the Mantel test in the face of heterogeneous dispersions: What null
-hypothesis are you testing? Ecological Monographs \textbf{83}:557--574.
-doi:\href{https://doi.org/10.1890/12-2010.1}{10.1890/12-2010.1}.
+22. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+the face of heterogeneous dispersions: What null hypothesis are you
+testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
+5{]};83:557--74. Available from:
+\url{http://doi.wiley.com/10.1890/12-2010.1}
 
 \hypertarget{ref-delong_comparing_1988}{}
-23. \textbf{DeLong ER}, \textbf{DeLong DM}, \textbf{Clarke-Pearson DL}.
-1988. Comparing the areas under two or more correlated receiver
-operating characteristic curves: A nonparametric approach. Biometrics
-\textbf{44}:837--845.
+23. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
+two or more correlated receiver operating characteristic curves: A
+nonparametric approach. Biometrics. 1988;44:837--45.
 
 \hypertarget{ref-benjamini_controlling_1995}{}
-24. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
-false discovery rate: A practical and powerful approach to multiple
-testing. Journal of the Royal Statistical Society Series B
-(Methodological) \textbf{57}:289--300.
+24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+practical and powerful approach to multiple testing. Journal of the
+Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
 

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add updated pdf

@@
 **Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.
+There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
 A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.
@@
 **Treatment and Time Differences**
-
 There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).

### 2017-01-13 submission/supplemental_outline_20161024.Rmd

@@
 
-**Table S2: Summary of Probability Assignment of Follow Up Individuals**
+**Table S2: Summary of OTUs in both Lesion and Initial/Follow Up Models**
 
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
 
-follow_up_prob_summary <- read.csv('../results/tables/follow_up_probability_summary.csv', stringsAsFactors = F, header = T)
+common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
 
-st3 <- as.data.frame(filter(follow_up_prob_summary, sampleType == "initial")[, "Yes"]) %>% 
-  mutate(follow_yes = filter(follow_up_prob_summary, sampleType == "followup")[, "Yes"]) %>% 
-  mutate(disease_free = filter(follow_up_prob_summary, sampleType == "initial")[, "disease_free"]) %>% 
-  mutate(diagnosis = filter(follow_up_prob_summary, sampleType == "initial")[, "Dx_Bin"])
+st2 <- select(common_vars_summary, otu) %>% 
+  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
+  mutate(Pvalue = common_vars_summary$Pvalue) %>% 
+  mutate(BH_corr = common_vars_summary$BH_corr)
 
-st3$disease_free[st3$disease_free == "n"] <- "No"
-st3$disease_free[st3$disease_free == "unknown"] <- "Unknown"
-st3$disease_free[st3$disease_free == "y"] <- "Yes"
+colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "P-value", "BH Corrected P-value")
 
-colnames(st3) <- c("Initial Positive Probability", "Follow Up Positive Probability", "Disease Free on Follow Up", "Initial Diagnosis")
-
-kable(st3, format = "markdown", align = 'c', padding = 2)
+kable(st2, format = "markdown", align = 'c', padding = 2)
 
@@
 \newpage
-
-**Table S3: Summary of OTUs in both Lesion and Initial/Follow Up Models**
-
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
-
-pvalue_common_imp_vars <- read.csv('../results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)
-
-st4 <- select(pvalue_common_imp_vars, otu) %>% 
-  mutate(Tax = createTaxaLabeller(select(pvalue_common_imp_vars, -Pvalue, -BH_corr))) %>% 
-  mutate(Pvalue = pvalue_common_imp_vars$Pvalue) %>% 
-  mutate(BH_corr = pvalue_common_imp_vars$BH_corr)
-
-colnames(st4) <- c("OTU", "Lowest Taxonomic ID", "P-value", "BH Corrected P-value")
-
-kable(st4, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S4: Summary of P-values for Specific CRC OTUs between Initial and Follow Ups**
-
-```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
-
-pvalue_adn_crc_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
-
-st5 <- cbind(tax_id = c("Fusobacterium Nucleatum", "Parvimonas Micra", 
-                        "Peptostreptococcus Assacharolytica", "Porphyromonas Stomatis"), 
-             crc_pvalue = format(pvalue_adn_crc_summary$crc_pvalue, digits = 3), 
-             crc_bh = format(pvalue_adn_crc_summary$crc_BH, digits = 3), 
-             adn_pvalue = format(pvalue_adn_crc_summary$adn_pvalue, digits = 3), 
-             adn_bh = format(pvalue_adn_crc_summary$adn_BH, digits = 3))
-
-colnames(st5) <- c("Taxonomic ID", "CRC P-value", "CRC BH Corrected P-value", 
-                   "Adenoma P-value", "Adenoma BH Corrected P-value")
-
-kable(st5, format = "markdown", align = 'c', padding = 2)
-
-```

### 2017-01-15 submission/manuscript_outline_20161024.Rmd
updated summary added to discussion

@@
 There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-The whole community structure before and after surgery are visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
 When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
@@
 The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.
-Interestingly, identification of the most important variables and reducing the variables considered to only these increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
 
@@
 
-**Treatment and Time Differences**
+***Treatment and Time Differences***
 There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).
-Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
+Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
+Time elapsed between the collection of the follow up sample from the initial sample, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
 
@@
 ## Discussion
-In our training set we show that the overall community structure as measured by different alpha diversity metrics, shows very little change between controls and those with either adenoma or carcinoma [Table S1].
-With respect to our test set there was very little difference in magnitude of change in the thetayc distance metric between those with adenoma or carcinoma [Figure 1a].
-In contrast, FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].
-An NMDS showed that there was very little observable change between initial and follow up for the adenoma group but there was one for the carcinoma group [Figure 1c & 1d].
-This cursory information is suggestive that treatment of carcinoma, had the largest response.
+There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].
+Yet based on thetayc there was a significant difference between inital and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.
+The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].
+There was also a significant difference in the overall bacterial community structure for those in the carcinoma group [Figure 1d] before and after surgery but not for the adenoma group [Figure 1c].
+Investigation of the OTU relative abundance before and after surgery found no single OTU to be significantly different [Figure S1].
 
-We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).
-Our middle training model, based on AUC, from 100 80/20 (train/test) splits was similar to the full training data model.
-It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].
-Using the full training data model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].
-There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.
-The overall sensitivity for lesion detection in the intial samples was `r #format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r #format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.
-Although there was a decrease in overall probability of an adenoma or carcinoma only `r #followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r #sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.
+Interestingly, when previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.
+This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.
 
-We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S3].
-Many of the OTUs identified classified to normal flora bacterium [Table S3].
-Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.
-Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].
-Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.
-We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].
-This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.
-It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.
-By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.
-That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.
+We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or chaning due to surgery.
+We found that the OTUs that made up most important variables to the model overwhelmingly belonged to comensal bacteria.
+With only the lesion model having a single OTU from previously associated cancer bacteria (Porphyromonas asaccharolytica).
+Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
+Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and all belonged to regular residents of our gut community.
+
+There was no difference for the majority of models tested for differences in positive probablity based on whether chemotherapy or radiation was received.
+There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.
+These results would indicate that the findings described on specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.
 

### 2017-01-15 submission/manuscript_outline_20161024.Rmd
complete discussion additions

@@
 
-Interestingly, when previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.
+Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.
 This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.
@@
 
-This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool (**insert citation**).
-Based on previous work by Jobin, et al.
-(**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for this model.
-Similarily, Porphyromonas has also been implicated in colorectal cancer (**insert citation**).
-Interestingly, many of the other OTUs had taxonomic identification for resident gut microbes.
-This could suggest that changes to the resident microbiome are important to the initiation of adenoma or carcinoma formation (**insert citation**) and provide support for the hypothesis that an initial change in the bacterial microbiome could pave the way for more inflammatory species: whether by creation of a new niche for oral microbes (**insert citation**) or allowing for a bloom of existing pro-inflammatory residents (**insert citation**).
+This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.
+Based on previous work by Arthur, et al.
+[@arthur_intestinal_2012]  it may not be surprising to see E.coli as one of the most important OTUs and one that was common to both models.
+Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.
+This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.
+These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].
 
-Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.
-One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.
-This has been suggested to be the case in a previous study (**insert Baxter Study**).
-It is also possible that these specific bacteria play a major role in the progression to carcinoma but may not be as important in the initiation of an adenoma, which would be supported by our data [Figure 5].
-Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.
-Another potential reason why we did not identify the "usual suspects"" is that these bacteria may not change much between initial and follow up samples in those with an identified lesion.
-That is to say that the bacteria are consistently present even after removal of the lesion by surgery.
-Finally, it is likely that within our test set there was not enough indviduals in which detection  was made or relative abundance high enough for these bacteria to be significant using a paired wilcoxson test.
+Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
+One potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.
+An alternative reason would be that even if they were reduced in relative abundance they were still present.
+Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
+Finally, it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
+It is also possible that all three of these potential explanations could have played a role.
+Regardless, our observations would suggest that an individual's resident bacteria have a role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery.
+It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).
+In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].
 
-One limitation in this study is that we do not know whether individuals in our test set eventually had a subsequent CRC diagnosis.
+One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
 This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
-Another limitation is that we do not know if adding modern tests such as the stool DNA test (**insert citation**) could help improve our overall AUC.
+Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
 Another limitation is that this study drew heavily from those with caucasian ancestry.
@@
 
-By adding patient data such as age, BMI, etc.
-to the model and showing that it can successfully help to predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.
-Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
+Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.
+This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
+Another potential reason is that the model itself may be overfit since the total number of samples is not that large.
+However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.
+Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
 
 Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.
-The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures associated with these lesions.
-It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day diagnose CRC with a high level of accuracy.
+The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
+It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day stratify CRC disease risk with a high level of accuracy.
 

### 2017-01-15 submission/manuscript_outline_20161024.Rmd
update Analysis overview and reproducible methods sections

@@
 We also investigated whether a model could be created that could identify before and after surgery samples.
-The training set utilized the `r length(rownames(good_metaf))`-person cohort that was previously used for testing of the lesion modeld.
+The training set utilized the `r length(rownames(good_metaf))`-person cohort that was previously used for testing of the lesion model.
 The creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.
@@
 Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.
-All OTUs used in the lesion model were also analyzed using a paired wilcoxson test.
+All OTUs used in the lesion model were then also analyzed using a paired wilcoxson test.
+We then investigated the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.
+Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
 The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
@@
 This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.
-Finally, in order to investigate the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.
-Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
+Finally, a list of common OTUs were found for the two different models used.
 
@@
 A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
+Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005).
 

### 2017-01-15 submission/manuscript_outline_20161024.tex
add updated pdf

@@
 2.15e-05). There was no significant difference in any indivdiual OTU
-between samples before and after surgery (P-value \textgreater{} 0.125.
+between samples before and after surgery (P-value \textgreater{} 0.125).
 A model with a total of 37 variables was able to classify lesion with an
@@
 versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
-community structure before and after surgery are visualized on NMDS
+community structure before and after surgery were visualized on NMDS
 graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
@@
 100 20 repeated 10-fold cross validations. Interestingly, identification
-of the most important variables and reducing the variables considered to
-only these increased the AUC in the lesion model (0.791 - 0.847) and
+of the most important variables and reducing the models to only these
+factors increased the AUC in the lesion model (0.791 - 0.847) and
 initial and follow up model (0.651 - 0.79).
@@
 
-\textbf{Treatment and Time Differences}
-
-There was no difference in the amount of change in positive probability
-for either the full or reduced lesion model for either chemotherapy
-(P-value = 0.821 and 0.821) or radiation therapy (P-value = 0.69 and
-0.981). Although the initial follow up model was similar there was a
-significant decrease in positive probability for those treated with
-chemotherapy (P-value = 7.04e-04 and 5.07e-03Time of follow up sample
-from initial sampling, did not have a significant difference between
+\textbf{\emph{Treatment and Time Differences}} There was no difference
+in the amount of change in positive probability for either the full or
+reduced lesion model for either chemotherapy (P-value = 0.821 and 0.821)
+or radiation therapy (P-value = 0.69 and 0.981). Although the initial
+follow up model was similar there was a significant decrease in positive
+probability for those treated with chemotherapy (P-value = 7.04e-04 and
+5.07e-03). Time elapsed between the collection of the follow up sample
+from the initial sample, did not have a significant difference between
 adenoma and carcinoma (uncorrected P-value = 0.784).
@@
 
-In our training set we show that the overall community structure as
-measured by different alpha diversity metrics, shows very little change
-between controls and those with either adenoma or carcinoma {[}Table
-S1{]}. With respect to our test set there was very little difference in
-magnitude of change in the thetayc distance metric between those with
-adenoma or carcinoma {[}Figure 1a{]}. In contrast, FIT had a large
-change in the initial and follow up samples in the carcinoma group
-versus the adenoma {[}Figure 1b{]}. An NMDS showed that there was very
-little observable change between initial and follow up for the adenoma
-group but there was one for the carcinoma group {[}Figure 1c \& 1d{]}.
-This cursory information is suggestive that treatment of carcinoma, had
-the largest response.
+There was no difference in alpha diversity metrics between the initial
+and follow up samples {[}Table S1{]}. Yet based on thetayc there was a
+significant difference between inital and follow up samples {[}Figure
+1a{]} with carcinoma samples being more dissimilar to their initial
+samples then adenoma samples. The change in FIT was also different
+between initial and follow up for those with adenoma or carcinoma
+{[}Figure 1b{]}. There was also a significant difference in the overall
+bacterial community structure for those in the carcinoma group {[}Figure
+1d{]} before and after surgery but not for the adenoma group {[}Figure
+1c{]}. Investigation of the OTU relative abundance before and after
+surgery found no single OTU to be significantly different {[}Figure
+S1{]}.
 
-We next created a model that incorporated both patient metadata, FIT,
-and the bacterial microbiome to be able to predict lesions (adenoma or
-carcinoma). Our middle training model, based on AUC, from 100 80/20
-(train/test) splits was similar to the full training data model. It's
-10-fold cross validated AUC was similar to it's test set AUC which was
-not the case for both the best and worse training model {[}Figure 3{]}.
-Using the full training data model we predicted the probability of a
-lesion in the inital and follow up samples {[}Figure 4{]}. There was a
-signficant decrease in positive probability regardless of whether the
-sample was a carcinoma or adenoma. The overall sensitivity for lesion
-detection in the intial samples was and for follow ups was . Although
-there was a decrease in overall probability of an adenoma or carcinoma
-only were below the 0.5 threshold out of the total individuals who were
-diagnosed as not having a carcinoma on follow up.
+Interestingly, when only previously associated cancer bacteria were
+investigated only 2 out of the 4 had a significant decrease in relative
+abundance between initial and follow up for the carcinoma group and 0
+out of the 4 were significant for the adenoma group. This data suggests
+that what may change after surgery is not necessarily any one specific
+bacterium nor any addition of new or depressed bacterium but rather that
+a shift of the bacterial community as a whole occurs with the existing
+members of this microbiome.
 
-We then investigated which OTUs could potentially be more important in
-our model {[}Figure 5 \& Table S3{]}. Many of the OTUs identified
-classified to normal flora bacterium {[}Table S3{]}. Only a single OTU
-though was significant after multiple comparison correction and the
-lowest taxonomic identification of was to . Although there was a
-difference in the relative abundance at initial and follow up these
-values were not drastically different from the relative abundance values
-observed in the control individuals of the training set {[}Figure 5{]}.
-Although we were interested in what we could use to classify those with
-either adenoma or carcinoma versus normal. We found that the traditional
-bacteria associated with CRC were higher in magnitude in the carcinoma
-group and there were significant differences in some of these OTUs
-between the initial and follow up samples {[}Figure 5 and Tabl S5{]}.
-This research provides evidence that it is possible to use bacterial
-microbiome data to create a highily sensitive model, that is reactive to
-therapy, for detection of adenoma or carcinoma. It accomplishes this by
-using a unique sample set in which before and after surgery stool
-samples are available for assessment. By using these types of samples we
-are not only able to show sensitivity of lesion prediction but also able
-to show that this model is reactive. That is to say that after surgery
-for removal of the adenoma or carcinoma it decreases the positive
-probability to reflect a lower likelihood of the individual having an
-adenoma or carcinoma.
+We next created a model that incorporated FIT and the bacterial
+microbiome to either be able to classify lesions (adenoma or carcinoma)
+or initial and follow up samples in order to confirm that the community
+was what was responding or chaning due to surgery. We found that the
+OTUs that made up most important variables to the model overwhelmingly
+belonged to comensal bacteria. With only the lesion model having a
+single OTU from previously associated cancer bacteria (Porphyromonas
+asaccharolytica). Using only these important OTUs and FIT both models
+(lesion and initial and follow up) had a significantly decreased
+positive probability of either lesion or being an initial sample on
+follow up {[}Figure 4 \& S4{]}. Confirming the importance of the changes
+of comensal bacteria to these classifications was that a total of 14
+OTUs were common to both models and all belonged to regular residents of
+our gut community.
+
+There was no difference for the majority of models tested for
+differences in positive probablity based on whether chemotherapy or
+radiation was received. There was also no difference in the thetayc
+distance metric based on length of time between initial and follow up
+sample between adenoma and carcinoma. These results would indicate that
+the findings described on specific to the surgical intervention and that
+differences observed between carcinoma and adenoma samples can not be
+simply attributed to collection time between samples.
 
 This study builds upon previous work from numerous labs that have looked
-into the bacterial microbiome as a potential screening tool
-(\textbf{insert citation}). Based on previous work by Jobin, et al.
-(\textbf{insert citation}) it may not be surprising to see E.coli in the
-top 5 OTUs for this model. Similarily, Porphyromonas has also been
-implicated in colorectal cancer (\textbf{insert citation}).
-Interestingly, many of the other OTUs had taxonomic identification for
-resident gut microbes. This could suggest that changes to the resident
-microbiome are important to the initiation of adenoma or carcinoma
-formation (\textbf{insert citation}) and provide support for the
-hypothesis that an initial change in the bacterial microbiome could pave
-the way for more inflammatory species: whether by creation of a new
-niche for oral microbes (\textbf{insert citation}) or allowing for a
-bloom of existing pro-inflammatory residents (\textbf{insert citation}).
+into the bacterial microbiome as a potential screening tool {[}13,14{]}
+by exploring what happens to the bacterial community after surgical
+removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
+it may not be surprising to see E.coli as one of the most important OTUs
+and one that was common to both models. Interestingly, many of the most
+important OTUs had taxonomic identification for resident gut microbes.
+This could suggest that the bacterial community is one of the first
+components that could change during the pathogenesis of disease. These
+bacterial microbiome changes could be the first step in allowing more
+inflammatory bacterium to gain a foothold within the colon {[}19{]}.
 
-Naturally, it is curious that normal staples of many screening studies
-such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
-present in the majority of the training models. One potential
-explanation for this is that FIT provides the same information to the
-model as these three organisms and so the model uses FIT preferentially
-over them. This has been suggested to be the case in a previous study
-(\textbf{insert Baxter Study}). It is also possible that these specific
-bacteria play a major role in the progression to carcinoma but may not
-be as important in the initiation of an adenoma, which would be
-supported by our data {[}Figure 5{]}. Regardless, our study does not
-argue against the importance of these bacterium in CRC initiation or
-pathogenesis but rather that the model does not utilize these specific
-bacteria for prediction purposes. Another potential reason why we did
-not identify the ``usual suspects''" is that these bacteria may not
-change much between initial and follow up samples in those with an
-identified lesion. That is to say that the bacteria are consistently
-present even after removal of the lesion by surgery. Finally, it is
-likely that within our test set there was not enough indviduals in which
-detection was made or relative abundance high enough for these bacteria
-to be significant using a paired wilcoxson test.
+Curiously, we observed that the typical CRC associated bacteria were not
+predictive within our models. One potential explanation is that even
+with surgery and a shift of the bacterial community these specific
+bacteria still persisted within the colon. An alternative reason would
+be that even if they were reduced in relative abundance they were still
+present. Another potential explanation is that they were not present in
+enough individuals to be able to classify those with and without disease
+with a high degree of accuracy. Finally, it is possible that our Random
+Forest models were able to get the same information from measures such
+as FIT or other OTUs. It is also possible that all three of these
+potential explanations could have played a role. Regardless, our
+observations would suggest that an individual's resident bacteria have a
+role to play in disease initiation and could change in a way that allows
+predictive models to lower the positive probability of a lesion after
+surgery. It should be stated that our study does not argue against the
+importance of these CRC associated bacteria in the pathogenesis of
+disease but rather that the models do not utilize these specific
+bacteria for classification purposes (lesion or initial and follow up).
+In fact, it is possible that these CRC associated bacteria are important
+in the transition from adenoma to carcinoma and would be one explanation
+as to why in our data we not only see high initial relative abundances,
+in certain individuals, but also large decreases in relative abundance
+in those with carcinoma but not in those with adenoma after surgery
+{[}Figure 2{]}.
 
-One limitation in this study is that we do not know whether individuals
-in our test set eventually had a subsequent CRC diagnosis. This
-information would help to strengthen the case for our Random Forest
-based model keeping a number of individuals above the cutoff threshold
-even though at follow up they were diagnosed as no longer having a
-lesion. Another limitation is that we do not know if adding modern tests
-such as the stool DNA test (\textbf{insert citation}) could help improve
-our overall AUC. Another limitation is that this study drew heavily from
-those with caucasian ancestry. The results may not be immediately
-representative of those with either Asian or African ancestry. Finally,
-although our training and test set are relatively large we still run the
-risk of overfitting or having a model that may not be immediately
-extrapolateable to other populations. We've done our best to safeguard
-against this by not only running 10-fold cross validation but also
-having over 100 different 80/20 splits to try and mimic the type of
-variation that might be expected to occur.
+One limitation of our study is that we do not know whether individuals
+who were still classified as positive by the lesion model eventually had
+a subsequent CRC diagnosis. This information would help to strengthen
+the case for our Random Forest based model keeping a number of
+individuals above the cutoff threshold even though at follow up they
+were diagnosed as no longer having a lesion. Another limitation is that
+we do not know if adding modern tests such as the stool DNA test
+{[}20{]} could help improve our overall AUC. Another limitation is that
+this study drew heavily from those with caucasian ancestry. The results
+may not be immediately representative of those with either Asian or
+African ancestry. Finally, although our training and test set are
+relatively large we still run the risk of overfitting or having a model
+that may not be immediately extrapolateable to other populations. We've
+done our best to safeguard against this by not only running 10-fold
+cross validation but also having over 100 different 80/20 splits to try
+and mimic the type of variation that might be expected to occur.
 
-By adding patient data such as age, BMI, etc. to the model and showing
-that it can successfully help to predict both carcinoma and adenoma our
-study provides further data that these patient factors in conjunction
-with the bacterial microbiome could potentially influence CRC and
-perhaps have a role in formation of adenomas. Further studies need to be
-carried out to verify our findings since not only are we dealing with
-stool, which could be very different than the communities present on the
-actual tissue, but also are dealing with correlations that may not be
-representative of the true pathogensis of disease.
+Within figure 3 the initial and follow up model showed better test AUC
+results then the training set AUC. This may have occured because the
+training AUC that was determined from 20 repeated 10 fold cross
+validation removed samples at random and did not take into account that
+they were matched samples. Another potential reason is that the model
+itself may be overfit since the total number of samples is not that
+large. However, the lesion model did not suffer from these discrepencies
+and similar conclusions can be drawn solely from this model. Regardless,
+further independent studies will need to be carried out to verify our
+findings since not only are we dealing with stool, which could be very
+different than the communities present on the actual tissue, but also
+are dealing with correlations that may not be representative of the true
+pathogensis of disease.
 
@@
 after either adenoma or carcinoma removal provides evidence that there
-are specific signatures associated with these lesions. It also shows
-that these algorithms can not only successfully react to successful
-treatment regimens but also may be able to one day diagnose CRC with a
-high level of accuracy.
+are specific signatures, mostly attributable to commensal organisms,
+associated with these lesions. It also shows that these algorithms can
+not only successfully react to successful treatment regimens but also
+may be able to one day stratify CRC disease risk with a high level of
+accuracy.
 
@@
 The bacterial microbiome has been shown to nomralize within this time
-period {[}18{]}. Kept apart from this training set were a total of 67
+period {[}21{]}. Kept apart from this training set were a total of 67
 individuals that not only had a sample as described previoulsy but also
@@
 FIT was analyzed as previously published using both OC FIT-CHEK and
-OC-Auto Micro 80 automated system (Polymedco Inc.) {[}19{]}. 16S rRNA
+OC-Auto Micro 80 automated system (Polymedco Inc.) {[}22{]}. 16S rRNA
 gene sequencing was completed as previously described by Kozich, et al.
-{[}20{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
+{[}23{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
 kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described {[}20{]}. The general processing workflow
+has been previously described {[}23{]}. The general processing workflow
 using mothur is as follows: Paired-end reads were first merged into
@@
 
-\textbf{\emph{Lesion Model Creation}} The Random Forest {[}21{]}
+\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
 algorithm was used to create the model used for prediction of lesion
@@
 surgery samples. The training set utilized the 67-person cohort that was
-previously used for testing of the lesion modeld. The creation of this
+previously used for testing of the lesion model. The creation of this
 model and optimization of the mtry hyperparameter was completed using
@@
 used for all statisitical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}22{]} in the vegan package
+community structure utilized PERMANOVA {[}25{]} in the vegan package
 (v2.4.1) while comparisons between ROC curves utilized the method by
-DeLong et al. {[}23{]} executed by the pROC (v1.8) package. Comparisons
+DeLong et al. {[}26{]} executed by the pROC (v1.8) package. Comparisons
 between probabilities as well as overall amount of OTU between initial
@@
 multiple comparison testing was needed a Benjamini-Hochberg (BH)
-correction was applied {[}24{]} and a corrected P-value of less than
+correction was applied {[}27{]} and a corrected P-value of less than
 0.05 was considered significant. Unless otherwise stated the P-values
@@
 initial and follow up samples were analyzed for differences in alpha and
-beta diversity. All OTUs used in the lesion model were also analyzed
-using a paired wilcoxson test. The lesion model was then tested for
-accuracy in prediction and whether it reduced the positive probability
-of lesion after surgery. The most important OTUs for this were used to
-build an updated model and this reduced feature model was assessed for
-it's similarity to the original model. We then used the initial follow
-up model to assess whether this model could classify samples better then
-the lesion model. The most important OTUs were then identified from this
-model and used to create a reduced feature initial follow up model. This
-reduced feature model, as was done with the lesion model, was compared
-to the full model for loss of accuracy. Finally, in order to investigate
-the relative abundance of specific bacteria, that have been previously
+beta diversity. All OTUs used in the lesion model were then also
+analyzed using a paired wilcoxson test. We then investigated the
+relative abundance of specific bacteria, that have been previously
 associated with CRC, we selected OTUs that taxonomically classified to
@@
 test if there were any differences based on whether the individual had
-an adenoma or carcinoma.
+an adenoma or carcinoma. The lesion model was then tested for accuracy
+in prediction and whether it reduced the positive probability of lesion
+after surgery. The most important OTUs for this were used to build an
+updated model and this reduced feature model was assessed for it's
+similarity to the original model. We then used the initial follow up
+model to assess whether this model could classify samples better then
+the lesion model. The most important OTUs were then identified from this
+model and used to create a reduced feature initial follow up model. This
+reduced feature model, as was done with the lesion model, was compared
+to the full model for loss of accuracy. Finally, a list of common OTUs
+were found for the two different models used.
 
@@
 description of how the data were processed and analyzed can be found at
-\url{https://github.com/SchlossLab/Sze_followUps_2017}.
+\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
+have been deposited into the NCBI Sequence Read Archive (SRP062005).
 
@@
 
+\hypertarget{ref-arthur_intestinal_2012}{}
+18. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+T-J, et al. Intestinal inflammation targets cancer-inducing activity of
+the microbiota. Science (New York, N.Y.). 2012;338:120--3.
+
+\hypertarget{ref-flynn_metabolic_2016}{}
+19. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
+
+\hypertarget{ref-cotter_long-term_2016}{}
+20. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
+et al. Long-Term Follow-up of Patients Having False Positive
+Multi-target Stool DNA Tests after Negative Screening Colonoscopy: The
+LONG-HAUL Cohort Study. Cancer Epidemiology, Biomarkers \& Prevention: A
+Publication of the American Association for Cancer Research, Cosponsored
+by the American Society of Preventive Oncology. 2016;
+
 \hypertarget{ref-obrien_impact_2013}{}
-18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+21. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-daly_evaluation_2013}{}
-19. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
+22. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
 for colorectal cancer screening. Journal of Primary Care \& Community
@@
 \hypertarget{ref-kozich_development_2013}{}
-20. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-21. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+24. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-22. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+25. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-delong_comparing_1988}{}
-23. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
+26. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
 two or more correlated receiver operating characteristic curves: A
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+27. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
correct centering problem

@@
   pdf_document:
+    keep_tex: true
     includes:
       in_header: header.tex
-    keep_tex: yes
-  html_document: default
 csl: microbiome.csl
@@
 # Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment
-
 ```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
@@
 ```
-\begin{center}  
 \vspace{25mm}
 
+\begin{center}
 Running Title: Human Microbiome and Colorectal Cancer
@@
 
-Marc A Sze^1^, Nielson T Baxter^2^, Mack T Ruffin IV^3^, Mary AM Rogers^4^, and Patrick D Schloss^1^${^\dagger}$
+Marc A Sze${^1}$, Nielson T Baxter${^2}$, Mack T Ruffin IV${^3}$, Mary AM Rogers${^2}$, and Patrick D Schloss${^1}$${^\dagger}$
 
@@
 
-^1^Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI  
-^2^Department of Internal Medicine, Division of Infectious Disease, University of Michigan, Ann Arbor, MI  
-^3^Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA  
-^4^Department of Internal Medicine, University of Michigan, Ann Arbor, MI
+$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+
+$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI
+
+$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+
 \end{center}

### 2017-01-16 submission/manuscript_outline_20161024.tex
correct centering problem

@@
 
-\textbackslash{}begin\{center\}\\
 \vspace{25mm}
 
+\begin{center}
 Running Title: Human Microbiome and Colorectal Cancer
@@
 
-Marc A Sze\textsuperscript{1}, Nielson T Baxter\textsuperscript{2}, Mack
-T Ruffin IV\textsuperscript{3}, Mary AM Rogers\textsuperscript{4}, and
-Patrick D Schloss\textsuperscript{1}\({^\dagger}\)
+Marc A Sze${^1}$, Nielson T Baxter${^2}$, Mack T Ruffin IV${^3}$, Mary AM Rogers${^2}$, and Patrick D Schloss${^1}$${^\dagger}$
 
@@
 
-\(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
 
-\textsuperscript{1}Department of Microbiology and Immunology, University
-of Michigan, Ann Arbor, MI\\
-\textsuperscript{2}Department of Internal Medicine, Division of
-Infectious Disease, University of Michigan, Ann Arbor, MI\\
-\textsuperscript{3}Department of Family Medicine and Community Medicine,
-Penn State Hershey Medical Center, Hershey, PA\\
-\textsuperscript{4}Department of Internal Medicine, University of
-Michigan, Ann Arbor, MI \textbackslash{}end\{center\}
+$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+
+$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI
+
+$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+
+\end{center}
 

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
abstract cleaning

@@
 
-# Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment
+# Differences in the fecal Microbiome Before and After Colorectal Cancer Treatment
 ```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
@@
 **Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.
-Previous research suggests a link between stool bacterial microbiome and CRC.
+Previous research suggests a link between fecal bacterial microbiome and CRC.
 The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e.
@@
 We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.
-**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT  (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.
-Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+Both models had a significant decrease in the positive probability of a lesion or being a before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and all of these classified to commensals with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
 **Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.
@@
 This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
-Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
+Another limitation is that we do not know if adding modern tests such as the fecal DNA test [@cotter_long-term_2016] could help improve our overall AUC.
 Another limitation is that this study drew heavily from those with caucasian ancestry.
@@
 However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.
-Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
+Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with fecal, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
 
@@
 The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].
-In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
+In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
 Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.

### 2017-01-16 submission/manuscript_outline_20161024.tex
abstract cleaning

@@
 
-\section{Differences in the Stool Microbiome Before and After Colorectal
+\section{Differences in the fecal Microbiome Before and After Colorectal
 Cancer
-Treatment}\label{differences-in-the-stool-microbiome-before-and-after-colorectal-cancer-treatment}
+Treatment}\label{differences-in-the-fecal-microbiome-before-and-after-colorectal-cancer-treatment}
 
@@
 mitigating deaths due to the disease. Previous research suggests a link
-between stool bacterial microbiome and CRC. The overall objective was to
+between fecal bacterial microbiome and CRC. The overall objective was to
 investigate the changes in the bacterial microbiome after surgery in
@@
 similar to their pre-surgery sample then those with carcinoma (P-value =
-0.00198) and this was also reflected in FIT as well (P-value =
-2.15e-05). There was no significant difference in any indivdiual OTU
-between samples before and after surgery (P-value \textgreater{} 0.125).
-A model with a total of 37 variables was able to classify lesion with an
-AUC range of 0.847 to 0.791 while the model to classify samples as
-before and after had 33 with an AUC range of 0.79 to 0.651 for 100 20
-repeated 10-fold cross-validated runs. Both models had a significant
-decrease in the positive probability of a lesion between individual's
-before versus after surgery samples (P-value = 1.91e-11 and 6.72e-12).
-In total there were 14 OTUs that were common to both models and were
-mostly commensals with largest representation from OTUs belonging to
-Bacteroides, Blautia, Streptococcus, and Clostridiales.\\
+0.00198) and this was also reflected in FIT (P-value = 2.15e-05). There
+was no significant difference in any indivdiual OTU between samples
+before and after surgery (P-value \textgreater{} 0.125). A model with a
+total of 37 variables was able to classify lesion (AUC = 0.847 - 0.791
+while the model to classify samples as before and after surgery had 33
+variables (AUC = 0.79 - 0.651). Both models had a significant decrease
+in the positive probability of a lesion or being a before surgery sample
+(P-value = 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were
+common to both models and all of these classified to commensals with the
+largest representation belonging to Bacteroides, Blautia, Streptococcus,
+and Clostridiales.\\
 \textbf{Conclusions:} Our data suggests that treatment not only
@@
 were diagnosed as no longer having a lesion. Another limitation is that
-we do not know if adding modern tests such as the stool DNA test
+we do not know if adding modern tests such as the fecal DNA test
 {[}20{]} could help improve our overall AUC. Another limitation is that
@@
 further independent studies will need to be carried out to verify our
-findings since not only are we dealing with stool, which could be very
+findings since not only are we dealing with fecal, which could be very
 different than the communities present on the actual tissue, but also
@@
 undergone surgery, radiation, or chemotherapy, had colorectal cancer
-before a baseline stool sample could be obtained, had IBD, a known
+before a baseline fecal sample could be obtained, had IBD, a known
 hereditary non-polyposis colorectal cancer, or Familial adenomatous

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline intro

@@
 The vast majority of CRCs are not inherited and the exact etiology of the disease has not been well worked out [@haggar_colorectal_2009].
-Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been an consistent additional increase in the incidence of CRC in the younger population.
+Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been an additional increase in the incidence of CRC in the younger population.
 
@@
 This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].
-Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at most susceptibility to reccurence.
+Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to reccurence.
 Concurrently, there has also been a lot of interest in new areas that could have a role in disease pathogenesis, such as the gut bacterial microbiome.
@@
 
-In this study we investigated what happened to the bacterial microbiome before and after surgery for both adenoma and carcinoma individuals.
+In this study we investigated what happened to the bacterial microbiome before and after surgery for indivdiuals with either adenoma and carcinoma.
 Our anlaysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest models and observed how these models well as specific OTUs within this model performed on before (initial) and after (follow up) surgery samples.
-We used these models to look for similar important OTUs to identify the most important ones for not only classifying initial and follow up samples but also lesion or normal.
+We also utilized Random Forest models and observed how these models as well as specific OTUs within this model performed on before (initial) and after (follow up) surgery samples.
+We used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
 

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline intro

@@
 non-invasive screening techniques have started to be put into consistent
-use {[}4,5{]} there has been an consistent additional increase in the
-incidence of CRC in the younger population.
+use {[}4,5{]} there has been an additional increase in the incidence of
+CRC in the younger population.
 
@@
 carry with it an increased chance of mortality due to this recurrence
-{[}9,10{]}. Therefore there has been a great amount of interest in early
-risk stratification tools {[}11,12{]} that can help identify those they
-may be at most susceptibility to reccurence. Concurrently, there has
-also been a lot of interest in new areas that could have a role in
+{[}9,10{]}. Therefore, there has been a great amount of interest in
+early risk stratification tools {[}11,12{]} that can help identify those
+that may be at most susceptibility to reccurence. Concurrently, there
+has also been a lot of interest in new areas that could have a role in
 disease pathogenesis, such as the gut bacterial microbiome.
@@
 In this study we investigated what happened to the bacterial microbiome
-before and after surgery for both adenoma and carcinoma individuals. Our
-anlaysis includes both alpha and beta diversity analysis along with
-investigation of individual operational taxonomic units (OTUs). We also
-utilized Random Forest models and observed how these models well as
-specific OTUs within this model performed on before (initial) and after
-(follow up) surgery samples. We used these models to look for similar
-important OTUs to identify the most important ones for not only
-classifying initial and follow up samples but also lesion or normal.
+before and after surgery for indivdiuals with either adenoma and
+carcinoma. Our anlaysis includes both alpha and beta diversity analysis
+along with investigation of individual operational taxonomic units
+(OTUs). We also utilized Random Forest models and observed how these
+models as well as specific OTUs within this model performed on before
+(initial) and after (follow up) surgery samples. We used these models to
+look for similar important OTUs to identify the crucial OTUs for not
+only classifying initial and follow up samples but also lesion or
+normal.
 

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline results

@@
 Both models had a significant decrease in the positive probability of a lesion or being a before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and all of these classified to commensals with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
 **Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.
@@
 
-***Bacterial Community and Fit Changes before and after Treatment***
+***Bacterial Community and FIT Changes before and after Surgery***
 Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
-There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+There was also a significant difference in change in FIT between initial and follow up samples (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
+The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
 When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
-There was no significant difference between initial and follow up samples for any single OTU [Figure S1].
+There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
+There was also no significant difference between initial and follow up samples for any single OTU [Figure S1].
 
@@
 We next examined whether there were differences in previously well described carcinoma associated OTUs.
-These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
-Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].
+These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
+First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there was no signficant differences in any of these OTUs in the adenoma samples [Table S4].
+Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].
 However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.
@@
 ***Full and Reduced Model Results***
-Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.
-The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.
+Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria; we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.
+The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.
 Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
@@
 The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].
-The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].
+The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].
 Interestingly, the test set AUC range for the initial and follow up performed much better then the training set AUCs.

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline results

@@
 (P-value = 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were
-common to both models and all of these classified to commensals with the
-largest representation belonging to Bacteroides, Blautia, Streptococcus,
-and Clostridiales.\\
+common to both models and the majority of these classified to commensal
+bacteria with the largest representation belonging to Bacteroides,
+Blautia, Streptococcus, and Clostridiales.\\
 \textbf{Conclusions:} Our data suggests that treatment not only
@@
 
-\textbf{\emph{Bacterial Community and Fit Changes before and after
-Treatment}} Based on thetayc distance metrics, comparing the initial to
+\textbf{\emph{Bacterial Community and FIT Changes before and after
+Surgery}} Based on thetayc distance metrics, comparing the initial to
 the follow up samples, there was a significant difference between the
 adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
-was also a significant difference in FIT between initial and follow up
-samples with the carcinoma group having a significant decrease in FIT
-versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
+was also a significant difference in change in FIT between initial and
+follow up samples (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
 community structure before and after surgery were visualized on NMDS
-graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
-{[}Figure 1d{]} (PERMANOVA = 0.997). When all initial and follow up
+graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.997) and carcinoma
+{[}Figure 1d{]} (PERMANOVA = 0.002). When all initial and follow up
 samples were compared to each other there was no significant overall
-difference between them (PERMANOVA = 0.085). There was also no
-significant difference between initial and follow up samples for
-observed OTUs, Shannon diversity, or evenness after correction for
-multiple comparisons {[}Table S1{]}. There was no significant difference
-between initial and follow up samples for any single OTU {[}Figure
-S1{]}.
+difference between them (PERMANOVA = 0.085). There was no significant
+difference between initial and follow up samples for observed OTUs,
+Shannon diversity, or evenness after correction for multiple comparisons
+{[}Table S1{]}. There was also no significant difference between initial
+and follow up samples for any single OTU {[}Figure S1{]}.
 
@@
 differences in previously well described carcinoma associated OTUs.
-These included the OTUs that aligned with Porphyromonas asaccharolytica
-(Otu000153), Fusobacterium nucleatum (Otu000226), Parvimonas micra
-(Otu000460), and Peptostreptococcus stomatis (Otu000653). First, the
-carcinoma samples showed a significant difference between initial and
-follow up samples for Peptostreptococcus stomatis (P-value = 0.0183) and
-Porphyromonas asaccharolytica (P-value = 0.0154) whereas there were no
-signficant differences in any of these OTUs in the adenoma samples
-{[}Table S4{]}. Second, when these OTUs were present, there was a clear
-magnitude difference in these specific OTUs based on whether they were
-from adenoma or carcinoma individuals {[}Figure 2{]}. However, only a
-small percentage of those with adenoma or carcinoma were positive for
-any of these OTUs.
+These included the OTUs that aligned with \emph{Porphyromonas
+asaccharolytica} (Otu000153), \emph{Fusobacterium nucleatum}
+(Otu000226), \emph{Parvimonas micra} (Otu000460), and
+\emph{Peptostreptococcus stomatis} (Otu000653). First, the carcinoma
+samples showed a significant difference between initial and follow up
+samples for \emph{Peptostreptococcus stomatis} (P-value = 0.0183) and
+\emph{Porphyromonas asaccharolytica} (P-value = 0.0154) whereas there
+was no signficant differences in any of these OTUs in the adenoma
+samples {[}Table S4{]}. Second, when these OTUs were present, there was
+a clear magnitude difference based on whether they were from adenoma or
+carcinoma {[}Figure 2{]}. However, only a small percentage of those with
+adenoma or carcinoma were positive for any of these OTUs.
 
@@
 observed between initial and follow up samples and only a small number
-of individuals were positive for previously associated CRC bacteria we
+of individuals were positive for previously associated CRC bacteria; we
 next investigated if we could create a model that could adequately
@@
 and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
-initial follow up model which had and AUC range of 0.451 to 0.67 after
+initial follow up model which had an AUC range of 0.451 to 0.67 after
 100 20 repeated 10-fold cross validations. Interestingly, identification
@@
 curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
-final lesion model used falls within the range of both the full and
+final lesion model used was within the range of both the full and
 reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline discussion

@@
 Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.
-This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.
+This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, nor any addition of new or depressed bacterium but rather that a shift of the bacterial community as a whole occurs with the existing members of this microbiome.
 
-We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or chaning due to surgery.
-We found that the OTUs that made up most important variables to the model overwhelmingly belonged to comensal bacteria.
-With only the lesion model having a single OTU from previously associated cancer bacteria (Porphyromonas asaccharolytica).
+We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or changing due to surgery.
+We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.
+With only the lesion model having a single OTU from previously associated cancer bacteria (*Porphyromonas asaccharolytica*).
 Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
-Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and all belonged to regular residents of our gut community.
+Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
 
@@
 There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.
-These results would indicate that the findings described on specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.
+These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.
 
@@
 Based on previous work by Arthur, et al.
-[@arthur_intestinal_2012]  it may not be surprising to see E.coli as one of the most important OTUs and one that was common to both models.
+[@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.
 Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.
@@
 Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
-Finally, it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
-It is also possible that all three of these potential explanations could have played a role.
-Regardless, our observations would suggest that an individual's resident bacteria have a role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery.
+It is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
+Finally, it is also possible that all three of these potential explanations could have played a role.
+Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
 It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).
@@
 One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
-This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
+This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
 Another limitation is that we do not know if adding modern tests such as the fecal DNA test [@cotter_long-term_2016] could help improve our overall AUC.
@@
 
-Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.
+Another interesting outcome was that Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.
 This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
-Another potential reason is that the model itself may be overfit since the total number of samples is not that large.
+Another potential reason is that the model itself may be overfit since the total number of samples was not that large.
 However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.
-Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with fecal, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
+Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
 

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline discussion

@@
 out of the 4 were significant for the adenoma group. This data suggests
-that what may change after surgery is not necessarily any one specific
-bacterium nor any addition of new or depressed bacterium but rather that
-a shift of the bacterial community as a whole occurs with the existing
-members of this microbiome.
+that the changes that may be important in the bacterial microbiome after
+surgery are not necessarily any one specific bacterium, a clear
+depression of CRC related bacteria, nor any addition of new or depressed
+bacterium but rather that a shift of the bacterial community as a whole
+occurs with the existing members of this microbiome.
 
@@
 or initial and follow up samples in order to confirm that the community
-was what was responding or chaning due to surgery. We found that the
-OTUs that made up most important variables to the model overwhelmingly
-belonged to comensal bacteria. With only the lesion model having a
-single OTU from previously associated cancer bacteria (Porphyromonas
-asaccharolytica). Using only these important OTUs and FIT both models
-(lesion and initial and follow up) had a significantly decreased
-positive probability of either lesion or being an initial sample on
-follow up {[}Figure 4 \& S4{]}. Confirming the importance of the changes
-of comensal bacteria to these classifications was that a total of 14
-OTUs were common to both models and all belonged to regular residents of
-our gut community.
+was what was responding or changing due to surgery. We found that the
+OTUs that made up the most important variables to the model
+overwhelmingly belonged to comensal bacteria. With only the lesion model
+having a single OTU from previously associated cancer bacteria
+(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
+and FIT both models (lesion and initial and follow up) had a
+significantly decreased positive probability of either lesion or being
+an initial sample on follow up {[}Figure 4 \& S4{]}. Confirming the
+importance of the changes of comensal bacteria to these classifications
+was that a total of 14 OTUs were common to both models and the vast
+majority belonged to regular residents of our gut community.
 
@@
 sample between adenoma and carcinoma. These results would indicate that
-the findings described on specific to the surgical intervention and that
-differences observed between carcinoma and adenoma samples can not be
-simply attributed to collection time between samples.
+the findings described were specific to the surgical intervention and
+that differences observed between carcinoma and adenoma samples can not
+be simply attributed to collection time between samples.
 
@@
 removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
-it may not be surprising to see E.coli as one of the most important OTUs
-and one that was common to both models. Interestingly, many of the most
-important OTUs had taxonomic identification for resident gut microbes.
-This could suggest that the bacterial community is one of the first
-components that could change during the pathogenesis of disease. These
-bacterial microbiome changes could be the first step in allowing more
-inflammatory bacterium to gain a foothold within the colon {[}19{]}.
+it may not be surprising to have E.coli as one of the most important
+OTUs and one that was common to both models. Interestingly, many of the
+most important OTUs had taxonomic identification for resident gut
+microbes. This could suggest that the bacterial community is one of the
+first components that could change during the pathogenesis of disease.
+These bacterial microbiome changes could be the first step in allowing
+more inflammatory bacterium to gain a foothold within the colon
+{[}19{]}.
 
@@
 enough individuals to be able to classify those with and without disease
-with a high degree of accuracy. Finally, it is possible that our Random
-Forest models were able to get the same information from measures such
-as FIT or other OTUs. It is also possible that all three of these
+with a high degree of accuracy. It is possible that our Random Forest
+models were able to get the same information from measures such as FIT
+or other OTUs. Finally, it is also possible that all three of these
 potential explanations could have played a role. Regardless, our
 observations would suggest that an individual's resident bacteria have a
-role to play in disease initiation and could change in a way that allows
-predictive models to lower the positive probability of a lesion after
-surgery. It should be stated that our study does not argue against the
-importance of these CRC associated bacteria in the pathogenesis of
-disease but rather that the models do not utilize these specific
-bacteria for classification purposes (lesion or initial and follow up).
-In fact, it is possible that these CRC associated bacteria are important
-in the transition from adenoma to carcinoma and would be one explanation
-as to why in our data we not only see high initial relative abundances,
-in certain individuals, but also large decreases in relative abundance
-in those with carcinoma but not in those with adenoma after surgery
-{[}Figure 2{]}.
+large role to play in disease initiation and could change in a way that
+allows predictive models to lower the positive probability of a lesion
+after surgery {[}Figure 4{]}. It should be stated that our study does
+not argue against the importance of these CRC associated bacteria in the
+pathogenesis of disease but rather that the models do not utilize these
+specific bacteria for classification purposes (lesion or initial and
+follow up). In fact, it is possible that these CRC associated bacteria
+are important in the transition from adenoma to carcinoma and would be
+one explanation as to why in our data we not only see high initial
+relative abundances, in certain individuals, but also large decreases in
+relative abundance in those with carcinoma but not in those with adenoma
+after surgery {[}Figure 2{]}.
 
@@
 a subsequent CRC diagnosis. This information would help to strengthen
-the case for our Random Forest based model keeping a number of
+the case for our Random Forest based model to have kept a number of
 individuals above the cutoff threshold even though at follow up they
@@
 
-Within figure 3 the initial and follow up model showed better test AUC
-results then the training set AUC. This may have occured because the
-training AUC that was determined from 20 repeated 10 fold cross
-validation removed samples at random and did not take into account that
-they were matched samples. Another potential reason is that the model
-itself may be overfit since the total number of samples is not that
-large. However, the lesion model did not suffer from these discrepencies
-and similar conclusions can be drawn solely from this model. Regardless,
-further independent studies will need to be carried out to verify our
-findings since not only are we dealing with fecal, which could be very
-different than the communities present on the actual tissue, but also
-are dealing with correlations that may not be representative of the true
-pathogensis of disease.
+Another interesting outcome was that Within figure 3 the initial and
+follow up model showed better test AUC results then the training set
+AUC. This may have occured because the training AUC that was determined
+from 20 repeated 10 fold cross validation removed samples at random and
+did not take into account that they were matched samples. Another
+potential reason is that the model itself may be overfit since the total
+number of samples was not that large. However, the lesion model did not
+suffer from these discrepencies and similar conclusions can be drawn
+solely from this model. Regardless, further independent studies will
+need to be carried out to verify our findings since not only are we
+dealing with feces, which could be very different than the communities
+present on the actual tissue, but also are dealing with correlations
+that may not be representative of the true pathogensis of disease.
 

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline methods and figure legends

@@
 In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
-Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
+Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
 The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].
-Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
+Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
 This study was approved by the University of Michigan Institutional Review Board.
@@
 
-***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
+***FIT and 16S rRNA Gene Sequencing***
 FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013].
@@
 This process has been previously described [@kozich_development_2013].
-The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
+The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
 The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.
@@
 ***Lesion Model Creation***
-The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.
+The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.
+This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.
 The model included data on FIT and the bacterial microbiome.
-Non-binary data was checked for near zero variance and auto correlation.
+Non-binary data was checked for near zero variance.
 Data columns that had near zero variance were removed.
-Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
@@
 We also investigated whether a model could be created that could identify before and after surgery samples.
-The training set utilized the `r length(rownames(good_metaf))`-person cohort that was previously used for testing of the lesion model.
-The creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.
+The main difference was that only the `r length(rownames(good_metaf))`-person cohort follow up data set was used at all stages of model building and classification.
+Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.
 
 ***Selection of Important OTUs***
-In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.
+In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.
 This final collated list of variables was what was considered the most important for the lesion or initial follow up models.
@@
 The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.
-Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al.
-[@delong_comparing_1988] executed by the pROC (v`r paste(packageVersion("pROC"))`) package.
-Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.
+Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
+Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired wilcoxson ranked sum test.
 Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
-Unless otherwise stated the P-values reported are those of the BH corrected ones.
+Unless otherwise stated the P-values reported are those of that were BH corrected.
 
 ***Analysis Overview***
-Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.
-Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.
-All OTUs used in the lesion model were then also analyzed using a paired wilcoxson test.
-We then investigated the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.
-Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
-The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
-The most important OTUs for this were used to build an updated model and this reduced feature model was assessed for it's similarity to the original model.
-We then used the initial follow up model to assess whether this model could classify samples better then the lesion model.
+Initial and follow up samples were analyzed for differences in alpha and beta diversity.
+Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.
+From here, all OTUs that were used in the lesion model were then analyzed using a paired wilcoxson test.
+We then investigated the relative abundance of specific previously CRC associated bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.
+We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
+From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
+The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.
+We then used the initial follow up model to assess whether it could classify samples better then the lesion model.
 The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.
@@
 
-**Figure 1: Changes and differences between the adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
-B) A significant difference was found between the adenoma and carcinoma group for FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 C) NMDS of the intial and follow up samples for the Adenoma group.
@@
 **Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
-There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+There was a significant difference in initial and follow up sample for the OTUs classfied as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
 B) Adenoma initial and follow up samples.
@@
 
-**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Initial/Follow up models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
+**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Initial and Follow up Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
 The blue line represents the reduced variable model using 100% of the data set.
@@
 
-**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up for Reduced Variable Models**  A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
-The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.
+**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Follow Up Reduced Variable Models**  A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
 
@@
 
-**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up for Full Variable Model** A) Positive probability adjustment of those with carcinoma from intial to follow up sample  B) Positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
-The dotted line represents the threshold used to make the decision of whether a sample was lesion positive or not.
+**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up Full Variable Models** A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
 

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline methods and figure legends

@@
 hereditary non-polyposis colorectal cancer, or Familial adenomatous
-polyposis. Samples used to build the model used for prediction were
+polyposis. Samples used to build the models for prediction were
 collected either prior to a colonoscopy or between 1 - 2 weeks after.
 The bacterial microbiome has been shown to nomralize within this time
-period {[}21{]}. Kept apart from this training set were a total of 67
-individuals that not only had a sample as described previoulsy but also
-a follow up sample between 188 - 546 days after surgery and treatment
-had been completed. This study was approved by the University of
-Michigan Institutional Review Board. All study participants provided
-informed consent and the study itself conformed to the guidelines set
-out by the Helsinki Declaration.
+period {[}21{]}. Our follow up data set had a total of 67 individuals
+that not only had a sample as described previoulsy but also a follow up
+sample between 188 - 546 days after surgery and treatment had been
+completed. This study was approved by the University of Michigan
+Institutional Review Board. All study participants provided informed
+consent and the study itself conformed to the guidelines set out by the
+Helsinki Declaration.
 
-\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
-FIT was analyzed as previously published using both OC FIT-CHEK and
-OC-Auto Micro 80 automated system (Polymedco Inc.) {[}22{]}. 16S rRNA
-gene sequencing was completed as previously described by Kozich, et al.
-{[}23{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
-kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
-system (Eppendorf). The V4 variable region was amplified and the
-resulting product was split between three sequencing runs with control,
-adenoma, and carcinoma evenly represented on each run. Each group was
-randomly assigned to avoid biases based on sample collection location.
+\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
+previously published using both OC FIT-CHEK and OC-Auto Micro 80
+automated system (Polymedco Inc.) {[}22{]}. 16S rRNA gene sequencing was
+completed as previously described by Kozich, et al. {[}23{]}. In brief,
+DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
+Laboratories) and an epMotion 5075 automated pipetting system
+(Eppendorf). The V4 variable region was amplified and the resulting
+product was split between three sequencing runs with control, adenoma,
+and carcinoma evenly represented on each run. Each group was randomly
+assigned to avoid biases based on sample collection location.
 
@@
 has been previously described {[}23{]}. The general processing workflow
-using mothur is as follows: Paired-end reads were first merged into
-contigs, quality filtered, aligned to the SILVA database, screening for
+using mothur was as follows: Paired-end reads were first merged into
+contigs, quality filtered, aligned to the SILVA database, screened for
 chimeras, classified with a naive Bayesian classifier using the
@@
 algorithm was used to create the model used for prediction of lesion
-(adenoma or carcinoma) for the 67 individuals with follow up samples.
-The model included data on FIT and the bacterial microbiome. Non-binary
-data was checked for near zero variance and auto correlation. Data
-columns that had near zero variance were removed. Columns that were
-correlated with each other over a Spearman correlation coefficient of
-0.75 had one of the two columns removed. This pre-processing was
-performed with the R package caret (v6.0.73). Optimization of the mtry
-hyperparameter involved taking the samples and making 100 80/20
-(train/test) splits in the data where control and lesion were equally
-represented in the 80 and 20 split, respectively. This 80\% portion was
-then split again into an 80/20 split, and run through 20 repeated
-10-fold cross validations to optimzie the model's AUC (Area Under the
-Curve of the Receiver Operator Characteristic). This resulting model was
-then tested on the 20\% of the data that was originally held out from
-this overall process. Once the ideal mtry was found the entire 490
-sample set was used to create the final Random Forest model on which
-testing on the 67-person cohort was completed. The default cutoff of 0.5
-was used as the threshold to classify individuals as positive or
-negative for lesion. The hyperparameter, mtry, defines the number of
-variables to investigate at each split before a new division of the data
-is created.
+(adenoma or carcinoma) with the main testing and training of the model
+completed on a data set of 490 individuals. This model was then applied
+to our follow up data set of 67 individuals. The model included data on
+FIT and the bacterial microbiome. Non-binary data was checked for near
+zero variance. Data columns that had near zero variance were removed.
+This pre-processing was performed with the R package caret (v6.0.73).
+Optimization of the mtry hyperparameter involved taking the samples and
+making 100 80/20 (train/test) splits in the data where control and
+lesion were equally represented in the 80 and 20 split, respectively.
+This 80\% portion was then split again into an 80/20 split, and run
+through 20 repeated 10-fold cross validations to optimzie the model's
+AUC (Area Under the Curve of the Receiver Operator Characteristic). This
+resulting model was then tested on the 20\% of the data that was
+originally held out from this overall process. Once the ideal mtry was
+found the entire 490 sample set was used to create the final Random
+Forest model on which testing on the 67-person cohort was completed. The
+default cutoff of 0.5 was used as the threshold to classify individuals
+as positive or negative for lesion. The hyperparameter, mtry, defines
+the number of variables to investigate at each split before a new
+division of the data is created.
 
@@
 whether a model could be created that could identify before and after
-surgery samples. The training set utilized the 67-person cohort that was
-previously used for testing of the lesion model. The creation of this
-model and optimization of the mtry hyperparameter was completed using
-the same procedure that was used to create the lesion model.
+surgery samples. The main difference was that only the 67-person cohort
+follow up data set was used at all stages of model building and
+classification. Other than this difference the creation of this model
+and optimization of the mtry hyperparameter was completed using the same
+procedure that was used to create the lesion model.
 
 \textbf{\emph{Selection of Important OTUs}} In order to assess which
-variables were most central to all the models we counted the number of
+variables were most important to all the models we counted the number of
 times a variable was present in the top 10\% of mean decrease in
@@
 community structure utilized PERMANOVA {[}25{]} in the vegan package
-(v2.4.1) while comparisons between ROC curves utilized the method by
-DeLong et al. {[}26{]} executed by the pROC (v1.8) package. Comparisons
-between probabilities as well as overall amount of OTU between initial
-and follow up samples utilized a paired wilcoxson ranked sum test. Where
-multiple comparison testing was needed a Benjamini-Hochberg (BH)
-correction was applied {[}27{]} and a corrected P-value of less than
-0.05 was considered significant. Unless otherwise stated the P-values
-reported are those of the BH corrected ones.
+(v2.4.1). Comparisons between probabilities as well as overall OTU
+differences between initial and follow up samples utilized a paired
+wilcoxson ranked sum test. Where multiple comparison testing was needed
+a Benjamini-Hochberg (BH) correction was applied {[}26{]} and a
+corrected P-value of less than 0.05 was considered significant. Unless
+otherwise stated the P-values reported are those of that were BH
+corrected.
 
-\textbf{\emph{Analysis Overview}} Differences in FIT between initial and
-follow ups for either adenoma or carcinoma were investigated. Next,
-initial and follow up samples were analyzed for differences in alpha and
-beta diversity. All OTUs used in the lesion model were then also
-analyzed using a paired wilcoxson test. We then investigated the
-relative abundance of specific bacteria, that have been previously
-associated with CRC, we selected OTUs that taxonomically classified to
-Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
-Assacharolytica, and Porphyromonas Stomatis. Specifically, we wanted to
-test if there were any differences based on whether the individual had
-an adenoma or carcinoma. The lesion model was then tested for accuracy
-in prediction and whether it reduced the positive probability of lesion
-after surgery. The most important OTUs for this were used to build an
-updated model and this reduced feature model was assessed for it's
+\textbf{\emph{Analysis Overview}} Initial and follow up samples were
+analyzed for differences in alpha and beta diversity. Next, differences
+in FIT between initial and follow ups for either adenoma or carcinoma
+were investigated. From here, all OTUs that were used in the lesion
+model were then analyzed using a paired wilcoxson test. We then
+investigated the relative abundance of specific previously CRC
+associated bacteria, specifically, OTUs that taxonomically classified to
+\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
+\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
+stomatis}. We wanted to test if there were any differences based on
+whether the individual had an adenoma or carcinoma. From here the lesion
+model was then tested for accuracy in prediction and whether it reduced
+the positive probability of lesion after surgery. The most important
+OTUs for this were used to build a reduced model and it was assessed for
 similarity to the original model. We then used the initial follow up
-model to assess whether this model could classify samples better then
-the lesion model. The most important OTUs were then identified from this
-model and used to create a reduced feature initial follow up model. This
-reduced feature model, as was done with the lesion model, was compared
-to the full model for loss of accuracy. Finally, a list of common OTUs
-were found for the two different models used.
+model to assess whether it could classify samples better then the lesion
+model. The most important OTUs were then identified from this model and
+used to create a reduced feature initial follow up model. This reduced
+feature model, as was done with the lesion model, was compared to the
+full model for loss of accuracy. Finally, a list of common OTUs were
+found for the two different models used.
 
@@
 
-\textbf{Figure 1: Changes and differences between the adenoma or
-carcinoma group.} A) No significant difference was found between the
-adenoma and carcinoma group for thetayc (P-value = 0.00198). B) A
-significant difference was found between the adenoma and carcinoma group
-for FIT (P-value = 2.15e-05). C) NMDS of the intial and follow up
-samples for the Adenoma group. D) NMDS of the initial and follow up
-samples for the Carcinoma group. For C) and D) the teal represents
-initial samples and the pink represents follow up samples.
+\textbf{Figure 1: General Differences between the Adenoma or Carcinoma
+Group.} A) A significant difference was found between the adenoma and
+carcinoma group for thetayc (P-value = 0.00198). B) A significant
+difference was found between the adenoma and carcinoma group for change
+in FIT (P-value = 2.15e-05). C) NMDS of the intial and follow up samples
+for the Adenoma group. D) NMDS of the initial and follow up samples for
+the Carcinoma group. For C) and D) the teal represents initial samples
+and the pink represents follow up samples.
 
@@
 was a significant difference in initial and follow up sample for the
-OTUs classfied as Peptostreptococcus stomatis (P-value = 0.0183) and
-Porphyromonas asaccharolytica (P-value = 0.0154). B) Adenoma initial and
-follow up samples. There were no significant differences between initial
-and follow up.
+OTUs classfied as \emph{Peptostreptococcus stomatis} (P-value = 0.0183)
+and \emph{Porphyromonas asaccharolytica} (P-value = 0.0154). B) Adenoma
+initial and follow up samples. There were no significant differences
+between initial and follow up.
 
 \textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
-for lesion and Initial/Follow up models.} The shaded areas represents
-the range of values of a 100 different 80/20 splits of the test set data
-using either all variables (grey) or reduced variable (red) models. The
-blue line represents the reduced variable model using 100\% of the data
-set. A) Lesion model. B) Initial/Follow up model
+for lesion and Initial and Follow up Models.} The shaded areas
+represents the range of values of a 100 different 80/20 splits of the
+test set data using either all variables (grey) or reduced variable
+(red) models. The blue line represents the reduced variable model using
+100\% of the data set. A) Lesion model. B) Initial/Follow up model
 
 \textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Initial and Follow Up for Reduced Variable
-Models} A) Positive probability adjustment of those with carcinoma from
-intial to follow up sample B) Positive probability adjustment of those
+Results for Lesion and Initial Follow Up Reduced Variable Models} A)
+Lesion positive probability adjustment of those with carcinoma from
+initial to follow up sample B) Initial follow up positive probability
+adjustment of those with carcinoma from initial to follow up sample C)
+Lesion positive probability adjustment of those with adenoma as well as
+those with SRN and the probability adjustment from initial to follow up
+sample. D) Initial follow up positive probability adjustment of those
 with adenoma as well as those with SRN and the probability adjustment
 from initial to follow up sample. The dotted line represents the
-threshold used to make the decision of whether a sample was lesion
-positive or not.
+threshold used to make the decision of whether a sample was positive or
+not.
 
@@
 \textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Initial and Follow Up for Full Variable Model} A)
-Positive probability adjustment of those with carcinoma from intial to
-follow up sample B) Positive probability adjustment of those with
-adenoma as well as those with SRN and the probability adjustment from
-initial to follow up sample. The dotted line represents the threshold
-used to make the decision of whether a sample was lesion positive or
+Results for Lesion and Initial and Follow Up Full Variable Models} A)
+Lesion positive probability adjustment of those with carcinoma from
+initial to follow up sample B) Initial follow up positive probability
+adjustment of those with carcinoma from initial to follow up sample C)
+Lesion positive probability adjustment of those with adenoma as well as
+those with SRN and the probability adjustment from initial to follow up
+sample. D) Initial follow up positive probability adjustment of those
+with adenoma as well as those with SRN and the probability adjustment
+from initial to follow up sample. The dotted line represents the
+threshold used to make the decision of whether a sample was positive or
 not.
@@
 
-\hypertarget{ref-delong_comparing_1988}{}
-26. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
-two or more correlated receiver operating characteristic curves: A
-nonparametric approach. Biometrics. 1988;44:837--45.
-
 \hypertarget{ref-benjamini_controlling_1995}{}
-27. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+26. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-01-18 submission/manuscript_outline_20161024.Rmd
improve flow - abstract

@@
 Previous research suggests a link between fecal bacterial microbiome and CRC.
-The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e.
+The overall objective of our study was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e.
 adenoma or carcinoma).
 Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.
-We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.
+We also wanted to investigate the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models which classify individuals as having a lesion or whether samples could be classified as before or after surgery.
 **Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT  (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before and after surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
 Both models had a significant decrease in the positive probability of a lesion or being a before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
 In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
-**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.
+**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion within our models but also causes detectable changes in the bacterial microbiome.
 Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

### 2017-01-18 submission/manuscript_outline_20161024.tex
improve flow - abstract

@@
 mitigating deaths due to the disease. Previous research suggests a link
-between fecal bacterial microbiome and CRC. The overall objective was to
-investigate the changes in the bacterial microbiome after surgery in
-patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
-wanted to identify what within the community was different within those
-undergoing surgical removal of lesion. We also investigated the use of
-the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
-models to either classify individuals as having a lesion or whether,
-based on the bacterial microbiome, the sample could be classified
-correctly as before or after surgery.\\
+between fecal bacterial microbiome and CRC. The overall objective of our
+study was to investigate the changes in the bacterial microbiome after
+surgery in patients with lesion (i.e.~adenoma or carcinoma).
+Specifically, we wanted to identify what within the community was
+different within those undergoing surgical removal of lesion. We also
+wanted to investigate the use of the bacterial microbiome and Fecal
+Immunoglobulin Test (FIT) to build models which classify individuals as
+having a lesion or whether samples could be classified as before or
+after surgery.\\
 \textbf{Results:} Adenoma individual's bacterial microbiome were more
@@
 before and after surgery (P-value \textgreater{} 0.125). A model with a
-total of 37 variables was able to classify lesion (AUC = 0.847 - 0.791
+total of 37 variables was able to classify lesion (AUC = 0.847 - 0.791)
 while the model to classify samples as before and after surgery had 33
@@
 \textbf{Conclusions:} Our data suggests that treatment not only
-significantly reduces the probability of having a colonic lesion but
-also causes detectable changes in the bacterial microbiome. Further
-surveillance of these individuals will enable us to determine whether
-models such as the one we present here can also be used to predict
-recurrence of colorectal cancer.
+significantly reduces the probability of having a colonic lesion within
+our models but also causes detectable changes in the bacterial
+microbiome. Further surveillance of these individuals will enable us to
+determine whether models such as the one we present here can also be
+used to predict recurrence of colorectal cancer.
 

### 2017-01-18 submission/manuscript_outline_20161024.Rmd
streamline - abstract part 2

@@
 Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.
-We also wanted to investigate the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models which classify individuals as having a lesion or whether samples could be classified as before or after surgery.
-**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT  (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+We also wanted to investigate the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models which could classify individuals as having a lesion or as before surgery.
+**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).
+Their change in FIT was also significantly different then those with carcinoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before and after surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-Both models had a significant decrease in the positive probability of a lesion or being a before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+The after surgery samples had a significant decrease in the model positive probability of a lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).
 **Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion within our models but also causes detectable changes in the bacterial microbiome.
-Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.
+Further surveillance of these individuals will enable us to determine whether models such as the ones we present could be used to predict recurrence of colorectal cancer.
 

### 2017-01-18 submission/manuscript_outline_20161024.tex
streamline - abstract part 2

@@
 wanted to investigate the use of the bacterial microbiome and Fecal
-Immunoglobulin Test (FIT) to build models which classify individuals as
-having a lesion or whether samples could be classified as before or
-after surgery.\\
+Immunoglobulin Test (FIT) to build models which could classify
+individuals as having a lesion or as before surgery.\\
 \textbf{Results:} Adenoma individual's bacterial microbiome were more
 similar to their pre-surgery sample then those with carcinoma (P-value =
-0.00198) and this was also reflected in FIT (P-value = 2.15e-05). There
-was no significant difference in any indivdiual OTU between samples
-before and after surgery (P-value \textgreater{} 0.125). A model with a
-total of 37 variables was able to classify lesion (AUC = 0.847 - 0.791)
-while the model to classify samples as before and after surgery had 33
-variables (AUC = 0.79 - 0.651). Both models had a significant decrease
-in the positive probability of a lesion or being a before surgery sample
-(P-value = 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were
-common to both models and the majority of these classified to commensal
-bacteria with the largest representation belonging to Bacteroides,
-Blautia, Streptococcus, and Clostridiales.\\
+0.00198). Their change in FIT was also significantly different then
+those with carcinoma (P-value = 2.15e-05). There was no significant
+difference in any indivdiual OTU between samples before and after
+surgery (P-value \textgreater{} 0.125). A model with a total of 37
+variables was able to classify lesion (AUC = 0.847 - 0.791) while the
+model to classify samples as before surgery had 33 variables (AUC = 0.79
+- 0.651). The after surgery samples had a significant decrease in the
+model positive probability of a lesion or before surgery sample (P-value
+= 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were common
+to both models and the majority of these classified to commensal
+bacteria (Bacteroides, Blautia, Streptococcus, and Clostridiales).\\
 \textbf{Conclusions:} Our data suggests that treatment not only
@@
 microbiome. Further surveillance of these individuals will enable us to
-determine whether models such as the one we present here can also be
-used to predict recurrence of colorectal cancer.
+determine whether models such as the ones we present could be used to
+predict recurrence of colorectal cancer.
 

### 2017-01-18 submission/manuscript_outline_20161024.Rmd
streamline -Introduction

@@
 
-In this study we investigated what happened to the bacterial microbiome before and after surgery for indivdiuals with either adenoma and carcinoma.
+In this study we investigated what happened to the bacterial microbiome before and after surgery for indivdiuals with either adenoma or carcinoma.
 Our anlaysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest models and observed how these models as well as specific OTUs within this model performed on before (initial) and after (follow up) surgery samples.
-We used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
+We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed before (initial) and after (follow up) surgery.
+We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
 

### 2017-01-18 submission/manuscript_outline_20161024.tex
streamline -Introduction

@@
 In this study we investigated what happened to the bacterial microbiome
-before and after surgery for indivdiuals with either adenoma and
+before and after surgery for indivdiuals with either adenoma or
 carcinoma. Our anlaysis includes both alpha and beta diversity analysis
@@
 (OTUs). We also utilized Random Forest models and observed how these
-models as well as specific OTUs within this model performed on before
-(initial) and after (follow up) surgery samples. We used these models to
-look for similar important OTUs to identify the crucial OTUs for not
-only classifying initial and follow up samples but also lesion or
-normal.
+models as well as specific OTUs within them performed before (initial)
+and after (follow up) surgery. We also used these models to look for
+similar important OTUs to identify the crucial OTUs for not only
+classifying initial and follow up samples but also lesion or normal.
 

### 2017-01-18 submission/manuscript_outline_20161024.Rmd
streamline -results

@@
 
-***Bacterial Community and FIT Changes before and after Surgery***
-Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
+***Bacterial Community and FIT***
+Based on the thetayc distance metric, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
 There was also a significant difference in change in FIT between initial and follow up samples (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+The whole community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
 When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
@@
 
-***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
+***Previously Associated Cancer Bacteria***
 We next examined whether there were differences in previously well described carcinoma associated OTUs.
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there was no signficant differences in any of these OTUs in the adenoma samples [Table S4].
+First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
 Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].
@@
 
-***Full and Reduced Model Results***
-Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria; we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.
-The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.
-Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+***Full and Reduced Model***
+Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria; we next investigated if we could create models that could adequately classify and adjust either lesion or before sample probability based on the bacterial community and FIT.
+The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the before sample model which had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
+Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and before sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
 
@@
 The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].
-Interestingly, the test set AUC range for the initial and follow up performed much better then the training set AUCs.
-Both the full and reduced initial and follow up models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.
+Interestingly, the test set AUC range for the before sample model performed much better then the training set AUCs.
+Both the full and reduced before sample models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.
 
-***Most Important Variables to the Models***
+***Most Important Variables***
 The reduced models were built based on the most important variables to the respective full model.
-For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial and follow up model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
+For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the before sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
 For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].
 
-***Positive Probability Prediction after Surgical Removal of Adenoma or Carcinoma***
-Regardless of model used there was a significant decrease in the positive probability of either the sample being lesion or an initial sample on follow up [Figure 4 & S4] (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+***Positive Probability after Lesion Removal***
+Regardless of model used there was a significant decrease in the positive probability of either the sample being lesion or a before sample on follow up [Figure 4 & S4] (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
 
-For the full and reduced lesion model there was a significant difference in the classification for the lesion model between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)`, respectively) but not for the initial follow up model (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)`).
-However, the lesion model correctly kept the one individual who still had a carcinoma on follow up above the cut off point [Figure 4a & S4a] for a positive call while the initial and follow up models did not [Figure 4b & S4b].
+For the full and reduced lesion model there was a significant difference in the classification for the lesion model between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)`, respectively) but not for the before sample model (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)`).
+However, the lesion model correctly kept the one individual who still had a carcinoma on follow up above the cut off threshold [Figure 4a & S4a] for a positive call while the before sample models did not [Figure 4b & S4b].
 
-***Common OTUs to both Lesion and Initial and Follow Up Models***
+***Common OTUs to both Models***
 There were a total of `r length(common_vars_summary$otu)` OTUs that were common to both models.
 Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.
-The majority of these OTUs had taxonomic identification to bacteria typically thought of as commensal [Table S2].
+The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
 
@@
 There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).
-Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
+Although the before sample model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
 Time elapsed between the collection of the follow up sample from the initial sample, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

### 2017-01-18 submission/manuscript_outline_20161024.tex
streamline -results

@@
 
-\textbf{\emph{Bacterial Community and FIT Changes before and after
-Surgery}} Based on thetayc distance metrics, comparing the initial to
-the follow up samples, there was a significant difference between the
-adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
-was also a significant difference in change in FIT between initial and
-follow up samples (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
-community structure before and after surgery were visualized on NMDS
-graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.997) and carcinoma
-{[}Figure 1d{]} (PERMANOVA = 0.002). When all initial and follow up
-samples were compared to each other there was no significant overall
-difference between them (PERMANOVA = 0.085). There was no significant
-difference between initial and follow up samples for observed OTUs,
-Shannon diversity, or evenness after correction for multiple comparisons
+\textbf{\emph{Bacterial Community and FIT}} Based on the thetayc
+distance metric, comparing the initial to the follow up samples, there
+was a significant difference between the adenoma and carcinoma groups
+(P-value = 0.00198) {[}Figure 1a{]}. There was also a significant
+difference in change in FIT between initial and follow up samples
+(P-value = 2.15e-05) {[}Figure 1b{]}. The whole community structure
+before and after surgery was visualized on NMDS graphs for both adenoma
+{[}Figure1c{]} (PERMANOVA = 0.997) and carcinoma {[}Figure 1d{]}
+(PERMANOVA = 0.002). When all initial and follow up samples were
+compared to each other there was no significant overall difference
+between them (PERMANOVA = 0.085). There was no significant difference
+between initial and follow up samples for observed OTUs, Shannon
+diversity, or evenness after correction for multiple comparisons
 {[}Table S1{]}. There was also no significant difference between initial
@@
 
-\textbf{\emph{Differences in Adenoma and Carcinoma of Previously
-Associated Cancer Bacteria}} We next examined whether there were
-differences in previously well described carcinoma associated OTUs.
-These included the OTUs that aligned with \emph{Porphyromonas
-asaccharolytica} (Otu000153), \emph{Fusobacterium nucleatum}
-(Otu000226), \emph{Parvimonas micra} (Otu000460), and
+\textbf{\emph{Previously Associated Cancer Bacteria}} We next examined
+whether there were differences in previously well described carcinoma
+associated OTUs. These included the OTUs that aligned with
+\emph{Porphyromonas asaccharolytica} (Otu000153), \emph{Fusobacterium
+nucleatum} (Otu000226), \emph{Parvimonas micra} (Otu000460), and
 \emph{Peptostreptococcus stomatis} (Otu000653). First, the carcinoma
@@
 \emph{Porphyromonas asaccharolytica} (P-value = 0.0154) whereas there
-was no signficant differences in any of these OTUs in the adenoma
+were no signficant differences in any of these OTUs in the adenoma
 samples {[}Table S4{]}. Second, when these OTUs were present, there was
@@
 
-\textbf{\emph{Full and Reduced Model Results}} Since differences were
-observed between initial and follow up samples and only a small number
-of individuals were positive for previously associated CRC bacteria; we
-next investigated if we could create a model that could adequately
-classify and adjust lesion probability based on the bacterial community
-and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
-initial follow up model which had an AUC range of 0.451 to 0.67 after
-100 20 repeated 10-fold cross validations. Interestingly, identification
-of the most important variables and reducing the models to only these
-factors increased the AUC in the lesion model (0.791 - 0.847) and
-initial and follow up model (0.651 - 0.79).
+\textbf{\emph{Full and Reduced Model}} Since differences were observed
+between initial and follow up samples and only a small number of
+individuals were positive for previously associated CRC bacteria; we
+next investigated if we could create models that could adequately
+classify and adjust either lesion or before sample probability based on
+the bacterial community and FIT. The lesion model had an AUC range of
+0.723 to 0.795 versus the before sample model which had an AUC range of
+0.451 to 0.67 after 100 iterations of 20 repeated 10-fold cross
+validations. Interestingly, identification of the most important
+variables and reducing the models to only these factors increased the
+AUC in the lesion model (0.791 - 0.847) and before sample model (0.651 -
+0.79).
 
@@
 reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
-range for the initial and follow up performed much better then the
-training set AUCs. Both the full and reduced initial and follow up
-models overlapped with each other {[}Figure 3b{]} there was a marked
-decrease in the ROC curve for the final model used.
+range for the before sample model performed much better then the
+training set AUCs. Both the full and reduced before sample models
+overlapped with each other {[}Figure 3b{]} there was a marked decrease
+in the ROC curve for the final model used.
 
-\textbf{\emph{Most Important Variables to the Models}} The reduced
-models were built based on the most important variables to the
-respective full model. For the lesion model there were a total of 37
-variables {[}Figure S2{]} whereas for the initial and follow up model
-there were a total of 33 variables {[}Figure S3{]}. For both models FIT
-resulted in the largest decrease in MDA {[}Figure S2a \& S3a{]}.
+\textbf{\emph{Most Important Variables}} The reduced models were built
+based on the most important variables to the respective full model. For
+the lesion model there were a total of 37 variables {[}Figure S2{]}
+whereas for the before sample model there were a total of 33 variables
+{[}Figure S3{]}. For both models FIT resulted in the largest decrease in
+MDA {[}Figure S2a \& S3a{]}.
 
-\textbf{\emph{Positive Probability Prediction after Surgical Removal of
-Adenoma or Carcinoma}} Regardless of model used there was a significant
-decrease in the positive probability of either the sample being lesion
-or an initial sample on follow up {[}Figure 4 \& S4{]} (full lesion
-P-value = 1.11e-11, reduced lesion P-value = 1.91e-11, initial and
-follow up P-value = 6.71e-12, and reduced initial and follow up P-value
-= 6.72e-12).
+\textbf{\emph{Positive Probability after Lesion Removal}} Regardless of
+model used there was a significant decrease in the positive probability
+of either the sample being lesion or a before sample on follow up
+{[}Figure 4 \& S4{]} (full lesion P-value = 1.11e-11, reduced lesion
+P-value = 1.91e-11, initial and follow up P-value = 6.71e-12, and
+reduced initial and follow up P-value = 6.72e-12).
 
@@
 in the classification for the lesion model between predicted and actual
-(P-value = 4.19e-10 and 6.98e-10, respectively) but not for the initial
-follow up model (P-value = 1.00 and 1.00). However, the lesion model
+(P-value = 4.19e-10 and 6.98e-10, respectively) but not for the before
+sample model (P-value = 1.00 and 1.00). However, the lesion model
 correctly kept the one individual who still had a carcinoma on follow up
-above the cut off point {[}Figure 4a \& S4a{]} for a positive call while
-the initial and follow up models did not {[}Figure 4b \& S4b{]}.
+above the cut off threshold {[}Figure 4a \& S4a{]} for a positive call
+while the before sample models did not {[}Figure 4b \& S4b{]}.
 
-\textbf{\emph{Common OTUs to both Lesion and Initial and Follow Up
-Models}} There were a total of 14 OTUs that were common to both models.
-Of these OTUs the most common taxonomic identifications were to Blautia,
-Bacteroides, Streptococcus, and Clostridiales. The majority of these
-OTUs had taxonomic identification to bacteria typically thought of as
-commensal {[}Table S2{]}.
+\textbf{\emph{Common OTUs to both Models}} There were a total of 14 OTUs
+that were common to both models. Of these OTUs the most common taxonomic
+identifications were to Blautia, Bacteroides, Streptococcus, and
+Clostridiales. The majority of these OTUs had classifications to
+bacteria typically thought of as commensal {[}Table S2{]}.
 
@@
 reduced lesion model for either chemotherapy (P-value = 0.821 and 0.821)
-or radiation therapy (P-value = 0.69 and 0.981). Although the initial
-follow up model was similar there was a significant decrease in positive
+or radiation therapy (P-value = 0.69 and 0.981). Although the before
+sample model was similar there was a significant decrease in positive
 probability for those treated with chemotherapy (P-value = 7.04e-04 and

### 2017-01-18 submission/manuscript_outline_20161024.tex
streamline -discussion

@@
 surgery are not necessarily any one specific bacterium, a clear
-depression of CRC related bacteria, nor any addition of new or depressed
-bacterium but rather that a shift of the bacterial community as a whole
-occurs with the existing members of this microbiome.
+depression of CRC related bacteria, or any addition of new or depressed
+bacterium. Instead it may be due to a shift of the bacterial community
+as a whole between existing members of the host's microbiome.
 
@@
 microbiome to either be able to classify lesions (adenoma or carcinoma)
-or initial and follow up samples in order to confirm that the community
-was what was responding or changing due to surgery. We found that the
-OTUs that made up the most important variables to the model
-overwhelmingly belonged to comensal bacteria. With only the lesion model
-having a single OTU from previously associated cancer bacteria
-(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
-and FIT both models (lesion and initial and follow up) had a
-significantly decreased positive probability of either lesion or being
-an initial sample on follow up {[}Figure 4 \& S4{]}. Confirming the
-importance of the changes of comensal bacteria to these classifications
-was that a total of 14 OTUs were common to both models and the vast
-majority belonged to regular residents of our gut community.
+or before samples in order to confirm that the community was what was
+responding or changing due to surgery. We found that the OTUs that made
+up the most important variables to the model overwhelmingly belonged to
+comensal bacteria. With only the lesion model having a single OTU from a
+previously associated cancer bacterium (\emph{Porphyromonas
+asaccharolytica}). Using only these important OTUs and FIT both models
+(lesion and before sample) had a significantly decreased positive
+probability of either lesion or being an initial sample on follow up
+{[}Figure 4 \& S4{]}. Further confirmation of the importance of the
+changes of comensal bacteria to these classifications was that a total
+of 14 OTUs were common to both models and the vast majority belonged to
+regular residents of our gut community.
 
@@
 Curiously, we observed that the typical CRC associated bacteria were not
-predictive within our models. One potential explanation is that even
-with surgery and a shift of the bacterial community these specific
-bacteria still persisted within the colon. An alternative reason would
-be that even if they were reduced in relative abundance they were still
-present. Another potential explanation is that they were not present in
-enough individuals to be able to classify those with and without disease
-with a high degree of accuracy. It is possible that our Random Forest
-models were able to get the same information from measures such as FIT
-or other OTUs. Finally, it is also possible that all three of these
-potential explanations could have played a role. Regardless, our
-observations would suggest that an individual's resident bacteria have a
-large role to play in disease initiation and could change in a way that
-allows predictive models to lower the positive probability of a lesion
-after surgery {[}Figure 4{]}. It should be stated that our study does
-not argue against the importance of these CRC associated bacteria in the
+predictive within our models. There are a number of reasons why this may
+have occured. First, one potential explanation is that even with surgery
+and a shift of the bacterial community these specific bacteria still
+persisted within the colon. Second, the bacteria even if they were
+reduced in relative abundance they were still present within the gut.
+Third, is that they were not present in enough individuals to be able to
+classify those with and without disease with a high degree of accuracy.
+Fourth, is that it is possible that our Random Forest models were able
+to get the same information from measures such as FIT or other OTUs.
+Finally, it is also possible that all three of these potential
+explanations could have played a role. Regardless, our observations
+would suggest that an individual's resident bacteria have a large role
+to play in disease initiation and could change in a way that allows
+predictive models to lower the positive probability of a lesion after
+surgery {[}Figure 4{]}. It should be noted that our study does not argue
+against the importance of these CRC associated bacteria in the
 pathogenesis of disease but rather that the models do not utilize these
-specific bacteria for classification purposes (lesion or initial and
-follow up). In fact, it is possible that these CRC associated bacteria
-are important in the transition from adenoma to carcinoma and would be
-one explanation as to why in our data we not only see high initial
-relative abundances, in certain individuals, but also large decreases in
-relative abundance in those with carcinoma but not in those with adenoma
-after surgery {[}Figure 2{]}.
+specific bacteria for classification purposes (lesion or before sample).
+In fact, it is possible that these CRC associated bacteria are important
+in the transition from adenoma to carcinoma and would be one explanation
+as to why in our data we not only see high initial relative abundances,
+in certain individuals, but also large decreases in relative abundance
+in those with carcinoma but not in those with adenoma after surgery
+{[}Figure 2{]}.
 
@@
 were diagnosed as no longer having a lesion. Another limitation is that
-we do not know if adding modern tests such as the fecal DNA test
+we do not know if adding modern tests such as the stool DNA test
 {[}20{]} could help improve our overall AUC. Another limitation is that
@@
 
-Another interesting outcome was that Within figure 3 the initial and
-follow up model showed better test AUC results then the training set
-AUC. This may have occured because the training AUC that was determined
-from 20 repeated 10 fold cross validation removed samples at random and
-did not take into account that they were matched samples. Another
-potential reason is that the model itself may be overfit since the total
-number of samples was not that large. However, the lesion model did not
-suffer from these discrepencies and similar conclusions can be drawn
-solely from this model. Regardless, further independent studies will
-need to be carried out to verify our findings since not only are we
-dealing with feces, which could be very different than the communities
-present on the actual tissue, but also are dealing with correlations
-that may not be representative of the true pathogensis of disease.
+Another interesting outcome was that Within figure 3 the before sample
+model showed better test AUC results then the training set AUC. This may
+have occured because the training AUC that was determined from 20
+repeated 10 fold cross validation removed samples at random and did not
+take into account that they were matched samples. Another potential
+reason is that the model itself may be overfit since the total number of
+samples was not that large. However, the lesion model did not suffer
+from these discrepencies and similar conclusions can be drawn solely
+from this model. Regardless, further independent studies will need to be
+carried out to verify our findings since not only are we dealing with
+feces, which could be very different than the communities present on the
+actual tissue, but also are dealing with correlations that may not be
+representative of the true pathogensis of disease.
 

### 2017-01-18 submission/manuscript_outline_20161024.Rmd

@@
 Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.
-This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, nor any addition of new or depressed bacterium but rather that a shift of the bacterial community as a whole occurs with the existing members of this microbiome.
+This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, or any addition of new or depressed bacterium.
+Instead it may be due to a shift of the bacterial community as a whole between existing members of the host's microbiome.
 
-We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or changing due to surgery.
+We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.
 We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.
-With only the lesion model having a single OTU from previously associated cancer bacteria (*Porphyromonas asaccharolytica*).
-Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
-Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
+With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
+Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
+Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
 
@@
 Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
-One potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.
-An alternative reason would be that even if they were reduced in relative abundance they were still present.
-Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
-It is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
+There are a number of reasons why this may have occured.
+First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.
+Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.
+Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
+Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
 Finally, it is also possible that all three of these potential explanations could have played a role.
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
-It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).
+It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).
 In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].
@@
 This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
-Another limitation is that we do not know if adding modern tests such as the fecal DNA test [@cotter_long-term_2016] could help improve our overall AUC.
+Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
 Another limitation is that this study drew heavily from those with caucasian ancestry.
@@
 
-Another interesting outcome was that Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.
+Another interesting outcome was that Within figure 3 the before sample model showed better test AUC results then the training set AUC.
 This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.

### 2017-01-19 submission/manuscript_outline_20161024.Rmd
streamline -discussion

@@
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].
-Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
+The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].
+Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
 This study was approved by the University of Michigan Institutional Review Board.
@@
 In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
-The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.
+The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.
 Each group was randomly assigned to avoid biases based on sample collection location.
@@
 The model included data on FIT and the bacterial microbiome.
-Non-binary data was checked for near zero variance.
-Data columns that had near zero variance were removed.
+Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
-Optimization of the mtry hyperparameter involved taking the samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.
-This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).
+Optimization of the mtry hyperparameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.
+Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyperparameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
 This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
-Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which testing on the `r length(rownames(good_metaf))`-person cohort was completed.
+Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
-The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created.
+The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.
 
-***Initial Follow Up Model Creation***
+***Before Sample Model Creation***
 We also investigated whether a model could be created that could identify before and after surgery samples.
-The main difference was that only the `r length(rownames(good_metaf))`-person cohort follow up data set was used at all stages of model building and classification.
+The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
 Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.
+Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.
 
 ***Selection of Important OTUs***
-In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.
-This final collated list of variables was what was considered the most important for the lesion or initial follow up models.
+In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.
+This final collated list of variables was what was considered the most important for the lesion or before sample models.
 
@@
 Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
-Unless otherwise stated the P-values reported are those of that were BH corrected.
+Unless otherwise stated the P-values reported are those that were BH corrected.
 
@@
 Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.
-From here, all OTUs that were used in the lesion model were then analyzed using a paired wilcoxson test.
-We then investigated the relative abundance of specific previously CRC associated bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.
+From here, all OTUs that were used in either model were then analyzed using a paired wilcoxson test.
+We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.
 We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
@@
 The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.
-We then used the initial follow up model to assess whether it could classify samples better then the lesion model.
-The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.
+We then used the before sample model to assess whether it could classify samples better then the lesion model.
+The most important OTUs were then identified from this model and used to create a reduced feature before sample model.
 This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.
@@
 
-***Reproducible methods.***
+***Reproducible Methods.***
 A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
-Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005).
+Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
 

### 2017-01-19 submission/manuscript_outline_20161024.tex
streamline -discussion

@@
 collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial microbiome has been shown to nomralize within this time
+The bacterial microbiome has been shown to normalize within this time
 period {[}21{]}. Our follow up data set had a total of 67 individuals
-that not only had a sample as described previoulsy but also a follow up
-sample between 188 - 546 days after surgery and treatment had been
-completed. This study was approved by the University of Michigan
-Institutional Review Board. All study participants provided informed
-consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+that not only had a sample as described but also a follow up sample
+between 188 - 546 days after surgery and treatment had been completed.
+This study was approved by the University of Michigan Institutional
+Review Board. All study participants provided informed consent and the
+study itself conformed to the guidelines set out by the Helsinki
+Declaration.
 
@@
 (Eppendorf). The V4 variable region was amplified and the resulting
-product was split between three sequencing runs with control, adenoma,
+product was split between three sequencing runs with normal, adenoma,
 and carcinoma evenly represented on each run. Each group was randomly
@@
 FIT and the bacterial microbiome. Non-binary data was checked for near
-zero variance. Data columns that had near zero variance were removed.
-This pre-processing was performed with the R package caret (v6.0.73).
+zero variance and OTUs that had near zero variance were removed. This
+pre-processing was performed with the R package caret (v6.0.73).
 Optimization of the mtry hyperparameter involved taking the samples and
-making 100 80/20 (train/test) splits in the data where control and
-lesion were equally represented in the 80 and 20 split, respectively.
-This 80\% portion was then split again into an 80/20 split, and run
-through 20 repeated 10-fold cross validations to optimzie the model's
-AUC (Area Under the Curve of the Receiver Operator Characteristic). This
-resulting model was then tested on the 20\% of the data that was
-originally held out from this overall process. Once the ideal mtry was
-found the entire 490 sample set was used to create the final Random
-Forest model on which testing on the 67-person cohort was completed. The
-default cutoff of 0.5 was used as the threshold to classify individuals
-as positive or negative for lesion. The hyperparameter, mtry, defines
-the number of variables to investigate at each split before a new
-division of the data is created.
+making 100 different 80/20 (train/test) splits of the data where normal
+and lesion were represented in the same proportion within both the whole
+data set and the 80/20 split. Each of these splits were then run through
+20 repeated 10-fold cross validations to optimize the mtry
+hyperparameter by maximizing the AUC (Area Under the Curve of the
+Receiver Operator Characteristic). This resulting model was then tested
+on the 20\% of the data that was originally held out from this overall
+process. Once the ideal mtry was found the entire 490 sample set was
+used to create the final Random Forest model on which classifications on
+the 67-person cohort was completed. The default cutoff of 0.5 was used
+as the threshold to classify individuals as positive or negative for
+lesion. The hyperparameter, mtry, defines the number of variables to
+investigate at each split before a new division of the data is created
+with the Random Forest model.
 
-\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
+\textbf{\emph{Before Sample Model Creation}} We also investigated
 whether a model could be created that could identify before and after
 surgery samples. The main difference was that only the 67-person cohort
-follow up data set was used at all stages of model building and
-classification. Other than this difference the creation of this model
-and optimization of the mtry hyperparameter was completed using the same
-procedure that was used to create the lesion model.
+was used at all stages of model building and classification. Other than
+this difference the creation of this model and optimization of the mtry
+hyperparameter was completed using the same procedure that was used to
+create the lesion model. Instead of classifying samples as positive or
+negative of lesion this model classified samples as positive or negative
+for being a before surgery sample.
 
@@
 times a variable was present in the top 10\% of mean decrease in
-accuracy (MDA) for each different 80/20 split model and then filtered
-this list to variables that were only present more than 50\% of the
-time. This final collated list of variables was what was considered the
-most important for the lesion or initial follow up models.
+accuracy (MDA) for each of the 100 different 80/20 split models and then
+filtered this list to variables that were only present more than 50\% of
+the time. This final collated list of variables was what was considered
+the most important for the lesion or before sample models.
 
@@
 corrected P-value of less than 0.05 was considered significant. Unless
-otherwise stated the P-values reported are those of that were BH
-corrected.
+otherwise stated the P-values reported are those that were BH corrected.
 
@@
 in FIT between initial and follow ups for either adenoma or carcinoma
-were investigated. From here, all OTUs that were used in the lesion
-model were then analyzed using a paired wilcoxson test. We then
-investigated the relative abundance of specific previously CRC
-associated bacteria, specifically, OTUs that taxonomically classified to
-\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
-\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
-stomatis}. We wanted to test if there were any differences based on
-whether the individual had an adenoma or carcinoma. From here the lesion
-model was then tested for accuracy in prediction and whether it reduced
-the positive probability of lesion after surgery. The most important
-OTUs for this were used to build a reduced model and it was assessed for
-similarity to the original model. We then used the initial follow up
-model to assess whether it could classify samples better then the lesion
-model. The most important OTUs were then identified from this model and
-used to create a reduced feature initial follow up model. This reduced
-feature model, as was done with the lesion model, was compared to the
-full model for loss of accuracy. Finally, a list of common OTUs were
-found for the two different models used.
+were investigated. From here, all OTUs that were used in either model
+were then analyzed using a paired wilcoxson test. We then investigated
+the relative abundance of specific previously associated CRC bacteria,
+specifically, OTUs that taxonomically classified to \emph{Fusobacterium
+nucleatum}, \emph{Parvimonas micra}, \emph{Peptostreptococcus
+assacharolytica}, and \emph{Porphyromonas stomatis}. We wanted to test
+if there were any differences based on whether the individual had an
+adenoma or carcinoma. From here the lesion model was then tested for
+accuracy in prediction and whether it reduced the positive probability
+of lesion after surgery. The most important OTUs for this were used to
+build a reduced model and it was assessed for similarity to the original
+model. We then used the before sample model to assess whether it could
+classify samples better then the lesion model. The most important OTUs
+were then identified from this model and used to create a reduced
+feature before sample model. This reduced feature model, as was done
+with the lesion model, was compared to the full model for loss of
+accuracy. Finally, a list of common OTUs were found for the two
+different models used.
 
-\textbf{\emph{Reproducible methods.}} A detailed and reproducible
+\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
 description of how the data were processed and analyzed can be found at
 \url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
-have been deposited into the NCBI Sequence Read Archive (SRP062005).
+have been deposited into the NCBI Sequence Read Archive (SRP062005) and
+the necessary metadata can be found at
+\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
+respective SRA study accession.
 

### 2017-01-19 submission/manuscript_outline_20161024.tex
change model names in figs from init follow to before sample

@@
 \textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
-for lesion and Initial and Follow up Models.} The shaded areas
-represents the range of values of a 100 different 80/20 splits of the
-test set data using either all variables (grey) or reduced variable
-(red) models. The blue line represents the reduced variable model using
-100\% of the data set. A) Lesion model. B) Initial/Follow up model
+for lesion and Before Sample Models.} The shaded areas represents the
+range of values of a 100 different 80/20 splits of the test set data
+using either all variables (grey) or reduced variable (red) models. The
+blue line represents the reduced variable model using 100\% of the data
+set. A) Lesion model. B) Defore sample model
 
 \textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Initial Follow Up Reduced Variable Models} A)
-Lesion positive probability adjustment of those with carcinoma from
-initial to follow up sample B) Initial follow up positive probability
-adjustment of those with carcinoma from initial to follow up sample C)
-Lesion positive probability adjustment of those with adenoma as well as
-those with SRN and the probability adjustment from initial to follow up
-sample. D) Initial follow up positive probability adjustment of those
-with adenoma as well as those with SRN and the probability adjustment
-from initial to follow up sample. The dotted line represents the
-threshold used to make the decision of whether a sample was positive or
-not.
+Results for Lesion and Before Sample Reduced Variable Models} A) Lesion
+positive probability adjustment of those with carcinoma from initial to
+follow up sample B) Initial follow up positive probability adjustment of
+those with carcinoma from initial to follow up sample C) Lesion positive
+probability adjustment of those with adenoma as well as those with SRN
+and the probability adjustment from initial to follow up sample. D)
+Initial follow up positive probability adjustment of those with adenoma
+as well as those with SRN and the probability adjustment from initial to
+follow up sample. The dotted line represents the threshold used to make
+the decision of whether a sample was positive or not.
 
@@
 
-\textbf{Figure S3: Summary of Important Variables in Initial Follow Up
+\textbf{Figure S3: Summary of Important Variables in Before Sample
 Model} A) MDA of the most important variables in the lesion model. The
@@
 \textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Initial and Follow Up Full Variable Models} A)
-Lesion positive probability adjustment of those with carcinoma from
-initial to follow up sample B) Initial follow up positive probability
-adjustment of those with carcinoma from initial to follow up sample C)
-Lesion positive probability adjustment of those with adenoma as well as
-those with SRN and the probability adjustment from initial to follow up
-sample. D) Initial follow up positive probability adjustment of those
-with adenoma as well as those with SRN and the probability adjustment
-from initial to follow up sample. The dotted line represents the
-threshold used to make the decision of whether a sample was positive or
-not.
+Results for Lesion and Before Sample Full Variable Models} A) Lesion
+positive probability adjustment of those with carcinoma from initial to
+follow up sample B) Initial follow up positive probability adjustment of
+those with carcinoma from initial to follow up sample C) Lesion positive
+probability adjustment of those with adenoma as well as those with SRN
+and the probability adjustment from initial to follow up sample. D)
+Initial follow up positive probability adjustment of those with adenoma
+as well as those with SRN and the probability adjustment from initial to
+follow up sample. The dotted line represents the threshold used to make
+the decision of whether a sample was positive or not.
 

### 2017-01-19 submission/manuscript_outline_20161024.Rmd

@@
 
-**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Initial and Follow up Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
+**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Before Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
 The blue line represents the reduced variable model using 100% of the data set.
 A) Lesion model.
-B) Initial/Follow up model
+B) Defore sample model
 
-**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Follow Up Reduced Variable Models**  A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Reduced Variable Models**  A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
 D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
@@
 
-**Figure S3: Summary of Important Variables in Initial Follow Up Model** A) MDA of the most important variables in the lesion model.
+**Figure S3: Summary of Important Variables in Before Sample Model** A) MDA of the most important variables in the lesion model.
 The black point represents the median and the different colors are the different runs up to 100.
@@
 
-**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial and Follow Up Full Variable Models** A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
 D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.

### 2017-01-20 submission/manuscript_outline_20161024.Rmd
add second SRA accession

@@
 A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
-Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
+Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
 

### 2017-01-20 submission/manuscript_outline_20161024.Rmd
abstract improvements

@@
 A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-The after surgery samples had a significant decrease in the model positive probability of a lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+There was a significant decrease in the model positive probability for the after surgery sample being either a lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
 In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).

### 2017-01-20 submission/manuscript_outline_20161024.tex
add second SRA accession
abstract improvements

@@
 model to classify samples as before surgery had 33 variables (AUC = 0.79
-- 0.651). The after surgery samples had a significant decrease in the
-model positive probability of a lesion or before surgery sample (P-value
-= 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were common
-to both models and the majority of these classified to commensal
-bacteria (Bacteroides, Blautia, Streptococcus, and Clostridiales).\\
+- 0.651). There was a significant decrease in the model positive
+probability for the after surgery sample being either a lesion or before
+surgery sample (P-value = 1.91e-11 and 6.72e-12). In total there were 14
+OTUs that were common to both models and the majority of these
+classified to commensal bacteria (Bacteroides, Blautia, Streptococcus,
+and Clostridiales).\\
 \textbf{Conclusions:} Our data suggests that treatment not only
@@
 \url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
-have been deposited into the NCBI Sequence Read Archive (SRP062005) and
-the necessary metadata can be found at
+have been deposited into the NCBI Sequence Read Archive (SRP062005 and
+SRP096978) and the necessary metadata can be found at
 \url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the

### 2017-01-20 submission/manuscript_outline_20161024.Rmd
clean up text edits

@@
 With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
-Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
+Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].
 Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
@@
 Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
-Finally, it is also possible that all three of these potential explanations could have played a role.
+Finally, it is also possible that all of these potential explanations could have played a role.
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
@@
 
-Another interesting outcome was that Within figure 3 the before sample model showed better test AUC results then the training set AUC.
+Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.
 This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.

### 2017-01-20 submission/manuscript_outline_20161024.tex
clean up text edits

@@
 (lesion and before sample) had a significantly decreased positive
-probability of either lesion or being an initial sample on follow up
+probability of either lesion or being a before sample on follow up
 {[}Figure 4 \& S4{]}. Further confirmation of the importance of the
@@
 to get the same information from measures such as FIT or other OTUs.
-Finally, it is also possible that all three of these potential
-explanations could have played a role. Regardless, our observations
-would suggest that an individual's resident bacteria have a large role
-to play in disease initiation and could change in a way that allows
-predictive models to lower the positive probability of a lesion after
-surgery {[}Figure 4{]}. It should be noted that our study does not argue
-against the importance of these CRC associated bacteria in the
-pathogenesis of disease but rather that the models do not utilize these
-specific bacteria for classification purposes (lesion or before sample).
-In fact, it is possible that these CRC associated bacteria are important
-in the transition from adenoma to carcinoma and would be one explanation
-as to why in our data we not only see high initial relative abundances,
-in certain individuals, but also large decreases in relative abundance
-in those with carcinoma but not in those with adenoma after surgery
+Finally, it is also possible that all of these potential explanations
+could have played a role. Regardless, our observations would suggest
+that an individual's resident bacteria have a large role to play in
+disease initiation and could change in a way that allows predictive
+models to lower the positive probability of a lesion after surgery
+{[}Figure 4{]}. It should be noted that our study does not argue against
+the importance of these CRC associated bacteria in the pathogenesis of
+disease but rather that the models do not utilize these specific
+bacteria for classification purposes (lesion or before sample). In fact,
+it is possible that these CRC associated bacteria are important in the
+transition from adenoma to carcinoma and would be one explanation as to
+why in our data we not only see high initial relative abundances, in
+certain individuals, but also large decreases in relative abundance in
+those with carcinoma but not in those with adenoma after surgery
 {[}Figure 2{]}.
@@
 
-Another interesting outcome was that Within figure 3 the before sample
+Another interesting outcome was that within figure 3 the before sample
 model showed better test AUC results then the training set AUC. This may

### 2017-01-25 submission/manuscript_outline_20161024.Rmd
correct error with BH calculation

@@
 source('../code/functions.R')
-source('../code/graphFunctions.R')
 
@@
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].
+First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples.
 Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].

### 2017-01-25 submission/manuscript_outline_20161024.tex
correct error with BH calculation

@@
 samples showed a significant difference between initial and follow up
-samples for \emph{Peptostreptococcus stomatis} (P-value = 0.0183) and
-\emph{Porphyromonas asaccharolytica} (P-value = 0.0154) whereas there
-were no signficant differences in any of these OTUs in the adenoma
-samples {[}Table S4{]}. Second, when these OTUs were present, there was
-a clear magnitude difference based on whether they were from adenoma or
-carcinoma {[}Figure 2{]}. However, only a small percentage of those with
-adenoma or carcinoma were positive for any of these OTUs.
+samples for \emph{Peptostreptococcus stomatis} (P-value = 0.00913),
+\emph{Parvimonas micra} (P-value = 0.0185), and \emph{Porphyromonas
+asaccharolytica} (P-value = 0.00913) whereas there were no signficant
+differences in any of these OTUs in the adenoma samples. Second, when
+these OTUs were present, there was a clear magnitude difference based on
+whether they were from adenoma or carcinoma {[}Figure 2{]}. However,
+only a small percentage of those with adenoma or carcinoma were positive
+for any of these OTUs.
 
@@
 Interestingly, when only previously associated cancer bacteria were
-investigated only 2 out of the 4 had a significant decrease in relative
+investigated only 3 out of the 4 had a significant decrease in relative
 abundance between initial and follow up for the carcinoma group and 0
@@
 was a significant difference in initial and follow up sample for the
-OTUs classfied as \emph{Peptostreptococcus stomatis} (P-value = 0.0183)
-and \emph{Porphyromonas asaccharolytica} (P-value = 0.0154). B) Adenoma
+OTUs classfied as \emph{Peptostreptococcus stomatis} (P-value = 0.00913)
+and \emph{Porphyromonas asaccharolytica} (P-value = 0.00913). B) Adenoma
 initial and follow up samples. There were no significant differences

### 2017-01-25 code/Run_potential_cancer_specific_OTUs.R
add BH comment

@@ -75,7 +75,7 @@ crc_select_data <- select(shared, Group, sampleType, one_of(good_counts_init)) %
 # Write out table for future use
 write.csv(crc_select_data, "results/tables/adn_crc_maybe_diff.csv", row.names = F)
 
-# Run statistics testing
+# Run statistics testing and BH correction
 test_data <- select(shared, Group, sampleType, Dx_Bin, one_of(good_counts_init))
 
 pvalue_summary <- cbind(

### 2017-01-25 code/Run_Supplemental_time_table.R
use rownames for Rmd clarity

@@ -29,11 +29,12 @@ summary_statistics <- bind_rows(
   c(filter(difference_table_treatment, dx == "adenoma") %>% 
       select(distance, time) %>% summarise_each(funs(mean, sd))), 
   c(filter(difference_table_treatment, dx == "cancer") %>% 
-      select(distance, time) %>% summarize_each(funs(mean, sd)))) %>% 
-  mutate(dx = c("adenoma", "cancer"))
+      select(distance, time) %>% summarize_each(funs(mean, sd)))) 
 
 colnames(summary_statistics) <- c(
-  "mean_thetayc_change", "mean_days", "sd_thetayc_change", "sd_days", "dx")
+  "mean_thetayc_change", "mean_days", "sd_thetayc_change", "sd_days")
+
+rownames(summary_statistics) <- c("adenoma", "carcinoma")
 
 pvalues_summary <- rbind(
   c("thetayc_change", wilcox.test(distance ~ dx, 
@@ -44,7 +45,7 @@ pvalues_summary <- rbind(
 colnames(pvalues_summary) <- c("test_values", "uncorrected_p_value")
 
 write.csv(pvalues_summary, "results/tables/time_pvalues.csv", row.names = F)
-write.csv(summary_statistics, "results/tables/time_summary_data.csv", row.names = F)
+write.csv(summary_statistics, "results/tables/time_summary_data.csv")
 write.csv(difference_table_treatment, "results/tables/time_datatable.csv", 
   row.names = F)
 

### 2017-01-25 submission/manuscript_outline_20161024.Rmd
update time analysis

@@
 
-loadLibs(c("dplyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
+loadLibs(c("dplyr", "reshape2", "scales", "knitr", "rmarkdown"))
 
@@
 time_pvalues <- read.csv("../results/tables/time_pvalues.csv", header = T, row.names = 1)
+time_summary <- read.csv("../results/tables/time_summary_data.csv", header = T, row.names = 1)
 all_OTU_pair_wilcox <- read.csv("../results/tables/OTU_paired_wilcoxson_test.csv", header = T)
@@
 Although the before sample model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
-Time elapsed between the collection of the follow up sample from the initial sample, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
+There was a significant difference in time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
 

### 2017-01-25 submission/manuscript_outline_20161024.tex
update time analysis

@@
 probability for those treated with chemotherapy (P-value = 7.04e-04 and
-5.07e-03). Time elapsed between the collection of the follow up sample
-from the initial sample, did not have a significant difference between
-adenoma and carcinoma (uncorrected P-value = 0.784).
+5.07e-03). There was a significant difference in time elapsed in the
+collection of the follow up sample between adenoma and carcinoma samples
+(uncorrected P-value = 7.59e-05) with time passed being lower for
+adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102 days).
 

### 2017-01-25 submission/manuscript_outline_20161024.tex
update with time imporvements

@@
 cross validation but also having over 100 different 80/20 splits to try
-and mimic the type of variation that might be expected to occur.
+and mimic the type of variation that might be expected to occur. The
+time difference in collection of sample between adenoma and carcinoma
+that was observed could have affected our observed results. However, our
+random forest based model, specifically the lesion trained model, were
+still able to adjust positive probabilities for both adenoma and
+carcinoma to similar amounts despite this.
 

### 2017-01-25 code/Run_Combine_Testing_pull_imp_OTUs.R
remove unneded graphfunctions.R

@@ -14,12 +14,10 @@
 
 ## Cutoff of 0.5 (default was used for this) for RF model
 
-
 ###Load needed Libraries and functions
 source('code/functions.R')
 
 loadLibs(c("dplyr", "reshape2", "scales", "caret", "pROC"))
-
 
 # Set up relevent environment variables
 imp_vars_list <- list()

### 2017-01-25 code/Run_Figure3.R

@@ -6,7 +6,6 @@
 
 ###Load needed Libraries and functions
 source('code/functions.R')
-source('code/graphFunctions.R')
 
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", 
            "wesanderson", "caret", "pROC"))

### 2017-01-25 code/Run_Figure4.R

@@ -6,7 +6,6 @@
 
 ###Load needed Libraries and functions
 source('code/functions.R')
-source('code/graphFunctions.R')
 
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
            "gridExtra", "scales", "wesanderson", "caret"))

### 2017-01-25 code/Run_IF_best_model.R

@@ -4,7 +4,6 @@
 
 #Load needed libraries
 source('code/functions.R')
-source('code/graphFunctions.R')
 
 loadLibs(c("dplyr", "caret","scales", "wesanderson", "randomForest", "pROC"))
 

### 2017-01-25 code/Run_combine_IF_aggregate_model.R

@@ -5,7 +5,6 @@
 
 ###Load needed Libraries and functions
 source('code/functions.R')
-source('code/graphFunctions.R')
 
 loadLibs(c("dplyr", "ggplot2", "reshape2", "gridExtra", "scales", 
            "wesanderson", "caret", "pROC"))

### 2017-01-25 code/Run_reduce_feature_IF_model.R

@@ -5,7 +5,6 @@
 
 #Load needed libraries
 source('code/functions.R')
-source('code/graphFunctions.R')
 
 loadLibs(c("randomForest", "dplyr", "ggplot2", "reshape2", 
            "gridExtra", "scales", "wesanderson", "caret", "doMC"))

### 2017-01-25 code/old/Run_Get_Imp_OTUs_noFIT.R

@@ -6,6 +6,7 @@
 
 ###Load needed Libraries and functions
 source('code/functions.R')
+source('code/graphFunctions.R')
 
 loadLibs(c("randomForest", "dplyr", "ggplot2", "reshape2", 
   "gridExtra", "scales", "wesanderson", "caret", "doMC"))
@@ -31,7 +32,12 @@ shared <- read.delim('data/process/final.0.03.subsample.shared',
 good_metaf_select <- data_frame(
   sample = c(good_metaf$initial, good_metaf$followUp)) %>% 
   mutate(lesion = c(good_metaf$lesion, good_metaf$lesion_follow), 
-         fit_result = c(good_metaf$fit_result, good_metaf$fit_followUp))
+         Hx_Prev = rep(good_metaf$Hx_Prev, 2), 
+         Hx_Fam_CRC = rep(good_metaf$Hx_Fam_CRC, 2), 
+         White = rep(good_metaf$White, 2), 
+         BMI = rep(good_metaf$BMI, 2), 
+         Age = rep(good_metaf$Age, 2), 
+         Gender = rep(good_metaf$Gender, 2))
 
 test_data <- inner_join(good_metaf_select, shared, 
   by = c("sample" = "Group"))
@@ -41,23 +47,27 @@ rm(shared)
 # Filter and use only specific data
 
 test_data <- test_data %>% 
-  select(sample, lesion, fit_result, contains("Otu0")) %>% 
-  mutate(lesion = c(rep(1, 67), rep(0, 67)))
+  select(sample, lesion, Hx_Prev, Hx_Fam_CRC, White, 
+    BMI, Age, Gender, contains("Otu0")) %>% 
+  mutate(lesion = factor(lesion), Gender = factor(Gender)) %>% 
+  mutate(Hx_Prev = factor(Hx_Prev), Hx_Fam_CRC = factor(Hx_Fam_CRC), 
+    White = factor(White))
 
 #Filter out rows that are not all complete
 
 test_data <- test_data[complete.cases(test_data), ]
 
 # Need to create dummy variables for those that are factor classes
-#temp_dummy <- dummyVars(~lesion,  data = test_data)
-#useable_dummy_vars <- as.data.frame(predict(temp_dummy, 
-#  newdata = test_data))
+temp_dummy <- dummyVars(~lesion + Gender + Hx_Prev + Hx_Fam_CRC + White, 
+  data = test_data)
+useable_dummy_vars <- as.data.frame(predict(temp_dummy, 
+  newdata = test_data))
 
[…]

### 2017-01-25 code/reference_run_RF.R

@@ -5,9 +5,8 @@
 
 #Load needed libraries
 source('code/functions.R')
-source('code/graphFunctions.R')
 
-loadLibs(c("dplyr", "caret","scales", "wesanderson", "doMC"))
+loadLibs(c("dplyr", "caret","scales", "doMC"))
 
 load("exploratory/RF_model_setup.RData")
 
@@ -21,10 +20,6 @@ i = 1
 #               Model Training and Parameter Tuning                             #
 #                                                                               #
 #################################################################################
-
-# Create the tune grid to be used
-#tunegrid <- expand.grid(
- # .mtry=c(1:441), .ntree=c(250, 500, 1000, 1500, 2000))
 
 # Call number of processors to use
 registerDoMC(cores = 4)

### 2017-01-25 code/Run_Get_Imp_OTUs_noFIT.R
move old code

@@ -1,252 +1,0 @@
-### Try to identify the important OTUs 
-### Specifically try and identify which OTUs are involved
-### with the decrease in positive probability between intial and follow up
-## Marc Sze
-
-
-###Load needed Libraries and functions
-source('code/functions.R')
-source('code/graphFunctions.R')
-
-loadLibs(c("randomForest", "dplyr", "ggplot2", "reshape2", 
-  "gridExtra", "scales", "wesanderson", "caret", "doMC"))
-
-# Read in data tables
-good_metaf <- read.csv(
-  "results/tables/mod_metadata/good_metaf_final.csv", 
-  stringsAsFactors = F, header = T) %>% 
-  mutate(lesion_follow = ifelse(Disease_Free == "n", 1, 0))
-
-shared <- read.delim('data/process/final.0.03.subsample.shared', 
-  header=T, sep='\t') %>% select(-label, -numOtus)
-
-
-#################################################################################
-#                                                                               #
-#                                                                               #
-#               Data Clean up for better and faster modeling                    #
-#                                                                               #
-#################################################################################
-
-#Create an insync version to merge with filtered_shared
-good_metaf_select <- data_frame(
-  sample = c(good_metaf$initial, good_metaf$followUp)) %>% 
-  mutate(lesion = c(good_metaf$lesion, good_metaf$lesion_follow), 
-         Hx_Prev = rep(good_metaf$Hx_Prev, 2), 
-         Hx_Fam_CRC = rep(good_metaf$Hx_Fam_CRC, 2), 
-         White = rep(good_metaf$White, 2), 
-         BMI = rep(good_metaf$BMI, 2), 
-         Age = rep(good_metaf$Age, 2), 
-         Gender = rep(good_metaf$Gender, 2))
-
-test_data <- inner_join(good_metaf_select, shared, 
-  by = c("sample" = "Group"))
-
-rm(shared)
-
-# Filter and use only specific data
-
-test_data <- test_data %>% 
-  select(sample, lesion, Hx_Prev, Hx_Fam_CRC, White, 
-    BMI, Age, Gender, contains("Otu0")) %>% 
-  mutate(lesion = factor(lesion), Gender = factor(Gender)) %>% 
-  mutate(Hx_Prev = factor(Hx_Prev), Hx_Fam_CRC = factor(Hx_Fam_CRC), 
-    White = factor(White))
-
-#Filter out rows that are not all complete
-
-test_data <- test_data[complete.cases(test_data), ]
-
-# Need to create dummy variables for those that are factor classes
[…]

### 2017-01-25 submission/manuscript_outline_20161024.Rmd

@@
 We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
+The time difference in collection of sample between adenoma and carcinoma that was observed could have affected our observed results.
+However, our random forest based model, specifically the lesion trained model, were still able to adjust positive probabilities for both adenoma and carcinoma to similar amounts despite this.
 

### 2017-01-25 submission/supplemental_outline_20161024.Rmd
align font and text to manuscript

@@
 bibliography:
-csl: mbio.csl
+csl: microbiome.csl
 output: 
@@
     keep_tex: true
+    includes:
+      in_header: header.tex
 fontsize: 12pt

### 2017-02-06 submission/manuscript_outline_20161024.Rmd

@@
 Their change in FIT was also significantly different then those with carcinoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
+There was no significant difference in any individual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
 A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
@@
 This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].
-Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to reccurence.
-Concurrently, there has also been a lot of interest in new areas that could have a role in disease pathogenesis, such as the gut bacterial microbiome.
+Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to recurrence.
+Concurrently, there has also been a lot of interest in new areas that could have a role in disease parthenogenesis, such as the gut bacterial microbiome.
 
@@
 There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
-A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
-These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC.
+A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the parthenogenesis of CRC.
+These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
 However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.
 
-In this study we investigated what happened to the bacterial microbiome before and after surgery for indivdiuals with either adenoma or carcinoma.
-Our anlaysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
+In this study we investigated what happened to the bacterial microbiome before and after surgery for individuals with either adenoma or carcinoma.
+Our analysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
 We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed before (initial) and after (follow up) surgery.
@@
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples.
+First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no significant differences in any of these OTUs in the adenoma samples.
 Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].
@@
 There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].
-Yet based on thetayc there was a significant difference between inital and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.
+Yet based on thetayc there was a significant difference between initial and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.
 The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].
@@
 We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.
-We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.
+We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to commensal bacteria.
 With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
 Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].
-Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
+Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
 
-There was no difference for the majority of models tested for differences in positive probablity based on whether chemotherapy or radiation was received.
+There was no difference for the majority of models tested for differences in positive probability based on whether chemotherapy or radiation was received.
 There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.
@@
 Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.
-This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.
+This could suggest that the bacterial community is one of the first components that could change during the parthenogenesis of disease.
 These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].
@@
 Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
-There are a number of reasons why this may have occured.
+There are a number of reasons why this may have occurred.
 First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.
@@
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
-It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).
+It should be noted that our study does not argue against the importance of these CRC associated bacteria in the parthenogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).
 In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].
@@
 Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
-Another limitation is that this study drew heavily from those with caucasian ancestry.
+Another limitation is that this study drew heavily from those with Caucasian ancestry.
 The results may not be immediately representative of those with either Asian or African ancestry.
-Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.
+Finally, although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be immediately extrapolate-able to other populations.
 We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
@@
 Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.
-This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
-Another potential reason is that the model itself may be overfit since the total number of samples was not that large.
-However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.
-Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
+This may have occurred because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
+Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.
+However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.
+Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true parthenogenesis of disease.
 
@@
 The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
-The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.
+The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.
 
@@
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
-Optimization of the mtry hyperparameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.
-Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyperparameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
+Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.
+Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
 This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
@@
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
-The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.
+The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.
 
@@
 The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
-Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.
+Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure that was used to create the lesion model.
 Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.
@@
 ***Statistical Analysis***
-The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.
+The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.
 Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
-Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired wilcoxson ranked sum test.
+Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.
 Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
@@
 Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.
-From here, all OTUs that were used in either model were then analyzed using a paired wilcoxson test.
+From here, all OTUs that were used in either model were then analyzed using a paired Wilcoxson test.
 We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.
@@
 B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-C) NMDS of the intial and follow up samples for the Adenoma group.
+C) NMDS of the initial and follow up samples for the Adenoma group.
 D) NMDS of the initial and follow up samples for the Carcinoma group.
@@
 **Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
-There was a significant difference in initial and follow up sample for the OTUs classfied as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+There was a significant difference in initial and follow up sample for the OTUs classified as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
 B) Adenoma initial and follow up samples.
@@
 A) Lesion model.
-B) Defore sample model
+B) Before sample model
 
@@
 
-All authors declare that they do not have any relevent competing interests to report.
+All authors declare that they do not have any relevant competing interests to report.
 

### 2017-02-07 submission/manuscript_outline_20161024.Rmd

@@
 
-**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.
-Previous research suggests a link between fecal bacterial microbiome and CRC.
-The overall objective of our study was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e.
+**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.
+The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after surgery in patients with lesions (i.e.
 adenoma or carcinoma).
-Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.
-We also wanted to investigate the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models which could classify individuals as having a lesion or as before surgery.
-**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).
-Their change in FIT was also significantly different then those with carcinoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-There was no significant difference in any individual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-There was a significant decrease in the model positive probability for the after surgery sample being either a lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+Specifically, we wanted to identify what within the community was different before and after surgical removal of said lesion.
+
+**Results:** The bacterial microbiome in the pre and post surgery samples of the individuals with adenoma were similar while those with carcinoma were not (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).
+There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+The post surgery sample for both models had a decrease in the positive probability for either lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
 In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).
-**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion within our models but also causes detectable changes in the bacterial microbiome.
-Further surveillance of these individuals will enable us to determine whether models such as the ones we present could be used to predict recurrence of colorectal cancer.
+**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.
+With individuals with carcinoma having more drastic changes to the overall community then those with adenoma.
+However, changes to commensal bacteria were common to individuals with either adenoma or carcinoma suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.
 
@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; detection; risk factors
+bacterial microbiome; colorectal cancer; polyps; FIT; post surgery; risk factors
 

### 2017-02-07 submission/manuscript_outline_20161024.Rmd
make intro more about follow ups

@@
 However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].
-It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007].
-The vast majority of CRCs are not inherited and the exact etiology of the disease has not been well worked out [@haggar_colorectal_2009].
-Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been an additional increase in the incidence of CRC in the younger population.
 
-This increased incidence of CRC in the younger population is concerning since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].
-This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].
-Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to recurrence.
-Concurrently, there has also been a lot of interest in new areas that could have a role in disease parthenogenesis, such as the gut bacterial microbiome.
+Recently, there has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
+There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
+A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
+These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
+Yet there remains limited information on the bacterial microbiome before and after surgery for removal of lesion (adenoma or carcinoma) and whether the community changes at all.
 
-There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
-A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the parthenogenesis of CRC.
-These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
-However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.
-
-In this study we investigated what happened to the bacterial microbiome before and after surgery for individuals with either adenoma or carcinoma.
-Our analysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed before (initial) and after (follow up) surgery.
+In this study we tested the hypothesis of whether that the bacterial microbiome changes after surgery for individuals with a lesion.
+Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
+We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed pre (initial) and post (follow up) surgery.
 We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
@@
 Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.
-This could suggest that the bacterial community is one of the first components that could change during the parthenogenesis of disease.
+This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.
 These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].
@@
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
-It should be noted that our study does not argue against the importance of these CRC associated bacteria in the parthenogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).
+It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).
 In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].
@@
 However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.
-Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true parthenogenesis of disease.
+Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.
 

### 2017-02-07 submission/manuscript_outline_20161024.Rmd
explain CRC associated OTUs better -Results

@@
 ***Bacterial Community and FIT***
-Based on the thetayc distance metric, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
-There was also a significant difference in change in FIT between initial and follow up samples (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-The whole community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].
-There was also no significant difference between initial and follow up samples for any single OTU [Figure S1].
+We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.
+What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
+We also found that there were larger changes fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
+The broad shift in bacterial community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric [Table S1].
+We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lession, adenoma only, or carcinoma only [Figure S1].
 
 ***Previously Associated Cancer Bacteria***
-We next examined whether there were differences in previously well described carcinoma associated OTUs.
+Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].
+So we next examined whether there were differences in previously well described carcinoma associated OTUs.
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no significant differences in any of these OTUs in the adenoma samples.
-Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].
-However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.
+There was a difference in initial and follow up samples for lesion and carcinoma for relative abundance of *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+In contrast, there was no differences relative abundance in any of these OTUs for those with adenoma [Figure 2].
+We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these OTUs [Figure 2].
 

### 2017-02-07 submission/manuscript_outline_20161024.Rmd

@@
 ***Full and Reduced Model***
-Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria; we next investigated if we could create models that could adequately classify and adjust either lesion or before sample probability based on the bacterial community and FIT.
-The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the before sample model which had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
-Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and before sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-
+We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma.
+In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre versus post samples based on the bacterial community and FIT.
+The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while the initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
+By identifying the most important variables for each respective model and then reducing them to only these factors we were able to increase the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
 The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].
 The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].
-Interestingly, the test set AUC range for the before sample model performed much better then the training set AUCs.
-Both the full and reduced before sample models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.
+Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.
+Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final model used.
 
-***Most Important Variables***
-The reduced models were built based on the most important variables to the respective full model.
+***Common OTUs to both Models***
+The reduced models were built based on the most important variables to either classification model.
 For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the before sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
 For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].
+There were a total of `r length(common_vars_summary$otu)` OTUs that were common to both models.
+Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.
+The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
 
@@
 However, the lesion model correctly kept the one individual who still had a carcinoma on follow up above the cut off threshold [Figure 4a & S4a] for a positive call while the before sample models did not [Figure 4b & S4b].
-
-***Common OTUs to both Models***
-There were a total of `r length(common_vars_summary$otu)` OTUs that were common to both models.
-Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.
-The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
 

### 2017-02-09 submission/manuscript_outline_20161024.Rmd
better flow for results

@@
 The reduced models were built based on the most important variables to either classification model.
-For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the before sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
+For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
 For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].
@@
 ***Positive Probability after Lesion Removal***
-Regardless of model used there was a significant decrease in the positive probability of either the sample being lesion or a before sample on follow up [Figure 4 & S4] (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+If there were common OTUs for individuals with adenoma and carcinoma that were different versus normal controls we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or a before sample.
+This is what we observed regardless of model used (lesion or before sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
 
-For the full and reduced lesion model there was a significant difference in the classification for the lesion model between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)`, respectively) but not for the before sample model (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)`).
-However, the lesion model correctly kept the one individual who still had a carcinoma on follow up above the cut off threshold [Figure 4a & S4a] for a positive call while the before sample models did not [Figure 4b & S4b].
+Although there was a decrease in positive probability for all models there were differences on whether the models classified the samples as still having lesion.
+This was most evident for the full and reduced lesion model where there was a large difference between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)`, respectively).
+However, the before sample model did not have this disparity between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)`).
+Even though the lesion model varied greatly between predicted and actual it was able to correctly keep the one individual who still had a carcinoma on follow up above the cut off threshold [Figure 4a & S4a] for a positive call while the before sample models did not [Figure 4b & S4b].
 
 ***Treatment and Time Differences***
+After observing these changes in positive probability We wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
 There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).
-Although the before sample model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
-There was a significant difference in time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
+Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
+Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.
+Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
 

### 2017-02-09 submission/manuscript_outline_20161024.Rmd
discussion improvements pt1

@@
 What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
-We also found that there were larger changes fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-The broad shift in bacterial community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
+We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
+The broad shift in bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
 Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric [Table S1].
-We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lession, adenoma only, or carcinoma only [Figure S1].
+When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric tested [Table S1].
+We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lesion, adenoma, or carcinoma [Figure S1].
 
@@
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-There was a difference in initial and follow up samples for lesion and carcinoma for relative abundance of *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
-In contrast, there was no differences relative abundance in any of these OTUs for those with adenoma [Figure 2].
-We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these OTUs [Figure 2].
+There was a difference in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2].
+We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these respective OTUs [Figure 2].
 
@@
 By identifying the most important variables for each respective model and then reducing them to only these factors we were able to increase the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].
+The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap with each other [Figure 3a].
 The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].
 Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.
-Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final model used.
+Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final before sample model used.
 
@@
 ***Treatment and Time Differences***
-After observing these changes in positive probability We wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
+After observing these changes in positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
 There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).
@@
 ## Discussion
-There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].
-Yet based on thetayc there was a significant difference between initial and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.
-The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].
-There was also a significant difference in the overall bacterial community structure for those in the carcinoma group [Figure 1d] before and after surgery but not for the adenoma group [Figure 1c].
-Investigation of the OTU relative abundance before and after surgery found no single OTU to be significantly different [Figure S1].
+From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
+There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for indivdiuals with carcinoma versus adenoma [Figure 1].
+However, there were no differences between initial and follow up samples for Shannon Diversity, oberseved OTUs, or evenness regardless of whether the indidivual had an adenoma or carcinoma [Table S1].
+There was alos no differences in relative abundnace of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].
 

### 2017-02-12 submission/manuscript_outline_20161024.Rmd
incorporate updated SRA and corrected RF runs into manuscript

@@
 There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
 The post surgery sample for both models had a decrease in the positive probability for either lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`, and `r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`).
 **Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.
@@
 For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].
-There were a total of `r length(common_vars_summary$otu)` OTUs that were common to both models.
-Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.
+There were a total of `r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`, and `r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`.
 The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
@@
 Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
-Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.
+Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible con-founder.
 Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
@@
 From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
-There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for indivdiuals with carcinoma versus adenoma [Figure 1].
-However, there were no differences between initial and follow up samples for Shannon Diversity, oberseved OTUs, or evenness regardless of whether the indidivual had an adenoma or carcinoma [Table S1].
-There was alos no differences in relative abundnace of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].
+There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
+However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].
+There was also no differences in relative abundance of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].
 
-Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.
-This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, or any addition of new or depressed bacterium.
-Instead it may be due to a shift of the bacterial community as a whole between existing members of the host's microbiome.
+Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested cancer causing microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
+This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.
 
-We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.
-We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to commensal bacteria.
+We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.
+What we found was that the commonly associated CRC bacteria were not higly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
 With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
-Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].
+Using only these important OTUs and FIT both models (lesion and before sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
 Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
 
+*Problematic paragraph need to amend*
+
 There was no difference for the majority of models tested for differences in positive probability based on whether chemotherapy or radiation was received.
-There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.
+There was a difference in the length of time between initial and follow up sample between adenoma and carcinoma.
 These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.

### 2017-02-12 submission/manuscript_outline_20161024.tex
incorporate updated SRA and corrected RF runs into manuscript

@@
 \textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
-health problem with early detection being used as a key component in
-mitigating deaths due to the disease. Previous research suggests a link
-between fecal bacterial microbiome and CRC. The overall objective of our
-study was to investigate the changes in the bacterial microbiome after
-surgery in patients with lesion (i.e.~adenoma or carcinoma).
-Specifically, we wanted to identify what within the community was
-different within those undergoing surgical removal of lesion. We also
-wanted to investigate the use of the bacterial microbiome and Fecal
-Immunoglobulin Test (FIT) to build models which could classify
-individuals as having a lesion or as before surgery.\\
-\textbf{Results:} Adenoma individual's bacterial microbiome were more
-similar to their pre-surgery sample then those with carcinoma (P-value =
-0.00198). Their change in FIT was also significantly different then
-those with carcinoma (P-value = 2.15e-05). There was no significant
-difference in any indivdiual OTU between samples before and after
-surgery (P-value \textgreater{} 0.125). A model with a total of 37
-variables was able to classify lesion (AUC = 0.847 - 0.791) while the
-model to classify samples as before surgery had 33 variables (AUC = 0.79
-- 0.651). There was a significant decrease in the model positive
-probability for the after surgery sample being either a lesion or before
-surgery sample (P-value = 1.91e-11 and 6.72e-12). In total there were 14
-OTUs that were common to both models and the majority of these
-classified to commensal bacteria (Bacteroides, Blautia, Streptococcus,
-and Clostridiales).\\
-\textbf{Conclusions:} Our data suggests that treatment not only
-significantly reduces the probability of having a colonic lesion within
-our models but also causes detectable changes in the bacterial
-microbiome. Further surveillance of these individuals will enable us to
-determine whether models such as the ones we present could be used to
-predict recurrence of colorectal cancer.
+health problem with previous research suggesting that a link may exist
+between the fecal bacterial microbiome and CRC. The overall objective of
+our study was to test the hypothesis that changes in the bacterial
+microbiome occur after surgery in patients with lesions (i.e.~adenoma or
+carcinoma). Specifically, we wanted to identify what within the
+community was different before and after surgical removal of said
+lesion.
+
+\textbf{Results:} The bacterial microbiome in the pre and post surgery
+samples of the individuals with adenoma were similar while those with
+carcinoma were not (P-value = 0.000472). There was no difference in the
+relative abundance of any OTU between the pre and post surgery samples
+(P-value \textgreater{} 0.151). A model with a total of 53 variables was
+able to classify lesion (AUC = 0.811 - 0.866) while a model built to
+classify samples as before surgery had 70 variables (AUC = 0.641 -
+0.805). The post surgery sample for both models had a decrease in the
+positive probability for either lesion or before surgery sample (P-value
+= 3.28e-04 and 6.71e-12). In total there were 23 OTUs that were common
+to both models and the majority of these classified to commensal
+bacteria (Bacteroides, Clostridiales, Blautia, and Ruminococcaceae).\\
+\textbf{Conclusions:} Our data supports the hypothesis that there are
+differences in the bacterial microbiome between pre and post surgical
+samples. With individuals with carcinoma having more drastic changes to
+the overall community then those with adenoma. However, changes to
+commensal bacteria were common to individuals with either adenoma or
+carcinoma suggesting that these bacteria may be central to initial polyp
+formation and transition to carcinoma.
 
@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; detection; risk
+bacterial microbiome; colorectal cancer; polyps; FIT; post surgery; risk
 factors
@@
 in screening {[}1{]}. However, despite this improvement there are still
-approximately 50,000 deaths from the disease a year {[}2{]}. It is
-estimated that around 5-10\% of all CRCs can be explained by autosomal
-dominant inheritance {[}3{]}. The vast majority of CRCs are not
-inherited and the exact etiology of the disease has not been well worked
-out {[}2{]}. Although many risk factors have been identified {[}2{]} and
-non-invasive screening techniques have started to be put into consistent
-use {[}4,5{]} there has been an additional increase in the incidence of
-CRC in the younger population.
+approximately 50,000 deaths from the disease a year {[}2{]}.
 
-This increased incidence of CRC in the younger population is concerning
-since having either an adenoma or carcinoma increases ones risk for
-future adenomas or carcinomas {[}6--8{]}. This increased risk can also
-carry with it an increased chance of mortality due to this recurrence
-{[}9,10{]}. Therefore, there has been a great amount of interest in
-early risk stratification tools {[}11,12{]} that can help identify those
-that may be at most susceptibility to reccurence. Concurrently, there
-has also been a lot of interest in new areas that could have a role in
-disease pathogenesis, such as the gut bacterial microbiome.
+Recently, there has been promising work on the bacterial microbiome and
+it's ability to be able to complement existing screening methods such as
+Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
+{[}3,4{]}. There has also been research into how this microbiome could
+be altered directly on tumor tissue itself {[}5{]}. A few studies have
+also shown how this microbiome {[}6{]} or specific members within it
+{[}7{]} could be directly involved with the pathogenesis of CRC. These
+studies have helped to provide a tantalizing link between the bacterial
+microbiome and CRC. Yet there remains limited information on the
+bacterial microbiome before and after surgery for removal of lesion
+(adenoma or carcinoma) and whether the community changes at all.
 
-There has been promising work on the bacterial microbiome and it's
-ability to be able to complement existing screening methods such as
-Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
-{[}13,14{]}. There has also been research into how this microbiome could
-be altered directly on tumor tissue itself {[}15{]}. A few studies have
-now even shown how this microbiome {[}16{]} or specific members within
-it {[}17{]} could be directly involved with the pathogenesis of CRC.
-These studies have helped to provide a tantilzing link between the
-bacterial microbiome and CRC. However, at this present time there
-remains limited information on the bacterial microbiome before and after
-successful surgery for removal of the adenoma or carcinoma and whether
-it changes at all.
-
-In this study we investigated what happened to the bacterial microbiome
-before and after surgery for indivdiuals with either adenoma or
-carcinoma. Our anlaysis includes both alpha and beta diversity analysis
-along with investigation of individual operational taxonomic units
-(OTUs). We also utilized Random Forest models and observed how these
-models as well as specific OTUs within them performed before (initial)
-and after (follow up) surgery. We also used these models to look for
-similar important OTUs to identify the crucial OTUs for not only
-classifying initial and follow up samples but also lesion or normal.
+In this study we tested the hypothesis of whether that the bacterial
+microbiome changes after surgery for individuals with a lesion. Our
+analysis included both alpha and beta diversity analysis along with
+investigation of individual operational taxonomic units (OTUs). We also
+utilized Random Forest models and observed how these models as well as
+specific OTUs within them performed pre (initial) and post (follow up)
+surgery. We also used these models to look for similar important OTUs to
+identify the crucial OTUs for not only classifying initial and follow up
+samples but also lesion or normal.
 
@@
 
-\textbf{\emph{Bacterial Community and FIT}} Based on the thetayc
-distance metric, comparing the initial to the follow up samples, there
-was a significant difference between the adenoma and carcinoma groups
-(P-value = 0.00198) {[}Figure 1a{]}. There was also a significant
-difference in change in FIT between initial and follow up samples
-(P-value = 2.15e-05) {[}Figure 1b{]}. The whole community structure
-before and after surgery was visualized on NMDS graphs for both adenoma
-{[}Figure1c{]} (PERMANOVA = 0.997) and carcinoma {[}Figure 1d{]}
-(PERMANOVA = 0.002). When all initial and follow up samples were
-compared to each other there was no significant overall difference
-between them (PERMANOVA = 0.085). There was no significant difference
-between initial and follow up samples for observed OTUs, Shannon
-diversity, or evenness after correction for multiple comparisons
-{[}Table S1{]}. There was also no significant difference between initial
-and follow up samples for any single OTU {[}Figure S1{]}.
+\textbf{\emph{Bacterial Community and FIT}} We first wanted to test
+whether there were any broad differences between initial and follow up
+samples based on lesion being either adenoma or carcinoma. What we found
+was that the bacterial community in those with carcinoma were more
+dissimilar (as measured by thetayc) to their initial sample then those
+with adenoma (P-value = 0.000472) {[}Figure 1a{]}. We also found that
+there were larger changes in fecal blood (measured by FIT) for those
+with carcinoma versus adenoma (P-value = 2.15e-05) {[}Figure 1b{]}. The
+broad shift in bacterial community structure before and after surgery
+was visualized using NMDS for both adenoma {[}Figure1c{]} (PERMANOVA =
+0.999) and carcinoma {[}Figure 1d{]} (PERMANOVA = 0.005). Interestingly,
+when initial and follow up samples were compared to each other,
+regardless of whether they were adenoma or carcinoma (lesion), there was
+no significant overall difference between them (PERMANOVA = 0.09). When
+investigating more broad alpha diversity metrics there was no difference
+found between initial and follow up samples for lesion, adenoma, or
+carcinoma for any metric tested {[}Table S1{]}. We also observed that
+there was no difference in OTU relative abundance between initial and
+follow up samples for lesion, adenoma, or carcinoma {[}Figure S1{]}.
 
-\textbf{\emph{Previously Associated Cancer Bacteria}} We next examined
-whether there were differences in previously well described carcinoma
-associated OTUs. These included the OTUs that aligned with
-\emph{Porphyromonas asaccharolytica} (Otu000153), \emph{Fusobacterium
-nucleatum} (Otu000226), \emph{Parvimonas micra} (Otu000460), and
-\emph{Peptostreptococcus stomatis} (Otu000653). First, the carcinoma
-samples showed a significant difference between initial and follow up
-samples for \emph{Peptostreptococcus stomatis} (P-value = 0.00913),
-\emph{Parvimonas micra} (P-value = 0.0185), and \emph{Porphyromonas
-asaccharolytica} (P-value = 0.00913) whereas there were no signficant
-differences in any of these OTUs in the adenoma samples. Second, when
-these OTUs were present, there was a clear magnitude difference based on
-whether they were from adenoma or carcinoma {[}Figure 2{]}. However,
-only a small percentage of those with adenoma or carcinoma were positive
-for any of these OTUs.
+\textbf{\emph{Previously Associated Cancer Bacteria}} Previous
+literature has suggested that a number of oral microbes may be important
+in CRC pathogenesis {[}8{]}. So we next examined whether there were
+differences in previously well described carcinoma associated OTUs.
+These included the OTUs that aligned with \emph{Porphyromonas
+asaccharolytica} (Otu000202), \emph{Fusobacterium nucleatum}
+(Otu000442), \emph{Parvimonas micra} (Otu001273), and
+\emph{Peptostreptococcus stomatis} (Otu001682). There was a difference
+in relative abundance in initial and follow up samples for lesion and
+carcinoma for \emph{Parvimonas micra} (P-value = 0.0116), and
+\emph{Porphyromonas asaccharolytica} (P-value = 0.00842). In contrast,
+there was no difference in relative abundance in any of these OTUs for
+those with adenoma {[}Figure 2{]}. We also observed that only a small
+percentage of those with adenoma or carcinoma were positive or had an
+appreciable relative abundance of these respective OTUs {[}Figure 2{]}.
 
-\textbf{\emph{Full and Reduced Model}} Since differences were observed
-between initial and follow up samples and only a small number of
-individuals were positive for previously associated CRC bacteria; we
-next investigated if we could create models that could adequately
-classify and adjust either lesion or before sample probability based on
-the bacterial community and FIT. The lesion model had an AUC range of
-0.723 to 0.795 versus the before sample model which had an AUC range of
-0.451 to 0.67 after 100 iterations of 20 repeated 10-fold cross
-validations. Interestingly, identification of the most important
-variables and reducing the models to only these factors increased the
-AUC in the lesion model (0.791 - 0.847) and before sample model (0.651 -
-0.79).
-
+\textbf{\emph{Full and Reduced Model}} We next wanted to identify if
+there were any common bacterial microbiome changes in individuals with
+adenoma or carcinoma. In order to investigate this we created two
+different models: one to classify lesion versus normal and one to
+classify pre versus post samples based on the bacterial community and
+FIT. The lesion model had an AUC range of 0.73 to 0.797 while the
+initial sample model had an AUC range of 0.485 to 0.686 after 100
+iterations of 20 repeated 10-fold cross validations. By identifying the
+most important variables for each respective model and then reducing
+them to only these factors we were able to increase the AUC in the
+lesion model (0.811 - 0.866) and initial sample model (0.641 - 0.805).\\
 The test set AUC range for the full and reduced lesion model were
 similar to that reported for the training set AUC ranges and the ROC
-curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
-final lesion model used was within the range of both the full and
+curve ranges overlap with each other {[}Figure 3a{]}. The ROC curve for
+the final lesion model used was within the range of both the full and
 reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
-range for the before sample model performed much better then the
-training set AUCs. Both the full and reduced before sample models
-overlapped with each other {[}Figure 3b{]} there was a marked decrease
-in the ROC curve for the final model used.
+range for the initial sample model performed much better then the
+training set AUCs. Both the full and reduced initial sample models
+overlapped with each other {[}Figure 3b{]} but there was a marked
+decrease in the ROC curve for the final before sample model used.
 
-\textbf{\emph{Most Important Variables}} The reduced models were built
-based on the most important variables to the respective full model. For
-the lesion model there were a total of 37 variables {[}Figure S2{]}
-whereas for the before sample model there were a total of 33 variables
+\textbf{\emph{Common OTUs to both Models}} The reduced models were built
+based on the most important variables to either classification model.
+For the lesion model there were a total of 53 variables {[}Figure S2{]}
+whereas for the initial sample model there were a total of 70 variables
 {[}Figure S3{]}. For both models FIT resulted in the largest decrease in
-MDA {[}Figure S2a \& S3a{]}.
+MDA {[}Figure S2a \& S3a{]}. There were a total of Bacteroides,
+Clostridiales, Blautia, and Ruminococcaceae. The majority of these OTUs
+had classifications to bacteria typically thought of as commensal
+{[}Table S2{]}.
 
-\textbf{\emph{Positive Probability after Lesion Removal}} Regardless of
-model used there was a significant decrease in the positive probability
-of either the sample being lesion or a before sample on follow up
-{[}Figure 4 \& S4{]} (full lesion P-value = 1.11e-11, reduced lesion
-P-value = 1.91e-11, initial and follow up P-value = 6.71e-12, and
-reduced initial and follow up P-value = 6.72e-12).
+\textbf{\emph{Positive Probability after Lesion Removal}} If there were
+common OTUs for individuals with adenoma and carcinoma that were
+different versus normal controls we would expect to find a decrease in
+the positive probability of the follow up sample to be either lesion or
+a before sample. This is what we observed regardless of model used
+(lesion or before sample) or whether it was built on the full or reduced
+variable data set {[}Figure 4 \& S4{]} (full lesion P-value = 1.62e-04,
+reduced lesion P-value = 3.28e-04, initial and follow up P-value =
+6.71e-12, and reduced initial and follow up P-value = 6.71e-12).
 
-For the full and reduced lesion model there was a significant difference
-in the classification for the lesion model between predicted and actual
-(P-value = 4.19e-10 and 6.98e-10, respectively) but not for the before
-sample model (P-value = 1.00 and 1.00). However, the lesion model
-correctly kept the one individual who still had a carcinoma on follow up
-above the cut off threshold {[}Figure 4a \& S4a{]} for a positive call
-while the before sample models did not {[}Figure 4b \& S4b{]}.
+Although there was a decrease in positive probability for all models
+there were differences on whether the models classified the samples as
+still having lesion. This was most evident for the full and reduced
+lesion model where there was a large difference between predicted and
+actual (P-value = 4.88e-01 and 4.7e-01, respectively). However, the
+before sample model did not have this disparity between predicted and
+actual (P-value = 1.00 and 1.00). Even though the lesion model varied
+greatly between predicted and actual it was able to correctly keep the
+one individual who still had a carcinoma on follow up above the cut off
+threshold {[}Figure 4a \& S4a{]} for a positive call while the before
+sample models did not {[}Figure 4b \& S4b{]}.
 
-\textbf{\emph{Common OTUs to both Models}} There were a total of 14 OTUs
-that were common to both models. Of these OTUs the most common taxonomic
-identifications were to Blautia, Bacteroides, Streptococcus, and
-Clostridiales. The majority of these OTUs had classifications to
-bacteria typically thought of as commensal {[}Table S2{]}.
-
-\textbf{\emph{Treatment and Time Differences}} There was no difference
-in the amount of change in positive probability for either the full or
-reduced lesion model for either chemotherapy (P-value = 0.821 and 0.821)
-or radiation therapy (P-value = 0.69 and 0.981). Although the before
-sample model was similar there was a significant decrease in positive
-probability for those treated with chemotherapy (P-value = 7.04e-04 and
-5.07e-03). There was a significant difference in time elapsed in the
+\textbf{\emph{Treatment and Time Differences}} After observing these
+changes in positive probability we wanted to assess whether additional
+treatments, such as chemotherapy and radiation, could have an impact on
+the results that we observed. There was no difference in the amount of
+change in positive probability for either the full or reduced lesion
+model for either chemotherapy (P-value = 0.184 and 0.184) or radiation
+therapy (P-value = 0.942 and 0.839). Although the before sample model
+was similar there was a significant difference in decreased positive
+probability for those treated with chemotherapy (P-value = 1.27e-03 and
+6.18e-04). Finally, we wanted to know if the length of time between the
+initial and follow up sample could be a possible con-founder. Within our
+study there was a significant difference in the time elapsed in the
 collection of the follow up sample between adenoma and carcinoma samples
@@
 
-There was no difference in alpha diversity metrics between the initial
-and follow up samples {[}Table S1{]}. Yet based on thetayc there was a
-significant difference between inital and follow up samples {[}Figure
-1a{]} with carcinoma samples being more dissimilar to their initial
-samples then adenoma samples. The change in FIT was also different
-between initial and follow up for those with adenoma or carcinoma
-{[}Figure 1b{]}. There was also a significant difference in the overall
-bacterial community structure for those in the carcinoma group {[}Figure
-1d{]} before and after surgery but not for the adenoma group {[}Figure
-1c{]}. Investigation of the OTU relative abundance before and after
-surgery found no single OTU to be significantly different {[}Figure
-S1{]}.
+From our results there were some large observed differences in the
+bacterial microbiome between pre and post surgery samples based on
+whether the individual had an adenoma or carcinoma. There was much
+larger difference between initial and follow up samples based on the
+thetayc distance metric and in fecal blood as measured by FIT for
+individuals with carcinoma versus adenoma {[}Figure 1{]}. However, there
+were no differences between initial and follow up samples for Shannon
+Diversity, observed OTUs, or evenness regardless of whether the
+individual had an adenoma or carcinoma {[}Table S1{]}. There was also no
+differences in relative abundance of any specific OTU for lesion,
+adenoma, or carcinoma {[}Figure s1{]}.
 
-Interestingly, when only previously associated cancer bacteria were
-investigated only 3 out of the 4 had a significant decrease in relative
-abundance between initial and follow up for the carcinoma group and 0
-out of the 4 were significant for the adenoma group. This data suggests
-that the changes that may be important in the bacterial microbiome after
-surgery are not necessarily any one specific bacterium, a clear
-depression of CRC related bacteria, or any addition of new or depressed
-bacterium. Instead it may be due to a shift of the bacterial community
-as a whole between existing members of the host's microbiome.
+Although there were no differences when investigating all OTUs, when
+looking specifically at four OTUs that taxonomically classified to
+previously suggested cancer causing microbes we found that only 3/4 had
+a decrease in relative abundance between initial and follow up for those
+with carcinoma and 0/4 had differences for those with adenoma. This data
+would suggest that these specific OTUs may be important in the
+transition of an adenoma to a carcinoma but less so in the initiation of
+an adenoma from benign tissue.
 
@@
 microbiome to either be able to classify lesions (adenoma or carcinoma)
-or before samples in order to confirm that the community was what was
-responding or changing due to surgery. We found that the OTUs that made
-up the most important variables to the model overwhelmingly belonged to
-comensal bacteria. With only the lesion model having a single OTU from a
-previously associated cancer bacterium (\emph{Porphyromonas
-asaccharolytica}). Using only these important OTUs and FIT both models
-(lesion and before sample) had a significantly decreased positive
-probability of either lesion or being a before sample on follow up
-{[}Figure 4 \& S4{]}. Further confirmation of the importance of the
-changes of comensal bacteria to these classifications was that a total
-of 14 OTUs were common to both models and the vast majority belonged to
-regular residents of our gut community.
+or initial samples in order to find common OTUs in the community that
+change for both adenoma and carcinoma. What we found was that the
+commonly associated CRC bacteria were not higly represented within our
+models but rather that OTUs that made up the most important variables
+overwhelmingly belonged to commensal bacteria. With only the lesion
+model having a single OTU from a previously associated cancer bacterium
+(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
+and FIT both models (lesion and before sample) significantly decreased
+positive probability of either lesion or being an initial sample on
+follow up {[}Figure 4 \& S4{]}. Further confirmation of the importance
+of the changes of commensal bacteria to these classifications was that a
+total of 23 OTUs were common to both models and the vast majority
+belonged to regular residents of our gut community.
+
+\emph{Problematic paragraph need to amend}
 
 There was no difference for the majority of models tested for
-differences in positive probablity based on whether chemotherapy or
-radiation was received. There was also no difference in the thetayc
-distance metric based on length of time between initial and follow up
-sample between adenoma and carcinoma. These results would indicate that
-the findings described were specific to the surgical intervention and
-that differences observed between carcinoma and adenoma samples can not
-be simply attributed to collection time between samples.
+differences in positive probability based on whether chemotherapy or
+radiation was received. There was a difference in the length of time
+between initial and follow up sample between adenoma and carcinoma.
+These results would indicate that the findings described were specific
+to the surgical intervention and that differences observed between
+carcinoma and adenoma samples can not be simply attributed to collection
+time between samples.
 
 This study builds upon previous work from numerous labs that have looked
-into the bacterial microbiome as a potential screening tool {[}13,14{]}
-by exploring what happens to the bacterial community after surgical
-removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
-it may not be surprising to have E.coli as one of the most important
-OTUs and one that was common to both models. Interestingly, many of the
-most important OTUs had taxonomic identification for resident gut
-microbes. This could suggest that the bacterial community is one of the
-first components that could change during the pathogenesis of disease.
-These bacterial microbiome changes could be the first step in allowing
-more inflammatory bacterium to gain a foothold within the colon
-{[}19{]}.
+into the bacterial microbiome as a potential screening tool {[}3,4{]} by
+exploring what happens to the bacterial community after surgical removal
+of a lesion. Based on previous work by Arthur, et al. {[}9{]} it may not
+be surprising to have E.coli as one of the most important OTUs and one
+that was common to both models. Interestingly, many of the most
+important OTUs had taxonomic identification for resident gut microbes.
+This could suggest that the bacterial community is one of the first
+components that could change during the pathogenesis of disease. These
+bacterial microbiome changes could be the first step in allowing more
+inflammatory bacterium to gain a foothold within the colon {[}8{]}.
 
@@
 predictive within our models. There are a number of reasons why this may
-have occured. First, one potential explanation is that even with surgery
-and a shift of the bacterial community these specific bacteria still
-persisted within the colon. Second, the bacteria even if they were
+have occurred. First, one potential explanation is that even with
+surgery and a shift of the bacterial community these specific bacteria
+still persisted within the colon. Second, the bacteria even if they were
 reduced in relative abundance they were still present within the gut.
@@
 we do not know if adding modern tests such as the stool DNA test
-{[}20{]} could help improve our overall AUC. Another limitation is that
-this study drew heavily from those with caucasian ancestry. The results
+{[}10{]} could help improve our overall AUC. Another limitation is that
+this study drew heavily from those with Caucasian ancestry. The results
 may not be immediately representative of those with either Asian or
 African ancestry. Finally, although our training and test set are
-relatively large we still run the risk of overfitting or having a model
-that may not be immediately extrapolateable to other populations. We've
+relatively large we still run the risk of over-fitting or having a model
+that may not be immediately extrapolate-able to other populations. We've
 done our best to safeguard against this by not only running 10-fold
@@
 model showed better test AUC results then the training set AUC. This may
-have occured because the training AUC that was determined from 20
+have occurred because the training AUC that was determined from 20
 repeated 10 fold cross validation removed samples at random and did not
 take into account that they were matched samples. Another potential
-reason is that the model itself may be overfit since the total number of
-samples was not that large. However, the lesion model did not suffer
-from these discrepencies and similar conclusions can be drawn solely
+reason is that the model itself may be over-fit since the total number
+of samples was not that large. However, the lesion model did not suffer
+from these discrepancies and similar conclusions can be drawn solely
 from this model. Regardless, further independent studies will need to be
@@
 actual tissue, but also are dealing with correlations that may not be
-representative of the true pathogensis of disease.
+representative of the true pathogenesis of disease.
 
@@
 design of the study was similar to that reported in Baxter, et al
-{[}13{]}. In brief, study exclusion involved those who had already
+{[}3{]}. In brief, study exclusion involved those who had already
 undergone surgery, radiation, or chemotherapy, had colorectal cancer
@@
 The bacterial microbiome has been shown to normalize within this time
-period {[}21{]}. Our follow up data set had a total of 67 individuals
+period {[}11{]}. Our follow up data set had a total of 67 individuals
 that not only had a sample as described but also a follow up sample
@@
 previously published using both OC FIT-CHEK and OC-Auto Micro 80
-automated system (Polymedco Inc.) {[}22{]}. 16S rRNA gene sequencing was
-completed as previously described by Kozich, et al. {[}23{]}. In brief,
+automated system (Polymedco Inc.) {[}12{]}. 16S rRNA gene sequencing was
+completed as previously described by Kozich, et al. {[}13{]}. In brief,
 DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described {[}23{]}. The general processing workflow
+has been previously described {[}13{]}. The general processing workflow
 using mothur was as follows: Paired-end reads were first merged into
@@
 Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
-neighbor clustering algorithm. The numer of sequences for each sample
-was rarified to 10521 in an attempt to minimize uneven sampling.
+neighbor clustering algorithm. The number of sequences for each sample
+was rarefied to 10523 in an attempt to minimize uneven sampling.
 
-\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
+\textbf{\emph{Lesion Model Creation}} The Random Forest {[}14{]}
 algorithm was used to create the model used for prediction of lesion
@@
 pre-processing was performed with the R package caret (v6.0.73).
-Optimization of the mtry hyperparameter involved taking the samples and
+Optimization of the mtry hyper-parameter involved taking the samples and
 making 100 different 80/20 (train/test) splits of the data where normal
@@
 20 repeated 10-fold cross validations to optimize the mtry
-hyperparameter by maximizing the AUC (Area Under the Curve of the
+hyper-parameter by maximizing the AUC (Area Under the Curve of the
 Receiver Operator Characteristic). This resulting model was then tested
@@
 as the threshold to classify individuals as positive or negative for
-lesion. The hyperparameter, mtry, defines the number of variables to
+lesion. The hyper-parameter, mtry, defines the number of variables to
 investigate at each split before a new division of the data is created
@@
 this difference the creation of this model and optimization of the mtry
-hyperparameter was completed using the same procedure that was used to
+hyper-parameter was completed using the same procedure that was used to
 create the lesion model. Instead of classifying samples as positive or
@@
 \textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
-used for all statisitical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}25{]} in the vegan package
+used for all statistical analysis. Comparisons between bacterial
+community structure utilized PERMANOVA {[}15{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
 differences between initial and follow up samples utilized a paired
-wilcoxson ranked sum test. Where multiple comparison testing was needed
-a Benjamini-Hochberg (BH) correction was applied {[}26{]} and a
+Wilcoxson ranked sum test. Where multiple comparison testing was needed
+a Benjamini-Hochberg (BH) correction was applied {[}16{]} and a
 corrected P-value of less than 0.05 was considered significant. Unless
@@
 were investigated. From here, all OTUs that were used in either model
-were then analyzed using a paired wilcoxson test. We then investigated
+were then analyzed using a paired Wilcoxson test. We then investigated
 the relative abundance of specific previously associated CRC bacteria,
@@
 Group.} A) A significant difference was found between the adenoma and
-carcinoma group for thetayc (P-value = 0.00198). B) A significant
+carcinoma group for thetayc (P-value = 0.000472). B) A significant
 difference was found between the adenoma and carcinoma group for change
-in FIT (P-value = 2.15e-05). C) NMDS of the intial and follow up samples
-for the Adenoma group. D) NMDS of the initial and follow up samples for
-the Carcinoma group. For C) and D) the teal represents initial samples
-and the pink represents follow up samples.
+in FIT (P-value = 2.15e-05). C) NMDS of the initial and follow up
+samples for the Adenoma group. D) NMDS of the initial and follow up
+samples for the Carcinoma group. For C) and D) the teal represents
+initial samples and the pink represents follow up samples.
 
@@
 was a significant difference in initial and follow up sample for the
-OTUs classfied as \emph{Peptostreptococcus stomatis} (P-value = 0.00913)
-and \emph{Porphyromonas asaccharolytica} (P-value = 0.00913). B) Adenoma
+OTUs classified as \emph{Peptostreptococcus stomatis} (P-value = 0.0496)
+and \emph{Porphyromonas asaccharolytica} (P-value = 0.00842). B) Adenoma
 initial and follow up samples. There were no significant differences
@@
 blue line represents the reduced variable model using 100\% of the data
-set. A) Lesion model. B) Defore sample model
+set. A) Lesion model. B) Before sample model
 
@@
 
-All authors declare that they do not have any relevent competing
+All authors declare that they do not have any relevant competing
 interests to report.
@@
 
-\hypertarget{ref-green_very_2007}{}
-3. Green RC, Green JS, Buehler SK, Robb JD, Daftary D, Gallinger S, et
-al. Very high incidence of familial colorectal cancer in Newfoundland: A
-comparison with Ontario and 13 other population-based studies. Familial
-Cancer. 2007;6:53--62.
-
-\hypertarget{ref-liao_application_2013}{}
-4. Liao C-S, Lin Y-M, Chang H-C, Chen Y-H, Chong L-W, Chen C-H, et al.
-Application of quantitative estimates of fecal hemoglobin concentration
-for risk prediction of colorectal neoplasia. World Journal of
-Gastroenterology. 2013;19:8366--72.
-
-\hypertarget{ref-johnson_multi-target_2016}{}
-5. Johnson DH, Kisiel JB, Burger KN, Mahoney DW, Devens ME, Ahlquist DA,
-et al. Multi-target stool DNA test: Clinical performance and impact on
-yield and quality of colonoscopy for colorectal cancer screening.
-Gastrointestinal Endoscopy. 2016;
-
-\hypertarget{ref-laiyemo_short-_2013}{}
-6. Laiyemo AO, Doubeni C, Brim H, Ashktorab H, Schoen RE, Gupta S, et
-al. Short- and long-term risk of colorectal adenoma recurrence among
-whites and blacks. Gastrointestinal Endoscopy. 2013;77:447--54.
-
-\hypertarget{ref-matsuda_five-year_2009}{}
-7. Matsuda T, Fujii T, Sano Y, Kudo S-e, Oda Y, Igarashi M, et al.
-Five-year incidence of advanced neoplasia after initial colonoscopy in
-Japan: A multicenter retrospective cohort study. Japanese Journal of
-Clinical Oncology. 2009;39:435--42.
-
-\hypertarget{ref-ren_long-term_2016}{}
-8. Ren J, Kirkness CS, Kim M, Asche CV, Puli S. Long-term risk of
-colorectal cancer by gender after positive colonoscopy: Population-based
-cohort study. Current Medical Research and Opinion. 2016;32:1367--74.
-
-\hypertarget{ref-loberg_long-term_2014}{}
-9. Løberg M, Kalager M, Holme Ø, Hoff G, Adami H-O, Bretthauer M.
-Long-term colorectal-cancer mortality after adenoma removal. The New
-England Journal of Medicine. 2014;371:799--807.
-
-\hypertarget{ref-freeman_natural_2013}{}
-10. Freeman HJ. Natural history and long-term outcomes of patients
-treated for early stage colorectal cancer. Canadian Journal of
-Gastroenterology = Journal Canadien De Gastroenterologie.
-2013;27:409--13.
-
-\hypertarget{ref-lee_identification_2016}{}
-11. Lee JH, Lee JL, Park IJ, Lim S-B, Yu CS, Kim JC. Identification of
-Recurrence-Predictive Indicators in Stage I Colorectal Cancer. World
-Journal of Surgery. 2016;
-
-\hypertarget{ref-richards_evidence-based_2016}{}
-12. Richards CH, Ventham NT, Mansouri D, Wilson M, Ramsay G, Mackay CD,
-et al. An evidence-based treatment algorithm for colorectal polyp
-cancers: Results from the Scottish Screen-detected Polyp Cancer Study
-(SSPoCS). Gut. 2016;
-
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-13. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+3. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-zeller_potential_2014}{}
-14. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+4. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 \hypertarget{ref-dejea_microbiota_2014}{}
-15. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+5. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 \hypertarget{ref-zackular_manipulation_2016}{}
-16. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+6. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
 Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
@@
 \hypertarget{ref-arthur_microbial_2014}{}
-17. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+7. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
 McCafferty J, et al. Microbial genomic analysis reveals the essential
@@
 
+\hypertarget{ref-flynn_metabolic_2016}{}
+8. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
+
 \hypertarget{ref-arthur_intestinal_2012}{}
-18. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+9. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
 T-J, et al. Intestinal inflammation targets cancer-inducing activity of
@@
 
-\hypertarget{ref-flynn_metabolic_2016}{}
-19. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
-Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
-
 \hypertarget{ref-cotter_long-term_2016}{}
-20. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
+10. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
 et al. Long-Term Follow-up of Patients Having False Positive
@@
 \hypertarget{ref-obrien_impact_2013}{}
-21. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+11. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-daly_evaluation_2013}{}
-22. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
+12. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
 for colorectal cancer screening. Journal of Primary Care \& Community
@@
 \hypertarget{ref-kozich_development_2013}{}
-23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+13. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-24. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+14. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-25. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+15. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-26. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+16. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-02-14 submission/manuscript_outline_20161024.Rmd
updated manuscript text up to discussion

@@
 
-**Results:** The bacterial microbiome in the pre and post surgery samples of the individuals with adenoma were similar while those with carcinoma were not (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).
-There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-The post surgery sample for both models had a decrease in the positive probability for either lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`, and `r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`).
+**Results:** The bacterial microbiome in pre and post surgery samples of individuals with adenoma were more similar to each other than those with carcinoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
+A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify pre surgery samples had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
+The post surgery sample for both models had a decrease in the positive probability for either lesion or pre surgery sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (e.g.
+*`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).
+
 **Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.
-With individuals with carcinoma having more drastic changes to the overall community then those with adenoma.
-However, changes to commensal bacteria were common to individuals with either adenoma or carcinoma suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.
+With individuals with carcinoma having more drastic differences to the overall community then those with adenoma.
+Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.
 
@@
 
-Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].
+Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currenlty the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
 Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
@@
 These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
-Yet there remains limited information on the bacterial microbiome before and after surgery for removal of lesion (adenoma or carcinoma) and whether the community changes at all.
+Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgerical removal of lesion (adenoma or carcinoma) and whether the community changes at all.
 
-In this study we tested the hypothesis of whether that the bacterial microbiome changes after surgery for individuals with a lesion.
+In this study we tested the hypothesis that the bacterial microbiome changes after surgery for individuals with a lesion.
 Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
@@
 
-***Bacterial Community and FIT***
+***Bacterial Community and FIT:***
 We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.
-What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a].
-We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].
-The broad shift in bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).
-When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric tested [Table S1].
-We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lesion, adenoma, or carcinoma [Figure S1].
+What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a].
+We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1b].
+The broad shift in bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
+Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
+When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].
+We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].
 
-***Previously Associated Cancer Bacteria***
+***Cancer Associated Bacteria:***
 Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].
@@
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-There was a difference in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+There was a difference in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05).
 In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2].
-We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these respective OTUs [Figure 2].
+We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].
 
-***Full and Reduced Model***
+***Full and Reduced Model:***
 We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma.
-In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre versus post samples based on the bacterial community and FIT.
+In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre (initial) versus post (follow up) samples based on the bacterial community and FIT measurements.
 The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while the initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
@@
 The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap with each other [Figure 3a].
-The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].
+The ROC curve for the final lesion model used was within the range of the reduced lesion model [Figure 3a].
 Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.
@@
 
-***Common OTUs to both Models***
+***Common OTUs to both Models:***
 The reduced models were built based on the most important variables to either classification model.
 For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
-For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].
-There were a total of `r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`, and `r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`.
-The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
+For both models FIT measurement resulted in one of the largest decreases in MDA [Figure S2a & S3a].
+When we compared the two different reduced models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.
+Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.
+The vast majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
 
-***Positive Probability after Lesion Removal***
-If there were common OTUs for individuals with adenoma and carcinoma that were different versus normal controls we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or a before sample.
-This is what we observed regardless of model used (lesion or before sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (full lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, reduced lesion P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`, and reduced initial and follow up P-value = `r format(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).
+***Positive Probability after Lesion Removal:***
+If there were common OTUs for individuals with adenoma and carcinoma, that were different versus normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample.
+This is what we observed regardless of model used (lesion or initial sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
 
-Although there was a decrease in positive probability for all models there were differences on whether the models classified the samples as still having lesion.
-This was most evident for the full and reduced lesion model where there was a large difference between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"], scientific = TRUE, digits = 3)`, respectively).
-However, the before sample model did not have this disparity between predicted and actual (P-value = `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)` and `r format(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"], digits = 3, nsmall = 2)`).
-Even though the lesion model varied greatly between predicted and actual it was able to correctly keep the one individual who still had a carcinoma on follow up above the cut off threshold [Figure 4a & S4a] for a positive call while the before sample models did not [Figure 4b & S4b].
+When we seperated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+This lack of difference between the predicted and actual classifications were also observed for the initial sample model (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+Even though the lesion model was not as accurate in classifying samples as the initial sample model.
+It was able to correctly keep the one individual who still had a carcinoma on follow up above the cut off threshold for a positive call [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].
 
-***Treatment and Time Differences***
-After observing these changes in positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
-There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`).
-Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).
-Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible con-founder.
-Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
+***Treatment and Time Differences:***
+After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
+There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value `r ifelse(chemo_rad_summary['chemo_les', 'bh'] < 0.05 & chemo_rad_summary['chemo_red_les', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05) or radiation therapy (P-value `r ifelse(chemo_rad_summary['rad_les', 'bh'] < 0.05 & chemo_rad_summary['rad_red_les', 'bh'] < 0.05 , paste("<"), paste(">"))` 0.05).
+In contrast, we observed for the the before sample model a significant difference in decreased positive probability for those treated with chemotherapy (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
+All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].
+Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.
+Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
 
@@
 From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
-There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
+There was much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
 However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].
-There was also no differences in relative abundance of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].
+There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].
 
-Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested cancer causing microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
+Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested cancer causing microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
 This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.
@@
 
-*Problematic paragraph need to amend*
-
-There was no difference for the majority of models tested for differences in positive probability based on whether chemotherapy or radiation was received.
+For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].
 There was a difference in the length of time between initial and follow up sample between adenoma and carcinoma.
-These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.
+These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
 
@@
 There are a number of reasons why this may have occurred.
-First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.
-Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.
-Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
-Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.
-Finally, it is also possible that all of these potential explanations could have played a role.
+First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
+Second, is that it is possible that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.
+It is also possible that both of these explanations could have played a role.
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
@@
 
+Many of the common OTUs between the different models used had many OTUs that taxonomically classifed to potential butyrate producers [Table S2].
+Another batch of OTUs classifed to bacteria that can either degrade polyphenols or are inhibited by them.
+Both butyrate and polyphenols are thought to protective against cancer in part by reducing inflammation [@louis_gut_2014].
+These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.
+One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].
+High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].
+Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].
+Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the exisiting community but also have an effect on CRC risk and disease progression.
+
 One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
-This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
+This information would help to strengthen the case for our lesion model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
 Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
@@
 We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
-The time difference in collection of sample between adenoma and carcinoma that was observed could have affected our observed results.
-However, our random forest based model, specifically the lesion trained model, were still able to adjust positive probabilities for both adenoma and carcinoma to similar amounts despite this.
+The time difference in collection of sample between adenoma and carcinoma could have affected our observed results for the differences between individuals with adenoma or carcinoma.
+This confounding though does not affect the observations based on the overall lesion results.
 
@@
 
-Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.
-The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
-It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day stratify CRC disease risk with a high level of accuracy.
+Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.
+Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
+Our data provides evidence that commensal bacteria are important in the development of polyps and also potentially the transition from adenoma to carcinoma.
 

### 2017-02-14 submission/manuscript_outline_20161024.tex

@@
 
-\textbf{Results:} The bacterial microbiome in the pre and post surgery
-samples of the individuals with adenoma were similar while those with
-carcinoma were not (P-value = 0.000472). There was no difference in the
-relative abundance of any OTU between the pre and post surgery samples
-(P-value \textgreater{} 0.151). A model with a total of 53 variables was
-able to classify lesion (AUC = 0.811 - 0.866) while a model built to
-classify samples as before surgery had 70 variables (AUC = 0.641 -
-0.805). The post surgery sample for both models had a decrease in the
-positive probability for either lesion or before surgery sample (P-value
-= 3.28e-04 and 6.71e-12). In total there were 23 OTUs that were common
+\textbf{Results:} The bacterial microbiome in pre and post surgery
+samples of individuals with adenoma were more similar to each other than
+those with carcinoma (P-value \textless{} 0.05). There was no difference
+in the relative abundance of any OTU between the pre and post surgery
+samples (P-value \textgreater{} 0.05). A model with a total of 53
+variables was able to classify lesion (AUC = 0.811 - 0.866) while a
+model built to classify pre surgery samples had 70 variables (AUC =
+0.641 - 0.805). The post surgery sample for both models had a decrease
+in the positive probability for either lesion or pre surgery sample
+(P-value \textless{} 0.05). In total there were 23 OTUs that were common
 to both models and the majority of these classified to commensal
-bacteria (Bacteroides, Clostridiales, Blautia, and Ruminococcaceae).\\
+bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales}, \emph{Blautia},
+and \emph{Ruminococcaceae}).
+
 \textbf{Conclusions:} Our data supports the hypothesis that there are
 differences in the bacterial microbiome between pre and post surgical
-samples. With individuals with carcinoma having more drastic changes to
-the overall community then those with adenoma. However, changes to
-commensal bacteria were common to individuals with either adenoma or
-carcinoma suggesting that these bacteria may be central to initial polyp
-formation and transition to carcinoma.
+samples. With individuals with carcinoma having more drastic differences
+to the overall community then those with adenoma. Changes to commensal
+bacteria were some of the most important variables for model
+classification, suggesting that these bacteria may be central to initial
+polyp formation and transition to carcinoma.
 
@@
 Colorectal cancer (CRC) continues to be a leading cause of cancer
-related deaths and is the second most common cancer death among men aged
-40-79 years of age {[}1,2{]}. Over the last few years death due to the
-disease has seen a significant decrease, thanks mainly to improvements
-in screening {[}1{]}. However, despite this improvement there are still
-approximately 50,000 deaths from the disease a year {[}2{]}.
+related deaths and is currenlty the third most common cause of cancer
+deaths {[}1,2{]}. Over the last few years death due to the disease has
+seen a significant decrease, thanks mainly to improvements in screening
+{[}1{]}. However, despite this improvement there are still approximately
+50,000 deaths from the disease a year {[}2{]}.
 
@@
 studies have helped to provide a tantalizing link between the bacterial
-microbiome and CRC. Yet there remains limited information on the
-bacterial microbiome before and after surgery for removal of lesion
-(adenoma or carcinoma) and whether the community changes at all.
+microbiome and CRC. Although these studies suggest that the bacterial
+microbiome might change after treatment there remains limited
+information on the bacterial microbiome before and after surgerical
+removal of lesion (adenoma or carcinoma) and whether the community
+changes at all.
 
-In this study we tested the hypothesis of whether that the bacterial
-microbiome changes after surgery for individuals with a lesion. Our
-analysis included both alpha and beta diversity analysis along with
-investigation of individual operational taxonomic units (OTUs). We also
-utilized Random Forest models and observed how these models as well as
-specific OTUs within them performed pre (initial) and post (follow up)
-surgery. We also used these models to look for similar important OTUs to
-identify the crucial OTUs for not only classifying initial and follow up
-samples but also lesion or normal.
+In this study we tested the hypothesis that the bacterial microbiome
+changes after surgery for individuals with a lesion. Our analysis
+included both alpha and beta diversity analysis along with investigation
+of individual operational taxonomic units (OTUs). We also utilized
+Random Forest models and observed how these models as well as specific
+OTUs within them performed pre (initial) and post (follow up) surgery.
+We also used these models to look for similar important OTUs to identify
+the crucial OTUs for not only classifying initial and follow up samples
+but also lesion or normal.
 
@@
 
-\textbf{\emph{Bacterial Community and FIT}} We first wanted to test
+\textbf{\emph{Bacterial Community and FIT:}} We first wanted to test
 whether there were any broad differences between initial and follow up
@@
 dissimilar (as measured by thetayc) to their initial sample then those
-with adenoma (P-value = 0.000472) {[}Figure 1a{]}. We also found that
-there were larger changes in fecal blood (measured by FIT) for those
-with carcinoma versus adenoma (P-value = 2.15e-05) {[}Figure 1b{]}. The
-broad shift in bacterial community structure before and after surgery
-was visualized using NMDS for both adenoma {[}Figure1c{]} (PERMANOVA =
-0.999) and carcinoma {[}Figure 1d{]} (PERMANOVA = 0.005). Interestingly,
-when initial and follow up samples were compared to each other,
-regardless of whether they were adenoma or carcinoma (lesion), there was
-no significant overall difference between them (PERMANOVA = 0.09). When
+with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also found
+that there were larger changes in fecal blood (measured by FIT) for
+those with carcinoma versus adenoma (P-value \textless{} 0.0001)
+{[}Figure 1b{]}. The broad shift in bacterial community structure before
+and after surgery was visualized using NMDS for both adenoma
+{[}Figure1c{]} (PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure
+1d{]} (PERMANOVA \textless{} 0.05). Interestingly, when initial and
+follow up samples were compared to each other, regardless of whether
+they were adenoma or carcinoma (lesion), there was no significant
+overall difference between them (PERMANOVA \textgreater{} 0.05). When
 investigating more broad alpha diversity metrics there was no difference
-found between initial and follow up samples for lesion, adenoma, or
-carcinoma for any metric tested {[}Table S1{]}. We also observed that
-there was no difference in OTU relative abundance between initial and
-follow up samples for lesion, adenoma, or carcinoma {[}Figure S1{]}.
+found between initial and follow up samples for lesion, adenoma only, or
+carcinoma only for any metric tested {[}Table S1{]}. We also observed
+that there was no difference in OTU relative abundance between initial
+and follow up samples for lesion, adenoma only, or carcinoma only
+{[}Figure S1{]}.
 
-\textbf{\emph{Previously Associated Cancer Bacteria}} Previous
-literature has suggested that a number of oral microbes may be important
-in CRC pathogenesis {[}8{]}. So we next examined whether there were
-differences in previously well described carcinoma associated OTUs.
-These included the OTUs that aligned with \emph{Porphyromonas
-asaccharolytica} (Otu000202), \emph{Fusobacterium nucleatum}
-(Otu000442), \emph{Parvimonas micra} (Otu001273), and
-\emph{Peptostreptococcus stomatis} (Otu001682). There was a difference
-in relative abundance in initial and follow up samples for lesion and
-carcinoma for \emph{Parvimonas micra} (P-value = 0.0116), and
-\emph{Porphyromonas asaccharolytica} (P-value = 0.00842). In contrast,
-there was no difference in relative abundance in any of these OTUs for
-those with adenoma {[}Figure 2{]}. We also observed that only a small
-percentage of those with adenoma or carcinoma were positive or had an
-appreciable relative abundance of these respective OTUs {[}Figure 2{]}.
+\textbf{\emph{Cancer Associated Bacteria:}} Previous literature has
+suggested that a number of oral microbes may be important in CRC
+pathogenesis {[}8{]}. So we next examined whether there were differences
+in previously well described carcinoma associated OTUs. These included
+the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
+(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
+\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
+stomatis} (Otu001682). There was a difference in relative abundance in
+initial and follow up samples for lesion and carcinoma for
+\emph{Parvimonas micra} (P-value \textless{} 0.05), and
+\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05). In
+contrast, there was no difference in relative abundance in any of these
+OTUs for those with adenoma {[}Figure 2{]}. We also observed that only a
+small percentage of those with adenoma or carcinoma were positive or had
+an appreciable relative abundance of any of these respective OTUs
+{[}Figure 2{]}.
 
-\textbf{\emph{Full and Reduced Model}} We next wanted to identify if
+\textbf{\emph{Full and Reduced Model:}} We next wanted to identify if
 there were any common bacterial microbiome changes in individuals with
@@
 different models: one to classify lesion versus normal and one to
-classify pre versus post samples based on the bacterial community and
-FIT. The lesion model had an AUC range of 0.73 to 0.797 while the
-initial sample model had an AUC range of 0.485 to 0.686 after 100
-iterations of 20 repeated 10-fold cross validations. By identifying the
-most important variables for each respective model and then reducing
-them to only these factors we were able to increase the AUC in the
-lesion model (0.811 - 0.866) and initial sample model (0.641 - 0.805).\\
+classify pre (initial) versus post (follow up) samples based on the
+bacterial community and FIT measurements. The lesion model had an AUC
+range of 0.73 to 0.797 while the initial sample model had an AUC range
+of 0.485 to 0.686 after 100 iterations of 20 repeated 10-fold cross
+validations. By identifying the most important variables for each
+respective model and then reducing them to only these factors we were
+able to increase the AUC in the lesion model (0.811 - 0.866) and initial
+sample model (0.641 - 0.805).\\
 The test set AUC range for the full and reduced lesion model were
@@
 curve ranges overlap with each other {[}Figure 3a{]}. The ROC curve for
-the final lesion model used was within the range of both the full and
-reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
-range for the initial sample model performed much better then the
-training set AUCs. Both the full and reduced initial sample models
-overlapped with each other {[}Figure 3b{]} but there was a marked
-decrease in the ROC curve for the final before sample model used.
+the final lesion model used was within the range of the reduced lesion
+model {[}Figure 3a{]}. Interestingly, the test set AUC range for the
+initial sample model performed much better then the training set AUCs.
+Both the full and reduced initial sample models overlapped with each
+other {[}Figure 3b{]} but there was a marked decrease in the ROC curve
+for the final before sample model used.
 
-\textbf{\emph{Common OTUs to both Models}} The reduced models were built
-based on the most important variables to either classification model.
-For the lesion model there were a total of 53 variables {[}Figure S2{]}
-whereas for the initial sample model there were a total of 70 variables
-{[}Figure S3{]}. For both models FIT resulted in the largest decrease in
-MDA {[}Figure S2a \& S3a{]}. There were a total of Bacteroides,
-Clostridiales, Blautia, and Ruminococcaceae. The majority of these OTUs
-had classifications to bacteria typically thought of as commensal
+\textbf{\emph{Common OTUs to both Models:}} The reduced models were
+built based on the most important variables to either classification
+model. For the lesion model there were a total of 53 variables {[}Figure
+S2{]} whereas for the initial sample model there were a total of 70
+variables {[}Figure S3{]}. For both models FIT measurement resulted in
+one of the largest decreases in MDA {[}Figure S2a \& S3a{]}. When we
+compared the two different reduced models with each other there were a
+total of 23 common OTUs. Some of the most common taxonomic
+identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
+\emph{Blautia}, and \emph{Ruminococcaceae}. The vast majority of these
+OTUs had classifications to bacteria typically thought of as commensal
 {[}Table S2{]}.
 
-\textbf{\emph{Positive Probability after Lesion Removal}} If there were
-common OTUs for individuals with adenoma and carcinoma that were
-different versus normal controls we would expect to find a decrease in
+\textbf{\emph{Positive Probability after Lesion Removal:}} If there were
+common OTUs for individuals with adenoma and carcinoma, that were
+different versus normal controls, we would expect to find a decrease in
 the positive probability of the follow up sample to be either lesion or
-a before sample. This is what we observed regardless of model used
-(lesion or before sample) or whether it was built on the full or reduced
-variable data set {[}Figure 4 \& S4{]} (full lesion P-value = 1.62e-04,
-reduced lesion P-value = 3.28e-04, initial and follow up P-value =
-6.71e-12, and reduced initial and follow up P-value = 6.71e-12).
+an initial sample. This is what we observed regardless of model used
+(lesion or initial sample) or whether it was built on the full or
+reduced variable data set {[}Figure 4 \& S4{]} (P-value \textless{}
+0.001).
 
-Although there was a decrease in positive probability for all models
-there were differences on whether the models classified the samples as
-still having lesion. This was most evident for the full and reduced
-lesion model where there was a large difference between predicted and
-actual (P-value = 4.88e-01 and 4.7e-01, respectively). However, the
-before sample model did not have this disparity between predicted and
-actual (P-value = 1.00 and 1.00). Even though the lesion model varied
-greatly between predicted and actual it was able to correctly keep the
-one individual who still had a carcinoma on follow up above the cut off
-threshold {[}Figure 4a \& S4a{]} for a positive call while the before
-sample models did not {[}Figure 4b \& S4b{]}.
+When we seperated individuals based on whether they had an adenoma or
+carcinoma there was only a decrease in positive probability for the
+carcinoma group (P-value \textless{} 0.001) and not for the adenoma
+group (P-value \textgreater{} 0.05). We also observed that there were no
+significant differences in whether the models classified the samples as
+having lesion between the predicted and actual (P-value \textgreater{}
+0.05). This lack of difference between the predicted and actual
+classifications were also observed for the initial sample model (P-value
+\textgreater{} 0.05). Even though the lesion model was not as accurate
+in classifying samples as the initial sample model. It was able to
+correctly keep the one individual who still had a carcinoma on follow up
+above the cut off threshold for a positive call {[}Figure 4a \& S4a{]}
+while the initial sample model did not {[}Figure 4b \& S4b{]}.
 
-\textbf{\emph{Treatment and Time Differences}} After observing these
-changes in positive probability we wanted to assess whether additional
-treatments, such as chemotherapy and radiation, could have an impact on
-the results that we observed. There was no difference in the amount of
-change in positive probability for either the full or reduced lesion
-model for either chemotherapy (P-value = 0.184 and 0.184) or radiation
-therapy (P-value = 0.942 and 0.839). Although the before sample model
-was similar there was a significant difference in decreased positive
-probability for those treated with chemotherapy (P-value = 1.27e-03 and
-6.18e-04). Finally, we wanted to know if the length of time between the
-initial and follow up sample could be a possible con-founder. Within our
-study there was a significant difference in the time elapsed in the
-collection of the follow up sample between adenoma and carcinoma samples
-(uncorrected P-value = 7.59e-05) with time passed being lower for
-adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102 days).
+\textbf{\emph{Treatment and Time Differences:}} After observing these
+changes in the bacterial community and positive probability we wanted to
+assess whether additional treatments, such as chemotherapy and
+radiation, could have an impact on the results that we observed. There
+was no difference in the amount of change in positive probability for
+either the full or reduced lesion model for either chemotherapy (P-value
+\textgreater{} 0.05) or radiation therapy (P-value \textgreater{} 0.05).
+In contrast, we observed for the the before sample model a significant
+difference in decreased positive probability for those treated with
+chemotherapy (P-value \textless{} 0.05). All other variables that were
+tested showed no difference based on whether chemotherapy or radiation
+was used {[}Table S3{]}. Finally, we wanted to know if the length of
+time between the initial and follow up sample could be a possible
+confounder. Within our study there was a significant difference for the
+time elapsed in the collection of the follow up sample between adenoma
+and carcinoma (uncorrected P-value \textless{} 0.05), with time passed
+being less for adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102
+days).
 
@@
 whether the individual had an adenoma or carcinoma. There was much
-larger difference between initial and follow up samples based on the
+larger differences between initial and follow up samples based on the
 thetayc distance metric and in fecal blood as measured by FIT for
@@
 differences in relative abundance of any specific OTU for lesion,
-adenoma, or carcinoma {[}Figure s1{]}.
+adenoma only, or carcinoma only {[}Figure s1{]}.
 
@@
 looking specifically at four OTUs that taxonomically classified to
-previously suggested cancer causing microbes we found that only 3/4 had
+previously suggested cancer causing microbes we found that only 2/4 had
 a decrease in relative abundance between initial and follow up for those
@@
 
-\emph{Problematic paragraph need to amend}
-
-There was no difference for the majority of models tested for
-differences in positive probability based on whether chemotherapy or
-radiation was received. There was a difference in the length of time
+For the majority of tests performed there were no differences in the
+bacterial microbiome based on whether chemotherapy or radiation was
+received {[}Table S3{]}. There was a difference in the length of time
 between initial and follow up sample between adenoma and carcinoma.
 These results would indicate that the findings described were specific
-to the surgical intervention and that differences observed between
-carcinoma and adenoma samples can not be simply attributed to collection
-time between samples.
+to the surgical intervention and that some of the differences observed
+between carcinoma and adenoma samples could be due to differences in
+collection time between samples for the two different groups.
 
@@
 predictive within our models. There are a number of reasons why this may
-have occurred. First, one potential explanation is that even with
-surgery and a shift of the bacterial community these specific bacteria
-still persisted within the colon. Second, the bacteria even if they were
-reduced in relative abundance they were still present within the gut.
-Third, is that they were not present in enough individuals to be able to
-classify those with and without disease with a high degree of accuracy.
-Fourth, is that it is possible that our Random Forest models were able
-to get the same information from measures such as FIT or other OTUs.
-Finally, it is also possible that all of these potential explanations
-could have played a role. Regardless, our observations would suggest
-that an individual's resident bacteria have a large role to play in
-disease initiation and could change in a way that allows predictive
-models to lower the positive probability of a lesion after surgery
-{[}Figure 4{]}. It should be noted that our study does not argue against
-the importance of these CRC associated bacteria in the pathogenesis of
-disease but rather that the models do not utilize these specific
-bacteria for classification purposes (lesion or before sample). In fact,
-it is possible that these CRC associated bacteria are important in the
-transition from adenoma to carcinoma and would be one explanation as to
-why in our data we not only see high initial relative abundances, in
-certain individuals, but also large decreases in relative abundance in
-those with carcinoma but not in those with adenoma after surgery
+have occurred. First, is that they were not present in enough
+individuals to be able to classify those with and without disease with a
+high degree of accuracy. Second, is that it is possible that our Random
+Forest models were able to gather the same information from measures
+such as FIT or other OTUs. It is also possible that both of these
+explanations could have played a role. Regardless, our observations
+would suggest that an individual's resident bacteria have a large role
+to play in disease initiation and could change in a way that allows
+predictive models to lower the positive probability of a lesion after
+surgery {[}Figure 4{]}. It should be noted that our study does not argue
+against the importance of these CRC associated bacteria in the
+pathogenesis of disease but rather that the models do not utilize these
+specific bacteria for classification purposes (lesion or before sample).
+In fact, it is possible that these CRC associated bacteria are important
+in the transition from adenoma to carcinoma and would be one explanation
+as to why in our data we not only see high initial relative abundances,
+in certain individuals, but also large decreases in relative abundance
+in those with carcinoma but not in those with adenoma after surgery
 {[}Figure 2{]}.
+
+Many of the common OTUs between the different models used had many OTUs
+that taxonomically classifed to potential butyrate producers {[}Table
+S2{]}. Another batch of OTUs classifed to bacteria that can either
+degrade polyphenols or are inhibited by them. Both butyrate and
+polyphenols are thought to protective against cancer in part by reducing
+inflammation {[}10{]}. These protective compounds are derived from the
+breakdown of fiber, fruits, and vegetables by resident gut microbes. One
+example of this potential diet-microbiome-inflammation-polyp axis is
+that \emph{Bacteroides}, which was highly prevalent in our models, are
+known to be increased in those with high non-meat based protein
+consumption {[}11{]}. High protein consumption in general has been
+linked with an increased CRC risk {[}12{]}. Conversely,
+\emph{Bacteroides} are inhibited by polyphenols which are derived from
+fruits and vegetables {[}13{]}. Our data fits with the hypothesis that
+the microbial metabolites from breakdown products within our own diet
+could not only help to shape the exisiting community but also have an
+effect on CRC risk and disease progression.
 
@@
 a subsequent CRC diagnosis. This information would help to strengthen
-the case for our Random Forest based model to have kept a number of
-individuals above the cutoff threshold even though at follow up they
-were diagnosed as no longer having a lesion. Another limitation is that
-we do not know if adding modern tests such as the stool DNA test
-{[}10{]} could help improve our overall AUC. Another limitation is that
-this study drew heavily from those with Caucasian ancestry. The results
-may not be immediately representative of those with either Asian or
-African ancestry. Finally, although our training and test set are
-relatively large we still run the risk of over-fitting or having a model
-that may not be immediately extrapolate-able to other populations. We've
-done our best to safeguard against this by not only running 10-fold
-cross validation but also having over 100 different 80/20 splits to try
-and mimic the type of variation that might be expected to occur. The
-time difference in collection of sample between adenoma and carcinoma
-that was observed could have affected our observed results. However, our
-random forest based model, specifically the lesion trained model, were
-still able to adjust positive probabilities for both adenoma and
-carcinoma to similar amounts despite this.
+the case for our lesion model to have kept a number of individuals above
+the cutoff threshold even though at follow up they were diagnosed as no
+longer having a lesion. Another limitation is that we do not know if
+adding modern tests such as the stool DNA test {[}14{]} could help
+improve our overall AUC. Another limitation is that this study drew
+heavily from those with Caucasian ancestry. The results may not be
+immediately representative of those with either Asian or African
+ancestry. Finally, although our training and test set are relatively
+large we still run the risk of over-fitting or having a model that may
+not be immediately extrapolate-able to other populations. We've done our
+best to safeguard against this by not only running 10-fold cross
+validation but also having over 100 different 80/20 splits to try and
+mimic the type of variation that might be expected to occur. The time
+difference in collection of sample between adenoma and carcinoma could
+have affected our observed results for the differences between
+individuals with adenoma or carcinoma. This confounding though does not
+affect the observations based on the overall lesion results.
 
@@
 to the existing scientific knowledge on CRC and the bacterial
-microbiome. The ability for machine learning algorithms to take
-bacterial microbiome data and successfully lower positive probability
-after either adenoma or carcinoma removal provides evidence that there
-are specific signatures, mostly attributable to commensal organisms,
-associated with these lesions. It also shows that these algorithms can
-not only successfully react to successful treatment regimens but also
-may be able to one day stratify CRC disease risk with a high level of
-accuracy.
+microbiome: That there is a measurable difference in the bacterial
+community after surgical removal of lesion. Further, the ability for
+machine learning algorithms to take bacterial microbiome data and
+successfully lower positive probability after either adenoma or
+carcinoma removal provides evidence that there are specific signatures,
+mostly attributable to commensal organisms, associated with these
+lesions. Our data provides evidence that commensal bacteria are
+important in the development of polyps and also potentially the
+transition from adenoma to carcinoma.
 
@@
 The bacterial microbiome has been shown to normalize within this time
-period {[}11{]}. Our follow up data set had a total of 67 individuals
+period {[}15{]}. Our follow up data set had a total of 67 individuals
 that not only had a sample as described but also a follow up sample
@@
 previously published using both OC FIT-CHEK and OC-Auto Micro 80
-automated system (Polymedco Inc.) {[}12{]}. 16S rRNA gene sequencing was
-completed as previously described by Kozich, et al. {[}13{]}. In brief,
+automated system (Polymedco Inc.) {[}16{]}. 16S rRNA gene sequencing was
+completed as previously described by Kozich, et al. {[}17{]}. In brief,
 DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described {[}13{]}. The general processing workflow
+has been previously described {[}17{]}. The general processing workflow
 using mothur was as follows: Paired-end reads were first merged into
@@
 
-\textbf{\emph{Lesion Model Creation}} The Random Forest {[}14{]}
+\textbf{\emph{Lesion Model Creation}} The Random Forest {[}18{]}
 algorithm was used to create the model used for prediction of lesion
@@
 used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}15{]} in the vegan package
+community structure utilized PERMANOVA {[}19{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was needed
-a Benjamini-Hochberg (BH) correction was applied {[}16{]} and a
+a Benjamini-Hochberg (BH) correction was applied {[}20{]} and a
 corrected P-value of less than 0.05 was considered significant. Unless
@@
 
+\hypertarget{ref-louis_gut_2014}{}
+10. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
+metabolites and colorectal cancer. Nature Reviews Microbiology
+{[}Internet{]}. 2014 {[}cited 2017 Feb 14{]};12:661--72. Available from:
+\url{http://www.nature.com/doifinder/10.1038/nrmicro3344}
+
+\hypertarget{ref-zhu_intake_2016}{}
+11. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
+Proteins Substantially Increased the Relative Abundance of Genus
+Lactobacillus in Rat Feces. PloS One. 2016;11:e0152678.
+
+\hypertarget{ref-mu_colonic_2016}{}
+12. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
+Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
+Compared with a Normal-Protein Diet. The Journal of Nutrition.
+2016;146:474--83.
+
+\hypertarget{ref-ozdal_reciprocal_2016}{}
+13. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
+Reciprocal Interactions between Polyphenols and Gut Microbiota and
+Effects on Bioaccessibility. Nutrients {[}Internet{]}. 2016 {[}cited
+2017 Feb 14{]};8:78. Available from:
+\url{http://www.mdpi.com/2072-6643/8/2/78}
+
 \hypertarget{ref-cotter_long-term_2016}{}
-10. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
+14. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
 et al. Long-Term Follow-up of Patients Having False Positive
@@
 \hypertarget{ref-obrien_impact_2013}{}
-11. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+15. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-daly_evaluation_2013}{}
-12. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
+16. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
 for colorectal cancer screening. Journal of Primary Care \& Community
@@
 \hypertarget{ref-kozich_development_2013}{}
-13. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+17. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-14. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+18. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-15. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+19. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-16. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+20. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-02-14 submission/supplemental_outline_20161024.Rmd

@@
 
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 \newpage
+
+**Table S3: Summary of Comparisons based on Chemotherapy or Radiation Received**
+
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+
+chemo_rad_summary <- read.csv('../results/tables/probs_chemo_rad_pvalue_summary.csv', 
+                              header = T, stringsAsFactors = F)
+
+variable <- c(rep("Positive Probability", 8), rep("Thetayc", 2), rep(c("Sobs", "Shannon", "Evenness"), 2))
+comparison <- c(rep("Chemotherapy", 4), rep("Radiation", 4), "Chemotherapy", "Radiation", 
+                rep("Chemotherapy", 3), rep("Radiation", 3))
+
+st3 <- cbind(variable, comparison, format(chemo_rad_summary$pvalue, digits = 4), format(chemo_rad_summary$bh, digits = 4))
+
+colnames(st3) <- c("Variable Tested", "Treatment", "P-value", "BH Corrected P-value")
+
+kable(st3, format = "markdown", align = 'c', padding = 2)
+
+```

### 2017-02-16 submission/manuscript_outline_20161024.Rmd
version for lab review

@@
 
-Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currenlty the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
+Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
 Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
@@
 These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
-Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgerical removal of lesion (adenoma or carcinoma) and whether the community changes at all.
+Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgical removal of lesion (adenoma or carcinoma) and whether the community changes at all.
 
-In this study we tested the hypothesis that the bacterial microbiome changes after surgery for individuals with a lesion.
+In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after surgery for individuals with a lesion (adenoma or carcinoma).
 Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed pre (initial) and post (follow up) surgery.
-We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
+We also utilized Random Forest models and observed how these models, as well as specific OTUs within them, performed pre and post surgery.
+We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
 
@@
 ***Positive Probability after Lesion Removal:***
-If there were common OTUs for individuals with adenoma and carcinoma, that were different versus normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample.
+If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample.
 This is what we observed regardless of model used (lesion or initial sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
 
-When we seperated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
@@
 Even though the lesion model was not as accurate in classifying samples as the initial sample model.
-It was able to correctly keep the one individual who still had a carcinoma on follow up above the cut off threshold for a positive call [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].
+It was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].
 
@@
 ## Discussion
-From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
-There was much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
+From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
+There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
 However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].
 There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].
+Although there was a detectable change in the carcinoma group.
+It should be noted that whether this change was towards what wouldbe be expected for normal controls is not known.
 
@@
 We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.
-What we found was that the commonly associated CRC bacteria were not higly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
+What we found was that the commonly associated CRC bacteria were not highly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
 With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
-Using only these important OTUs and FIT both models (lesion and before sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
-Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.
+Using only these important OTUs and FIT, both models (lesion and initial sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
+Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community [Table S3].
 
@@
 First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
-Second, is that it is possible that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.
+Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.
 It is also possible that both of these explanations could have played a role.
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
-It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).
-In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].
+It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.
+In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].
 
-Many of the common OTUs between the different models used had many OTUs that taxonomically classifed to potential butyrate producers [Table S2].
-Another batch of OTUs classifed to bacteria that can either degrade polyphenols or are inhibited by them.
-Both butyrate and polyphenols are thought to protective against cancer in part by reducing inflammation [@louis_gut_2014].
+Many of the common OTUs between the different models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].
+Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.
+Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].
 These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.
@@
 Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].
-Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the exisiting community but also have an effect on CRC risk and disease progression.
+Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.
 
 One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
-This information would help to strengthen the case for our lesion model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
+This information would help to strengthen the case for our lesion model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
 Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
-Another limitation is that this study drew heavily from those with Caucasian ancestry.
-The results may not be immediately representative of those with either Asian or African ancestry.
-Finally, although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be immediately extrapolate-able to other populations.
+This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.
+Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.
 We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
-The time difference in collection of sample between adenoma and carcinoma could have affected our observed results for the differences between individuals with adenoma or carcinoma.
-This confounding though does not affect the observations based on the overall lesion results.
+The time difference in collection of sample between adenoma and carcinoma could have affected our observations for differences between individuals with adenoma or carcinoma.
+This confounding though would not affect the observations where these individuals are grouped together.
 
-Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.
-This may have occurred because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
+Interestingly, within figure 3 the before sample model showed better test AUC results then the training set AUC.
+This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
 Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.
-However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.
-Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.
+However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model alone.
+Further independent studies need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.
 
-Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.
+Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.
 Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
-Our data provides evidence that commensal bacteria are important in the development of polyps and also potentially the transition from adenoma to carcinoma.
+Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.
 
@@
 
-***Study Design and Patient Sampling***
-The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].
+***Study Design and Patient Sampling:***
+The sampling and design were similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].
 In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
@@
 
-***FIT and 16S rRNA Gene Sequencing***
+***FIT and 16S rRNA Gene Sequencing:***
 FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013].
@@
 [@kozich_development_2013].
-In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
+DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
 The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.
@@
 
-***Sequence Processing***
+***Sequence Processing:***
 The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.
@@
 
-***Lesion Model Creation***
-The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.
+***Lesion Model Creation:***
+The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
 This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.
-The model included data on FIT and the bacterial microbiome.
+In brief, the model included data on FIT and the bacterial microbiome.
 Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.
@@
 This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
-Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
+Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
@@
 
-***Before Sample Model Creation***
-We also investigated whether a model could be created that could identify before and after surgery samples.
+***Initial Sample Model Creation:***
+We also investigated whether a model could be created that could identify pre (initial) and post (follow up) surgery samples.
 The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
-Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure that was used to create the lesion model.
-Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.
+Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.
+Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial surgery sample.
 
-***Selection of Important OTUs***
-In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.
+***Selection of Important OTUs:***
+In order to assess which variables were most important to either model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.
 This final collated list of variables was what was considered the most important for the lesion or before sample models.
 
-***Statistical Analysis***
+***Statistical Analysis:***
 The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.
@@
 Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.
-Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
+Where multiple comparison testing was appropriate a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
 Unless otherwise stated the P-values reported are those that were BH corrected.
 
-***Analysis Overview***
-Initial and follow up samples were analyzed for differences in alpha and beta diversity.
-Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.
-From here, all OTUs that were used in either model were then analyzed using a paired Wilcoxson test.
-We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.
-We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
+***Analysis Overview:***
+We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
+This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.
 From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
 The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.
-We then used the before sample model to assess whether it could classify samples better then the lesion model.
-The most important OTUs were then identified from this model and used to create a reduced feature before sample model.
+We then used the initial sample model to assess whether it could classify samples better then the lesion model.
+The most important OTUs were then identified from this model and used to create a reduced feature initial sample model.
 This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.
-Finally, a list of common OTUs were found for the two different models used.
+A list of common OTUs were found for the two different models used.
+Finally, both lesion and initial sample models initial positive probabilites were tested against the follow up positive probabilities.
 
-***Reproducible Methods.***
+***Reproducible Methods:***
 A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
@@
 **Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
-B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
+B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
 C) NMDS of the initial and follow up samples for the Adenoma group.
@@
 
-**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples.
-There was a significant difference in initial and follow up sample for the OTUs classified as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
 B) Adenoma initial and follow up samples.
@@
 
-**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Before Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
-The blue line represents the reduced variable model using 100% of the data set.
+**Figure 3: Receiver Operating Characteristic Curve for Lesion and Initial Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
+The blue line represents the reduced variable model using 100% of the data set and what was used for subsequent classification.
 A) Lesion model.
@@
 
-**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Reduced Variable Models**  A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
-D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevent Neoplasia (SRN).
+D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.
 The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
@@
 
-**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of OTUs in Initial versus Follow Up**
+**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of All OTUs for Initial versus Follow Up**
 
@@
 
-**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
-D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.
+**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevent Neoplasia (SRN).
+D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.
 The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
 
-**Figure S5: Thetayc Graphed Against Time of Follow up Sample from Initial**
+**Figure S5: Thetayc Versus Time of Follow up Sample from Initial**
 
@@
 The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+We would also like to thank Amanda Elmore for reviewing and correcting code error and providing feedback on manuscript drafts.
+We would also like to thank Nicholas Lesniak for providing feedback on manuscript drafts.
 

### 2017-02-16 submission/manuscript_outline_20161024.tex

@@
 Colorectal cancer (CRC) continues to be a leading cause of cancer
-related deaths and is currenlty the third most common cause of cancer
+related deaths and is currently the third most common cause of cancer
 deaths {[}1,2{]}. Over the last few years death due to the disease has
@@
 microbiome might change after treatment there remains limited
-information on the bacterial microbiome before and after surgerical
+information on the bacterial microbiome before and after surgical
 removal of lesion (adenoma or carcinoma) and whether the community
@@
 In this study we tested the hypothesis that the bacterial microbiome
-changes after surgery for individuals with a lesion. Our analysis
+changes between pre (initial) and post (follow up) samples after surgery
+for individuals with a lesion (adenoma or carcinoma). Our analysis
 included both alpha and beta diversity analysis along with investigation
 of individual operational taxonomic units (OTUs). We also utilized
-Random Forest models and observed how these models as well as specific
-OTUs within them performed pre (initial) and post (follow up) surgery.
-We also used these models to look for similar important OTUs to identify
-the crucial OTUs for not only classifying initial and follow up samples
-but also lesion or normal.
+Random Forest models and observed how these models, as well as specific
+OTUs within them, performed pre and post surgery. We also investigated
+the models for similar important OTUs to identify the crucial OTUs for
+not only classifying initial and follow up samples but also lesion or
+normal.
 
@@
 \textbf{\emph{Positive Probability after Lesion Removal:}} If there were
-common OTUs for individuals with adenoma and carcinoma, that were
-different versus normal controls, we would expect to find a decrease in
-the positive probability of the follow up sample to be either lesion or
-an initial sample. This is what we observed regardless of model used
-(lesion or initial sample) or whether it was built on the full or
-reduced variable data set {[}Figure 4 \& S4{]} (P-value \textless{}
-0.001).
+common OTUs that could separate adenoma and carcinoma from normal
+controls, we would expect to find a decrease in the positive probability
+of the follow up sample to be either lesion or an initial sample. This
+is what we observed regardless of model used (lesion or initial sample)
+or whether it was built on the full or reduced variable data set
+{[}Figure 4 \& S4{]} (P-value \textless{} 0.001).
 
-When we seperated individuals based on whether they had an adenoma or
+When we separated individuals based on whether they had an adenoma or
 carcinoma there was only a decrease in positive probability for the
@@
 in classifying samples as the initial sample model. It was able to
-correctly keep the one individual who still had a carcinoma on follow up
-above the cut off threshold for a positive call {[}Figure 4a \& S4a{]}
-while the initial sample model did not {[}Figure 4b \& S4b{]}.
+correctly classify the one individual who still had a carcinoma on
+follow up {[}Figure 4a \& S4a{]} while the initial sample model did not
+{[}Figure 4b \& S4b{]}.
 
@@
 
-From our results there were some large observed differences in the
-bacterial microbiome between pre and post surgery samples based on
-whether the individual had an adenoma or carcinoma. There was much
-larger differences between initial and follow up samples based on the
-thetayc distance metric and in fecal blood as measured by FIT for
-individuals with carcinoma versus adenoma {[}Figure 1{]}. However, there
-were no differences between initial and follow up samples for Shannon
-Diversity, observed OTUs, or evenness regardless of whether the
-individual had an adenoma or carcinoma {[}Table S1{]}. There was also no
-differences in relative abundance of any specific OTU for lesion,
-adenoma only, or carcinoma only {[}Figure s1{]}.
+From our results there were large observed differences in the bacterial
+microbiome between pre and post surgery samples based on whether the
+individual had an adenoma or carcinoma. There were much larger
+differences between initial and follow up samples based on the thetayc
+distance metric and in fecal blood as measured by FIT for individuals
+with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
+differences between initial and follow up samples for Shannon Diversity,
+observed OTUs, or evenness regardless of whether the individual had an
+adenoma or carcinoma {[}Table S1{]}. There was also no differences in
+relative abundance of any specific OTU for lesion, adenoma only, or
+carcinoma only {[}Figure s1{]}. Although there was a detectable change
+in the carcinoma group. It should be noted that whether this change was
+towards what wouldbe be expected for normal controls is not known.
 
@@
 change for both adenoma and carcinoma. What we found was that the
-commonly associated CRC bacteria were not higly represented within our
+commonly associated CRC bacteria were not highly represented within our
 models but rather that OTUs that made up the most important variables
@@
 (\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
-and FIT both models (lesion and before sample) significantly decreased
+and FIT, both models (lesion and initial sample) significantly decreased
 positive probability of either lesion or being an initial sample on
@@
 total of 23 OTUs were common to both models and the vast majority
-belonged to regular residents of our gut community.
+belonged to regular residents of our gut community {[}Table S3{]}.
 
@@
 individuals to be able to classify those with and without disease with a
-high degree of accuracy. Second, is that it is possible that our Random
-Forest models were able to gather the same information from measures
-such as FIT or other OTUs. It is also possible that both of these
-explanations could have played a role. Regardless, our observations
-would suggest that an individual's resident bacteria have a large role
-to play in disease initiation and could change in a way that allows
-predictive models to lower the positive probability of a lesion after
-surgery {[}Figure 4{]}. It should be noted that our study does not argue
-against the importance of these CRC associated bacteria in the
-pathogenesis of disease but rather that the models do not utilize these
-specific bacteria for classification purposes (lesion or before sample).
-In fact, it is possible that these CRC associated bacteria are important
-in the transition from adenoma to carcinoma and would be one explanation
-as to why in our data we not only see high initial relative abundances,
-in certain individuals, but also large decreases in relative abundance
-in those with carcinoma but not in those with adenoma after surgery
-{[}Figure 2{]}.
+high degree of accuracy. Second, is that our Random Forest models were
+able to gather the same information from measures such as FIT or other
+OTUs. It is also possible that both of these explanations could have
+played a role. Regardless, our observations would suggest that an
+individual's resident bacteria have a large role to play in disease
+initiation and could change in a way that allows predictive models to
+lower the positive probability of a lesion after surgery {[}Figure 4{]}.
+It should be noted that our study does not argue against the importance
+of these CRC associated bacteria in the pathogenesis of disease but
+rather that they are not the main bacteria changing after surgical
+removal of lesion. In fact, it is possible that these CRC associated
+bacteria are important in the transition from adenoma to carcinoma and
+would be one explanation as to why in our data we not only see high
+initial relative abundances in carcinoma and not adenoma individuals but
+also large decreases in relative abundance in some of those with
+carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.
 
 Many of the common OTUs between the different models used had many OTUs
-that taxonomically classifed to potential butyrate producers {[}Table
-S2{]}. Another batch of OTUs classifed to bacteria that can either
+that taxonomically classified to potential butyrate producers {[}Table
+S2{]}. Another batch of OTUs classified to bacteria that can either
 degrade polyphenols or are inhibited by them. Both butyrate and
-polyphenols are thought to protective against cancer in part by reducing
-inflammation {[}10{]}. These protective compounds are derived from the
-breakdown of fiber, fruits, and vegetables by resident gut microbes. One
-example of this potential diet-microbiome-inflammation-polyp axis is
-that \emph{Bacteroides}, which was highly prevalent in our models, are
-known to be increased in those with high non-meat based protein
-consumption {[}11{]}. High protein consumption in general has been
-linked with an increased CRC risk {[}12{]}. Conversely,
-\emph{Bacteroides} are inhibited by polyphenols which are derived from
-fruits and vegetables {[}13{]}. Our data fits with the hypothesis that
-the microbial metabolites from breakdown products within our own diet
-could not only help to shape the exisiting community but also have an
-effect on CRC risk and disease progression.
+polyphenols are thought to be protective against cancer in part by
+reducing inflammation {[}10{]}. These protective compounds are derived
+from the breakdown of fiber, fruits, and vegetables by resident gut
+microbes. One example of this potential
+diet-microbiome-inflammation-polyp axis is that \emph{Bacteroides},
+which was highly prevalent in our models, are known to be increased in
+those with high non-meat based protein consumption {[}11{]}. High
+protein consumption in general has been linked with an increased CRC
+risk {[}12{]}. Conversely, \emph{Bacteroides} are inhibited by
+polyphenols which are derived from fruits and vegetables {[}13{]}. Our
+data fits with the hypothesis that the microbial metabolites from
+breakdown products within our own diet could not only help to shape the
+existing community but also have an effect on CRC risk and disease
+progression.
 
@@
 a subsequent CRC diagnosis. This information would help to strengthen
-the case for our lesion model to have kept a number of individuals above
-the cutoff threshold even though at follow up they were diagnosed as no
+the case for our lesion model keeping a number of individuals above the
+cutoff threshold even though at follow up they were diagnosed as no
 longer having a lesion. Another limitation is that we do not know if
 adding modern tests such as the stool DNA test {[}14{]} could help
-improve our overall AUC. Another limitation is that this study drew
-heavily from those with Caucasian ancestry. The results may not be
-immediately representative of those with either Asian or African
-ancestry. Finally, although our training and test set are relatively
-large we still run the risk of over-fitting or having a model that may
-not be immediately extrapolate-able to other populations. We've done our
-best to safeguard against this by not only running 10-fold cross
-validation but also having over 100 different 80/20 splits to try and
-mimic the type of variation that might be expected to occur. The time
-difference in collection of sample between adenoma and carcinoma could
-have affected our observed results for the differences between
-individuals with adenoma or carcinoma. This confounding though does not
-affect the observations based on the overall lesion results.
+improve our overall AUC. This study also drew heavily from those with
+Caucasian ancestry making it possible that the observations may not be
+representative of those with either Asian or African ancestry. Although
+our training and test set are relatively large we still run the risk of
+over-fitting or having a model that may not be representative of other
+populations. We've done our best to safeguard against this by not only
+running 10-fold cross validation but also having over 100 different
+80/20 splits to try and mimic the type of variation that might be
+expected to occur. The time difference in collection of sample between
+adenoma and carcinoma could have affected our observations for
+differences between individuals with adenoma or carcinoma. This
+confounding though would not affect the observations where these
+individuals are grouped together.
 
-Another interesting outcome was that within figure 3 the before sample
-model showed better test AUC results then the training set AUC. This may
-have occurred because the training AUC that was determined from 20
-repeated 10 fold cross validation removed samples at random and did not
-take into account that they were matched samples. Another potential
-reason is that the model itself may be over-fit since the total number
-of samples was not that large. However, the lesion model did not suffer
-from these discrepancies and similar conclusions can be drawn solely
-from this model. Regardless, further independent studies will need to be
-carried out to verify our findings since not only are we dealing with
-feces, which could be very different than the communities present on the
-actual tissue, but also are dealing with correlations that may not be
-representative of the true pathogenesis of disease.
+Interestingly, within figure 3 the before sample model showed better
+test AUC results then the training set AUC. This may have occurred
+because the training AUC determined from 20 repeated 10 fold cross
+validation removed samples at random and did not take into account that
+they were matched samples. Another potential reason is that the model
+itself may be over-fit since the total number of samples was not that
+large. However, the lesion model did not suffer from these discrepancies
+and similar conclusions can be drawn solely from this model alone.
+Further independent studies need to be carried out to verify our
+findings since not only are we dealing with feces, which could be very
+different than the communities present on the actual tissue, but also
+are dealing with correlations that may not be representative of the true
+pathogenesis of disease.
 
-Despite these limitations we think that these findings significantly add
-to the existing scientific knowledge on CRC and the bacterial
-microbiome: That there is a measurable difference in the bacterial
-community after surgical removal of lesion. Further, the ability for
-machine learning algorithms to take bacterial microbiome data and
-successfully lower positive probability after either adenoma or
-carcinoma removal provides evidence that there are specific signatures,
-mostly attributable to commensal organisms, associated with these
-lesions. Our data provides evidence that commensal bacteria are
-important in the development of polyps and also potentially the
-transition from adenoma to carcinoma.
+Despite these limitations our findings add to the existing scientific
+knowledge on CRC and the bacterial microbiome: That there is a
+measurable difference in the bacterial community after surgical removal
+of lesion. Further, the ability for machine learning algorithms to take
+bacterial microbiome data and successfully lower positive probability
+after either adenoma or carcinoma removal provides evidence that there
+are specific signatures, mostly attributable to commensal organisms,
+associated with these lesions. Our data provides evidence that commensal
+bacteria may be important in the development of polyps and also
+potentially the transition from adenoma to carcinoma.
 
@@
 
-\textbf{\emph{Study Design and Patient Sampling}} The sampling and
-design of the study was similar to that reported in Baxter, et al
-{[}3{]}. In brief, study exclusion involved those who had already
-undergone surgery, radiation, or chemotherapy, had colorectal cancer
-before a baseline fecal sample could be obtained, had IBD, a known
-hereditary non-polyposis colorectal cancer, or Familial adenomatous
-polyposis. Samples used to build the models for prediction were
-collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial microbiome has been shown to normalize within this time
-period {[}15{]}. Our follow up data set had a total of 67 individuals
-that not only had a sample as described but also a follow up sample
-between 188 - 546 days after surgery and treatment had been completed.
-This study was approved by the University of Michigan Institutional
-Review Board. All study participants provided informed consent and the
-study itself conformed to the guidelines set out by the Helsinki
-Declaration.
+\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
+design were similar to that reported in Baxter, et al {[}3{]}. In brief,
+study exclusion involved those who had already undergone surgery,
+radiation, or chemotherapy, had colorectal cancer before a baseline
+fecal sample could be obtained, had IBD, a known hereditary
+non-polyposis colorectal cancer, or Familial adenomatous polyposis.
+Samples used to build the models for prediction were collected either
+prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
+microbiome has been shown to normalize within this time period {[}15{]}.
+Our follow up data set had a total of 67 individuals that not only had a
+sample as described but also a follow up sample between 188 - 546 days
+after surgery and treatment had been completed. This study was approved
+by the University of Michigan Institutional Review Board. All study
+participants provided informed consent and the study itself conformed to
+the guidelines set out by the Helsinki Declaration.
 
-\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
+\textbf{\emph{FIT and 16S rRNA Gene Sequencing:}} FIT was analyzed as
 previously published using both OC FIT-CHEK and OC-Auto Micro 80
 automated system (Polymedco Inc.) {[}16{]}. 16S rRNA gene sequencing was
-completed as previously described by Kozich, et al. {[}17{]}. In brief,
-DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
-Laboratories) and an epMotion 5075 automated pipetting system
-(Eppendorf). The V4 variable region was amplified and the resulting
-product was split between three sequencing runs with normal, adenoma,
-and carcinoma evenly represented on each run. Each group was randomly
-assigned to avoid biases based on sample collection location.
+completed as previously described by Kozich, et al. {[}17{]}. DNA
+extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
+and an epMotion 5075 automated pipetting system (Eppendorf). The V4
+variable region was amplified and the resulting product was split
+between three sequencing runs with normal, adenoma, and carcinoma evenly
+represented on each run. Each group was randomly assigned to avoid
+biases based on sample collection location.
 
-\textbf{\emph{Sequence Processing}} The mothur software package
+\textbf{\emph{Sequence Processing:}} The mothur software package
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
@@
 
-\textbf{\emph{Lesion Model Creation}} The Random Forest {[}18{]}
+\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}18{]}
 algorithm was used to create the model used for prediction of lesion
 (adenoma or carcinoma) with the main testing and training of the model
-completed on a data set of 490 individuals. This model was then applied
-to our follow up data set of 67 individuals. The model included data on
-FIT and the bacterial microbiome. Non-binary data was checked for near
-zero variance and OTUs that had near zero variance were removed. This
-pre-processing was performed with the R package caret (v6.0.73).
-Optimization of the mtry hyper-parameter involved taking the samples and
-making 100 different 80/20 (train/test) splits of the data where normal
-and lesion were represented in the same proportion within both the whole
-data set and the 80/20 split. Each of these splits were then run through
-20 repeated 10-fold cross validations to optimize the mtry
-hyper-parameter by maximizing the AUC (Area Under the Curve of the
+completed on a data set of 423 individuals. This model was then applied
+to our follow up data set of 67 individuals. In brief, the model
+included data on FIT and the bacterial microbiome. Non-binary data was
+checked for near zero variance and OTUs that had near zero variance were
+removed. This pre-processing was performed with the R package caret
+(v6.0.73). Optimization of the mtry hyper-parameter involved taking the
+samples and making 100 different 80/20 (train/test) splits of the data
+where normal and lesion were represented in the same proportion within
+both the whole data set and the 80/20 split. Each of these splits were
+then run through 20 repeated 10-fold cross validations to optimize the
+mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
 Receiver Operator Characteristic). This resulting model was then tested
 on the 20\% of the data that was originally held out from this overall
-process. Once the ideal mtry was found the entire 490 sample set was
+process. Once the ideal mtry was found the entire 423 sample set was
 used to create the final Random Forest model on which classifications on
@@
 
-\textbf{\emph{Before Sample Model Creation}} We also investigated
-whether a model could be created that could identify before and after
-surgery samples. The main difference was that only the 67-person cohort
-was used at all stages of model building and classification. Other than
-this difference the creation of this model and optimization of the mtry
-hyper-parameter was completed using the same procedure that was used to
-create the lesion model. Instead of classifying samples as positive or
-negative of lesion this model classified samples as positive or negative
-for being a before surgery sample.
+\textbf{\emph{Initial Sample Model Creation:}} We also investigated
+whether a model could be created that could identify pre (initial) and
+post (follow up) surgery samples. The main difference was that only the
+67-person cohort was used at all stages of model building and
+classification. Other than this difference the creation of this model
+and optimization of the mtry hyper-parameter was completed using the
+same procedure as was used for the lesion model. Instead of classifying
+samples as positive or negative of lesion this model classified samples
+as positive or negative for being an initial surgery sample.
 
-\textbf{\emph{Selection of Important OTUs}} In order to assess which
-variables were most important to all the models we counted the number of
+\textbf{\emph{Selection of Important OTUs:}} In order to assess which
+variables were most important to either model we counted the number of
 times a variable was present in the top 10\% of mean decrease in
@@
 
-\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
-used for all statistical analysis. Comparisons between bacterial
+\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
+was used for all statistical analysis. Comparisons between bacterial
 community structure utilized PERMANOVA {[}19{]} in the vegan package
@@
 differences between initial and follow up samples utilized a paired
-Wilcoxson ranked sum test. Where multiple comparison testing was needed
-a Benjamini-Hochberg (BH) correction was applied {[}20{]} and a
-corrected P-value of less than 0.05 was considered significant. Unless
-otherwise stated the P-values reported are those that were BH corrected.
+Wilcoxson ranked sum test. Where multiple comparison testing was
+appropriate a Benjamini-Hochberg (BH) correction was applied {[}20{]}
+and a corrected P-value of less than 0.05 was considered significant.
+Unless otherwise stated the P-values reported are those that were BH
+corrected.
 
-\textbf{\emph{Analysis Overview}} Initial and follow up samples were
-analyzed for differences in alpha and beta diversity. Next, differences
-in FIT between initial and follow ups for either adenoma or carcinoma
-were investigated. From here, all OTUs that were used in either model
-were then analyzed using a paired Wilcoxson test. We then investigated
-the relative abundance of specific previously associated CRC bacteria,
-specifically, OTUs that taxonomically classified to \emph{Fusobacterium
-nucleatum}, \emph{Parvimonas micra}, \emph{Peptostreptococcus
-assacharolytica}, and \emph{Porphyromonas stomatis}. We wanted to test
-if there were any differences based on whether the individual had an
-adenoma or carcinoma. From here the lesion model was then tested for
-accuracy in prediction and whether it reduced the positive probability
-of lesion after surgery. The most important OTUs for this were used to
-build a reduced model and it was assessed for similarity to the original
-model. We then used the before sample model to assess whether it could
-classify samples better then the lesion model. The most important OTUs
-were then identified from this model and used to create a reduced
-feature before sample model. This reduced feature model, as was done
-with the lesion model, was compared to the full model for loss of
-accuracy. Finally, a list of common OTUs were found for the two
-different models used.
+\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
+any differences based on whether the individual had an adenoma or
+carcinoma. This was done by testing initial and follow up samples for
+differences in alpha and beta diversity, testing differences in FIT
+between initial and follow ups, testing all OTUs that were used in
+either built model, and investigating the relative abundance of specific
+previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
+\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
+\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
+the lesion model was then tested for accuracy in prediction and whether
+it reduced the positive probability of lesion after surgery. The most
+important OTUs for this were used to build a reduced model and it was
+assessed for similarity to the original model. We then used the initial
+sample model to assess whether it could classify samples better then the
+lesion model. The most important OTUs were then identified from this
+model and used to create a reduced feature initial sample model. This
+reduced feature model, as was done with the lesion model, was compared
+to the full model for loss of accuracy. A list of common OTUs were found
+for the two different models used. Finally, both lesion and initial
+sample models initial positive probabilites were tested against the
+follow up positive probabilities.
 
-\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
+\textbf{\emph{Reproducible Methods:}} A detailed and reproducible
 description of how the data were processed and analyzed can be found at
@@
 difference was found between the adenoma and carcinoma group for change
-in FIT (P-value = 2.15e-05). C) NMDS of the initial and follow up
-samples for the Adenoma group. D) NMDS of the initial and follow up
-samples for the Carcinoma group. For C) and D) the teal represents
-initial samples and the pink represents follow up samples.
+in FIT measurement (P-value = 2.15e-05). C) NMDS of the initial and
+follow up samples for the Adenoma group. D) NMDS of the initial and
+follow up samples for the Carcinoma group. For C) and D) the teal
+represents initial samples and the pink represents follow up samples.
 
 \textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
-Follow up Samples.} A) Carcinoma initial and follow up samples. There
-was a significant difference in initial and follow up sample for the
-OTUs classified as \emph{Peptostreptococcus stomatis} (P-value = 0.0496)
-and \emph{Porphyromonas asaccharolytica} (P-value = 0.00842). B) Adenoma
+Follow up Samples.} A) Carcinoma initial and follow up samples had an
+observed significant difference in initial and follow up sample for the
+OTUs classified as \emph{Parvimonas micra} (P-value = 0.0116) and
+\emph{Porphyromonas asaccharolytica} (P-value = 0.00842). B) Adenoma
 initial and follow up samples. There were no significant differences
@@
 
-\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
-for lesion and Before Sample Models.} The shaded areas represents the
-range of values of a 100 different 80/20 splits of the test set data
-using either all variables (grey) or reduced variable (red) models. The
-blue line represents the reduced variable model using 100\% of the data
-set. A) Lesion model. B) Before sample model
+\textbf{Figure 3: Receiver Operating Characteristic Curve for Lesion and
+Initial Sample Models.} The shaded areas represents the range of values
+of a 100 different 80/20 splits of the test set data using either all
+variables (grey) or reduced variable (red) models. The blue line
+represents the reduced variable model using 100\% of the data set and
+what was used for subsequent classification. A) Lesion model. B) Before
+sample model
 
 \textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Before Sample Reduced Variable Models} A) Lesion
-positive probability adjustment of those with carcinoma from initial to
-follow up sample B) Initial follow up positive probability adjustment of
-those with carcinoma from initial to follow up sample C) Lesion positive
-probability adjustment of those with adenoma as well as those with SRN
-and the probability adjustment from initial to follow up sample. D)
-Initial follow up positive probability adjustment of those with adenoma
-as well as those with SRN and the probability adjustment from initial to
-follow up sample. The dotted line represents the threshold used to make
-the decision of whether a sample was positive or not.
+Results for Lesion and Initial Sample Reduced Variable Models} A) Lesion
+positive probability change from initial to follow up sample in those
+with carcinoma B) Initial positive probability change from initial to
+follow up sample in those with carcinoma C) Lesion positive probability
+change from initial to follow up sample if those with adenoma or Screen
+Relevent Neoplasia (SRN). D) Initial positive probability change from
+initial to follow up sample in those with adenoma or SRN. The dotted
+line represents the threshold used to make the decision of whether a
+sample was positive or not.
 
@@
 \textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
-Analysis of OTUs in Initial versus Follow Up}
+Analysis of All OTUs for Initial versus Follow Up}
 
@@
 Results for Lesion and Before Sample Full Variable Models} A) Lesion
-positive probability adjustment of those with carcinoma from initial to
-follow up sample B) Initial follow up positive probability adjustment of
-those with carcinoma from initial to follow up sample C) Lesion positive
-probability adjustment of those with adenoma as well as those with SRN
-and the probability adjustment from initial to follow up sample. D)
-Initial follow up positive probability adjustment of those with adenoma
-as well as those with SRN and the probability adjustment from initial to
-follow up sample. The dotted line represents the threshold used to make
-the decision of whether a sample was positive or not.
+positive probability change from initial to follow up sample in those
+with carcinoma B) Initial positive probability change from initial to
+follow up sample in those with carcinoma C) Lesion positive probability
+change from initial to follow up sample in those with adenoma or Screen
+Relevent Neoplasia (SRN). D) Initial positive probability change from
+initial to follow up sample in those with adenoma or SRN. The dotted
+line represents the threshold used to make the decision of whether a
+sample was positive or not.
 
-\textbf{Figure S5: Thetayc Graphed Against Time of Follow up Sample from
-Initial}
+\textbf{Figure S5: Thetayc Versus Time of Follow up Sample from Initial}
 
@@
 The authors thank the Great Lakes-New England Early Detection Research
-Network for providing the fecal samples that were used in this study.
+Network for providing the fecal samples that were used in this study. We
+would also like to thank Amanda Elmore for reviewing and correcting code
+error and providing feedback on manuscript drafts. We would also like to
+thank Nicholas Lesniak for providing feedback on manuscript drafts.
 

### 2017-02-17 submission/manuscript_outline_20161024.tex
start incorporating Niel suggestions

@@
 our study was to test the hypothesis that changes in the bacterial
-microbiome occur after surgery in patients with lesions (i.e.~adenoma or
-carcinoma). Specifically, we wanted to identify what within the
-community was different before and after surgical removal of said
-lesion.
+microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
+Specifically, we wanted to identify what within the community was
+different before and after removal of said lesion.
 
 \textbf{Results:} The bacterial microbiome in pre and post surgery
-samples of individuals with adenoma were more similar to each other than
-those with carcinoma (P-value \textless{} 0.05). There was no difference
-in the relative abundance of any OTU between the pre and post surgery
-samples (P-value \textgreater{} 0.05). A model with a total of 53
-variables was able to classify lesion (AUC = 0.811 - 0.866) while a
-model built to classify pre surgery samples had 70 variables (AUC =
-0.641 - 0.805). The post surgery sample for both models had a decrease
-in the positive probability for either lesion or pre surgery sample
-(P-value \textless{} 0.05). In total there were 23 OTUs that were common
-to both models and the majority of these classified to commensal
-bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales}, \emph{Blautia},
-and \emph{Ruminococcaceae}).
+samples for the adenoma group were more similar to each other than the
+carcinoma group (P-value \textless{} 0.05). There was no difference for
+either the adenoma or carcinoma group in the relative abundance of any
+OTU between the pre and post lesion removal (P-value \textgreater{}
+0.05). A model built to classify model built to classify lesion had an
+AUC range of 0.811 - 0.866 while a model built to classify initial
+versus follow up samples had an AUC range of 0.641 - 0.805. The post
+removal sample for both models had a decrease in the positive
+probability for either lesion or initial sample (P-value \textless{}
+0.05). The lesion model used a total of 53 variables while the initial
+sample model used a total of 70 variables. A total of 23 OTUs were
+common to both models with the majority of these classifying to
+commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
+\emph{Blautia}, and \emph{Ruminococcaceae}).
 
 \textbf{Conclusions:} Our data supports the hypothesis that there are
-differences in the bacterial microbiome between pre and post surgical
-samples. With individuals with carcinoma having more drastic differences
-to the overall community then those with adenoma. Changes to commensal
-bacteria were some of the most important variables for model
+differences in the bacterial microbiome between pre and post lesion
+removal samples. With individuals with carcinoma having more drastic
+differences to the overall community then those with adenoma. Changes to
+commensal bacteria were some of the most important variables for model
 classification, suggesting that these bacteria may be central to initial
@@
 Recently, there has been promising work on the bacterial microbiome and
-it's ability to be able to complement existing screening methods such as
-Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
-{[}3,4{]}. There has also been research into how this microbiome could
-be altered directly on tumor tissue itself {[}5{]}. A few studies have
-also shown how this microbiome {[}6{]} or specific members within it
-{[}7{]} could be directly involved with the pathogenesis of CRC. These
-studies have helped to provide a tantalizing link between the bacterial
+it's ability to complement existing screening methods such as Fecal
+Immunoglobulin Test (FIT) or act alone as a screening tool {[}3,4{]}.
+There has also been research into how the bacterial microbiome could be
+altered directly on tumor tissue itself {[}5{]}. A few studies have also
+shown how this microbiome {[}6{]} or specific members within it {[}7{]}
+could be directly involved with the pathogenesis of CRC. These studies
+have helped to provide a tantalizing link between the bacterial
 microbiome and CRC. Although these studies suggest that the bacterial
@@
 In this study we tested the hypothesis that the bacterial microbiome
-changes between pre (initial) and post (follow up) samples after surgery
-for individuals with a lesion (adenoma or carcinoma). Our analysis
-included both alpha and beta diversity analysis along with investigation
-of individual operational taxonomic units (OTUs). We also utilized
-Random Forest models and observed how these models, as well as specific
-OTUs within them, performed pre and post surgery. We also investigated
-the models for similar important OTUs to identify the crucial OTUs for
-not only classifying initial and follow up samples but also lesion or
-normal.
+changes between pre (initial) and post (follow up) samples after removal
+of a lesion (adenoma or carcinoma). Our analysis included both alpha and
+beta diversity analysis along with investigation of individual
+operational taxonomic units (OTUs). We also utilized Random Forest to
+build models to classify either initial and follow up samples or lesion
+and normal samples and then subsequently observe how these models, as
+well as specific OTUs within them, performed pre and post surgery. We
+also investigated the models for similar important OTUs to identify the
+crucial OTUs for not only classifying initial and follow up samples but
+also lesion or normal.
 
@@
 \emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
-stomatis} (Otu001682). There was a difference in relative abundance in
+stomatis} (Otu001682). There was a decrease in relative abundance in
 initial and follow up samples for lesion and carcinoma for
@@
 
-\textbf{\emph{Full and Reduced Model:}} We next wanted to identify if
-there were any common bacterial microbiome changes in individuals with
-adenoma or carcinoma. In order to investigate this we created two
-different models: one to classify lesion versus normal and one to
-classify pre (initial) versus post (follow up) samples based on the
-bacterial community and FIT measurements. The lesion model had an AUC
-range of 0.73 to 0.797 while the initial sample model had an AUC range
-of 0.485 to 0.686 after 100 iterations of 20 repeated 10-fold cross
-validations. By identifying the most important variables for each
-respective model and then reducing them to only these factors we were
-able to increase the AUC in the lesion model (0.811 - 0.866) and initial
-sample model (0.641 - 0.805).\\
+\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
+were any common bacterial microbiome changes in individuals with adenoma
+or carcinoma. We first investigated this by creating a model to classify
+lesion versus normal based on the bacterial community and FIT
+measurements. This model had an AUC range of 0.811 - 0.866 after 100
+iterations of 20 repeated 10-fold cross validations. The ROC curve for
+the final lesion model used was within the observed range of the 100
+different iterations {[}Figure 3a{]}. There were a total of 53 variables
+{[}Figure S2{]}. For this model FIT measurement resulted in the largest
+decrease in MDA while the OTU with the largest MDA was Otu000251
+{[}Figure S2a{]}
+
+In order to investigate this we created two different models: one to
+classify lesion versus normal and one to classify pre (initial) versus
+post (follow up) samples based on the bacterial community and FIT
+measurements. T while the initial sample model had an AUC range of 0.485
+to 0.686 after 100 iterations of 20 repeated 10-fold cross validations.
+By identifying the most important variables for each respective model
+and then reducing them to only these factors we were able to increase
+the AUC in the lesion model and initial sample model (0.641 - 0.805).\\
 The test set AUC range for the full and reduced lesion model were
 similar to that reported for the training set AUC ranges and the ROC
-curve ranges overlap with each other {[}Figure 3a{]}. The ROC curve for
-the final lesion model used was within the range of the reduced lesion
-model {[}Figure 3a{]}. Interestingly, the test set AUC range for the
-initial sample model performed much better then the training set AUCs.
-Both the full and reduced initial sample models overlapped with each
-other {[}Figure 3b{]} but there was a marked decrease in the ROC curve
-for the final before sample model used.
+curve ranges overlap with each other {[}Figure 3a{]}. . Interestingly,
+the test set AUC range for the initial sample model performed much
+better then the training set AUCs. Both the full and reduced initial
+sample models overlapped with each other {[}Figure 3b{]} but there was a
+marked decrease in the ROC curve for the final before sample model used.
 
@@
 built based on the most important variables to either classification
-model. For the lesion model there were a total of 53 variables {[}Figure
-S2{]} whereas for the initial sample model there were a total of 70
+model. whereas for the initial sample model there were a total of 70
 variables {[}Figure S3{]}. For both models FIT measurement resulted in
@@
 in the carcinoma group. It should be noted that whether this change was
-towards what wouldbe be expected for normal controls is not known.
+towards what would be be expected for normal controls is not known.
 
@@
 for the two different models used. Finally, both lesion and initial
-sample models initial positive probabilites were tested against the
+sample models initial positive probabilities were tested against the
 follow up positive probabilities.
@@
 change from initial to follow up sample if those with adenoma or Screen
-Relevent Neoplasia (SRN). D) Initial positive probability change from
+Relevant Neoplasia (SRN). D) Initial positive probability change from
 initial to follow up sample in those with adenoma or SRN. The dotted
@@
 change from initial to follow up sample in those with adenoma or Screen
-Relevent Neoplasia (SRN). D) Initial positive probability change from
+Relevant Neoplasia (SRN). D) Initial positive probability change from
 initial to follow up sample in those with adenoma or SRN. The dotted

### 2017-02-17 submission/manuscript_outline_20161024.Rmd

@@
 **Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.
-The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after surgery in patients with lesions (i.e.
-adenoma or carcinoma).
-Specifically, we wanted to identify what within the community was different before and after surgical removal of said lesion.
+The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e.
+adenoma or carcinoma) removal.
+Specifically, we wanted to identify what within the community was different before and after removal of said lesion.
 
-**Results:** The bacterial microbiome in pre and post surgery samples of individuals with adenoma were more similar to each other than those with carcinoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
-There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify pre surgery samples had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-The post surgery sample for both models had a decrease in the positive probability for either lesion or pre surgery sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
-In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (e.g.
+**Results:** The bacterial microbiome in pre and post surgery samples for the adenoma group were more similar to each other than the carcinoma group (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
+A model built to classify model built to classify lesion had an AUC range of `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`.
+The post removal sample for both models had a decrease in the positive probability for either lesion or initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
+A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g.
 *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).
 
-**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.
+**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post lesion removal samples.
 With individuals with carcinoma having more drastic differences to the overall community then those with adenoma.
@@
 
-Recently, there has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
+Recently, there has been promising work on the bacterial microbiome and it's ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
+There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
 A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
@@
 
-In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after surgery for individuals with a lesion (adenoma or carcinoma).
+In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after removal of a lesion (adenoma or carcinoma).
 Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest models and observed how these models, as well as specific OTUs within them, performed pre and post surgery.
+We also utilized Random Forest to build models to classify either initial and follow up samples or lesion and normal samples and then subsequently observe how these models, as well as specific OTUs within them, performed pre and post surgery.
 We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
@@
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-There was a difference in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05).
+There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05).
 In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2].
@@
 
-***Full and Reduced Model:***
-We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma.
-In order to investigate this we created two different models: one to classify lesion versus normal and one to classify pre (initial) versus post (follow up) samples based on the bacterial community and FIT measurements.
-The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while the initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
-By identifying the most important variables for each respective model and then reducing them to only these factors we were able to increase the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial sample model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).
-The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap with each other [Figure 3a].
-The ROC curve for the final lesion model used was within the range of the reduced lesion model [Figure 3a].
-Interestingly, the test set AUC range for the initial sample model performed much better then the training set AUCs.
-Both the full and reduced initial sample models overlapped with each other [Figure 3b] but there was a marked decrease in the ROC curve for the final before sample model used.
+***The Lesion Model:***
+We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal control individuals.
+We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.
+This model had an AUC range of `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
+The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3a].
+There were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2].
+For this model FIT measurement resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure S2a].
+
+If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.
+This is what we observed for the lesion model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+The lesion model was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4].
+
+***The Initial Sample Model:***
+After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.
+The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
+The test set AUC range for this model performed better then the training set AUCs.
+There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3b].
+
+If there were common OTUs that could separate initial from follow up sample regardless of whether the lesion was adenoma or carcinoma we would expect to find a decrease in the positive probability of the follow up sample to be an initial sample.
+This is what we observed for the initial sample model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
 
@@
 The reduced models were built based on the most important variables to either classification model.
-For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
+whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
 For both models FIT measurement resulted in one of the largest decreases in MDA [Figure S2a & S3a].
@@
 Although there was a detectable change in the carcinoma group.
-It should be noted that whether this change was towards what wouldbe be expected for normal controls is not known.
+It should be noted that whether this change was towards what would be be expected for normal controls is not known.
 
@@
 A list of common OTUs were found for the two different models used.
-Finally, both lesion and initial sample models initial positive probabilites were tested against the follow up positive probabilities.
+Finally, both lesion and initial sample models initial positive probabilities were tested against the follow up positive probabilities.
 
@@
 
-**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevent Neoplasia (SRN).
+**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).
 D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.
@@
 
-**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevent Neoplasia (SRN).
+**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevant Neoplasia (SRN).
 D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.

### 2017-02-18 submission/manuscript_outline_20161024.Rmd
complete Niel edits

@@
 
-***Cancer Associated Bacteria:***
+***Carcinoma Associated Bacteria:***
 Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].
@@
 There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3b].
+There were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S2].
+For this model *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement resulted in the second largest decrease in MDA [Figure S3a].
 
@@
 This is what we observed for the initial sample model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+When we separated individuals based on whether they had an adenoma or carcinoma there was a decrease in positive probability for both the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+For this model there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 4].
 
 ***Common OTUs to both Models:***
-The reduced models were built based on the most important variables to either classification model.
-whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].
-For both models FIT measurement resulted in one of the largest decreases in MDA [Figure S2a & S3a].
-When we compared the two different reduced models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.
+We next wanted to compare the similarity between the OTU variables used in either model.
+The main purpose was to identify which OTUs were important not only for the classification of lesion but also for the classification of initial or follow up sample.
+Potentially, these specific OTUs are the most important with respect to the bacterial microbiome response to removal of lesion.
+When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.
 Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.
@@
 
-***Positive Probability after Lesion Removal:***
-If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample.
-This is what we observed regardless of model used (lesion or initial sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-
-When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
-We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-This lack of difference between the predicted and actual classifications were also observed for the initial sample model (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-Even though the lesion model was not as accurate in classifying samples as the initial sample model.
-It was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].
-
-***Treatment and Time Differences:***
+***Treatment Differences:***
 After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
-There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value `r ifelse(chemo_rad_summary['chemo_les', 'bh'] < 0.05 & chemo_rad_summary['chemo_red_les', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05) or radiation therapy (P-value `r ifelse(chemo_rad_summary['rad_les', 'bh'] < 0.05 & chemo_rad_summary['rad_red_les', 'bh'] < 0.05 , paste("<"), paste(">"))` 0.05).
-In contrast, we observed for the the before sample model a significant difference in decreased positive probability for those treated with chemotherapy (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
+There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
 All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].
-Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.
-Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
 
@@
 ## Discussion
+
 From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
@@
 There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].
-Although there was a detectable change in the carcinoma group.
-It should be noted that whether this change was towards what would be be expected for normal controls is not known.
+Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.
 
-Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested cancer causing microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
+Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested carcinoma associated microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
 This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.
 
-We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.
-What we found was that the commonly associated CRC bacteria were not highly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
-With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
-Using only these important OTUs and FIT, both models (lesion and initial sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].
-Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community [Table S3].
+We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).
+What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 4 & S4] but also that the commonly associated CRC bacteria were not highly represented within this model.
+Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
+With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
+We followed this up by creating a second model to classify initial versus follow up samples.
+We found that this model was able to accurately classify initial versus follow up samples [Figure 4 & S4].
+Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].
 
 For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].
-There was a difference in the length of time between initial and follow up sample between adenoma and carcinoma.
+Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
 These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
+For time differences, specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].
 
-This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.
+This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after surgical removal of a lesion.
 Based on previous work by Arthur, et al.
@@
 
-Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
+Curiously, we observed that the typical CRC associated bacteria were not predictive ewithin our models.
 There are a number of reasons why this may have occurred.

### 2017-02-18 submission/manuscript_outline_20161024.tex
complete Niel edits

@@
 
-\textbf{\emph{Cancer Associated Bacteria:}} Previous literature has
+\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
 suggested that a number of oral microbes may be important in CRC
@@
 were any common bacterial microbiome changes in individuals with adenoma
-or carcinoma. We first investigated this by creating a model to classify
-lesion versus normal based on the bacterial community and FIT
-measurements. This model had an AUC range of 0.811 - 0.866 after 100
-iterations of 20 repeated 10-fold cross validations. The ROC curve for
-the final lesion model used was within the observed range of the 100
-different iterations {[}Figure 3a{]}. There were a total of 53 variables
-{[}Figure S2{]}. For this model FIT measurement resulted in the largest
-decrease in MDA while the OTU with the largest MDA was Otu000251
-{[}Figure S2a{]}
+or carcinoma versus normal control individuals. We investigated this by
+creating a model to classify lesion versus normal based on the bacterial
+community and FIT measurements. This model had an AUC range of 0.811 -
+0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
+ROC curve for the final lesion model used was within the observed range
+of the 100 different test set AUC iterations {[}Figure 3a{]}. There were
+a total of 53 variables {[}Figure S2{]}. For this model FIT measurement
+resulted in the largest decrease in MDA while the OTU with the largest
+MDA was \emph{Ruminococcaceae} (Otu000251) {[}Figure S2a{]}.
 
-In order to investigate this we created two different models: one to
-classify lesion versus normal and one to classify pre (initial) versus
-post (follow up) samples based on the bacterial community and FIT
-measurements. T while the initial sample model had an AUC range of 0.485
-to 0.686 after 100 iterations of 20 repeated 10-fold cross validations.
-By identifying the most important variables for each respective model
-and then reducing them to only these factors we were able to increase
-the AUC in the lesion model and initial sample model (0.641 - 0.805).\\
-The test set AUC range for the full and reduced lesion model were
-similar to that reported for the training set AUC ranges and the ROC
-curve ranges overlap with each other {[}Figure 3a{]}. . Interestingly,
-the test set AUC range for the initial sample model performed much
-better then the training set AUCs. Both the full and reduced initial
-sample models overlapped with each other {[}Figure 3b{]} but there was a
-marked decrease in the ROC curve for the final before sample model used.
+If there were common OTUs that could separate adenoma and carcinoma from
+normal controls, we would expect to find a decrease in the positive
+probability of the follow up sample to be a lesion. This is what we
+observed for the lesion model {[}Figure 4{]} (P-value \textless{}
+0.001). When we separated individuals based on whether they had an
+adenoma or carcinoma there was only a decrease in positive probability
+for the carcinoma group (P-value \textless{} 0.001) and not for the
+adenoma group (P-value \textgreater{} 0.05). We also observed that there
+were no significant differences in whether the models classified the
+samples as having lesion between the predicted and actual (P-value
+\textgreater{} 0.05). The lesion model was able to correctly classify
+the one individual who still had a carcinoma on follow up {[}Figure
+4{]}.
 
-\textbf{\emph{Common OTUs to both Models:}} The reduced models were
-built based on the most important variables to either classification
-model. whereas for the initial sample model there were a total of 70
-variables {[}Figure S3{]}. For both models FIT measurement resulted in
-one of the largest decreases in MDA {[}Figure S2a \& S3a{]}. When we
-compared the two different reduced models with each other there were a
-total of 23 common OTUs. Some of the most common taxonomic
+\textbf{\emph{The Initial Sample Model:}} After building a model to
+classify based on lesion we wanted to then build a separate model
+specifically to be able to classify whether samples were initial or
+follow up samples based on the bacterial community and FIT measurements.
+The initial sample model had an AUC range of 0.485 to 0.686 after 100
+iterations of 20 repeated 10-fold cross validations. The test set AUC
+range for this model performed better then the training set AUCs. There
+was a marked decrease in the ROC curve for the final model used when
+compared to the 100 test set AUC iterations {[}Figure 3b{]}. There were
+a total of 70 variables {[}Figure S2{]}. For this model
+\emph{Ruminococcaceae} (Otu000278) resulted in the largest decrease in
+MDA while FIT measurement resulted in the second largest decrease in MDA
+{[}Figure S3a{]}.
+
+If there were common OTUs that could separate initial from follow up
+sample regardless of whether the lesion was adenoma or carcinoma we
+would expect to find a decrease in the positive probability of the
+follow up sample to be an initial sample. This is what we observed for
+the initial sample model {[}Figure 4{]} (P-value \textless{} 0.001).
+When we separated individuals based on whether they had an adenoma or
+carcinoma there was a decrease in positive probability for both the
+carcinoma group (P-value \textless{} 0.001) and for the adenoma group
+(P-value \textless{} 0.001). For this model there was no difference
+between the predicted and actual classifications (P-value \textgreater{}
+0.05) {[}Figure 4{]}.
+
+\textbf{\emph{Common OTUs to both Models:}} We next wanted to compare
+the similarity between the OTU variables used in either model. The main
+purpose was to identify which OTUs were important not only for the
+classification of lesion but also for the classification of initial or
+follow up sample. Potentially, these specific OTUs are the most
+important with respect to the bacterial microbiome response to removal
+of lesion. When we compared the two different models with each other
+there were a total of 23 common OTUs. Some of the most common taxonomic
 identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
@@
 
-\textbf{\emph{Positive Probability after Lesion Removal:}} If there were
-common OTUs that could separate adenoma and carcinoma from normal
-controls, we would expect to find a decrease in the positive probability
-of the follow up sample to be either lesion or an initial sample. This
-is what we observed regardless of model used (lesion or initial sample)
-or whether it was built on the full or reduced variable data set
-{[}Figure 4 \& S4{]} (P-value \textless{} 0.001).
-
-When we separated individuals based on whether they had an adenoma or
-carcinoma there was only a decrease in positive probability for the
-carcinoma group (P-value \textless{} 0.001) and not for the adenoma
-group (P-value \textgreater{} 0.05). We also observed that there were no
-significant differences in whether the models classified the samples as
-having lesion between the predicted and actual (P-value \textgreater{}
-0.05). This lack of difference between the predicted and actual
-classifications were also observed for the initial sample model (P-value
-\textgreater{} 0.05). Even though the lesion model was not as accurate
-in classifying samples as the initial sample model. It was able to
-correctly classify the one individual who still had a carcinoma on
-follow up {[}Figure 4a \& S4a{]} while the initial sample model did not
-{[}Figure 4b \& S4b{]}.
-
-\textbf{\emph{Treatment and Time Differences:}} After observing these
-changes in the bacterial community and positive probability we wanted to
-assess whether additional treatments, such as chemotherapy and
-radiation, could have an impact on the results that we observed. There
-was no difference in the amount of change in positive probability for
-either the full or reduced lesion model for either chemotherapy (P-value
-\textgreater{} 0.05) or radiation therapy (P-value \textgreater{} 0.05).
-In contrast, we observed for the the before sample model a significant
-difference in decreased positive probability for those treated with
-chemotherapy (P-value \textless{} 0.05). All other variables that were
-tested showed no difference based on whether chemotherapy or radiation
-was used {[}Table S3{]}. Finally, we wanted to know if the length of
-time between the initial and follow up sample could be a possible
-confounder. Within our study there was a significant difference for the
-time elapsed in the collection of the follow up sample between adenoma
-and carcinoma (uncorrected P-value \textless{} 0.05), with time passed
-being less for adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102
-days).
+\textbf{\emph{Treatment Differences:}} After observing these changes in
+the bacterial community and positive probability we wanted to assess
+whether additional treatments, such as chemotherapy and radiation, could
+have an impact on the results that we observed. There was only a
+significant difference for change in positive probability for those
+treated with chemotherapy for the initial sample model (P-value
+\textless{} 0.05). All other variables that were tested showed no
+difference based on whether chemotherapy or radiation was used {[}Table
+S3{]}.
 
@@
 carcinoma only {[}Figure s1{]}. Although there was a detectable change
-in the carcinoma group. It should be noted that whether this change was
-towards what would be be expected for normal controls is not known.
+in the carcinoma group and this change was towards what would be
+expected for normal controls, the follow up samples may not be a
+completely normal bacterial microbiome.
 
@@
 looking specifically at four OTUs that taxonomically classified to
-previously suggested cancer causing microbes we found that only 2/4 had
-a decrease in relative abundance between initial and follow up for those
-with carcinoma and 0/4 had differences for those with adenoma. This data
-would suggest that these specific OTUs may be important in the
+previously suggested carcinoma associated microbes we found that only
+2/4 had a decrease in relative abundance between initial and follow up
+for those with carcinoma and 0/4 had differences for those with adenoma.
+This data would suggest that these specific OTUs may be important in the
 transition of an adenoma to a carcinoma but less so in the initiation of
@@
 We next created a model that incorporated FIT and the bacterial
-microbiome to either be able to classify lesions (adenoma or carcinoma)
-or initial samples in order to find common OTUs in the community that
-change for both adenoma and carcinoma. What we found was that the
-commonly associated CRC bacteria were not highly represented within our
-models but rather that OTUs that made up the most important variables
-overwhelmingly belonged to commensal bacteria. With only the lesion
-model having a single OTU from a previously associated cancer bacterium
-(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
-and FIT, both models (lesion and initial sample) significantly decreased
-positive probability of either lesion or being an initial sample on
-follow up {[}Figure 4 \& S4{]}. Further confirmation of the importance
-of the changes of commensal bacteria to these classifications was that a
-total of 23 OTUs were common to both models and the vast majority
-belonged to regular residents of our gut community {[}Table S3{]}.
+microbiome to be able to classify lesions (adenoma or carcinoma). What
+we found was that not only was the positive probability of a lesion
+reduced after lesion removal {[}Figure 4 \& S4{]} but also that the
+commonly associated CRC bacteria were not highly represented within this
+model. Rather, OTUs that made up the most important variables
+overwhelmingly belonged to commensal bacteria. With only a single OTU
+from a previously associated cancer bacterium (\emph{Porphyromonas
+asaccharolytica}). We followed this up by creating a second model to
+classify initial versus follow up samples. We found that this model was
+able to accurately classify initial versus follow up samples {[}Figure 4
+\& S4{]}. Providing additional information on the importance of
+commensal bacteria was that there were a total of 23 OTUs in common to
+both models and the vast majority belonged to regular residents of our
+gut community {[}Table S3{]}.
 
@@
 bacterial microbiome based on whether chemotherapy or radiation was
-received {[}Table S3{]}. There was a difference in the length of time
-between initial and follow up sample between adenoma and carcinoma.
-These results would indicate that the findings described were specific
-to the surgical intervention and that some of the differences observed
-between carcinoma and adenoma samples could be due to differences in
-collection time between samples for the two different groups.
+received {[}Table S3{]}. Within our study there was a significant
+difference for the time elapsed in the collection of the follow up
+sample between adenoma and carcinoma (uncorrected P-value \textless{}
+0.05), with time passed being less for adenoma (253 +/- 41.3 days) then
+carcinoma (351 +/- 102 days). These results would indicate that the
+findings described were specific to the surgical intervention and that
+some of the differences observed between carcinoma and adenoma samples
+could be due to differences in collection time between samples for the
+two different groups. For time differences, specifically, it could
+confound the observation that carcinomas changed more than adenomas
+{[}Figure 1a \& 1d{]}.
 
 This study builds upon previous work from numerous labs that have looked
-into the bacterial microbiome as a potential screening tool {[}3,4{]} by
-exploring what happens to the bacterial community after surgical removal
-of a lesion. Based on previous work by Arthur, et al. {[}9{]} it may not
-be surprising to have E.coli as one of the most important OTUs and one
-that was common to both models. Interestingly, many of the most
-important OTUs had taxonomic identification for resident gut microbes.
-This could suggest that the bacterial community is one of the first
-components that could change during the pathogenesis of disease. These
-bacterial microbiome changes could be the first step in allowing more
-inflammatory bacterium to gain a foothold within the colon {[}8{]}.
+into the bacterial microbiome as a potential screening tool
+{[}3,4,9--11{]} by exploring what happens to the bacterial community
+after surgical removal of a lesion. Based on previous work by Arthur, et
+al. {[}12{]} it may not be surprising to have E.coli as one of the most
+important OTUs and one that was common to both models. Interestingly,
+many of the most important OTUs had taxonomic identification for
+resident gut microbes. This could suggest that the bacterial community
+is one of the first components that could change during the pathogenesis
+of disease. These bacterial microbiome changes could be the first step
+in allowing more inflammatory bacterium to gain a foothold within the
+colon {[}8{]}.
 
 Curiously, we observed that the typical CRC associated bacteria were not
-predictive within our models. There are a number of reasons why this may
-have occurred. First, is that they were not present in enough
+predictive ewithin our models. There are a number of reasons why this
+may have occurred. First, is that they were not present in enough
 individuals to be able to classify those with and without disease with a
@@
 polyphenols are thought to be protective against cancer in part by
-reducing inflammation {[}10{]}. These protective compounds are derived
+reducing inflammation {[}13{]}. These protective compounds are derived
 from the breakdown of fiber, fruits, and vegetables by resident gut
@@
 which was highly prevalent in our models, are known to be increased in
-those with high non-meat based protein consumption {[}11{]}. High
+those with high non-meat based protein consumption {[}14{]}. High
 protein consumption in general has been linked with an increased CRC
-risk {[}12{]}. Conversely, \emph{Bacteroides} are inhibited by
-polyphenols which are derived from fruits and vegetables {[}13{]}. Our
+risk {[}15{]}. Conversely, \emph{Bacteroides} are inhibited by
+polyphenols which are derived from fruits and vegetables {[}16{]}. Our
 data fits with the hypothesis that the microbial metabolites from
@@
 longer having a lesion. Another limitation is that we do not know if
-adding modern tests such as the stool DNA test {[}14{]} could help
+adding modern tests such as the stool DNA test {[}17{]} could help
 improve our overall AUC. This study also drew heavily from those with
@@
 prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
-microbiome has been shown to normalize within this time period {[}15{]}.
+microbiome has been shown to normalize within this time period {[}18{]}.
 Our follow up data set had a total of 67 individuals that not only had a
@@
 previously published using both OC FIT-CHEK and OC-Auto Micro 80
-automated system (Polymedco Inc.) {[}16{]}. 16S rRNA gene sequencing was
-completed as previously described by Kozich, et al. {[}17{]}. DNA
+automated system (Polymedco Inc.) {[}19{]}. 16S rRNA gene sequencing was
+completed as previously described by Kozich, et al. {[}20{]}. DNA
 extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described {[}17{]}. The general processing workflow
+has been previously described {[}20{]}. The general processing workflow
 using mothur was as follows: Paired-end reads were first merged into
@@
 
-\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}18{]}
+\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
 algorithm was used to create the model used for prediction of lesion
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}19{]} in the vegan package
+community structure utilized PERMANOVA {[}22{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate a Benjamini-Hochberg (BH) correction was applied {[}20{]}
+appropriate a Benjamini-Hochberg (BH) correction was applied {[}23{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
+\hypertarget{ref-yu_metagenomic_2017}{}
+9. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
+analysis of faecal microbiome as a tool towards targeted non-invasive
+biomarkers for colorectal cancer. Gut. 2017;66:70--8.
+
+\hypertarget{ref-zackular_human_2014}{}
+10. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+microbiome as a screening tool for colorectal cancer. Cancer Prevention
+Research (Philadelphia, Pa.). 2014;7:1112--21.
+
+\hypertarget{ref-warren_co-occurrence_2013}{}
+11. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
+et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
+Microbiome. 2013;1:16.
+
 \hypertarget{ref-arthur_intestinal_2012}{}
-9. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+12. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
 T-J, et al. Intestinal inflammation targets cancer-inducing activity of
@@
 \hypertarget{ref-louis_gut_2014}{}
-10. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
+13. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
 metabolites and colorectal cancer. Nature Reviews Microbiology
@@
 \hypertarget{ref-zhu_intake_2016}{}
-11. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
+14. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
 Proteins Substantially Increased the Relative Abundance of Genus
@@
 \hypertarget{ref-mu_colonic_2016}{}
-12. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
+15. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
 Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
@@
 \hypertarget{ref-ozdal_reciprocal_2016}{}
-13. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
+16. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
 Reciprocal Interactions between Polyphenols and Gut Microbiota and
@@
 \hypertarget{ref-cotter_long-term_2016}{}
-14. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
+17. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
 et al. Long-Term Follow-up of Patients Having False Positive
@@
 \hypertarget{ref-obrien_impact_2013}{}
-15. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-daly_evaluation_2013}{}
-16. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
+19. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
 for colorectal cancer screening. Journal of Primary Care \& Community
@@
 \hypertarget{ref-kozich_development_2013}{}
-17. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+20. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-18. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+21. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-19. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+22. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-20. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+23. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-02-20 code/Run_Figure4.R
merge and remove figs to streamline

@@ -1,6 +1,5 @@
-### Prediction of Follow Ups Analysis
-### How do the with and without FIT models do in correctly calling initial and follow up samples
-### P-value table of comparison and Figure 4 graph
+### Create Figure 4
+### Graph for the Initial Sample Model
 ## Marc Sze
 
 
@@ -8,138 +7,83 @@
 source('code/functions.R')
 
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
-           "gridExtra", "scales", "wesanderson", "caret"))
+           "gridExtra", "scales", "wesanderson"))
 
-# Load lesion model data
-follow_up_probability <- read.csv(
-  "results/tables/follow_up_probability_summary.csv", 
-  header = T, stringsAsFactors = F)
 
-red_follow_up_probability <- read.csv("results/tables/reduced_follow_up_probability_summary.csv", 
-                                      stringsAsFactors = F, header = T)
+### Load in needed data tables
+reduced_IF_model_roc <- read.csv("results/tables/reduced_IF_test_data_roc.csv", header = T)
+IF_MDA_data_summary <- read.csv("results/tables/reduced_IF_model_top_vars_MDA_Summary.csv", 
+                                header = T, stringsAsFactors = F)
 
-# Load IF model data
-IF_follow_up_probability <- read.csv(
-  "results/tables/IF_follow_up_probability_summary.csv", 
-  header = T, stringsAsFactors = F)
+IF_MDA_full <- read.csv("results/tables/reduced_IF_model_top_vars_MDA.csv", 
+                        header = T, stringsAsFactors = F)
 
 IF_red_follow_up_probability <- read.csv("results/tables/reduced_IF_follow_up_probability_summary.csv", 
-                                      stringsAsFactors = F, header = T)
+                                         stringsAsFactors = F, header = T)
 
+tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
 
-# Read in meta data tables
-good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", 
-                       stringsAsFactors = F, header = T) %>% 
-filter(!is.na(fit_followUp))
+# Convert taxa table to a data frame with columns for each taxonomic division
[…]

### 2017-02-20 code/Run_FigureS2.R

@@ -1,96 +1,22 @@
-### Figure S2
-### Description of variables in the lesion model
+### Figure S5
+### Distribution of P-values from Paired Wilcoxson Analysis
 ## Marc Sze
 
 #Load needed code and packages
 source('code/functions.R')
 
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
-           "gridExtra", "scales", "wesanderson"))
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
 #Read data needed
-lesion_MDA_data_summary <- read.csv("results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv", 
-                            header = T, stringsAsFactors = F)
+diff_table_time <- read.csv("results/tables/time_datatable.csv", header = T)
 
-lesion_MDA_full <- read.csv("results/tables/reduced_lesion_model_top_vars_MDA.csv", 
-                            header = T, stringsAsFactors = F)
+# Plot the change in distance versus time
+time_graph <- ggplot(diff_table_time, aes(x = time, y = distance)) + 
+  geom_point(aes(color = dx), size = 4) + theme_bw() + 
+  scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
+                     breaks = c("adenoma", "cancer"), labels = c("Adenoma", "Cancer")) + 
+  coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("Time from Initial (Days)") + 
+  theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"))
 
-tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
-
-occurance_data <- read.csv("results/tables/rf_wCV_imp_vars_summary.csv", header = T, stringsAsFactors = F)
-
-# Convert taxa table to a data frame with columns for each taxonomic division
-tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
-rownames(tax_df) <- rownames(tax)
-colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
-tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
-rm(tax)
-
[…]

### 2017-02-20 code/Run_probs_comparison.R

@@ -1,0 +1,108 @@
+### Prediction of Follow Ups Analysis
+### How do the with and without FIT models do in correctly calling initial and follow up samples
+### P-value table of comparison and Figure 4 graph
+## Marc Sze
+
+
+###Load needed Libraries and functions
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson", "caret"))
+
+# Load lesion model data
+follow_up_probability <- read.csv(
+  "results/tables/follow_up_probability_summary.csv", 
+  header = T, stringsAsFactors = F)
+
+red_follow_up_probability <- read.csv("results/tables/reduced_follow_up_probability_summary.csv", 
+                                      stringsAsFactors = F, header = T)
+
+# Load IF model data
+IF_follow_up_probability <- read.csv(
+  "results/tables/IF_follow_up_probability_summary.csv", 
+  header = T, stringsAsFactors = F)
+
+IF_red_follow_up_probability <- read.csv("results/tables/reduced_IF_follow_up_probability_summary.csv", 
+                                      stringsAsFactors = F, header = T)
+
+
+# Read in meta data tables
+good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", 
+                       stringsAsFactors = F, header = T) %>% 
+filter(!is.na(fit_followUp))
+
+
+# create tables to hold wilcoxson paired tests pvalues with BH correction
+lesion_wilcox_pvalue_summary <- getProb_PairedWilcox(follow_up_probability)
+lesion_red_wilcox_pvalue_summary <- getProb_PairedWilcox(red_follow_up_probability)
+IF_wilcox_pvalue_summary <- getProb_PairedWilcox(IF_follow_up_probability)
+IF_red_wilcox_pvalue_summary <- getProb_PairedWilcox(IF_red_follow_up_probability)
+
+all_wilcox_summary <- rbind(lesion_wilcox_pvalue_summary, 
+                    lesion_red_wilcox_pvalue_summary, 
+                    IF_wilcox_pvalue_summary, 
+                    IF_red_wilcox_pvalue_summary)
+
+all_wilcox_summary <- as.data.frame(all_wilcox_summary) %>% 
+  mutate(model_type = c(rep("lesion", 4), rep("red_lesion", 4), rep("IF", 4), rep("red_IF", 4))) %>% 
[…]

### 2017-02-20 code/old/Run_FigureS2.R

@@ -1,0 +1,96 @@
+### Figure S2
+### Description of variables in the lesion model
+## Marc Sze
+
+#Load needed code and packages
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson"))
+
+#Read data needed
+lesion_MDA_data_summary <- read.csv("results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv", 
+                            header = T, stringsAsFactors = F)
+
+lesion_MDA_full <- read.csv("results/tables/reduced_lesion_model_top_vars_MDA.csv", 
+                            header = T, stringsAsFactors = F)
+
+tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
+
+occurance_data <- read.csv("results/tables/rf_wCV_imp_vars_summary.csv", header = T, stringsAsFactors = F)
+
+# Convert taxa table to a data frame with columns for each taxonomic division
+tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
+rownames(tax_df) <- rownames(tax)
+colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
+tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
+rm(tax)
+
+# Order the data file from most to least important based on mean MDA
+lesion_MDA_data_summary <- lesion_MDA_data_summary[order(lesion_MDA_data_summary$mean_MDA, decreasing = TRUE), ]
+
+# Pull OTUs that are only in the MDA data
+OTU_IDs <- unique(filter(lesion_MDA_data_summary, variable != "fit_result")[, "variable"])
+select_tax_df <- tax_df[OTU_IDs, ]
+low_tax_ID <- gsub("2", "", gsub("_unclassified", "", createTaxaLabeller(select_tax_df)))
+
+# create labels for factor values with low taxonomy
+graph_labels <- c("FIT", paste(gsub("2", "", low_tax_ID), " (", names(low_tax_ID), ")", sep = ""))
+OTU_names <- names(low_tax_ID)
+
+test <- c()
+for(i in 1:length(low_tax_ID)){
+  
[…]

### 2017-02-20 code/old/Run_FigureS3.R

@@ -1,0 +1,100 @@
+### Figure S3
+### Description of variables in IF models
+## Marc Sze
+
+
+#Load needed code and packages
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson"))
+
+#Read data needed
+IF_MDA_data_summary <- read.csv("results/tables/reduced_IF_model_top_vars_MDA_Summary.csv", 
+                                    header = T, stringsAsFactors = F)
+
+IF_MDA_full <- read.csv("results/tables/reduced_IF_model_top_vars_MDA.csv", 
+                            header = T, stringsAsFactors = F)
+
+tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
+
+occurance_data <- read.csv("results/tables/IF_rf_wCV_imp_vars_summary.csv", header = T, stringsAsFactors = F)
+
+# Convert taxa table to a data frame with columns for each taxonomic division
+tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
+rownames(tax_df) <- rownames(tax)
+colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
+tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
+rm(tax)
+
+# Order the data file from most to least important based on mean MDA
+IF_MDA_data_summary <- IF_MDA_data_summary[order(IF_MDA_data_summary$mean_MDA, decreasing = TRUE), ]
+
+# Pull OTUs that are only in the MDA data
+OTU_IDs <- unique(IF_MDA_data_summary$variable)
+select_tax_df <- tax_df[OTU_IDs, ]
+low_tax_ID <- gsub("2", "", gsub("_unclassified", "", createTaxaLabeller(select_tax_df)))
+
+# create labels for factor values with low taxonomy
+graph_labels <- c(paste(gsub("2", "", low_tax_ID), " (", names(low_tax_ID), ")", sep = ""))
+OTU_names <- names(low_tax_ID)
+
+test <- c()
+for(i in 1:length(low_tax_ID)){
+  if(is.na(low_tax_ID[i])){
+    
+    test <- c(test, bquote(paste("FIT", sep = "")))
+  } else {
+    
[…]

### 2017-02-20 code/old/Run_FigureS4.R

@@ -1,0 +1,129 @@
+### Figure S4
+### How do the full models of lesion and IF look
+## Marc Sze
+
+
+###Load needed Libraries and functions
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson", "caret"))
+
+# Load lesion model data
+follow_up_probability <- read.csv(
+  "results/tables/follow_up_probability_summary.csv", 
+  header = T, stringsAsFactors = F)
+
+# Load IF model data
+IF_follow_up_probability <- read.csv(
+  "results/tables/IF_follow_up_probability_summary.csv", 
+  header = T, stringsAsFactors = F)
+
+# Read in meta data tables
+good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", 
+                       stringsAsFactors = F, header = T) %>% 
+  filter(!is.na(fit_followUp))
+
+
+accuracy_plot <- grid.arrange(
+  # Graph the lesion carcinoma data only
+  filter(follow_up_probability, diagnosis != "adenoma") %>% 
+    ggplot(aes(factor(sampleType, 
+                      levels = c("initial", "followup")), Yes, group = factor(EDRN))) + 
+    geom_point(aes(color=factor(disease_free, 
+                                levels = c("n", "y", "unknown"))), size = 2) + 
+    geom_line(linetype = 2, alpha = 0.3) + 
+    scale_color_manual(name = "Cancer Free", 
+                       label = c("No", "Yes", "Unknown"),  
+                       values = wes_palette("GrandBudapest")) + 
+    scale_x_discrete(breaks = c("initial", "followup"), 
+                     labels = c("Initial", "Follow Up")) + 
+    coord_cartesian(ylim = c(0, 1)) + 
+    geom_hline(aes(yintercept = 0.5), linetype = 2) + 
+    ggtitle("A") + ylab("Lesion Postive Probability") + 
+    xlab("") + theme_bw() + theme(
+      axis.title = element_text(face="bold"), 
+      legend.title = element_text(face="bold"), 
+      legend.position = c(0.25, 0.15), 
+      plot.title = element_text(face="bold", hjust = 0)), 
+  
+  # Graph the lesion adenoma data only
+  filter(follow_up_probability, diagnosis == "adenoma") %>%
[…]

### 2017-02-20 code/old/Run_deprecated_Figure3.R

@@ -1,0 +1,53 @@
+### Create Figure 3
+### Graph to show AUC of full data with chosen model
+### Graph to show how min, middle, and max performed on 20% test data
+## Marc Sze
+
+
+###Load needed Libraries and functions
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", 
+           "wesanderson", "caret", "pROC"))
+
+### Load in needed data tables
+lesion_model_roc <- read.csv("results/tables/test_data_roc.csv", header = T)
+reduced_lesion_model_roc <- read.csv("results/tables/reduced_test_data_roc.csv", header = T)
+lesion_auc_info <- read.csv("results/tables/auc_summary.csv", 
+                            header = T, row.names = 1)
+
+
+IF_model_roc <- read.csv("results/tables/IF_test_data_roc.csv", header = T)
+reduced_IF_model_roc <- read.csv("results/tables/reduced_IF_test_data_roc.csv", header = T)
+
+roc_curve_graph <- grid.arrange(
+  # Graph showing all variable and reduced variable lesion model
+  filter(lesion_model_roc, run != "middle_roc" & run != "full_roc") %>% 
+    ggplot(aes(x = sensitivities, y = specificities)) + 
+    geom_polygon(alpha = 0.25) + 
+    geom_line(aes(group = run), size = 1.25, color = "black", alpha = 0.5) + 
+    geom_polygon(data = filter(reduced_lesion_model_roc, run != "full_roc"), alpha = 0.25, fill = "red") + 
+    geom_line(data = filter(reduced_lesion_model_roc, run != "full_roc"), 
+              aes(group = run), size = 1.25, color = "red", alpha = 0.5) + 
+    geom_line(data = filter(reduced_lesion_model_roc, run == "full_roc"), 
+              size = 1.5, color = "cornflowerblue") + 
+    scale_x_continuous(trans = "reverse") + theme_bw() + geom_abline(intercept = 1, linetype = 2, size = 1) + 
+    ggtitle("A") + xlab("Sensitivity") + ylab("Specificity") + theme(plot.title = element_text(face= "bold")), 
+  
+  # Graph showing all variable and reduced variable IF model
+  filter(IF_model_roc, run != "full_roc") %>% ggplot(aes(x = sensitivities, y = specificities)) + 
+    geom_polygon(alpha = 0.25) + 
[…]

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
update fig references in results section

@@
 
-# Lesion Model (Full)
-lesion_top10_varsMDA <- read.csv('../results/tables/lesion_model_top_vars_MDA_Summary.csv', header = T)
+# Lesion Model 
+lesion_top10_varsMDA <- read.csv('../results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T)
 lesion_imp_vars <- read.csv('../results/tables/rf_otu_tax.csv', header = T, row.names = 1)
-lesion_AUC_run_summary <- read.csv('../results/tables/ROC_model_summary.csv', header = T)
+lesion_AUC_run_summary <- read.csv('../results/tables/Reduced_ROC_model_summary.csv', header = T)
 
-# Lesion Model (Reduced)
-reduced_lesion_AUC_run_summary <- read.csv('../results/tables/Reduced_ROC_model_summary.csv', header = T)
-
-# IF Model (Full)
-IF_top10_varsMDA <- read.csv('../results/tables/IF_model_top_vars_MDA_Summary.csv', header = T)
+# IF Model
+IF_top10_varsMDA <- read.csv('../results/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T)
 IF_imp_vars <- read.csv('../results/tables/if_rf_otu_tax.csv', header = T, row.names = 1)
-IF_AUC_run_summary <- read.csv('../results/tables/IF_ROC_model_summary.csv', header = T)
-
-# IF Model (Reduced)
-reduced_IF_AUC_run_summary <- read.csv('../results/tables/reduced_IF_ROC_model_summary.csv', header = T)
+IF_AUC_run_summary <- read.csv('../results/tables/reduced_IF_ROC_model_summary.csv', header = T)
 
@@
 There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A model built to classify model built to classify lesion had an AUC range of `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`.
+A model built to classify model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
 The post removal sample for both models had a decrease in the positive probability for either lesion or initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
@@
 These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05).
-In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2].
+There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A].
+In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2B].
 We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].
@@
 We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.
-This model had an AUC range of `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
-The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3a].
-There were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2].
-For this model FIT measurement resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure S2a].
+This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
+The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].
+There were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure 3B].
+For this model FIT measurement resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
 
 If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.
-This is what we observed for the lesion model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+This is what we observed for the lesion model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group [Figure 3D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-The lesion model was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4].
+The lesion model was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 3C].
 
@@
 The test set AUC range for this model performed better then the training set AUCs.
-There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3b].
-There were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S2].
-For this model *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement resulted in the second largest decrease in MDA [Figure S3a].
+There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].
+There were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure 4B].
+For this model *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement resulted in the sixth largest decrease in MDA [Figure 4B].
 
 If there were common OTUs that could separate initial from follow up sample regardless of whether the lesion was adenoma or carcinoma we would expect to find a decrease in the positive probability of the follow up sample to be an initial sample.
-This is what we observed for the initial sample model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-When we separated individuals based on whether they had an adenoma or carcinoma there was a decrease in positive probability for both the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-For this model there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 4].
+This is what we observed for the initial sample model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+When we separated individuals based on whether they had an adenoma or carcinoma there was a decrease in positive probability for both the carcinoma group [Figure 4C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and for the adenoma group [Figure 4D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+For this model there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
 

### 2017-02-20 submission/manuscript_outline_20161024.tex
update fig references in results section

@@
 \emph{Parvimonas micra} (P-value \textless{} 0.05), and
-\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05). In
-contrast, there was no difference in relative abundance in any of these
-OTUs for those with adenoma {[}Figure 2{]}. We also observed that only a
-small percentage of those with adenoma or carcinoma were positive or had
-an appreciable relative abundance of any of these respective OTUs
-{[}Figure 2{]}.
+\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05)
+{[}Figure 2A{]}. In contrast, there was no difference in relative
+abundance in any of these OTUs for those with adenoma {[}Figure 2B{]}.
+We also observed that only a small percentage of those with adenoma or
+carcinoma were positive or had an appreciable relative abundance of any
+of these respective OTUs {[}Figure 2{]}.
 
@@
 ROC curve for the final lesion model used was within the observed range
-of the 100 different test set AUC iterations {[}Figure 3a{]}. There were
-a total of 53 variables {[}Figure S2{]}. For this model FIT measurement
+of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
+a total of 53 variables {[}Figure 3B{]}. For this model FIT measurement
 resulted in the largest decrease in MDA while the OTU with the largest
-MDA was \emph{Ruminococcaceae} (Otu000251) {[}Figure S2a{]}.
+MDA was \emph{Lachnospiraceae} (Otu000015) {[}Figure 3B{]}.
 
@@
 probability of the follow up sample to be a lesion. This is what we
-observed for the lesion model {[}Figure 4{]} (P-value \textless{}
-0.001). When we separated individuals based on whether they had an
-adenoma or carcinoma there was only a decrease in positive probability
-for the carcinoma group (P-value \textless{} 0.001) and not for the
-adenoma group (P-value \textgreater{} 0.05). We also observed that there
-were no significant differences in whether the models classified the
-samples as having lesion between the predicted and actual (P-value
-\textgreater{} 0.05). The lesion model was able to correctly classify
-the one individual who still had a carcinoma on follow up {[}Figure
-4{]}.
+observed for the lesion model (P-value \textless{} 0.001). When we
+separated individuals based on whether they had an adenoma or carcinoma
+there was only a decrease in positive probability for the carcinoma
+group {[}Figure 3C{]} (P-value \textless{} 0.001) and not for the
+adenoma group {[}Figure 3D{]} (P-value \textgreater{} 0.05). We also
+observed that there were no significant differences in whether the
+models classified the samples as having lesion between the predicted and
+actual (P-value \textgreater{} 0.05). The lesion model was able to
+correctly classify the one individual who still had a carcinoma on
+follow up {[}Figure 3C{]}.
 
@@
 follow up samples based on the bacterial community and FIT measurements.
-The initial sample model had an AUC range of 0.485 to 0.686 after 100
+The initial sample model had an AUC range of 0.641 to 0.805 after 100
 iterations of 20 repeated 10-fold cross validations. The test set AUC
@@
 was a marked decrease in the ROC curve for the final model used when
-compared to the 100 test set AUC iterations {[}Figure 3b{]}. There were
-a total of 70 variables {[}Figure S2{]}. For this model
-\emph{Ruminococcaceae} (Otu000278) resulted in the largest decrease in
-MDA while FIT measurement resulted in the second largest decrease in MDA
-{[}Figure S3a{]}.
+compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
+a total of 70 variables {[}Figure 4B{]}. For this model
+\emph{Pseudomonas} (Otu000438) resulted in the largest decrease in MDA
+while FIT measurement resulted in the sixth largest decrease in MDA
+{[}Figure 4B{]}.
 
@@
 follow up sample to be an initial sample. This is what we observed for
-the initial sample model {[}Figure 4{]} (P-value \textless{} 0.001).
-When we separated individuals based on whether they had an adenoma or
-carcinoma there was a decrease in positive probability for both the
-carcinoma group (P-value \textless{} 0.001) and for the adenoma group
-(P-value \textless{} 0.001). For this model there was no difference
-between the predicted and actual classifications (P-value \textgreater{}
-0.05) {[}Figure 4{]}.
+the initial sample model (P-value \textless{} 0.001). When we separated
+individuals based on whether they had an adenoma or carcinoma there was
+a decrease in positive probability for both the carcinoma group
+{[}Figure 4C{]} (P-value \textless{} 0.001) and for the adenoma group
+{[}Figure 4D{]} (P-value \textless{} 0.001). For this model there was no
+difference between the predicted and actual classifications (P-value
+\textgreater{} 0.05).
 

### 2017-02-20 code/Run_FigureS3.R
unification of repo

@@ -1,100 +1,0 @@
-### Figure S3
-### Description of variables in IF models
-## Marc Sze
-
-
-#Load needed code and packages
-source('code/functions.R')
-
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson"))
-
-#Read data needed
-IF_MDA_data_summary <- read.csv("results/tables/reduced_IF_model_top_vars_MDA_Summary.csv", 
-                                    header = T, stringsAsFactors = F)
-
-IF_MDA_full <- read.csv("results/tables/reduced_IF_model_top_vars_MDA.csv", 
-                            header = T, stringsAsFactors = F)
-
-tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
-
-occurance_data <- read.csv("results/tables/IF_rf_wCV_imp_vars_summary.csv", header = T, stringsAsFactors = F)
-
-# Convert taxa table to a data frame with columns for each taxonomic division
-tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
-rownames(tax_df) <- rownames(tax)
-colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
-tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
-rm(tax)
-
-# Order the data file from most to least important based on mean MDA
-IF_MDA_data_summary <- IF_MDA_data_summary[order(IF_MDA_data_summary$mean_MDA, decreasing = TRUE), ]
-
-# Pull OTUs that are only in the MDA data
-OTU_IDs <- unique(IF_MDA_data_summary$variable)
-select_tax_df <- tax_df[OTU_IDs, ]
-low_tax_ID <- gsub("2", "", gsub("_unclassified", "", createTaxaLabeller(select_tax_df)))
-
-# create labels for factor values with low taxonomy
-graph_labels <- c(paste(gsub("2", "", low_tax_ID), " (", names(low_tax_ID), ")", sep = ""))
-OTU_names <- names(low_tax_ID)
-
-test <- c()
-for(i in 1:length(low_tax_ID)){
-  if(is.na(low_tax_ID[i])){
-    
-    test <- c(test, bquote(paste("FIT", sep = "")))
-  } else {
-    
[…]

### 2017-02-20 code/Run_FigureS4.R

@@ -1,129 +1,0 @@
-### Figure S4
-### How do the full models of lesion and IF look
-## Marc Sze
-
-
-###Load needed Libraries and functions
-source('code/functions.R')
-
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
-           "gridExtra", "scales", "wesanderson", "caret"))
-
-# Load lesion model data
-follow_up_probability <- read.csv(
-  "results/tables/follow_up_probability_summary.csv", 
-  header = T, stringsAsFactors = F)
-
-# Load IF model data
-IF_follow_up_probability <- read.csv(
-  "results/tables/IF_follow_up_probability_summary.csv", 
-  header = T, stringsAsFactors = F)
-
-# Read in meta data tables
-good_metaf <- read.csv("results/tables/mod_metadata/good_metaf_final.csv", 
-                       stringsAsFactors = F, header = T) %>% 
-  filter(!is.na(fit_followUp))
-
-
-accuracy_plot <- grid.arrange(
-  # Graph the lesion carcinoma data only
-  filter(follow_up_probability, diagnosis != "adenoma") %>% 
-    ggplot(aes(factor(sampleType, 
-                      levels = c("initial", "followup")), Yes, group = factor(EDRN))) + 
-    geom_point(aes(color=factor(disease_free, 
-                                levels = c("n", "y", "unknown"))), size = 2) + 
-    geom_line(linetype = 2, alpha = 0.3) + 
-    scale_color_manual(name = "Cancer Free", 
-                       label = c("No", "Yes", "Unknown"),  
-                       values = wes_palette("GrandBudapest")) + 
-    scale_x_discrete(breaks = c("initial", "followup"), 
-                     labels = c("Initial", "Follow Up")) + 
-    coord_cartesian(ylim = c(0, 1)) + 
-    geom_hline(aes(yintercept = 0.5), linetype = 2) + 
-    ggtitle("A") + ylab("Lesion Postive Probability") + 
-    xlab("") + theme_bw() + theme(
-      axis.title = element_text(face="bold"), 
-      legend.title = element_text(face="bold"), 
-      legend.position = c(0.25, 0.15), 
-      plot.title = element_text(face="bold", hjust = 0)), 
-  
-  # Graph the lesion adenoma data only
-  filter(follow_up_probability, diagnosis == "adenoma") %>%
[…]

### 2017-02-20 code/Run_FigureS5.R

@@ -1,22 +1,0 @@
-### Figure S5
-### Distribution of P-values from Paired Wilcoxson Analysis
-## Marc Sze
-
-#Load needed code and packages
-source('code/functions.R')
-
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
-
-#Read data needed
-diff_table_time <- read.csv("results/tables/time_datatable.csv", header = T)
-
-# Plot the change in distance versus time
-time_graph <- ggplot(diff_table_time, aes(x = time, y = distance)) + 
-  geom_point(aes(color = dx), size = 4) + theme_bw() + 
-  scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
-                     breaks = c("adenoma", "cancer"), labels = c("Adenoma", "Cancer")) + 
-  coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("Time from Initial (Days)") + 
-  theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"))
-
-ggsave(file = "results/figures/FigureS5.pdf", time_graph, 
-       width=6, height = 8, dpi = 300)

### 2017-02-20 submission/manuscript_outline_20161024.Rmd

@@
 The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].
-There were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure 3B].
-For this model FIT measurement resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
+There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].
+The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
 
@@
 There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].
-There were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure 4B].
-For this model *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement resulted in the sixth largest decrease in MDA [Figure 4B].
+There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].
+The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].
 
@@
 We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).
-What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 4 & S4] but also that the commonly associated CRC bacteria were not highly represented within this model.
+What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 3 & 4] but also that the commonly associated CRC bacteria were not highly represented within this model.
 Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
@@
 We followed this up by creating a second model to classify initial versus follow up samples.
-We found that this model was able to accurately classify initial versus follow up samples [Figure 4 & S4].
+We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].
 Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].
@@
 It is also possible that both of these explanations could have played a role.
-Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].
+Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].
 It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.
@@
 
-Interestingly, within figure 3 the before sample model showed better test AUC results then the training set AUC.
+Interestingly, within figure 4 the initial sample model showed better test AUC results then the training set AUC.
 This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
@@
 This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.
+
 In brief, the model included data on FIT and the bacterial microbiome.
@@
 This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
+Next, in order to assess which variables were most important to the model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.
+This final collated list of variables was what was considered the most important for the model.
+This reduced data set was then run through the mtry optimization again.
 Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
+
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
@@
 ***Initial Sample Model Creation:***
-We also investigated whether a model could be created that could identify pre (initial) and post (follow up) surgery samples.
+We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal.
 The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
 Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.
-Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial surgery sample.
-
-***Selection of Important OTUs:***
-In order to assess which variables were most important to either model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.
-This final collated list of variables was what was considered the most important for the lesion or before sample models.
+Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial to lesion removal sample.
 
@@
 From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
-The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.
-We then used the initial sample model to assess whether it could classify samples better then the lesion model.
-The most important OTUs were then identified from this model and used to create a reduced feature initial sample model.
-This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.
-A list of common OTUs were found for the two different models used.
-Finally, both lesion and initial sample models initial positive probabilities were tested against the follow up positive probabilities.
+We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.
+Finally, a list of common OTUs were found for the two two different models used.
 
@@
 
-**Figure 3: Receiver Operating Characteristic Curve for Lesion and Initial Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.
-The blue line represents the reduced variable model using 100% of the data set and what was used for subsequent classification.
-A) Lesion model.
-B) Before sample model
+**Figure 3: The Lesion Model.** A) ROC Curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
+B) Summary of Important Variables.
+MDA of the most important variables in the lesion model.
+The black point represents the median and the different colors are the different runs up to 100.
+C) Positive probability change from initial to follow up sample in those with carcinomaA) Lesion model.
+D) Positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN)
 
-**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).
-D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.
-The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
+**Figure 4: The Initial Sample Model.**  A) ROC Curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
+B) Summary of Important Variables.
+MDA of the most important variables in the initial sample model.
+The black point represents the median and the different colors are the different runs up to 100.
+C) Positive probability change from initial to follow up sample in those with carcinoma.
+D) Positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).
 
@@
 
-**Figure S2: Summary of Important Variables in the Lesion Model** A) MDA of the most important variables in the lesion model.
-The black point represents the median and the different colors are the different runs up to 100.
-B) The total number of appearances of each variable in the 100 different lesion models.
-The cutoff of 50% was used to assess importance.
-
-**Figure S3: Summary of Important Variables in Before Sample Model** A) MDA of the most important variables in the lesion model.
-The black point represents the median and the different colors are the different runs up to 100.
-B) The total number of appearances of each variable in the 100 different lesion models.
-The cutoff of 50% was used to assess importance.
-
-**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevant Neoplasia (SRN).
-D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.
-The dotted line represents the threshold used to make the decision of whether a sample was positive or not.
-
-**Figure S5: Thetayc Versus Time of Follow up Sample from Initial**
+**Figure S2: Thetayc Versus Time of Follow up Sample from Initial**
 

### 2017-02-20 submission/manuscript_outline_20161024.tex

@@
 of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
-a total of 53 variables {[}Figure 3B{]}. For this model FIT measurement
-resulted in the largest decrease in MDA while the OTU with the largest
-MDA was \emph{Lachnospiraceae} (Otu000015) {[}Figure 3B{]}.
+a total of 53 variables that were used in this model {[}Figure 3B{]}.
+The FIT measurement for fecal blood resulted in the largest decrease in
+MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
+(Otu000015) {[}Figure 3B{]}.
 
@@
 compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
-a total of 70 variables {[}Figure 4B{]}. For this model
-\emph{Pseudomonas} (Otu000438) resulted in the largest decrease in MDA
-while FIT measurement resulted in the sixth largest decrease in MDA
-{[}Figure 4B{]}.
+a total of 70 variables that were used for this model {[}Figure 4B{]}.
+The variable that resulted in the largest MDA was \emph{Pseudomonas}
+(Otu000438) resulted in the largest decrease in MDA while FIT
+measurement for fecal blood resulted in the sixth largest decrease in
+MDA {[}Figure 4B{]}.
 
@@
 we found was that not only was the positive probability of a lesion
-reduced after lesion removal {[}Figure 4 \& S4{]} but also that the
+reduced after lesion removal {[}Figure 3 \& 4{]} but also that the
 commonly associated CRC bacteria were not highly represented within this
@@
 classify initial versus follow up samples. We found that this model was
-able to accurately classify initial versus follow up samples {[}Figure 4
-\& S4{]}. Providing additional information on the importance of
-commensal bacteria was that there were a total of 23 OTUs in common to
-both models and the vast majority belonged to regular residents of our
-gut community {[}Table S3{]}.
+able to accurately classify initial versus follow up samples {[}Figure 3
+\& 4{]}. Providing additional information on the importance of commensal
+bacteria was that there were a total of 23 OTUs in common to both models
+and the vast majority belonged to regular residents of our gut community
+{[}Table S3{]}.
 
@@
 initiation and could change in a way that allows predictive models to
-lower the positive probability of a lesion after surgery {[}Figure 4{]}.
-It should be noted that our study does not argue against the importance
-of these CRC associated bacteria in the pathogenesis of disease but
-rather that they are not the main bacteria changing after surgical
-removal of lesion. In fact, it is possible that these CRC associated
-bacteria are important in the transition from adenoma to carcinoma and
-would be one explanation as to why in our data we not only see high
-initial relative abundances in carcinoma and not adenoma individuals but
-also large decreases in relative abundance in some of those with
-carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.
+lower the positive probability of a lesion after surgery {[}Figure 3C \&
+3D{]}. It should be noted that our study does not argue against the
+importance of these CRC associated bacteria in the pathogenesis of
+disease but rather that they are not the main bacteria changing after
+surgical removal of lesion. In fact, it is possible that these CRC
+associated bacteria are important in the transition from adenoma to
+carcinoma and would be one explanation as to why in our data we not only
+see high initial relative abundances in carcinoma and not adenoma
+individuals but also large decreases in relative abundance in some of
+those with carcinoma but not in those with adenoma after surgery
+{[}Figure 2{]}.
 
@@
 
-Interestingly, within figure 3 the before sample model showed better
+Interestingly, within figure 4 the initial sample model showed better
 test AUC results then the training set AUC. This may have occurred
@@
 completed on a data set of 423 individuals. This model was then applied
-to our follow up data set of 67 individuals. In brief, the model
-included data on FIT and the bacterial microbiome. Non-binary data was
-checked for near zero variance and OTUs that had near zero variance were
-removed. This pre-processing was performed with the R package caret
-(v6.0.73). Optimization of the mtry hyper-parameter involved taking the
-samples and making 100 different 80/20 (train/test) splits of the data
-where normal and lesion were represented in the same proportion within
-both the whole data set and the 80/20 split. Each of these splits were
-then run through 20 repeated 10-fold cross validations to optimize the
-mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
-Receiver Operator Characteristic). This resulting model was then tested
-on the 20\% of the data that was originally held out from this overall
-process. Once the ideal mtry was found the entire 423 sample set was
-used to create the final Random Forest model on which classifications on
-the 67-person cohort was completed. The default cutoff of 0.5 was used
-as the threshold to classify individuals as positive or negative for
-lesion. The hyper-parameter, mtry, defines the number of variables to
-investigate at each split before a new division of the data is created
-with the Random Forest model.
+to our follow up data set of 67 individuals.
+
+In brief, the model included data on FIT and the bacterial microbiome.
+Non-binary data was checked for near zero variance and OTUs that had
+near zero variance were removed. This pre-processing was performed with
+the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
+involved taking the samples and making 100 different 80/20 (train/test)
+splits of the data where normal and lesion were represented in the same
+proportion within both the whole data set and the 80/20 split. Each of
+these splits were then run through 20 repeated 10-fold cross validations
+to optimize the mtry hyper-parameter by maximizing the AUC (Area Under
+the Curve of the Receiver Operator Characteristic). This resulting model
+was then tested on the 20\% of the data that was originally held out
+from this overall process. Next, in order to assess which variables were
+most important to the model we counted the number of times a variable
+was present in the top 10\% of mean decrease in accuracy (MDA) for each
+of the 100 different 80/20 split models and then filtered this list to
+variables that were only present more than 50\% of the time. This final
+collated list of variables was what was considered the most important
+for the model. This reduced data set was then run through the mtry
+optimization again. Once the ideal mtry was found the entire 423 sample
+set was used to create the final Random Forest model on which
+classifications on the 67-person cohort was completed.
+
+The default cutoff of 0.5 was used as the threshold to classify
+individuals as positive or negative for lesion. The hyper-parameter,
+mtry, defines the number of variables to investigate at each split
+before a new division of the data is created with the Random Forest
+model.
 
@@
 whether a model could be created that could identify pre (initial) and
-post (follow up) surgery samples. The main difference was that only the
+post (follow up) lesion removal. The main difference was that only the
 67-person cohort was used at all stages of model building and
@@
 samples as positive or negative of lesion this model classified samples
-as positive or negative for being an initial surgery sample.
-
-\textbf{\emph{Selection of Important OTUs:}} In order to assess which
-variables were most important to either model we counted the number of
-times a variable was present in the top 10\% of mean decrease in
-accuracy (MDA) for each of the 100 different 80/20 split models and then
-filtered this list to variables that were only present more than 50\% of
-the time. This final collated list of variables was what was considered
-the most important for the lesion or before sample models.
+as positive or negative for being an initial to lesion removal sample.
 
@@
 the lesion model was then tested for accuracy in prediction and whether
-it reduced the positive probability of lesion after surgery. The most
-important OTUs for this were used to build a reduced model and it was
-assessed for similarity to the original model. We then used the initial
-sample model to assess whether it could classify samples better then the
-lesion model. The most important OTUs were then identified from this
-model and used to create a reduced feature initial sample model. This
-reduced feature model, as was done with the lesion model, was compared
-to the full model for loss of accuracy. A list of common OTUs were found
-for the two different models used. Finally, both lesion and initial
-sample models initial positive probabilities were tested against the
-follow up positive probabilities.
+it reduced the positive probability of lesion after surgery. We then
+used the initial sample model to assess whether it could classify
+samples better then the lesion model and whether it could reduce the
+positive probability of an initial sample in the follow up samples.
+Finally, a list of common OTUs were found for the two two different
+models used.
 
@@
 
-\textbf{Figure 3: Receiver Operating Characteristic Curve for Lesion and
-Initial Sample Models.} The shaded areas represents the range of values
-of a 100 different 80/20 splits of the test set data using either all
-variables (grey) or reduced variable (red) models. The blue line
-represents the reduced variable model using 100\% of the data set and
-what was used for subsequent classification. A) Lesion model. B) Before
-sample model
+\textbf{Figure 3: The Lesion Model.} A) ROC Curve: The shaded areas
+represents the range of values of a 100 different 80/20 splits of the
+test set data and the blue line represents the model using 100\% of the
+data set and what was used for subsequent classification. B) Summary of
+Important Variables. MDA of the most important variables in the lesion
+model. The black point represents the median and the different colors
+are the different runs up to 100. C) Positive probability change from
+initial to follow up sample in those with carcinomaA) Lesion model. D)
+Positive probability change from initial to follow up sample if those
+with adenoma or Screen Relevant Neoplasia (SRN)
 
-\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Initial Sample Reduced Variable Models} A) Lesion
-positive probability change from initial to follow up sample in those
-with carcinoma B) Initial positive probability change from initial to
-follow up sample in those with carcinoma C) Lesion positive probability
-change from initial to follow up sample if those with adenoma or Screen
-Relevant Neoplasia (SRN). D) Initial positive probability change from
-initial to follow up sample in those with adenoma or SRN. The dotted
-line represents the threshold used to make the decision of whether a
-sample was positive or not.
+\textbf{Figure 4: The Initial Sample Model.} A) ROC Curve: The shaded
+areas represents the range of values of a 100 different 80/20 splits of
+the test set data and the blue line represents the model using 100\% of
+the data set and what was used for subsequent classification. B) Summary
+of Important Variables. MDA of the most important variables in the
+initial sample model. The black point represents the median and the
+different colors are the different runs up to 100. C) Positive
+probability change from initial to follow up sample in those with
+carcinoma. D) Positive probability change from initial to follow up
+sample if those with adenoma or Screen Relevant Neoplasia (SRN).
 
@@
 
-\textbf{Figure S2: Summary of Important Variables in the Lesion Model}
-A) MDA of the most important variables in the lesion model. The black
-point represents the median and the different colors are the different
-runs up to 100. B) The total number of appearances of each variable in
-the 100 different lesion models. The cutoff of 50\% was used to assess
-importance.
-
-\textbf{Figure S3: Summary of Important Variables in Before Sample
-Model} A) MDA of the most important variables in the lesion model. The
-black point represents the median and the different colors are the
-different runs up to 100. B) The total number of appearances of each
-variable in the 100 different lesion models. The cutoff of 50\% was used
-to assess importance.
-
-\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
-Results for Lesion and Before Sample Full Variable Models} A) Lesion
-positive probability change from initial to follow up sample in those
-with carcinoma B) Initial positive probability change from initial to
-follow up sample in those with carcinoma C) Lesion positive probability
-change from initial to follow up sample in those with adenoma or Screen
-Relevant Neoplasia (SRN). D) Initial positive probability change from
-initial to follow up sample in those with adenoma or SRN. The dotted
-line represents the threshold used to make the decision of whether a
-sample was positive or not.
-
-\textbf{Figure S5: Thetayc Versus Time of Follow up Sample from Initial}
+\textbf{Figure S2: Thetayc Versus Time of Follow up Sample from Initial}
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
abstract corrections

@@
 
-**Results:** The bacterial microbiome in pre and post surgery samples for the adenoma group were more similar to each other than the carcinoma group (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+**Results:** The bacterial microbiome in pre and post lesion removal samples for the adenoma group were more similar to each other than the carcinoma group (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
 There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
 A model built to classify model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
-The post removal sample for both models had a decrease in the positive probability for either lesion or initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+Follow up samples had a decrease in the positive probability (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+This decrease in positive probability was also observed for the initial sample model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.

### 2017-02-20 submission/manuscript_outline_20161024.tex
abstract corrections

@@
 
-\textbf{Results:} The bacterial microbiome in pre and post surgery
-samples for the adenoma group were more similar to each other than the
-carcinoma group (P-value \textless{} 0.05). There was no difference for
-either the adenoma or carcinoma group in the relative abundance of any
-OTU between the pre and post lesion removal (P-value \textgreater{}
-0.05). A model built to classify model built to classify lesion had an
-AUC range of 0.811 - 0.866 while a model built to classify initial
-versus follow up samples had an AUC range of 0.641 - 0.805. The post
-removal sample for both models had a decrease in the positive
-probability for either lesion or initial sample (P-value \textless{}
-0.05). The lesion model used a total of 53 variables while the initial
-sample model used a total of 70 variables. A total of 23 OTUs were
-common to both models with the majority of these classifying to
-commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
-\emph{Blautia}, and \emph{Ruminococcaceae}).
+\textbf{Results:} The bacterial microbiome in pre and post lesion
+removal samples for the adenoma group were more similar to each other
+than the carcinoma group (P-value \textless{} 0.05). There was no
+difference for either the adenoma or carcinoma group in the relative
+abundance of any OTU between the pre and post lesion removal (P-value
+\textgreater{} 0.05). A model built to classify model built to classify
+lesion had an AUC range of 0.811 - 0.866 while a model built to classify
+initial versus follow up samples had an AUC range of 0.641 - 0.805.
+Follow up samples had a decrease in the positive probability (P-value
+\textless{} 0.05). This decrease in positive probability was also
+observed for the initial sample model (P-value \textless{} 0.05). The
+lesion model used a total of 53 variables while the initial sample model
+used a total of 70 variables. A total of 23 OTUs were common to both
+models with the majority of these classifying to commensal bacteria
+(e.g. \emph{Bacteroides}, \emph{Clostridiales}, \emph{Blautia}, and
+\emph{Ruminococcaceae}).
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
title changes

@@
 
-# Differences in the fecal Microbiome Before and After Colorectal Cancer Treatment
+# The Fecal Microbiome Before and After Treatment for Colorectal Adenoma or Carcinoma
 ```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
@@
 \begin{center}
-Running Title: Human Microbiome and Colorectal Cancer
+Running Title: Human Microbiome before and after Colorectal Cancer
 

### 2017-02-20 submission/manuscript_outline_20161024.tex
title changes

@@
 
-\section{Differences in the fecal Microbiome Before and After Colorectal
-Cancer
-Treatment}\label{differences-in-the-fecal-microbiome-before-and-after-colorectal-cancer-treatment}
+\section{The Fecal Microbiome Before and After Treatment for Colorectal
+Adenoma or
+Carcinoma}\label{the-fecal-microbiome-before-and-after-treatment-for-colorectal-adenoma-or-carcinoma}
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
background corrections

@@
 There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
-A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
+A few studies have also shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
 These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
@@
 
-In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after removal of a lesion (adenoma or carcinoma).
+In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples removal of lesion (adenoma or carcinoma).
 Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest to build models to classify either initial and follow up samples or lesion and normal samples and then subsequently observe how these models, as well as specific OTUs within them, performed pre and post surgery.
-We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.
+We also utilized Random Forest to build models to classify either lesion and normal samples or initial and follow up samples and then subsequently observe how these models, as well as specific OTUs within them, performed pre and post lesion removal.
+We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normalsamples but also initial and follow up.
 

### 2017-02-20 submission/manuscript_outline_20161024.tex
background corrections

@@
 altered directly on tumor tissue itself {[}5{]}. A few studies have also
-shown how this microbiome {[}6{]} or specific members within it {[}7{]}
-could be directly involved with the pathogenesis of CRC. These studies
-have helped to provide a tantalizing link between the bacterial
-microbiome and CRC. Although these studies suggest that the bacterial
-microbiome might change after treatment there remains limited
+shown how this bacterial microbiome {[}6{]} or specific members within
+it {[}7{]} could be directly involved with the pathogenesis of CRC.
+These studies have helped to provide a tantalizing link between the
+bacterial microbiome and CRC. Although these studies suggest that the
+bacterial microbiome might change after treatment there remains limited
 information on the bacterial microbiome before and after surgical
@@
 In this study we tested the hypothesis that the bacterial microbiome
-changes between pre (initial) and post (follow up) samples after removal
-of a lesion (adenoma or carcinoma). Our analysis included both alpha and
-beta diversity analysis along with investigation of individual
-operational taxonomic units (OTUs). We also utilized Random Forest to
-build models to classify either initial and follow up samples or lesion
-and normal samples and then subsequently observe how these models, as
-well as specific OTUs within them, performed pre and post surgery. We
+changes between pre (initial) and post (follow up) samples removal of
+lesion (adenoma or carcinoma). Our analysis included both alpha and beta
+diversity analysis along with investigation of individual operational
+taxonomic units (OTUs). We also utilized Random Forest to build models
+to classify either lesion and normal samples or initial and follow up
+samples and then subsequently observe how these models, as well as
+specific OTUs within them, performed pre and post lesion removal. We
 also investigated the models for similar important OTUs to identify the
-crucial OTUs for not only classifying initial and follow up samples but
-also lesion or normal.
+crucial OTUs for not only classifying lesion or normalsamples but also
+initial and follow up.
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
results text corrections

@@
 When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group [Figure 3D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
-We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-The lesion model was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 3C].
+We also observed that there were no significant differences between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+The lesion model was also able to correctly classify the one individual who still had a carcinoma on follow up [Figure 3C].
 
 ***The Initial Sample Model:***
-After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.
+After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.
 The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
@@
 There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].
-The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].
+The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].
 

### 2017-02-20 submission/manuscript_outline_20161024.tex
results text corrections

@@
 adenoma group {[}Figure 3D{]} (P-value \textgreater{} 0.05). We also
-observed that there were no significant differences in whether the
-models classified the samples as having lesion between the predicted and
-actual (P-value \textgreater{} 0.05). The lesion model was able to
-correctly classify the one individual who still had a carcinoma on
-follow up {[}Figure 3C{]}.
+observed that there were no significant differences between the
+predicted and actual calls (P-value \textgreater{} 0.05). The lesion
+model was also able to correctly classify the one individual who still
+had a carcinoma on follow up {[}Figure 3C{]}.
 
@@
 classify based on lesion we wanted to then build a separate model
-specifically to be able to classify whether samples were initial or
-follow up samples based on the bacterial community and FIT measurements.
-The initial sample model had an AUC range of 0.641 to 0.805 after 100
-iterations of 20 repeated 10-fold cross validations. The test set AUC
-range for this model performed better then the training set AUCs. There
-was a marked decrease in the ROC curve for the final model used when
-compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
-a total of 70 variables that were used for this model {[}Figure 4B{]}.
-The variable that resulted in the largest MDA was \emph{Pseudomonas}
-(Otu000438) resulted in the largest decrease in MDA while FIT
-measurement for fecal blood resulted in the sixth largest decrease in
-MDA {[}Figure 4B{]}.
+specifically to be able to classify whether samples were initial samples
+based on the bacterial community and FIT measurements. The initial
+sample model had an AUC range of 0.641 to 0.805 after 100 iterations of
+20 repeated 10-fold cross validations. The test set AUC range for this
+model performed better then the training set AUCs. There was a marked
+decrease in the ROC curve for the final model used when compared to the
+100 test set AUC iterations {[}Figure 4A{]}. There were a total of 70
+variables that were used for this model {[}Figure 4B{]}. The variable
+that resulted in the largest MDA was \emph{Pseudomonas} (Otu000438)
+while FIT measurement for fecal blood resulted in the sixth largest
+decrease in MDA {[}Figure 4B{]}.
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
discussion text corrections

@@
 
-From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.
+From our results there were large observed differences in the bacterial microbiome between pre and post lesion removal samples based on whether the individual had an adenoma or carcinoma.
 There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
 However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].
-There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].
-Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.
+There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].
 
@@
 We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].
-Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].
+Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].
+Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.
 
@@
 These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
-For time differences, specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].
+Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].
 
@@
 
-Curiously, we observed that the typical CRC associated bacteria were not predictive ewithin our models.
+Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
 There are a number of reasons why this may have occurred.
@@
 
-Many of the common OTUs between the different models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].
+Many of the common OTUs between the two models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].
 Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.
@@
 
-Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.
+Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma or carcinoma removal.
 Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.

### 2017-02-20 submission/manuscript_outline_20161024.tex
discussion text corrections

@@
 From our results there were large observed differences in the bacterial
-microbiome between pre and post surgery samples based on whether the
-individual had an adenoma or carcinoma. There were much larger
+microbiome between pre and post lesion removal samples based on whether
+the individual had an adenoma or carcinoma. There were much larger
 differences between initial and follow up samples based on the thetayc
@@
 relative abundance of any specific OTU for lesion, adenoma only, or
-carcinoma only {[}Figure s1{]}. Although there was a detectable change
-in the carcinoma group and this change was towards what would be
-expected for normal controls, the follow up samples may not be a
-completely normal bacterial microbiome.
+carcinoma only {[}Figure S1{]}.
 
@@
 and the vast majority belonged to regular residents of our gut community
-{[}Table S3{]}.
+{[}Table S2{]}. Although there was a detectable change in the carcinoma
+group and this change was towards what would be expected for normal
+controls, the follow up samples may not be a completely normal bacterial
+microbiome.
 
@@
 could be due to differences in collection time between samples for the
-two different groups. For time differences, specifically, it could
-confound the observation that carcinomas changed more than adenomas
-{[}Figure 1a \& 1d{]}.
+two different groups. Specifically, it could confound the observation
+that carcinomas changed more than adenomas {[}Figure 1a \& 1d{]}.
 
@@
 Curiously, we observed that the typical CRC associated bacteria were not
-predictive ewithin our models. There are a number of reasons why this
-may have occurred. First, is that they were not present in enough
+predictive within our models. There are a number of reasons why this may
+have occurred. First, is that they were not present in enough
 individuals to be able to classify those with and without disease with a
@@
 
-Many of the common OTUs between the different models used had many OTUs
-that taxonomically classified to potential butyrate producers {[}Table
-S2{]}. Another batch of OTUs classified to bacteria that can either
-degrade polyphenols or are inhibited by them. Both butyrate and
-polyphenols are thought to be protective against cancer in part by
-reducing inflammation {[}13{]}. These protective compounds are derived
-from the breakdown of fiber, fruits, and vegetables by resident gut
-microbes. One example of this potential
-diet-microbiome-inflammation-polyp axis is that \emph{Bacteroides},
-which was highly prevalent in our models, are known to be increased in
-those with high non-meat based protein consumption {[}14{]}. High
-protein consumption in general has been linked with an increased CRC
-risk {[}15{]}. Conversely, \emph{Bacteroides} are inhibited by
-polyphenols which are derived from fruits and vegetables {[}16{]}. Our
-data fits with the hypothesis that the microbial metabolites from
-breakdown products within our own diet could not only help to shape the
-existing community but also have an effect on CRC risk and disease
-progression.
+Many of the common OTUs between the two models used had many OTUs that
+taxonomically classified to potential butyrate producers {[}Table S2{]}.
+Another batch of OTUs classified to bacteria that can either degrade
+polyphenols or are inhibited by them. Both butyrate and polyphenols are
+thought to be protective against cancer in part by reducing inflammation
+{[}13{]}. These protective compounds are derived from the breakdown of
+fiber, fruits, and vegetables by resident gut microbes. One example of
+this potential diet-microbiome-inflammation-polyp axis is that
+\emph{Bacteroides}, which was highly prevalent in our models, are known
+to be increased in those with high non-meat based protein consumption
+{[}14{]}. High protein consumption in general has been linked with an
+increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
+inhibited by polyphenols which are derived from fruits and vegetables
+{[}16{]}. Our data fits with the hypothesis that the microbial
+metabolites from breakdown products within our own diet could not only
+help to shape the existing community but also have an effect on CRC risk
+and disease progression.
 
@@
 knowledge on CRC and the bacterial microbiome: That there is a
-measurable difference in the bacterial community after surgical removal
-of lesion. Further, the ability for machine learning algorithms to take
-bacterial microbiome data and successfully lower positive probability
-after either adenoma or carcinoma removal provides evidence that there
-are specific signatures, mostly attributable to commensal organisms,
-associated with these lesions. Our data provides evidence that commensal
-bacteria may be important in the development of polyps and also
-potentially the transition from adenoma to carcinoma.
+measurable difference in the bacterial community after adenoma or
+carcinoma removal. Further, the ability for machine learning algorithms
+to take bacterial microbiome data and successfully lower positive
+probability after either adenoma or carcinoma removal provides evidence
+that there are specific signatures, mostly attributable to commensal
+organisms, associated with these lesions. Our data provides evidence
+that commensal bacteria may be important in the development of polyps
+and also potentially the transition from adenoma to carcinoma.
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
methods text corrections

@@
 The sampling and design were similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].
-In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.
+In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
 The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].
-Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed.
+Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after lesion removal and treatment had been completed.
 This study was approved by the University of Michigan Institutional Review Board.
@@
 ***Lesion Model Creation:***
-The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
+The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
 This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.
@@
 The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
-The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.
+The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.
 
 ***Initial Sample Model Creation:***
-We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal.
+We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal samples from each other.
 The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
@@
 We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
-This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.
+This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.
 From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.

### 2017-02-20 submission/manuscript_outline_20161024.tex
methods text corrections

@@
 fecal sample could be obtained, had IBD, a known hereditary
-non-polyposis colorectal cancer, or Familial adenomatous polyposis.
+non-polyposis colorectal cancer, or familial adenomatous polyposis.
 Samples used to build the models for prediction were collected either
@@
 sample as described but also a follow up sample between 188 - 546 days
-after surgery and treatment had been completed. This study was approved
-by the University of Michigan Institutional Review Board. All study
-participants provided informed consent and the study itself conformed to
-the guidelines set out by the Helsinki Declaration.
+after lesion removal and treatment had been completed. This study was
+approved by the University of Michigan Institutional Review Board. All
+study participants provided informed consent and the study itself
+conformed to the guidelines set out by the Helsinki Declaration.
 
@@
 algorithm was used to create the model used for prediction of lesion
-(adenoma or carcinoma) with the main testing and training of the model
+(adenoma or carcinoma) with the main training and testing of the model
 completed on a data set of 423 individuals. This model was then applied
@@
 mtry, defines the number of variables to investigate at each split
-before a new division of the data is created with the Random Forest
+before a new division of the data was created with the Random Forest
 model.
@@
 whether a model could be created that could identify pre (initial) and
-post (follow up) lesion removal. The main difference was that only the
-67-person cohort was used at all stages of model building and
-classification. Other than this difference the creation of this model
-and optimization of the mtry hyper-parameter was completed using the
-same procedure as was used for the lesion model. Instead of classifying
-samples as positive or negative of lesion this model classified samples
-as positive or negative for being an initial to lesion removal sample.
+post (follow up) lesion removal samples from each other. The main
+difference was that only the 67-person cohort was used at all stages of
+model building and classification. Other than this difference the
+creation of this model and optimization of the mtry hyper-parameter was
+completed using the same procedure as was used for the lesion model.
+Instead of classifying samples as positive or negative of lesion this
+model classified samples as positive or negative for being an initial to
+lesion removal sample.
 
@@
 differences in alpha and beta diversity, testing differences in FIT
-between initial and follow ups, testing all OTUs that were used in
-either built model, and investigating the relative abundance of specific
-previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
-\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
-\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
-the lesion model was then tested for accuracy in prediction and whether
-it reduced the positive probability of lesion after surgery. We then
-used the initial sample model to assess whether it could classify
-samples better then the lesion model and whether it could reduce the
-positive probability of an initial sample in the follow up samples.
-Finally, a list of common OTUs were found for the two two different
-models used.
+between initial and follow ups, testing all OTUs, and investigating the
+relative abundance of specific previously associated CRC bacteria
+(\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
+\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
+stomatis}) based on adenoma and carcinoma. From here the lesion model
+was then tested for accuracy in prediction and whether it reduced the
+positive probability of lesion after surgery. We then used the initial
+sample model to assess whether it could classify samples better then the
+lesion model and whether it could reduce the positive probability of an
+initial sample in the follow up samples. Finally, a list of common OTUs
+were found for the two two different models used.
 

### 2017-02-20 submission/manuscript_outline_20161024.Rmd
figure legend text corrections

@@
 
-**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+**Figure 1: General Differences between the Adenoma and Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
 B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
@@
 
-**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
+**Figure 2: Previously Associated CRC Bacteria in Initial and Follow Up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
 B) Adenoma initial and follow up samples.
-There were no significant differences between initial and follow up.
+There were no significant differences between initial and follow up (P-value = `r format(main_crc_bugs_summary["porp", "adn_BH"], digits = 3)`.
 
-**Figure 3: The Lesion Model.** A) ROC Curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure 3: The Lesion Model.** A) ROC curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Summary of Important Variables.
 MDA of the most important variables in the lesion model.
-The black point represents the median and the different colors are the different runs up to 100.
-C) Positive probability change from initial to follow up sample in those with carcinomaA) Lesion model.
-D) Positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN)
+The black point represents the mean and the different colors are the values of each different run up to 100.
+C) Positive probability change from initial to follow up sample in those with carcinoma.
+D) Positive probability change from initial to follow up sample if those with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
 
-**Figure 4: The Initial Sample Model.**  A) ROC Curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure 4: The Initial Sample Model.**  A) ROC curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Summary of Important Variables.
 MDA of the most important variables in the initial sample model.
-The black point represents the median and the different colors are the different runs up to 100.
+The black point represents the mean and the different colors are the values of each different run up to 100.
 C) Positive probability change from initial to follow up sample in those with carcinoma.
-D) Positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).
+D) Positive probability change from initial to follow up sample of those with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
 
@@
 ### Availability of data and material
+
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
+Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
 

### 2017-02-20 submission/manuscript_outline_20161024.tex

@@
 
-\textbf{Figure 1: General Differences between the Adenoma or Carcinoma
+\textbf{Figure 1: General Differences between the Adenoma and Carcinoma
 Group.} A) A significant difference was found between the adenoma and
@@
 \textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
-Follow up Samples.} A) Carcinoma initial and follow up samples had an
+Follow Up Samples.} A) Carcinoma initial and follow up samples had an
 observed significant difference in initial and follow up sample for the
@@
 initial and follow up samples. There were no significant differences
-between initial and follow up.
+between initial and follow up (P-value = 0.37.
 
-\textbf{Figure 3: The Lesion Model.} A) ROC Curve: The shaded areas
+\textbf{Figure 3: The Lesion Model.} A) ROC curve: The shaded areas
 represents the range of values of a 100 different 80/20 splits of the
@@
 Important Variables. MDA of the most important variables in the lesion
-model. The black point represents the median and the different colors
-are the different runs up to 100. C) Positive probability change from
-initial to follow up sample in those with carcinomaA) Lesion model. D)
+model. The black point represents the mean and the different colors are
+the values of each different run up to 100. C) Positive probability
+change from initial to follow up sample in those with carcinoma. D)
 Positive probability change from initial to follow up sample if those
-with adenoma or Screen Relevant Neoplasia (SRN)
+with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
 
-\textbf{Figure 4: The Initial Sample Model.} A) ROC Curve: The shaded
+\textbf{Figure 4: The Initial Sample Model.} A) ROC curve: The shaded
 areas represents the range of values of a 100 different 80/20 splits of
@@
 of Important Variables. MDA of the most important variables in the
-initial sample model. The black point represents the median and the
-different colors are the different runs up to 100. C) Positive
-probability change from initial to follow up sample in those with
-carcinoma. D) Positive probability change from initial to follow up
-sample if those with adenoma or Screen Relevant Neoplasia (SRN).
+initial sample model. The black point represents the mean and the
+different colors are the values of each different run up to 100. C)
+Positive probability change from initial to follow up sample in those
+with carcinoma. D) Positive probability change from initial to follow up
+sample of those with adenoma or advanced adenoma (Screen Relevant
+Neoplasia (SRN)).
 
@@
 material}\label{availability-of-data-and-material}
+
+A detailed and reproducible description of how the data were processed
+and analyzed can be found at
+\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
+have been deposited into the NCBI Sequence Read Archive (SRP062005 and
+SRP096978) and the necessary metadata can be found at
+\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
+respective SRA study accession.
 

### 2017-02-21 submission/manuscript_outline_20161024.Rmd

@@
 
-**Results:** The bacterial microbiome in pre and post lesion removal samples for the adenoma group were more similar to each other than the carcinoma group (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
 There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A model built to classify model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
-Follow up samples had a decrease in the positive probability (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
-This decrease in positive probability was also observed for the initial sample model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
+Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.
+The initial sample model had a decrease in positive probability for the follow up samples tobe an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
@@
 
-**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post lesion removal samples.
-With individuals with carcinoma having more drastic differences to the overall community then those with adenoma.
+**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal cancer lesion removal.
+Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.
 Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.
@@
 Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
-Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
-However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].
+Over the last few years rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
+However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
 
-Recently, there has been promising work on the bacterial microbiome and it's ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
+Over the last few years there has been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
 A few studies have also shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
 These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
-Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgical removal of lesion (adenoma or carcinoma) and whether the community changes at all.
+Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
+Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.
 
-In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples removal of lesion (adenoma or carcinoma).
-Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).
-We also utilized Random Forest to build models to classify either lesion and normal samples or initial and follow up samples and then subsequently observe how these models, as well as specific OTUs within them, performed pre and post lesion removal.
-We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normalsamples but also initial and follow up.
+In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).
+We analyzed changes in alpha and beta diversity as well as specific Operational Taxonomic Units (OTUs) relative abundance.
+We utilized Random Forest to build two models.
+The first model was built to classify either lesion and normal samples while the second was built to classify initial and follow up samples.
+Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.
+We then investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.
 

### 2017-02-22 submission/manuscript_outline_20161024.Rmd
incorporate Amanda suggestions results and discussion

@@
 ***Bacterial Community and FIT:***
-We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.
-What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a].
+Within our `r length(rownames(good_metaf))` person cohort we first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`).
+We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a].
 We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1b].
-The broad shift in bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
-Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
-When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].
-We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].
+The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
+Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
+When investigating alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].
+We also observed that there was no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].
 
@@
 Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].
-So we next examined whether there were differences in previously well described carcinoma associated OTUs.
-These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A].
-In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2B].
-We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].
+So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
+We first observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance for any of these respective OTUs [Figure 2].
+Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A].
+In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].
 
 ***The Lesion Model:***
-We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal control individuals.
-We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.
+We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal controls.
+We investigated this by creating a model to classify samples as lesion versus normal based on the bacterial community and FIT measurements.
 This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
@@
 ***The Initial Sample Model:***
-After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.
+After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.
 The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
@@
 There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
+This suggests that those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.
 All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].
@@
 ## Discussion
-
-From our results there were large observed differences in the bacterial microbiome between pre and post lesion removal samples based on whether the individual had an adenoma or carcinoma.
-There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].
-However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].
-There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].
-
-Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested carcinoma associated microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
-This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.
-
-We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).
-What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 3 & 4] but also that the commonly associated CRC bacteria were not highly represented within this model.
-Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.
-With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).
-We followed this up by creating a second model to classify initial versus follow up samples.
-We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].
-Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].
-Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.
-
-For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].
-Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
-These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
-Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].
 
@@
 These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].
+
+From our results there were large observed differences in the bacterial microbiome in samples before and after lesion removal based on whether the individual had an adenoma or carcinoma.
+In individuals with carcinoma compared to adenoma there were much larger differences between initial and follow up samples based on the community beta diversity and in fecal blood as measured by FIT [Figure 1].
+However, there were no differences between initial and follow up samples for any alpha diversity metric measured regardless of whether the individual had an adenoma or carcinoma [Table S1].
+There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].
+
+Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested carcinoma associated microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
+This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.
+
+We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).
+Based on this model we found that the follow up samples were closer to normal then the initial samples due to a decrease in positive probability for lesion and that the commonly associated CRC bacteria were not highly represented within this model with the exception of *Porphyromonas asaccharolytica* [Figure 3].
+Although there was a detectable change and this change was towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.
+
+After creating the lesion mode we then created a second model to classify initial versus follow up samples.
+We found that this model was able to accurately classify initial versus follow up samples suggesting that regardless of adenoma or carcinoma there are distinct common changes within the bacterial microbiome that occurs after lesion removal [Figure 4].
+Both models had OTUs that overwhelmingly belonged to commensal bacteria.
+Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].
+
+Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
+These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
+Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].
+This confounding though would not affect the observations where these individuals were grouped together.
 
@@
 We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
-The time difference in collection of sample between adenoma and carcinoma could have affected our observations for differences between individuals with adenoma or carcinoma.
-This confounding though would not affect the observations where these individuals are grouped together.
 
-Interestingly, within figure 4 the initial sample model showed better test AUC results then the training set AUC.
+Interestingly, within the initial sample model the test data performed better than the training data.
 This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
 Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.
-However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model alone.
-Further independent studies need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.
+However, the lesion model did not suffer from these discrepancies.
+Further independent studies need to be carried out to verify our findings and models since we are dealing with correlations that may not be truly representative of the pathogenesis of disease.
 

### 2017-02-22 submission/manuscript_outline_20161024.tex
incorporate Amanda suggestions results and discussion

@@
 
-\textbf{Results:} The bacterial microbiome in pre and post lesion
-removal samples for the adenoma group were more similar to each other
-than the carcinoma group (P-value \textless{} 0.05). There was no
-difference for either the adenoma or carcinoma group in the relative
-abundance of any OTU between the pre and post lesion removal (P-value
-\textgreater{} 0.05). A model built to classify model built to classify
-lesion had an AUC range of 0.811 - 0.866 while a model built to classify
-initial versus follow up samples had an AUC range of 0.641 - 0.805.
-Follow up samples had a decrease in the positive probability (P-value
-\textless{} 0.05). This decrease in positive probability was also
-observed for the initial sample model (P-value \textless{} 0.05). The
-lesion model used a total of 53 variables while the initial sample model
-used a total of 70 variables. A total of 23 OTUs were common to both
-models with the majority of these classifying to commensal bacteria
-(e.g. \emph{Bacteroides}, \emph{Clostridiales}, \emph{Blautia}, and
-\emph{Ruminococcaceae}).
+\textbf{Results:} The bacterial microbiome changed more in response to
+lesion removal in carcinoma cases compared to adenoma cases (P-value
+\textless{} 0.05). There was no difference for either the adenoma or
+carcinoma group in the relative abundance of any OTU between the pre and
+post lesion removal (P-value \textgreater{} 0.05). A model built to
+classify lesion had an AUC range of 0.811 - 0.866 while a model built to
+classify initial versus follow up samples had an AUC range of 0.641 -
+0.805. Follow up samples had a decrease in the positive probability of
+lesion (P-value \textless{} 0.05) suggesting a movement towards a more
+normal bacterial community. The initial sample model had a decrease in
+positive probability for the follow up samples tobe an initial sample
+(P-value \textless{} 0.05). The lesion model used a total of 53
+variables while the initial sample model used a total of 70 variables. A
+total of 23 OTUs were common to both models with the majority of these
+classifying to commensal bacteria (e.g. \emph{Bacteroides},
+\emph{Clostridiales}, \emph{Blautia}, and \emph{Ruminococcaceae}).
 
 \textbf{Conclusions:} Our data supports the hypothesis that there are
-differences in the bacterial microbiome between pre and post lesion
-removal samples. With individuals with carcinoma having more drastic
-differences to the overall community then those with adenoma. Changes to
-commensal bacteria were some of the most important variables for model
-classification, suggesting that these bacteria may be central to initial
-polyp formation and transition to carcinoma.
+changes in the bacterial microbiome following colorectal cancer lesion
+removal. Individuals with carcinoma have more drastic differences to the
+overall community then those with adenoma. Changes to commensal bacteria
+were some of the most important variables for model classification,
+suggesting that these bacteria may be central to initial polyp formation
+and transition to carcinoma.
 
@@
 related deaths and is currently the third most common cause of cancer
-deaths {[}1,2{]}. Over the last few years death due to the disease has
+deaths {[}1,2{]}. Over the last few years rate of disease mortality has
 seen a significant decrease, thanks mainly to improvements in screening
 {[}1{]}. However, despite this improvement there are still approximately
-50,000 deaths from the disease a year {[}2{]}.
+50,000 deaths from the disease per year {[}2{]}.
 
-Recently, there has been promising work on the bacterial microbiome and
-it's ability to complement existing screening methods such as Fecal
-Immunoglobulin Test (FIT) or act alone as a screening tool {[}3,4{]}.
-There has also been research into how the bacterial microbiome could be
-altered directly on tumor tissue itself {[}5{]}. A few studies have also
-shown how this bacterial microbiome {[}6{]} or specific members within
-it {[}7{]} could be directly involved with the pathogenesis of CRC.
-These studies have helped to provide a tantalizing link between the
-bacterial microbiome and CRC. Although these studies suggest that the
-bacterial microbiome might change after treatment there remains limited
-information on the bacterial microbiome before and after surgical
-removal of lesion (adenoma or carcinoma) and whether the community
-changes at all.
+Over the last few years there has been research into how the bacterial
+microbiome could be altered directly on tumor tissue itself {[}3{]}. A
+few studies have also shown how this bacterial microbiome {[}4{]} or
+specific members within it {[}5{]} could be directly involved with the
+pathogenesis of CRC. These studies have helped to provide a tantalizing
+link between the bacterial microbiome and CRC. Building off of these
+findings, there has been promising work on the bacterial microbiome and
+its ability to complement existing screening methods such as Fecal
+Immunoglobulin Test (FIT) or act alone as a screening tool {[}6,7{]}.
+Although these studies suggest that the bacterial microbiome might
+change after treatment there remains limited information on the
+bacterial microbiome response to lesion (adenoma or carcinoma) removal.
 
 In this study we tested the hypothesis that the bacterial microbiome
-changes between pre (initial) and post (follow up) samples removal of
-lesion (adenoma or carcinoma). Our analysis included both alpha and beta
-diversity analysis along with investigation of individual operational
-taxonomic units (OTUs). We also utilized Random Forest to build models
-to classify either lesion and normal samples or initial and follow up
-samples and then subsequently observe how these models, as well as
-specific OTUs within them, performed pre and post lesion removal. We
-also investigated the models for similar important OTUs to identify the
-crucial OTUs for not only classifying lesion or normalsamples but also
-initial and follow up.
+changes between pre (initial) and post (follow up) removal of lesion
+(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
+as well as specific Operational Taxonomic Units (OTUs) relative
+abundance. We utilized Random Forest to build two models. The first
+model was built to classify either lesion and normal samples while the
+second was built to classify initial and follow up samples. Subsequent
+observations on how these models, as well as specific OTUs within them,
+performed pre and post lesion removal helped inform us as to whether
+initial and follow up samples were changing and whether it was towards a
+more normal bacterial microbiome or not. We then investigated the two
+models for similar important OTUs to identify the crucial OTUs for not
+only classifying lesion or normal samples but also initial and follow
+up.
 
@@
 
-\textbf{\emph{Bacterial Community and FIT:}} We first wanted to test
-whether there were any broad differences between initial and follow up
-samples based on lesion being either adenoma or carcinoma. What we found
-was that the bacterial community in those with carcinoma were more
-dissimilar (as measured by thetayc) to their initial sample then those
-with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also found
-that there were larger changes in fecal blood (measured by FIT) for
-those with carcinoma versus adenoma (P-value \textless{} 0.0001)
-{[}Figure 1b{]}. The broad shift in bacterial community structure before
-and after surgery was visualized using NMDS for both adenoma
-{[}Figure1c{]} (PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure
-1d{]} (PERMANOVA \textless{} 0.05). Interestingly, when initial and
-follow up samples were compared to each other, regardless of whether
-they were adenoma or carcinoma (lesion), there was no significant
-overall difference between them (PERMANOVA \textgreater{} 0.05). When
-investigating more broad alpha diversity metrics there was no difference
-found between initial and follow up samples for lesion, adenoma only, or
-carcinoma only for any metric tested {[}Table S1{]}. We also observed
-that there was no difference in OTU relative abundance between initial
-and follow up samples for lesion, adenoma only, or carcinoma only
-{[}Figure S1{]}.
+\textbf{\emph{Bacterial Community and FIT:}} Within our 67 person cohort
+we first wanted to test whether there were any broad differences between
+initial and follow up samples based on lesion being either adenoma (n =
+41) or carcinoma (n = 26). We found that the bacterial community in
+those with carcinoma were more dissimilar to their initial sample than
+those with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also
+found that there were larger changes in fecal blood (measured by FIT)
+for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
+{[}Figure 1b{]}. The bacterial community structure before and after
+surgery was visualized using NMDS for both adenoma {[}Figure1c{]}
+(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1d{]} (PERMANOVA
+\textless{} 0.05). Interestingly, when initial and follow up samples
+were compared, regardless of whether the lesions were adenoma or
+carcinoma, there was no significant overall difference in beta diversity
+(PERMANOVA \textgreater{} 0.05). When investigating alpha diversity
+metrics there was no difference found between initial and follow up
+samples for lesion, adenoma only, or carcinoma only for any metric
+tested {[}Table S1{]}. We also observed that there was no difference in
+the relative abundance of any OTU between initial and follow up samples
+for lesion, adenoma only, or carcinoma only {[}Figure S1{]}.
 
@@
 suggested that a number of oral microbes may be important in CRC
-pathogenesis {[}8{]}. So we next examined whether there were differences
-in previously well described carcinoma associated OTUs. These included
-the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
-(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
-\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
-stomatis} (Otu001682). There was a decrease in relative abundance in
-initial and follow up samples for lesion and carcinoma for
-\emph{Parvimonas micra} (P-value \textless{} 0.05), and
-\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05)
-{[}Figure 2A{]}. In contrast, there was no difference in relative
-abundance in any of these OTUs for those with adenoma {[}Figure 2B{]}.
-We also observed that only a small percentage of those with adenoma or
-carcinoma were positive or had an appreciable relative abundance of any
-of these respective OTUs {[}Figure 2{]}.
+pathogenesis {[}8{]}. So we next examined whether there were changes in
+previously well described carcinoma associated OTUs, such as
+\emph{Porphyromonas asaccharolytica} (Otu000202), \emph{Fusobacterium
+nucleatum} (Otu000442), \emph{Parvimonas micra} (Otu001273), and
+\emph{Peptostreptococcus stomatis} (Otu001682). We first observed that
+only a small percentage of those with adenoma or carcinoma were positive
+or had an appreciable relative abundance for any of these respective
+OTUs {[}Figure 2{]}. Despite this, those with carcinoma had a decrease
+in relative abundance from initial to follow up for \emph{Parvimonas
+micra} (P-value \textless{} 0.05) and \emph{Porphyromonas
+asaccharolytica} (P-value \textless{} 0.05) {[}Figure 2A{]}. In
+contrast, there was no difference in relative abundance in any of these
+OTUs when considering only those with adenoma {[}Figure 2B{]}.
 
@@
 were any common bacterial microbiome changes in individuals with adenoma
-or carcinoma versus normal control individuals. We investigated this by
-creating a model to classify lesion versus normal based on the bacterial
+or carcinoma versus normal controls. We investigated this by creating a
+model to classify samples as lesion versus normal based on the bacterial
 community and FIT measurements. This model had an AUC range of 0.811 -
@@
 \textbf{\emph{The Initial Sample Model:}} After building a model to
-classify based on lesion we wanted to then build a separate model
-specifically to be able to classify whether samples were initial samples
-based on the bacterial community and FIT measurements. The initial
-sample model had an AUC range of 0.641 to 0.805 after 100 iterations of
-20 repeated 10-fold cross validations. The test set AUC range for this
-model performed better then the training set AUCs. There was a marked
-decrease in the ROC curve for the final model used when compared to the
-100 test set AUC iterations {[}Figure 4A{]}. There were a total of 70
-variables that were used for this model {[}Figure 4B{]}. The variable
-that resulted in the largest MDA was \emph{Pseudomonas} (Otu000438)
-while FIT measurement for fecal blood resulted in the sixth largest
-decrease in MDA {[}Figure 4B{]}.
+classify based on lesion we built a separate model specifically to be
+able to classify whether samples were initial samples based on the
+bacterial community and FIT measurements. The initial sample model had
+an AUC range of 0.641 to 0.805 after 100 iterations of 20 repeated
+10-fold cross validations. The test set AUC range for this model
+performed better then the training set AUCs. There was a marked decrease
+in the ROC curve for the final model used when compared to the 100 test
+set AUC iterations {[}Figure 4A{]}. There were a total of 70 variables
+that were used for this model {[}Figure 4B{]}. The variable that
+resulted in the largest MDA was \emph{Pseudomonas} (Otu000438) while FIT
+measurement for fecal blood resulted in the sixth largest decrease in
+MDA {[}Figure 4B{]}.
 
@@
 treated with chemotherapy for the initial sample model (P-value
-\textless{} 0.05). All other variables that were tested showed no
-difference based on whether chemotherapy or radiation was used {[}Table
-S3{]}.
+\textless{} 0.05). This suggests that those treated for chemotherapy my
+have had a larger change from the initial sample then those without such
+treatment. All other variables that were tested showed no difference
+based on whether chemotherapy or radiation was used {[}Table S3{]}.
 
@@
 
+This study builds upon previous work from numerous labs that have looked
+into the bacterial microbiome as a potential screening tool
+{[}6,7,9--11{]} by exploring what happens to the bacterial community
+after surgical removal of a lesion. Based on previous work by Arthur, et
+al. {[}12{]} it may not be surprising to have E.coli as one of the most
+important OTUs and one that was common to both models. Interestingly,
+many of the most important OTUs had taxonomic identification for
+resident gut microbes. This could suggest that the bacterial community
+is one of the first components that could change during the pathogenesis
+of disease. These bacterial microbiome changes could be the first step
+in allowing more inflammatory bacterium to gain a foothold within the
+colon {[}8{]}.
+
 From our results there were large observed differences in the bacterial
-microbiome between pre and post lesion removal samples based on whether
-the individual had an adenoma or carcinoma. There were much larger
-differences between initial and follow up samples based on the thetayc
-distance metric and in fecal blood as measured by FIT for individuals
-with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
-differences between initial and follow up samples for Shannon Diversity,
-observed OTUs, or evenness regardless of whether the individual had an
+microbiome in samples before and after lesion removal based on whether
+the individual had an adenoma or carcinoma. In individuals with
+carcinoma compared to adenoma there were much larger differences between
+initial and follow up samples based on the community beta diversity and
+in fecal blood as measured by FIT {[}Figure 1{]}. However, there were no
+differences between initial and follow up samples for any alpha
+diversity metric measured regardless of whether the individual had an
 adenoma or carcinoma {[}Table S1{]}. There was also no differences in
@@
 We next created a model that incorporated FIT and the bacterial
-microbiome to be able to classify lesions (adenoma or carcinoma). What
-we found was that not only was the positive probability of a lesion
-reduced after lesion removal {[}Figure 3 \& 4{]} but also that the
-commonly associated CRC bacteria were not highly represented within this
-model. Rather, OTUs that made up the most important variables
-overwhelmingly belonged to commensal bacteria. With only a single OTU
-from a previously associated cancer bacterium (\emph{Porphyromonas
-asaccharolytica}). We followed this up by creating a second model to
+microbiome to be able to classify lesions (adenoma or carcinoma). Based
+on this model we found that the follow up samples were closer to normal
+then the initial samples due to a decrease in positive probability for
+lesion and that the commonly associated CRC bacteria were not highly
+represented within this model with the exception of \emph{Porphyromonas
+asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
+and this change was towards what would be expected for normal controls,
+it should be noted that the follow up samples may not be a completely
+normal bacterial microbiome.
+
+After creating the lesion mode we then created a second model to
 classify initial versus follow up samples. We found that this model was
-able to accurately classify initial versus follow up samples {[}Figure 3
-\& 4{]}. Providing additional information on the importance of commensal
-bacteria was that there were a total of 23 OTUs in common to both models
-and the vast majority belonged to regular residents of our gut community
-{[}Table S2{]}. Although there was a detectable change in the carcinoma
-group and this change was towards what would be expected for normal
-controls, the follow up samples may not be a completely normal bacterial
-microbiome.
+able to accurately classify initial versus follow up samples suggesting
+that regardless of adenoma or carcinoma there are distinct common
+changes within the bacterial microbiome that occurs after lesion removal
+{[}Figure 4{]}. Both models had OTUs that overwhelmingly belonged to
+commensal bacteria. Providing additional information on the importance
+of commensal bacteria was that there were a total of 23 OTUs in common
+to both models and the vast majority belonged to regular residents of
+our gut community {[}Table S2{]}.
 
-For the majority of tests performed there were no differences in the
-bacterial microbiome based on whether chemotherapy or radiation was
-received {[}Table S3{]}. Within our study there was a significant
-difference for the time elapsed in the collection of the follow up
-sample between adenoma and carcinoma (uncorrected P-value \textless{}
-0.05), with time passed being less for adenoma (253 +/- 41.3 days) then
-carcinoma (351 +/- 102 days). These results would indicate that the
-findings described were specific to the surgical intervention and that
-some of the differences observed between carcinoma and adenoma samples
-could be due to differences in collection time between samples for the
-two different groups. Specifically, it could confound the observation
-that carcinomas changed more than adenomas {[}Figure 1a \& 1d{]}.
-
-This study builds upon previous work from numerous labs that have looked
-into the bacterial microbiome as a potential screening tool
-{[}3,4,9--11{]} by exploring what happens to the bacterial community
-after surgical removal of a lesion. Based on previous work by Arthur, et
-al. {[}12{]} it may not be surprising to have E.coli as one of the most
-important OTUs and one that was common to both models. Interestingly,
-many of the most important OTUs had taxonomic identification for
-resident gut microbes. This could suggest that the bacterial community
-is one of the first components that could change during the pathogenesis
-of disease. These bacterial microbiome changes could be the first step
-in allowing more inflammatory bacterium to gain a foothold within the
-colon {[}8{]}.
+Within our study there was a significant difference for the time elapsed
+in the collection of the follow up sample between adenoma and carcinoma
+(uncorrected P-value \textless{} 0.05), with time passed being less for
+adenoma (253 +/- 41.3 days) than carcinoma (351 +/- 102 days). These
+results would indicate that the findings described were specific to the
+surgical intervention and that some of the differences observed between
+carcinoma and adenoma samples could be due to differences in collection
+time between samples for the two different groups. Specifically, it
+could confound the observation that carcinomas changed more than
+adenomas {[}Figure 1a \& 1d{]}. This confounding though would not affect
+the observations where these individuals were grouped together.
 
@@
 80/20 splits to try and mimic the type of variation that might be
-expected to occur. The time difference in collection of sample between
-adenoma and carcinoma could have affected our observations for
-differences between individuals with adenoma or carcinoma. This
-confounding though would not affect the observations where these
-individuals are grouped together.
+expected to occur.
 
-Interestingly, within figure 4 the initial sample model showed better
-test AUC results then the training set AUC. This may have occurred
-because the training AUC determined from 20 repeated 10 fold cross
-validation removed samples at random and did not take into account that
-they were matched samples. Another potential reason is that the model
-itself may be over-fit since the total number of samples was not that
-large. However, the lesion model did not suffer from these discrepancies
-and similar conclusions can be drawn solely from this model alone.
+Interestingly, within the initial sample model the test data performed
+better than the training data. This may have occurred because the
+training AUC determined from 20 repeated 10 fold cross validation
+removed samples at random and did not take into account that they were
+matched samples. Another potential reason is that the model itself may
+be over-fit since the total number of samples was not that large.
+However, the lesion model did not suffer from these discrepancies.
 Further independent studies need to be carried out to verify our
-findings since not only are we dealing with feces, which could be very
-different than the communities present on the actual tissue, but also
-are dealing with correlations that may not be representative of the true
-pathogenesis of disease.
+findings and models since we are dealing with correlations that may not
+be truly representative of the pathogenesis of disease.
 
@@
 \textbf{\emph{Study Design and Patient Sampling:}} The sampling and
-design were similar to that reported in Baxter, et al {[}3{]}. In brief,
+design were similar to that reported in Baxter, et al {[}6{]}. In brief,
 study exclusion involved those who had already undergone surgery,
@@
 
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-3. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
-improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine. 2016;8:37.
-
-\hypertarget{ref-zeller_potential_2014}{}
-4. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
-Potential of fecal microbiota for early-stage detection of colorectal
-cancer. Molecular Systems Biology. 2014;10:766.
-
 \hypertarget{ref-dejea_microbiota_2014}{}
-5. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+3. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 \hypertarget{ref-zackular_manipulation_2016}{}
-6. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+4. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
 Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
@@
 \hypertarget{ref-arthur_microbial_2014}{}
-7. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+5. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
 McCafferty J, et al. Microbial genomic analysis reveals the essential
@@
 Communications. 2014;5:4724.
+
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+6. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. 2016;8:37.
+
+\hypertarget{ref-zeller_potential_2014}{}
+7. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+Potential of fecal microbiota for early-stage detection of colorectal
+cancer. Molecular Systems Biology. 2014;10:766.
 

### 2017-02-22 submission/manuscript_outline_20161024.tex
read through edits to streamline Amanda suggestions

@@
 \textless{} 0.05). There was no difference for either the adenoma or
-carcinoma group in the relative abundance of any OTU between the pre and
+carcinoma group in the relative abundance of any OTU between pre and
 post lesion removal (P-value \textgreater{} 0.05). A model built to
 classify lesion had an AUC range of 0.811 - 0.866 while a model built to
-classify initial versus follow up samples had an AUC range of 0.641 -
-0.805. Follow up samples had a decrease in the positive probability of
-lesion (P-value \textless{} 0.05) suggesting a movement towards a more
-normal bacterial community. The initial sample model had a decrease in
-positive probability for the follow up samples tobe an initial sample
-(P-value \textless{} 0.05). The lesion model used a total of 53
-variables while the initial sample model used a total of 70 variables. A
-total of 23 OTUs were common to both models with the majority of these
-classifying to commensal bacteria (e.g. \emph{Bacteroides},
-\emph{Clostridiales}, \emph{Blautia}, and \emph{Ruminococcaceae}).
+classify initial samples had an AUC range of 0.641 - 0.805. Follow up
+samples had a decrease in the positive probability of lesion (P-value
+\textless{} 0.05) suggesting a movement towards a more normal bacterial
+community. The initial sample model had a decrease in positive
+probability for the follow up samples to be an initial sample (P-value
+\textless{} 0.05). The lesion model used a total of 53 variables while
+the initial sample model used a total of 70 variables. A total of 23
+OTUs were common to both models with the majority of these classifying
+to commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
+\emph{Blautia}, and \emph{Ruminococcaceae}).
 
@@
 related deaths and is currently the third most common cause of cancer
-deaths {[}1,2{]}. Over the last few years rate of disease mortality has
-seen a significant decrease, thanks mainly to improvements in screening
-{[}1{]}. However, despite this improvement there are still approximately
-50,000 deaths from the disease per year {[}2{]}.
+deaths {[}1,2{]}. Over the last few years the rate of disease mortality
+has seen a significant decrease, thanks mainly to improvements in
+screening {[}1{]}. However, despite this improvement there are still
+approximately 50,000 deaths from the disease per year {[}2{]}.
 
-Over the last few years there has been research into how the bacterial
-microbiome could be altered directly on tumor tissue itself {[}3{]}. A
-few studies have also shown how this bacterial microbiome {[}4{]} or
-specific members within it {[}5{]} could be directly involved with the
-pathogenesis of CRC. These studies have helped to provide a tantalizing
-link between the bacterial microbiome and CRC. Building off of these
-findings, there has been promising work on the bacterial microbiome and
-its ability to complement existing screening methods such as Fecal
-Immunoglobulin Test (FIT) or act alone as a screening tool {[}6,7{]}.
-Although these studies suggest that the bacterial microbiome might
-change after treatment there remains limited information on the
+Over the last few years a few studies have shown how this bacterial
+microbiome {[}3{]} or specific members within it {[}4{]} could be
+directly involved with the pathogenesis of CRC. There has also been
+research into how the bacterial microbiome could be altered directly on
+tumor tissue itself {[}5{]}. These studies have helped to provide a
+tantalizing link between the bacterial microbiome and CRC. Building off
+of these findings, there has been promising work on the bacterial
+microbiome and its ability to complement existing screening methods such
+as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
+{[}6,7{]}. Although these studies suggest that the bacterial microbiome
+might change after treatment there remains limited information on the
 bacterial microbiome response to lesion (adenoma or carcinoma) removal.
@@
 abundance. We utilized Random Forest to build two models. The first
-model was built to classify either lesion and normal samples while the
-second was built to classify initial and follow up samples. Subsequent
+model was built to classify lesion and normal samples while the second
+was built to classify initial and follow up samples. Subsequent
 observations on how these models, as well as specific OTUs within them,
@@
 initial and follow up samples were changing and whether it was towards a
-more normal bacterial microbiome or not. We then investigated the two
+more normal bacterial microbiome or not. We also investigated the two
 models for similar important OTUs to identify the crucial OTUs for not
@@
 those with carcinoma were more dissimilar to their initial sample than
-those with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also
+those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. We also
 found that there were larger changes in fecal blood (measured by FIT)
 for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
-{[}Figure 1b{]}. The bacterial community structure before and after
-surgery was visualized using NMDS for both adenoma {[}Figure1c{]}
-(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1d{]} (PERMANOVA
+{[}Figure 1B{]}. The bacterial community structure before and after
+surgery was visualized using NMDS for both adenoma {[}Figure 1C{]}
+(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1D{]} (PERMANOVA
 \textless{} 0.05). Interestingly, when initial and follow up samples
@@
 treated with chemotherapy for the initial sample model (P-value
-\textless{} 0.05). This suggests that those treated for chemotherapy my
-have had a larger change from the initial sample then those without such
-treatment. All other variables that were tested showed no difference
-based on whether chemotherapy or radiation was used {[}Table S3{]}.
+\textless{} 0.05). This suggests that follow up samples for those
+treated for chemotherapy my have had a larger change from the initial
+sample then those without such treatment. All other variables that were
+tested showed no difference based on whether chemotherapy or radiation
+was used {[}Table S3{]}.
 
@@
 could confound the observation that carcinomas changed more than
-adenomas {[}Figure 1a \& 1d{]}. This confounding though would not affect
-the observations where these individuals were grouped together.
+adenomas {[}Figure 1A \& 1D{]}. This confounding though would not affect
+the observations where these individuals were grouped together {[}Figure
+3 \& 4{]}.
 
@@
 
-Many of the common OTUs between the two models used had many OTUs that
+Many of the common OTUs between the two models had OTUs that
 taxonomically classified to potential butyrate producers {[}Table S2{]}.
@@
 Further independent studies need to be carried out to verify our
-findings and models since we are dealing with correlations that may not
-be truly representative of the pathogenesis of disease.
+findings since we are dealing with correlations that may not be truly
+representative of the pathogenesis of disease.
 
@@
 
+\hypertarget{ref-zackular_manipulation_2016}{}
+3. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
+
+\hypertarget{ref-arthur_microbial_2014}{}
+4. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+McCafferty J, et al. Microbial genomic analysis reveals the essential
+role of inflammation in bacteria-induced colorectal cancer. Nature
+Communications. 2014;5:4724.
+
 \hypertarget{ref-dejea_microbiota_2014}{}
-3. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+5. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 Sciences of the United States of America. 2014;111:18321--6.
-
-\hypertarget{ref-zackular_manipulation_2016}{}
-4. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
-Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
-
-\hypertarget{ref-arthur_microbial_2014}{}
-5. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
-McCafferty J, et al. Microbial genomic analysis reveals the essential
-role of inflammation in bacteria-induced colorectal cancer. Nature
-Communications. 2014;5:4724.
 

### 2017-02-22 submission/manuscript_outline_20161024.Rmd

@@
 **Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
-There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
+There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
+A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
 Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.
-The initial sample model had a decrease in positive probability for the follow up samples tobe an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+The initial sample model had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
@@
 Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
-Over the last few years rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
+Over the last few years the rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
 However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
 
-Over the last few years there has been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
-A few studies have also shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
+Over the last few years a few studies have shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
+There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
 These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
@@
 We utilized Random Forest to build two models.
-The first model was built to classify either lesion and normal samples while the second was built to classify initial and follow up samples.
+The first model was built to classify lesion and normal samples while the second was built to classify initial and follow up samples.
 Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.
-We then investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.
+We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.
 
@@
 Within our `r length(rownames(good_metaf))` person cohort we first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`).
-We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a].
-We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1b].
-The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
+We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1A].
+We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1B].
+The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
 Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
@@
 There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
-This suggests that those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.
+This suggests that follow up samples for those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.
 All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].
@@
 These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
-Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].
-This confounding though would not affect the observations where these individuals were grouped together.
+Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1D].
+This confounding though would not affect the observations where these individuals were grouped together [Figure 3 & 4].
 
@@
 
-Many of the common OTUs between the two models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].
+Many of the common OTUs between the two models had OTUs that taxonomically classified to potential butyrate producers [Table S2].
 Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.
@@
 However, the lesion model did not suffer from these discrepancies.
-Further independent studies need to be carried out to verify our findings and models since we are dealing with correlations that may not be truly representative of the pathogenesis of disease.
+Further independent studies need to be carried out to verify our findings since we are dealing with correlations that may not be truly representative of the pathogenesis of disease.
 

### 2017-02-23 code/Run_Figure1.R
add uniformity to colors

@@ -29,10 +29,10 @@ differences_graph <- grid.arrange(
     stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
     stat_summary(fun.y = mean, colour = "black", geom = "line") + 
     scale_color_manual(name = "Lesion Type", 
-                       values = wes_palette("GrandBudapest"), 
+                       values = c('#00FFFF', '#0000FF', '#F1BB7B'), 
                        breaks = c("adenoma", "adv_adenoma", "cancer"), 
                        labels = c("Adenoma", "SRN", "Carcinoma")) + 
-    coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + 
+    coord_cartesian(ylim = c(0, 1)) + ylab(expression(paste(theta[YC], " Distance"))) + 
     xlab("") + theme_bw() + ggtitle("A") + 
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
           plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "lines"), 
@@ -51,10 +51,10 @@ differences_graph <- grid.arrange(
     stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
     stat_summary(fun.y = mean, colour = "black", geom = "line") + 
     scale_color_manual(name = "Lesion Type", 
-                       values = wes_palette("GrandBudapest"), 
+                       values = c('#00FFFF', '#0000FF', '#F1BB7B'), 
                        breaks = c("adenoma", "adv_adenoma", "cancer"), 
                        labels = c("Adenoma", "SRN", "Carcinoma")) + 
-    ylab("Change in FIT from Follow up to Initial") + 
+    ylab("Change in FIT from Follow Up to Initial") + 
     xlab("") + theme_bw() + ggtitle("B") + 
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
           plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "lines"), 
@@ -66,32 +66,40 @@ differences_graph <- grid.arrange(
               x = 0.90, y = -2250, size = 2.5), 
   
   # Adenoma Only initial and follow up
-  ggplot(data = thetayc_adn_IF, 
-         aes(x=NMDS1, y=NMDS2)) + geom_point(aes(color=samples)) + 
[…]

### 2017-02-23 code/Run_Figure2.R

@@ -36,7 +36,7 @@ crc_specific <- grid.arrange(
     geom_point(aes(color = factor(Disease_Free, levels = c("n", "y", "unknown")))) + 
     facet_wrap(~labs, labeller = label_parsed, scales = "free_y") + 
     theme_bw() + ylab("% Relative Abundance") + xlab("") + ggtitle("A") +  
-    scale_colour_manual(name = "Cancer Free", 
+    scale_colour_manual(name = "Cancer Free\n on Follow Up", 
                         label = c("No", "Yes", "Unknown"),  
                         values = wes_palette("GrandBudapest")) + 
     scale_x_discrete(

### 2017-02-23 code/Run_Figure3.R

@@ -86,8 +86,9 @@ Lesion_plot <- grid.arrange(
                       levels = c("initial", "followup")), Yes, group = factor(EDRN))) + 
     geom_point(aes(color=factor(disease_free, 
                                 levels = c("n", "y", "unknown"))), size = 2) + 
-    geom_line(linetype = 2, alpha = 0.3) + 
-    scale_color_manual(name = "Cancer Free", 
+    geom_line(aes(color=factor(disease_free, 
+                               levels = c("n", "y", "unknown"))), linetype = 2, alpha = 0.6) + 
+    scale_color_manual(name = "Cancer Free\n on Follow Up", 
                        label = c("No", "Yes", "Unknown"),  
                        values = wes_palette("GrandBudapest")) + 
     scale_x_discrete(breaks = c("initial", "followup"), 
@@ -98,8 +99,9 @@ Lesion_plot <- grid.arrange(
     xlab("") + theme_bw() + theme(
       axis.title = element_text(face="bold"), 
       legend.title = element_text(face="bold"), 
-      legend.position = c(0.25, 0.15), 
-      plot.title = element_text(face="bold", hjust = 0)), 
+      legend.position = c(0.25, 0.20), 
+      plot.title = element_text(face="bold", hjust = 0)) + 
+    annotate("text", label = paste("Carcinoma"), x = 1.5, y = 1.02, size = 2.5), 
   
     # Graph the lesion adenoma data only
   filter(red_follow_up_probability, diagnosis == "adenoma") %>%
@@ -120,7 +122,8 @@ Lesion_plot <- grid.arrange(
       axis.title = element_text(face="bold"), 
       legend.title = element_text(face="bold"), 
       legend.position = c(0.20, 0.12), 
-      plot.title = element_text(face="bold", hjust = 0))
+      plot.title = element_text(face="bold", hjust = 0)) + 
+    annotate("text", label = paste("Adenoma"), x = 1.5, y = 1.02, size = 2.5)
   
 )
 

### 2017-02-23 submission/manuscript_outline_20161024.Rmd

@@
 adenoma or carcinoma) removal.
-Specifically, we wanted to identify what within the community was different before and after removal of said lesion.
+Specifically, we wanted to identify components within the community that were different before and after removal of said lesion.
 
@@
 There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
-Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.
-The initial sample model had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.
+A second model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
@@
 
-Over the last few years a few studies have shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
+Over the last few years studies have shown how the bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
 There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
@@
 In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).
-We analyzed changes in alpha and beta diversity as well as specific Operational Taxonomic Units (OTUs) relative abundance.
+We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
 We utilized Random Forest to build two models.
-The first model was built to classify lesion and normal samples while the second was built to classify initial and follow up samples.
+The first model was built to classify lesion and normal (non-lesion) samples while the second was built to classify initial and follow up samples.
 Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.
@@
 So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-We first observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance for any of these respective OTUs [Figure 2].
+We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].
 Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A].
@@
 ***The Lesion Model:***
-We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal controls.
+We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma and carcinoma versus normal controls.
 We investigated this by creating a model to classify samples as lesion versus normal based on the bacterial community and FIT measurements.
@@
 ***The Initial Sample Model:***
-After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.
+After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial (before lesion was removed) samples based on the bacterial community and FIT measurements.
 The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
@@
 
-This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after surgical removal of a lesion.
+This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after lesion removal.
 Based on previous work by Arthur, et al.
@@
 
-We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).
+We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma and carcinoma).
 Based on this model we found that the follow up samples were closer to normal then the initial samples due to a decrease in positive probability for lesion and that the commonly associated CRC bacteria were not highly represented within this model with the exception of *Porphyromonas asaccharolytica* [Figure 3].
-Although there was a detectable change and this change was towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.
+Although there was a detectable change towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.
 
@@
 Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].
-It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.
+It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after removal of lesion.
 In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].
@@
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].
+The bacterial microbiome has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013].
 Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after lesion removal and treatment had been completed.
@@
 ***Lesion Model Creation:***
-The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
+The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on an independent data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
 This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.
+It should be noted that all individuals with an adenoma or carcinoma were grouped together to form the lesion group and the model was not created to find differences between normal, adenoma, and carcinoma but rather differences between both adenoma and carcinoma versus normal.
 
@@
 Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.
-Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial to lesion removal sample.
+Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial sample prior to lesion removal.
 
@@
 Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.
-Where multiple comparison testing was appropriate a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
+Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
 Unless otherwise stated the P-values reported are those that were BH corrected.
@@
 We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.
-Finally, a list of common OTUs were found for the two two different models used.
+Finally, a list of common OTUs were found for the two different models used.
 
@@
 **Figure 1: General Differences between the Adenoma and Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).
 B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-C) NMDS of the initial and follow up samples for the Adenoma group.
-D) NMDS of the initial and follow up samples for the Carcinoma group.
-For C) and D) the teal represents initial samples and the pink represents follow up samples.
+Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).
+C) NMDS of the initial and follow up samples for the adenoma group.
+D) NMDS of the initial and follow up samples for the carcinoma group.
 

### 2017-02-23 submission/manuscript_outline_20161024.tex

@@
 microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
-Specifically, we wanted to identify what within the community was
-different before and after removal of said lesion.
+Specifically, we wanted to identify components within the community that
+were different before and after removal of said lesion.
 
@@
 post lesion removal (P-value \textgreater{} 0.05). A model built to
-classify lesion had an AUC range of 0.811 - 0.866 while a model built to
-classify initial samples had an AUC range of 0.641 - 0.805. Follow up
-samples had a decrease in the positive probability of lesion (P-value
+classify lesion had an AUC range of 0.811 - 0.866 and follow up samples
+had a decrease in the positive probability of lesion (P-value
 \textless{} 0.05) suggesting a movement towards a more normal bacterial
-community. The initial sample model had a decrease in positive
-probability for the follow up samples to be an initial sample (P-value
-\textless{} 0.05). The lesion model used a total of 53 variables while
-the initial sample model used a total of 70 variables. A total of 23
-OTUs were common to both models with the majority of these classifying
-to commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
+community. A second model built to classify initial samples had an AUC
+range of 0.641 - 0.805 and had a decrease in positive probability for
+the follow up samples to be an initial sample (P-value \textless{}
+0.05). The lesion model used a total of 53 variables while the initial
+sample model used a total of 70 variables. A total of 23 OTUs were
+common to both models with the majority of these classifying to
+commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
 \emph{Blautia}, and \emph{Ruminococcaceae}).
@@
 
-Over the last few years a few studies have shown how this bacterial
-microbiome {[}3{]} or specific members within it {[}4{]} could be
-directly involved with the pathogenesis of CRC. There has also been
-research into how the bacterial microbiome could be altered directly on
-tumor tissue itself {[}5{]}. These studies have helped to provide a
-tantalizing link between the bacterial microbiome and CRC. Building off
-of these findings, there has been promising work on the bacterial
-microbiome and its ability to complement existing screening methods such
-as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
-{[}6,7{]}. Although these studies suggest that the bacterial microbiome
-might change after treatment there remains limited information on the
-bacterial microbiome response to lesion (adenoma or carcinoma) removal.
+Over the last few years studies have shown how the bacterial microbiome
+{[}3{]} or specific members within it {[}4{]} could be directly involved
+with the pathogenesis of CRC. There has also been research into how the
+bacterial microbiome could be altered directly on tumor tissue itself
+{[}5{]}. These studies have helped to provide a tantalizing link between
+the bacterial microbiome and CRC. Building off of these findings, there
+has been promising work on the bacterial microbiome and its ability to
+complement existing screening methods such as Fecal Immunoglobulin Test
+(FIT) or act alone as a screening tool {[}6,7{]}. Although these studies
+suggest that the bacterial microbiome might change after treatment there
+remains limited information on the bacterial microbiome response to
+lesion (adenoma or carcinoma) removal.
 
@@
 (adenoma or carcinoma). We analyzed changes in alpha and beta diversity
-as well as specific Operational Taxonomic Units (OTUs) relative
-abundance. We utilized Random Forest to build two models. The first
-model was built to classify lesion and normal samples while the second
-was built to classify initial and follow up samples. Subsequent
-observations on how these models, as well as specific OTUs within them,
-performed pre and post lesion removal helped inform us as to whether
-initial and follow up samples were changing and whether it was towards a
-more normal bacterial microbiome or not. We also investigated the two
-models for similar important OTUs to identify the crucial OTUs for not
-only classifying lesion or normal samples but also initial and follow
-up.
+as well as the relative abundance of specific Operational Taxonomic
+Units (OTUs). We utilized Random Forest to build two models. The first
+model was built to classify lesion and normal (non-lesion) samples while
+the second was built to classify initial and follow up samples.
+Subsequent observations on how these models, as well as specific OTUs
+within them, performed pre and post lesion removal helped inform us as
+to whether initial and follow up samples were changing and whether it
+was towards a more normal bacterial microbiome or not. We also
+investigated the two models for similar important OTUs to identify the
+crucial OTUs for not only classifying lesion or normal samples but also
+initial and follow up.
 
@@
 only a small percentage of those with adenoma or carcinoma were positive
-or had an appreciable relative abundance for any of these respective
-OTUs {[}Figure 2{]}. Despite this, those with carcinoma had a decrease
-in relative abundance from initial to follow up for \emph{Parvimonas
-micra} (P-value \textless{} 0.05) and \emph{Porphyromonas
-asaccharolytica} (P-value \textless{} 0.05) {[}Figure 2A{]}. In
-contrast, there was no difference in relative abundance in any of these
-OTUs when considering only those with adenoma {[}Figure 2B{]}.
+or had a relative abundance above 0.1\% for any of these respective OTUs
+{[}Figure 2{]}. Despite this, those with carcinoma had a decrease in
+relative abundance from initial to follow up for \emph{Parvimonas micra}
+(P-value \textless{} 0.05) and \emph{Porphyromonas asaccharolytica}
+(P-value \textless{} 0.05) {[}Figure 2A{]}. In contrast, there was no
+difference in relative abundance in any of these OTUs when considering
+only those with adenoma {[}Figure 2B{]}.
 
@@
 were any common bacterial microbiome changes in individuals with adenoma
-or carcinoma versus normal controls. We investigated this by creating a
+and carcinoma versus normal controls. We investigated this by creating a
 model to classify samples as lesion versus normal based on the bacterial
@@
 classify based on lesion we built a separate model specifically to be
-able to classify whether samples were initial samples based on the
-bacterial community and FIT measurements. The initial sample model had
-an AUC range of 0.641 to 0.805 after 100 iterations of 20 repeated
-10-fold cross validations. The test set AUC range for this model
-performed better then the training set AUCs. There was a marked decrease
-in the ROC curve for the final model used when compared to the 100 test
-set AUC iterations {[}Figure 4A{]}. There were a total of 70 variables
-that were used for this model {[}Figure 4B{]}. The variable that
-resulted in the largest MDA was \emph{Pseudomonas} (Otu000438) while FIT
-measurement for fecal blood resulted in the sixth largest decrease in
-MDA {[}Figure 4B{]}.
+able to classify whether samples were initial (before lesion was
+removed) samples based on the bacterial community and FIT measurements.
+The initial sample model had an AUC range of 0.641 to 0.805 after 100
+iterations of 20 repeated 10-fold cross validations. The test set AUC
+range for this model performed better then the training set AUCs. There
+was a marked decrease in the ROC curve for the final model used when
+compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
+a total of 70 variables that were used for this model {[}Figure 4B{]}.
+The variable that resulted in the largest MDA was \emph{Pseudomonas}
+(Otu000438) while FIT measurement for fecal blood resulted in the sixth
+largest decrease in MDA {[}Figure 4B{]}.
 
@@
 {[}6,7,9--11{]} by exploring what happens to the bacterial community
-after surgical removal of a lesion. Based on previous work by Arthur, et
-al. {[}12{]} it may not be surprising to have E.coli as one of the most
-important OTUs and one that was common to both models. Interestingly,
-many of the most important OTUs had taxonomic identification for
-resident gut microbes. This could suggest that the bacterial community
-is one of the first components that could change during the pathogenesis
-of disease. These bacterial microbiome changes could be the first step
-in allowing more inflammatory bacterium to gain a foothold within the
-colon {[}8{]}.
+after lesion removal. Based on previous work by Arthur, et al. {[}12{]}
+it may not be surprising to have E.coli as one of the most important
+OTUs and one that was common to both models. Interestingly, many of the
+most important OTUs had taxonomic identification for resident gut
+microbes. This could suggest that the bacterial community is one of the
+first components that could change during the pathogenesis of disease.
+These bacterial microbiome changes could be the first step in allowing
+more inflammatory bacterium to gain a foothold within the colon {[}8{]}.
 
@@
 We next created a model that incorporated FIT and the bacterial
-microbiome to be able to classify lesions (adenoma or carcinoma). Based
+microbiome to be able to classify lesions (adenoma and carcinoma). Based
 on this model we found that the follow up samples were closer to normal
@@
 asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
-and this change was towards what would be expected for normal controls,
-it should be noted that the follow up samples may not be a completely
-normal bacterial microbiome.
+towards what would be expected for normal controls, it should be noted
+that the follow up samples may not be a completely normal bacterial
+microbiome.
 
@@
 disease but rather that they are not the main bacteria changing after
-surgical removal of lesion. In fact, it is possible that these CRC
-associated bacteria are important in the transition from adenoma to
-carcinoma and would be one explanation as to why in our data we not only
-see high initial relative abundances in carcinoma and not adenoma
-individuals but also large decreases in relative abundance in some of
-those with carcinoma but not in those with adenoma after surgery
-{[}Figure 2{]}.
+removal of lesion. In fact, it is possible that these CRC associated
+bacteria are important in the transition from adenoma to carcinoma and
+would be one explanation as to why in our data we not only see high
+initial relative abundances in carcinoma and not adenoma individuals but
+also large decreases in relative abundance in some of those with
+carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.
 
@@
 prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
-microbiome has been shown to normalize within this time period {[}18{]}.
-Our follow up data set had a total of 67 individuals that not only had a
-sample as described but also a follow up sample between 188 - 546 days
-after lesion removal and treatment had been completed. This study was
-approved by the University of Michigan Institutional Review Board. All
-study participants provided informed consent and the study itself
-conformed to the guidelines set out by the Helsinki Declaration.
+microbiome has been shown to normalize back to a pre-colonscopy
+community within this time period {[}18{]}. Our follow up data set had a
+total of 67 individuals that not only had a sample as described but also
+a follow up sample between 188 - 546 days after lesion removal and
+treatment had been completed. This study was approved by the University
+of Michigan Institutional Review Board. All study participants provided
+informed consent and the study itself conformed to the guidelines set
+out by the Helsinki Declaration.
 
@@
 (adenoma or carcinoma) with the main training and testing of the model
-completed on a data set of 423 individuals. This model was then applied
-to our follow up data set of 67 individuals.
+completed on an independent data set of 423 individuals. This model was
+then applied to our follow up data set of 67 individuals. It should be
+noted that all individuals with an adenoma or carcinoma were grouped
+together to form the lesion group and the model was not created to find
+differences between normal, adenoma, and carcinoma but rather
+differences between both adenoma and carcinoma versus normal.
 
@@
 Instead of classifying samples as positive or negative of lesion this
-model classified samples as positive or negative for being an initial to
-lesion removal sample.
+model classified samples as positive or negative for being an initial
+sample prior to lesion removal.
 
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate a Benjamini-Hochberg (BH) correction was applied {[}23{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}23{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 initial sample in the follow up samples. Finally, a list of common OTUs
-were found for the two two different models used.
+were found for the two different models used.
 
@@
 Group.} A) A significant difference was found between the adenoma and
-carcinoma group for thetayc (P-value = 0.000472). B) A significant
-difference was found between the adenoma and carcinoma group for change
-in FIT measurement (P-value = 2.15e-05). C) NMDS of the initial and
-follow up samples for the Adenoma group. D) NMDS of the initial and
-follow up samples for the Carcinoma group. For C) and D) the teal
-represents initial samples and the pink represents follow up samples.
+carcinoma group for thetayc (P-value = 0.000472). Advanced adenomas are
+denoted as Screen Relevant Neoplasia (SRN). B) A significant difference
+was found between the adenoma and carcinoma group for change in FIT
+measurement (P-value = 2.15e-05). Advanced adenomas are denoted as
+Screen Relevant Neoplasia (SRN). C) NMDS of the initial and follow up
+samples for the adenoma group. D) NMDS of the initial and follow up
+samples for the carcinoma group.
 

### 2017-02-23 submission/manuscript_outline_20161024.Rmd
read through after all additions

@@
 The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e.
-adenoma or carcinoma) removal.
+adenoma and carcinoma) removal.
 Specifically, we wanted to identify components within the community that were different before and after removal of said lesion.
@@
 
-**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal cancer lesion removal.
+**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal lesion removal.
 Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.
@@
 Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.
+Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma and carcinoma) removal.
 
-In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).
+In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion.
 We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
 We utilized Random Forest to build two models.
-The first model was built to classify lesion and normal (non-lesion) samples while the second was built to classify initial and follow up samples.
+The first model was built to classify lesion versus non-lesion (normal) samples while the second was built to classify initial versus follow up samples.
 Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.
-We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.
+We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial versus follow up.
 
@@
 It is also possible that both of these explanations could have played a role.
-Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].
+Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after removal [Figure 3C & 3D].
 It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after removal of lesion.
-In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].
+In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after lesion removal [Figure 2].
 
@@
 
-Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma or carcinoma removal.
-Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
+Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma and carcinoma removal.
+Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after adenoma and carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
 Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.
@@
 We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
-This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.
+This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma or carcinoma.
 From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
@@
 
+The University of Michigan Institutional Review Board approved this study, and all subjects provided informed consent. This study conformed to the guidelines of the Helsinki Declaration.
+
 ### Consent for publication
+
+Not applicable.
 

### 2017-02-27 submission/manuscript_outline_20161024.tex

@@
 our study was to test the hypothesis that changes in the bacterial
-microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
+microbiome occur after lesion (i.e.~adenoma and carcinoma) removal.
 Specifically, we wanted to identify components within the community that
@@
 \textbf{Conclusions:} Our data supports the hypothesis that there are
-changes in the bacterial microbiome following colorectal cancer lesion
-removal. Individuals with carcinoma have more drastic differences to the
-overall community then those with adenoma. Changes to commensal bacteria
-were some of the most important variables for model classification,
+changes in the bacterial microbiome following colorectal lesion removal.
+Individuals with carcinoma have more drastic differences to the overall
+community then those with adenoma. Changes to commensal bacteria were
+some of the most important variables for model classification,
 suggesting that these bacteria may be central to initial polyp formation
@@
 remains limited information on the bacterial microbiome response to
-lesion (adenoma or carcinoma) removal.
+lesion (adenoma and carcinoma) removal.
 
 In this study we tested the hypothesis that the bacterial microbiome
-changes between pre (initial) and post (follow up) removal of lesion
-(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
-as well as the relative abundance of specific Operational Taxonomic
-Units (OTUs). We utilized Random Forest to build two models. The first
-model was built to classify lesion and normal (non-lesion) samples while
-the second was built to classify initial and follow up samples.
-Subsequent observations on how these models, as well as specific OTUs
-within them, performed pre and post lesion removal helped inform us as
-to whether initial and follow up samples were changing and whether it
-was towards a more normal bacterial microbiome or not. We also
-investigated the two models for similar important OTUs to identify the
-crucial OTUs for not only classifying lesion or normal samples but also
-initial and follow up.
+changes between pre (initial) and post (follow up) removal of lesion. We
+analyzed changes in alpha and beta diversity as well as the relative
+abundance of specific Operational Taxonomic Units (OTUs). We utilized
+Random Forest to build two models. The first model was built to classify
+lesion versus non-lesion (normal) samples while the second was built to
+classify initial versus follow up samples. Subsequent observations on
+how these models, as well as specific OTUs within them, performed pre
+and post lesion removal helped inform us as to whether initial and
+follow up samples were changing and whether it was towards a more normal
+bacterial microbiome or not. We also investigated the two models for
+similar important OTUs to identify the crucial OTUs for not only
+classifying lesion or normal samples but also initial versus follow up.
 
@@
 initiation and could change in a way that allows predictive models to
-lower the positive probability of a lesion after surgery {[}Figure 3C \&
+lower the positive probability of a lesion after removal {[}Figure 3C \&
 3D{]}. It should be noted that our study does not argue against the
@@
 also large decreases in relative abundance in some of those with
-carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.
+carcinoma but not in those with adenoma after lesion removal {[}Figure
+2{]}.
 
@@
 knowledge on CRC and the bacterial microbiome: That there is a
-measurable difference in the bacterial community after adenoma or
+measurable difference in the bacterial community after adenoma and
 carcinoma removal. Further, the ability for machine learning algorithms
 to take bacterial microbiome data and successfully lower positive
-probability after either adenoma or carcinoma removal provides evidence
-that there are specific signatures, mostly attributable to commensal
+probability after adenoma and carcinoma removal provides evidence that
+there are specific signatures, mostly attributable to commensal
 organisms, associated with these lesions. Our data provides evidence
@@
 \emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
-stomatis}) based on adenoma and carcinoma. From here the lesion model
-was then tested for accuracy in prediction and whether it reduced the
+stomatis}) based on adenoma or carcinoma. From here the lesion model was
+then tested for accuracy in prediction and whether it reduced the
 positive probability of lesion after surgery. We then used the initial
@@
 
+The University of Michigan Institutional Review Board approved this
+study, and all subjects provided informed consent. This study conformed
+to the guidelines of the Helsinki Declaration.
+
 \subsubsection{Consent for publication}\label{consent-for-publication}
+
+Not applicable.
 

### 2017-02-27 submission/supplemental_outline_20161024.Rmd

@@
 
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
+loadLibs(c("tidyr", "dplyr", "scales", "knitr", "rmarkdown"))
 
@@
 
+lesion_top10_varsMDA <- read.csv('../results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+IF_top10_varsMDA <- read.csv('../results/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+
+# Create selection vector
+common_vars <- common_vars_summary$otu
+
+# Create MDA vectors
+lesion_MDA <- filter(lesion_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
+
+initial_MDA <- filter(IF_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
+
 st2 <- select(common_vars_summary, otu) %>% 
   mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
-  mutate(Pvalue = common_vars_summary$Pvalue) %>% 
-  mutate(BH_corr = common_vars_summary$BH_corr)
+  mutate(lesion_MDA = lesion_MDA$mean_MDA) %>% 
+  mutate(initial_MDA = initial_MDA$mean_MDA) %>% arrange(desc(lesion_MDA))
 
-colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "P-value", "BH Corrected P-value")
+colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Lesion Model MDA", "Initial Model MDA")
 

### 2017-03-06 submission/manuscript_outline_20161024.tex
re render after FIT removal

@@
 post lesion removal (P-value \textgreater{} 0.05). A model built to
-classify lesion had an AUC range of 0.811 - 0.866 and follow up samples
+classify lesion had an AUC range of 0.692 - 0.761 and follow up samples
 had a decrease in the positive probability of lesion (P-value
-\textless{} 0.05) suggesting a movement towards a more normal bacterial
-community. A second model built to classify initial samples had an AUC
-range of 0.641 - 0.805 and had a decrease in positive probability for
-the follow up samples to be an initial sample (P-value \textless{}
-0.05). The lesion model used a total of 53 variables while the initial
-sample model used a total of 70 variables. A total of 23 OTUs were
-common to both models with the majority of these classifying to
-commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
-\emph{Blautia}, and \emph{Ruminococcaceae}).
+\textgreater{} 0.05) suggesting a movement towards a more normal
+bacterial community. A second model built to classify initial samples
+had an AUC range of 0.657 - 0.796 and had a decrease in positive
+probability for the follow up samples to be an initial sample (P-value
+\textless{} 0.05). The lesion model used a total of 54 variables while
+the initial sample model used a total of 70 variables. A total of 32
+OTUs were common to both models with the majority of these classifying
+to commensal bacteria (e.g. \emph{Lachnospiraceae},
+\emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}).
 
@@
 model to classify samples as lesion versus normal based on the bacterial
-community and FIT measurements. This model had an AUC range of 0.811 -
-0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
+community and FIT measurements. This model had an AUC range of 0.692 -
+0.761 after 100 iterations of 20 repeated 10-fold cross validations. The
 ROC curve for the final lesion model used was within the observed range
 of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
-a total of 53 variables that were used in this model {[}Figure 3B{]}.
+a total of 54 variables that were used in this model {[}Figure 3B{]}.
 The FIT measurement for fecal blood resulted in the largest decrease in
@@
 probability of the follow up sample to be a lesion. This is what we
-observed for the lesion model (P-value \textless{} 0.001). When we
+observed for the lesion model (P-value \textgreater{} 0.001). When we
 separated individuals based on whether they had an adenoma or carcinoma
 there was only a decrease in positive probability for the carcinoma
-group {[}Figure 3C{]} (P-value \textless{} 0.001) and not for the
+group {[}Figure 3C{]} (P-value \textgreater{} 0.001) and not for the
 adenoma group {[}Figure 3D{]} (P-value \textgreater{} 0.05). We also
 observed that there were no significant differences between the
-predicted and actual calls (P-value \textgreater{} 0.05). The lesion
-model was also able to correctly classify the one individual who still
-had a carcinoma on follow up {[}Figure 3C{]}.
+predicted and actual calls (P-value \textless{} 0.05). The lesion model
+was also able to correctly classify the one individual who still had a
+carcinoma on follow up {[}Figure 3C{]}.
 
@@
 removed) samples based on the bacterial community and FIT measurements.
-The initial sample model had an AUC range of 0.641 to 0.805 after 100
+The initial sample model had an AUC range of 0.657 to 0.796 after 100
 iterations of 20 repeated 10-fold cross validations. The test set AUC
@@
 a total of 70 variables that were used for this model {[}Figure 4B{]}.
-The variable that resulted in the largest MDA was \emph{Pseudomonas}
-(Otu000438) while FIT measurement for fecal blood resulted in the sixth
+The variable that resulted in the largest MDA was \emph{Ruminococcaceae}
+(Otu000278) while FIT measurement for fecal blood resulted in the sixth
 largest decrease in MDA {[}Figure 4B{]}.
@@
 of lesion. When we compared the two different models with each other
-there were a total of 23 common OTUs. Some of the most common taxonomic
-identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
-\emph{Blautia}, and \emph{Ruminococcaceae}. The vast majority of these
-OTUs had classifications to bacteria typically thought of as commensal
-{[}Table S2{]}.
+there were a total of 32 common OTUs. Some of the most common taxonomic
+identifications belonged to \emph{Lachnospiraceae},
+\emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}. The
+vast majority of these OTUs had classifications to bacteria typically
+thought of as commensal {[}Table S2{]}.
 
@@
 commensal bacteria. Providing additional information on the importance
-of commensal bacteria was that there were a total of 23 OTUs in common
+of commensal bacteria was that there were a total of 32 OTUs in common
 to both models and the vast majority belonged to regular residents of

### 2017-03-06 submission/manuscript_outline_20161024.Rmd
abstract edits to more active voice

@@
 ---
-title: ""
-bibliography: reference.bib
+title: ''
+csl: microbiome.csl
+fontsize: 12pt
 output:
+  word_document: default
   pdf_document:
-    keep_tex: true
     includes:
       in_header: header.tex
-csl: microbiome.csl
-fontsize: 12pt
+    keep_tex: yes
 geometry: margin=1.0in
+bibliography: reference.bib
 ---
@@
 
-**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.
-The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e.
-adenoma and carcinoma) removal.
-Specifically, we wanted to identify components within the community that were different before and after removal of said lesion.
+**Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.
+Our objective was to test the hypothesis that changes in the bacterial community occur after treatment for adenoma or carcinoma.
+Specifically, we wanted to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).
 
-**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
-There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.
-A second model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+**Results:** There was a larger change in the bacterial community in response to lesion removal for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+Yet no difference was found in the relative abundance of any OTU before and after lesion removal for adenoma and carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
+A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples decreased in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting a movement towards a normal bacterial community.
+An initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
@@
 
-**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal lesion removal.
+**Conclusions:** Our data supports the hypothesis that the bacterial community changes after treatment.
 Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.
-Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.
+Commensal bacteria were crucial for accurate model classification, suggesting that these bacteria may be important to initial polyp formation and transition to carcinoma.
 
@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; post surgery; risk factors
+bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factors
 

### 2017-03-06 submission/manuscript_outline_20161024.Rmd
Background section changes to active voice

@@
 
-Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
-Over the last few years the rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
+Colorectal cancer (CRC) is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
+The rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
 However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
 
-Over the last few years studies have shown how the bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.
-There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].
-These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.
-Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma and carcinoma) removal.
+Recent studies in humans have shown that both the bacterial community and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].
+Further, Dejea, et al.
+observed that bacterial communities are altered between normal and tumor tissue [@dejea_microbiota_2014].
+Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.
+Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.
+Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
+Although these studies show how our gut bacteria can impact CRC progression via a changed community or invations by more inflammatory bacteria [@flynn_metabolic_2016].
+They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).
 
-In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion.
+In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.
 We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
-We utilized Random Forest to build two models.
-The first model was built to classify lesion versus non-lesion (normal) samples while the second was built to classify initial versus follow up samples.
-Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.
-We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial versus follow up.
+We utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.
+Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.
+We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.
+This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
 

### 2017-03-06 submission/manuscript_outline_20161024.Rmd
small text change

@@
 **Results:** There was a larger change in the bacterial community in response to lesion removal for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
-Yet no difference was found in the relative abundance of any OTU before and after lesion removal for adenoma and carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
+Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma and carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
 A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples decreased in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting a movement towards a normal bacterial community.

### 2017-03-06 submission/manuscript_outline_20161024.Rmd

@@
 **Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.
-Our objective was to test the hypothesis that changes in the bacterial community occur after treatment for adenoma or carcinoma.
-Specifically, we wanted to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).
+This study tested the hypothesis that changes in the bacterial community occur after treatment for adenoma or carcinoma.
+Specifically, we tried to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).
 
 **Results:** There was a larger change in the bacterial community in response to lesion removal for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
-Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma and carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
+Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
 A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples decreased in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting a movement towards a normal bacterial community.

### 2017-03-07 submission/manuscript_outline_20161024.Rmd
make sure new paths all work

@@
 output:
-  word_document: default
   pdf_document:
@@
 # Extra data and meta data needed
-good_metaf <- read.csv('../results/tables/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
-metaI <- read.csv('../results/tables/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
+good_metaf <- read.csv('../data/process/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
+metaI <- read.csv('../data/process/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
 total_sub_Seqs <- read.delim('../data/process/final.0.03.subsample.shared', header = T) %>% 
@@
 # initial microbiome and FIT description data 
-change_summary <- read.csv('../results/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
-alpha_summary <- read.csv('../results/tables/alpha_table_summary.csv', header = T, row.names = 1)
-beta_summary <- read.csv('../results/tables/beta_diver_summary.csv', header = T, row.names = 1)
-time_pvalues <- read.csv("../results/tables/time_pvalues.csv", header = T, row.names = 1)
-time_summary <- read.csv("../results/tables/time_summary_data.csv", header = T, row.names = 1)
-all_OTU_pair_wilcox <- read.csv("../results/tables/OTU_paired_wilcoxson_test.csv", header = T)
+change_summary <- read.csv('../data/process/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
+alpha_summary <- read.csv('../data/process/tables/alpha_table_summary.csv', header = T, row.names = 1)
+beta_summary <- read.csv('../data/process/tables/beta_diver_summary.csv', header = T, row.names = 1)
+time_pvalues <- read.csv("../data/process/tables/time_pvalues.csv", header = T, row.names = 1)
+time_summary <- read.csv("../data/process/tables/time_summary_data.csv", header = T, row.names = 1)
+all_OTU_pair_wilcox <- read.csv("../data/process/tables/OTU_paired_wilcoxson_test.csv", header = T)
 
 # Main CRC bacteria differences between Adenoma and Carcinoma
-main_crc_bugs_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
-crc_bug_info <- read.csv("../results/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)
+main_crc_bugs_summary <- read.csv('../data/process/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
+crc_bug_info <- read.csv("../data/process/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)
 
 # Lesion Model 
-lesion_top10_varsMDA <- read.csv('../results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T)
-lesion_imp_vars <- read.csv('../results/tables/rf_otu_tax.csv', header = T, row.names = 1)
-lesion_AUC_run_summary <- read.csv('../results/tables/Reduced_ROC_model_summary.csv', header = T)
+lesion_top10_varsMDA <- read.csv('../data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T)
+lesion_imp_vars <- read.csv('../data/process/tables/rf_otu_tax.csv', header = T, row.names = 1)
+lesion_AUC_run_summary <- read.csv('../data/process/tables/Reduced_ROC_model_summary.csv', header = T)
 
 # IF Model
-IF_top10_varsMDA <- read.csv('../results/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T)
-IF_imp_vars <- read.csv('../results/tables/if_rf_otu_tax.csv', header = T, row.names = 1)
-IF_AUC_run_summary <- read.csv('../results/tables/reduced_IF_ROC_model_summary.csv', header = T)
+IF_top10_varsMDA <- read.csv('../data/process/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T)
+IF_imp_vars <- read.csv('../data/process/tables/if_rf_otu_tax.csv', header = T, row.names = 1)
+IF_AUC_run_summary <- read.csv('../data/process/tables/reduced_IF_ROC_model_summary.csv', header = T)
 
 # All model information
-all_probs_paired_wilcox <- read.csv('../results/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
-all_pred_stat_summary <- read.csv('../results/tables/all_models_summary_info.csv', header = T, stringsAsFactors = F)
-all_confu_matrix <- read.csv('../results/tables/all_models_confusion_summary.csv', header = T, stringsAsFactors = F)
+all_probs_paired_wilcox <- read.csv('../data/process/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
+all_pred_stat_summary <- read.csv('../data/process/tables/all_models_summary_info.csv', header = T, stringsAsFactors = F)
+all_confu_matrix <- read.csv('../data/process/tables/all_models_confusion_summary.csv', header = T, stringsAsFactors = F)
 
 # Common Variable Info
-common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
+common_vars_summary <- read.csv('../data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
 
 #Therapy Info
-chemo_rad_summary <- read.csv('../results/tables/probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
+chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
 

### 2017-03-07 submission/manuscript_outline_20161024.tex

@@
 
-\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
-health problem with previous research suggesting that a link may exist
-between the fecal bacterial microbiome and CRC. The overall objective of
-our study was to test the hypothesis that changes in the bacterial
-microbiome occur after lesion (i.e.~adenoma and carcinoma) removal.
-Specifically, we wanted to identify components within the community that
-were different before and after removal of said lesion.
+\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
+problem and research suggests a correlation between the fecal bacterial
+microbiome and CRC. This study tested the hypothesis that changes in the
+bacterial community occur after treatment for adenoma or carcinoma.
+Specifically, we tried to identify components within the community that
+were different before and after removal of lesion (adenoma or
+carcinoma).
 
-\textbf{Results:} The bacterial microbiome changed more in response to
-lesion removal in carcinoma cases compared to adenoma cases (P-value
-\textless{} 0.05). There was no difference for either the adenoma or
-carcinoma group in the relative abundance of any OTU between pre and
-post lesion removal (P-value \textgreater{} 0.05). A model built to
-classify lesion had an AUC range of 0.692 - 0.761 and follow up samples
-had a decrease in the positive probability of lesion (P-value
-\textgreater{} 0.05) suggesting a movement towards a more normal
-bacterial community. A second model built to classify initial samples
-had an AUC range of 0.657 - 0.796 and had a decrease in positive
-probability for the follow up samples to be an initial sample (P-value
-\textless{} 0.05). The lesion model used a total of 54 variables while
-the initial sample model used a total of 70 variables. A total of 32
-OTUs were common to both models with the majority of these classifying
-to commensal bacteria (e.g. \emph{Lachnospiraceae},
+\textbf{Results:} There was a larger change in the bacterial community
+in response to lesion removal for carcinoma versus adenoma cases
+(P-value \textless{} 0.05). Yet no difference was found in the relative
+abundance of any OTU before and after treatment for adenoma or carcinoma
+groups (P-value \textgreater{} 0.05). A lesion model had an AUC range of
+0.692 - 0.761 and follow up samples decreased in the positive
+probability of lesion versus initial samples (P-value \textgreater{}
+0.05); suggesting a movement towards a normal bacterial community. An
+initial sample model had an AUC range of 0.657 - 0.796 and had a
+decrease positive probability for the follow up samples to be an initial
+sample (P-value \textless{} 0.05). The lesion model used a total of 54
+variables while the initial sample model used a total of 70 variables. A
+total of 32 OTUs were common to both models with the majority of these
+classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
 \emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}).
 
-\textbf{Conclusions:} Our data supports the hypothesis that there are
-changes in the bacterial microbiome following colorectal lesion removal.
-Individuals with carcinoma have more drastic differences to the overall
-community then those with adenoma. Changes to commensal bacteria were
-some of the most important variables for model classification,
-suggesting that these bacteria may be central to initial polyp formation
-and transition to carcinoma.
+\textbf{Conclusions:} Our data supports the hypothesis that the
+bacterial community changes after treatment. Individuals with carcinoma
+have more drastic differences to the overall community then those with
+adenoma. Commensal bacteria were crucial for accurate model
+classification, suggesting that these bacteria may be important to
+initial polyp formation and transition to carcinoma.
 
@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; post surgery; risk
+bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk
 factors
@@
 
-Colorectal cancer (CRC) continues to be a leading cause of cancer
-related deaths and is currently the third most common cause of cancer
-deaths {[}1,2{]}. Over the last few years the rate of disease mortality
-has seen a significant decrease, thanks mainly to improvements in
-screening {[}1{]}. However, despite this improvement there are still
-approximately 50,000 deaths from the disease per year {[}2{]}.
+Colorectal cancer (CRC) is currently the third most common cause of
+cancer deaths {[}1,2{]}. The rate of disease mortality has seen a
+significant decrease, thanks mainly to improvements in screening
+{[}1{]}. However, despite this improvement there are still approximately
+50,000 deaths from the disease per year {[}2{]}.
 
-Over the last few years studies have shown how the bacterial microbiome
-{[}3{]} or specific members within it {[}4{]} could be directly involved
-with the pathogenesis of CRC. There has also been research into how the
-bacterial microbiome could be altered directly on tumor tissue itself
-{[}5{]}. These studies have helped to provide a tantalizing link between
-the bacterial microbiome and CRC. Building off of these findings, there
-has been promising work on the bacterial microbiome and its ability to
-complement existing screening methods such as Fecal Immunoglobulin Test
-(FIT) or act alone as a screening tool {[}6,7{]}. Although these studies
-suggest that the bacterial microbiome might change after treatment there
-remains limited information on the bacterial microbiome response to
-lesion (adenoma and carcinoma) removal.
+Recent studies in humans have shown that both the bacterial community
+and specific members within it correlate with CRC pathogenesis
+{[}3,4{]}. Further, Dejea, et al. observed that bacterial communities
+are altered between normal and tumor tissue {[}5{]}. Mouse models of CRC
+have further demonstrated the importance of the microbiome, both on a
+community {[}3,6{]} and species level {[}4{]}, for tumorgenesis.
+Collectively, these studies provide a tantalizing link between our gut
+bacteria and CRC and suggest that biomarkers using our microbes could be
+developed. Indeed, builidng models using 16S rRNA gene sequencing along
+with clinical tests such as Fecal Immunoglobulin Test (FIT) result in
+good predictions of CRC {[}7,8{]}. Although these studies show how our
+gut bacteria can impact CRC progression via a changed community or
+invations by more inflammatory bacteria {[}9{]}. They provide very
+little information as to whether these changed communities rebound after
+successful treatment of lesion (adenoma or carcinoma).
 
-In this study we tested the hypothesis that the bacterial microbiome
-changes between pre (initial) and post (follow up) removal of lesion. We
-analyzed changes in alpha and beta diversity as well as the relative
-abundance of specific Operational Taxonomic Units (OTUs). We utilized
-Random Forest to build two models. The first model was built to classify
-lesion versus non-lesion (normal) samples while the second was built to
-classify initial versus follow up samples. Subsequent observations on
-how these models, as well as specific OTUs within them, performed pre
-and post lesion removal helped inform us as to whether initial and
-follow up samples were changing and whether it was towards a more normal
-bacterial microbiome or not. We also investigated the two models for
-similar important OTUs to identify the crucial OTUs for not only
-classifying lesion or normal samples but also initial versus follow up.
+In this study we tested the hypothesis that there are detectable changes
+to the bacterial community between pre- (initial) and post- (follow up)
+treatment of lesion. We analyzed changes in alpha and beta diversity as
+well as the relative abundance of specific Operational Taxonomic Units
+(OTUs). We utilized Random Forest to build two models: The first was
+built to classify lesion versus non-lesion (normal) while the second was
+built to classify initial versus follow up samples. Subsequent
+observations on how these models and OTUs within them performed before
+and after treatment helped inform us as to whether initial and follow up
+samples were changing and whether it was towards a normal community. We
+also investigated the two models for similar important OTUs to identify
+the crucial OTUs involved with both classifying lesion or normal and
+initial versus follow up samples. This study will provide evidence as to
+whether treatment can influence the community and if the CRC microbiome,
+identified in previous studies, persists.
 
@@
 suggested that a number of oral microbes may be important in CRC
-pathogenesis {[}8{]}. So we next examined whether there were changes in
+pathogenesis {[}9{]}. So we next examined whether there were changes in
 previously well described carcinoma associated OTUs, such as
@@
 into the bacterial microbiome as a potential screening tool
-{[}6,7,9--11{]} by exploring what happens to the bacterial community
-after lesion removal. Based on previous work by Arthur, et al. {[}12{]}
+{[}7,8,10--12{]} by exploring what happens to the bacterial community
+after lesion removal. Based on previous work by Arthur, et al. {[}13{]}
 it may not be surprising to have E.coli as one of the most important
@@
 These bacterial microbiome changes could be the first step in allowing
-more inflammatory bacterium to gain a foothold within the colon {[}8{]}.
+more inflammatory bacterium to gain a foothold within the colon {[}9{]}.
 
@@
 thought to be protective against cancer in part by reducing inflammation
-{[}13{]}. These protective compounds are derived from the breakdown of
+{[}14{]}. These protective compounds are derived from the breakdown of
 fiber, fruits, and vegetables by resident gut microbes. One example of
@@
 to be increased in those with high non-meat based protein consumption
-{[}14{]}. High protein consumption in general has been linked with an
-increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
+{[}15{]}. High protein consumption in general has been linked with an
+increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
 inhibited by polyphenols which are derived from fruits and vegetables
-{[}16{]}. Our data fits with the hypothesis that the microbial
+{[}17{]}. Our data fits with the hypothesis that the microbial
 metabolites from breakdown products within our own diet could not only
@@
 longer having a lesion. Another limitation is that we do not know if
-adding modern tests such as the stool DNA test {[}17{]} could help
+adding modern tests such as the stool DNA test {[}18{]} could help
 improve our overall AUC. This study also drew heavily from those with
@@
 \textbf{\emph{Study Design and Patient Sampling:}} The sampling and
-design were similar to that reported in Baxter, et al {[}6{]}. In brief,
+design were similar to that reported in Baxter, et al {[}7{]}. In brief,
 study exclusion involved those who had already undergone surgery,
@@
 microbiome has been shown to normalize back to a pre-colonscopy
-community within this time period {[}18{]}. Our follow up data set had a
+community within this time period {[}19{]}. Our follow up data set had a
 total of 67 individuals that not only had a sample as described but also
@@
 previously published using both OC FIT-CHEK and OC-Auto Micro 80
-automated system (Polymedco Inc.) {[}19{]}. 16S rRNA gene sequencing was
-completed as previously described by Kozich, et al. {[}20{]}. DNA
+automated system (Polymedco Inc.) {[}20{]}. 16S rRNA gene sequencing was
+completed as previously described by Kozich, et al. {[}21{]}. DNA
 extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described {[}20{]}. The general processing workflow
+has been previously described {[}21{]}. The general processing workflow
 using mothur was as follows: Paired-end reads were first merged into
@@
 
-\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
+\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}22{]}
 algorithm was used to create the model used for prediction of lesion
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}22{]} in the vegan package
+community structure utilized PERMANOVA {[}23{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}23{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
+\hypertarget{ref-zackular_gut_2013}{}
+6. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen GY,
+et al. The gut microbiome modulates colon tumorigenesis. mBio.
+2013;4:e00692--00613.
+
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-6. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+7. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-zeller_potential_2014}{}
-7. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+8. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 \hypertarget{ref-flynn_metabolic_2016}{}
-8. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+9. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
 Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
@@
 \hypertarget{ref-yu_metagenomic_2017}{}
-9. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
+10. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
 analysis of faecal microbiome as a tool towards targeted non-invasive
@@
 \hypertarget{ref-zackular_human_2014}{}
-10. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+11. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-warren_co-occurrence_2013}{}
-11. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
+12. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
 et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
@@
 \hypertarget{ref-arthur_intestinal_2012}{}
-12. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+13. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
 T-J, et al. Intestinal inflammation targets cancer-inducing activity of
@@
 \hypertarget{ref-louis_gut_2014}{}
-13. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
+14. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
 metabolites and colorectal cancer. Nature Reviews Microbiology
@@
 \hypertarget{ref-zhu_intake_2016}{}
-14. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
+15. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
 Proteins Substantially Increased the Relative Abundance of Genus
@@
 \hypertarget{ref-mu_colonic_2016}{}
-15. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
+16. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
 Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
@@
 \hypertarget{ref-ozdal_reciprocal_2016}{}
-16. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
+17. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
 Reciprocal Interactions between Polyphenols and Gut Microbiota and
@@
 \hypertarget{ref-cotter_long-term_2016}{}
-17. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
+18. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
 et al. Long-Term Follow-up of Patients Having False Positive
@@
 \hypertarget{ref-obrien_impact_2013}{}
-18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+19. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-daly_evaluation_2013}{}
-19. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
+20. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
 for colorectal cancer screening. Journal of Primary Care \& Community
@@
 \hypertarget{ref-kozich_development_2013}{}
-20. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+21. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-21. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+22. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-22. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+23. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-23. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-03-07 submission/supplemental_outline_20161024.Rmd

@@
 
-alpha_table <- read.csv("../results/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
+alpha_table <- read.csv("../data/process/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
 supplemental_table1 <- separate(alpha_table, X, c("comparison", "alpha_metric"), sep="_")
@@
 
-common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
+common_vars_summary <- read.csv('../data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
 
-lesion_top10_varsMDA <- read.csv('../results/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
-IF_top10_varsMDA <- read.csv('../results/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+lesion_top10_varsMDA <- read.csv('../data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+IF_top10_varsMDA <- read.csv('../data/process/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
 
@@
 
-chemo_rad_summary <- read.csv('../results/tables/probs_chemo_rad_pvalue_summary.csv', 
+chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv', 
                               header = T, stringsAsFactors = F)

### 2017-03-07 code/Run_Figure1.R
remove FIT from figure

@@ -37,41 +37,20 @@ differences_graph <- grid.arrange(
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
           plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "lines"), 
           legend.title = element_text(face="bold"), 
-          legend.position = c(0.75, 0.2), 
+          legend.position = c(0.75, 0.20), 
           title = element_text(face="bold", hjust = 0), 
           panel.grid.major = element_blank(), 
-          panel.grid.minor = element_blank()), 
-  
-  # Difference from fit between adenoma and cancer
-  ggplot(difference_table_treatment, 
-         aes(factor(dx, levels = c("adenoma", "cancer"), labels = c("Adenoma", "Carcinoma")), 
-             fit_difference*-1, group = 1)) + 
-    geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
-    stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
-    stat_summary(fun.y = mean, colour = "black", geom = "line") + 
-    scale_color_manual(name = "Lesion Type", 
-                       values = c('#00FFFF', '#0000FF', '#F1BB7B'), 
-                       breaks = c("adenoma", "adv_adenoma", "cancer"), 
-                       labels = c("Adenoma", "SRN", "Carcinoma")) + 
-    ylab("Change in FIT from Follow Up to Initial") + 
-    xlab("") + theme_bw() + ggtitle("B") + 
-    theme(axis.title = element_text(face="bold", hjust = 0.5), 
-          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "lines"), 
-          legend.title = element_text(face="bold"), 
-          legend.position = "none", 
-          title = element_text(face="bold", hjust = 0), 
-          panel.grid.major = element_blank(), 
-          panel.grid.minor = element_blank()), 
+          panel.grid.minor = element_blank()) + coord_fixed(ratio = 1.8), 
   
   # Adenoma Only initial and follow up
   mutate(thetayc_adn_IF, samples = factor(samples, levels = c("follow_up", "initial"), labels = c("Follow Up", "Initial"))) %>% 
   ggplot(aes(x=NMDS1, y=NMDS2)) + geom_point(aes(color=samples)) + 
[…]

### 2017-03-07 submission/manuscript_outline_20161024.Rmd
modify abstract
additional background changes

@@
 **Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.
-This study tested the hypothesis that changes in the bacterial community occur after treatment for adenoma or carcinoma.
+This study tested the hypothesis that treatment for adenoma or carcinoma result in changes to the bacterial community.
 Specifically, we tried to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).
@@
 We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
-We utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.
+We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].
+We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.
 Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.
@@
 ***Carcinoma Associated Bacteria:***
-Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].
 So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).

### 2017-03-08 code/Run_Test_Chemo_Rad.R
add effect sizes to chemo/rads comparison table

@@ -17,48 +17,13 @@ data_list <- list(
                           header = T, stringsAsFactors = F)
 )
 
-
-# Test for differences in positive probability changes
-pvalue_table <- as.data.frame(matrix(nrow = 12, ncol = 2, dimnames = list(
-  rown = c("chemo_red_les", "chemo_red_IF", "rad_red_les", "rad_red_IF", "chemo_b", "rad_b", 
-           "chemo_sobs", "chemo_shannon", "chemo_shannoneven", 
-           "rads_sobs", "rads_shannon", "rads_shannoneven"), coln = c("pvalue", "bh"))))
-
-for(i in 1:length(data_list)){
-  
-  pvalue_table[i, "pvalue"] <- wilcox.test(
-    filter(data_list[[i]], sampleType == "initial", chemo == "yes")[, "Yes"] - 
-                filter(data_list[[i]], sampleType == "followup", chemo == "yes")[, "Yes"], 
-              filter(data_list[[i]], sampleType == "initial", chemo == "no")[, "Yes"] - 
-                filter(data_list[[i]], sampleType == "followup", chemo == "no")[, "Yes"])$p.value
-  
-  pvalue_table[i+2, "pvalue"] <- wilcox.test(
-    filter(data_list[[i]], sampleType == "initial", rads == "yes")[, "Yes"] - 
-                filter(data_list[[i]], sampleType == "followup", rads == "yes")[, "Yes"], 
-              filter(data_list[[i]], sampleType == "initial", rads == "no")[, "Yes"] - 
-                filter(data_list[[i]], sampleType == "followup", rads == "no")[, "Yes"])$p.value
-}
-
-
-# Test thetayc value differences between treatment type
-
 difference_table_treatment <- read.csv("data/process/tables/difference_table.csv", 
                                        header = T, stringsAsFactors = F)
 
-pvalue_table["chemo_b", "pvalue"] <- wilcox.test(
-  filter(difference_table_treatment, chemo == "yes")[, "distance"], 
-  filter(difference_table_treatment, chemo == "no")[, "distance"])$p.value
-
-pvalue_table["rad_b", "pvalue"] <- wilcox.test(
-  filter(difference_table_treatment, rads == "yes")[, "distance"], 
-  filter(difference_table_treatment, rads == "no")[, "distance"])$p.value
-
-
[…]

### 2017-03-08 submission/manuscript_outline_20161024.Rmd
update to include effect sizes of chemo/rads treatment

@@
 We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.
+Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionaly treatments significantly lowered any of the metrics examined versus those who received only lesion removal.
 This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
@@
 
-***Carcinoma Associated Bacteria:***
+***Carcinoma Associated Bacteria:*** (Anna recommends moving this to the supplement)
 So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
@@
 
-***Treatment Differences:***
+***Treatment Differences:*** (after meeting with Anna an idea I have is that it might be a good idea to tie all the findings together and relate it back to treatment)
 After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.

### 2017-03-12 submission/manuscript_outline_20161024.tex

@@
 \usepackage{comment}
+\usepackage{lscape}
 
@@
 problem and research suggests a correlation between the fecal bacterial
-microbiome and CRC. This study tested the hypothesis that changes in the
-bacterial community occur after treatment for adenoma or carcinoma.
+microbiome and CRC. This study tested the hypothesis that treatment for
+adenoma or carcinoma result in changes to the bacterial community.
 Specifically, we tried to identify components within the community that
@@
 well as the relative abundance of specific Operational Taxonomic Units
-(OTUs). We utilized Random Forest to build two models: The first was
-built to classify lesion versus non-lesion (normal) while the second was
-built to classify initial versus follow up samples. Subsequent
-observations on how these models and OTUs within them performed before
-and after treatment helped inform us as to whether initial and follow up
-samples were changing and whether it was towards a normal community. We
-also investigated the two models for similar important OTUs to identify
-the crucial OTUs involved with both classifying lesion or normal and
-initial versus follow up samples. This study will provide evidence as to
-whether treatment can influence the community and if the CRC microbiome,
-identified in previous studies, persists.
+(OTUs). We then investigated several important genera from oral microbes
+that have been suggested to be important in CRC pathogenesis {[}9{]}. We
+next utilized Random Forest to build two models: The first was built to
+classify lesion versus non-lesion (normal) while the second was built to
+classify initial versus follow up samples. Subsequent observations on
+how these models and OTUs within them performed before and after
+treatment helped inform us as to whether initial and follow up samples
+were changing and whether it was towards a normal community. We also
+investigated the two models for similar important OTUs to identify the
+crucial OTUs involved with both classifying lesion or normal and initial
+versus follow up samples. Finally, since treatment also included
+chemotherapy and radiation therapy along with lesion removal, we tested
+if these additionaly treatments significantly lowered any of the metrics
+examined versus those who received only lesion removal. This study will
+provide evidence as to whether treatment can influence the community and
+if the CRC microbiome, identified in previous studies, persists.
 
@@
 
-\textbf{\emph{Bacterial Community and FIT:}} Within our 67 person cohort
-we first wanted to test whether there were any broad differences between
-initial and follow up samples based on lesion being either adenoma (n =
-41) or carcinoma (n = 26). We found that the bacterial community in
-those with carcinoma were more dissimilar to their initial sample than
-those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. We also
-found that there were larger changes in fecal blood (measured by FIT)
-for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
-{[}Figure 1B{]}. The bacterial community structure before and after
-surgery was visualized using NMDS for both adenoma {[}Figure 1C{]}
-(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1D{]} (PERMANOVA
-\textless{} 0.05). Interestingly, when initial and follow up samples
-were compared, regardless of whether the lesions were adenoma or
-carcinoma, there was no significant overall difference in beta diversity
-(PERMANOVA \textgreater{} 0.05). When investigating alpha diversity
-metrics there was no difference found between initial and follow up
-samples for lesion, adenoma only, or carcinoma only for any metric
-tested {[}Table S1{]}. We also observed that there was no difference in
-the relative abundance of any OTU between initial and follow up samples
-for lesion, adenoma only, or carcinoma only {[}Figure S1{]}.
+\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
+tested whether those with adenoma (n = 41) or carcinoma (n = 26) had any
+broad differences between their initial and follow up samples. We found
+that those with carcinoma had a more dissimilar bacterial community
+between their initial and follow up sample than those with adenoma
+(P-value \textless{} 0.001) {[}Figure 1A{]}. The bacterial community
+structure before and after surgery was visualized using NMDS for both
+adenoma {[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05) and carcinoma
+{[}Figure 1C{]} (PERMANOVA \textless{} 0.05). Interestingly, when
+initial and follow up samples were compared, regardless of whether the
+lesions were adenoma or carcinoma, there was no significant overall
+difference in beta diversity (PERMANOVA \textgreater{} 0.05). There was
+no difference between initial and follow up samples when investigating
+alpha diversity metrics for lesion, adenoma only, or carcinoma only for
+any metric tested {[}Table S1{]}. Additionally, there was also no
+difference in the relative abundance of any OTU between initial and
+follow up samples for lesion, adenoma only, or carcinoma only {[}Figure
+S1{]}.
 
-\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
-suggested that a number of oral microbes may be important in CRC
-pathogenesis {[}9{]}. So we next examined whether there were changes in
-previously well described carcinoma associated OTUs, such as
-\emph{Porphyromonas asaccharolytica} (Otu000202), \emph{Fusobacterium
-nucleatum} (Otu000442), \emph{Parvimonas micra} (Otu001273), and
-\emph{Peptostreptococcus stomatis} (Otu001682). We first observed that
-only a small percentage of those with adenoma or carcinoma were positive
-or had a relative abundance above 0.1\% for any of these respective OTUs
-{[}Figure 2{]}. Despite this, those with carcinoma had a decrease in
-relative abundance from initial to follow up for \emph{Parvimonas micra}
-(P-value \textless{} 0.05) and \emph{Porphyromonas asaccharolytica}
-(P-value \textless{} 0.05) {[}Figure 2A{]}. In contrast, there was no
-difference in relative abundance in any of these OTUs when considering
-only those with adenoma {[}Figure 2B{]}.
+\textbf{\emph{Carcinoma Associated Genera:}} For carcinoma {[}Figure
+S2A{]} (P-value \textgreater{} 0.05) but not adenoma {[}Figure S2B{]}
+(P-value \textgreater{} 0.05) there was a significant decrease in
+\emph{Porphyromonas}, \emph{Parvimonas}, and \emph{Peptostreptococcus}
+between initial and follow up samples {[}Figure S2{]}.
+\emph{Fusobacterium} for both adenoma and carcinoma had no significant
+differences between initial and follow up samples {[}FIgure S2{]}
+(P-value \textgreater{} 0.05). Although there were significant
+differences, only a small percentage of those with adenoma or carcinoma
+were positive for any of these specific genera {[}Figure S2{]}.
 
-\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
-were any common bacterial microbiome changes in individuals with adenoma
-and carcinoma versus normal controls. We investigated this by creating a
-model to classify samples as lesion versus normal based on the bacterial
-community and FIT measurements. This model had an AUC range of 0.692 -
-0.761 after 100 iterations of 20 repeated 10-fold cross validations. The
+\textbf{\emph{The Lesion Model:}} We tested whether follow up samples
+were more normal than initial samples by building a model that
+classified lesion versus normal. This model had an AUC range of 0.692 -
+0.761 after 100 iterations of 20 repeated 10-fold corss validations. The
 ROC curve for the final lesion model used was within the observed range
-of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
-a total of 54 variables that were used in this model {[}Figure 3B{]}.
-The FIT measurement for fecal blood resulted in the largest decrease in
-MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
-(Otu000015) {[}Figure 3B{]}.
+of the 100-different test set AUC iterations {[}Figure 2A{]}. There was
+a total of 54 OTUs that were used in this model {[}Figure 2B{]}.
+\emph{Lachnospiraceae} (Otu000015) resulted in the largest decrease in
+MDA {[}FIgure 2B{]}. Heavily represented genera within the model
+included OTUs from \emph{Lachnospiraceae}, \emph{Bacteroides},
+\emph{Roseburia}, \emph{Blautia}, and \emph{Ruminococcus}.
 
-If there were common OTUs that could separate adenoma and carcinoma from
-normal controls, we would expect to find a decrease in the positive
-probability of the follow up sample to be a lesion. This is what we
-observed for the lesion model (P-value \textgreater{} 0.001). When we
-separated individuals based on whether they had an adenoma or carcinoma
-there was only a decrease in positive probability for the carcinoma
-group {[}Figure 3C{]} (P-value \textgreater{} 0.001) and not for the
-adenoma group {[}Figure 3D{]} (P-value \textgreater{} 0.05). We also
-observed that there were no significant differences between the
-predicted and actual calls (P-value \textless{} 0.05). The lesion model
-was also able to correctly classify the one individual who still had a
-carcinoma on follow up {[}Figure 3C{]}.
+If there was movement towards a more normal community after treatment of
+lesion, we would expect to find a decrease in the positive probability
+of the follow up sample to be a lesion. We observed no such decrease in
+positive probability for either those with adenoma or carcinoma
+{[}FIgure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
+significant difference between the predicted and actual calls (P-value
+\textless{} 0.05). This model correctly classified the one individual
+who still had carcinoma on follow up and this individual's positive
+probability of lesion increased between their initial and follow up
+sample {[}Figure 2C{]}.
 
-\textbf{\emph{The Initial Sample Model:}} After building a model to
-classify based on lesion we built a separate model specifically to be
-able to classify whether samples were initial (before lesion was
-removed) samples based on the bacterial community and FIT measurements.
-The initial sample model had an AUC range of 0.657 to 0.796 after 100
-iterations of 20 repeated 10-fold cross validations. The test set AUC
-range for this model performed better then the training set AUCs. There
-was a marked decrease in the ROC curve for the final model used when
-compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
-a total of 70 variables that were used for this model {[}Figure 4B{]}.
-The variable that resulted in the largest MDA was \emph{Ruminococcaceae}
-(Otu000278) while FIT measurement for fecal blood resulted in the sixth
-largest decrease in MDA {[}Figure 4B{]}.
+\textbf{\emph{The Treatment Model:}} To test if there were differences
+that could be used to classify initial and follow up treatment samples
+we built a model to classify initial samples. This treatment model had
+an AUC range of 0.657 to 0.796. after 100 iterations of 20 repeated
+10-fold cross validations. The test set AUC range for this model
+performed better than the training set AUCs. There was a marked decrease
+in the ROC curve for the final model used when compared to the 100 test
+set AUC iterations {[}Figure 3A{]}. There was a total of 70 OTUs that
+were used for this model {[}Figure 3B{]}. The variable that resulted in
+the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
+Similarly, to the lesion model heavily represented genera included OTUs
+from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
+\emph{Blautia}, and \emph{Ruminococcus}.
 
-If there were common OTUs that could separate initial from follow up
-sample regardless of whether the lesion was adenoma or carcinoma we
-would expect to find a decrease in the positive probability of the
-follow up sample to be an initial sample. This is what we observed for
-the initial sample model (P-value \textless{} 0.001). When we separated
-individuals based on whether they had an adenoma or carcinoma there was
-a decrease in positive probability for both the carcinoma group
-{[}Figure 4C{]} (P-value \textless{} 0.001) and for the adenoma group
-{[}Figure 4D{]} (P-value \textless{} 0.001). For this model there was no
+If there were changes between intial and follow up samples we would
+expect the positive probability of being an initial sample to be
+decreased in the follow up. The is what we observed for the treatment
+model (P-value \textless{} 0.001). When we separated lesion into adenoma
+and carcinoma there was a decrease in positive arobability for both the
+carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
+{[}FIgure 3D{]} \textless{} 0.001). For this model, there was no
 difference between the predicted and actual classifications (P-value
@@
 
-\textbf{\emph{Common OTUs to both Models:}} We next wanted to compare
-the similarity between the OTU variables used in either model. The main
-purpose was to identify which OTUs were important not only for the
-classification of lesion but also for the classification of initial or
-follow up sample. Potentially, these specific OTUs are the most
-important with respect to the bacterial microbiome response to removal
-of lesion. When we compared the two different models with each other
-there were a total of 32 common OTUs. Some of the most common taxonomic
-identifications belonged to \emph{Lachnospiraceae},
-\emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}. The
-vast majority of these OTUs had classifications to bacteria typically
+\textbf{\emph{Common OTUs to both Models:}} We next wanted to know what
+predictors within the lesion model were also in the initial sample
+model. The main purpose was to identify which TOUs could be important in
+lesion formation and were impacted by treatment. When we compared the
+two different models with each other there were a total of 32 common
+OTUs. Some of the most common taxonomic identifications belonged to
+\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
+\emph{Blautia}, \emph{Anaerostipes}, \emph{Dorea}, and
+\emph{Ruminococcus}. These along withthe vast majority of the OTUs that
+were common between models had classifications to bacteria typically
 thought of as commensal {[}Table S2{]}.
 
-\textbf{\emph{Treatment Differences:}} After observing these changes in
-the bacterial community and positive probability we wanted to assess
-whether additional treatments, such as chemotherapy and radiation, could
-have an impact on the results that we observed. There was only a
-significant difference for change in positive probability for those
-treated with chemotherapy for the initial sample model (P-value
-\textless{} 0.05). This suggests that follow up samples for those
-treated for chemotherapy my have had a larger change from the initial
-sample then those without such treatment. All other variables that were
-tested showed no difference based on whether chemotherapy or radiation
-was used {[}Table S3{]}.
+\textbf{\emph{Chemotherapy and Radiation Differences:}} After observing
+these changes due to treatment we assessed whether chemotherapy or
+radiation impacted the observed results. Only the treatment model had a
+significant change in positive probability for those treated with
+chemotherapy or radiation versus those who received neither {[}Figure 4A
+\& Table S3{]} (P-value \textless{} 0.05). Using the 32 common OTUs we
+tested how chemotherapy or radiation affects important members of the
+community. FOr both chemotherapy or radiation, the directionaly of
+change of these OTUs were similar {[}Figure 4B \& 4C{]}. After multiple
+comparison correction, only \emph{Blautia} (Otu000006) remained
+significantly different between those who did and did not receive
+chemotherapy. This data suggests that follow up samples from those
+treated with either chemotherapy or radiation may have had a larger
+change from initial sample due to affects on the bacterial community.
 
@@
 
-This study builds upon previous work from numerous labs that have looked
-into the bacterial microbiome as a potential screening tool
-{[}7,8,10--12{]} by exploring what happens to the bacterial community
-after lesion removal. Based on previous work by Arthur, et al. {[}13{]}
-it may not be surprising to have E.coli as one of the most important
-OTUs and one that was common to both models. Interestingly, many of the
-most important OTUs had taxonomic identification for resident gut
-microbes. This could suggest that the bacterial community is one of the
-first components that could change during the pathogenesis of disease.
-These bacterial microbiome changes could be the first step in allowing
-more inflammatory bacterium to gain a foothold within the colon {[}9{]}.
+This study builds upon previous work from numerous labs that have
+considered both how the bacterial community between those with and
+without CRC differ and how it might be used as an early screening tool
+{[}7,8,10--12{]}. Here we describe how the bacterial community changes
+after treatment and which OTUs may be most important to lesion
+classification. Interestingly, many of the most important OTUs for both
+treatment and lesion classification had taxonomic identification for
+resident gut microbes which could suggest that members within the
+commensal community may be the first that change during pathogenesis of
+disease. Further, these changes could be the first step in allowing more
+inflammatory bacterium to gain a foothold within the colon {[}9{]}.
 
-From our results there were large observed differences in the bacterial
-microbiome in samples before and after lesion removal based on whether
-the individual had an adenoma or carcinoma. In individuals with
-carcinoma compared to adenoma there were much larger differences between
-initial and follow up samples based on the community beta diversity and
-in fecal blood as measured by FIT {[}Figure 1{]}. However, there were no
-differences between initial and follow up samples for any alpha
-diversity metric measured regardless of whether the individual had an
-adenoma or carcinoma {[}Table S1{]}. There was also no differences in
-relative abundance of any specific OTU for lesion, adenoma only, or
-carcinoma only {[}Figure S1{]}.
-
-Although there were no differences when investigating all OTUs, when
-looking specifically at four OTUs that taxonomically classified to
-previously suggested carcinoma associated microbes we found that only
-2/4 had a decrease in relative abundance between initial and follow up
-for those with carcinoma and 0/4 had differences for those with adenoma.
-This data would suggest that these specific OTUs may be important in the
-transition of an adenoma to a carcinoma but less so in the initiation of
-an adenoma from benign tissue.
-
-We next created a model that incorporated FIT and the bacterial
-microbiome to be able to classify lesions (adenoma and carcinoma). Based
-on this model we found that the follow up samples were closer to normal
-then the initial samples due to a decrease in positive probability for
-lesion and that the commonly associated CRC bacteria were not highly
-represented within this model with the exception of \emph{Porphyromonas
-asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
-towards what would be expected for normal controls, it should be noted
-that the follow up samples may not be a completely normal bacterial
-microbiome.
-
-After creating the lesion mode we then created a second model to
-classify initial versus follow up samples. We found that this model was
-able to accurately classify initial versus follow up samples suggesting
-that regardless of adenoma or carcinoma there are distinct common
-changes within the bacterial microbiome that occurs after lesion removal
-{[}Figure 4{]}. Both models had OTUs that overwhelmingly belonged to
-commensal bacteria. Providing additional information on the importance
-of commensal bacteria was that there were a total of 32 OTUs in common
-to both models and the vast majority belonged to regular residents of
-our gut community {[}Table S2{]}.
-
-Within our study there was a significant difference for the time elapsed
-in the collection of the follow up sample between adenoma and carcinoma
-(uncorrected P-value \textless{} 0.05), with time passed being less for
-adenoma (253 +/- 41.3 days) than carcinoma (351 +/- 102 days). These
-results would indicate that the findings described were specific to the
-surgical intervention and that some of the differences observed between
-carcinoma and adenoma samples could be due to differences in collection
-time between samples for the two different groups. Specifically, it
-could confound the observation that carcinomas changed more than
-adenomas {[}Figure 1A \& 1D{]}. This confounding though would not affect
-the observations where these individuals were grouped together {[}Figure
-3 \& 4{]}.
+Unlike previous studies on the microbiome and CRC, ours focuses on
+identifying commonalities within both adenoma and carcinoma groups
+before and after treatment. Although there were differences for genera
+associated with specific bacterium linked with CRC {[}Figure S2{]}.
+These changes were not consistent across lesion. Instead, most of our
+results provide support that the first members of the community to
+change and potentially stay changed even after treatment are those that
+are commensal bacteria {[}Figure 2-3 \& Table S3{]}. Additionally,
+treatment with chemotherapy or radiation changed these bacteria and
+provided a larger decrease in positive probability then removal of
+lesion.
 
 Curiously, we observed that the typical CRC associated bacteria were not
-predictive within our models. There are a number of reasons why this may
+predictive within our models. There are several reasons why this may
 have occurred. First, is that they were not present in enough
@@
 high degree of accuracy. Second, is that our Random Forest models were
-able to gather the same information from measures such as FIT or other
-OTUs. It is also possible that both of these explanations could have
-played a role. Regardless, our observations would suggest that an
-individual's resident bacteria have a large role to play in disease
-initiation and could change in a way that allows predictive models to
-lower the positive probability of a lesion after removal {[}Figure 3C \&
-3D{]}. It should be noted that our study does not argue against the
-importance of these CRC associated bacteria in the pathogenesis of
-disease but rather that they are not the main bacteria changing after
-removal of lesion. In fact, it is possible that these CRC associated
-bacteria are important in the transition from adenoma to carcinoma and
-would be one explanation as to why in our data we not only see high
-initial relative abundances in carcinoma and not adenoma individuals but
-also large decreases in relative abundance in some of those with
-carcinoma but not in those with adenoma after lesion removal {[}Figure
-2{]}.
+able to gather the same information from other OTUs. Third, it is also
+possible that changes in commensal bacteria are the common differences
+across adenoma and carcinoma. Finally, it is also possible that all
+these explanations could have played a role. Our observations though,
+would suggest that an individual's resident bacteria have a large role
+to play in disease initiation and could change in a way that allows
+predictive models to lower the positive probability of wither lesion or
+initial sample before treatment {[}Figure 2 \& 3{]}. It should be noted
+that our study does not argue against the importance of these CRC
+associated bacteria in the pathogenesis of disease but rather that they
+are not the main bacteria changing after treatment. It is possible that
+these CRC associated bacteria are important in the transition from
+adenoma to carcinoma and would be one explanation as to why in our data
+we not only see high initial relative abundances in carcinoma and not
+adenoma individuals but also large decreases in relative abundance in
+some of those with carcinoma but not in those with adenoma after lesion
+removal {[}Figure S2{]}.
 
-Many of the common OTUs between the two models had OTUs that
-taxonomically classified to potential butyrate producers {[}Table S2{]}.
-Another batch of OTUs classified to bacteria that can either degrade
-polyphenols or are inhibited by them. Both butyrate and polyphenols are
-thought to be protective against cancer in part by reducing inflammation
-{[}14{]}. These protective compounds are derived from the breakdown of
-fiber, fruits, and vegetables by resident gut microbes. One example of
-this potential diet-microbiome-inflammation-polyp axis is that
+Many of the common OTUs that we identified taxnomically classified to
+potential butyrate producers {[}Table S2{]}. Another batch of OTUs
+classified to bacteria that can either degrade polyphenols or are
+inhibited by them. Both butyrate and polyphenols are thought to be
+protective against cancer in part by reducing inflammation {[}13{]}.
+These protective compounds are derived from the breakdown of fiber,
+fruits, and vegetables by resident gut microbes. One example of this
+potential diet-microbiome-inflammation-polyp axis is that
 \emph{Bacteroides}, which was highly prevalent in our models, are known
 to be increased in those with high non-meat based protein consumption
-{[}15{]}. High protein consumption in general has been linked with an
-increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
+{[}14{]}. High protein consumption in general has been linked with an
+increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
 inhibited by polyphenols which are derived from fruits and vegetables
-{[}17{]}. Our data fits with the hypothesis that the microbial
+{[}16{]}. Our data fits with the hypothesis that the microbial
 metabolites from breakdown products within our own diet could not only
@@
 
-One limitation of our study is that we do not know whether individuals
-who were still classified as positive by the lesion model eventually had
-a subsequent CRC diagnosis. This information would help to strengthen
-the case for our lesion model keeping a number of individuals above the
-cutoff threshold even though at follow up they were diagnosed as no
-longer having a lesion. Another limitation is that we do not know if
-adding modern tests such as the stool DNA test {[}18{]} could help
-improve our overall AUC. This study also drew heavily from those with
-Caucasian ancestry making it possible that the observations may not be
+A limitation, in our study, was that there was a significant difference
+in the time elapsed in the collection of the follow up sample between
+adenoma and carcinoma (uncorrected P-value \textless{} 0.05), with time
+passed being less for adenoma (253 +/- 41.3 days) than carcinoma (351
++/- 102 days). These results would indicate that some of the differences
+observed between the carcinoma and adenoma groups could be due to
+differences in collection time. Specifically, it could confound the
+observation that carcinomas changed more than adenomas {[}Figure 1A \&
+1C{]}. This confounding though would not affect the observations where
+these individuals were grouped together {[}Figures 2-4{]}.
+
+Another limitation was that we do not know whether individuals who were
+still classified as positive by the lesion model eventually had a
+subsequent CRC diagnosis. This information would help to strengthen the
+case for this model keeping several individuals above the cutoff
+threshold even though at follow up they were diagnosed as no longer
+having a lesion. This study also drew heavily from those with Caucasian
+ancestry making it possible that the observations may not be
 representative of those with either Asian or African ancestry. Although
@@
 
-Interestingly, within the initial sample model the test data performed
-better than the training data. This may have occurred because the
-training AUC determined from 20 repeated 10 fold cross validation
-removed samples at random and did not take into account that they were
-matched samples. Another potential reason is that the model itself may
-be over-fit since the total number of samples was not that large.
-However, the lesion model did not suffer from these discrepancies.
-Further independent studies need to be carried out to verify our
-findings since we are dealing with correlations that may not be truly
-representative of the pathogenesis of disease.
+Interestingly, within the treatment model the test data performed better
+than the training data. This may have occurred because the training AUC
+determined from 20 repeated 10-fold cross validation removed samples at
+random and did not consider that they were matched samples. Another
+potential reason is that the model itself may have been over-fit since
+the total number of samples was not that large. However, the lesion
+model did not suffer from these discrepancies. Further independent
+studies need to be carried out to verify our findings on lesion and
+treatment changes due to these limitations.
 
-Despite these limitations our findings add to the existing scientific
-knowledge on CRC and the bacterial microbiome: That there is a
-measurable difference in the bacterial community after adenoma and
-carcinoma removal. Further, the ability for machine learning algorithms
-to take bacterial microbiome data and successfully lower positive
-probability after adenoma and carcinoma removal provides evidence that
-there are specific signatures, mostly attributable to commensal
-organisms, associated with these lesions. Our data provides evidence
-that commensal bacteria may be important in the development of polyps
-and also potentially the transition from adenoma to carcinoma.
+Despite these shortcomings our findings add to the existing scientific
+knowledge on CRC and the microbiome: That there is a measurable
+difference in the bacterial community after adenoma and carcinoma
+treatment. Further, the ability for machine learning algorithms to take
+OTU data and successfully lower positive probability of lesion or
+initial sample after treatment provides evidence that there are specific
+signatures, mostly attributable to commensal organisms, associated with
+both treatment and lesion. Our data provides evidence that commensal
+bacteria may be important in the development of polyps and potentially
+the transition from adenoma to carcinoma.
 
@@
 
-\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
-design were similar to that reported in Baxter, et al {[}7{]}. In brief,
-study exclusion involved those who had already undergone surgery,
-radiation, or chemotherapy, had colorectal cancer before a baseline
-fecal sample could be obtained, had IBD, a known hereditary
-non-polyposis colorectal cancer, or familial adenomatous polyposis.
-Samples used to build the models for prediction were collected either
-prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
-microbiome has been shown to normalize back to a pre-colonscopy
-community within this time period {[}19{]}. Our follow up data set had a
-total of 67 individuals that not only had a sample as described but also
-a follow up sample between 188 - 546 days after lesion removal and
-treatment had been completed. This study was approved by the University
-of Michigan Institutional Review Board. All study participants provided
-informed consent and the study itself conformed to the guidelines set
-out by the Helsinki Declaration.
+\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
+have been previously reported in Baxter, et al {[}7{]}. Briefly, study
+exclusion involved those who had already undergone surgery, radiation,
+or chemotherapy, had colorectal cancer before a baseline fecal sample
+could be obtained, had IBD, a known hereditary non-polyposis colorectal
+cancer, or familial adenomatous polyposis. Samples used to build the
+models for prediction were collected either prior to a colonoscopy or
+between 1 - 2 weeks after. The bacterial community has been shown to
+normalize back to a pre-colonscopy community within this time period
+{[}17{]}. Our study cohort consisted of 67 individuals with an initial
+sample as described and a follow up sample obtained between 188 - 546
+days after treatment of lesion. This study was approved by the
+University of Michigan Institutional Review Board. All study
+participants provided informed consent and the study itself conformed to
+the guidelines set out by the Helsinki Declaration.
 
-\textbf{\emph{FIT and 16S rRNA Gene Sequencing:}} FIT was analyzed as
-previously published using both OC FIT-CHEK and OC-Auto Micro 80
-automated system (Polymedco Inc.) {[}20{]}. 16S rRNA gene sequencing was
-completed as previously described by Kozich, et al. {[}21{]}. DNA
-extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
-and an epMotion 5075 automated pipetting system (Eppendorf). The V4
-variable region was amplified and the resulting product was split
-between three sequencing runs with normal, adenoma, and carcinoma evenly
-represented on each run. Each group was randomly assigned to avoid
-biases based on sample collection location.
+\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
+described by Kozich, et al. {[}18{]}. DNA extraction used the 96-well
+Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
+automated pipetting system (Eppendorf). The V4 variable region was
+amplified and the resulting product was split between three sequencing
+runs with normal, adenoma, and carcinoma evenly represented on each run.
+Each group was randomly assigned to avoid biases based on sample
+collection location. The initial and follow up samples were sequenced on
+the same run.
 
 \textbf{\emph{Sequence Processing:}} The mothur software package
-(v1.37.5) was used to process the 16S rRNA gene sequences. This process
-has been previously described {[}21{]}. The general processing workflow
-using mothur was as follows: Paired-end reads were first merged into
-contigs, quality filtered, aligned to the SILVA database, screened for
-chimeras, classified with a naive Bayesian classifier using the
-Ribosomal Database Project (RDP), and clustered into Operational
-Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
-neighbor clustering algorithm. The number of sequences for each sample
-was rarefied to 10523 in an attempt to minimize uneven sampling.
+(v1.37.5) was used to process the 16S rRNA gene sequences and has been
+previously described {[}18{]}. The general workflow using mothur was:
+Paired-end reads were first merged into contigs, quality filtered,
+aligned to the SILVA database, screened for chimeras, classified with a
+naive Bayesian classifier using the Ribosomal Database Project (RDP),
+and clustered into Operational Taxonomic Units (OTUs) using a 97\%
+similarity cutoff with an average neighbor clustering algorithm. The
+number of sequences for each sample was rarefied to 10523 to minimize
+uneven sampling.
 
-\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}22{]}
+\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}19{]}
 algorithm was used to create the model used for prediction of lesion
@@
 completed on an independent data set of 423 individuals. This model was
-then applied to our follow up data set of 67 individuals. It should be
-noted that all individuals with an adenoma or carcinoma were grouped
-together to form the lesion group and the model was not created to find
-differences between normal, adenoma, and carcinoma but rather
-differences between both adenoma and carcinoma versus normal.
+then applied to our 67-person cohort. It should be noted that all
+individuals with an adenoma or carcinoma were grouped together to form
+the lesion group and the model was not created to find differences
+between normal, adenoma, and carcinoma but rather differences between
+both adenoma and carcinoma versus normal.
 
-In brief, the model included data on FIT and the bacterial microbiome.
+The model included only OTU data obtained from 16S rRNA sequencing.
 Non-binary data was checked for near zero variance and OTUs that had
@@
 the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
-involved taking the samples and making 100 different 80/20 (train/test)
-splits of the data where normal and lesion were represented in the same
-proportion within both the whole data set and the 80/20 split. Each of
-these splits were then run through 20 repeated 10-fold cross validations
-to optimize the mtry hyper-parameter by maximizing the AUC (Area Under
-the Curve of the Receiver Operator Characteristic). This resulting model
-was then tested on the 20\% of the data that was originally held out
-from this overall process. Next, in order to assess which variables were
-most important to the model we counted the number of times a variable
-was present in the top 10\% of mean decrease in accuracy (MDA) for each
-of the 100 different 80/20 split models and then filtered this list to
-variables that were only present more than 50\% of the time. This final
-collated list of variables was what was considered the most important
-for the model. This reduced data set was then run through the mtry
-optimization again. Once the ideal mtry was found the entire 423 sample
-set was used to create the final Random Forest model on which
-classifications on the 67-person cohort was completed.
+involved making 100 different 80/20 (train/test) splits of the data
+where normal and lesion were represented in the same proportion within
+both the whole data set and the 80/20 split. For each a 20 repeated
+10-fold cross validation was performed on 80\% component to optimize the
+mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
+Receiver Operator Characteristic). The resulting model was then tested
+on the hold out data obtained from the 20\% component. Assessment of the
+most important OTUs to the model involved counting the number of times
+an OTU was present in the top 10\% of mean decrease in accuracy (MDA)
+for each of the 100 different splits run. This was then followed with
+filtering of this list to variables that were only present in more than
+50\% of these 100 runs. The final collated list of variables was then
+run through the mtry optimization again. Once the ideal mtry was found
+the entire 423 sample set was used to create the final Random Forest
+model on which classifications on the 67-person cohort was completed.
 
@@
 
-\textbf{\emph{Initial Sample Model Creation:}} We also investigated
-whether a model could be created that could identify pre (initial) and
-post (follow up) lesion removal samples from each other. The main
-difference was that only the 67-person cohort was used at all stages of
-model building and classification. Other than this difference the
-creation of this model and optimization of the mtry hyper-parameter was
-completed using the same procedure as was used for the lesion model.
-Instead of classifying samples as positive or negative of lesion this
-model classified samples as positive or negative for being an initial
-sample prior to lesion removal.
+\textbf{\emph{Treatment Model Creation:}} We also investigated whether a
+model could be created that could identify pre- (initial) and post-
+(follow up) treatment samples. The main difference was that only the
+67-person cohort was used at all stages of model building and
+classification. Other than this difference the creation of this model
+and optimization of the mtry hyper-parameter was completed using the
+same procedure as was used for the lesion model. Instead of classifying
+samples as positive or negative of lesion this model classified samples
+as positive or negative for being an initial sample prior to treatment.
 
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}23{]} in the vegan package
+community structure utilized PERMANOVA {[}20{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}21{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
-\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
-any differences based on whether the individual had an adenoma or
-carcinoma. This was done by testing initial and follow up samples for
-differences in alpha and beta diversity, testing differences in FIT
-between initial and follow ups, testing all OTUs, and investigating the
-relative abundance of specific previously associated CRC bacteria
-(\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
-\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
-stomatis}) based on adenoma or carcinoma. From here the lesion model was
-then tested for accuracy in prediction and whether it reduced the
-positive probability of lesion after surgery. We then used the initial
-sample model to assess whether it could classify samples better then the
-lesion model and whether it could reduce the positive probability of an
-initial sample in the follow up samples. Finally, a list of common OTUs
-were found for the two different models used.
+\textbf{\emph{Analysis Overview:}} We first tested for any differences
+based on whether the individual had an adenoma or carcinoma. This was
+done by testing initial and follow up samples for differences in alpha
+and beta diversity, testing all OTUs, and investigating the relative
+abundance of genera from previously associated CRC bacteria
+(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
+\emph{Porphyromonas}). Next, the lesion model was tested for accuracy in
+prediction and whether it reduced the positive probability of lesion in
+follow up samples. We then used the treatment model to assess whether it
+could classify samples better than the lesion model and whether it could
+reduce the positive probability of an initial sample in the follow up
+samples. Common OTUs were found for the two different models used to
+assess which were important for both models. Finally, differences
+between those receiving chemotherapy and radiation versus those who
+received neither were tested.
 
@@
 carcinoma group for thetayc (P-value = 0.000472). Advanced adenomas are
-denoted as Screen Relevant Neoplasia (SRN). B) A significant difference
-was found between the adenoma and carcinoma group for change in FIT
-measurement (P-value = 2.15e-05). Advanced adenomas are denoted as
-Screen Relevant Neoplasia (SRN). C) NMDS of the initial and follow up
-samples for the adenoma group. D) NMDS of the initial and follow up
-samples for the carcinoma group.
+denoted as Screen Relevant Neoplasia (SRN). B) NMDS of the initial and
+follow up samples for the adenoma group. C) NMDS of the initial and
+follow up samples for the carcinoma group.
 
-\textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
-Follow Up Samples.} A) Carcinoma initial and follow up samples had an
-observed significant difference in initial and follow up sample for the
-OTUs classified as \emph{Parvimonas micra} (P-value = 0.0116) and
-\emph{Porphyromonas asaccharolytica} (P-value = 0.00842). B) Adenoma
-initial and follow up samples. There were no significant differences
-between initial and follow up (P-value = 0.37.
-
-\textbf{Figure 3: The Lesion Model.} A) ROC curve: The shaded areas
-represents the range of values of a 100 different 80/20 splits of the
+\textbf{Figure 2: The Lesion Model.} A) ROC curve: The shaded areas
+represent the range of values of a 100 different 80/20 splits of the
 test set data and the blue line represents the model using 100\% of the
@@
 
-\textbf{Figure 4: The Initial Sample Model.} A) ROC curve: The shaded
-areas represents the range of values of a 100 different 80/20 splits of
-the test set data and the blue line represents the model using 100\% of
-the data set and what was used for subsequent classification. B) Summary
-of Important Variables. MDA of the most important variables in the
-initial sample model. The black point represents the mean and the
-different colors are the values of each different run up to 100. C)
-Positive probability change from initial to follow up sample in those
-with carcinoma. D) Positive probability change from initial to follow up
+\textbf{Figure 3: The Treatment Model.} A) ROC curve: The shaded areas
+represent the range of values of a 100 different 80/20 splits of the
+test set data and the blue line represents the model using 100\% of the
+data set and what was used for subsequent classification. B) Summary of
+Important Variables. MDA of the most important variables in the initial
+sample model. The black point represents the mean and the different
+colors are the values of each different run up to 100. C) Positive
+probability change from initial to follow up sample in those with
+carcinoma. D) Positive probability change from initial to follow up
 sample of those with adenoma or advanced adenoma (Screen Relevant
 Neoplasia (SRN)).
+
+\textbf{Figure 4: Chemotherapy and Radiation Therapy Effects on
+Community.} A) Treatment model initial sample positive probability
+reduction for chemotherapy. B) Treatment model initial sample positive
+probability reduction for radiation therapy. C) Common OTUs based on
+whether chemotherapy was received or not. The * denotes a P-value
+\textless{} 0.05 after multiple comparison correction. D) Common OTUs
+based on whether radiation therapy was received or not.
 
@@
 
-\textbf{Figure S2: Thetayc Versus Time of Follow up Sample from Initial}
+\textbf{Figure S2: Previously Associated CRC Bacteria in Initial and
+Follow Up Samples.} A) Carcinoma initial and follow up samples had an
+observed significant difference in initial and follow up sample for the
+OTUs classified as \emph{Parvimonas} (P-value = 0.0059),
+\emph{Porphyromonas} (P-value = 0.225), \emph{Peptostreptococcus}
+(P-value = 0.00424). B) Adenoma initial and follow up samples. There
+were no significant differences between initial and follow up (P-value =
+0.881.
 
@@
 
-\hypertarget{ref-arthur_intestinal_2012}{}
-13. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
-T-J, et al. Intestinal inflammation targets cancer-inducing activity of
-the microbiota. Science (New York, N.Y.). 2012;338:120--3.
-
 \hypertarget{ref-louis_gut_2014}{}
-14. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
+13. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
 metabolites and colorectal cancer. Nature Reviews Microbiology
@@
 \hypertarget{ref-zhu_intake_2016}{}
-15. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
+14. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
 Proteins Substantially Increased the Relative Abundance of Genus
@@
 \hypertarget{ref-mu_colonic_2016}{}
-16. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
+15. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
 Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
@@
 \hypertarget{ref-ozdal_reciprocal_2016}{}
-17. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
+16. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
 Reciprocal Interactions between Polyphenols and Gut Microbiota and
@@
 
-\hypertarget{ref-cotter_long-term_2016}{}
-18. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
-et al. Long-Term Follow-up of Patients Having False Positive
-Multi-target Stool DNA Tests after Negative Screening Colonoscopy: The
-LONG-HAUL Cohort Study. Cancer Epidemiology, Biomarkers \& Prevention: A
-Publication of the American Association for Cancer Research, Cosponsored
-by the American Society of Preventive Oncology. 2016;
-
 \hypertarget{ref-obrien_impact_2013}{}
-19. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+17. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
 
-\hypertarget{ref-daly_evaluation_2013}{}
-20. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
-for colorectal cancer screening. Journal of Primary Care \& Community
-Health. 2013;4:245--50.
-
 \hypertarget{ref-kozich_development_2013}{}
-21. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+18. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-22. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+19. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-23. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+20. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+21. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-03-14 submission/manuscript_outline_20161024.Rmd
manuscript update

--- 2017-03-07 submission/manuscript_outline_20161024.Rmd
@@
     keep_tex: yes
+  word_document: default
 geometry: margin=1.0in
@@
 chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
+chemo_rad_common_vars_summary <- read.csv('../data/process/tables/chemo_rads_treatment_pvalue_summary.csv', header = T, stringsAsFactors = F)
 
@@
 
-**Results:** There was a larger change in the bacterial community in response to lesion removal for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
 Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples decreased in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting a movement towards a normal bacterial community.
-An initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had no difference in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that the lesion associated community persists after treatment.
+A treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that there are members within the community that repond to treatment.
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
-A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g.
-*`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).
+A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g.
+*Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*).
 
-**Conclusions:** Our data supports the hypothesis that the bacterial community changes after treatment.
+**Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.
 Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.
@@
 Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Although these studies show how our gut bacteria can impact CRC progression via a changed community or invations by more inflammatory bacteria [@flynn_metabolic_2016].
+Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].
 They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).
 
-In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.
+In this study, we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.
 We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
 We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].
-We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.
-Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.
-We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.
-This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
+We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up treatment samples.
+Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether they were changing after treatment and whether it was towards a normal community.
+We also investigated the two models for similar important OTUs to identify which were important for classifying both lesion versus normal and initial versus follow up treatment samples.
+Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionally treatments significantly lowered any of the metrics examined versus those who received only lesion removal.
+This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
 
@@
 
-***Bacterial Community and FIT:***
-Within our `r length(rownames(good_metaf))` person cohort we first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`).
-We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1A].
-We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1B].
-The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
+***The Bacterial Community:***
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`) had any broad differences between their initial and follow up samples.
+We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1A].
+The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
 Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
-When investigating alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].
-We also observed that there was no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].
+There was no difference between initial and follow up samples when investigating alpha diversity metrics for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].
+Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].
 
-***Carcinoma Associated Bacteria:***
-So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).
-We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].
-Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A].
-In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].
+***Carcinoma Associated Genera:*** 
+For carcinoma [Figure S2A] (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05 & main_crc_bugs_summary["porp", "crc_BH"] < 0.05 & main_crc_bugs_summary["pept", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) but not adenoma [Figure S2B] (P-value `r ifelse(main_crc_bugs_summary["parv", "adn_BH"] < 0.05 & main_crc_bugs_summary["porp", "adn_BH"] < 0.05 & main_crc_bugs_summary["pept", "adn_BH"] < 0.05, paste("<"), paste(">"))` 0.05) there was a significant decrease in *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* between initial and follow up samples [Figure S2].
+*Fusobacterium* for both adenoma and carcinoma had no significant differences between initial and follow up samples [FIgure S2] (P-value `r ifelse(main_crc_bugs_summary["fn", "crc_BH"] < 0.05 & main_crc_bugs_summary["fn", "adn_BH"] < 0.05, paste("<"), paste(">"))` 0.05).
+Although there were significant differences, only a small percentage of those with adenoma or carcinoma were positive for any of these specific genera [Figure S2].
 
 ***The Lesion Model:***
-We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma and carcinoma versus normal controls.
-We investigated this by creating a model to classify samples as lesion versus normal based on the bacterial community and FIT measurements.
+We tested whether follow up samples were more normal than initial samples by building a model based solely on OTUs alone that classified lesion versus normal.
 This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
-The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].
-There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].
-The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
+The ROC curve for the final lesion model used was within the observed range of the 100-different test set AUC iterations [Figure 2A].
+There was a total of `r length(lesion_top10_varsMDA$variable)` OTUs that were used in this model [Figure 2B].
+*`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [Figure 2B].
+Heavily represented genera within the model included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
 
-If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.
-This is what we observed for the lesion model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group [Figure 3D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
-We also observed that there were no significant differences between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-The lesion model was also able to correctly classify the one individual who still had a carcinoma on follow up [Figure 3C].
+If there was movement towards a more normal community after treatment of lesion, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.
+We observed no such decrease in positive probability for either those with adenoma or carcinoma [Figure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+We also observed a significant difference between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+However, this model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].
 
-***The Initial Sample Model:***
-After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial (before lesion was removed) samples based on the bacterial community and FIT measurements.
-The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
-The test set AUC range for this model performed better then the training set AUCs.
-There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].
-There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].
-The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].
+***The Treatment Model:***
+To test if there were differences that could be used to classify initial and follow up treatment samples we built a model to classify initial samples.
+This treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
+after 100 iterations of 20 repeated 10-fold cross validations.
+The test set AUC range for this model performed better than the training set AUCs.
+There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3A].
+There was a total of `r length(IF_top10_varsMDA$variable)` OTUs that were used for this model [Figure 3B].
+The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
+Similar to the lesion model, heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
 
-If there were common OTUs that could separate initial from follow up sample regardless of whether the lesion was adenoma or carcinoma we would expect to find a decrease in the positive probability of the follow up sample to be an initial sample.
-This is what we observed for the initial sample model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-When we separated individuals based on whether they had an adenoma or carcinoma there was a decrease in positive probability for both the carcinoma group [Figure 4C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and for the adenoma group [Figure 4D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-For this model there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+If there were changes between initial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.
+The is what we observed for the treatment model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+When we separated lesion into adenoma and carcinoma there was a decrease in positive probability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [Figure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+For this model, there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
 
 ***Common OTUs to both Models:***
-We next wanted to compare the similarity between the OTU variables used in either model.
-The main purpose was to identify which OTUs were important not only for the classification of lesion but also for the classification of initial or follow up sample.
-Potentially, these specific OTUs are the most important with respect to the bacterial microbiome response to removal of lesion.
+We next wanted to know what predictors within the lesion model were also in the initial sample model.
+The main purpose was to identify which OTUs could be important in lesion formation and were impacted by treatment.
 When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.
-Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.
-The vast majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].
+Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000061")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000037")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000006")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000015")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000167")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.
+These along withthe vast majority of the OTUs that were common between models had classifications to bacteria typically thought of as commensal [Table S2].
 
-***Treatment Differences:***
-After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.
-There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
-This suggests that follow up samples for those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.
-All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].
+***Chemotherapy and Radiation Differences:*** 
+After observing these changes due to treatment we assessed whether chemotherapy or radiation impacted the observed results.
+Only the treatment model had a significant change in positive probability for those treated with chemotherapy or radiation versus those who received neither [Figure 4A & Table S3] (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
+Additionally, there was no difference between the change in positive probability for chemotherapy and radiation therapy (Uncorrected P-value `r ifelse(chemo_rad_summary['chemo_v_rads_IF', 'pvalue'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
+Using the `r length(common_vars_summary$otu)` common OTUs we tested how chemotherapy or radiation affects important members of the community.
+For both chemotherapy and radiation, the directionality of change of these OTUs were similar [Figure 4B & 4C].
+After multiple comparison correction, only *`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "lowest_ID"]`* (`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "otu"]`) remained significantly different between those who did and did not receive chemotherapy.
+This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to effects on the bacterial community.
 
@@
 ## Discussion
+This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
+Here we describe how the bacterial community changes after treatment and which OTUs may be most important to lesion classification.
+Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes.
+This suggest that members within the commensal community may be the first that change during CRC pathogenesis.
+These changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].
 
-This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after lesion removal.
-Based on previous work by Arthur, et al.
-[@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.
-Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.
-This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.
-These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].
-
-From our results there were large observed differences in the bacterial microbiome in samples before and after lesion removal based on whether the individual had an adenoma or carcinoma.
-In individuals with carcinoma compared to adenoma there were much larger differences between initial and follow up samples based on the community beta diversity and in fecal blood as measured by FIT [Figure 1].
-However, there were no differences between initial and follow up samples for any alpha diversity metric measured regardless of whether the individual had an adenoma or carcinoma [Table S1].
-There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].
-
-Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested carcinoma associated microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.
-This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.
-
-We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma and carcinoma).
-Based on this model we found that the follow up samples were closer to normal then the initial samples due to a decrease in positive probability for lesion and that the commonly associated CRC bacteria were not highly represented within this model with the exception of *Porphyromonas asaccharolytica* [Figure 3].
-Although there was a detectable change towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.
-
-After creating the lesion mode we then created a second model to classify initial versus follow up samples.
-We found that this model was able to accurately classify initial versus follow up samples suggesting that regardless of adenoma or carcinoma there are distinct common changes within the bacterial microbiome that occurs after lesion removal [Figure 4].
-Both models had OTUs that overwhelmingly belonged to commensal bacteria.
-Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].
-
-Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
-These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.
-Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1D].
-This confounding though would not affect the observations where these individuals were grouped together [Figure 3 & 4].
+Unlike previous studies on the microbiome and CRC, ours focuses on identifying commonalities within both adenoma and carcinoma groups before and after treatment.
+Although there were differences for genera associated with specific bacterium linked with CRC [Figure S2].
+These changes were not consistent across lesion.
+Instead, most of our results provide support that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria [Figure 2-3 & Table S3].
+Additionally, treatment with chemotherapy or radiation changed these bacteria and provided a larger decrease in positive probability then removal of lesion.
 
 Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
-There are a number of reasons why this may have occurred.
+There are several reasons why this may have occurred.
 First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
-Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.
-It is also possible that both of these explanations could have played a role.
-Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after removal [Figure 3C & 3D].
-It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after removal of lesion.
-In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after lesion removal [Figure 2].
+Second, is that our Random Forest models were able to gather the same information from other OTUs.
+Third, is that changes in commensal bacteria are the common differences across adenoma and carcinoma.
+Finally, it is also possible that all these explanations could have played a role.
+Our observations though, would suggest that an individual's resident bacteria have a large role to play in polyp formation and could change in a way that allows predictive models to lower the positive probability of lesion or initial sample before treatment [Figure 2 & 3].
+It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after treatment.
+It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after treatment [Figure S2].
 
-Many of the common OTUs between the two models had OTUs that taxonomically classified to potential butyrate producers [Table S2].
+Many of the common OTUs that we identified taxnomically classified to potential butyrate producers [Table S2].
 Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.
@@
 
-One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
-This information would help to strengthen the case for our lesion model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
-Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.
+A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
+These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time.
+Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1C].
+This confounding though would not affect the observations where these individuals were grouped together [Figures 2-4].
+
+Another limitation was that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
+This information would help to strengthen the case for this model keeping several individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
 This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.
@@
 
-Interestingly, within the initial sample model the test data performed better than the training data.
-This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.
-Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.
+Interestingly, within the treatment model the test data performed better than the training data.
+This may have occurred because the training AUC determined from 20 repeated 10-fold cross validation removed samples at random and did not consider that they were matched samples.
+Another potential reason is that the model itself may have been over-fit since the total number of samples was not that large.
 However, the lesion model did not suffer from these discrepancies.
-Further independent studies need to be carried out to verify our findings since we are dealing with correlations that may not be truly representative of the pathogenesis of disease.
+Further independent studies need to be carried out to verify our findings on lesion and treatment changes due to these limitations.
 
-Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma and carcinoma removal.
-Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after adenoma and carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.
-Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.
+Despite these shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma and carcinoma treatment.
+Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of lesion or initial sample after treatment provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with both treatment and lesion.
+Our data provides evidence that commensal bacteria may be important in the development of polyps and potentially the transition from adenoma to carcinoma.
 
@@
 ***Study Design and Patient Sampling:***
-The sampling and design were similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].
-In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
+Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016].
+Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial microbiome has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013].
-Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after lesion removal and treatment had been completed.
+The bacterial community has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013].
+Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion.
 This study was approved by the University of Michigan Institutional Review Board.
@@
 
-***FIT and 16S rRNA Gene Sequencing:***
-FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013].
-16S rRNA gene sequencing was completed as previously described by Kozich, et al.
+***16S rRNA Gene Sequencing:***
+Sequencing was completed as described by Kozich, et al.
 [@kozich_development_2013].
-DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
+DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
 The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.
 Each group was randomly assigned to avoid biases based on sample collection location.
+The initial and follow up samples were sequenced on the same run.
 
 ***Sequence Processing:***
-The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.
-This process has been previously described [@kozich_development_2013].
-The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
-The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.
+The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013].
+The general workflow using mothur was: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
+The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize uneven sampling.
 
@@
 The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on an independent data set of  `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
-This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.
+This model was then applied to our `r length(rownames(good_metaf))`-person cohort.
 It should be noted that all individuals with an adenoma or carcinoma were grouped together to form the lesion group and the model was not created to find differences between normal, adenoma, and carcinoma but rather differences between both adenoma and carcinoma versus normal.
 
-In brief, the model included data on FIT and the bacterial microbiome.
+The model included only OTU data obtained from 16S rRNA sequencing.
 Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
-Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.
-Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
-This resulting model was then tested on the 20% of the data that was originally held out from this overall process.
-Next, in order to assess which variables were most important to the model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.
-This final collated list of variables was what was considered the most important for the model.
-This reduced data set was then run through the mtry optimization again.
+Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.
+For each a 20 repeated 10-fold cross validation was performed on 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
+The resulting model was then tested on the hold out data obtained from the 20% component.
+Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run.
+This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs.
+The final collated list of variables was then run through the mtry optimization again.
 Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
@@
 
-***Initial Sample Model Creation:***
-We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal samples from each other.
+***Treatment Model Creation:***
+We also investigated whether a model could be created that could identify pre- (initial) and post- (follow up) treatment samples.
 The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
 Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.
-Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial sample prior to lesion removal.
+Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial sample prior to treatment.
 
@@
 ***Analysis Overview:***
-We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.
-This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma or carcinoma.
-From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.
-We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.
-Finally, a list of common OTUs were found for the two different models used.
+We first tested for any differences based on whether the individual had an adenoma or carcinoma.
+This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing all OTUs, and investigating the relative abundance of genera from previously associated CRC bacteria (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*).
+Next, the lesion model was tested for accuracy in prediction and whether it reduced the positive probability of lesion in follow up samples.
+We then used the treatment model to assess whether it could classify samples better than the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.
+Common OTUs were found for the two different models used to assess which were important for both models.
+Finally, differences between those receiving chemotherapy and radiation versus those who received neither were tested.
 
@@
 Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).
-B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
-Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).
-C) NMDS of the initial and follow up samples for the adenoma group.
-D) NMDS of the initial and follow up samples for the carcinoma group.
+B) NMDS of the initial and follow up samples for the adenoma group.
+C) NMDS of the initial and follow up samples for the carcinoma group.
 
-**Figure 2: Previously Associated CRC Bacteria in Initial and Follow Up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).
-B) Adenoma initial and follow up samples.
-There were no significant differences between initial and follow up (P-value = `r format(main_crc_bugs_summary["porp", "adn_BH"], digits = 3)`.
-
-**Figure 3: The Lesion Model.** A) ROC curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure 2: The Lesion Model.** A) ROC curve: The shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Summary of Important Variables.
@@
 
-**Figure 4: The Initial Sample Model.**  A) ROC curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure 3: The Treatment Model.**  A) ROC curve: The shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Summary of Important Variables.
@@
 
+**Figure 4: Chemotherapy and Radiation Therapy Effects on Community.**  A) Treatment model initial sample positive probability reduction for chemotherapy.
+B) Treatment model initial sample positive probability reduction for radiation therapy.
+C) Common OTUs based on whether chemotherapy was received or not.
+The * denotes a P-value < 0.05 after multiple comparison correction.
+D) Common OTUs based on whether radiation therapy was received or not.
+
 \newpage
@@
 
-**Figure S2: Thetayc Versus Time of Follow up Sample from Initial**
+**Figure S2: Previously Associated CRC Bacteria in Initial and Follow Up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), *Porphyromonas* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`), *Peptostreptococcus* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`).
+B) Adenoma initial and follow up samples.
+There were no significant differences between initial and follow up (P-value = `r format(main_crc_bugs_summary["porp", "adn_BH"], digits = 3)`.
 

### 2017-03-14 submission/manuscript_outline_20161024.tex
manuscript update

@@
 \textbf{Results:} There was a larger change in the bacterial community
-in response to lesion removal for carcinoma versus adenoma cases
-(P-value \textless{} 0.05). Yet no difference was found in the relative
-abundance of any OTU before and after treatment for adenoma or carcinoma
-groups (P-value \textgreater{} 0.05). A lesion model had an AUC range of
-0.692 - 0.761 and follow up samples decreased in the positive
+in response to treatment for carcinoma versus adenoma cases (P-value
+\textless{} 0.05). Yet no difference was found in the relative abundance
+of any OTU before and after treatment for adenoma or carcinoma groups
+(P-value \textgreater{} 0.05). A lesion model had an AUC range of 0.692
+- 0.761 and follow up samples had no difference in the positive
 probability of lesion versus initial samples (P-value \textgreater{}
-0.05); suggesting a movement towards a normal bacterial community. An
-initial sample model had an AUC range of 0.657 - 0.796 and had a
+0.05); suggesting that the lesion associated community persists after
+treatment. A treatment model had an AUC range of 0.657 - 0.796 and had a
 decrease positive probability for the follow up samples to be an initial
-sample (P-value \textless{} 0.05). The lesion model used a total of 54
-variables while the initial sample model used a total of 70 variables. A
-total of 32 OTUs were common to both models with the majority of these
+sample (P-value \textless{} 0.05); suggesting that there are members
+within the community that repond to treatment. The lesion model used a
+total of 54 variables while the initial sample model used a total of 70
+variables. A total of 32 OTUs were common to both models with many
 classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
-\emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}).
+\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
+\emph{Ruminococcus}).
 
-\textbf{Conclusions:} Our data supports the hypothesis that the
-bacterial community changes after treatment. Individuals with carcinoma
-have more drastic differences to the overall community then those with
-adenoma. Commensal bacteria were crucial for accurate model
+\textbf{Conclusions:} Our data partially supports the hypothesis that
+the bacterial community changes after treatment. Individuals with
+carcinoma have more drastic differences to the overall community then
+those with adenoma. Commensal bacteria were crucial for accurate model
 classification, suggesting that these bacteria may be important to
@@
 gut bacteria can impact CRC progression via a changed community or
-invations by more inflammatory bacteria {[}9{]}. They provide very
-little information as to whether these changed communities rebound after
+invasion by more inflammatory bacteria {[}9{]}. They provide very little
+information as to whether these changed communities rebound after
 successful treatment of lesion (adenoma or carcinoma).
 
-In this study we tested the hypothesis that there are detectable changes
-to the bacterial community between pre- (initial) and post- (follow up)
-treatment of lesion. We analyzed changes in alpha and beta diversity as
-well as the relative abundance of specific Operational Taxonomic Units
-(OTUs). We then investigated several important genera from oral microbes
-that have been suggested to be important in CRC pathogenesis {[}9{]}. We
-next utilized Random Forest to build two models: The first was built to
-classify lesion versus non-lesion (normal) while the second was built to
-classify initial versus follow up samples. Subsequent observations on
-how these models and OTUs within them performed before and after
-treatment helped inform us as to whether initial and follow up samples
-were changing and whether it was towards a normal community. We also
-investigated the two models for similar important OTUs to identify the
-crucial OTUs involved with both classifying lesion or normal and initial
-versus follow up samples. Finally, since treatment also included
-chemotherapy and radiation therapy along with lesion removal, we tested
-if these additionaly treatments significantly lowered any of the metrics
-examined versus those who received only lesion removal. This study will
-provide evidence as to whether treatment can influence the community and
-if the CRC microbiome, identified in previous studies, persists.
+In this study, we tested the hypothesis that there are detectable
+changes to the bacterial community between pre- (initial) and post-
+(follow up) treatment of lesion. We analyzed changes in alpha and beta
+diversity as well as the relative abundance of specific Operational
+Taxonomic Units (OTUs). We then investigated several important genera
+from oral microbes that have been suggested to be important in CRC
+pathogenesis {[}9{]}. We next utilized Random Forest to build two
+models: The first was built to classify lesion versus non-lesion
+(normal) while the second was built to classify initial versus follow up
+treatment samples. Subsequent observations on how these models and OTUs
+within them performed before and after treatment helped inform us as to
+whether they were changing after treatment and whether it was towards a
+normal community. We also investigated the two models for similar
+important OTUs to identify which were important for classifying both
+lesion versus normal and initial versus follow up treatment samples.
+Finally, since treatment also included chemotherapy and radiation
+therapy along with lesion removal, we tested if these additionally
+treatments significantly lowered any of the metrics examined versus
+those who received only lesion removal. This study helps to provide
+evidence as to whether treatment can influence the community and if the
+CRC microbiome, identified in previous studies, persists.
 
@@
 \textbf{\emph{The Lesion Model:}} We tested whether follow up samples
-were more normal than initial samples by building a model that
-classified lesion versus normal. This model had an AUC range of 0.692 -
-0.761 after 100 iterations of 20 repeated 10-fold corss validations. The
-ROC curve for the final lesion model used was within the observed range
-of the 100-different test set AUC iterations {[}Figure 2A{]}. There was
-a total of 54 OTUs that were used in this model {[}Figure 2B{]}.
-\emph{Lachnospiraceae} (Otu000015) resulted in the largest decrease in
-MDA {[}FIgure 2B{]}. Heavily represented genera within the model
-included OTUs from \emph{Lachnospiraceae}, \emph{Bacteroides},
-\emph{Roseburia}, \emph{Blautia}, and \emph{Ruminococcus}.
+were more normal than initial samples by building a model based solely
+on OTUs alone that classified lesion versus normal. This model had an
+AUC range of 0.692 - 0.761 after 100 iterations of 20 repeated 10-fold
+cross validations. The ROC curve for the final lesion model used was
+within the observed range of the 100-different test set AUC iterations
+{[}Figure 2A{]}. There was a total of 54 OTUs that were used in this
+model {[}Figure 2B{]}. \emph{Lachnospiraceae} (Otu000015) resulted in
+the largest decrease in MDA {[}Figure 2B{]}. Heavily represented genera
+within the model included OTUs from \emph{Lachnospiraceae},
+\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
+\emph{Ruminococcus}.
 
@@
 positive probability for either those with adenoma or carcinoma
-{[}FIgure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
+{[}Figure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
 significant difference between the predicted and actual calls (P-value
-\textless{} 0.05). This model correctly classified the one individual
-who still had carcinoma on follow up and this individual's positive
-probability of lesion increased between their initial and follow up
-sample {[}Figure 2C{]}.
+\textless{} 0.05). However, this model correctly classified the one
+individual who still had carcinoma on follow up and this individual's
+positive probability of lesion increased between their initial and
+follow up sample {[}Figure 2C{]}.
 
@@
 the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
-Similarly, to the lesion model heavily represented genera included OTUs
+Similar to the lesion model, heavily represented genera included OTUs
 from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
@@
 
-If there were changes between intial and follow up samples we would
+If there were changes between initial and follow up samples we would
 expect the positive probability of being an initial sample to be
@@
 model (P-value \textless{} 0.001). When we separated lesion into adenoma
-and carcinoma there was a decrease in positive arobability for both the
+and carcinoma there was a decrease in positive probability for both the
 carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
-{[}FIgure 3D{]} \textless{} 0.001). For this model, there was no
+{[}Figure 3D{]} \textless{} 0.001). For this model, there was no
 difference between the predicted and actual classifications (P-value
@@
 predictors within the lesion model were also in the initial sample
-model. The main purpose was to identify which TOUs could be important in
+model. The main purpose was to identify which OTUs could be important in
 lesion formation and were impacted by treatment. When we compared the
@@
 chemotherapy or radiation versus those who received neither {[}Figure 4A
-\& Table S3{]} (P-value \textless{} 0.05). Using the 32 common OTUs we
-tested how chemotherapy or radiation affects important members of the
-community. FOr both chemotherapy or radiation, the directionaly of
-change of these OTUs were similar {[}Figure 4B \& 4C{]}. After multiple
-comparison correction, only \emph{Blautia} (Otu000006) remained
-significantly different between those who did and did not receive
-chemotherapy. This data suggests that follow up samples from those
-treated with either chemotherapy or radiation may have had a larger
-change from initial sample due to affects on the bacterial community.
+\& Table S3{]} (P-value \textless{} 0.05). Additionally, there was no
+difference between the change in positive probability for chemotherapy
+and radiation therapy (Uncorrected P-value \textgreater{} 0.05). Using
+the 32 common OTUs we tested how chemotherapy or radiation affects
+important members of the community. For both chemotherapy and radiation,
+the directionality of change of these OTUs were similar {[}Figure 4B \&
+4C{]}. After multiple comparison correction, only \emph{Blautia}
+(Otu000006) remained significantly different between those who did and
+did not receive chemotherapy. This data suggests that follow up samples
+from those treated with either chemotherapy or radiation may have had a
+larger change from initial sample due to effects on the bacterial
+community.
 
@@
 treatment and lesion classification had taxonomic identification for
-resident gut microbes which could suggest that members within the
-commensal community may be the first that change during pathogenesis of
-disease. Further, these changes could be the first step in allowing more
-inflammatory bacterium to gain a foothold within the colon {[}9{]}.
+resident gut microbes. This suggest that members within the commensal
+community may be the first that change during CRC pathogenesis. These
+changes, in turn, could be the first step in allowing more inflammatory
+bacteria to gain a foothold within the colon {[}9{]}.
 
@@
 high degree of accuracy. Second, is that our Random Forest models were
-able to gather the same information from other OTUs. Third, it is also
-possible that changes in commensal bacteria are the common differences
-across adenoma and carcinoma. Finally, it is also possible that all
-these explanations could have played a role. Our observations though,
-would suggest that an individual's resident bacteria have a large role
-to play in disease initiation and could change in a way that allows
-predictive models to lower the positive probability of wither lesion or
-initial sample before treatment {[}Figure 2 \& 3{]}. It should be noted
-that our study does not argue against the importance of these CRC
-associated bacteria in the pathogenesis of disease but rather that they
-are not the main bacteria changing after treatment. It is possible that
-these CRC associated bacteria are important in the transition from
-adenoma to carcinoma and would be one explanation as to why in our data
-we not only see high initial relative abundances in carcinoma and not
-adenoma individuals but also large decreases in relative abundance in
-some of those with carcinoma but not in those with adenoma after lesion
-removal {[}Figure S2{]}.
+able to gather the same information from other OTUs. Third, is that
+changes in commensal bacteria are the common differences across adenoma
+and carcinoma. Finally, it is also possible that all these explanations
+could have played a role. Our observations though, would suggest that an
+individual's resident bacteria have a large role to play in polyp
+formation and could change in a way that allows predictive models to
+lower the positive probability of lesion or initial sample before
+treatment {[}Figure 2 \& 3{]}. It should be noted that our study does
+not argue against the importance of these CRC associated bacteria in the
+pathogenesis of disease but rather that they are not the main bacteria
+changing after treatment. It is possible that these CRC associated
+bacteria are important in the transition from adenoma to carcinoma and
+would be one explanation as to why in our data we not only see high
+initial relative abundances in carcinoma and not adenoma individuals but
+also large decreases in relative abundance in some of those with
+carcinoma but not in those with adenoma after treatment {[}Figure S2{]}.
 

### 2017-03-15 code/crc_reference_run_RF.R
crc specific model

@@ -1,0 +1,54 @@
+### Build the best lesion model possible
+### Try XG-Boost, RF, Logit (GLM), C5.0, SVM
+### Find the best based on Jenna Wiens suggestions on test and training
+## Marc Sze
+
+#Load needed libraries
+source('code/functions.R')
+
+loadLibs(c("dplyr", "caret","scales", "doMC"))
+
+load("exploratory/crc_RF_model_setup.RData")
+
+# Set i variable
+
+i = 1
+
+#################################################################################
+#                                                                               #
+#                                                                               #
+#               Model Training and Parameter Tuning                             #
+#                                                                               #
+#################################################################################
+
+# Call number of processors to use
+registerDoMC(cores = 4)
+
+#Set up lists to store the data
+test_tune_list <- list()
+test_predictions <- list()
+
+#Get test data
+train_test_data <- test_data[eighty_twenty_splits[, i], ]
+  
+#Train the model
+train_name <- paste("data_split", i, sep = "")
+
+set.seed(3457)
+test_tune_list[[paste("data_split", i, sep = "")]] <- assign(train_name, 
+  train(lesion ~ ., data = train_test_data, 
+        method = "rf", 
+        ntree = 2000, 
+        trControl = fitControl, 
+        metric = "ROC", 
+        verbose = FALSE))
+   
+test_test_data <- test_data[-eighty_twenty_splits[, i], ]
+  
+test_predictions[[paste("data_split", i, sep = "")]] <- 
+predict(test_tune_list[[paste("data_split", i, sep = "")]], 
+  test_test_data)
+
+
+# Save image with data and relevant parameters
+save.image(paste("exploratory/crc_RF_model_", i, ".RData", sep=""))

### 2017-03-15 code/setup_crc_RF_test.R

@@ -1,0 +1,104 @@
+### Build the best lesion model possible
+### Find the best based on Jenna Wiens suggestions on test and training
+## Marc Sze
+
+#Load needed libraries
+source('code/functions.R')
+
+loadLibs(c("dplyr", "caret","scales", "doMC"))
+
+# Read in necessary data frames
+
+shared <- read.delim('data/process/final.0.03.subsample.shared', 
+  header=T, sep='\t') %>% select(Group, contains("Otu0"))
+
+metaI <- read.csv("data/process/mod_metadata/metaI_final.csv", 
+  stringsAsFactors = F, header = T) 
+
+good_metaf <- read.csv('data/process/mod_metadata/good_metaf_final.csv', 
+                       header = T, stringsAsFactors = F) %>% select(initial)
+
+metaI <- filter(metaI, !(sample %in% good_metaf$initial))
+
+#################################################################################
+#                                                                               #
+#                                                                               #
+#               Data Clean up for better and faster modeling                    #
+#                                                                               #
+#################################################################################
+
+# Remove follow up samples and join metadata with microbiome data
+test_data <- inner_join(metaI, shared, by = c("sample" = "Group")) %>% 
+  filter(dx == "cancer" | dx == "normal")
+rm(shared)
+
+# Filter and use only specific data
+
+test_data <- test_data %>% 
+  select(sample, contains("Otu0"))
+
+#Filter out rows that are not all complete
+
+test_data <- test_data[complete.cases(test_data), ]
+
+#Reduce MetaI to match the test_data
+metaI <- filter(metaI, dx == "cancer" | dx == "normal")
+
+rownames(test_data) <- test_data$sample
+test_data <- select(test_data, -sample)
+
+# Remove those with near zero variance 
+  # Similar to removal of singletons or removal based on percentage in sample
+  # Using near zero variance should cover the above two so could use only this instead
+
+nzv <- nearZeroVar(test_data)
+    # By default, nearZeroVar will return the positions of the variables that are flagged to be problematic
+    # using the saveMetrics argument will output a table with more in-depth info
+    # Group stays in this case (because it is numberic) 
+        # but maybe better in future to transfer this to row names if not a numberic label
+
+test_data <- test_data[, -nzv]
+
+# Add the lesion variable to the test_data
+test_data <- cbind(lesion = factor(metaI$lesion, 
+  levels = c(0, 1), labels = c("No", "Yes")), test_data)
+
[…]

### 2017-03-15 submission/manuscript_outline_20161024.Rmd

@@
 output:
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-  word_document: default
 geometry: margin=1.0in
@@
 
-**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
+**Results:** There was a larger change in the bacterial community in response to lesion removal for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
 Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had no difference in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that the lesion associated community persists after treatment.
-A treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that there are members within the community that repond to treatment.
+A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples decreased in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting a movement towards a normal bacterial community.
+An initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
-A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g.
-*Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*).
+A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g.
+*`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).
 
-**Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.
+**Conclusions:** Our data supports the hypothesis that the bacterial community changes after treatment.
 Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.
@@
 Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].
+Although these studies show how our gut bacteria can impact CRC progression via a changed community or invations by more inflammatory bacteria [@flynn_metabolic_2016].
 They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).
 
-In this study, we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.
+In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.
 We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
 We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].
-We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up treatment samples.
-Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether they were changing after treatment and whether it was towards a normal community.
-We also investigated the two models for similar important OTUs to identify which were important for classifying both lesion versus normal and initial versus follow up treatment samples.
-Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionally treatments significantly lowered any of the metrics examined versus those who received only lesion removal.
-This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
+We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.
+Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.
+We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.
+Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionaly treatments significantly lowered any of the metrics examined versus those who received only lesion removal.
+This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
 
@@
 ***The Lesion Model:***
-We tested whether follow up samples were more normal than initial samples by building a model based solely on OTUs alone that classified lesion versus normal.
-This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
+We tested whether follow up samples were more normal than initial samples by building a model that classified lesion versus normal.
+This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold corss validations.
 The ROC curve for the final lesion model used was within the observed range of the 100-different test set AUC iterations [Figure 2A].
 There was a total of `r length(lesion_top10_varsMDA$variable)` OTUs that were used in this model [Figure 2B].
-*`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [Figure 2B].
+*`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [FIgure 2B].
 Heavily represented genera within the model included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
@@
 If there was movement towards a more normal community after treatment of lesion, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.
-We observed no such decrease in positive probability for either those with adenoma or carcinoma [Figure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
+We observed no such decrease in positive probability for either those with adenoma or carcinoma [FIgure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
 We also observed a significant difference between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-However, this model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].
+This model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].
 
@@
 The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
-Similar to the lesion model, heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
+Similarly, to the lesion model heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
 
-If there were changes between initial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.
+If there were changes between intial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.
 The is what we observed for the treatment model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-When we separated lesion into adenoma and carcinoma there was a decrease in positive probability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [Figure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
+When we separated lesion into adenoma and carcinoma there was a decrease in positive arobability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [FIgure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
 For this model, there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
@@
 We next wanted to know what predictors within the lesion model were also in the initial sample model.
-The main purpose was to identify which OTUs could be important in lesion formation and were impacted by treatment.
+The main purpose was to identify which TOUs could be important in lesion formation and were impacted by treatment.
 When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.
@@
 Only the treatment model had a significant change in positive probability for those treated with chemotherapy or radiation versus those who received neither [Figure 4A & Table S3] (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
-Additionally, there was no difference between the change in positive probability for chemotherapy and radiation therapy (Uncorrected P-value `r ifelse(chemo_rad_summary['chemo_v_rads_IF', 'pvalue'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
 Using the `r length(common_vars_summary$otu)` common OTUs we tested how chemotherapy or radiation affects important members of the community.
-For both chemotherapy and radiation, the directionality of change of these OTUs were similar [Figure 4B & 4C].
+FOr both chemotherapy or radiation, the directionaly of change of these OTUs were similar [Figure 4B & 4C].
 After multiple comparison correction, only *`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "lowest_ID"]`* (`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "otu"]`) remained significantly different between those who did and did not receive chemotherapy.
-This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to effects on the bacterial community.
+This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to affects on the bacterial community.
 
@@
 Here we describe how the bacterial community changes after treatment and which OTUs may be most important to lesion classification.
-Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes.
-This suggest that members within the commensal community may be the first that change during CRC pathogenesis.
-These changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].
+Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes which could suggest that members within the commensal community may be the first that change during pathogenesis of disease.
+Further, these changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].
 
@@
 Second, is that our Random Forest models were able to gather the same information from other OTUs.
-Third, is that changes in commensal bacteria are the common differences across adenoma and carcinoma.
+Third, it is also possible that changes in commensal bacteria are the common differences across adenoma and carcinoma.
 Finally, it is also possible that all these explanations could have played a role.
-Our observations though, would suggest that an individual's resident bacteria have a large role to play in polyp formation and could change in a way that allows predictive models to lower the positive probability of lesion or initial sample before treatment [Figure 2 & 3].
+Our observations though, would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of wither lesion or initial sample before treatment [Figure 2 & 3].
 It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after treatment.
-It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after treatment [Figure S2].
+It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after lesion removal [Figure S2].
 

### 2017-03-15 submission/supplemental_outline_20161024.Rmd

@@
 \newpage
+\newcommand{\blandscape}{\begin{landscape}}
+\newcommand{\elandscape}{\end{landscape}}
+
+\blandscape
 
@@
 
-variable <- c(rep("Positive Probability", 8), rep("Thetayc", 2), rep(c("Sobs", "Shannon", "Evenness"), 2))
-comparison <- c(rep("Chemotherapy", 4), rep("Radiation", 4), "Chemotherapy", "Radiation", 
-                rep("Chemotherapy", 3), rep("Radiation", 3))
+variable <- c(rep("sobs", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4))
+comparison <- rep(c("Chemotherapy", "Radiation"), 6)
 
-st3 <- cbind(variable, comparison, format(chemo_rad_summary$pvalue, digits = 4), format(chemo_rad_summary$bh, digits = 4))
+st3 <- cbind(variable, comparison, format(chemo_rad_summary$chemo_rad_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$removal_only_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$cr_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$ro_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$pvalue, scientific = TRUE, digits = 2), format(chemo_rad_summary$bh, scientific = TRUE, digits = 2))
 
-colnames(st3) <- c("Variable Tested", "Treatment", "P-value", "BH Corrected P-value")
+colnames(st3) <- c("Variable", "Additive Treatment", "Additive Treatment Mean", "Removal Mean", "Additive Treatment SD", "Removal SD", "P-value", "BH Corrected")
 
@@
 ```
+
+\elandscape

### 2017-03-21 results/tables/Table1.Rmd
add data for new direction

---
title: ""
output: pdf_document
---

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../../code/functions.R')

loadLibs(c("tidyr", "dplyr", "scales", "knitr"))
```

**Table 1: Demographic Data of Training Cohort**

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}

metadata <- read.csv("../../data/process/mod_metadata/metaI_final.csv", header = T, stringsAsFactors = F)
test_cohort <- read.csv("../../data/process/mod_metadata/good_metaf_final.csv", header = T, stringsAsFactors = F)

metadata <- filter(metadata, !sample %in% c(test_cohort$initial, test_cohort$followUp)) %>% 
  select(Dx_Bin, dx, Age, Gender, BMI, White) %>% mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin), Dx_Bin = gsub("High Risk Normal", "Normal", Dx_Bin))

t1 <- as.data.frame(matrix(nrow = 5, ncol = 4, 
                           dimnames = 
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
                                  coln = c("Normal", "Adenoma", "SRN", "Carcinoma"))))

data_filter1 <- c("normal", "adenoma", "adenoma", "cancer")
data_filter2 <- c("Normal", "Adenoma", "adv_adenoma", "Cancer")

# Generate the summary stats for each group
for(i in 1:length(colnames(t1))){

  t1["Dx_Bin", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100 
}

# Generate SD for applicable metrics
age_sd <- c()
BMI_sd <- c()

for(i in 1:length(colnames(t1))){

  age_sd <- c(age_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
}

# Create combined table
# "u\00B1" is to add a +/- symbol where applicable

ttemptable <- as.data.frame(t(t1)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(age_sd, digits = 2, nmsall = 2)), 
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
                                        White = format(White, digits =2, nsmall = 2))

ft1 <- cbind(Group = c("Normal", "Adenoma", "SRN", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
             ttemptable$BMI, ttemptable$White)

age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")

colnames(ft1) <- c("Group", "n", age_label, "Gender (F%)", bmi_label, "Caucasian (%)")

test <- t(ft1)
colnames(test) <- test["Group", ]
test <- test[-1, ]

kable(test, format = "markdown", align = 'c', padding = 2)

```

### 2017-03-21 code/Run_Figure1.R

@@ -12,6 +12,7 @@ loadLibs(c("dplyr", "ggplot2", "reshape2",
 difference_table_treatment <- read.csv("data/process/tables/difference_table.csv", header = T)
 diff_summary_table <- read.csv("data/process/tables/change_theta_fit_summary.csv", header = T, row.names = 1)
 thetayc_adn_IF <- read.csv("data/process/tables/thetayc_adn_IF.csv", header = T)
+thetayc_srn_IF <- read.csv("data/process/tables/thetayc_srn_IF.csv", header = T)
 thetayc_crc_IF <- read.csv("data/process/tables/thetayc_crc_IF.csv", header = T)
 thetayc_summary_table <- read.csv("data/process/tables/beta_diver_summary.csv", header = T, row.names = 1)
 
@@ -23,13 +24,13 @@ thetayc_summary_table <- read.csv("data/process/tables/beta_diver_summary.csv", 
 differences_graph <- grid.arrange(
   # Difference from bacterial community structure between adenoma and cancer
   ggplot(difference_table_treatment, 
-         aes(factor(dx, levels = c("adenoma", "cancer"), labels = c("Adenoma", "Carcinoma")), 
-             distance, group = 1)) + 
+         aes(factor(Dx_Bin, levels = c("adenoma", "adv_adenoma", "cancer"), 
+                    labels = c("Adenoma", "SRN", "Carcinoma")), distance, group = 1)) + 
     geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
     stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
     stat_summary(fun.y = mean, colour = "black", geom = "line") + 
     scale_color_manual(name = "Lesion Type", 
-                       values = c('#00FFFF', '#0000FF', '#F1BB7B'), 
+                       values = c('#228B22', '#FFD700', '#DC143C'), 
                        breaks = c("adenoma", "adv_adenoma", "cancer"), 
                        labels = c("Adenoma", "SRN", "Carcinoma")) + 
     coord_cartesian(ylim = c(0, 1)) + ylab(expression(paste(theta[YC], " Distance"))) + 
@@ -37,20 +38,22 @@ differences_graph <- grid.arrange(
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
           plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "lines"), 
           legend.title = element_text(face="bold"), 
[…]

### 2017-03-21 code/Run_Figure2.R

@@ -1,74 +1,82 @@
-### Create Figure 6 graph
-### Show that CRC specific OTUs may be the difference between Adenoma and CRC
+### Create Figure 2 graph
+### Show results for initial and follow up samples based on testing
 ## Marc Sze
 
-# Load needed functions
+###Load needed Libraries and functions
 source('code/functions.R')
 
-# Load needed packages
-loadLibs(c("dplyr", "ggplot2", "gridExtra", "scales", "wesanderson"))
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", 
+           "gridExtra", "scales", "wesanderson"))
 
-# Load needed data
-graph_data <- read.csv("data/process/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)
-
-# Add column to jitter values and italicize names
-graph_data <- mutate(graph_data, 
-    j_values = jitter(rel.abund, 3))
+### Load in needed data tables
 
 
-# Create the figure
+adn_red_follow_up_probability <- read.csv("data/process/tables/adn_reduced_follow_up_probability_summary.csv", 
+                                      stringsAsFactors = F, header = T)
+srn_red_follow_up_probability <- read.csv("data/process/tables/srn_reduced_follow_up_probability_summary.csv", 
+                                         stringsAsFactors = F, header = T)
+crc_red_follow_up_probability <- read.csv("data/process/tables/crc_reduced_follow_up_probability_summary.csv", 
+                                          stringsAsFactors = F, header = T) %>% 
+  mutate(followup_crc = ifelse(disease_free == "n", "Yes", "No"))
 
-crc_specific <- grid.arrange(  
-  # Cancer OTUs graph
-  filter(graph_data, Dx_Bin == "cancer") %>% 
-    ggplot(aes(factor(sampleType, levels = c("initial", "followup")), 
-               j_values*100, group = factor(EDRN))) + 
-    geom_line(aes(color = factor(Disease_free, levels = c("n", "y", "unknown")))) + 
-    geom_point(aes(color = factor(Disease_free, levels = c("n", "y", "unknown")))) + 
-    facet_wrap(~Genus, scales = "free_y") + 
-    theme_bw() + ylab("% Relative Abundance") + xlab("") + ggtitle("A") +  
-    scale_colour_manual(name = "Cancer Free\n on Follow Up", 
[…]

### 2017-03-21 code/Run_Figure3.R

@@ -1,5 +1,5 @@
-### Create Figure 3
-### Graph for the lesion model
+### Create Figure 2 graph
+### Show results for initial and follow up samples based on testing
 ## Marc Sze
 
 ###Load needed Libraries and functions
@@ -9,124 +9,70 @@ loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2",
            "gridExtra", "scales", "wesanderson"))
 
 ### Load in needed data tables
-reduced_lesion_model_roc <- read.csv("data/process/tables/reduced_test_data_roc.csv", header = T)
-lesion_MDA_data_summary <- read.csv("data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv", 
-                                    header = T, stringsAsFactors = F)
 
-lesion_MDA_full <- read.csv("data/process/tables/reduced_lesion_model_top_vars_MDA.csv", 
-                            header = T, stringsAsFactors = F)
 
-red_follow_up_probability <- read.csv("data/process/tables/reduced_follow_up_probability_summary.csv", 
+adn_red_follow_up_probability <- read.csv("data/process/tables/adn_reduced_follow_up_probability_summary.csv", 
                                       stringsAsFactors = F, header = T)
-
-tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
-
-# Convert taxa table to a data frame with columns for each taxonomic division
-tax_df <- data.frame(do.call('rbind', strsplit(as.character(tax$Taxonomy), ';')))
-rownames(tax_df) <- rownames(tax)
-colnames(tax_df) <- c("Domain", "Phyla", "Order", "Class", "Family", "Genus")
-tax_df <- as.data.frame(apply(tax_df, 2, function(x) gsub("\\(\\d{2}\\d?\\)", "", x)))
-rm(tax)
-
-# Generate files needed for the MDA graph
-
-# Order the data file from most to least important based on mean MDA
-lesion_MDA_data_summary <- lesion_MDA_data_summary[order(lesion_MDA_data_summary$mean_MDA, decreasing = TRUE), ]
-
-# Pull OTUs that are only in the MDA data
-OTU_IDs <- unique(filter(lesion_MDA_data_summary, variable != "fit_result")[, "variable"])
-select_tax_df <- tax_df[OTU_IDs, ]
-low_tax_ID <- gsub("_", " ", gsub("2", "", gsub("_unclassified", "", createTaxaLabeller(select_tax_df))))
-
[…]

### 2017-03-21 code/Run_Figure4.R

@@ -1,7 +1,6 @@
-### Create Figure 4
-### Graph for the Initial Sample Model
+### Create Figure 3 graph
+### Show results for initial and follow up samples based on testing
 ## Marc Sze
-
 
 ###Load needed Libraries and functions
 source('code/functions.R')

### 2017-03-21 code/Run_FigureS2.R

@@ -1,5 +1,5 @@
-### Figure S5
-### Distribution of P-values from Paired Wilcoxson Analysis
+### Figure S2
+### ROC Curves summary for three models used
 ## Marc Sze
 
 #Load needed code and packages
@@ -8,15 +8,53 @@ source('code/functions.R')
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
 #Read data needed
-diff_table_time <- read.csv("data/process/tables/time_datatable.csv", header = T)
+adn_model_roc <- read.csv("data/process/tables/adn_reduced_test_data_roc.csv", header = T)
+srn_model_roc <- read.csv("data/process/tables/srn_reduced_test_data_roc.csv", header = T)
+crc_model_roc <- read.csv("data/process/tables/crc_reduced_test_data_roc.csv", header = T)
 
-# Plot the change in distance versus time
-time_graph <- ggplot(diff_table_time, aes(x = time, y = distance)) + 
-  geom_point(aes(color = dx), size = 4) + theme_bw() + 
-  scale_color_manual(name = "Lesion Type", values = wes_palette("GrandBudapest"), 
-                     breaks = c("adenoma", "cancer"), labels = c("Adenoma", "Cancer")) + 
-  coord_cartesian(ylim = c(0, 1)) + ylab("Thetayc Distance") + xlab("Time from Initial (Days)") + 
-  theme(axis.title = element_text(face="bold"), legend.title = element_text(face="bold"))
 
-ggsave(file = "results/figures/FigureS2.pdf", time_graph, 
+# Create the graph
+
+rocs_graph <- grid.arrange(
+  # Adenoma ROC curve information
+  filter(adn_model_roc, run != "middle_roc" & run != "full_roc") %>% 
+    ggplot(aes(x = sensitivities, y = specificities)) + 
+    geom_polygon(data = filter(adn_model_roc, run != "full_roc"), alpha = 0.5, fill = '#76EE00') + 
+    geom_line(data = filter(adn_model_roc, run != "full_roc"), 
+              aes(group = run), size = 1.25, color = '#76EE00') + 
+    geom_line(data = filter(adn_model_roc, run == "full_roc"), 
+              size = 1.5, color = '#006400') + 
+    scale_x_continuous(trans = "reverse") + theme_bw() + 
[…]

### 2017-03-21 code/Run_treatment_common_otus.R

@@ -9,15 +9,16 @@ source('code/functions.R')
 loadLibs(c("dplyr", "tidyr", "ggplot2", "gridExtra"))
 
 # Read in data tables
-chemo_rad_stats_summary <- read.csv('data/process/tables/probs_chemo_rad_pvalue_summary.csv', 
-                              header = T, stringsAsFactors = F) %>% rename(comparison = X)
+chemo_rad_stats_summary <- read.csv('data/process/tables/crc_probs_chemo_rad_pvalue_summary.csv', 
+                              header = T, stringsAsFactors = F, row.names = 1)
 
 good_metaf <- read.csv('data/process/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
 
-common_vars_summary <- read.csv('data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', 
+common_vars_summary <- read.csv('data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', 
                                 header = T, row.names = 1, stringsAsFactors = F)
 
-chemo_rad_summary <- read.csv('data/process/tables/chemo_rad_summary.csv', header = T, stringsAsFactors = F)
+chemo_rad_summary <- read.csv('data/process/tables/crc_chemo_rad_summary.csv', 
+                              header = T, stringsAsFactors = F, row.names = 1)
 
 shared <- read.delim('data/process/final.shared', header = T, stringsAsFactors = F) %>% 
   select(-label, -numOtus) %>% mutate(Group = as.character(Group))
@@ -36,17 +37,46 @@ shared <- filter(shared, Group %in% samples_to_keep) %>%
 
 # Select only OTUs that were common
 shared <- shared %>% select(one_of(common_vars_summary$otu)) %>% 
-  mutate(sampleType = c(rep("initial", length(good_metaf$initial)), rep("followups", length(good_metaf$followUp))))
+  mutate(sampleType = c(rep("initial", length(good_metaf$initial)), rep("followups", length(good_metaf$followUp))), 
+         Dx_Bin = rep(good_metaf$Dx_Bin, 2), 
+         chemo = rep(good_metaf$chemo_received, 2), 
+         rads = rep(good_metaf$radiation_received, 2)) %>% 
+  filter(Dx_Bin == "cancer")
 
-# Calculate P-values and write out to csv
[…]

### 2017-03-21 submission/supplemental_outline_20161024.Rmd

@@
 chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv', 
-                              header = T, stringsAsFactors = F)
+                              header = T, stringsAsFactors = F) %>% filter(X != "chemo_v_rads_IF") %>% 
+  separate(X, c("comparison", "variable"), sep = "_") %>% 
+  mutate(comparison = ifelse(comparison == "rads" | comparison == "rad", invisible("Radiation"), invisible("Chemotherapy")), 
+         variable = c(rep("sobs", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4)))
 
-variable <- c(rep("sobs", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4))
-comparison <- rep(c("Chemotherapy", "Radiation"), 6)
-
-st3 <- cbind(variable, comparison, format(chemo_rad_summary$chemo_rad_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$removal_only_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$cr_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$ro_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$pvalue, scientific = TRUE, digits = 2), format(chemo_rad_summary$bh, scientific = TRUE, digits = 2))
+st3 <- cbind(chemo_rad_summary$variable, chemo_rad_summary$comparison, format(chemo_rad_summary$chemo_rad_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$removal_only_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$cr_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$ro_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$pvalue, scientific = TRUE, digits = 2), format(chemo_rad_summary$bh, scientific = TRUE, digits = 2))
 

### 2017-03-23 results/tables/Table2.Rmd
changes to manuscript up to results section

---
title: ""
output: pdf_document
---

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../../code/functions.R')

loadLibs(c("tidyr", "dplyr", "scales", "knitr"))
```

**Table 2: Demographic Data of Pre and Post Treatment Cohort**

```{r table2, echo=FALSE, message=FALSE, warning=FALSE}

test_cohort <- read.csv("../../data/process/mod_metadata/good_metaf_final.csv", header = T, stringsAsFactors = F) %>% 
  select(Dx_Bin, dx, Age, Gender, BMI, White)

t2 <- as.data.frame(matrix(nrow = 5, ncol = 3, 
                           dimnames = 
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
                                  coln = c("Adenoma", "SRN", "Carcinoma"))))

data_filter1 <- c("adenoma", "adenoma", "cancer")
data_filter2 <- c("adenoma", "adv_adenoma", "cancer")

# Generate the summary stats for each group
for(i in 1:length(colnames(t2))){

  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100 
}

# Generate SD for applicable metrics
age_sd <- c()
BMI_sd <- c()

for(i in 1:length(colnames(t2))){

  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
}

# Create combined table
# "u\00B1" is to add a +/- symbol where applicable

ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(age_sd, digits = 2, nmsall = 2)), 
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
                                        White = format(White, digits =2, nsmall = 2))

ft1 <- cbind(Group = c("Adenoma", "SRN", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
             ttemptable$BMI, ttemptable$White)

ft1 <- cbind(Group = c("Normal", "Adenoma", "SRN", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
             ttemptable$BMI, ttemptable$White)

age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")

colnames(ft1) <- c("Group", "n", age_label, "Gender (F%)", bmi_label, "Caucasian (%)")

test <- t(ft1)
colnames(test) <- test["Group", ]
test <- test[-1, ]

kable(test, format = "markdown", align = 'c', padding = 2)
```

### 2017-03-23 code/Run_treatment_common_otus.R

@@ -6,7 +6,7 @@
 ###Load needed Libraries and functions
 source('code/functions.R')
 
-loadLibs(c("dplyr", "tidyr", "ggplot2", "gridExtra"))
+loadLibs(c("dplyr", "tidyr"))
 
 # Read in data tables
 chemo_rad_stats_summary <- read.csv('data/process/tables/crc_probs_chemo_rad_pvalue_summary.csv', 
@@ -36,17 +36,17 @@ shared <- filter(shared, Group %in% samples_to_keep) %>%
   slice(match(samples_to_keep, Group))
 
 # Select only OTUs that were common
-shared <- shared %>% select(one_of(common_vars_summary$otu)) %>% 
+crc_shared <- shared %>% select(one_of(common_vars_summary$otu)) %>% 
   mutate(sampleType = c(rep("initial", length(good_metaf$initial)), rep("followups", length(good_metaf$followUp))), 
          Dx_Bin = rep(good_metaf$Dx_Bin, 2), 
          chemo = rep(good_metaf$chemo_received, 2), 
          rads = rep(good_metaf$radiation_received, 2)) %>% 
   filter(Dx_Bin == "cancer")
 
-test_data <- ((filter(shared, sampleType == "initial") %>% select(contains("Otu0"))) - 
-  (filter(shared, sampleType == "followups") %>% select(contains("Otu0")))) %>% 
-  mutate(chemo = shared$chemo[shared$sampleType == "initial"], 
-         rads = shared$rads[shared$sampleType == "initial"])
+test_data <- ((filter(crc_shared, sampleType == "initial") %>% select(contains("Otu0"))) - 
+  (filter(crc_shared, sampleType == "followups") %>% select(contains("Otu0")))) %>% 
+  mutate(chemo = crc_shared$chemo[crc_shared$sampleType == "initial"], 
+         rads = crc_shared$rads[crc_shared$sampleType == "initial"])
 
 # Create lowest ID labels
 common_lowest_IDs <- common_vars_summary$lowest_ID
@@ -56,14 +56,17 @@ for(i in 1:length(common_lowest_IDs)){
   table_labels <- c(table_labels, rep(common_lowest_IDs[i], length(good_metaf$EDRN)))
 }
 
+
 # Add a lowest taxa ID to test data table
 graph_ids <- c()
 otus <- unique(common_vars_summary$otu)
 
 for(i in 1:length(otus)){
   
[…]

### 2017-03-23 submission/manuscript_outline_20161024.Rmd

--- 2017-03-14 submission/manuscript_outline_20161024.Rmd
@@
 
-# Main CRC bacteria differences between Adenoma and Carcinoma
-main_crc_bugs_summary <- read.csv('../data/process/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
-crc_bug_info <- read.csv("../data/process/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)
+# Adenoma Model 
+adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
+adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
+adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)
 
-# Lesion Model 
-lesion_top10_varsMDA <- read.csv('../data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T)
-lesion_imp_vars <- read.csv('../data/process/tables/rf_otu_tax.csv', header = T, row.names = 1)
-lesion_AUC_run_summary <- read.csv('../data/process/tables/Reduced_ROC_model_summary.csv', header = T)
+# SRN Model
+srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T)
+srn_imp_vars <- read.csv('../data/process/tables/srn_rf_otu_tax.csv', header = T, row.names = 1)
+srn_AUC_run_summary <- read.csv('../data/process/tables/srn_Reduced_ROC_model_summary.csv', header = T)
 
-# IF Model
-IF_top10_varsMDA <- read.csv('../data/process/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T)
-IF_imp_vars <- read.csv('../data/process/tables/if_rf_otu_tax.csv', header = T, row.names = 1)
-IF_AUC_run_summary <- read.csv('../data/process/tables/reduced_IF_ROC_model_summary.csv', header = T)
+# CRC Model
+crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
+crc_imp_vars <- read.csv('../data/process/tables/crc_rf_otu_tax.csv', header = T, row.names = 1)
+crc_AUC_run_summary <- read.csv('../data/process/tables/crc_Reduced_ROC_model_summary.csv', header = T)
 
 # All model information
-all_probs_paired_wilcox <- read.csv('../data/process/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
-all_pred_stat_summary <- read.csv('../data/process/tables/all_models_summary_info.csv', header = T, stringsAsFactors = F)
-all_confu_matrix <- read.csv('../data/process/tables/all_models_confusion_summary.csv', header = T, stringsAsFactors = F)
+all_probs_paired_wilcox <- read.csv('../data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
+all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F)
+all_confu_matrix <- read.csv('../data/process/tables/all_crc_srn_adn_models_confusion_summary.csv', header = T, stringsAsFactors = F)
 
 # Common Variable Info
-common_vars_summary <- read.csv('../data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
+common_vars_summary <- read.csv('../data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
 
 #Therapy Info
-chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
+chemo_rad_summary <- read.csv('../data/process/tables/crc_probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
 chemo_rad_common_vars_summary <- read.csv('../data/process/tables/chemo_rads_treatment_pvalue_summary.csv', header = T, stringsAsFactors = F)
+
+comb_adn_surg_summary <- read.csv('../data/process/tables/adn_combined_probs_surgery_pvalue_summary.csv', header = T, stringsAsFactors = F, row.names = 1)
+
+comb_adn_surg_common_vars_summary <- read.csv('../data/process/tables/all_adn_surg_pvalue_summary.csv', header = T, stringsAsFactors = F)
 
@@
 **Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.
-This study tested the hypothesis that treatment for adenoma or carcinoma result in changes to the bacterial community.
-Specifically, we tried to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).
+This study tested the hypothesis that treatment for adenoma or carcinoma results in changes to the bacterial community.
+Specifically, we tried to identify components within the community that were different before and after treatment of adenoma, advanced adenoma (Screen Relevant Neoplasia (SRN)), and carcinoma.
 
-**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).
-Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).
-A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had no difference in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that the lesion associated community persists after treatment.
-A treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that there are members within the community that repond to treatment.
-The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.
-A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g.
-*Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*).
+**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma or SRN cases *(P-value )*.
+But there was a trend for increasingly less community similarity between initial and follow up samples from adenoma to SRN to carcinoma.
+Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups *(P-value )*.
+Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability *(P-value)*; providing further evidence that treatment has the greatest effect in those with carcinoma.
+The adenoma model used a total of *model n* OTUs, the SRN model used a total of *model n* OTUs, and the carcinoma model used a total of *model n* OTUs.
+A total of *common n* OTUs were common to both models with many classifying to commensal bacteria (e.g.
+*Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
+Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma *(Pvalue )*
 
 **Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.
-Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.
-Commensal bacteria were crucial for accurate model classification, suggesting that these bacteria may be important to initial polyp formation and transition to carcinoma.
+Individuals with carcinoma had more drastic differences to the overall community then those with adenoma.
+Common OTUs to all models were overwhelmingly from commensal bacteria, suggesting that these bacteria may be important to initial polyp formation, development of advanced adenoma, and transition to carcinoma.
 
@@
 
-Recent studies in humans have shown that both the bacterial community and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].
-Further, Dejea, et al.
-observed that bacterial communities are altered between normal and tumor tissue [@dejea_microbiota_2014].
+Recent studies in humans have shown that both the microbiome and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].
+Further, bacterial communities have been observed to be altered between normal and tumor tissue [@dejea_microbiota_2014].
 Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.
@@
 Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].
-They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).
+They provide very little information as to whether these communities change and rebound towards normal after successful treatment of adenoma, advanced adenoma (Screen Relevent Neoplasia (SRN)), or carcinoma.
 
-In this study, we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.
-We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).
-We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].
-We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up treatment samples.
-Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether they were changing after treatment and whether it was towards a normal community.
-We also investigated the two models for similar important OTUs to identify which were important for classifying both lesion versus normal and initial versus follow up treatment samples.
-Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionally treatments significantly lowered any of the metrics examined versus those who received only lesion removal.
-This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
+Providing an answer to this question is important because it has far reaching implications on both how the bacterial community causes the formation of more polyps [@zackular_manipulation_2016; @arthur_microbial_2014] and the ability to be able to use the microbiome as a predictive screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
+Understanding polyp formation and transition to SRN and then carinoma is crucial to being able to understand how to prevent CRC occurence.
+Response of the community to treatment is also eqaully important to predictive models designed for screening purposes since an unresponsive community would provide little additional information for important events, such as recurrence [@hassan_efficacy_2016].
+
+Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes to the microbiome in those with adenoma, SRN, and carincoma.
+First, we assessed differences between initial and follow up samples in adenoma, SRN, or carinoma using alpha or beta diversity metrics.
+Second, we explored whether models built to classify adenoma, SRN, or carcinoma versus normal were able to identify specific community members that differed between initial and follow up.
+We also used these models to assess whether changes in the community were toward a more normal micorbiome.
+Finally, we assessed both whether surgery for adenomas and SRN provided larger community changes or whether chemotherapy or radiation provided additive changes to the microbiome over surgerical resection.
+This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists after such interventions.
 
@@
 ***The Bacterial Community:*** 
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`) had any broad differences between their initial and follow up samples.
-We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1A].
-The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
-Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).
-There was no difference between initial and follow up samples when investigating alpha diversity metrics for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].
-Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma *(n = )*, SRN *(n = )*, or carcinoma *(n = )* had any broad differences between their initial and follow up samples.
+We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma *(P-value 0.001)* [Figure 1A].
+Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma *(mean and SD)* to SRN *(mean and SD)* to carcinoma *(mean and SD)* [Figure 1A].
+The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] *(PERMANOVA  0.05)*, SRN [FIgure 1C] *(PERMANOVA 0.05)*, and carcinoma [Figure 1C] *(PERMANOVA  0.05)*.
+Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity *(PERMANOVA  0.05)*.
+There was no difference between initial and follow up samples when investigating alpha diversity metrics for adenoma, SRN, or carcinoma for any metric tested [Table S1].
+Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for adenoma, SRN, or carcinoma only [Figure S1].
 
-***Carcinoma Associated Genera:*** 
-For carcinoma [Figure S2A] (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05 & main_crc_bugs_summary["porp", "crc_BH"] < 0.05 & main_crc_bugs_summary["pept", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) but not adenoma [Figure S2B] (P-value `r ifelse(main_crc_bugs_summary["parv", "adn_BH"] < 0.05 & main_crc_bugs_summary["porp", "adn_BH"] < 0.05 & main_crc_bugs_summary["pept", "adn_BH"] < 0.05, paste("<"), paste(">"))` 0.05) there was a significant decrease in *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* between initial and follow up samples [Figure S2].
-*Fusobacterium* for both adenoma and carcinoma had no significant differences between initial and follow up samples [FIgure S2] (P-value `r ifelse(main_crc_bugs_summary["fn", "crc_BH"] < 0.05 & main_crc_bugs_summary["fn", "adn_BH"] < 0.05, paste("<"), paste(">"))` 0.05).
-Although there were significant differences, only a small percentage of those with adenoma or carcinoma were positive for any of these specific genera [Figure S2].
+***Adenoma Model***
+The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2A].
+There was a total of *blank* OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A].
+There was a significant difference between the actual and model predicted group stratification calls *(P-value)*.
+There was no significant decrease in the positive probability of adenoma between initial and follow up samples *(P_value)* [Figure 2A].
 
-***The Lesion Model:***
-We tested whether follow up samples were more normal than initial samples by building a model based solely on OTUs alone that classified lesion versus normal.
-This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.
-The ROC curve for the final lesion model used was within the observed range of the 100-different test set AUC iterations [Figure 2A].
-There was a total of `r length(lesion_top10_varsMDA$variable)` OTUs that were used in this model [Figure 2B].
-*`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [Figure 2B].
-Heavily represented genera within the model included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
+***SRN Model***
+The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2B].
+There was a total of *blank* OTUs in the SRN model [Figure S3B].
+Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal.
+Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.
+There was no significant decrease in the positive probability of SRN between initial and follow up samples *(P_value)* [Figure 2B].
 
-If there was movement towards a more normal community after treatment of lesion, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.
-We observed no such decrease in positive probability for either those with adenoma or carcinoma [Figure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).
-We also observed a significant difference between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
-However, this model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].
+***Carcinoma Model***
+The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2C].
+Interestingly, the AUCs improved from adenoma to SRN to carcinoma [Figure 2].
+There was a total of *blank* OTUs in the carinoma model [Figure S3C].
+Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* appear to be important for carcinoma classification [Figure S3C].
+Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.
+There was a significant decrease in the positive probability of carcinoma between initial and follow up samples *(P_value)* [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.
 
-***The Treatment Model:***
-To test if there were differences that could be used to classify initial and follow up treatment samples we built a model to classify initial samples.
-This treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.
-after 100 iterations of 20 repeated 10-fold cross validations.
-The test set AUC range for this model performed better than the training set AUCs.
-There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3A].
-There was a total of `r length(IF_top10_varsMDA$variable)` OTUs that were used for this model [Figure 3B].
-The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].
-Similar to the lesion model, heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.
+***Adenoma, SRN, and Carcinoma Common OTUs***
+We next wanted to know what predictors within the adenoma, SRN, and carcinoma models were similar to each other.
+The main purpose was to identify which OTUs could be important at all three stages of disease.
+When we compared the three different models with each other there were a total of *number* common OTUs.
+Some of the most common taxonomic identifications belonged to *one*, *two*, *three*, *four*, *five*, *six*, and *seven*.
+These along with the vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].
 
-If there were changes between initial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.
-The is what we observed for the treatment model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-When we separated lesion into adenoma and carcinoma there was a decrease in positive probability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [Figure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).
-For this model, there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).
+***Treatment Affects on Community***
+After observing these these changes from treatment we assessed whether chemotherapy or radiation, in the carcinoma group, and surgery, in the adenoma group, impacted the observed results.
+In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone [Table S3].
+For the adenoma group there was no difference between those that received surgery and those that did not [Table S4].
+For surgery adenoma and SRN were combined due to the low number of surgery occurances.
+There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups *(P-value)*.
+This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.
 
-***Common OTUs to both Models:***
-We next wanted to know what predictors within the lesion model were also in the initial sample model.
-The main purpose was to identify which OTUs could be important in lesion formation and were impacted by treatment.
-When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.
-Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000061")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000037")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000006")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000015")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000167")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.
-These along withthe vast majority of the OTUs that were common between models had classifications to bacteria typically thought of as commensal [Table S2].
-
-***Chemotherapy and Radiation Differences:*** 
-After observing these changes due to treatment we assessed whether chemotherapy or radiation impacted the observed results.
-Only the treatment model had a significant change in positive probability for those treated with chemotherapy or radiation versus those who received neither [Figure 4A & Table S3] (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
-Additionally, there was no difference between the change in positive probability for chemotherapy and radiation therapy (Uncorrected P-value `r ifelse(chemo_rad_summary['chemo_v_rads_IF', 'pvalue'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).
-Using the `r length(common_vars_summary$otu)` common OTUs we tested how chemotherapy or radiation affects important members of the community.
-For both chemotherapy and radiation, the directionality of change of these OTUs were similar [Figure 4B & 4C].
-After multiple comparison correction, only *`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "lowest_ID"]`* (`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "otu"]`) remained significantly different between those who did and did not receive chemotherapy.
-This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to effects on the bacterial community.
+#Need to make a table with the adenoma results
 

### 2017-03-23 submission/supplemental_outline_20161024.Rmd

@@
 
-**Table S2: Summary of OTUs in both Lesion and Initial/Follow Up Models**
+**Table S2: Summary of OTUs in Adenoma, SRN, and Carcinoma Models**
 
@@
 
-common_vars_summary <- read.csv('../data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
+common_vars_summary <- read.csv('../data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)
 
-lesion_top10_varsMDA <- read.csv('../data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
-IF_top10_varsMDA <- read.csv('../data/process/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+
+srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+
+crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
 
@@
 # Create MDA vectors
-lesion_MDA <- filter(lesion_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
-
-initial_MDA <- filter(IF_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
+adn_MDA <- filter(adn_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
+srn_MDA <- filter(srn_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
+crc_MDA <- filter(crc_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
 
@@
   mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
-  mutate(lesion_MDA = lesion_MDA$mean_MDA) %>% 
-  mutate(initial_MDA = initial_MDA$mean_MDA) %>% arrange(desc(lesion_MDA))
+  mutate(adn_MDA = adn_MDA$mean_MDA, srn_MDA = srn_MDA$mean_MDA, crc_MDA = crc_MDA$mean_MDA) %>% 
+  arrange(desc(adn_MDA), desc(srn_MDA), desc(crc_MDA))
 
-colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Lesion Model MDA", "Initial Model MDA")
+colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Adenoma Model MDA", "SRN Model MDA", "Carcinoma Model MDA")
 

### 2017-03-24 submission/manuscript_outline_20161024.Rmd
add discussion to method changes

@@
 This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
-Here we describe how the bacterial community changes after treatment and which OTUs may be most important to lesion classification.
-Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes.
-This suggest that members within the commensal community may be the first that change during CRC pathogenesis.
-These changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].
+Here we show that the bacterial community changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide an additive change.
+Although some of the important OTUs classified to genera from bacteria considered the usual suspects (e.g.
+*Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.
+The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, SRN, and carcinoma models.
+This suggests that members within the commensal community may be the first that change during CRC pathogenesis.
+These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].
 
-Unlike previous studies on the microbiome and CRC, ours focuses on identifying commonalities within both adenoma and carcinoma groups before and after treatment.
-Although there were differences for genera associated with specific bacterium linked with CRC [Figure S2].
-These changes were not consistent across lesion.
-Instead, most of our results provide support that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria [Figure 2-3 & Table S3].
-Additionally, treatment with chemotherapy or radiation changed these bacteria and provided a larger decrease in positive probability then removal of lesion.
+Unlike previous studies on the microbiome and CRC, ours focuses both on identifying commonalities and differences within adenoma, SRN, and carcinoma groups before and after treatment.
+Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3].
+The majority of important OTUs taxonmically classified to commensal bacteria [Figure S3].
+Although these changes may be subtle, due to the lack of signifcant difference in the bacterial community before and after treatment in adenoma and SRN [Figure 1], they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
 
-Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.
-There are several reasons why this may have occurred.
-First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.
-Second, is that our Random Forest models were able to gather the same information from other OTUs.
-Third, is that changes in commensal bacteria are the common differences across adenoma and carcinoma.
-Finally, it is also possible that all these explanations could have played a role.
-Our observations though, would suggest that an individual's resident bacteria have a large role to play in polyp formation and could change in a way that allows predictive models to lower the positive probability of lesion or initial sample before treatment [Figure 2 & 3].
-It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after treatment.
-It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after treatment [Figure S2].
-
-Many of the common OTUs that we identified taxnomically classified to potential butyrate producers [Table S2].
-Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.
-Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].
+Many of the common OTUs that we identified taxnomically classified to potential butyrate producers (e.g.
+*Clostridiales*, *Roseburia*,  and *Anaerostipes*)[Table S2].
+Other OTUs classified to bacteria that are inhibited by polyphenols (e.g.
+*Bacteroides*).
+Both butyrate and polyphenols are thought to be protective against cancer, in part by reducing inflammation [@louis_gut_2014].
 These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.
@@
 
-A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).
+A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma *(uncorrected P-value 0.05)*, with time passed being less for adenoma *( days)* and SRN *( days)* than carcinoma *( days)*.
 These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time.
 Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1C].
-This confounding though would not affect the observations where these individuals were grouped together [Figures 2-4].
+However, there are two reasons that this may not be the case.
+First, the SRN group did not have a significant difference and the collection time of their follow up sample was less than the adenoma group.
+Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].
 
-Another limitation was that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.
-This information would help to strengthen the case for this model keeping several individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.
+Another limitation was that we do not know whether individuals who were still classified as positive by the carcinoma model eventually had a subsequent CRC recurrence.
+This information would help to strengthen the case for this model keeping numerous individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having carcinoma.
 This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.
@@
 
-Interestingly, within the treatment model the test data performed better than the training data.
-This may have occurred because the training AUC determined from 20 repeated 10-fold cross validation removed samples at random and did not consider that they were matched samples.
-Another potential reason is that the model itself may have been over-fit since the total number of samples was not that large.
-However, the lesion model did not suffer from these discrepancies.
-Further independent studies need to be carried out to verify our findings on lesion and treatment changes due to these limitations.
-
-Despite these shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma and carcinoma treatment.
-Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of lesion or initial sample after treatment provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with both treatment and lesion.
+Despite these shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma, SRN or carcinoma treatment.
+Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of carcinoma after treatment provides evidence that there are specific signatures, attributable to both inflammtory and resident commensal organisms, associated with treatment.
 Our data provides evidence that commensal bacteria may be important in the development of polyps and potentially the transition from adenoma to carcinoma.
@@
 The bacterial community has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013].
-Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion.
+Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].
+Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 2].
 This study was approved by the University of Michigan Institutional Review Board.

### 2017-03-24 code/Run_FigureS1.R

@@ -10,12 +10,12 @@ loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesa
 #Read data needed
 paired_table <- read.csv("data/process/tables/OTU_paired_wilcoxson_test.csv", header = T, stringsAsFactors = F)
 
-label_names <- c("Lesion", "Adenoma Only", "CRC Only")
-names(label_names) <- c("ALL", "adn", "crc")
+label_names <- c("Adenoma", "SRN", "Carcinoma")
+names(label_names) <- c("adn", "srn", "crc")
 
 pvalue_distribution_paired <- ggplot(paired_table, aes(x=BH_corrected)) + geom_histogram() + 
   geom_vline(aes(xintercept = 0.05, color = "red"), linetype = 2, size = 1) + 
-  facet_wrap(~factor(analysis, levels = c("ALL", "adn", "crc")), 
+  facet_wrap(~factor(analysis, levels = c("adn", "srn", "crc")), 
              labeller = as_labeller(label_names)) + theme_bw() + 
   xlab("Benjamini-Hochberg Corrected P-values") + ylab("Counts") + 
   theme(legend.position = "none")

### 2017-03-24 submission/manuscript_outline_20161024.Rmd
wrapping up mansucript text

@@
 There was a significant decrease in the positive probability of carcinoma between initial and follow up samples *(P_value)* [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.
+The one indivdiual still positive for carcinoma after treatment had an increase in carcinoma positive proability on follow up [Figure 2C].
 
@@
 This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.
-
-#Need to make a table with the adenoma results
 
@@
 
-**Figure 1: General Differences between the Adenoma and Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+**Figure 1: General Differences between Adenoma, SRN, and Carcinoma Groups After Treatment.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
 Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).
@@
 
-**Figure 2: The Lesion Model.** A) ROC curve: The shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
-B) Summary of Important Variables.
-MDA of the most important variables in the lesion model.
-The black point represents the mean and the different colors are the values of each different run up to 100.
-C) Positive probability change from initial to follow up sample in those with carcinoma.
-D) Positive probability change from initial to follow up sample if those with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
+**Figure 2: Treatment Response Based on Models Built for Adenoma, SRN, or Carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
+B) Positive probability change from initial to follow up sample in those with SRN.
+C) Positive probability change from initial to follow up sample in those with carcinoma..
 
-**Figure 3: The Treatment Model.**  A) ROC curve: The shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.
-B) Summary of Important Variables.
-MDA of the most important variables in the initial sample model.
-The black point represents the mean and the different colors are the values of each different run up to 100.
-C) Positive probability change from initial to follow up sample in those with carcinoma.
+**Figure 3: The Treatment Model.**  B)   C) Positive probability change from initial to follow up sample in those with carcinoma.
 D) Positive probability change from initial to follow up sample of those with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
-
-**Figure 4: Chemotherapy and Radiation Therapy Effects on Community.**  A) Treatment model initial sample positive probability reduction for chemotherapy.
-B) Treatment model initial sample positive probability reduction for radiation therapy.
-C) Common OTUs based on whether chemotherapy was received or not.
-The * denotes a P-value < 0.05 after multiple comparison correction.
-D) Common OTUs based on whether radiation therapy was received or not.
 
@@
 
-**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of All OTUs for Initial versus Follow Up**
+**Table 1: Demographic Data of Training Cohort**
 
-**Figure S2: Previously Associated CRC Bacteria in Initial and Follow Up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), *Porphyromonas* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`), *Peptostreptococcus* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`).
-B) Adenoma initial and follow up samples.
-There were no significant differences between initial and follow up (P-value = `r format(main_crc_bugs_summary["porp", "adn_BH"], digits = 3)`.
+**Table 2: Demographic Data of Pre and Post Treatment Cohort**
+
+\newpage
+
+**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of All OTUs Before and After Treatment**
+
+**Figure S2: ROC Curves of the Adenoma, SRN, and Carcinoma Models.**  A) Adenoma ROC curve: The light greenshaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+B) SRN ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
+C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.
+
+**Figure S3: Summary of Important Variables for the Adenoma, SRN, and Carcinoma Models.**  A) MDA of the most important variables in the adenoma model.
+The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
+B) Summary of Important Variables in the SRN model.
+MDA of the most important variables in the SRN model.
+The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs.
+C) MDA of the most important variables in the carcinoma model.
+The dark red point represents the mean and the lighter redpoints are the value of each of the 100 different runs.
 

### 2017-03-24 submission/manuscript_outline_20161024.tex
wrapping up mansucript text

@@
 microbiome and CRC. This study tested the hypothesis that treatment for
-adenoma or carcinoma result in changes to the bacterial community.
+adenoma or carcinoma results in changes to the bacterial community.
 Specifically, we tried to identify components within the community that
-were different before and after removal of lesion (adenoma or
-carcinoma).
+were different before and after treatment of adenoma, advanced adenoma
+(Screen Relevant Neoplasia (SRN)), and carcinoma.
 
 \textbf{Results:} There was a larger change in the bacterial community
-in response to treatment for carcinoma versus adenoma cases (P-value
-\textless{} 0.05). Yet no difference was found in the relative abundance
-of any OTU before and after treatment for adenoma or carcinoma groups
-(P-value \textgreater{} 0.05). A lesion model had an AUC range of 0.692
-- 0.761 and follow up samples had no difference in the positive
-probability of lesion versus initial samples (P-value \textgreater{}
-0.05); suggesting that the lesion associated community persists after
-treatment. A treatment model had an AUC range of 0.657 - 0.796 and had a
-decrease positive probability for the follow up samples to be an initial
-sample (P-value \textless{} 0.05); suggesting that there are members
-within the community that repond to treatment. The lesion model used a
-total of 54 variables while the initial sample model used a total of 70
-variables. A total of 32 OTUs were common to both models with many
-classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
-\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
-\emph{Ruminococcus}).
+in response to treatment for carcinoma versus adenoma or SRN cases
+\emph{(P-value )}. But there was a trend for increasingly less community
+similarity between initial and follow up samples from adenoma to SRN to
+carcinoma. Despite this, no difference was found in the relative
+abundance of any OTU before and after treatment for adenoma, SRN, or
+carcinoma groups \emph{(P-value )}. Using Random Forest models to assess
+whether changes in follow up samples were towards a normal community
+showed that only those with carcinoma had a significant decrease in
+positive probability \emph{(P-value)}; providing further evidence that
+treatment has the greatest effect in those with carcinoma. The adenoma
+model used a total of \emph{model n} OTUs, the SRN model used a total of
+\emph{model n} OTUs, and the carcinoma model used a total of \emph{model
+n} OTUs. A total of \emph{common n} OTUs were common to both models with
+many classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
+\emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
+\emph{Dorea}). Both chemotherapy and radiation did not provide any
+additional changes to the bacterial community in those treated for
+carcinoma \emph{(Pvalue )}
 
@@
 the bacterial community changes after treatment. Individuals with
-carcinoma have more drastic differences to the overall community then
-those with adenoma. Commensal bacteria were crucial for accurate model
-classification, suggesting that these bacteria may be important to
-initial polyp formation and transition to carcinoma.
+carcinoma had more drastic differences to the overall community then
+those with adenoma. Common OTUs to all models were overwhelmingly from
+commensal bacteria, suggesting that these bacteria may be important to
+initial polyp formation, development of advanced adenoma, and transition
+to carcinoma.
 
@@
 
-Recent studies in humans have shown that both the bacterial community
-and specific members within it correlate with CRC pathogenesis
-{[}3,4{]}. Further, Dejea, et al. observed that bacterial communities
-are altered between normal and tumor tissue {[}5{]}. Mouse models of CRC
-have further demonstrated the importance of the microbiome, both on a
-community {[}3,6{]} and species level {[}4{]}, for tumorgenesis.
-Collectively, these studies provide a tantalizing link between our gut
-bacteria and CRC and suggest that biomarkers using our microbes could be
-developed. Indeed, builidng models using 16S rRNA gene sequencing along
-with clinical tests such as Fecal Immunoglobulin Test (FIT) result in
-good predictions of CRC {[}7,8{]}. Although these studies show how our
-gut bacteria can impact CRC progression via a changed community or
-invasion by more inflammatory bacteria {[}9{]}. They provide very little
-information as to whether these changed communities rebound after
-successful treatment of lesion (adenoma or carcinoma).
+Recent studies in humans have shown that both the microbiome and
+specific members within it correlate with CRC pathogenesis {[}3,4{]}.
+Further, bacterial communities have been observed to be altered between
+normal and tumor tissue {[}5{]}. Mouse models of CRC have further
+demonstrated the importance of the microbiome, both on a community
+{[}3,6{]} and species level {[}4{]}, for tumorgenesis. Collectively,
+these studies provide a tantalizing link between our gut bacteria and
+CRC and suggest that biomarkers using our microbes could be developed.
+Indeed, builidng models using 16S rRNA gene sequencing along with
+clinical tests such as Fecal Immunoglobulin Test (FIT) result in good
+predictions of CRC {[}7,8{]}. Although these studies show how our gut
+bacteria can impact CRC progression via a changed community or invasion
+by more inflammatory bacteria {[}9{]}. They provide very little
+information as to whether these communities change and rebound towards
+normal after successful treatment of adenoma, advanced adenoma (Screen
+Relevent Neoplasia (SRN)), or carcinoma.
 
-In this study, we tested the hypothesis that there are detectable
-changes to the bacterial community between pre- (initial) and post-
-(follow up) treatment of lesion. We analyzed changes in alpha and beta
-diversity as well as the relative abundance of specific Operational
-Taxonomic Units (OTUs). We then investigated several important genera
-from oral microbes that have been suggested to be important in CRC
-pathogenesis {[}9{]}. We next utilized Random Forest to build two
-models: The first was built to classify lesion versus non-lesion
-(normal) while the second was built to classify initial versus follow up
-treatment samples. Subsequent observations on how these models and OTUs
-within them performed before and after treatment helped inform us as to
-whether they were changing after treatment and whether it was towards a
-normal community. We also investigated the two models for similar
-important OTUs to identify which were important for classifying both
-lesion versus normal and initial versus follow up treatment samples.
-Finally, since treatment also included chemotherapy and radiation
-therapy along with lesion removal, we tested if these additionally
-treatments significantly lowered any of the metrics examined versus
-those who received only lesion removal. This study helps to provide
-evidence as to whether treatment can influence the community and if the
-CRC microbiome, identified in previous studies, persists.
+Providing an answer to this question is important because it has far
+reaching implications on both how the bacterial community causes the
+formation of more polyps {[}3,4{]} and the ability to be able to use the
+microbiome as a predictive screening tool {[}7,8{]}. Understanding polyp
+formation and transition to SRN and then carinoma is crucial to being
+able to understand how to prevent CRC occurence. Response of the
+community to treatment is also eqaully important to predictive models
+designed for screening purposes since an unresponsive community would
+provide little additional information for important events, such as
+recurrence {[}10{]}.
+
+Using pre- (initial) and post- (follow up) treatment samples we tested
+the hypothesis that treatment causes detectable changes to the
+microbiome in those with adenoma, SRN, and carincoma. First, we assessed
+differences between initial and follow up samples in adenoma, SRN, or
+carinoma using alpha or beta diversity metrics. Second, we explored
+whether models built to classify adenoma, SRN, or carcinoma versus
+normal were able to identify specific community members that differed
+between initial and follow up. We also used these models to assess
+whether changes in the community were toward a more normal micorbiome.
+Finally, we assessed both whether surgery for adenomas and SRN provided
+larger community changes or whether chemotherapy or radiation provided
+additive changes to the microbiome over surgerical resection. This study
+helps to provide evidence as to whether treatment can influence the
+community and if the CRC microbiome, identified in previous studies,
+persists after such interventions.
 
@@
 \textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
-tested whether those with adenoma (n = 41) or carcinoma (n = 26) had any
-broad differences between their initial and follow up samples. We found
-that those with carcinoma had a more dissimilar bacterial community
-between their initial and follow up sample than those with adenoma
-(P-value \textless{} 0.001) {[}Figure 1A{]}. The bacterial community
-structure before and after surgery was visualized using NMDS for both
-adenoma {[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05) and carcinoma
-{[}Figure 1C{]} (PERMANOVA \textless{} 0.05). Interestingly, when
-initial and follow up samples were compared, regardless of whether the
-lesions were adenoma or carcinoma, there was no significant overall
-difference in beta diversity (PERMANOVA \textgreater{} 0.05). There was
-no difference between initial and follow up samples when investigating
-alpha diversity metrics for lesion, adenoma only, or carcinoma only for
-any metric tested {[}Table S1{]}. Additionally, there was also no
-difference in the relative abundance of any OTU between initial and
-follow up samples for lesion, adenoma only, or carcinoma only {[}Figure
-S1{]}.
+tested whether those with adenoma \emph{(n = )}, SRN \emph{(n = )}, or
+carcinoma \emph{(n = )} had any broad differences between their initial
+and follow up samples. We found that those with carcinoma had a more
+dissimilar bacterial community between their initial and follow up
+sample than those with adenoma \emph{(P-value 0.001)} {[}Figure 1A{]}.
+Although no significant differences were observed between SRN and
+carcinoma there was an increase in the dissimilarity of the initial and
+follow up samples from adenoma \emph{(mean and SD)} to SRN \emph{(mean
+and SD)} to carcinoma \emph{(mean and SD)} {[}Figure 1A{]}. The
+bacterial community structure before and after surgery was visualized
+using NMDS for adenoma {[}Figure 1B{]} \emph{(PERMANOVA 0.05)}, SRN
+{[}FIgure 1C{]} \emph{(PERMANOVA 0.05)}, and carcinoma {[}Figure 1C{]}
+\emph{(PERMANOVA 0.05)}. Interestingly, when initial and follow up
+samples were compared, regardless of whether the lesions were adenoma or
+carcinoma, there was no significant overall difference in beta diversity
+\emph{(PERMANOVA 0.05)}. There was no difference between initial and
+follow up samples when investigating alpha diversity metrics for
+adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
+Additionally, there was also no difference in the relative abundance of
+any OTU between initial and follow up samples for adenoma, SRN, or
+carcinoma only {[}Figure S1{]}.
 
-\textbf{\emph{Carcinoma Associated Genera:}} For carcinoma {[}Figure
-S2A{]} (P-value \textgreater{} 0.05) but not adenoma {[}Figure S2B{]}
-(P-value \textgreater{} 0.05) there was a significant decrease in
-\emph{Porphyromonas}, \emph{Parvimonas}, and \emph{Peptostreptococcus}
-between initial and follow up samples {[}Figure S2{]}.
-\emph{Fusobacterium} for both adenoma and carcinoma had no significant
-differences between initial and follow up samples {[}FIgure S2{]}
-(P-value \textgreater{} 0.05). Although there were significant
-differences, only a small percentage of those with adenoma or carcinoma
-were positive for any of these specific genera {[}Figure S2{]}.
+\textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
+20 repeated 10 fold cross-validation was \emph{from to this} with the
+AUC of the model used being \emph{AUC value} {[}Figure S2A{]}. There was
+a total of \emph{blank} OTUs in this model with the vast majority
+classifying to bacteria typically thought of as commensal {[}Figure
+S3A{]}. There was a significant difference between the actual and model
+predicted group stratification calls \emph{(P-value)}. There was no
+significant decrease in the positive probability of adenoma between
+initial and follow up samples \emph{(P\_value)} {[}Figure 2A{]}.
 
-\textbf{\emph{The Lesion Model:}} We tested whether follow up samples
-were more normal than initial samples by building a model based solely
-on OTUs alone that classified lesion versus normal. This model had an
-AUC range of 0.692 - 0.761 after 100 iterations of 20 repeated 10-fold
-cross validations. The ROC curve for the final lesion model used was
-within the observed range of the 100-different test set AUC iterations
-{[}Figure 2A{]}. There was a total of 54 OTUs that were used in this
-model {[}Figure 2B{]}. \emph{Lachnospiraceae} (Otu000015) resulted in
-the largest decrease in MDA {[}Figure 2B{]}. Heavily represented genera
-within the model included OTUs from \emph{Lachnospiraceae},
-\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
-\emph{Ruminococcus}.
+\textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
+repeated 10 fold cross-validation was \emph{from to this} with the AUC
+of the model used being \emph{AUC value} {[}Figure S2B{]}. There was a
+total of \emph{blank} OTUs in the SRN model {[}Figure S3B{]}. Similar to
+the adenoma model the vast majority of OTUs classified to bacteria
+typically thought of as commensal. Also similar to the adenoma model
+there was a significant difference between the actual and model
+predicted group stratification calls \emph{(P-value)}. There was no
+significant decrease in the positive probability of SRN between initial
+and follow up samples \emph{(P\_value)} {[}Figure 2B{]}.
 
-If there was movement towards a more normal community after treatment of
-lesion, we would expect to find a decrease in the positive probability
-of the follow up sample to be a lesion. We observed no such decrease in
-positive probability for either those with adenoma or carcinoma
-{[}Figure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
-significant difference between the predicted and actual calls (P-value
-\textless{} 0.05). However, this model correctly classified the one
-individual who still had carcinoma on follow up and this individual's
-positive probability of lesion increased between their initial and
-follow up sample {[}Figure 2C{]}.
+\textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
+of 20 repeated 10 fold cross-validation was \emph{from to this} with the
+AUC of the model used being \emph{AUC value} {[}Figure S2C{]}.
+Interestingly, the AUCs improved from adenoma to SRN to carcinoma
+{[}Figure 2{]}. There was a total of \emph{blank} OTUs in the carinoma
+model {[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
+majority of OTUs classified to bacteria typically thought of as
+commensal but OTUs that also classified to \emph{Fusobacterium},
+\emph{Porphyromonas}, and \emph{Parvimonas} appear to be important for
+carcinoma classification {[}Figure S3C{]}. Also similar to the adenoma
+and SRN models there was a significant difference between the actual and
+model predicted group stratification calls \emph{(P-value)}. There was a
+significant decrease in the positive probability of carcinoma between
+initial and follow up samples \emph{(P\_value)} {[}Figure 2C{]};
+suggesting that the carcinoma samples changed towards normal after
+treatment, unlike either adenoma or SRN. The one indivdiual still
+positive for carcinoma after treatment had an increase in carcinoma
+positive proability on follow up {[}Figure 2C{]}.
 
-\textbf{\emph{The Treatment Model:}} To test if there were differences
-that could be used to classify initial and follow up treatment samples
-we built a model to classify initial samples. This treatment model had
-an AUC range of 0.657 to 0.796. after 100 iterations of 20 repeated
-10-fold cross validations. The test set AUC range for this model
-performed better than the training set AUCs. There was a marked decrease
-in the ROC curve for the final model used when compared to the 100 test
-set AUC iterations {[}Figure 3A{]}. There was a total of 70 OTUs that
-were used for this model {[}Figure 3B{]}. The variable that resulted in
-the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
-Similar to the lesion model, heavily represented genera included OTUs
-from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
-\emph{Blautia}, and \emph{Ruminococcus}.
+\textbf{\emph{Adenoma, SRN, and Carcinoma Common OTUs}} We next wanted
+to know what predictors within the adenoma, SRN, and carcinoma models
+were similar to each other. The main purpose was to identify which OTUs
+could be important at all three stages of disease. When we compared the
+three different models with each other there were a total of
+\emph{number} common OTUs. Some of the most common taxonomic
+identifications belonged to \emph{one}, \emph{two}, \emph{three},
+\emph{four}, \emph{five}, \emph{six}, and \emph{seven}. These along with
+the vast majority of the OTUs that were common between these models had
+classifications to bacteria typically thought of as commensal {[}Table
+S2{]}.
 
-If there were changes between initial and follow up samples we would
-expect the positive probability of being an initial sample to be
-decreased in the follow up. The is what we observed for the treatment
-model (P-value \textless{} 0.001). When we separated lesion into adenoma
-and carcinoma there was a decrease in positive probability for both the
-carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
-{[}Figure 3D{]} \textless{} 0.001). For this model, there was no
-difference between the predicted and actual classifications (P-value
-\textgreater{} 0.05).
-
-\textbf{\emph{Common OTUs to both Models:}} We next wanted to know what
-predictors within the lesion model were also in the initial sample
-model. The main purpose was to identify which OTUs could be important in
-lesion formation and were impacted by treatment. When we compared the
-two different models with each other there were a total of 32 common
-OTUs. Some of the most common taxonomic identifications belonged to
-\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
-\emph{Blautia}, \emph{Anaerostipes}, \emph{Dorea}, and
-\emph{Ruminococcus}. These along withthe vast majority of the OTUs that
-were common between models had classifications to bacteria typically
-thought of as commensal {[}Table S2{]}.
-
-\textbf{\emph{Chemotherapy and Radiation Differences:}} After observing
-these changes due to treatment we assessed whether chemotherapy or
-radiation impacted the observed results. Only the treatment model had a
-significant change in positive probability for those treated with
-chemotherapy or radiation versus those who received neither {[}Figure 4A
-\& Table S3{]} (P-value \textless{} 0.05). Additionally, there was no
-difference between the change in positive probability for chemotherapy
-and radiation therapy (Uncorrected P-value \textgreater{} 0.05). Using
-the 32 common OTUs we tested how chemotherapy or radiation affects
-important members of the community. For both chemotherapy and radiation,
-the directionality of change of these OTUs were similar {[}Figure 4B \&
-4C{]}. After multiple comparison correction, only \emph{Blautia}
-(Otu000006) remained significantly different between those who did and
-did not receive chemotherapy. This data suggests that follow up samples
-from those treated with either chemotherapy or radiation may have had a
-larger change from initial sample due to effects on the bacterial
-community.
+\textbf{\emph{Treatment Affects on Community}} After observing these
+these changes from treatment we assessed whether chemotherapy or
+radiation, in the carcinoma group, and surgery, in the adenoma group,
+impacted the observed results. In the carcinoma group neither
+chemotherapy or radiation provided any additive change from initial
+sample over surgery alone {[}Table S3{]}. For the adenoma group there
+was no difference between those that received surgery and those that did
+not {[}Table S4{]}. For surgery adenoma and SRN were combined due to the
+low number of surgery occurances. There was no difference in the
+proportion of those receiving surgery between the adenoma and SRN groups
+\emph{(P-value)}. This data suggests that microbiome changes observed in
+the carcinoma group were mostly a result from surgery and not from
+chemotherapy or radiation.
 
@@
 without CRC differ and how it might be used as an early screening tool
-{[}7,8,10--12{]}. Here we describe how the bacterial community changes
-after treatment and which OTUs may be most important to lesion
-classification. Interestingly, many of the most important OTUs for both
-treatment and lesion classification had taxonomic identification for
-resident gut microbes. This suggest that members within the commensal
-community may be the first that change during CRC pathogenesis. These
-changes, in turn, could be the first step in allowing more inflammatory
-bacteria to gain a foothold within the colon {[}9{]}.
+{[}7,8,11--13{]}. Here we show that the bacterial community changes
+towards normal after treatment for carcinoma and that chemotherapy and
+radiation did not provide an additive change. Although some of the
+important OTUs classified to genera from bacteria considered the usual
+suspects (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
+\emph{Parvimonas}) many did not. The majority of important OTUs had
+taxonomic classifications for resident gut microbes and were common for
+the adenoma, SRN, and carcinoma models. This suggests that members
+within the commensal community may be the first that change during CRC
+pathogenesis. These subtle changes, in turn, could be the first step in
+allowing more inflammatory bacteria to gain a foothold within the colon
+{[}9{]}.
 
-Unlike previous studies on the microbiome and CRC, ours focuses on
-identifying commonalities within both adenoma and carcinoma groups
-before and after treatment. Although there were differences for genera
-associated with specific bacterium linked with CRC {[}Figure S2{]}.
-These changes were not consistent across lesion. Instead, most of our
-results provide support that the first members of the community to
+Unlike previous studies on the microbiome and CRC, ours focuses both on
+identifying commonalities and differences within adenoma, SRN, and
+carcinoma groups before and after treatment. Although there were
+differences for genera associated with specific bacterium linked with
+CRC {[}Figure S3{]}. The majority of important OTUs taxonmically
+classified to commensal bacteria {[}Figure S3{]}. Although these changes
+may be subtle, due to the lack of signifcant difference in the bacterial
+community before and after treatment in adenoma and SRN {[}Figure 1{]},
+they support the hypothesis that the first members of the community to
 change and potentially stay changed even after treatment are those that
-are commensal bacteria {[}Figure 2-3 \& Table S3{]}. Additionally,
-treatment with chemotherapy or radiation changed these bacteria and
-provided a larger decrease in positive probability then removal of
-lesion.
-
-Curiously, we observed that the typical CRC associated bacteria were not
-predictive within our models. There are several reasons why this may
-have occurred. First, is that they were not present in enough
-individuals to be able to classify those with and without disease with a
-high degree of accuracy. Second, is that our Random Forest models were
-able to gather the same information from other OTUs. Third, is that
-changes in commensal bacteria are the common differences across adenoma
-and carcinoma. Finally, it is also possible that all these explanations
-could have played a role. Our observations though, would suggest that an
-individual's resident bacteria have a large role to play in polyp
-formation and could change in a way that allows predictive models to
-lower the positive probability of lesion or initial sample before
-treatment {[}Figure 2 \& 3{]}. It should be noted that our study does
-not argue against the importance of these CRC associated bacteria in the
-pathogenesis of disease but rather that they are not the main bacteria
-changing after treatment. It is possible that these CRC associated
-bacteria are important in the transition from adenoma to carcinoma and
-would be one explanation as to why in our data we not only see high
-initial relative abundances in carcinoma and not adenoma individuals but
-also large decreases in relative abundance in some of those with
-carcinoma but not in those with adenoma after treatment {[}Figure S2{]}.
+are commensal bacteria.
 
 Many of the common OTUs that we identified taxnomically classified to
-potential butyrate producers {[}Table S2{]}. Another batch of OTUs
-classified to bacteria that can either degrade polyphenols or are
-inhibited by them. Both butyrate and polyphenols are thought to be
-protective against cancer in part by reducing inflammation {[}13{]}.
+potential butyrate producers (e.g. \emph{Clostridiales},
+\emph{Roseburia}, and \emph{Anaerostipes}){[}Table S2{]}. Other OTUs
+classified to bacteria that are inhibited by polyphenols (e.g.
+\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
+protective against cancer, in part by reducing inflammation {[}14{]}.
 These protective compounds are derived from the breakdown of fiber,
@@
 in the time elapsed in the collection of the follow up sample between
-adenoma and carcinoma (uncorrected P-value \textless{} 0.05), with time
-passed being less for adenoma (253 +/- 41.3 days) than carcinoma (351
-+/- 102 days). These results would indicate that some of the differences
-observed between the carcinoma and adenoma groups could be due to
-differences in collection time. Specifically, it could confound the
-observation that carcinomas changed more than adenomas {[}Figure 1A \&
-1C{]}. This confounding though would not affect the observations where
-these individuals were grouped together {[}Figures 2-4{]}.
+adenoma or SRN versus carcinoma \emph{(uncorrected P-value 0.05)}, with
+time passed being less for adenoma \emph{( days)} and SRN \emph{( days)}
+than carcinoma \emph{( days)}. These results would indicate that some of
+the differences observed between the carcinoma and adenoma groups could
+be due to differences in collection time. Specifically, it could
+confound the observation that carcinomas changed more than adenomas
+{[}Figure 1A \& 1C{]}. However, there are two reasons that this may not
+be the case. First, the SRN group did not have a significant difference
+and the collection time of their follow up sample was less than the
+adenoma group. Second, this confounding would not affect the
+observations where models were used since they were built using a
+different cohort {[}Figures 2 \& S2-S3{]}.
 
 Another limitation was that we do not know whether individuals who were
-still classified as positive by the lesion model eventually had a
-subsequent CRC diagnosis. This information would help to strengthen the
-case for this model keeping several individuals above the cutoff
+still classified as positive by the carcinoma model eventually had a
+subsequent CRC recurrence. This information would help to strengthen the
+case for this model keeping numerous individuals above the cutoff
 threshold even though at follow up they were diagnosed as no longer
-having a lesion. This study also drew heavily from those with Caucasian
+having carcinoma. This study also drew heavily from those with Caucasian
 ancestry making it possible that the observations may not be
@@
 
-Interestingly, within the treatment model the test data performed better
-than the training data. This may have occurred because the training AUC
-determined from 20 repeated 10-fold cross validation removed samples at
-random and did not consider that they were matched samples. Another
-potential reason is that the model itself may have been over-fit since
-the total number of samples was not that large. However, the lesion
-model did not suffer from these discrepancies. Further independent
-studies need to be carried out to verify our findings on lesion and
-treatment changes due to these limitations.
-
 Despite these shortcomings our findings add to the existing scientific
 knowledge on CRC and the microbiome: That there is a measurable
-difference in the bacterial community after adenoma and carcinoma
+difference in the bacterial community after adenoma, SRN or carcinoma
 treatment. Further, the ability for machine learning algorithms to take
-OTU data and successfully lower positive probability of lesion or
-initial sample after treatment provides evidence that there are specific
-signatures, mostly attributable to commensal organisms, associated with
-both treatment and lesion. Our data provides evidence that commensal
+OTU data and successfully lower positive probability of carcinoma after
+treatment provides evidence that there are specific signatures,
+attributable to both inflammtory and resident commensal organisms,
+associated with treatment. Our data provides evidence that commensal
 bacteria may be important in the development of polyps and potentially
@@
 normalize back to a pre-colonscopy community within this time period
-{[}17{]}. Our study cohort consisted of 67 individuals with an initial
-sample as described and a follow up sample obtained between 188 - 546
-days after treatment of lesion. This study was approved by the
-University of Michigan Institutional Review Board. All study
-participants provided informed consent and the study itself conformed to
-the guidelines set out by the Helsinki Declaration.
+{[}18{]}. Our training cohort consisted of a total of 423 individuals
+{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
+initial sample as described and a follow up sample obtained between 188
+- 546 days after treatment of lesion {[}Table 2{]}. This study was
+approved by the University of Michigan Institutional Review Board. All
+study participants provided informed consent and the study itself
+conformed to the guidelines set out by the Helsinki Declaration.
 
 \textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
-described by Kozich, et al. {[}18{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}19{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}18{]}. The general workflow using mothur was:
+previously described {[}19{]}. The general workflow using mothur was:
 Paired-end reads were first merged into contigs, quality filtered,
@@
 
-\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}19{]}
+\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}20{]}
 algorithm was used to create the model used for prediction of lesion
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}20{]} in the vegan package
+community structure utilized PERMANOVA {[}21{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}21{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}22{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
-\textbf{Figure 1: General Differences between the Adenoma and Carcinoma
-Group.} A) A significant difference was found between the adenoma and
-carcinoma group for thetayc (P-value = 0.000472). Advanced adenomas are
-denoted as Screen Relevant Neoplasia (SRN). B) NMDS of the initial and
-follow up samples for the adenoma group. C) NMDS of the initial and
-follow up samples for the carcinoma group.
+\textbf{Figure 1: General Differences between Adenoma, SRN, and
+Carcinoma Groups After Treatment.} A) A significant difference was found
+between the adenoma and carcinoma group for thetayc (P-value = NULL).
+Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN). B)
+NMDS of the initial and follow up samples for the adenoma group. C) NMDS
+of the initial and follow up samples for the carcinoma group.
 
-\textbf{Figure 2: The Lesion Model.} A) ROC curve: The shaded areas
-represent the range of values of a 100 different 80/20 splits of the
-test set data and the blue line represents the model using 100\% of the
-data set and what was used for subsequent classification. B) Summary of
-Important Variables. MDA of the most important variables in the lesion
-model. The black point represents the mean and the different colors are
-the values of each different run up to 100. C) Positive probability
+\textbf{Figure 2: Treatment Response Based on Models Built for Adenoma,
+SRN, or Carcinoma.} A) Positive probability change from initial to
+follow up sample in those with adenoma. B) Positive probability change
+from initial to follow up sample in those with SRN. C) Positive
+probability change from initial to follow up sample in those with
+carcinoma..
+
+\textbf{Figure 3: The Treatment Model.} B) C) Positive probability
 change from initial to follow up sample in those with carcinoma. D)
-Positive probability change from initial to follow up sample if those
+Positive probability change from initial to follow up sample of those
 with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
 
-\textbf{Figure 3: The Treatment Model.} A) ROC curve: The shaded areas
-represent the range of values of a 100 different 80/20 splits of the
-test set data and the blue line represents the model using 100\% of the
-data set and what was used for subsequent classification. B) Summary of
-Important Variables. MDA of the most important variables in the initial
-sample model. The black point represents the mean and the different
-colors are the values of each different run up to 100. C) Positive
-probability change from initial to follow up sample in those with
-carcinoma. D) Positive probability change from initial to follow up
-sample of those with adenoma or advanced adenoma (Screen Relevant
-Neoplasia (SRN)).
+\newpage
 
-\textbf{Figure 4: Chemotherapy and Radiation Therapy Effects on
-Community.} A) Treatment model initial sample positive probability
-reduction for chemotherapy. B) Treatment model initial sample positive
-probability reduction for radiation therapy. C) Common OTUs based on
-whether chemotherapy was received or not. The * denotes a P-value
-\textless{} 0.05 after multiple comparison correction. D) Common OTUs
-based on whether radiation therapy was received or not.
+\textbf{Table 1: Demographic Data of Training Cohort}
+
+\textbf{Table 2: Demographic Data of Pre and Post Treatment Cohort}
 
@@
 \textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
-Analysis of All OTUs for Initial versus Follow Up}
+Analysis of All OTUs Before and After Treatment}
 
-\textbf{Figure S2: Previously Associated CRC Bacteria in Initial and
-Follow Up Samples.} A) Carcinoma initial and follow up samples had an
-observed significant difference in initial and follow up sample for the
-OTUs classified as \emph{Parvimonas} (P-value = 0.0059),
-\emph{Porphyromonas} (P-value = 0.225), \emph{Peptostreptococcus}
-(P-value = 0.00424). B) Adenoma initial and follow up samples. There
-were no significant differences between initial and follow up (P-value =
-0.881.
+\textbf{Figure S2: ROC Curves of the Adenoma, SRN, and Carcinoma
+Models.} A) Adenoma ROC curve: The light greenshaded areas represent the
+range of values of a 100 different 80/20 splits of the test set data and
+the dark green line represents the model using 100\% of the data set and
+what was used for subsequent classification. B) SRN ROC curve: The light
+yellow shaded areas represent the range of values of a 100 different
+80/20 splits of the test set data and the dark yellow line represents
+the model using 100\% of the data set and what was used for subsequent
+classification. C) Carcinoma ROC curve: The light red shaded areas
+represent the range of values of a 100 different 80/20 splits of the
+test set data and the dark red line represents the model using 100\% of
+the data set and what was used for subsequent classification.
+
+\textbf{Figure S3: Summary of Important Variables for the Adenoma, SRN,
+and Carcinoma Models.} A) MDA of the most important variables in the
+adenoma model. The dark green point represents the mean and the lighter
+green points are the value of each of the 100 different runs. B) Summary
+of Important Variables in the SRN model. MDA of the most important
+variables in the SRN model. The dark yellow point represents the mean
+and the lighter yellow points are the value of each of the 100 different
+runs. C) MDA of the most important variables in the carcinoma model. The
+dark red point represents the mean and the lighter redpoints are the
+value of each of the 100 different runs.
 
@@
 
+\hypertarget{ref-hassan_efficacy_2016}{}
+10. Hassan C, Repici A, Sharma P, Correale L, Zullo A, Bretthauer M, et
+al. Efficacy and safety of endoscopic resection of large colorectal
+polyps: A systematic review and meta-analysis. Gut. 2016;65:806--20.
+
 \hypertarget{ref-yu_metagenomic_2017}{}
-10. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
+11. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
 analysis of faecal microbiome as a tool towards targeted non-invasive
@@
 \hypertarget{ref-zackular_human_2014}{}
-11. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+12. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-warren_co-occurrence_2013}{}
-12. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
+13. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
 et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
@@
 \hypertarget{ref-obrien_impact_2013}{}
-17. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-18. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+19. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-19. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+20. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-20. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+21. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-21. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+22. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-03-24 code/Run_adn_crc_Test_Chemo_Rad.R

@@ -45,8 +45,6 @@ alpha_data <- mutate(alpha_data,
 adn_test <- (filter(alpha_data, sampleType == "initial" & Dx_Bin != "cancer")[, c("sobs", "shannon", "shannoneven")] - 
   filter(alpha_data, sampleType == "followups" & Dx_Bin != "cancer")[, c("sobs", "shannon", "shannoneven")]) %>% 
   mutate(distance = filter(difference_table_treatment, Dx_Bin != "cancer")[, "distance"], 
-         red_adn = (filter(data_list[["red_adn_probs"]], sampleType == "initial" & Dx_Bin != "cancer")[, "Yes"] - 
-           filter(data_list[["red_adn_probs"]], sampleType == "followup" & Dx_Bin != "cancer")[, "Yes"]), 
          chemo = filter(good_metaf, Dx_Bin != "cancer")[, "chemo_received"], 
          rads = filter(good_metaf, Dx_Bin != "cancer")[, "radiation_received"], 
          surgery = filter(good_metaf, Dx_Bin !="cancer")[, "Surgery"], 
@@ -65,13 +63,13 @@ crc_test <- (filter(alpha_data, sampleType == "initial" & Dx_Bin == "cancer")[, 
 
 
 # Create variables for actual analysis to be automated
-adn_tests <- c("sobs", "shannon", "shannoneven", "distance", "red_adn")
+adn_tests <- c("sobs", "shannon", "shannoneven", "distance", "red_adn", "red_srn")
 
 crc_tests <- c("sobs", "sobs", "shannon", "shannon", "shannoneven", "shannoneven", 
                "distance", "distance", "red_crc", "red_crc")
 
-adn_mean_table <- as.data.frame(matrix(nrow = 5, ncol = 6, dimnames = list(
-  rown = c("sobs", "shannon", "even", "dist", "red_adn"), 
+adn_mean_table <- as.data.frame(matrix(nrow = 6, ncol = 6, dimnames = list(
+  rown = c("sobs", "shannon", "even", "dist", "red_adn", "red_srn"), 
   coln = c("surg_mean", "no_surg_mean", "surg_sd", "no_surg_sd", "pvalue", "bh"))))
 
 crc_mean_table <- as.data.frame(matrix(nrow = 10, ncol = 6, dimnames = list(
[…]

### 2017-03-24 code/Run_adn_srn_crc_probs_comparison.R

@@ -36,7 +36,7 @@ adn_wilcox_pvalue_summary <- getProb_PairedWilcox(
   extra_specifics = "adenoma")
 
 srn_wilcox_pvalue_summary <- getProb_PairedWilcox(
-  adn_follow_up_probability, 
+  srn_follow_up_probability, 
   rown = "adv_adenoma", 
   not_group = c("cancer", "adenoma"), 
   extra_specifics = "adv_adenoma")
@@ -69,9 +69,7 @@ tempList <- list(
 model_summary_info <- as.data.frame(c())
 
 model_summary_info <- 
-  cbind(info = rownames(get_confusion_data(tempList[[1]], good_metaf, to_filter1 = "cancer", 
-                                                 to_filter2 = "cancer")), 
-              rbind(
+  cbind(rbind(
                 get_confusion_data(tempList[["red_adn"]], good_metaf, to_filter1 = "cancer", 
                                    to_filter2 = "adv_adenoma"), 
                 get_confusion_data(tempList[["red_srn"]], good_metaf, to_filter1 = "cancer", 

### 2017-03-24 submission/manuscript_outline_20161024.Rmd
latest revised draft

@@
 adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)
+adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)
 
@@
 srn_AUC_run_summary <- read.csv('../data/process/tables/srn_Reduced_ROC_model_summary.csv', header = T)
+srn_general_summary <- read.csv('../data/process/tables/srn_reduced_auc_summary.csv', header = T, row.names = 1)
 
@@
 crc_AUC_run_summary <- read.csv('../data/process/tables/crc_Reduced_ROC_model_summary.csv', header = T)
+crc_general_summary <- read.csv('../data/process/tables/crc_reduced_auc_summary.csv', header = T, row.names = 1)
 
 # All model information
-all_probs_paired_wilcox <- read.csv('../data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
-all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F)
+all_probs_paired_wilcox <- read.csv('../data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F, row.names = 4)
+all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F) %>% rename(comparison = X)
 all_confu_matrix <- read.csv('../data/process/tables/all_crc_srn_adn_models_confusion_summary.csv', header = T, stringsAsFactors = F)
@@
 
-**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma or SRN cases *(P-value )*.
+**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus SRN cases (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 But there was a trend for increasingly less community similarity between initial and follow up samples from adenoma to SRN to carcinoma.
-Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups *(P-value )*.
-Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability *(P-value)*; providing further evidence that treatment has the greatest effect in those with carcinoma.
-The adenoma model used a total of *model n* OTUs, the SRN model used a total of *model n* OTUs, and the carcinoma model used a total of *model n* OTUs.
-A total of *common n* OTUs were common to both models with many classifying to commensal bacteria (e.g.
+Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma.
+The adenoma model used a total of `r length(rownames(adn_imp_vars))` OTUs, the SRN model used a total of `r length(rownames(srn_imp_vars))` OTUs, and the carcinoma model used a total of `r length(rownames(crc_imp_vars))` OTUs.
+A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g.
 *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
-Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma *(Pvalue )*
+Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (Pvalue `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
@@
 ***The Bacterial Community:*** 
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma *(n = )*, SRN *(n = )*, or carcinoma *(n = )* had any broad differences between their initial and follow up samples.
-We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma *(P-value 0.001)* [Figure 1A].
-Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma *(mean and SD)* to SRN *(mean and SD)* to carcinoma *(mean and SD)* [Figure 1A].
-The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] *(PERMANOVA  0.05)*, SRN [FIgure 1C] *(PERMANOVA 0.05)*, and carcinoma [Figure 1C] *(PERMANOVA  0.05)*.
-Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity *(PERMANOVA  0.05)*.
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma (n = `r change_summary["thetayc", "adn_n"]`), SRN (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between their initial and follow up samples.
+We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
+Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to SRN (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].
+The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), SRN [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 There was no difference between initial and follow up samples when investigating alpha diversity metrics for adenoma, SRN, or carcinoma for any metric tested [Table S1].
@@
 ***Adenoma Model***
-The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2A].
-There was a total of *blank* OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A].
-There was a significant difference between the actual and model predicted group stratification calls *(P-value)*.
-There was no significant decrease in the positive probability of adenoma between initial and follow up samples *(P_value)* [Figure 2A].
+The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification havin an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].
+There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A].
+There was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was no significant decrease in the positive probability of adenoma between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].
 
 ***SRN Model***
-The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2B].
-There was a total of *blank* OTUs in the SRN model [Figure S3B].
+The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].
+There was a total of `r length(rownames(srn_imp_vars))` OTUs in the SRN model [Figure S3B].
 Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal.
-Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.
-There was no significant decrease in the positive probability of SRN between initial and follow up samples *(P_value)* [Figure 2B].
+Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was no significant decrease in the positive probability of SRN between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].
 
 ***Carcinoma Model***
-The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2C].
+The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].
 Interestingly, the AUCs improved from adenoma to SRN to carcinoma [Figure 2].
-There was a total of *blank* OTUs in the carinoma model [Figure S3C].
+There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carinoma model [Figure S3C].
 Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* appear to be important for carcinoma classification [Figure S3C].
-Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.
-There was a significant decrease in the positive probability of carcinoma between initial and follow up samples *(P_value)* [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.
+Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was a significant decrease in the positive probability of carcinoma between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.
 The one indivdiual still positive for carcinoma after treatment had an increase in carcinoma positive proability on follow up [Figure 2C].
@@
 The main purpose was to identify which OTUs could be important at all three stages of disease.
-When we compared the three different models with each other there were a total of *number* common OTUs.
-Some of the most common taxonomic identifications belonged to *one*, *two*, *three*, *four*, *five*, *six*, and *seven*.
+When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.
+Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*.
 These along with the vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].
@@
 After observing these these changes from treatment we assessed whether chemotherapy or radiation, in the carcinoma group, and surgery, in the adenoma group, impacted the observed results.
-In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone [Table S3].
-For the adenoma group there was no difference between those that received surgery and those that did not [Table S4].
-For surgery adenoma and SRN were combined due to the low number of surgery occurances.
-There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups *(P-value)*.
+In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S3].
+For the adenoma group there was a single difference in observed OTUs (sobs) between those that received surgery and those that did not (P-value `r ifelse(min(comb_adn_surg_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S4].
+For the surgery comparison, adenoma and SRN were combined due to the low number of surgery occurances in these two groups.
+There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups (P-value `r ifelse(comb_adn_surg_summary["prop_surg_adn_srn", "pvalue"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.
@@
 Many of the common OTUs that we identified taxnomically classified to potential butyrate producers (e.g.
-*Clostridiales*, *Roseburia*,  and *Anaerostipes*)[Table S2].
+*Clostridiales*, *Roseburia*,  and *Anaerostipes*) [Table S2].
 Other OTUs classified to bacteria that are inhibited by polyphenols (e.g.
@@
 
-A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma *(uncorrected P-value 0.05)*, with time passed being less for adenoma *( days)* and SRN *( days)* than carcinoma *( days)*.
+A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and SRN (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).
 These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time.
-Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1C].
+Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1].
 However, there are two reasons that this may not be the case.

### 2017-03-24 code/Run_adn_srn_crc_probs_comparison.R

@@ -109,7 +109,7 @@ write.csv(all_wilcox_summary,
           "data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv", row.names = F)
 
 write.csv(model_summary_info, 
-          "data/process/tables/all_crc_srn_adn_models_summary_info.csv", row.names = F)
+          "data/process/tables/all_crc_srn_adn_models_summary_info.csv")
 
 write.csv(confusion_counts_summary, "data/process/tables/all_crc_srn_adn_models_confusion_summary.csv", row.names = F)
 

### 2017-03-24 submission/manuscript_outline_20161024.tex

@@
 \textbf{Results:} There was a larger change in the bacterial community
-in response to treatment for carcinoma versus adenoma or SRN cases
-\emph{(P-value )}. But there was a trend for increasingly less community
-similarity between initial and follow up samples from adenoma to SRN to
-carcinoma. Despite this, no difference was found in the relative
-abundance of any OTU before and after treatment for adenoma, SRN, or
-carcinoma groups \emph{(P-value )}. Using Random Forest models to assess
-whether changes in follow up samples were towards a normal community
-showed that only those with carcinoma had a significant decrease in
-positive probability \emph{(P-value)}; providing further evidence that
-treatment has the greatest effect in those with carcinoma. The adenoma
-model used a total of \emph{model n} OTUs, the SRN model used a total of
-\emph{model n} OTUs, and the carcinoma model used a total of \emph{model
-n} OTUs. A total of \emph{common n} OTUs were common to both models with
-many classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
+in response to treatment for carcinoma versus adenoma (P-value
+\textless{} 0.05) but not carcinoma versus SRN cases (P-value
+\textgreater{} 0.05). But there was a trend for increasingly less
+community similarity between initial and follow up samples from adenoma
+to SRN to carcinoma. Despite this, no difference was found in the
+relative abundance of any OTU before and after treatment for adenoma,
+SRN, or carcinoma groups (P-value \textgreater{} 0.05). Using Random
+Forest models to assess whether changes in follow up samples were
+towards a normal community showed that only those with carcinoma had a
+significant decrease in positive probability (P-value \textless{} 0.05);
+providing further evidence that treatment has the greatest effect in
+those with carcinoma. The adenoma model used a total of 62 OTUs, the SRN
+model used a total of 61 OTUs, and the carcinoma model used a total of
+59 OTUs. A total of 26 OTUs were common to both models with many
+classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
 \emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
@@
 additional changes to the bacterial community in those treated for
-carcinoma \emph{(Pvalue )}
+carcinoma (Pvalue \textgreater{} 0.05).
 
@@
 \textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
-tested whether those with adenoma \emph{(n = )}, SRN \emph{(n = )}, or
-carcinoma \emph{(n = )} had any broad differences between their initial
-and follow up samples. We found that those with carcinoma had a more
-dissimilar bacterial community between their initial and follow up
-sample than those with adenoma \emph{(P-value 0.001)} {[}Figure 1A{]}.
-Although no significant differences were observed between SRN and
-carcinoma there was an increase in the dissimilarity of the initial and
-follow up samples from adenoma \emph{(mean and SD)} to SRN \emph{(mean
-and SD)} to carcinoma \emph{(mean and SD)} {[}Figure 1A{]}. The
-bacterial community structure before and after surgery was visualized
-using NMDS for adenoma {[}Figure 1B{]} \emph{(PERMANOVA 0.05)}, SRN
-{[}FIgure 1C{]} \emph{(PERMANOVA 0.05)}, and carcinoma {[}Figure 1C{]}
-\emph{(PERMANOVA 0.05)}. Interestingly, when initial and follow up
+tested whether those with adenoma (n = 22), SRN (n = 19), or carcinoma
+(n = 26) had any broad differences between their initial and follow up
+samples. We found that those with carcinoma had a more dissimilar
+bacterial community between their initial and follow up sample than
+those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. Although
+no significant differences were observed between SRN and carcinoma there
+was an increase in the dissimilarity of the initial and follow up
+samples from adenoma (0.55 ± 0.21 (mean ± SD)) to SRN (0.65 ± 0.25) to
+carcinoma (0.78 ± 0.15) {[}Figure 1A{]}. The bacterial community
+structure before and after surgery was visualized using NMDS for adenoma
+{[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05), SRN {[}Figure 1C{]}
+(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
+(PERMANOVA \textless{} 0.05). Interestingly, when initial and follow up
 samples were compared, regardless of whether the lesions were adenoma or
 carcinoma, there was no significant overall difference in beta diversity
-\emph{(PERMANOVA 0.05)}. There was no difference between initial and
-follow up samples when investigating alpha diversity metrics for
+(PERMANOVA \textgreater{} 0.05). There was no difference between initial
+and follow up samples when investigating alpha diversity metrics for
 adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
@@
 \textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
-20 repeated 10 fold cross-validation was \emph{from to this} with the
-AUC of the model used being \emph{AUC value} {[}Figure S2A{]}. There was
-a total of \emph{blank} OTUs in this model with the vast majority
+20 repeated 10 fold cross-validation was 0.62 - 0.72 with the AUC of the
+model used for classification havin an AUC of 0.65 {[}Figure S2A{]}.
+There was a total of 62 OTUs in this model with the vast majority
 classifying to bacteria typically thought of as commensal {[}Figure
 S3A{]}. There was a significant difference between the actual and model
-predicted group stratification calls \emph{(P-value)}. There was no
-significant decrease in the positive probability of adenoma between
-initial and follow up samples \emph{(P\_value)} {[}Figure 2A{]}.
+predicted group stratification calls (P-value \textless{} 0.05). There
+was no significant decrease in the positive probability of adenoma
+between initial and follow up samples (P-value \textgreater{} 0.05)
+{[}Figure 2A{]}.
 
 \textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
-repeated 10 fold cross-validation was \emph{from to this} with the AUC
-of the model used being \emph{AUC value} {[}Figure S2B{]}. There was a
-total of \emph{blank} OTUs in the SRN model {[}Figure S3B{]}. Similar to
-the adenoma model the vast majority of OTUs classified to bacteria
+repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
+model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
+There was a total of 61 OTUs in the SRN model {[}Figure S3B{]}. Similar
+to the adenoma model the vast majority of OTUs classified to bacteria
 typically thought of as commensal. Also similar to the adenoma model
 there was a significant difference between the actual and model
-predicted group stratification calls \emph{(P-value)}. There was no
-significant decrease in the positive probability of SRN between initial
-and follow up samples \emph{(P\_value)} {[}Figure 2B{]}.
+predicted group stratification calls (P-value \textless{} 0.05). There
+was no significant decrease in the positive probability of SRN between
+initial and follow up samples (P-value \textgreater{} 0.05) {[}Figure
+2B{]}.
 
 \textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
-of 20 repeated 10 fold cross-validation was \emph{from to this} with the
-AUC of the model used being \emph{AUC value} {[}Figure S2C{]}.
+of 20 repeated 10 fold cross-validation was 0.84 - 0.9 with the AUC of
+the model used for classification being 0.88 {[}Figure S2C{]}.
 Interestingly, the AUCs improved from adenoma to SRN to carcinoma
-{[}Figure 2{]}. There was a total of \emph{blank} OTUs in the carinoma
-model {[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
+{[}Figure 2{]}. There was a total of 59 OTUs in the carinoma model
+{[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
 majority of OTUs classified to bacteria typically thought of as
@@
 and SRN models there was a significant difference between the actual and
-model predicted group stratification calls \emph{(P-value)}. There was a
-significant decrease in the positive probability of carcinoma between
-initial and follow up samples \emph{(P\_value)} {[}Figure 2C{]};
-suggesting that the carcinoma samples changed towards normal after
-treatment, unlike either adenoma or SRN. The one indivdiual still
-positive for carcinoma after treatment had an increase in carcinoma
-positive proability on follow up {[}Figure 2C{]}.
+model predicted group stratification calls (P-value \textless{} 0.05).
+There was a significant decrease in the positive probability of
+carcinoma between initial and follow up samples (P-value \textless{}
+0.05) {[}Figure 2C{]}; suggesting that the carcinoma samples changed
+towards normal after treatment, unlike either adenoma or SRN. The one
+indivdiual still positive for carcinoma after treatment had an increase
+in carcinoma positive proability on follow up {[}Figure 2C{]}.
 
@@
 could be important at all three stages of disease. When we compared the
-three different models with each other there were a total of
-\emph{number} common OTUs. Some of the most common taxonomic
-identifications belonged to \emph{one}, \emph{two}, \emph{three},
-\emph{four}, \emph{five}, \emph{six}, and \emph{seven}. These along with
-the vast majority of the OTUs that were common between these models had
+three different models with each other there were a total of 26 common
+OTUs. Some of the most common taxonomic identifications belonged to
+\emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
+\emph{Lachnospiraceae}, and \emph{Dorea}. These along with the vast
+majority of the OTUs that were common between these models had
 classifications to bacteria typically thought of as commensal {[}Table
@@
 chemotherapy or radiation provided any additive change from initial
-sample over surgery alone {[}Table S3{]}. For the adenoma group there
-was no difference between those that received surgery and those that did
-not {[}Table S4{]}. For surgery adenoma and SRN were combined due to the
-low number of surgery occurances. There was no difference in the
+sample over surgery alone (P-value \textgreater{} 0.05) {[}Table S3{]}.
+For the adenoma group there was a single difference in observed OTUs
+(sobs) between those that received surgery and those that did not
+(P-value \textless{} 0.05) {[}Table S4{]}. For the surgery comparison,
+adenoma and SRN were combined due to the low number of surgery
+occurances in these two groups. There was no difference in the
 proportion of those receiving surgery between the adenoma and SRN groups
-\emph{(P-value)}. This data suggests that microbiome changes observed in
-the carcinoma group were mostly a result from surgery and not from
-chemotherapy or radiation.
+(P-value \textgreater{} 0.05). This data suggests that microbiome
+changes observed in the carcinoma group were mostly a result from
+surgery and not from chemotherapy or radiation.
 
@@
 potential butyrate producers (e.g. \emph{Clostridiales},
-\emph{Roseburia}, and \emph{Anaerostipes}){[}Table S2{]}. Other OTUs
+\emph{Roseburia}, and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
 classified to bacteria that are inhibited by polyphenols (e.g.
@@
 in the time elapsed in the collection of the follow up sample between
-adenoma or SRN versus carcinoma \emph{(uncorrected P-value 0.05)}, with
-time passed being less for adenoma \emph{( days)} and SRN \emph{( days)}
-than carcinoma \emph{( days)}. These results would indicate that some of
-the differences observed between the carcinoma and adenoma groups could
-be due to differences in collection time. Specifically, it could
-confound the observation that carcinomas changed more than adenomas
-{[}Figure 1A \& 1C{]}. However, there are two reasons that this may not
-be the case. First, the SRN group did not have a significant difference
-and the collection time of their follow up sample was less than the
-adenoma group. Second, this confounding would not affect the
-observations where models were used since they were built using a
-different cohort {[}Figures 2 \& S2-S3{]}.
+adenoma or SRN versus carcinoma (P-value \textless{} 0.05), with time
+passed being less for adenoma (255 ± 42 days) and SRN (250 ± 41) than
+carcinoma (351 ± 102). These results would indicate that some of the
+differences observed between the carcinoma and adenoma groups could be
+due to differences in collection time. Specifically, it could confound
+the observation that carcinomas changed more than adenomas {[}Figure
+1{]}. However, there are two reasons that this may not be the case.
+First, the SRN group did not have a significant difference and the
+collection time of their follow up sample was less than the adenoma
+group. Second, this confounding would not affect the observations where
+models were used since they were built using a different cohort
+{[}Figures 2 \& S2-S3{]}.
 

### 2017-03-24 submission/supplemental_outline_20161024.Rmd

@@
 \newpage
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-
-\blandscape
 
@@
 \elandscape
+
+\newpage
+
+**Table S4: Summary of Adenoma based on Surgery Received**
+
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+
+adn_surg_summary <- read.csv('../data/process/tables/adn_combined_probs_surgery_pvalue_summary.csv', 
+                              header = T, stringsAsFactors = F) %>% 
+  filter(X != "prop_surg_adn_srn") %>% 
+  mutate(variable = c("sobs", "Shannon", "Evenness", "Thetayc", "Adenoma Only\nPositive Probability", "SRN Only\nPositive Probablity"))
+
+st4 <- cbind(adn_surg_summary$variable, format(adn_surg_summary$surg_mean, digits = 1, nsmall = 1), format(adn_surg_summary$no_surg_mean, digits = 1, nsmall = 1), format(adn_surg_summary$surg_sd, digits = 1, nsmall = 1), format(adn_surg_summary$no_surg_sd, digits = 1, nsmall = 1), format(adn_surg_summary$pvalue, scientific = TRUE, digits = 2), format(adn_surg_summary$bh, scientific = TRUE, digits = 2))
+
+colnames(st4) <- c("Variable", "Surgery Mean", "No Surgery Mean", "Surgery SD", "No Surgery SD", "P-value", "BH Corrected")
+
+kable(st4, format = "markdown", align = 'c', padding = 2)
+
+```

### 2017-03-29 submission/manuscript_outline_20161024.Rmd
incorporate Amanda edits

--- 2017-03-24 submission/manuscript_outline_20161024.Rmd
@@
 **Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.
-This study tested the hypothesis that treatment for adenoma or carcinoma results in changes to the bacterial community.
-Specifically, we tried to identify components within the community that were different before and after treatment of adenoma, advanced adenoma (Screen Relevant Neoplasia (SRN)), and carcinoma.
+Despite this, very little is known about what happens to the microbiome after treatment for an adenoma or carcinoma.
+This study tested the hypothesis that treatment for adenoma or carcinoma results in changes towards a normal bacterial community.
+Specifically, we tried to identify components within the community that were different before and after treatment of adenoma, advanced adenoma, and carcinoma.
 
-**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus SRN cases (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-But there was a trend for increasingly less community similarity between initial and follow up samples from adenoma to SRN to carcinoma.
-Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma.
+**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus advanced adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was a trend for increasingly less community similarity, between samples pre- and post-treatment from adenoma to advanced adenoma to carcinoma.
+Despite this, no difference was found in the relative abundance of any specific OTU before and after treatment for adenoma, advanced adenoma, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Using Random Forest models to assess whether changes in post-treatment samples were towards a normal community, only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma.
 The adenoma model used a total of `r length(rownames(adn_imp_vars))` OTUs, the SRN model used a total of `r length(rownames(srn_imp_vars))` OTUs, and the carcinoma model used a total of `r length(rownames(crc_imp_vars))` OTUs.
-A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g.
+A total of `r length(common_vars_summary$otu)` OTUs were common to all three models with many classifying to commensal bacteria (e.g.
 *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
@@
 
-**Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.
-Individuals with carcinoma had more drastic differences to the overall community then those with adenoma.
-Common OTUs to all models were overwhelmingly from commensal bacteria, suggesting that these bacteria may be important to initial polyp formation, development of advanced adenoma, and transition to carcinoma.
+**Conclusions:** Our data partially supports the hypothesis that the microbiome changes after treatment towards a normal community.
+Individuals with carcinoma had more drastic differences to the overall community than those with adenoma.
+Common OTUs to all models were overwhelmingly from commensal bacteria, suggesting that these bacteria may be important in initial polyp formation, development of advanced adenoma, and transition to carcinoma.
 
@@
 However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
+Some of these deaths are due to disease recurrence, with current estimates that 20-30% of those undergoing resection having a CRC recurrence [@hellinger_reoperation_2006].
+This is important since approximately 30-50% of those who have a recurrence will die of CRC [@ryuk_predictive_2014].
+To reduce the burden on the healthcare system, finding ways to better stratify those at highest risk of recurrence is of great importance.
 
-Recent studies in humans have shown that both the microbiome and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].
-Further, bacterial communities have been observed to be altered between normal and tumor tissue [@dejea_microbiota_2014].
-Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.
-Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.
-Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].
-They provide very little information as to whether these communities change and rebound towards normal after successful treatment of adenoma, advanced adenoma (Screen Relevent Neoplasia (SRN)), or carcinoma.
+Recent studies in humans and mouse models have shown that altered membership and structure of the gut microbiome  correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014; @zackular_gut_2013].
+Further, bacterial communities on the mucosa are altered between normal and tumor tissue [@dejea_microbiota_2014].
+Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest the possibility of using biomarkers to diagnose disease.
+Indeed, models created with microbiome data and clinical tests, such as Fecal Immunoglobulin Test (FIT), result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
+While these studies show how changed microbial communities or invasion by inflammatory species can impact CRC progression  [@flynn_metabolic_2016], they provide very little information as to whether these communities change and rebound towards normal after successful treatment of adenoma, advanced adenoma, or carcinoma.
 
-Providing an answer to this question is important because it has far reaching implications on both how the bacterial community causes the formation of more polyps [@zackular_manipulation_2016; @arthur_microbial_2014] and the ability to be able to use the microbiome as a predictive screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Understanding polyp formation and transition to SRN and then carinoma is crucial to being able to understand how to prevent CRC occurence.
-Response of the community to treatment is also eqaully important to predictive models designed for screening purposes since an unresponsive community would provide little additional information for important events, such as recurrence [@hassan_efficacy_2016].
+Providing an answer to this question is important because it has far reaching implications on both how the bacterial community causes the formation of more polyps [@zackular_manipulation_2016; @arthur_microbial_2014] and the ability to use the microbiome as a predictive screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
+Understanding polyp formation and transition to advanced adenoma and then carcinoma is crucial to being able to understand how to prevent CRC occurrence.
+Response of the community to treatment is also equally important to predictive models designed for screening purposes since an unresponsive community would provide little additional information for important events, such as recurrence [@hassan_efficacy_2016].
 
-Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes to the microbiome in those with adenoma, SRN, and carincoma.
-First, we assessed differences between initial and follow up samples in adenoma, SRN, or carinoma using alpha or beta diversity metrics.
-Second, we explored whether models built to classify adenoma, SRN, or carcinoma versus normal were able to identify specific community members that differed between initial and follow up.
-We also used these models to assess whether changes in the community were toward a more normal micorbiome.
-Finally, we assessed both whether surgery for adenomas and SRN provided larger community changes or whether chemotherapy or radiation provided additive changes to the microbiome over surgerical resection.
-This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists after such interventions.
+Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes to the microbiome in those with adenoma, advanced adenoma, and carincoma.
+First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics.
+Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between initial and follow up.
+We also used these models to assess whether changes in the community were toward a more normal microbiome.
+Finally, we assessed both whether surgery for adenomas and SRN provided larger community changes or whether chemotherapy or radiation provided additive changes to the microbiome over surgical resection.
+This study aims to provide evidence as to whether the altered CRC microbiome persists or shifts back towards normal after such interventions.
 
@@
 ***The Bacterial Community:*** 
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma (n = `r change_summary["thetayc", "adn_n"]`), SRN (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between their initial and follow up samples.
-We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
-Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to SRN (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].
-The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), SRN [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was no difference between initial and follow up samples when investigating alpha diversity metrics for adenoma, SRN, or carcinoma for any metric tested [Table S1].
-Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for adenoma, SRN, or carcinoma only [Figure S1].
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- or post-treatment.
+We found that carcinoma patients had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
+Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of the samples pre- and post-treatment from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].
+The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Interestingly, when pre- and post-treatment samples were compared, regardless of whether they were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was no difference between pre- and post-treatment samples when investigating alpha diversity metrics for adenoma, advanced adenoma, or carcinoma [Table S1].
+Additionally, there was also no difference in the relative abundance of any specific OTU between pre- and post-treatment samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].
 
-***Adenoma Model***
-The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification havin an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].
+***Adenoma Model:***
+Using normal and adenoma individauls from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal.
+The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].
 There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A].
-There was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was no significant decrease in the positive probability of adenoma between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].
+There was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was also no significant decrease in the positive probability of adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].
 
-***SRN Model***
+***Advanced Adenoma Model:***
+Using normal and advanced adenoma individuals from a separate cohort we created an advanced adenoma model to help assess whether those with advanced adenoma in this cohort changed towards normal.
 The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].
-There was a total of `r length(rownames(srn_imp_vars))` OTUs in the SRN model [Figure S3B].
+There was a total of `r length(rownames(srn_imp_vars))` OTUs in the advanced adenoma model [Figure S3B].
 Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal.
-Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was no significant decrease in the positive probability of SRN between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].
+Also similar to the adenoma model there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05) and no significant decrease in the positive probability of advanced adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].
 
-***Carcinoma Model***
+***Carcinoma Model:***
+Using normal and carcinoma individuals from a separate cohort we created a carcinoma model to help assess whether those with carcinoma in this cohort changed towards normal.
 The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].
-Interestingly, the AUCs improved from adenoma to SRN to carcinoma [Figure 2].
-There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carinoma model [Figure S3C].
-Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* appear to be important for carcinoma classification [Figure S3C].
-Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was a significant decrease in the positive probability of carcinoma between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.
-The one indivdiual still positive for carcinoma after treatment had an increase in carcinoma positive proability on follow up [Figure 2C].
+Interestingly, the AUCs improved from adenoma to advanced adenoma to carcinoma [Figure 2].
+There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carcinoma model [Figure S3C].
+Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* are also important for carcinoma classification [Figure S3C].
+Also, like the adenoma and advanced adenoma models there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was a significant decrease in the positive probability of carcinoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or advanced adenoma.
+The one individual still positive for carcinoma after treatment had an increase in carcinoma positive proability between their pre- and post-treatment sample [Figure 2C].
 
-***Adenoma, SRN, and Carcinoma Common OTUs***
-We next wanted to know what predictors within the adenoma, SRN, and carcinoma models were similar to each other.
-The main purpose was to identify which OTUs could be important at all three stages of disease.
+***Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:***
+Next, to identify which OTUs are important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models.
 When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.
 Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*.
-These along with the vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].
+The vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].
 
-***Treatment Affects on Community***
-After observing these these changes from treatment we assessed whether chemotherapy or radiation, in the carcinoma group, and surgery, in the adenoma group, impacted the observed results.
-In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S3].
-For the adenoma group there was a single difference in observed OTUs (sobs) between those that received surgery and those that did not (P-value `r ifelse(min(comb_adn_surg_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S4].
-For the surgery comparison, adenoma and SRN were combined due to the low number of surgery occurances in these two groups.
-There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups (P-value `r ifelse(comb_adn_surg_summary["prop_surg_adn_srn", "pvalue"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.
+***Treatment Affects on Community:***
+After observing these changes from treatment we assessed the possible confounding impact of chemotherapy or radiation, in the carcinoma group, and surgical resection, in the adenoma group, on the observed results.
+In the carcinoma group neither chemotherapy nor radiation provided any additional change from initial sample over surgery alone (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S3].
+For the adenoma group there was a single difference in observed OTUs (sobs) between those that received surgerical resection and those that had regular polyp removal (P-value `r ifelse(min(comb_adn_surg_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S4].
+For the surgical resection comparison, adenoma and advanced adenoma were combined due to the low number of resection occurances in these two groups.
+Using a fisher exact test there was no difference in the proportion of those receiving surgerical resection between the adenoma and advanced adenoma groups (P-value `r ifelse(comb_adn_surg_summary["prop_surg_adn_srn", "pvalue"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+This data suggests that the microbiome changes observed in the carcinoma group were mostly a result from surgerical resection and not from chemotherapy or radiation.
 
@@
 ## Discussion
-This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
-Here we show that the bacterial community changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide an additive change.
-Although some of the important OTUs classified to genera from bacteria considered the usual suspects (e.g.
+This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differs and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
+Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide an additive change.
+Although some of the important OTUs classified to genera from bacteria considered the "usual suspects"" (e.g.
 *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.
-The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, SRN, and carcinoma models.
+The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models.
 This suggests that members within the commensal community may be the first that change during CRC pathogenesis.
@@
 
-Unlike previous studies on the microbiome and CRC, ours focuses both on identifying commonalities and differences within adenoma, SRN, and carcinoma groups before and after treatment.
-Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3].
-The majority of important OTUs taxonmically classified to commensal bacteria [Figure S3].
-Although these changes may be subtle, due to the lack of signifcant difference in the bacterial community before and after treatment in adenoma and SRN [Figure 1], they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
+Unlike previous studies on the microbiome and CRC, ours focuses on comparing the microbiome during recovery from treatment in adenoma, advanced adenoma, and carcinoma groups.
+Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], the majority of important OTUs taxonmically classified to commensal bacteria [Figure S3].
+Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
 
-Many of the common OTUs that we identified taxnomically classified to potential butyrate producers (e.g.
+Many of the common OTUs that we identified taxonomically classified to potential butyrate producers (e.g.
 *Clostridiales*, *Roseburia*,  and *Anaerostipes*) [Table S2].
@@
 Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.
+Within this context the commensal community may be an important modifiable risk factor for monitoring and preventing CRC recurrence after treatment.
 
-A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and SRN (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).
+A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).
 These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time.
@@
 However, there are two reasons that this may not be the case.
-First, the SRN group did not have a significant difference and the collection time of their follow up sample was less than the adenoma group.
+First, the advanced adenoma group had a higher dissimilarity than adenoma but lower dissimilarity then carcinoma and the collection time to their follow up samples was less than the adenoma group.
 Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].
 
-Another limitation was that we do not know whether individuals who were still classified as positive by the carcinoma model eventually had a subsequent CRC recurrence.
+Another limitation of this study was that it drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.
+Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.
+We have done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
+
+Building off of our results, an area for future research, is that within our study we do not know whether individuals who were still classified as positive by the carcinoma model eventually had a subsequent CRC recurrence.
 This information would help to strengthen the case for this model keeping numerous individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having carcinoma.
-This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.
-Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.
-We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
+It would also provide additional evidence that the microbiome could be used as a risk stratification tool in monitoring recurrence risk and whether different interventions could potentially change this community and lower the probability of future reccurence.
 
-Despite these shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma, SRN or carcinoma treatment.
-Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of carcinoma after treatment provides evidence that there are specific signatures, attributable to both inflammtory and resident commensal organisms, associated with treatment.
-Our data provides evidence that commensal bacteria may be important in the development of polyps and potentially the transition from adenoma to carcinoma.
+Despite the stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment.
+Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of carcinoma after treatment provides evidence that there are specific signatures, attributable to both inflammatory and resident commensal organisms, associated with treatment.
+Our data provides evidence that commensal bacteria may be important in the development of polyps, potential transition of advanced adenoma to carcinoma, and recovery of the microbiome in CRC following treatment.
 
@@
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
-The bacterial community has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013].
+The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013].
 Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].
@@
 Each group was randomly assigned to avoid biases based on sample collection location.
-The initial and follow up samples were sequenced on the same run.
+The pre- and post-treatment samples were sequenced on the same run.
 
@@
 
-***Lesion Model Creation:***
-The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on an independent data set of  `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
-This model was then applied to our `r length(rownames(good_metaf))`-person cohort.
-It should be noted that all individuals with an adenoma or carcinoma were grouped together to form the lesion group and the model was not created to find differences between normal, adenoma, and carcinoma but rather differences between both adenoma and carcinoma versus normal.
+***Model Building:***
+The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used.
+The adenoma model classified normal versus adenoma, advanced adenoma was normal versus advanced adenoma, and carcinoma was normal versus carcinoma.
+The toal number of individuals in this data set was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
+There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` indivdiuals in the carcinoma model [Table 1].
+Each model was then applied to our `r length(rownames(good_metaf))`-person cohort testing prediction of adenoma intial (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])`) versus adenoma follow up (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])`), advanced adenoma initial (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])`) versus advanced adenoma follow up (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])`), carcinoma initial (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])`) versus carcinoma follow up (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")[, "followUp"])`).
 
@@
 This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
-Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.
-For each a 20 repeated 10-fold cross validation was performed on 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
+Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where normal and adenoma, normal and advanced adenoma, or normal and carcinoma were represented in the same proportion within both the whole data set and the 80/20 split.
+For each of the different splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
 The resulting model was then tested on the hold out data obtained from the 20% component.
+All three models had an optimized mtry of `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
+
 Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run.
@@
 The final collated list of variables was then run through the mtry optimization again.
-Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
+Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
+For all three models the final optimized mtry was `r `ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`
 
@@
 
-***Treatment Model Creation:***
-We also investigated whether a model could be created that could identify pre- (initial) and post- (follow up) treatment samples.
-The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.
-Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.
-Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial sample prior to treatment.
-
 ***Statistical Analysis:***
@@
 Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
-Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.
+Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxson ranked sum test.
 Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
@@
 ***Analysis Overview:***
-We first tested for any differences based on whether the individual had an adenoma or carcinoma.
-This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing all OTUs, and investigating the relative abundance of genera from previously associated CRC bacteria (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*).
-Next, the lesion model was tested for accuracy in prediction and whether it reduced the positive probability of lesion in follow up samples.
-We then used the treatment model to assess whether it could classify samples better than the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.
-Common OTUs were found for the two different models used to assess which were important for both models.
-Finally, differences between those receiving chemotherapy and radiation versus those who received neither were tested.
+We first tested whether there were any differences between pre- and post-treatment samples in alpha and beta diversity based on adenoma, SRN, or carcinoma.
+We then tested all OTUs for differences between pre- and post-treatment samples.
+We next used our specific models for adenoma, SRN, and carcinoma to test classification accuracy, response towards a normal microbiome, and common OTUs used across models.
+Finally, for the adenoma group differences between those that received surgery or not was tested while for the carcinoma group differences between those receiving chemotherapy and radiation was tested.
 
@@
 
-**Figure 1: General Differences between Adenoma, SRN, and Carcinoma Groups After Treatment.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
-Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).
-B) NMDS of the initial and follow up samples for the adenoma group.
-C) NMDS of the initial and follow up samples for the carcinoma group.
+**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+B) NMDS of the pre- and post-treatment samples for the adenoma group.
+C) NMDS of the pre- and post-treatment samples for the advanced adenoma group.
+D) NMDS of the pre- and post-treatment samples for the carcinoma group.
 
@@
 C) Positive probability change from initial to follow up sample in those with carcinoma..
-
-**Figure 3: The Treatment Model.**  B)   C) Positive probability change from initial to follow up sample in those with carcinoma.
-D) Positive probability change from initial to follow up sample of those with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
 
@@
 
-**Figure S2: ROC Curves of the Adenoma, SRN, and Carcinoma Models.**  A) Adenoma ROC curve: The light greenshaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.**  A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) SRN ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S3: Summary of Important Variables for the Adenoma, SRN, and Carcinoma Models.**  A) MDA of the most important variables in the adenoma model.
+**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.**  A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
@@
 C) MDA of the most important variables in the carcinoma model.
-The dark red point represents the mean and the lighter redpoints are the value of each of the 100 different runs.
+The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.
 

### 2017-03-29 submission/manuscript_outline_20161024.tex
incorporate Amanda edits

@@
 problem and research suggests a correlation between the fecal bacterial
-microbiome and CRC. This study tested the hypothesis that treatment for
-adenoma or carcinoma results in changes to the bacterial community.
-Specifically, we tried to identify components within the community that
-were different before and after treatment of adenoma, advanced adenoma
-(Screen Relevant Neoplasia (SRN)), and carcinoma.
+microbiome and CRC. Despite this, very little is known about what
+happens to the microbiome after treatment for an adenoma or carcinoma.
+This study tested the hypothesis that treatment for adenoma or carcinoma
+results in changes towards a normal bacterial community. Specifically,
+we tried to identify components within the community that were different
+before and after treatment of adenoma, advanced adenoma, and carcinoma.
 
@@
 in response to treatment for carcinoma versus adenoma (P-value
-\textless{} 0.05) but not carcinoma versus SRN cases (P-value
-\textgreater{} 0.05). But there was a trend for increasingly less
-community similarity between initial and follow up samples from adenoma
-to SRN to carcinoma. Despite this, no difference was found in the
-relative abundance of any OTU before and after treatment for adenoma,
-SRN, or carcinoma groups (P-value \textgreater{} 0.05). Using Random
-Forest models to assess whether changes in follow up samples were
-towards a normal community showed that only those with carcinoma had a
-significant decrease in positive probability (P-value \textless{} 0.05);
-providing further evidence that treatment has the greatest effect in
-those with carcinoma. The adenoma model used a total of 62 OTUs, the SRN
-model used a total of 61 OTUs, and the carcinoma model used a total of
-59 OTUs. A total of 26 OTUs were common to both models with many
-classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
-\emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
-\emph{Dorea}). Both chemotherapy and radiation did not provide any
-additional changes to the bacterial community in those treated for
-carcinoma (Pvalue \textgreater{} 0.05).
+\textless{} 0.05) but not carcinoma versus advanced adenoma (P-value
+\textgreater{} 0.05). There was a trend for increasingly less community
+similarity, between samples pre- and post-treatment from adenoma to
+advanced adenoma to carcinoma. Despite this, no difference was found in
+the relative abundance of any specific OTU before and after treatment
+for adenoma, advanced adenoma, or carcinoma groups (P-value
+\textgreater{} 0.05). Using Random Forest models to assess whether
+changes in post-treatment samples were towards a normal community, only
+those with carcinoma had a significant decrease in positive probability
+(P-value \textless{} 0.05); providing further evidence that treatment
+has the greatest effect in those with carcinoma. The adenoma model used
+a total of 62 OTUs, the SRN model used a total of 61 OTUs, and the
+carcinoma model used a total of 59 OTUs. A total of 26 OTUs were common
+to all three models with many classifying to commensal bacteria (e.g.
+\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Anaerostipes},
+\emph{Blautia}, and \emph{Dorea}). Both chemotherapy and radiation did
+not provide any additional changes to the bacterial community in those
+treated for carcinoma (Pvalue \textgreater{} 0.05).
 
 \textbf{Conclusions:} Our data partially supports the hypothesis that
-the bacterial community changes after treatment. Individuals with
-carcinoma had more drastic differences to the overall community then
-those with adenoma. Common OTUs to all models were overwhelmingly from
-commensal bacteria, suggesting that these bacteria may be important to
-initial polyp formation, development of advanced adenoma, and transition
-to carcinoma.
+the microbiome changes after treatment towards a normal community.
+Individuals with carcinoma had more drastic differences to the overall
+community than those with adenoma. Common OTUs to all models were
+overwhelmingly from commensal bacteria, suggesting that these bacteria
+may be important in initial polyp formation, development of advanced
+adenoma, and transition to carcinoma.
 
@@
 {[}1{]}. However, despite this improvement there are still approximately
-50,000 deaths from the disease per year {[}2{]}.
+50,000 deaths from the disease per year {[}2{]}. Some of these deaths
+are due to disease recurrence, with current estimates that 20-30\% of
+those undergoing resection having a CRC recurrence {[}3{]}. This is
+important since approximately 30-50\% of those who have a recurrence
+will die of CRC {[}4{]}. To reduce the burden on the healthcare system,
+finding ways to better stratify those at highest risk of recurrence is
+of great importance.
 
-Recent studies in humans have shown that both the microbiome and
-specific members within it correlate with CRC pathogenesis {[}3,4{]}.
-Further, bacterial communities have been observed to be altered between
-normal and tumor tissue {[}5{]}. Mouse models of CRC have further
-demonstrated the importance of the microbiome, both on a community
-{[}3,6{]} and species level {[}4{]}, for tumorgenesis. Collectively,
-these studies provide a tantalizing link between our gut bacteria and
-CRC and suggest that biomarkers using our microbes could be developed.
-Indeed, builidng models using 16S rRNA gene sequencing along with
-clinical tests such as Fecal Immunoglobulin Test (FIT) result in good
-predictions of CRC {[}7,8{]}. Although these studies show how our gut
-bacteria can impact CRC progression via a changed community or invasion
-by more inflammatory bacteria {[}9{]}. They provide very little
-information as to whether these communities change and rebound towards
-normal after successful treatment of adenoma, advanced adenoma (Screen
-Relevent Neoplasia (SRN)), or carcinoma.
+Recent studies in humans and mouse models have shown that altered
+membership and structure of the gut microbiome correlate with CRC
+pathogenesis {[}5--7{]}. Further, bacterial communities on the mucosa
+are altered between normal and tumor tissue {[}8{]}. Collectively, these
+studies provide a tantalizing link between our gut bacteria and CRC and
+suggest the possibility of using biomarkers to diagnose disease. Indeed,
+models created with microbiome data and clinical tests, such as Fecal
+Immunoglobulin Test (FIT), result in good predictions of CRC {[}9,10{]}.
+While these studies show how changed microbial communities or invasion
+by inflammatory species can impact CRC progression {[}11{]}, they
+provide very little information as to whether these communities change
+and rebound towards normal after successful treatment of adenoma,
+advanced adenoma, or carcinoma.
 
@@
 reaching implications on both how the bacterial community causes the
-formation of more polyps {[}3,4{]} and the ability to be able to use the
-microbiome as a predictive screening tool {[}7,8{]}. Understanding polyp
-formation and transition to SRN and then carinoma is crucial to being
-able to understand how to prevent CRC occurence. Response of the
-community to treatment is also eqaully important to predictive models
+formation of more polyps {[}5,6{]} and the ability to use the microbiome
+as a predictive screening tool {[}9,10{]}. Understanding polyp formation
+and transition to advanced adenoma and then carcinoma is crucial to
+being able to understand how to prevent CRC occurrence. Response of the
+community to treatment is also equally important to predictive models
 designed for screening purposes since an unresponsive community would
 provide little additional information for important events, such as
-recurrence {[}10{]}.
+recurrence {[}12{]}.
 
@@
 the hypothesis that treatment causes detectable changes to the
-microbiome in those with adenoma, SRN, and carincoma. First, we assessed
-differences between initial and follow up samples in adenoma, SRN, or
-carinoma using alpha or beta diversity metrics. Second, we explored
-whether models built to classify adenoma, SRN, or carcinoma versus
-normal were able to identify specific community members that differed
-between initial and follow up. We also used these models to assess
-whether changes in the community were toward a more normal micorbiome.
-Finally, we assessed both whether surgery for adenomas and SRN provided
-larger community changes or whether chemotherapy or radiation provided
-additive changes to the microbiome over surgerical resection. This study
-helps to provide evidence as to whether treatment can influence the
-community and if the CRC microbiome, identified in previous studies,
-persists after such interventions.
+microbiome in those with adenoma, advanced adenoma, and carincoma.
+First, we assessed differences between samples pre- and post-treatment
+in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity
+metrics. Second, we explored whether models built to classify adenoma,
+advanced adenoma, or carcinoma versus normal were able to identify
+specific community members that differed between initial and follow up.
+We also used these models to assess whether changes in the community
+were toward a more normal microbiome. Finally, we assessed both whether
+surgery for adenomas and SRN provided larger community changes or
+whether chemotherapy or radiation provided additive changes to the
+microbiome over surgical resection. This study aims to provide evidence
+as to whether the altered CRC microbiome persists or shifts back towards
+normal after such interventions.
 
@@
 \textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
-tested whether those with adenoma (n = 22), SRN (n = 19), or carcinoma
-(n = 26) had any broad differences between their initial and follow up
-samples. We found that those with carcinoma had a more dissimilar
-bacterial community between their initial and follow up sample than
-those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. Although
-no significant differences were observed between SRN and carcinoma there
-was an increase in the dissimilarity of the initial and follow up
-samples from adenoma (0.55 ± 0.21 (mean ± SD)) to SRN (0.65 ± 0.25) to
-carcinoma (0.78 ± 0.15) {[}Figure 1A{]}. The bacterial community
-structure before and after surgery was visualized using NMDS for adenoma
-{[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05), SRN {[}Figure 1C{]}
+tested whether the microbiome in patients with adenoma (n = 22),
+advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
+differences between pre- or post-treatment. We found that carcinoma
+patients had a more dissimilar bacterial community between their initial
+and follow up sample than those with adenoma (P-value \textless{} 0.001)
+{[}Figure 1A{]}. Although no significant differences were observed
+between advanced adenoma and carcinoma there was an increase in the
+dissimilarity of the samples pre- and post-treatment from adenoma (0.55
+± 0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78
+± 0.15) {[}Figure 1A{]}. The bacterial community structure before and
+after surgery was visualized using NMDS for adenoma {[}Figure 1B{]}
+(PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
 (PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
-(PERMANOVA \textless{} 0.05). Interestingly, when initial and follow up
-samples were compared, regardless of whether the lesions were adenoma or
-carcinoma, there was no significant overall difference in beta diversity
-(PERMANOVA \textgreater{} 0.05). There was no difference between initial
-and follow up samples when investigating alpha diversity metrics for
-adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
-Additionally, there was also no difference in the relative abundance of
-any OTU between initial and follow up samples for adenoma, SRN, or
-carcinoma only {[}Figure S1{]}.
+(PERMANOVA \textless{} 0.05). Interestingly, when pre- and
+post-treatment samples were compared, regardless of whether they were
+adenoma or carcinoma, there was no significant overall difference in
+beta diversity (PERMANOVA \textgreater{} 0.05). There was no difference
+between pre- and post-treatment samples when investigating alpha
+diversity metrics for adenoma, advanced adenoma, or carcinoma {[}Table
+S1{]}. Additionally, there was also no difference in the relative
+abundance of any specific OTU between pre- and post-treatment samples
+for adenoma, advanced adenoma, or carcinoma only {[}Figure S1{]}.
 
-\textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
-20 repeated 10 fold cross-validation was 0.62 - 0.72 with the AUC of the
-model used for classification havin an AUC of 0.65 {[}Figure S2A{]}.
-There was a total of 62 OTUs in this model with the vast majority
-classifying to bacteria typically thought of as commensal {[}Figure
-S3A{]}. There was a significant difference between the actual and model
-predicted group stratification calls (P-value \textless{} 0.05). There
-was no significant decrease in the positive probability of adenoma
-between initial and follow up samples (P-value \textgreater{} 0.05)
-{[}Figure 2A{]}.
+\textbf{\emph{Adenoma Model:}} Using normal and adenoma individauls from
+a separate cohort we created an adenoma model to help assess whether
+those with adenoma in this cohort changed towards normal. The range of
+model AUC's from 100 runs of 20 repeated 10-fold cross-validation was
+0.62 - 0.72 with the AUC of the model used for classification having an
+AUC of 0.65 {[}Figure S2A{]}. There was a total of 62 OTUs in this model
+with the vast majority classifying to bacteria typically thought of as
+commensal {[}Figure S3A{]}. There was a significant difference between
+the actual and predicted disease classification (P-value \textless{}
+0.05). There was also no significant decrease in the positive
+probability of adenoma between pre- and post-treatment samples (P-value
+\textgreater{} 0.05) {[}Figure 2A{]}.
 
-\textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
-repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
+\textbf{\emph{Advanced Adenoma Model:}} Using normal and advanced
+adenoma individuals from a separate cohort we created an advanced
+adenoma model to help assess whether those with advanced adenoma in this
+cohort changed towards normal. The range of model AUC's from 100 runs of
+20 repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
 model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
-There was a total of 61 OTUs in the SRN model {[}Figure S3B{]}. Similar
-to the adenoma model the vast majority of OTUs classified to bacteria
-typically thought of as commensal. Also similar to the adenoma model
-there was a significant difference between the actual and model
-predicted group stratification calls (P-value \textless{} 0.05). There
-was no significant decrease in the positive probability of SRN between
-initial and follow up samples (P-value \textgreater{} 0.05) {[}Figure
-2B{]}.
+There was a total of 61 OTUs in the advanced adenoma model {[}Figure
+S3B{]}. Similar to the adenoma model the vast majority of OTUs
+classified to bacteria typically thought of as commensal. Also similar
+to the adenoma model there was a significant difference between the
+actual and predicted disease classification (P-value \textless{} 0.05)
+and no significant decrease in the positive probability of advanced
+adenoma between pre- and post-treatment samples (P-value \textgreater{}
+0.05) {[}Figure 2B{]}.
 
-\textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
-of 20 repeated 10 fold cross-validation was 0.84 - 0.9 with the AUC of
-the model used for classification being 0.88 {[}Figure S2C{]}.
-Interestingly, the AUCs improved from adenoma to SRN to carcinoma
-{[}Figure 2{]}. There was a total of 59 OTUs in the carinoma model
-{[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
-majority of OTUs classified to bacteria typically thought of as
-commensal but OTUs that also classified to \emph{Fusobacterium},
-\emph{Porphyromonas}, and \emph{Parvimonas} appear to be important for
-carcinoma classification {[}Figure S3C{]}. Also similar to the adenoma
-and SRN models there was a significant difference between the actual and
-model predicted group stratification calls (P-value \textless{} 0.05).
-There was a significant decrease in the positive probability of
-carcinoma between initial and follow up samples (P-value \textless{}
-0.05) {[}Figure 2C{]}; suggesting that the carcinoma samples changed
-towards normal after treatment, unlike either adenoma or SRN. The one
-indivdiual still positive for carcinoma after treatment had an increase
-in carcinoma positive proability on follow up {[}Figure 2C{]}.
+\textbf{\emph{Carcinoma Model:}} Using normal and carcinoma individuals
+from a separate cohort we created a carcinoma model to help assess
+whether those with carcinoma in this cohort changed towards normal. The
+range of model AUC's from 100 runs of 20 repeated 10 fold
+cross-validation was 0.84 - 0.9 with the AUC of the model used for
+classification being 0.88 {[}Figure S2C{]}. Interestingly, the AUCs
+improved from adenoma to advanced adenoma to carcinoma {[}Figure 2{]}.
+There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
+Similar to the adenoma and SRN models the vast majority of OTUs
+classified to bacteria typically thought of as commensal but OTUs that
+also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
+\emph{Parvimonas} are also important for carcinoma classification
+{[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
+there was a significant difference between the actual and predicted
+disease classification (P-value \textless{} 0.05). There was a
+significant decrease in the positive probability of carcinoma between
+pre- and post-treatment samples (P-value \textless{} 0.05) {[}Figure
+2C{]}; suggesting that the carcinoma samples changed towards normal
+after treatment, unlike either adenoma or advanced adenoma. The one
+individual still positive for carcinoma after treatment had an increase
+in carcinoma positive proability between their pre- and post-treatment
+sample {[}Figure 2C{]}.
 
-\textbf{\emph{Adenoma, SRN, and Carcinoma Common OTUs}} We next wanted
-to know what predictors within the adenoma, SRN, and carcinoma models
-were similar to each other. The main purpose was to identify which OTUs
-could be important at all three stages of disease. When we compared the
-three different models with each other there were a total of 26 common
-OTUs. Some of the most common taxonomic identifications belonged to
-\emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
-\emph{Lachnospiraceae}, and \emph{Dorea}. These along with the vast
-majority of the OTUs that were common between these models had
-classifications to bacteria typically thought of as commensal {[}Table
-S2{]}.
+\textbf{\emph{Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:}}
+Next, to identify which OTUs are important at all stages of disease, we
+identified common predictive OTUs within the adenoma, advanced adenoma,
+and carcinoma models. When we compared the three different models with
+each other there were a total of 26 common OTUs. Some of the most common
+taxonomic identifications belonged to \emph{Bacteroides},
+\emph{Blautia}, \emph{Anaerostipes}, \emph{Lachnospiraceae}, and
+\emph{Dorea}. The vast majority of the OTUs that were common between
+these models had classifications to bacteria typically thought of as
+commensal {[}Table S2{]}.
 
-\textbf{\emph{Treatment Affects on Community}} After observing these
-these changes from treatment we assessed whether chemotherapy or
-radiation, in the carcinoma group, and surgery, in the adenoma group,
-impacted the observed results. In the carcinoma group neither
-chemotherapy or radiation provided any additive change from initial
-sample over surgery alone (P-value \textgreater{} 0.05) {[}Table S3{]}.
-For the adenoma group there was a single difference in observed OTUs
-(sobs) between those that received surgery and those that did not
-(P-value \textless{} 0.05) {[}Table S4{]}. For the surgery comparison,
-adenoma and SRN were combined due to the low number of surgery
-occurances in these two groups. There was no difference in the
-proportion of those receiving surgery between the adenoma and SRN groups
-(P-value \textgreater{} 0.05). This data suggests that microbiome
-changes observed in the carcinoma group were mostly a result from
-surgery and not from chemotherapy or radiation.
+\textbf{\emph{Treatment Affects on Community:}} After observing these
+changes from treatment we assessed the possible confounding impact of
+chemotherapy or radiation, in the carcinoma group, and surgical
+resection, in the adenoma group, on the observed results. In the
+carcinoma group neither chemotherapy nor radiation provided any
+additional change from initial sample over surgery alone (P-value
+\textgreater{} 0.05) {[}Table S3{]}. For the adenoma group there was a
+single difference in observed OTUs (sobs) between those that received
+surgerical resection and those that had regular polyp removal (P-value
+\textless{} 0.05) {[}Table S4{]}. For the surgical resection comparison,
+adenoma and advanced adenoma were combined due to the low number of
+resection occurances in these two groups. Using a fisher exact test
+there was no difference in the proportion of those receiving surgerical
+resection between the adenoma and advanced adenoma groups (P-value
+\textgreater{} 0.05). This data suggests that the microbiome changes
+observed in the carcinoma group were mostly a result from surgerical
+resection and not from chemotherapy or radiation.
 
@@
 considered both how the bacterial community between those with and
-without CRC differ and how it might be used as an early screening tool
-{[}7,8,11--13{]}. Here we show that the bacterial community changes
-towards normal after treatment for carcinoma and that chemotherapy and
-radiation did not provide an additive change. Although some of the
-important OTUs classified to genera from bacteria considered the usual
-suspects (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
-\emph{Parvimonas}) many did not. The majority of important OTUs had
-taxonomic classifications for resident gut microbes and were common for
-the adenoma, SRN, and carcinoma models. This suggests that members
-within the commensal community may be the first that change during CRC
-pathogenesis. These subtle changes, in turn, could be the first step in
-allowing more inflammatory bacteria to gain a foothold within the colon
-{[}9{]}.
+without CRC differs and how it might be used as an early screening tool
+{[}9,10,13--15{]}. Here we show that the microbiome changes towards
+normal after treatment for carcinoma and that chemotherapy and radiation
+did not provide an additive change. Although some of the important OTUs
+classified to genera from bacteria considered the ``usual suspects''"
+(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
+many did not. The majority of important OTUs had taxonomic
+classifications for resident gut microbes and were common for the
+adenoma, advanced adenoma, and carcinoma models. This suggests that
+members within the commensal community may be the first that change
+during CRC pathogenesis. These subtle changes, in turn, could be the
+first step in allowing more inflammatory bacteria to gain a foothold
+within the colon {[}11{]}.
 
-Unlike previous studies on the microbiome and CRC, ours focuses both on
-identifying commonalities and differences within adenoma, SRN, and
-carcinoma groups before and after treatment. Although there were
-differences for genera associated with specific bacterium linked with
-CRC {[}Figure S3{]}. The majority of important OTUs taxonmically
-classified to commensal bacteria {[}Figure S3{]}. Although these changes
-may be subtle, due to the lack of signifcant difference in the bacterial
-community before and after treatment in adenoma and SRN {[}Figure 1{]},
-they support the hypothesis that the first members of the community to
-change and potentially stay changed even after treatment are those that
-are commensal bacteria.
+Unlike previous studies on the microbiome and CRC, ours focuses on
+comparing the microbiome during recovery from treatment in adenoma,
+advanced adenoma, and carcinoma groups. Although there were differences
+for genera associated with specific bacterium linked with CRC {[}Figure
+S3{]}, the majority of important OTUs taxonmically classified to
+commensal bacteria {[}Figure S3{]}. Although these changes may be
+subtle, due to the lack of significant differences in the bacterial
+community pre- and post-treatment in adenoma and advanced adenoma
+{[}Figure 1{]}; they support the hypothesis that the first members of
+the community to change and potentially stay changed even after
+treatment are those that are commensal bacteria.
 
-Many of the common OTUs that we identified taxnomically classified to
+Many of the common OTUs that we identified taxonomically classified to
 potential butyrate producers (e.g. \emph{Clostridiales},
@@
 \emph{Bacteroides}). Both butyrate and polyphenols are thought to be
-protective against cancer, in part by reducing inflammation {[}14{]}.
+protective against cancer, in part by reducing inflammation {[}16{]}.
 These protective compounds are derived from the breakdown of fiber,
@@
 to be increased in those with high non-meat based protein consumption
-{[}15{]}. High protein consumption in general has been linked with an
-increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
+{[}17{]}. High protein consumption in general has been linked with an
+increased CRC risk {[}18{]}. Conversely, \emph{Bacteroides} are
 inhibited by polyphenols which are derived from fruits and vegetables
-{[}17{]}. Our data fits with the hypothesis that the microbial
+{[}19{]}. Our data fits with the hypothesis that the microbial
 metabolites from breakdown products within our own diet could not only
 help to shape the existing community but also have an effect on CRC risk
-and disease progression.
+and disease progression. Within this context the commensal community may
+be an important modifiable risk factor for monitoring and preventing CRC
+recurrence after treatment.
 
@@
 in the time elapsed in the collection of the follow up sample between
-adenoma or SRN versus carcinoma (P-value \textless{} 0.05), with time
-passed being less for adenoma (255 ± 42 days) and SRN (250 ± 41) than
-carcinoma (351 ± 102). These results would indicate that some of the
-differences observed between the carcinoma and adenoma groups could be
-due to differences in collection time. Specifically, it could confound
-the observation that carcinomas changed more than adenomas {[}Figure
-1{]}. However, there are two reasons that this may not be the case.
-First, the SRN group did not have a significant difference and the
-collection time of their follow up sample was less than the adenoma
+adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
+with time passed being less for adenoma (255 ± 42 days) and advanced
+adenoma (250 ± 41) than carcinoma (351 ± 102). These results would
+indicate that some of the differences observed between the carcinoma and
+adenoma groups could be due to differences in collection time.
+Specifically, it could confound the observation that carcinomas changed
+more than adenomas {[}Figure 1{]}. However, there are two reasons that
+this may not be the case. First, the advanced adenoma group had a higher
+dissimilarity than adenoma but lower dissimilarity then carcinoma and
+the collection time to their follow up samples was less than the adenoma
 group. Second, this confounding would not affect the observations where
@@
 
-Another limitation was that we do not know whether individuals who were
-still classified as positive by the carcinoma model eventually had a
-subsequent CRC recurrence. This information would help to strengthen the
-case for this model keeping numerous individuals above the cutoff
-threshold even though at follow up they were diagnosed as no longer
-having carcinoma. This study also drew heavily from those with Caucasian
-ancestry making it possible that the observations may not be
-representative of those with either Asian or African ancestry. Although
-our training and test set are relatively large we still run the risk of
-over-fitting or having a model that may not be representative of other
-populations. We've done our best to safeguard against this by not only
-running 10-fold cross validation but also having over 100 different
-80/20 splits to try and mimic the type of variation that might be
-expected to occur.
+Another limitation of this study was that it drew heavily from those
+with Caucasian ancestry making it possible that the observations may not
+be representative of those with either Asian or African ancestry.
+Although our training and test set are relatively large we still run the
+risk of over-fitting or having a model that may not be representative of
+other populations. We have done our best to safeguard against this by
+not only running 10-fold cross validation but also having over 100
+different 80/20 splits to try and mimic the type of variation that might
+be expected to occur.
 
-Despite these shortcomings our findings add to the existing scientific
-knowledge on CRC and the microbiome: That there is a measurable
-difference in the bacterial community after adenoma, SRN or carcinoma
-treatment. Further, the ability for machine learning algorithms to take
-OTU data and successfully lower positive probability of carcinoma after
-treatment provides evidence that there are specific signatures,
-attributable to both inflammtory and resident commensal organisms,
-associated with treatment. Our data provides evidence that commensal
-bacteria may be important in the development of polyps and potentially
-the transition from adenoma to carcinoma.
+Building off of our results, an area for future research, is that within
+our study we do not know whether individuals who were still classified
+as positive by the carcinoma model eventually had a subsequent CRC
+recurrence. This information would help to strengthen the case for this
+model keeping numerous individuals above the cutoff threshold even
+though at follow up they were diagnosed as no longer having carcinoma.
+It would also provide additional evidence that the microbiome could be
+used as a risk stratification tool in monitoring recurrence risk and
+whether different interventions could potentially change this community
+and lower the probability of future reccurence.
+
+Despite the stated shortcomings our findings add to the existing
+scientific knowledge on CRC and the microbiome: That there is a
+measurable difference in the bacterial community after adenoma, advanced
+adenoma, or carcinoma treatment. Further, the ability for machine
+learning algorithms to take OTU data and successfully lower positive
+probability of carcinoma after treatment provides evidence that there
+are specific signatures, attributable to both inflammatory and resident
+commensal organisms, associated with treatment. Our data provides
+evidence that commensal bacteria may be important in the development of
+polyps, potential transition of advanced adenoma to carcinoma, and
+recovery of the microbiome in CRC following treatment.
 
@@
 \textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
-have been previously reported in Baxter, et al {[}7{]}. Briefly, study
+have been previously reported in Baxter, et al {[}9{]}. Briefly, study
 exclusion involved those who had already undergone surgery, radiation,
@@
 between 1 - 2 weeks after. The bacterial community has been shown to
-normalize back to a pre-colonscopy community within this time period
-{[}18{]}. Our training cohort consisted of a total of 423 individuals
+normalize back to a pre-colonoscopy community within this time period
+{[}20{]}. Our training cohort consisted of a total of 423 individuals
 {[}Table 1{]}. Our study cohort consisted of 67 individuals with an
@@
 \textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
-described by Kozich, et al. {[}19{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}21{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 Each group was randomly assigned to avoid biases based on sample
-collection location. The initial and follow up samples were sequenced on
-the same run.
+collection location. The pre- and post-treatment samples were sequenced
+on the same run.
 
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}19{]}. The general workflow using mothur was:
+previously described {[}21{]}. The general workflow using mothur was:
 Paired-end reads were first merged into contigs, quality filtered,
@@
 
-\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}20{]}
-algorithm was used to create the model used for prediction of lesion
-(adenoma or carcinoma) with the main training and testing of the model
-completed on an independent data set of 423 individuals. This model was
-then applied to our 67-person cohort. It should be noted that all
-individuals with an adenoma or carcinoma were grouped together to form
-the lesion group and the model was not created to find differences
-between normal, adenoma, and carcinoma but rather differences between
-both adenoma and carcinoma versus normal.
+\textbf{\emph{Model Building:}} The Random Forest {[}22{]} algorithm was
+used to create the model used to create the three models used. The
+adenoma model classified normal versus adenoma, advanced adenoma was
+normal versus advanced adenoma, and carcinoma was normal versus
+carcinoma. The toal number of individuals in this data set was 423
+individuals. There were a total of 239 individuals in the adenoma model,
+262 individuals in the advanced adenoma model, and 266 indivdiuals in
+the carcinoma model {[}Table 1{]}. Each model was then applied to our
+67-person cohort testing prediction of adenoma intial (adenoma n = 22)
+versus adenoma follow up (adenoma n = 0), advanced adenoma initial
+(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
+adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
+follow up (carcinoma n = 1).
 
@@
 involved making 100 different 80/20 (train/test) splits of the data
-where normal and lesion were represented in the same proportion within
-both the whole data set and the 80/20 split. For each a 20 repeated
-10-fold cross validation was performed on 80\% component to optimize the
-mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
-Receiver Operator Characteristic). The resulting model was then tested
-on the hold out data obtained from the 20\% component. Assessment of the
-most important OTUs to the model involved counting the number of times
-an OTU was present in the top 10\% of mean decrease in accuracy (MDA)
-for each of the 100 different splits run. This was then followed with
-filtering of this list to variables that were only present in more than
-50\% of these 100 runs. The final collated list of variables was then
-run through the mtry optimization again. Once the ideal mtry was found
-the entire 423 sample set was used to create the final Random Forest
-model on which classifications on the 67-person cohort was completed.
+where normal and adenoma, normal and advanced adenoma, or normal and
+carcinoma were represented in the same proportion within both the whole
+data set and the 80/20 split. For each of the different splits, 20
+repeated 10-fold cross validation was performed on the 80\% component to
+optimize the mtry hyper-parameter by maximizing the AUC (Area Under the
+Curve of the Receiver Operator Characteristic). The resulting model was
+then tested on the hold out data obtained from the 20\% component. All
+three models had an optimized mtry of 2.
+
+Assessment of the most important OTUs to the model involved counting the
+number of times an OTU was present in the top 10\% of mean decrease in
+accuracy (MDA) for each of the 100 different splits run. This was then
+followed with filtering of this list to variables that were only present
+in more than 50\% of these 100 runs. The final collated list of
+variables was then run through the mtry optimization again. Once the
+ideal mtry was found the entire sample set specific to normal versus
+adenoma, normal versus advanced adenoma, or normal versus carcinoma was
+used to create the final Random Forest model on which classifications on
+the 67-person cohort was completed. For all three models the final
+optimized mtry was
+\texttt{r}ifelse(as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
+\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} ==
+as.data.frame(count(srn\_AUC\_run\_summary, best\_mtry)
+\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} \&
+as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
+\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} ==
+as.data.frame(count(crc\_AUC\_run\_summary, best\_mtry)
+\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]},
+as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
+\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]}, ``not the
+same'')`
 
@@
 
-\textbf{\emph{Treatment Model Creation:}} We also investigated whether a
-model could be created that could identify pre- (initial) and post-
-(follow up) treatment samples. The main difference was that only the
-67-person cohort was used at all stages of model building and
-classification. Other than this difference the creation of this model
-and optimization of the mtry hyper-parameter was completed using the
-same procedure as was used for the lesion model. Instead of classifying
-samples as positive or negative of lesion this model classified samples
-as positive or negative for being an initial sample prior to treatment.
-
 \textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}21{]} in the vegan package
+community structure utilized PERMANOVA {[}23{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
-differences between initial and follow up samples utilized a paired
+differences between pre- and post-treatment samples utilized a paired
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}22{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
-\textbf{\emph{Analysis Overview:}} We first tested for any differences
-based on whether the individual had an adenoma or carcinoma. This was
-done by testing initial and follow up samples for differences in alpha
-and beta diversity, testing all OTUs, and investigating the relative
-abundance of genera from previously associated CRC bacteria
-(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
-\emph{Porphyromonas}). Next, the lesion model was tested for accuracy in
-prediction and whether it reduced the positive probability of lesion in
-follow up samples. We then used the treatment model to assess whether it
-could classify samples better than the lesion model and whether it could
-reduce the positive probability of an initial sample in the follow up
-samples. Common OTUs were found for the two different models used to
-assess which were important for both models. Finally, differences
-between those receiving chemotherapy and radiation versus those who
-received neither were tested.
+\textbf{\emph{Analysis Overview:}} We first tested whether there were
+any differences between pre- and post-treatment samples in alpha and
+beta diversity based on adenoma, SRN, or carcinoma. We then tested all
+OTUs for differences between pre- and post-treatment samples. We next
+used our specific models for adenoma, SRN, and carcinoma to test
+classification accuracy, response towards a normal microbiome, and
+common OTUs used across models. Finally, for the adenoma group
+differences between those that received surgery or not was tested while
+for the carcinoma group differences between those receiving chemotherapy
+and radiation was tested.
 
@@
 
-\textbf{Figure 1: General Differences between Adenoma, SRN, and
-Carcinoma Groups After Treatment.} A) A significant difference was found
-between the adenoma and carcinoma group for thetayc (P-value = NULL).
-Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN). B)
-NMDS of the initial and follow up samples for the adenoma group. C) NMDS
-of the initial and follow up samples for the carcinoma group.
+\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
+and Carcinoma Groups After Treatment.} A) A significant difference was
+found between the adenoma and carcinoma group for thetayc (P-value =
+NULL). B) NMDS of the pre- and post-treatment samples for the adenoma
+group. C) NMDS of the pre- and post-treatment samples for the advanced
+adenoma group. D) NMDS of the pre- and post-treatment samples for the
+carcinoma group.
 
@@
 carcinoma..
-
-\textbf{Figure 3: The Treatment Model.} B) C) Positive probability
-change from initial to follow up sample in those with carcinoma. D)
-Positive probability change from initial to follow up sample of those
-with adenoma or advanced adenoma (Screen Relevant Neoplasia (SRN)).
 
@@
 
-\textbf{Figure S2: ROC Curves of the Adenoma, SRN, and Carcinoma
-Models.} A) Adenoma ROC curve: The light greenshaded areas represent the
-range of values of a 100 different 80/20 splits of the test set data and
-the dark green line represents the model using 100\% of the data set and
-what was used for subsequent classification. B) SRN ROC curve: The light
-yellow shaded areas represent the range of values of a 100 different
-80/20 splits of the test set data and the dark yellow line represents
-the model using 100\% of the data set and what was used for subsequent
-classification. C) Carcinoma ROC curve: The light red shaded areas
+\textbf{Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and
+Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
 represent the range of values of a 100 different 80/20 splits of the
-test set data and the dark red line represents the model using 100\% of
-the data set and what was used for subsequent classification.
+test set data and the dark green line represents the model using 100\%
+of the data set and what was used for subsequent classification. B) SRN
+ROC curve: The light yellow shaded areas represent the range of values
+of a 100 different 80/20 splits of the test set data and the dark yellow
+line represents the model using 100\% of the data set and what was used
+for subsequent classification. C) Carcinoma ROC curve: The light red
+shaded areas represent the range of values of a 100 different 80/20
+splits of the test set data and the dark red line represents the model
+using 100\% of the data set and what was used for subsequent
+classification.
 
-\textbf{Figure S3: Summary of Important Variables for the Adenoma, SRN,
-and Carcinoma Models.} A) MDA of the most important variables in the
-adenoma model. The dark green point represents the mean and the lighter
-green points are the value of each of the 100 different runs. B) Summary
-of Important Variables in the SRN model. MDA of the most important
-variables in the SRN model. The dark yellow point represents the mean
-and the lighter yellow points are the value of each of the 100 different
-runs. C) MDA of the most important variables in the carcinoma model. The
-dark red point represents the mean and the lighter redpoints are the
-value of each of the 100 different runs.
+\textbf{Figure S3: Summary of Important Variables for the Adenoma,
+Advanced Adenoma, and Carcinoma Models.} A) MDA of the most important
+variables in the adenoma model. The dark green point represents the mean
+and the lighter green points are the value of each of the 100 different
+runs. B) Summary of Important Variables in the SRN model. MDA of the
+most important variables in the SRN model. The dark yellow point
+represents the mean and the lighter yellow points are the value of each
+of the 100 different runs. C) MDA of the most important variables in the
+carcinoma model. The dark red point represents the mean and the lighter
+red points are the value of each of the 100 different runs.
 
@@
 
+\hypertarget{ref-hellinger_reoperation_2006}{}
+3. Hellinger MD, Santiago CA. Reoperation for recurrent colorectal
+cancer. Clinics in Colon and Rectal Surgery. 2006;19:228--36.
+
+\hypertarget{ref-ryuk_predictive_2014}{}
+4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, et al.
+Predictive factors and the prognosis of recurrence of colorectal cancer
+within 2 years after curative resection. Annals of Surgical Treatment
+and Research. 2014;86:143--51.
+
 \hypertarget{ref-zackular_manipulation_2016}{}
-3. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+5. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
 Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
@@
 \hypertarget{ref-arthur_microbial_2014}{}
-4. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+6. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
 McCafferty J, et al. Microbial genomic analysis reveals the essential
@@
 
+\hypertarget{ref-zackular_gut_2013}{}
+7. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen GY,
+et al. The gut microbiome modulates colon tumorigenesis. mBio.
+2013;4:e00692--00613.
+
 \hypertarget{ref-dejea_microbiota_2014}{}
-5. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+8. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 
-\hypertarget{ref-zackular_gut_2013}{}
-6. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen GY,
-et al. The gut microbiome modulates colon tumorigenesis. mBio.
-2013;4:e00692--00613.
-
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-7. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+9. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-zeller_potential_2014}{}
-8. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+10. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 \hypertarget{ref-flynn_metabolic_2016}{}
-9. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+11. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
 Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
@@
 \hypertarget{ref-hassan_efficacy_2016}{}
-10. Hassan C, Repici A, Sharma P, Correale L, Zullo A, Bretthauer M, et
+12. Hassan C, Repici A, Sharma P, Correale L, Zullo A, Bretthauer M, et
 al. Efficacy and safety of endoscopic resection of large colorectal
@@
 \hypertarget{ref-yu_metagenomic_2017}{}
-11. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
+13. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
 analysis of faecal microbiome as a tool towards targeted non-invasive
@@
 \hypertarget{ref-zackular_human_2014}{}
-12. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+14. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-warren_co-occurrence_2013}{}
-13. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
+15. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
 et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
@@
 \hypertarget{ref-louis_gut_2014}{}
-14. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
+16. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
 metabolites and colorectal cancer. Nature Reviews Microbiology
@@
 \hypertarget{ref-zhu_intake_2016}{}
-15. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
+17. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
 Proteins Substantially Increased the Relative Abundance of Genus
@@
 \hypertarget{ref-mu_colonic_2016}{}
-16. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
+18. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
 Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
@@
 \hypertarget{ref-ozdal_reciprocal_2016}{}
-17. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
+19. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
 Reciprocal Interactions between Polyphenols and Gut Microbiota and
@@
 \hypertarget{ref-obrien_impact_2013}{}
-18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+20. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-19. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+21. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-20. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+22. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-21. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+23. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-22. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-03-29 submission/manuscript_outline_20161024.Rmd
change SRN to advanced adenoma

@@
 **Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.**  A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
-B) SRN ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
+B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
 C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
-B) Summary of Important Variables in the SRN model.
+B) Summary of Important Variables in the advanced adenoma model.
 MDA of the most important variables in the SRN model.

### 2017-03-29 code/Run_Figure1.R

@@ -25,14 +25,14 @@ differences_graph <- grid.arrange(
   # Difference from bacterial community structure between adenoma and cancer
   ggplot(difference_table_treatment, 
          aes(factor(Dx_Bin, levels = c("adenoma", "adv_adenoma", "cancer"), 
-                    labels = c("Adenoma", "SRN", "Carcinoma")), distance, group = 1)) + 
+                    labels = c("Adenoma", "Advanced\nAdenoma", "Carcinoma")), distance, group = 1)) + 
     geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
     stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
     stat_summary(fun.y = mean, colour = "black", geom = "line") + 
     scale_color_manual(name = "Lesion Type", 
                        values = c('#228B22', '#FFD700', '#DC143C'), 
                        breaks = c("adenoma", "adv_adenoma", "cancer"), 
-                       labels = c("Adenoma", "SRN", "Carcinoma")) + 
+                       labels = c("Adenoma", "Advanced\nAdenoma", "Carcinoma")) + 
     coord_cartesian(ylim = c(0, 1)) + ylab(expression(paste(theta[YC], " Distance"))) + 
     xlab("") + theme_bw() + ggtitle("A") + 
     theme(axis.title = element_text(face="bold", hjust = 0.5), 

### 2017-03-29 code/Run_Figure2.R

@@ -45,7 +45,7 @@ Lesion_plot <- grid.arrange(
     geom_line(color = '#FFD700') + 
     coord_cartesian(ylim = c(0, 0.75)) + 
     geom_hline(aes(yintercept = 0.5), linetype = 2) + 
-    ggtitle("B") + ylab("SRN Postive Probability") + xlab("") + theme_bw() + 
+    ggtitle("B") + ylab("Advanced Adenoma Postive Probability") + xlab("") + theme_bw() + 
     theme(
       axis.title = element_text(face="bold"), 
       plot.title = element_text(face="bold", hjust = 0)), 

### 2017-03-29 code/Run_FigureS1.R

@@ -10,7 +10,7 @@ loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesa
 #Read data needed
 paired_table <- read.csv("data/process/tables/OTU_paired_wilcoxson_test.csv", header = T, stringsAsFactors = F)
 
-label_names <- c("Adenoma", "SRN", "Carcinoma")
+label_names <- c("Adenoma", "Advanced Adenoma", "Carcinoma")
 names(label_names) <- c("adn", "srn", "crc")
 
 pvalue_distribution_paired <- ggplot(paired_table, aes(x=BH_corrected)) + geom_histogram() + 

### 2017-03-29 submission/manuscript_outline_20161024.tex

@@
 test set data and the dark green line represents the model using 100\%
-of the data set and what was used for subsequent classification. B) SRN
-ROC curve: The light yellow shaded areas represent the range of values
-of a 100 different 80/20 splits of the test set data and the dark yellow
-line represents the model using 100\% of the data set and what was used
-for subsequent classification. C) Carcinoma ROC curve: The light red
-shaded areas represent the range of values of a 100 different 80/20
-splits of the test set data and the dark red line represents the model
-using 100\% of the data set and what was used for subsequent
-classification.
+of the data set and what was used for subsequent classification. B)
+Advanced Adenoma ROC curve: The light yellow shaded areas represent the
+range of values of a 100 different 80/20 splits of the test set data and
+the dark yellow line represents the model using 100\% of the data set
+and what was used for subsequent classification. C) Carcinoma ROC curve:
+The light red shaded areas represent the range of values of a 100
+different 80/20 splits of the test set data and the dark red line
+represents the model using 100\% of the data set and what was used for
+subsequent classification.
 
@@
 and the lighter green points are the value of each of the 100 different
-runs. B) Summary of Important Variables in the SRN model. MDA of the
-most important variables in the SRN model. The dark yellow point
-represents the mean and the lighter yellow points are the value of each
-of the 100 different runs. C) MDA of the most important variables in the
-carcinoma model. The dark red point represents the mean and the lighter
-red points are the value of each of the 100 different runs.
+runs. B) Summary of Important Variables in the advanced adenoma model.
+MDA of the most important variables in the SRN model. The dark yellow
+point represents the mean and the lighter yellow points are the value of
+each of the 100 different runs. C) MDA of the most important variables
+in the carcinoma model. The dark red point represents the mean and the
+lighter red points are the value of each of the 100 different runs.
 

### 2017-03-31 submission/manuscript_outline_20161024.Rmd
final edits maybe

@@
 # Adenoma Model 
-adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
+adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv', header = T)
 adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
@@
 *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
-Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (Pvalue `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
@@
 
-Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes to the microbiome in those with adenoma, advanced adenoma, and carincoma.
+Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes towards a normal microbiome in those with adenoma, advanced adenoma, and carincoma.
 First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics.
-Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between initial and follow up.
+Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between pre- and post-treatment samples.
 We also used these models to assess whether changes in the community were toward a more normal microbiome.
-Finally, we assessed both whether surgery for adenomas and SRN provided larger community changes or whether chemotherapy or radiation provided additive changes to the microbiome over surgical resection.
+Finally, we assessed both whether surgery for adenomas and advanced adenomas provided larger community changes and whether chemotherapy or radiation provided additional changes to the microbiome over surgical resection.
 This study aims to provide evidence as to whether the altered CRC microbiome persists or shifts back towards normal after such interventions.
@@
 ***The Bacterial Community:*** 
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- or post-treatment.
-We found that carcinoma patients had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
-Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of the samples pre- and post-treatment from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].
-The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples.
+We found that carcinoma patients had a more dissimilar bacterial community between these samples than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
+Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of pre- and post-treatment samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].
+The bacterial community structure pre- post-treatment was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 Interestingly, when pre- and post-treatment samples were compared, regardless of whether they were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 There was no difference between pre- and post-treatment samples when investigating alpha diversity metrics for adenoma, advanced adenoma, or carcinoma [Table S1].
-Additionally, there was also no difference in the relative abundance of any specific OTU between pre- and post-treatment samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].
+Additionally, there was also no difference in the relative abundance of any specific OTU between these samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].
 
 ***Adenoma Model:***
-Using normal and adenoma individauls from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal.
+Using normal and adenoma individuals from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal.
 The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].
@@
 Using normal and advanced adenoma individuals from a separate cohort we created an advanced adenoma model to help assess whether those with advanced adenoma in this cohort changed towards normal.
-The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].
+The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].
 There was a total of `r length(rownames(srn_imp_vars))` OTUs in the advanced adenoma model [Figure S3B].
@@
 Using normal and carcinoma individuals from a separate cohort we created a carcinoma model to help assess whether those with carcinoma in this cohort changed towards normal.
-The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].
+The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].
 Interestingly, the AUCs improved from adenoma to advanced adenoma to carcinoma [Figure 2].
 There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carcinoma model [Figure S3C].
-Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* are also important for carcinoma classification [Figure S3C].
+Similar to the adenoma and advanced adenoma models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* were also important for carcinoma classification [Figure S3C].
 Also, like the adenoma and advanced adenoma models there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 
-***Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:***
-Next, to identify which OTUs are important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models.
+***Common OTUs:***
+Next, to identify which OTUs were important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models.
 When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.
@@
 The vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].
+The rank of these OTUs, based on median mean decrease in accuracy (MDA), varied higly depending on whether they were in the adenoma, advanced adenoma, or carcinoma model [Figure 3].
+As an example, *Bacteroides* (OTU85) was ranked `r filter(crc_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ in the carcinoma model but in both the adenoma and advanced adenoma model ranked `r filter(adn_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ and `r filter(srn_varsMDA, otu == "Otu000085") %>% select(rank)`^st^ respectively [Figure 3].
 
@@
 This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differs and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
-Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide an additive change.
-Although some of the important OTUs classified to genera from bacteria considered the "usual suspects"" (e.g.
+Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide any additional change.
+Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g.
 *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.
@@
 Unlike previous studies on the microbiome and CRC, ours focuses on comparing the microbiome during recovery from treatment in adenoma, advanced adenoma, and carcinoma groups.
-Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], the majority of important OTUs taxonmically classified to commensal bacteria [Figure S3].
+Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3].
 Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
@@
 
-A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).
+A limitation in our study was that there was a significant difference in the time elapsed in the collection of the post-treatment sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).
 These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time.
@@
 Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
-For all three models the final optimized mtry was `r `ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`
+For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`
 
@@
 Unless otherwise stated the P-values reported are those that were BH corrected.
+Model rank importance was determined by obtining the median MDA from the 100 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.
 
@@
 C) Positive probability change from initial to follow up sample in those with carcinoma..
+
+**Figure 3: Common OTUs Model Rank Based on Median MDA.** For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
 

### 2017-03-31 code/Run_FigureS3.R

@@ -8,7 +8,7 @@ source('code/functions.R')
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
 #Read data needed
-adn_MDA_data_summary <- read.csv("data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv", 
+adn_MDA_data_summary <- read.csv("data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv", 
                                     header = T, stringsAsFactors = F)
 
 adn_MDA_full <- read.csv("data/process/tables/adn_reduced_lesion_model_top_vars_MDA.csv", 
@@ -40,10 +40,10 @@ rm(tax)
 # Generate the files needed for the MDA graph (ADN)
 
 # Order the data file from most to least important based on mean MDA
-adn_MDA_data_summary <- adn_MDA_data_summary[order(adn_MDA_data_summary$mean_MDA, decreasing = TRUE), ]
+adn_MDA_data_summary <- adn_MDA_data_summary[order(adn_MDA_data_summary$median_MDA, decreasing = TRUE), ]
 
 # Pull OTUs that are only in the MDA data
-OTU_IDs <- unique(filter(adn_MDA_data_summary, variable != "fit_result")[, "variable"])
+OTU_IDs <- unique(filter(adn_MDA_data_summary, otu != "fit_result")[, "otu"])
 select_tax_df <- tax_df[OTU_IDs, ]
 low_tax_ID <- gsub("_", " ", gsub("2", "", gsub("_unclassified", "", createTaxaLabeller(select_tax_df))))
 
@@ -63,10 +63,10 @@ adn_labels <- do.call(expression, test)
 # Generate the files needed for the MDA graph (SRN)
 
 # Order the data file from most to least important based on mean MDA
-srn_MDA_data_summary <- srn_MDA_data_summary[order(srn_MDA_data_summary$mean_MDA, decreasing = TRUE), ]
+srn_MDA_data_summary <- srn_MDA_data_summary[order(srn_MDA_data_summary$median_MDA, decreasing = TRUE), ]
 
 # Pull OTUs that are only in the MDA data
-OTU_IDs <- unique(filter(srn_MDA_data_summary, variable != "fit_result")[, "variable"])
[…]

### 2017-03-31 code/Run_combine_aggregate_reduced_adn_model.R

@@ -157,14 +157,15 @@ for(i in 1:length(top_vars_MDA_by_run)){
   rm(tempData)
 }
 
-# "1" pulls the value of mean or sd from the data frame
-MDA_vars_summary <- cbind(
-  mean_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(mean)))[, 1], 
-  sd_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(sd)))[, 1], 
-  variable = rownames(top_vars_MDA_by_run))
+# Generate Median and IQR for the data
+MDA_vars_summary <- as.data.frame(t(top_vars_MDA_by_run)) %>% summarise_each(funs(median, IQR)) %>% 
+  gather(key = tempName) %>% separate(tempName, c("otu", "measurement")) %>% 
+  spread(key = measurement, value = value) %>% rename(median_MDA = median, median_IQR = IQR) %>% 
+  arrange(desc(median_MDA)) %>% 
+  mutate(rank = seq(1:length(rownames(top_vars_MDA_by_run))))
 
-write.csv(MDA_vars_summary[order(MDA_vars_summary[, "mean_MDA"], decreasing = TRUE), ], 
-          "data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv", row.names = F)
+write.csv(MDA_vars_summary, 
+          "data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv", row.names = F)
 
 adn_model_top_vars_MDA_full_data <- 
   mutate(top_vars_MDA_by_run, variables = rownames(top_vars_MDA_by_run)) %>% 

### 2017-03-31 code/Run_combine_aggregate_reduced_srn_model.R

@@ -157,13 +157,14 @@ for(i in 1:length(top_vars_MDA_by_run)){
   rm(tempData)
 }
 
-# "1" pulls the value of mean or sd from the data frame
-MDA_vars_summary <- cbind(
-  mean_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(mean)))[, 1], 
-  sd_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(sd)))[, 1], 
-  variable = rownames(top_vars_MDA_by_run))
+# Generate Median and IQR for the data
+MDA_vars_summary <- as.data.frame(t(top_vars_MDA_by_run)) %>% summarise_each(funs(median, IQR)) %>% 
+  gather(key = tempName) %>% separate(tempName, c("otu", "measurement")) %>% 
+  spread(key = measurement, value = value) %>% rename(median_MDA = median, median_IQR = IQR) %>% 
+  arrange(desc(median_MDA)) %>% 
+  mutate(rank = seq(1:length(rownames(top_vars_MDA_by_run))))
 
-write.csv(MDA_vars_summary[order(MDA_vars_summary[, "mean_MDA"], decreasing = TRUE), ], 
+write.csv(MDA_vars_summary, 
           "data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv", row.names = F)
 
 adn_model_top_vars_MDA_full_data <- 

### 2017-03-31 code/common_otu_graph.R

@@ -1,0 +1,65 @@
+### Plot for FHCRC
+### Common OTUs
+### Marc Sze
+
+source('code/functions.R')
+
+loadLibs(c("tidyr", "dplyr", "scales", "ggplot2"))
+
+# Load needed data
+common_vars_summary <- read.csv('data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', 
+                                header = T, row.names = 1, stringsAsFactors = F)
+
+adn_varsMDA <- read.csv('data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F)
+
+srn_varsMDA <- read.csv('data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F)
+
+crc_varsMDA <- read.csv('data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F)
+
+# Create selection vector
+common_vars <- common_vars_summary$otu
+
+# Create MDA vectors
+adn_rank <- (filter(adn_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)))[, "rank"]
+srn_rank <- (filter(srn_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)))[, "rank"]
+crc_rank <- (filter(crc_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)))[, "rank"]
+
+# Create a combined table for graphing
+combined_table <- select(common_vars_summary, otu) %>% 
+  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID), 
+         adn_rank = adn_rank, srn_rank = srn_rank, crc_rank = crc_rank) %>% 
+  gather(key = model, value = rank, adn_rank, srn_rank, crc_rank)
+
+
+otu_num <- as.numeric(gsub("Otu", "", common_vars))
+
+test <- c()
+for(i in 1:length(common_vars)){
+  
+  test <- c(test, 
+            bquote(paste(italic(.(gsub("_", " ", combined_table$Tax[i]))) ~ "(OTU", .(otu_num[i]), ")", sep = "")))
+}
+
+common_labels <- do.call(expression, test)
+
+# Create Graph
+test_graph <- ggplot(combined_table, 
+                     aes(factor(otu, levels = common_vars), rank)) + 
[…]

### 2017-03-31 results/tables/Table1.Rmd

@@
 output: pdf_document
+
 ---
@@
 
-ft1 <- cbind(Group = c("Normal", "Adenoma", "SRN", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
+ft1 <- cbind(Group = c("Normal", "Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
              ttemptable$BMI, ttemptable$White)
@@
 
-colnames(ft1) <- c("Group", "n", age_label, "Gender (F%)", bmi_label, "Caucasian (%)")
+colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)")
 

### 2017-03-31 results/tables/Table2.Rmd
edits prior to Amanda suggestions

@@
 
-ft1 <- cbind(Group = c("Adenoma", "SRN", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
-             ttemptable$BMI, ttemptable$White)
-
-ft1 <- cbind(Group = c("Normal", "Adenoma", "SRN", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
+ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
              ttemptable$BMI, ttemptable$White)

### 2017-03-31 submission/supplemental_outline_20161024.Rmd

@@
 
-**Table S2: Summary of OTUs in Adenoma, SRN, and Carcinoma Models**
+**Table S2: Summary of OTUs in Adenoma, Advanced Adenoma, and Carcinoma Models**
 
@@
 
-adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
 
@@
 # Create MDA vectors
-adn_MDA <- filter(adn_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
-srn_MDA <- filter(srn_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
-crc_MDA <- filter(crc_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
+adn_MDA <- filter(adn_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)) %>% select(median_MDA)
+srn_MDA <- filter(srn_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)) %>% select(median_MDA)
+crc_MDA <- filter(crc_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)) %>% select(median_MDA)
 
@@
   mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
-  mutate(adn_MDA = adn_MDA$mean_MDA, srn_MDA = srn_MDA$mean_MDA, crc_MDA = crc_MDA$mean_MDA) %>% 
+  mutate(adn_MDA = adn_MDA$median_MDA, srn_MDA = srn_MDA$median_MDA, crc_MDA = crc_MDA$median_MDA) %>% 
   arrange(desc(adn_MDA), desc(srn_MDA), desc(crc_MDA))

### 2017-04-10 submission/manuscript.Rmd
Editing with Pat

--- 2017-03-31 submission/manuscript_outline_20161024.Rmd
@@
   pdf_document:
-    includes:
-      in_header: header.tex
-    keep_tex: yes
+. includes:
+. . in_header: header.tex
+. keep_tex: yes
   word_document: default
@@
 
-$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI
+$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, M.
 
-$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, P.
 
-\end{center}
+\end{center.
 
@@
 
-## Abstract
+## Abstract [ edit at end ]
 
-**Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.
-Despite this, very little is known about what happens to the microbiome after treatment for an adenoma or carcinoma.
-This study tested the hypothesis that treatment for adenoma or carcinoma results in changes towards a normal bacterial community.
-Specifically, we tried to identify components within the community that were different before and after treatment of adenoma, advanced adenoma, and carcinoma.
+**Background.** Colorectal cancer (CRC) is a worldwide health problem.
+Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiome after treatment for an adenoma or carcinoma.
+This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with with disease to those associated with a normal colon.
+We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), and carcinoma (N=`r change_summary["thetayc", "crc_n"]`).
 
-**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus advanced adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was a trend for increasingly less community similarity, between samples pre- and post-treatment from adenoma to advanced adenoma to carcinoma.
-Despite this, no difference was found in the relative abundance of any specific OTU before and after treatment for adenoma, advanced adenoma, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+**Results.** There were large changes the the bacterial communities associated with treatment across the three groups.
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r change_summary["thetayc", "Pvalue_adnVcrc"]`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value>0.05).
+Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value=`r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups.
+The OTUs that were shared across the models for the three diagnosis groups were dominated by lineages that are commonly considered to be commensals (e.g.
+*Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
+
+4. Models suggest that the positive predictive value from diagnostic models decrease following treatment of carcinomas
+
+Finally, we used a larger cohort that contained individuals with
+
+Using diagnostic Random Forest models that were trained to differentiate between individuals with normal colons or those with adenomas, advanced adenomas, or carcinomas we measured the change in the positive probability of having one of the three diagnoses.
+
 Using Random Forest models to assess whether changes in post-treatment samples were towards a normal community, only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma.
-The adenoma model used a total of `r length(rownames(adn_imp_vars))` OTUs, the SRN model used a total of `r length(rownames(srn_imp_vars))` OTUs, and the carcinoma model used a total of `r length(rownames(crc_imp_vars))` OTUs.
-A total of `r length(common_vars_summary$otu)` OTUs were common to all three models with many classifying to commensal bacteria (e.g.
-*Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
+
+5. CRC-associated bacterial populations are unique and they decline in relative abundance with treatment
+
+6. Among the patients with carcinoma, the type of treatment did not significantly alter the amount of change to their bacterial communities.
+
 Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-**Conclusions:** Our data partially supports the hypothesis that the microbiome changes after treatment towards a normal community.
+**Conclusions.** Our data partially supports the hypothesis that the microbiome changes after treatment towards a normal community.
 Individuals with carcinoma had more drastic differences to the overall community than those with adenoma.
@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factors
+bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factor. .
 
@@
 
-Colorectal cancer (CRC) is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].
+Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009].
 The rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
-However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
-Some of these deaths are due to disease recurrence, with current estimates that 20-30% of those undergoing resection having a CRC recurrence [@hellinger_reoperation_2006].
-This is important since approximately 30-50% of those who have a recurrence will die of CRC [@ryuk_predictive_2014].
-To reduce the burden on the healthcare system, finding ways to better stratify those at highest risk of recurrence is of great importance.
+Despite this improvement, there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
+Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014].
+Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.
 
-Recent studies in humans and mouse models have shown that altered membership and structure of the gut microbiome  correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014; @zackular_gut_2013].
+[ 	fix: the following paragraph is mixing two things - mechanistic studies and biomarker studies.
+they need to be split up so that you establish that there is a mechanistic link and that we can use biomarkers to diagnose people with adenomas and carcinomas from those with normal colons.
+]
+
+Recent studies in humans and mouse models have demonstrated a connection between the  that altered membership and structure of the gut microbiome correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014; @zackular_gut_2013].
 Further, bacterial communities on the mucosa are altered between normal and tumor tissue [@dejea_microbiota_2014].
 Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest the possibility of using biomarkers to diagnose disease.
+
 Indeed, models created with microbiome data and clinical tests, such as Fecal Immunoglobulin Test (FIT), result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
-While these studies show how changed microbial communities or invasion by inflammatory species can impact CRC progression  [@flynn_metabolic_2016], they provide very little information as to whether these communities change and rebound towards normal after successful treatment of adenoma, advanced adenoma, or carcinoma.
+Although these studies show how changed microbial communities or invasion by inflammatory species can impact CRC progression [@flynn_metabolic_2016], they provide very little information as to whether treatment to remove lesions affects the presence of bacterial populations associated with those lesions.
 
@@
 
-Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes towards a normal microbiome in those with adenoma, advanced adenoma, and carincoma.
+Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes towards a normal microbiome in those with adenoma, advanced adenoma, and carcinoma.
 First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics.
@@
 ***The Bacterial Community:***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples.
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples.
 We found that carcinoma patients had a more dissimilar bacterial community between these samples than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
@@
 As an example, *Bacteroides* (OTU85) was ranked `r filter(crc_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ in the carcinoma model but in both the adenoma and advanced adenoma model ranked `r filter(adn_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ and `r filter(srn_varsMDA, otu == "Otu000085") %>% select(rank)`^st^ respectively [Figure 3].
+.
 
@@
 Many of the common OTUs that we identified taxonomically classified to potential butyrate producers (e.g.
-*Clostridiales*, *Roseburia*,  and *Anaerostipes*) [Table S2].
+*Clostridiales*, *Roseburia*.
+and *Anaerostipes*) [Table S2].
 Other OTUs classified to bacteria that are inhibited by polyphenols (e.g.
@@
 This information would help to strengthen the case for this model keeping numerous individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having carcinoma.
-It would also provide additional evidence that the microbiome could be used as a risk stratification tool in monitoring recurrence risk and whether different interventions could potentially change this community and lower the probability of future reccurence.
+It would also provide additional evidence that the microbiome could be used as a risk stratification tool in monitoring recurrence risk and whether different interventions could potentially change this community and lower the probability of future recurrence.
 
@@
 
-**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.*.
+A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
 B) NMDS of the pre- and post-treatment samples for the adenoma group.
@@
 D) NMDS of the pre- and post-treatment samples for the carcinoma group.
+
+A: Convert mean to median w/ IQR. Add 0.00 to y-labels
 
@@
 
-**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.**  A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.*.
+A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.**  A) MDA of the most important variables in the adenoma model.
+**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.*.
+A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.

### 2017-04-10 submission/manuscript.Rmd

@@
   pdf_document:
-. includes:
-. . in_header: header.tex
-. keep_tex: yes
+  includes:
+    in_header: header.tex
+  keep_tex: yes
   word_document: default
@@
 
-## Abstract [ edit at end ]
+## Abstract
 
@@
 Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups.
-The OTUs that were shared across the models for the three diagnosis groups were dominated by lineages that are commonly considered to be commensals (e.g.
-*Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).
+However, across the three models, there was little overlap between the OTUs that were indicative of treatment.
+Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent a treatment were more likely to have OTUs associated with normal colons.
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses.
+Although we could clearly differentiate pre- and post-treatment communities from the three diagnosis groups, only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a carcinoma (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Finally, tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with chemotherapy and those treated with chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-4. Models suggest that the positive predictive value from diagnostic models decrease following treatment of carcinomas
-
-Finally, we used a larger cohort that contained individuals with
-
-Using diagnostic Random Forest models that were trained to differentiate between individuals with normal colons or those with adenomas, advanced adenomas, or carcinomas we measured the change in the positive probability of having one of the three diagnoses.
-
-Using Random Forest models to assess whether changes in post-treatment samples were towards a normal community, only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma.
-
-5. CRC-associated bacterial populations are unique and they decline in relative abundance with treatment
-
-6. Among the patients with carcinoma, the type of treatment did not significantly alter the amount of change to their bacterial communities.
-
-Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-
-**Conclusions.** Our data partially supports the hypothesis that the microbiome changes after treatment towards a normal community.
-Individuals with carcinoma had more drastic differences to the overall community than those with adenoma.
-Common OTUs to all models were overwhelmingly from commensal bacteria, suggesting that these bacteria may be important in initial polyp formation, development of advanced adenoma, and transition to carcinoma.
+**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiome will yield improvements in disease management.
 
@@
 Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009].
-The rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].
+Disease mortality has significantly decreased, thanks to improvements in screening [@jemal_cancer_2010].
 Despite this improvement, there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
@@
 
-[ 	fix: the following paragraph is mixing two things - mechanistic studies and biomarker studies.
-they need to be split up so that you establish that there is a mechanistic link and that we can use biomarkers to diagnose people with adenomas and carcinomas from those with normal colons.
-]
+There is growing evidence that the gut microbiota is involved in the progression of CRC.
+Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [sears-ros/etbf into min mice, garrett-into min mice, han?
+tissue culture, @arthur_microbial_2014].
+Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; ; @zackular_gut_2013; baxter human feces into mice].
+Collectively, these studies hypothesize that the microbiota is altering the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
 
-Recent studies in humans and mouse models have demonstrated a connection between the  that altered membership and structure of the gut microbiome correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014; @zackular_gut_2013].
-Further, bacterial communities on the mucosa are altered between normal and tumor tissue [@dejea_microbiota_2014].
-Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest the possibility of using biomarkers to diagnose disease.
+Building on this evidence, several human-based studies have identified unique signatures of colonic lesions.
+One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014, garrett, jobin - science].
+Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; zackular ].
+These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions.
+They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions.
+Gut bacteria have a role in tumorigenesis and there is promise that these populations may be useful biomarkers for aiding in the early detection of disease.
 
-Indeed, models created with microbiome data and clinical tests, such as Fecal Immunoglobulin Test (FIT), result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Although these studies show how changed microbial communities or invasion by inflammatory species can impact CRC progression [@flynn_metabolic_2016], they provide very little information as to whether treatment to remove lesions affects the presence of bacterial populations associated with those lesions.
+Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota.
+If the community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence.
+To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner?
+If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?
 
-Providing an answer to this question is important because it has far reaching implications on both how the bacterial community causes the formation of more polyps [@zackular_manipulation_2016; @arthur_microbial_2014] and the ability to use the microbiome as a predictive screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Understanding polyp formation and transition to advanced adenoma and then carcinoma is crucial to being able to understand how to prevent CRC occurrence.
-Response of the community to treatment is also equally important to predictive models designed for screening purposes since an unresponsive community would provide little additional information for important events, such as recurrence [@hassan_efficacy_2016].
-
-Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes towards a normal microbiome in those with adenoma, advanced adenoma, and carcinoma.
-First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics.
-Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between pre- and post-treatment samples.
-We also used these models to assess whether changes in the community were toward a more normal microbiome.
-Finally, we assessed both whether surgery for adenomas and advanced adenomas provided larger community changes and whether chemotherapy or radiation provided additional changes to the microbiome over surgical resection.
-This study aims to provide evidence as to whether the altered CRC microbiome persists or shifts back towards normal after such interventions.
+We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre- and post-treatment.
+We used both classical community analysis to compare the alpha and beta-diversity of communities pre- and post treatment.
+Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group.
+Finally, we measured the predictive probability of diagnosis models to determine whether treatment would alter the probability of being diagnosed as having a normal colon or the colon of the original diagnosis group.
+Ultimately, we found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas.
+Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence.
 

### 2017-04-11 submission/manuscript.tex

--- 2017-03-29 submission/manuscript_outline_20161024.tex
@@
 not provide any additional changes to the bacterial community in those
-treated for carcinoma (Pvalue \textgreater{} 0.05).
+treated for carcinoma (P-value \textgreater{} 0.05).
 
@@
 Using pre- (initial) and post- (follow up) treatment samples we tested
-the hypothesis that treatment causes detectable changes to the
+the hypothesis that treatment causes detectable changes towards a normal
 microbiome in those with adenoma, advanced adenoma, and carincoma.
@@
 advanced adenoma, or carcinoma versus normal were able to identify
-specific community members that differed between initial and follow up.
-We also used these models to assess whether changes in the community
-were toward a more normal microbiome. Finally, we assessed both whether
-surgery for adenomas and SRN provided larger community changes or
-whether chemotherapy or radiation provided additive changes to the
-microbiome over surgical resection. This study aims to provide evidence
-as to whether the altered CRC microbiome persists or shifts back towards
-normal after such interventions.
+specific community members that differed between pre- and post-treatment
+samples. We also used these models to assess whether changes in the
+community were toward a more normal microbiome. Finally, we assessed
+both whether surgery for adenomas and advanced adenomas provided larger
+community changes and whether chemotherapy or radiation provided
+additional changes to the microbiome over surgical resection. This study
+aims to provide evidence as to whether the altered CRC microbiome
+persists or shifts back towards normal after such interventions.
 
@@
 advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
-differences between pre- or post-treatment. We found that carcinoma
-patients had a more dissimilar bacterial community between their initial
-and follow up sample than those with adenoma (P-value \textless{} 0.001)
+differences between pre- and post-treatment samples. We found that
+carcinoma patients had a more dissimilar bacterial community between
+these samples than those with adenoma (P-value \textless{} 0.001)
 {[}Figure 1A{]}. Although no significant differences were observed
 between advanced adenoma and carcinoma there was an increase in the
-dissimilarity of the samples pre- and post-treatment from adenoma (0.55
-± 0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78
-± 0.15) {[}Figure 1A{]}. The bacterial community structure before and
-after surgery was visualized using NMDS for adenoma {[}Figure 1B{]}
+dissimilarity of pre- and post-treatment samples from adenoma (0.55 ±
+0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78 ±
+0.15) {[}Figure 1A{]}. The bacterial community structure pre-
+post-treatment was visualized using NMDS for adenoma {[}Figure 1B{]}
 (PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
@@
 S1{]}. Additionally, there was also no difference in the relative
-abundance of any specific OTU between pre- and post-treatment samples
-for adenoma, advanced adenoma, or carcinoma only {[}Figure S1{]}.
+abundance of any specific OTU between these samples for adenoma,
+advanced adenoma, or carcinoma only {[}Figure S1{]}.
 
-\textbf{\emph{Adenoma Model:}} Using normal and adenoma individauls from
+\textbf{\emph{Adenoma Model:}} Using normal and adenoma individuals from
 a separate cohort we created an adenoma model to help assess whether
@@
 cohort changed towards normal. The range of model AUC's from 100 runs of
-20 repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
+20 repeated 10-fold cross-validation was 0.68 - 0.77 with the AUC of the
 model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
@@
 whether those with carcinoma in this cohort changed towards normal. The
-range of model AUC's from 100 runs of 20 repeated 10 fold
+range of model AUC's from 100 runs of 20 repeated 10-fold
 cross-validation was 0.84 - 0.9 with the AUC of the model used for
@@
 There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
-Similar to the adenoma and SRN models the vast majority of OTUs
-classified to bacteria typically thought of as commensal but OTUs that
-also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
-\emph{Parvimonas} are also important for carcinoma classification
+Similar to the adenoma and advanced adenoma models the vast majority of
+OTUs classified to bacteria typically thought of as commensal but OTUs
+that also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
+\emph{Parvimonas} were also important for carcinoma classification
 {[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
@@
 
-\textbf{\emph{Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:}}
-Next, to identify which OTUs are important at all stages of disease, we
-identified common predictive OTUs within the adenoma, advanced adenoma,
-and carcinoma models. When we compared the three different models with
-each other there were a total of 26 common OTUs. Some of the most common
-taxonomic identifications belonged to \emph{Bacteroides},
-\emph{Blautia}, \emph{Anaerostipes}, \emph{Lachnospiraceae}, and
-\emph{Dorea}. The vast majority of the OTUs that were common between
-these models had classifications to bacteria typically thought of as
-commensal {[}Table S2{]}.
+\textbf{\emph{Common OTUs:}} Next, to identify which OTUs were important
+at all stages of disease, we identified common predictive OTUs within
+the adenoma, advanced adenoma, and carcinoma models. When we compared
+the three different models with each other there were a total of 26
+common OTUs. Some of the most common taxonomic identifications belonged
+to \emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
+\emph{Lachnospiraceae}, and \emph{Dorea}. The vast majority of the OTUs
+that were common between these models had classifications to bacteria
+typically thought of as commensal {[}Table S2{]}. The rank of these
+OTUs, based on median mean decrease in accuracy (MDA), varied higly
+depending on whether they were in the adenoma, advanced adenoma, or
+carcinoma model {[}Figure 3{]}. As an example, \emph{Bacteroides}
+(OTU85) was ranked 14\textsuperscript{th} in the carcinoma model but in
+both the adenoma and advanced adenoma model ranked
+35\textsuperscript{th} and 41\textsuperscript{st} respectively {[}Figure
+3{]}.
 
@@
 normal after treatment for carcinoma and that chemotherapy and radiation
-did not provide an additive change. Although some of the important OTUs
-classified to genera from bacteria considered the ``usual suspects''"
-(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
-many did not. The majority of important OTUs had taxonomic
-classifications for resident gut microbes and were common for the
-adenoma, advanced adenoma, and carcinoma models. This suggests that
+did not provide any additional change. Although some of the important
+OTUs classified to genera from bacteria considered the ``usual
+suspects'' (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
+\emph{Parvimonas}) many did not. The majority of important OTUs had
+taxonomic classifications for resident gut microbes and were common for
+the adenoma, advanced adenoma, and carcinoma models. This suggests that
 members within the commensal community may be the first that change
@@
 for genera associated with specific bacterium linked with CRC {[}Figure
-S3{]}, the majority of important OTUs taxonmically classified to
-commensal bacteria {[}Figure S3{]}. Although these changes may be
-subtle, due to the lack of significant differences in the bacterial
-community pre- and post-treatment in adenoma and advanced adenoma
-{[}Figure 1{]}; they support the hypothesis that the first members of
-the community to change and potentially stay changed even after
-treatment are those that are commensal bacteria.
+S3{]}, most of the important OTUs taxonomically classified to commensal
+bacteria {[}Figure S3{]}. Although these changes may be subtle, due to
+the lack of significant differences in the bacterial community pre- and
+post-treatment in adenoma and advanced adenoma {[}Figure 1{]}; they
+support the hypothesis that the first members of the community to change
+and potentially stay changed even after treatment are those that are
+commensal bacteria.
 
@@
 
-A limitation, in our study, was that there was a significant difference
-in the time elapsed in the collection of the follow up sample between
+A limitation in our study was that there was a significant difference in
+the time elapsed in the collection of the post-treatment sample between
 adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
@@
 the 67-person cohort was completed. For all three models the final
-optimized mtry was
-\texttt{r}ifelse(as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
-\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} ==
-as.data.frame(count(srn\_AUC\_run\_summary, best\_mtry)
-\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} \&
-as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
-\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} ==
-as.data.frame(count(crc\_AUC\_run\_summary, best\_mtry)
-\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]},
-as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
-\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]}, ``not the
-same'')`
+optimized mtry was 2
 
@@
 Unless otherwise stated the P-values reported are those that were BH
-corrected.
+corrected. Model rank importance was determined by obtining the median
+MDA from the 100 20 repeated 10-fold cross validation and then ranking
+from largest to smallest MDA.
 
@@
 carcinoma..
+
+\textbf{Figure 3: Common OTUs Model Rank Based on Median MDA.} For each
+common OTU the lowest taxonomic identification and importance rank for
+each model run is shown.
 

### 2017-04-11 submission/manuscript.Rmd
update references

@@
 There is growing evidence that the gut microbiota is involved in the progression of CRC.
-Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [sears-ros/etbf into min mice, garrett-into min mice, han?
-tissue culture, @arthur_microbial_2014].
-Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; ; @zackular_gut_2013; baxter human feces into mice].
+Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014].
+Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; ; @zackular_gut_2013; @baxter_structure_2014].
 Collectively, these studies hypothesize that the microbiota is altering the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
@@
 Building on this evidence, several human-based studies have identified unique signatures of colonic lesions.
-One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014, garrett, jobin - science].
-Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; zackular ].
+One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012].
+Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].
 These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions.

### 2017-04-11 submission/manuscript.Rmd

@@
 
-***The Bacterial Community:***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples.
-We found that carcinoma patients had a more dissimilar bacterial community between these samples than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].
-Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of pre- and post-treatment samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].
-The bacterial community structure pre- post-treatment was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-Interestingly, when pre- and post-treatment samples were compared, regardless of whether they were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was no difference between pre- and post-treatment samples when investigating alpha diversity metrics for adenoma, advanced adenoma, or carcinoma [Table S1].
-Additionally, there was also no difference in the relative abundance of any specific OTU between these samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].
+***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1).
+The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A].
+We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001).
+There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P>0.05).
+Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1BCD].
+We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`).
+Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value>0.05) [Table S1].
 
-***Adenoma Model:***
-Using normal and adenoma individuals from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal.
-The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].
-There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A].
-There was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was also no significant decrease in the positive probability of adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].
+***The effects of treatment are not consistent across treatment groups.***
+We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group.
+First, we sought to identify individual OTUs that could account for the change in overall community structure.
+However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1].
+It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test.
+To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
+Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs.
+Despite this restriction, the models performed well (adenoma AUC=, advanced adenoma AUC=, carcinoma AUC= ).
+Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
+These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.
 
-***Advanced Adenoma Model:***
-Using normal and advanced adenoma individuals from a separate cohort we created an advanced adenoma model to help assess whether those with advanced adenoma in this cohort changed towards normal.
-The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].
-There was a total of `r length(rownames(srn_imp_vars))` OTUs in the advanced adenoma model [Figure S3B].
-Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal.
-Also similar to the adenoma model there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05) and no significant decrease in the positive probability of advanced adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].
+***Post-treatment samples from patients with carcinoma share***
+Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons.
+To test this, we used an expanded cohort of 423 individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
+We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon.
+The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2).
+The OTUs that were most important in the models for differentiating between patients with normal colons and those with adenomas or advanced adenomas were dominated by populations commonly considered as commensals (e.g.
+*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, etc.
+[Figure?]).
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure ?].
+Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
+A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon.
+There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 2].
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased.
+In fact, only 3 of the 26 patients (12%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment.
+One of those was re-diagnosed with carcinoma on the follow up visit.
+These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon
 
-***Carcinoma Model:***
-Using normal and carcinoma individuals from a separate cohort we created a carcinoma model to help assess whether those with carcinoma in this cohort changed towards normal.
-The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].
-Interestingly, the AUCs improved from adenoma to advanced adenoma to carcinoma [Figure 2].
-There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carcinoma model [Figure S3C].
-Similar to the adenoma and advanced adenoma models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* were also important for carcinoma classification [Figure S3C].
-Also, like the adenoma and advanced adenoma models there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was a significant decrease in the positive probability of carcinoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or advanced adenoma.
-The one individual still positive for carcinoma after treatment had an increase in carcinoma positive proability between their pre- and post-treatment sample [Figure 2C].
-
-***Common OTUs:***
-Next, to identify which OTUs were important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models.
-When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.
-Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*.
-The vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].
-The rank of these OTUs, based on median mean decrease in accuracy (MDA), varied higly depending on whether they were in the adenoma, advanced adenoma, or carcinoma model [Figure 3].
-As an example, *Bacteroides* (OTU85) was ranked `r filter(crc_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ in the carcinoma model but in both the adenoma and advanced adenoma model ranked `r filter(adn_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ and `r filter(srn_varsMDA, otu == "Otu000085") %>% select(rank)`^st^ respectively [Figure 3].
-.
-
-***Treatment Affects on Community:***
-After observing these changes from treatment we assessed the possible confounding impact of chemotherapy or radiation, in the carcinoma group, and surgical resection, in the adenoma group, on the observed results.
-In the carcinoma group neither chemotherapy nor radiation provided any additional change from initial sample over surgery alone (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S3].
-For the adenoma group there was a single difference in observed OTUs (sobs) between those that received surgerical resection and those that had regular polyp removal (P-value `r ifelse(min(comb_adn_surg_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S4].
-For the surgical resection comparison, adenoma and advanced adenoma were combined due to the low number of resection occurances in these two groups.
-Using a fisher exact test there was no difference in the proportion of those receiving surgerical resection between the adenoma and advanced adenoma groups (P-value `r ifelse(comb_adn_surg_summary["prop_surg_adn_srn", "pvalue"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-This data suggests that the microbiome changes observed in the carcinoma group were mostly a result from surgerical resection and not from chemotherapy or radiation.
+***Difficult to identify effects of specific treatments on the change in the microbiota.***
+The type of treatment that the patients received varied across diagnosis groups.
+Those with adenomas and advanced adenomas received surgical resection (adenoma, N=4; advanced adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=4; advanced adenoma, N=4) and those with carcinomas received surgical resection (N=~14), surgical resection with chemotherapy (N=~6), and surgical resection with chemotherapy and radiation (N=~6).
+We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power.
+We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P<0.XXX).
+Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P=0.XX).
+Finally, the change in the positive probability was not significantly different between the two treatment groups (P=0.XX).
+Due to the relatively small number of samples in each treatment group, it is difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota
 

### 2017-04-12 code/Run_change_theta_Fit.R
general updates

@@ -5,7 +5,7 @@
 # Load needed functions and libraries
 source('code/functions.R')
 
-loadLibs("dplyr")
+loadLibs(c("dplyr", "aod"))
 
 # Load needed data
 thetaCompTotal <- read.dist('data/process/final.thetayc.0.03.lt.ave.dist')
@@ -75,6 +75,12 @@ change_theta_fit_summary['fit', ] <- c(
   Pvalue_srnVcrc = wilcox.test(
     filter(difference_table_treatment, Dx_Bin == "adv_adenoma")[, "fit_difference"], 
     filter(difference_table_treatment, Dx_Bin == "cancer")[, "fit_difference"])$p.value)
+
+
+# Perform a trend test 
+
+
+
 
 # Save table for later use
 write.csv(change_theta_fit_summary, 

### 2017-04-12 code/old/common_all_models.R

@@ -1,0 +1,27 @@
+# Rough comparison of common OTUs
+### Most concerned with commonalities between models (treatment and normal)
+### Marc Sze
+
+
+source('code/functions.R')
+
+loadLibs(c("tidyr", "dplyr", "scales", "knitr", "rmarkdown"))
+
+adn_treat <- read.csv("data/process/tables/adn_treatment_imp_vars_summary.csv")
+srn_treat <- read.csv("data/process/tables/srn_treatment_imp_vars_summary.csv")
+crc_treat <- read.csv("data/process/tables/crc_treatment_imp_vars_summary.csv")
+common_norm <- read.csv('data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', 
+                                                       header = T, row.names = 1, stringsAsFactors = F)
+
+
+
+adn_otus <- as.character(adn_treat$Variable[1:round(length(adn_treat$Variable)*0.10)])
+srn_otus <- as.character(srn_treat$Variable[1:round(length(srn_treat$Variable)*0.10)])
+crc_otus <- as.character(crc_treat$Variable[1:round(length(crc_treat$Variable)*0.10)])
+
+
+length(rownames(filter(common_norm, otu %in% adn_otus) %>% select(otu)))
+length(rownames(filter(common_norm, otu %in% srn_otus) %>% select(otu)))
+length(rownames(filter(common_norm, otu %in% crc_otus) %>% select(otu)))
+
+

### 2017-04-12 results/tables/Table2.Rmd
flip demo tables

@@
 
-**Table 2: Demographic Data of Pre and Post Treatment Cohort**
+**Table 2: Demographic Data of Training Cohort**
 
-```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
 
-test_cohort <- read.csv("../../data/process/mod_metadata/good_metaf_final.csv", header = T, stringsAsFactors = F) %>% 
-  select(Dx_Bin, dx, Age, Gender, BMI, White)
+metadata <- read.csv("../../data/process/mod_metadata/metaI_final.csv", header = T, stringsAsFactors = F)
+test_cohort <- read.csv("../../data/process/mod_metadata/good_metaf_final.csv", header = T, stringsAsFactors = F)
 
-t2 <- as.data.frame(matrix(nrow = 5, ncol = 3, 
+metadata <- filter(metadata, !sample %in% c(test_cohort$initial, test_cohort$followUp)) %>% 
+  select(Dx_Bin, dx, Age, Gender, BMI, White) %>% mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin), Dx_Bin = gsub("High Risk Normal", "Normal", Dx_Bin))
+
+t1 <- as.data.frame(matrix(nrow = 5, ncol = 4, 
                            dimnames = 
                              list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
-                                  coln = c("Adenoma", "SRN", "Carcinoma"))))
+                                  coln = c("Normal", "Adenoma", "SRN", "Carcinoma"))))
 
-data_filter1 <- c("adenoma", "adenoma", "cancer")
-data_filter2 <- c("adenoma", "adv_adenoma", "cancer")
+data_filter1 <- c("normal", "adenoma", "adenoma", "cancer")
+data_filter2 <- c("Normal", "Adenoma", "adv_adenoma", "Cancer")
 
 # Generate the summary stats for each group
-for(i in 1:length(colnames(t2))){
+for(i in 1:length(colnames(t1))){
 
-  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
-  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
-  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
-  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
-  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
+  t1["Dx_Bin", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
+  t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
+  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
+  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
+  t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
 }
@@
 
-for(i in 1:length(colnames(t2))){
+for(i in 1:length(colnames(t1))){
 
-  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
-  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
+  age_sd <- c(age_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
+  BMI_sd <- c(BMI_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
 }
@@
 
-ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
+ttemptable <- as.data.frame(t(t1)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                     label = "\u00B1", 
@@
 
-ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
+ft1 <- cbind(Group = c("Normal", "Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
              ttemptable$BMI, ttemptable$White)
@@
 kable(test, format = "markdown", align = 'c', padding = 2)
+
 ```

### 2017-04-12 submission/supplemental.Rmd
try to fit table on one page (not possible)

--- 2017-03-31 submission/supplemental_outline_20161024.Rmd
@@
 \newpage
+
 \newcommand{\blandscape}{\begin{landscape}}
@@
 
-**Table S2: Summary of OTUs in Adenoma, Advanced Adenoma, and Carcinoma Models**
+**Table S2: Summary of Common OTUs for All Diagnosis Groups**
 
@@
 common_vars <- common_vars_summary$otu
+common_num <- as.numeric(gsub("Otu", "", common_vars_summary$otu))
 
@@
   mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
-  mutate(adn_MDA = adn_MDA$median_MDA, srn_MDA = srn_MDA$median_MDA, crc_MDA = crc_MDA$median_MDA) %>% 
+  mutate(otu = paste("OTU", common_num, sep = "")) %>% 
+  mutate(adn_MDA = round(adn_MDA$median_MDA, digits = 2), srn_MDA = round(srn_MDA$median_MDA, digits = 2), crc_MDA = round(crc_MDA$median_MDA, digits = 2)) %>% 
   arrange(desc(adn_MDA), desc(srn_MDA), desc(crc_MDA))
 
-colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Adenoma Model MDA", "SRN Model MDA", "Carcinoma Model MDA")
+colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Adenoma Median MDA", "Advanced Adenoma Median MDA", "Carcinoma Median MDA")
 
-kable(st2, format = "markdown", align = 'c', padding = 2)
+kable(st2, format = "markdown", align = 'c', padding = 0)
 

### 2017-04-12 code/Run_Figure3.R
Figure changes and chemo comparison streamlining

@@ -73,7 +73,7 @@ Lesion_plot <- grid.arrange(
              label = paste("Carcinoma"),x = 1.5, y = 0.77, size = 4), nrow = 1)
   
 # Save figures and write necessary tables
-ggsave(file = "results/figures/Figure2.pdf", Lesion_plot, 
+ggsave(file = "results/figures/Figure3.pdf", Lesion_plot, 
        width=13, height = 10, dpi = 300)
 
 

### 2017-04-12 code/old/Adns_combine_reduced_agg_treat_model.R

@@ -6,7 +6,7 @@
 ###Load needed Libraries and functions
 source('code/functions.R')
 
-loadLibs(c("dplyr", "scales", "caret", "pROC", "reshape2"))
+loadLibs(c("dplyr", "scales", "caret", "pROC", "tidyr"))
 
 
 # Load needed data
@@ -125,24 +125,23 @@ for(i in 1:length(top_vars_MDA_by_run)){
   rm(tempData)
 }
 
-# Get Taxa
-tax_df[vars_to_keep, ]
-
 
 # "1" pulls the value of mean or sd from the data frame
-MDA_vars_summary <- as.data.frame(cbind(
-  mean_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(mean)))[, 1], 
-  sd_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(sd)))[, 1])) %>% 
-  mutate(variable = rownames(top_vars_MDA_by_run)) %>% 
-  mutate(tax_id = as.character(tax_df[variable, "Genus"]))
+MDA_vars_summary <- as.data.frame(t(top_vars_MDA_by_run)) %>% summarise_each(funs(median, IQR)) %>% 
+  gather(key = tempName) %>% separate(tempName, c("otu", "measurement")) %>% 
+  spread(key = measurement, value = value) %>% rename(median_MDA = median, median_IQR = IQR) %>% 
+  mutate(tax_id = as.character(tax_df[otu, "Genus"])) %>% 
+  arrange(desc(median_MDA)) %>% 
+  mutate(rank = seq(1:length(rownames(top_vars_MDA_by_run))))
 
 
-write.csv(MDA_vars_summary[order(MDA_vars_summary[, "mean_MDA"], decreasing = TRUE), ], 
+write.csv(MDA_vars_summary, 
           "data/process/tables/reduced_adn_treatment_top_vars_MDA_Summary.csv", row.names = F)
 
-adn_model_top_vars_MDA_full_data <- 
-  mutate(top_vars_MDA_by_run, variables = rownames(top_vars_MDA_by_run)) %>% 
-  melt(id = c("variables"))
+adn_model_top_vars_MDA_full_data <- mutate(top_vars_MDA_by_run, otu = rownames(top_vars_MDA_by_run)) %>% 
+  slice(match(MDA_vars_summary$otu, otu)) %>% 
+  gather(key = variable, value = value, -otu)
+  
 
[…]

### 2017-04-12 code/old/CRC_combine_reduced_agg_treat_model.R

@@ -6,7 +6,7 @@
 ###Load needed Libraries and functions
 source('code/functions.R')
 
-loadLibs(c("dplyr", "scales", "caret", "pROC", "reshape2"))
+loadLibs(c("dplyr", "scales", "caret", "pROC", "tidyr"))
 
 
 # Load needed data
@@ -125,18 +125,19 @@ for(i in 1:length(top_vars_MDA_by_run)){
 }
 
 # "1" pulls the value of mean or sd from the data frame
-MDA_vars_summary <- as.data.frame(cbind(
-  mean_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(mean)))[, 1], 
-  sd_MDA = t(summarise_each(as.data.frame(t(top_vars_MDA_by_run)), funs(sd)))[, 1])) %>% 
-  mutate(variable = rownames(top_vars_MDA_by_run)) %>% 
-  mutate(tax_id = as.character(tax_df[variable, "Genus"]))
+MDA_vars_summary <- as.data.frame(t(top_vars_MDA_by_run)) %>% summarise_each(funs(median, IQR)) %>% 
+  gather(key = tempName) %>% separate(tempName, c("otu", "measurement")) %>% 
+  spread(key = measurement, value = value) %>% rename(median_MDA = median, median_IQR = IQR) %>% 
+  mutate(tax_id = as.character(tax_df[otu, "Genus"])) %>% 
+  arrange(desc(median_MDA)) %>% 
+  mutate(rank = seq(1:length(rownames(top_vars_MDA_by_run))))
 
-write.csv(MDA_vars_summary[order(MDA_vars_summary[, "mean_MDA"], decreasing = TRUE), ], 
+write.csv(MDA_vars_summary, 
           "data/process/tables/reduced_crc_treatment_top_vars_MDA_Summary.csv", row.names = F)
 
-crc_model_top_vars_MDA_full_data <- 
-  mutate(top_vars_MDA_by_run, variables = rownames(top_vars_MDA_by_run)) %>% 
-  melt(id = c("variables"))
+crc_model_top_vars_MDA_full_data <- mutate(top_vars_MDA_by_run, otu = rownames(top_vars_MDA_by_run)) %>% 
+  slice(match(MDA_vars_summary$otu, otu)) %>% 
+  gather(key = variable, value = value, -otu)
 
 write.csv(crc_model_top_vars_MDA_full_data, 
[…]

### 2017-04-12 code/treatment_otu_graph.R

@@ -1,0 +1,119 @@
+### Plot for FHCRC
+### Common OTUs
+### Marc Sze
+
+source('code/functions.R')
+
+loadLibs(c("tidyr", "dplyr", "scales", "ggplot2", "gridExtra"))
+
+# Load needed data
+adn_treat_varsMDA <- read.csv('data/process/tables/reduced_adn_treatment_top_vars_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F) %>% 
+  mutate(rank = seq(1:10), tax_id = gsub("_unclassified", "", tax_id))
+
+srn_treat_varsMDA <- read.csv('data/process/tables/reduced_srn_treatment_top_vars_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F) %>% 
+  mutate(rank = seq(1:10), tax_id = gsub("_unclassified", "", tax_id))
+
+crc_treat_varsMDA <- read.csv('data/process/tables/reduced_crc_treatment_top_vars_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F) %>% 
+  mutate(rank = seq(1:10), tax_id = gsub("_unclassified", "", tax_id))
+
+
+adn_data <- read.csv('data/process/tables/reduced_adn_treatment_top_vars_MDA_full_data.csv', 
+                     header = T, stringsAsFactors = F)
+
+srn_data <- read.csv('data/process/tables/reduced_srn_treatment_top_vars_MDA_full_data.csv', 
+                     header = T, stringsAsFactors = F)
+crc_data <- read.csv('data/process/tables/reduced_crc_treatment_top_vars_MDA_full_data.csv', 
+                     header = T, stringsAsFactors = F)
+
+# Create adn labels
+otu_num <- as.numeric(gsub("Otu", "", adn_treat_varsMDA$otu))
+
+test <- c()
+for(i in 1:length(otu_num)){
+  
+  test <- c(test, 
+            bquote(paste(italic(.(gsub("_", " ", adn_treat_varsMDA$tax_id[i]))) ~ "(OTU", .(otu_num[i]), ")", sep = "")))
+}
+
+adn_labels <- do.call(expression, test)
+adn_axis <- paste(adn_treat_varsMDA$tax_id, " (OTU", otu_num, ")", sep="")
+
+
+# Create srn labels
+otu_num <- as.numeric(gsub("Otu", "", srn_treat_varsMDA$otu))
+
+test <- c()
+for(i in 1:length(otu_num)){
+  
+  test <- c(test, 
[…]

### 2017-04-13 code/old/common_all_models.R
add venn diagram

@@ -5,23 +5,98 @@
 
 source('code/functions.R')
 
-loadLibs(c("tidyr", "dplyr", "scales", "knitr", "rmarkdown"))
-
-adn_treat <- read.csv("data/process/tables/adn_treatment_imp_vars_summary.csv")
-srn_treat <- read.csv("data/process/tables/srn_treatment_imp_vars_summary.csv")
-crc_treat <- read.csv("data/process/tables/crc_treatment_imp_vars_summary.csv")
-common_norm <- read.csv('data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', 
-                                                       header = T, row.names = 1, stringsAsFactors = F)
+loadLibs(c("tidyr", "dplyr", "scales", "VennDiagram", "ggplot2"))
 
 
+# Load necessary data
+adn_data <- read.csv("data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv", 
+                     header = T, stringsAsFactors = F)
+srn_data <- read.csv("data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv", 
+                      header = T, stringsAsFactors = F)
+crc_data <- read.csv("data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv", 
+                      header = T, stringsAsFactors = F)
 
-adn_otus <- as.character(adn_treat$Variable[1:round(length(adn_treat$Variable)*0.10)])
-srn_otus <- as.character(srn_treat$Variable[1:round(length(srn_treat$Variable)*0.10)])
-crc_otus <- as.character(crc_treat$Variable[1:round(length(crc_treat$Variable)*0.10)])
+#Generate vectors of OTUs for each diagnosis
+adn_OTUs <- adn_data$otu
+srn_OTUs <- srn_data$otu
+crc_data <- crc_data$otu
 
 
-length(rownames(filter(common_norm, otu %in% adn_otus) %>% select(otu)))
-length(rownames(filter(common_norm, otu %in% srn_otus) %>% select(otu)))
-length(rownames(filter(common_norm, otu %in% crc_otus) %>% select(otu)))
+test <- calculate.overlap(list(adn_OTUs, srn_OTUs, crc_data))
+
+test2 <- as.data.frame.list(lapply(test, function(x) length(x)))
+
+
+venn_graph <- as.data.frame(cbind(g1 = c(4.5, 5.4, 6.2), g2 = c(7, 5.5, 7))) %>% 
+  mutate(group = c("adn", "srn", "crc")) %>% 
[…]

### 2017-04-13 results/tables/Table1.Rmd

@@
 output: pdf_document
-
 ---
@@
 
-**Table 1: Demographic Data of Training Cohort**
+**Table 1: Demographic Data of Pre and Post Treatment Cohort**
 
-```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
+```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
 
-metadata <- read.csv("../../data/process/mod_metadata/metaI_final.csv", header = T, stringsAsFactors = F)
-test_cohort <- read.csv("../../data/process/mod_metadata/good_metaf_final.csv", header = T, stringsAsFactors = F)
+test_cohort <- read.csv("../../data/process/mod_metadata/good_metaf_final.csv", header = T, stringsAsFactors = F) %>% 
+  select(Dx_Bin, dx, Age, Gender, BMI, White)
 
-metadata <- filter(metadata, !sample %in% c(test_cohort$initial, test_cohort$followUp)) %>% 
-  select(Dx_Bin, dx, Age, Gender, BMI, White) %>% mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin), Dx_Bin = gsub("High Risk Normal", "Normal", Dx_Bin))
-
-t1 <- as.data.frame(matrix(nrow = 5, ncol = 4, 
+t2 <- as.data.frame(matrix(nrow = 5, ncol = 3, 
                            dimnames = 
                              list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
-                                  coln = c("Normal", "Adenoma", "SRN", "Carcinoma"))))
+                                  coln = c("Adenoma", "SRN", "Carcinoma"))))
 
-data_filter1 <- c("normal", "adenoma", "adenoma", "cancer")
-data_filter2 <- c("Normal", "Adenoma", "adv_adenoma", "Cancer")
+data_filter1 <- c("adenoma", "adenoma", "cancer")
+data_filter2 <- c("adenoma", "adv_adenoma", "cancer")
 
 # Generate the summary stats for each group
-for(i in 1:length(colnames(t1))){
+for(i in 1:length(colnames(t2))){
 
-  t1["Dx_Bin", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
-  t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
-  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
-  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
-  t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
+  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
+  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
+  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
+  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
+  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
 }
@@
 
-for(i in 1:length(colnames(t1))){
+for(i in 1:length(colnames(t2))){
 
-  age_sd <- c(age_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
-  BMI_sd <- c(BMI_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
+  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
+  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
 }
@@
 
-ttemptable <- as.data.frame(t(t1)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
+ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                     label = "\u00B1", 
@@
 
-ft1 <- cbind(Group = c("Normal", "Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
+ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
              ttemptable$BMI, ttemptable$White)
@@
 kable(test, format = "markdown", align = 'c', padding = 2)
-
 ```

### 2017-04-14 submission/manuscript.Rmd

@@
 ---
-title: ''
+title: ""
+bibliography: reference.bib
+output:
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: header.tex
 csl: microbiome.csl
 fontsize: 12pt
-output:
-  pdf_document:
-  includes:
-    in_header: header.tex
-  keep_tex: yes
-  word_document: default
 geometry: margin=1.0in
-bibliography: reference.bib
 ---
@@
 
-# initial microbiome and FIT description data
+# initial microbiome changes
 change_summary <- read.csv('../data/process/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
@@
 all_OTU_pair_wilcox <- read.csv("../data/process/tables/OTU_paired_wilcoxson_test.csv", header = T)
+
+# Treatment model data
+adn_treat_AUC_summary <- read.csv('../data/process/tables/reduced_adn_treatment_ROC_model_summary.csv', 
+                                  stringsAsFactors = F, header = T)
+srn_treat_AUC_summary <- read.csv('../data/process/tables/reduced_srn_treatment_ROC_model_summary.csv', 
+                                  stringsAsFactors = F, header = T)
+crc_treat_AUC_summary <- read.csv('../data/process/tables/reduced_crc_treatment_ROC_model_summary.csv', 
+                                  stringsAsFactors = F, header = T)
+
+adn_treat_imp_vars <- read.csv('../data/process/tables/reduced_adn_treatment_top_vars_MDA_Summary.csv', 
+                                  stringsAsFactors = F, header = T)
+srn_treat_imp_vars <- read.csv('../data/process/tables/reduced_srn_treatment_top_vars_MDA_Summary.csv', 
+                                  stringsAsFactors = F, header = T)
+crc_treat_imp_vars <- read.csv('../data/process/tables/reduced_crc_treatment_top_vars_MDA_Summary.csv', 
+                                  stringsAsFactors = F, header = T)
 
@@
 crc_general_summary <- read.csv('../data/process/tables/crc_reduced_auc_summary.csv', header = T, row.names = 1)
+crc_probs <- read.csv('../data/process/tables/crc_reduced_follow_up_probability_summary.csv', header = T)
 
@@
 
-\end{center.
+\end{center}
 
@@
 **Results.** There were large changes the the bacterial communities associated with treatment across the three groups.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r change_summary["thetayc", "Pvalue_adnVcrc"]`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value>0.05).
-Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value=`r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value>0.05).
+Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups.
@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factor. .
+bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factor.
 
 \newpage
+\linenumbers
 
@@
 There is growing evidence that the gut microbiota is involved in the progression of CRC.
-Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014].
-Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; ; @zackular_gut_2013; @baxter_structure_2014].
+Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009].
+Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
 Collectively, these studies hypothesize that the microbiota is altering the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
 
-Building on this evidence, several human-based studies have identified unique signatures of colonic lesions.
+Building on this evidence, several human-based studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015].
 One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012].
@@
 Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs.
-Despite this restriction, the models performed well (adenoma AUC=, advanced adenoma AUC=, carcinoma AUC= ).
+Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
@@
 The OTUs that were most important in the models for differentiating between patients with normal colons and those with adenomas or advanced adenomas were dominated by populations commonly considered as commensals (e.g.
-*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, etc.
-[Figure?]).
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure ?].
+*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figure S3].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
 A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon.
-There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 2].
+There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 3].
 The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased.
-In fact, only 3 of the 26 patients (12%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment.
+In fact, only `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment.
 One of those was re-diagnosed with carcinoma on the follow up visit.
@@
 The type of treatment that the patients received varied across diagnosis groups.
-Those with adenomas and advanced adenomas received surgical resection (adenoma, N=4; advanced adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=4; advanced adenoma, N=4) and those with carcinomas received surgical resection (N=~14), surgical resection with chemotherapy (N=~6), and surgical resection with chemotherapy and radiation (N=~6).
+Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`).
 We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power.
-We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P<0.XXX).
-Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P=0.XX).
-Finally, the change in the positive probability was not significantly different between the two treatment groups (P=0.XX).
+We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P= `r chemo_rad_summary["dist", "bh"]`).
+Finally, the change in the positive probability was not significantly different between the two treatment groups (P=`r chemo_rad_summary["red_crc", "bh"]`).
 Due to the relatively small number of samples in each treatment group, it is difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota
@@
 A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+Solid black points represent the median value for each diagnosis group.
 B) NMDS of the pre- and post-treatment samples for the adenoma group.
@@
 D) NMDS of the pre- and post-treatment samples for the carcinoma group.
-
-A: Convert mean to median w/ IQR. Add 0.00 to y-labels
 

### 2017-04-14 submission/manuscript.tex

@@
 
-$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI
+$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, M.
 
-$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, P.
 
@@
 
-\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
-problem and research suggests a correlation between the fecal bacterial
-microbiome and CRC. Despite this, very little is known about what
-happens to the microbiome after treatment for an adenoma or carcinoma.
-This study tested the hypothesis that treatment for adenoma or carcinoma
-results in changes towards a normal bacterial community. Specifically,
-we tried to identify components within the community that were different
-before and after treatment of adenoma, advanced adenoma, and carcinoma.
+\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
+problem. Despite growing evidence that members of the gut microbiota can
+drive tumorigenesis, little is known about what happens to the
+microbiome after treatment for an adenoma or carcinoma. This study
+tested the hypothesis that treatment for adenoma or carcinoma alters the
+abundance of bacterial populations associated with with disease to those
+associated with a normal colon. We tested this hypothesis by sequencing
+the 16S rRNA genes in the feces of 67 individuals before and after
+treatment for adenoma (N=22), advanced adenoma (N=19), and carcinoma
+(N=26).
 
-\textbf{Results:} There was a larger change in the bacterial community
-in response to treatment for carcinoma versus adenoma (P-value
-\textless{} 0.05) but not carcinoma versus advanced adenoma (P-value
-\textgreater{} 0.05). There was a trend for increasingly less community
-similarity, between samples pre- and post-treatment from adenoma to
-advanced adenoma to carcinoma. Despite this, no difference was found in
-the relative abundance of any specific OTU before and after treatment
-for adenoma, advanced adenoma, or carcinoma groups (P-value
-\textgreater{} 0.05). Using Random Forest models to assess whether
-changes in post-treatment samples were towards a normal community, only
-those with carcinoma had a significant decrease in positive probability
-(P-value \textless{} 0.05); providing further evidence that treatment
-has the greatest effect in those with carcinoma. The adenoma model used
-a total of 62 OTUs, the SRN model used a total of 61 OTUs, and the
-carcinoma model used a total of 59 OTUs. A total of 26 OTUs were common
-to all three models with many classifying to commensal bacteria (e.g.
-\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Anaerostipes},
-\emph{Blautia}, and \emph{Dorea}). Both chemotherapy and radiation did
-not provide any additional changes to the bacterial community in those
-treated for carcinoma (P-value \textgreater{} 0.05).
+\textbf{Results.} There were large changes the the bacterial communities
+associated with treatment across the three groups. The communities from
+patients with carcinomas changed significantly more than those with
+adenoma following treatment (P-value=5.4e-05); however, there was not a
+significant difference between those with advanced adenoma and those
+with adenoma or carcinoma (P-value\textgreater{}0.05). Although
+treatment brought about large intrapersonal changes, the change in the
+abundance of individual OTUs to treatment was not consistent within
+diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
+of OTUs across patients and diagnosis groups was patchy, we used the
+Random Forest machine learning algorithm to identify groups of OTUs that
+allowed us to successfully distinguish between pre- and post-treatment
+samples for each of the diagnosis groups. However, across the three
+models, there was little overlap between the OTUs that were indicative
+of treatment. Next, we used a larger cohort that contained individuals
+with normal colons and those with adenomas, advanced adenomas, and
+carcinomas to determine whether individuals who underwent a treatment
+were more likely to have OTUs associated with normal colons. We again
+built Random Forest models and measured the change in the positive
+probability of having one of the three diagnoses. Although we could
+clearly differentiate pre- and post-treatment communities from the three
+diagnosis groups, only those patients that initially had carcinomas
+experienced a significant decrease in positive probability of having a
+carcinoma (P-value \textless{} 0.05). Finally, tested whether the type
+of treatment impacted the microbiota of those diagnosed with carcinomas
+and were unable to detect any significant differences in characteristics
+of these communities between individuals treated with chemotherapy and
+those treated with chemotherapy and radiation (P-value \textgreater{}
+0.05).
 
-\textbf{Conclusions:} Our data partially supports the hypothesis that
-the microbiome changes after treatment towards a normal community.
-Individuals with carcinoma had more drastic differences to the overall
-community than those with adenoma. Common OTUs to all models were
-overwhelmingly from commensal bacteria, suggesting that these bacteria
-may be important in initial polyp formation, development of advanced
-adenoma, and transition to carcinoma.
+\textbf{Conclusions.} Further exploration of the relationship between
+diagnosis, treatment, and the impact on the microbiome will yield
+improvements in disease management.
 
@@
 bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk
-factors
+factor.
 
@@
 
+\linenumbers
+
 \subsection{Background}\label{background}
 
-Colorectal cancer (CRC) is currently the third most common cause of
-cancer deaths {[}1,2{]}. The rate of disease mortality has seen a
-significant decrease, thanks mainly to improvements in screening
-{[}1{]}. However, despite this improvement there are still approximately
-50,000 deaths from the disease per year {[}2{]}. Some of these deaths
-are due to disease recurrence, with current estimates that 20-30\% of
-those undergoing resection having a CRC recurrence {[}3{]}. This is
-important since approximately 30-50\% of those who have a recurrence
-will die of CRC {[}4{]}. To reduce the burden on the healthcare system,
-finding ways to better stratify those at highest risk of recurrence is
-of great importance.
+Colorectal cancer (CRC) is the third most common cause of cancer deaths
+in the United States {[}1,2{]}. Disease mortality has significantly
+decreased, thanks to improvements in screening {[}1{]}. Despite this
+improvement, there are still approximately 50,000 deaths from the
+disease per year {[}2{]}. Current estimates indicate that 20-30\% of
+those who undergo treatment will experience recurrence and 30-50\% of
+those patients will die {[}3,4{]}. Identification of methods to assess
+patients' risk of recurrence is of great importance to reduce mortality
+and healthcare costs.
 
-Recent studies in humans and mouse models have shown that altered
-membership and structure of the gut microbiome correlate with CRC
-pathogenesis {[}5--7{]}. Further, bacterial communities on the mucosa
-are altered between normal and tumor tissue {[}8{]}. Collectively, these
-studies provide a tantalizing link between our gut bacteria and CRC and
-suggest the possibility of using biomarkers to diagnose disease. Indeed,
-models created with microbiome data and clinical tests, such as Fecal
-Immunoglobulin Test (FIT), result in good predictions of CRC {[}9,10{]}.
-While these studies show how changed microbial communities or invasion
-by inflammatory species can impact CRC progression {[}11{]}, they
-provide very little information as to whether these communities change
-and rebound towards normal after successful treatment of adenoma,
-advanced adenoma, or carcinoma.
+There is growing evidence that the gut microbiota is involved in the
+progression of CRC. Mouse-based studies have identified populations of
+\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
+\emph{Fusobacterium nucleatum} that alter disease progression
+{[}5--9{]}. Furthermore, studies that alter the structure of the
+microbiota through the use of antibiotics or inoculation of germ free
+mice with human feces has shown that varying community compositions can
+result in varied tumor burden {[}10--12{]}. Collectively, these studies
+hypothesize that the microbiota is altering the amount of inflammation
+in the colon and with it the rate of tumorigenesis {[}13{]}.
 
-Providing an answer to this question is important because it has far
-reaching implications on both how the bacterial community causes the
-formation of more polyps {[}5,6{]} and the ability to use the microbiome
-as a predictive screening tool {[}9,10{]}. Understanding polyp formation
-and transition to advanced adenoma and then carcinoma is crucial to
-being able to understand how to prevent CRC occurrence. Response of the
-community to treatment is also equally important to predictive models
-designed for screening purposes since an unresponsive community would
-provide little additional information for important events, such as
-recurrence {[}12{]}.
+Building on this evidence, several human-based studies have identified
+unique signatures of colonic lesions {[}14--19{]}. One line of research
+has identified community-level differences between those bacteria that
+are found on and adjacent to colonic lesions and have supported a role
+for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
+\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
+have proposed feces-based biomarkers that could be used to diagnose the
+presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
+have associated \emph{Fusobacterium nucleatum} and other oral pathogens
+with lesions. They have also noted that the loss of bacteria generally
+thought to produce short chain fatty acids, which can suppress
+inflammation, is associated with colonic lesions. Gut bacteria have a
+role in tumorigenesis and there is promise that these populations may be
+useful biomarkers for aiding in the early detection of disease.
 
-Using pre- (initial) and post- (follow up) treatment samples we tested
-the hypothesis that treatment causes detectable changes towards a normal
-microbiome in those with adenoma, advanced adenoma, and carincoma.
-First, we assessed differences between samples pre- and post-treatment
-in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity
-metrics. Second, we explored whether models built to classify adenoma,
-advanced adenoma, or carcinoma versus normal were able to identify
-specific community members that differed between pre- and post-treatment
-samples. We also used these models to assess whether changes in the
-community were toward a more normal microbiome. Finally, we assessed
-both whether surgery for adenomas and advanced adenomas provided larger
-community changes and whether chemotherapy or radiation provided
-additional changes to the microbiome over surgical resection. This study
-aims to provide evidence as to whether the altered CRC microbiome
-persists or shifts back towards normal after such interventions.
+Despite advances in understanding the role between the gut microbiota
+and colonic tumorigenesis, we still do not understand how treatments
+including resection, chemotherapy, and/or radiation affect the
+composition of the gut microbiota. If the community drives tumorigenesis
+then one would hypothesize that treatment to remove a lesion would
+affect the microbiota and risk of recurrence. To test this hypothesis,
+we addressed two related questions: Does treatment affect the colonic
+microbiota in a predictable manner? If so, does the treatment alter the
+community to more closely resemble that of individuals with normal
+colons?
+
+We answered these questions by sequencing the V4 region of 16S rRNA
+genes amplified from fecal samples of individuals with adenoma, advanced
+adenoma, and carcinomas pre- and post-treatment. We used both classical
+community analysis to compare the alpha and beta-diversity of
+communities pre- and post treatment. Next, we generated Random Forest
+models to identify bacterial populations that were indicative of
+treatment for each diagnosis group. Finally, we measured the predictive
+probability of diagnosis models to determine whether treatment would
+alter the probability of being diagnosed as having a normal colon or the
+colon of the original diagnosis group. Ultimately, we found that
+treatment alters the composition of the gut microbiota; however, we only
+observed a shift from diseased to normal colon for those initially
+diagnosed with carcinomas. Understanding how the community responds to
+these treatments could be a valuable tool for identifying biomarkers to
+quantify the risk of recurrence.
 
@@
 
-\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
-tested whether the microbiome in patients with adenoma (n = 22),
-advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
-differences between pre- and post-treatment samples. We found that
-carcinoma patients had a more dissimilar bacterial community between
-these samples than those with adenoma (P-value \textless{} 0.001)
-{[}Figure 1A{]}. Although no significant differences were observed
-between advanced adenoma and carcinoma there was an increase in the
-dissimilarity of pre- and post-treatment samples from adenoma (0.55 ±
-0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78 ±
-0.15) {[}Figure 1A{]}. The bacterial community structure pre-
-post-treatment was visualized using NMDS for adenoma {[}Figure 1B{]}
-(PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
-(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
-(PERMANOVA \textless{} 0.05). Interestingly, when pre- and
-post-treatment samples were compared, regardless of whether they were
-adenoma or carcinoma, there was no significant overall difference in
-beta diversity (PERMANOVA \textgreater{} 0.05). There was no difference
-between pre- and post-treatment samples when investigating alpha
-diversity metrics for adenoma, advanced adenoma, or carcinoma {[}Table
-S1{]}. Additionally, there was also no difference in the relative
-abundance of any specific OTU between these samples for adenoma,
-advanced adenoma, or carcinoma only {[}Figure S1{]}.
+\textbf{\emph{Treatment alters the bacterial community structure of
+patients diagnosed with colonic lesions.}} Within our 67-person cohort
+we tested whether the microbiome of patients with adenoma (N=22),
+advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
+between pre- and post-treatment samples (Table 1). The structure of the
+microbial communities of the pre- and post-treatment samples differed
+{[}Figure 1A{]}. We found that the communities obtained pre- and
+post-treament among the patients with carcinomas changed significantly
+more than those patients with adenoma (P-value \textless{} 0.001). There
+were no significant differences in the amount of change observed between
+the patients with adenoma and advanced adenoma or between the patients
+with advanced adenoma and carcinoma (P\textgreater{}0.05). Next, we
+tested whether there was a consistent direction in the change in the
+community structure between the pre and post-treatment samples for each
+of the diagnosis groups {[}Figure 1BCD{]}. We only observed a consistent
+shift in community structure for the patients with carcinoma when using
+a PERMANOVA test (adenoma P-value0.999, advanced adenoma P-value0.945,
+and carcinoma P-value0.005). Finally, we measured the number of observed
+OTUs, Shannon evenness, and Shannon diversity in the pre and
+post-treatment samples and did not observe a significant change for any
+of the diagnosis groups (P-value\textgreater{}0.05) {[}Table S1{]}.
 
-\textbf{\emph{Adenoma Model:}} Using normal and adenoma individuals from
-a separate cohort we created an adenoma model to help assess whether
-those with adenoma in this cohort changed towards normal. The range of
-model AUC's from 100 runs of 20 repeated 10-fold cross-validation was
-0.62 - 0.72 with the AUC of the model used for classification having an
-AUC of 0.65 {[}Figure S2A{]}. There was a total of 62 OTUs in this model
-with the vast majority classifying to bacteria typically thought of as
-commensal {[}Figure S3A{]}. There was a significant difference between
-the actual and predicted disease classification (P-value \textless{}
-0.05). There was also no significant decrease in the positive
-probability of adenoma between pre- and post-treatment samples (P-value
-\textgreater{} 0.05) {[}Figure 2A{]}.
+\textbf{\emph{The effects of treatment are not consistent across
+treatment groups.}} We used two approaches to identify those bacterial
+populations that change between the two samples for each diagnosis
+group. First, we sought to identify individual OTUs that could account
+for the change in overall community structure. However, using a paired
+Wilcoxon test we were unable to identify any OTUs that were
+significantly different in the pre and post-treatment groups {[}Figure
+S1{]}. It is likely that high inter-individual variation and the patchy
+distribution of OTUs across individuals limited the statistical power of
+the test. To overcome these problems we developed Random Forest models
+to identify collections of OTUs that would allow us to differentiate
+between pre and post-treatment samples from each of the diagnosis
+groups. Because of the relatively small number of subject in each group,
+we restricted our models to only incorporate 10 OTUs. Despite this
+restriction, the models performed well (adenoma AUC=0.69 - 0.92,
+advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
+the 10 OTUs that were used for each model had little overlap with each
+other {[}Figure 2{]}. These results support the earlier community-wide
+analysis where we observed that the treatment had an impact on the
+overall community structure; however, the effect of treatment was not
+consistent across patients and diagnosis groups.
 
-\textbf{\emph{Advanced Adenoma Model:}} Using normal and advanced
-adenoma individuals from a separate cohort we created an advanced
-adenoma model to help assess whether those with advanced adenoma in this
-cohort changed towards normal. The range of model AUC's from 100 runs of
-20 repeated 10-fold cross-validation was 0.68 - 0.77 with the AUC of the
-model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
-There was a total of 61 OTUs in the advanced adenoma model {[}Figure
-S3B{]}. Similar to the adenoma model the vast majority of OTUs
-classified to bacteria typically thought of as commensal. Also similar
-to the adenoma model there was a significant difference between the
-actual and predicted disease classification (P-value \textless{} 0.05)
-and no significant decrease in the positive probability of advanced
-adenoma between pre- and post-treatment samples (P-value \textgreater{}
-0.05) {[}Figure 2B{]}.
+\textbf{\emph{Post-treatment samples from patients with carcinoma
+share}} Next, we determined whether treatment changed the microbiota in
+a way that the post-treatment communities resembled that of patients
+with normal colons. To test this, we used an expanded cohort of 423
+individuals that were diagnosed under the same protocol as having normal
+colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
+We then constructed Random Forest models to classify the pre and
+post-treatment samples as having their original diagnosis or having a
+normal colon. The models performed well (adenoma AUC=0.62 - 0.72,
+advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
+The OTUs that were most important in the models for differentiating
+between patients with normal colons and those with adenomas or advanced
+adenomas were dominated by populations commonly considered as commensals
+(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
+\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
+\emph{Roseburia}) {[}Figure S3{]}. Although many of these OTUs were also
+included in the model differentiating between patients with normal
+colons and those with carcinoma, this model also included OTUs
+affiliated with populations that have previously been associated with
+carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
+\emph{Parvimonas}) {[}Figure S3{]}. Finally, we applied these three
+models to the pre and post-treatment samples for each diagnosis group
+and quantified the change in the positive probability of the model. A
+decrease in the positive probability would indicate that the colon more
+closely resembled that of a patient with a normal colon. There was not a
+significant change in the positive probability for the adenoma or
+advanced adenoma groups {[}Figure 3{]}. The positive probability for the
+pre and post-treatment samples from patients diagnosed with carcinoma
+significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
+that were initially diagnosed with a carcinoma had a higher positive
+probability after treatment. One of those was re-diagnosed with
+carcinoma on the follow up visit. These results indicate that although
+there are changes in the microbiota associated with treatment, those
+experienced by the patients in the carcinoma group were the only ones
+where the change was directed towards what would be found in a normal
+colon
 
-\textbf{\emph{Carcinoma Model:}} Using normal and carcinoma individuals
-from a separate cohort we created a carcinoma model to help assess
-whether those with carcinoma in this cohort changed towards normal. The
-range of model AUC's from 100 runs of 20 repeated 10-fold
-cross-validation was 0.84 - 0.9 with the AUC of the model used for
-classification being 0.88 {[}Figure S2C{]}. Interestingly, the AUCs
-improved from adenoma to advanced adenoma to carcinoma {[}Figure 2{]}.
-There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
-Similar to the adenoma and advanced adenoma models the vast majority of
-OTUs classified to bacteria typically thought of as commensal but OTUs
-that also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
-\emph{Parvimonas} were also important for carcinoma classification
-{[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
-there was a significant difference between the actual and predicted
-disease classification (P-value \textless{} 0.05). There was a
-significant decrease in the positive probability of carcinoma between
-pre- and post-treatment samples (P-value \textless{} 0.05) {[}Figure
-2C{]}; suggesting that the carcinoma samples changed towards normal
-after treatment, unlike either adenoma or advanced adenoma. The one
-individual still positive for carcinoma after treatment had an increase
-in carcinoma positive proability between their pre- and post-treatment
-sample {[}Figure 2C{]}.
-
-\textbf{\emph{Common OTUs:}} Next, to identify which OTUs were important
-at all stages of disease, we identified common predictive OTUs within
-the adenoma, advanced adenoma, and carcinoma models. When we compared
-the three different models with each other there were a total of 26
-common OTUs. Some of the most common taxonomic identifications belonged
-to \emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
-\emph{Lachnospiraceae}, and \emph{Dorea}. The vast majority of the OTUs
-that were common between these models had classifications to bacteria
-typically thought of as commensal {[}Table S2{]}. The rank of these
-OTUs, based on median mean decrease in accuracy (MDA), varied higly
-depending on whether they were in the adenoma, advanced adenoma, or
-carcinoma model {[}Figure 3{]}. As an example, \emph{Bacteroides}
-(OTU85) was ranked 14\textsuperscript{th} in the carcinoma model but in
-both the adenoma and advanced adenoma model ranked
-35\textsuperscript{th} and 41\textsuperscript{st} respectively {[}Figure
-3{]}.
-
-\textbf{\emph{Treatment Affects on Community:}} After observing these
-changes from treatment we assessed the possible confounding impact of
-chemotherapy or radiation, in the carcinoma group, and surgical
-resection, in the adenoma group, on the observed results. In the
-carcinoma group neither chemotherapy nor radiation provided any
-additional change from initial sample over surgery alone (P-value
-\textgreater{} 0.05) {[}Table S3{]}. For the adenoma group there was a
-single difference in observed OTUs (sobs) between those that received
-surgerical resection and those that had regular polyp removal (P-value
-\textless{} 0.05) {[}Table S4{]}. For the surgical resection comparison,
-adenoma and advanced adenoma were combined due to the low number of
-resection occurances in these two groups. Using a fisher exact test
-there was no difference in the proportion of those receiving surgerical
-resection between the adenoma and advanced adenoma groups (P-value
-\textgreater{} 0.05). This data suggests that the microbiome changes
-observed in the carcinoma group were mostly a result from surgerical
-resection and not from chemotherapy or radiation.
+\textbf{\emph{Difficult to identify effects of specific treatments on
+the change in the microbiota.}} The type of treatment that the patients
+received varied across diagnosis groups. Those with adenomas and
+advanced adenomas received surgical resection (adenoma, N=4; advanced
+adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
+advanced adenoma, N=15) and those with carcinomas received surgical
+resection (N=12), surgical resection with chemotherapy (N=9), and
+surgical resection with chemotherapy and radiation (N=5). We focused on
+the patients with carcinoma and pooled those patients that received
+chemotherapy with those that received chemotherapy and radiation to
+improve our statistical power. We did not observe a significant
+difference in the effect of these treatments on the number of observed
+OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
+0.05). Furthermore, there was not a significant difference in the effect
+of the treatments on the amount of change in the community structure (P=
+1). Finally, the change in the positive probability was not
+significantly different between the two treatment groups (P=1). Due to
+the relatively small number of samples in each treatment group, it is
+difficult to make a definitive statement regarding the specific type of
+treatment on the amount of change in the structure of the microbiota
 
@@
 without CRC differs and how it might be used as an early screening tool
-{[}9,10,13--15{]}. Here we show that the microbiome changes towards
-normal after treatment for carcinoma and that chemotherapy and radiation
-did not provide any additional change. Although some of the important
-OTUs classified to genera from bacteria considered the ``usual
-suspects'' (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
-\emph{Parvimonas}) many did not. The majority of important OTUs had
-taxonomic classifications for resident gut microbes and were common for
-the adenoma, advanced adenoma, and carcinoma models. This suggests that
+{[}23--27{]}. Here we show that the microbiome changes towards normal
+after treatment for carcinoma and that chemotherapy and radiation did
+not provide any additional change. Although some of the important OTUs
+classified to genera from bacteria considered the ``usual suspects''
+(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
+many did not. The majority of important OTUs had taxonomic
+classifications for resident gut microbes and were common for the
+adenoma, advanced adenoma, and carcinoma models. This suggests that
 members within the commensal community may be the first that change
@@
 first step in allowing more inflammatory bacteria to gain a foothold
-within the colon {[}11{]}.
+within the colon {[}13{]}.
 
@@
 potential butyrate producers (e.g. \emph{Clostridiales},
-\emph{Roseburia}, and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
+\emph{Roseburia}. and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
 classified to bacteria that are inhibited by polyphenols (e.g.
 \emph{Bacteroides}). Both butyrate and polyphenols are thought to be
-protective against cancer, in part by reducing inflammation {[}16{]}.
+protective against cancer, in part by reducing inflammation {[}28{]}.
 These protective compounds are derived from the breakdown of fiber,
@@
 to be increased in those with high non-meat based protein consumption
-{[}17{]}. High protein consumption in general has been linked with an
-increased CRC risk {[}18{]}. Conversely, \emph{Bacteroides} are
+{[}29{]}. High protein consumption in general has been linked with an
+increased CRC risk {[}30{]}. Conversely, \emph{Bacteroides} are
 inhibited by polyphenols which are derived from fruits and vegetables
-{[}19{]}. Our data fits with the hypothesis that the microbial
+{[}31{]}. Our data fits with the hypothesis that the microbial
 metabolites from breakdown products within our own diet could not only
@@
 whether different interventions could potentially change this community
-and lower the probability of future reccurence.
+and lower the probability of future recurrence.
 
@@
 \textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
-have been previously reported in Baxter, et al {[}9{]}. Briefly, study
+have been previously reported in Baxter, et al {[}23{]}. Briefly, study
 exclusion involved those who had already undergone surgery, radiation,
@@
 normalize back to a pre-colonoscopy community within this time period
-{[}20{]}. Our training cohort consisted of a total of 423 individuals
+{[}32{]}. Our training cohort consisted of a total of 423 individuals
 {[}Table 1{]}. Our study cohort consisted of 67 individuals with an
@@
 \textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
-described by Kozich, et al. {[}21{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}33{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}21{]}. The general workflow using mothur was:
+previously described {[}33{]}. The general workflow using mothur was:
 Paired-end reads were first merged into contigs, quality filtered,
@@
 
-\textbf{\emph{Model Building:}} The Random Forest {[}22{]} algorithm was
+\textbf{\emph{Model Building:}} The Random Forest {[}34{]} algorithm was
 used to create the model used to create the three models used. The
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}23{]} in the vegan package
+community structure utilized PERMANOVA {[}35{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
-\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
-and Carcinoma Groups After Treatment.} A) A significant difference was
+**Figure 1: General Differences between Adenoma, Advanced Adenoma, and
+Carcinoma Groups After Treatment.*. A) A significant difference was
 found between the adenoma and carcinoma group for thetayc (P-value =
-NULL). B) NMDS of the pre- and post-treatment samples for the adenoma
+NULL). Solid black points represent the median value for each diagnosis
+group. B) NMDS of the pre- and post-treatment samples for the adenoma
 group. C) NMDS of the pre- and post-treatment samples for the advanced
@@
 
-\textbf{Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and
-Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
-represent the range of values of a 100 different 80/20 splits of the
-test set data and the dark green line represents the model using 100\%
-of the data set and what was used for subsequent classification. B)
-Advanced Adenoma ROC curve: The light yellow shaded areas represent the
-range of values of a 100 different 80/20 splits of the test set data and
-the dark yellow line represents the model using 100\% of the data set
-and what was used for subsequent classification. C) Carcinoma ROC curve:
-The light red shaded areas represent the range of values of a 100
-different 80/20 splits of the test set data and the dark red line
+**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma
+Models.*. A) Adenoma ROC curve: The light green shaded areas represent
+the range of values of a 100 different 80/20 splits of the test set data
+and the dark green line represents the model using 100\% of the data set
+and what was used for subsequent classification. B) Advanced Adenoma ROC
+curve: The light yellow shaded areas represent the range of values of a
+100 different 80/20 splits of the test set data and the dark yellow line
 represents the model using 100\% of the data set and what was used for
-subsequent classification.
+subsequent classification. C) Carcinoma ROC curve: The light red shaded
+areas represent the range of values of a 100 different 80/20 splits of
+the test set data and the dark red line represents the model using 100\%
+of the data set and what was used for subsequent classification.
 
-\textbf{Figure S3: Summary of Important Variables for the Adenoma,
-Advanced Adenoma, and Carcinoma Models.} A) MDA of the most important
-variables in the adenoma model. The dark green point represents the mean
-and the lighter green points are the value of each of the 100 different
-runs. B) Summary of Important Variables in the advanced adenoma model.
-MDA of the most important variables in the SRN model. The dark yellow
-point represents the mean and the lighter yellow points are the value of
-each of the 100 different runs. C) MDA of the most important variables
-in the carcinoma model. The dark red point represents the mean and the
-lighter red points are the value of each of the 100 different runs.
+**Figure S3: Summary of Important Variables for the Adenoma, Advanced
+Adenoma, and Carcinoma Models.*. A) MDA of the most important variables
+in the adenoma model. The dark green point represents the mean and the
+lighter green points are the value of each of the 100 different runs. B)
+Summary of Important Variables in the advanced adenoma model. MDA of the
+most important variables in the SRN model. The dark yellow point
+represents the mean and the lighter yellow points are the value of each
+of the 100 different runs. C) MDA of the most important variables in the
+carcinoma model. The dark red point represents the mean and the lighter
+red points are the value of each of the 100 different runs.
 
@@
 
-\hypertarget{ref-zackular_manipulation_2016}{}
-5. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
-Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
+\hypertarget{ref-goodwin_polyamine_2011}{}
+5. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
+TR, et al. Polyamine catabolism contributes to enterotoxigenic
+Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
+National Academy of Sciences of the United States of America.
+2011;108:15354--9.
+
+\hypertarget{ref-abed_fap2_2016}{}
+6. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
+Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
+Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
+Microbe. 2016;20:215--25.
 
 \hypertarget{ref-arthur_microbial_2014}{}
-6. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+7. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
 McCafferty J, et al. Microbial genomic analysis reveals the essential
@@
 
+\hypertarget{ref-kostic_fusobacterium_2013}{}
+8. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
+et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
+modulates the tumor-immune microenvironment. Cell Host \& Microbe.
+2013;14:207--15.
+
+\hypertarget{ref-wu_human_2009}{}
+9. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
+human colonic commensal promotes colon tumorigenesis via activation of T
+helper type 17 T cell responses. Nature Medicine. 2009;15:1016--22.
+
+\hypertarget{ref-zackular_manipulation_2016}{}
+10. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
+
 \hypertarget{ref-zackular_gut_2013}{}
-7. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen GY,
-et al. The gut microbiome modulates colon tumorigenesis. mBio.
+11. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
+GY, et al. The gut microbiome modulates colon tumorigenesis. mBio.
 2013;4:e00692--00613.
 
+\hypertarget{ref-baxter_structure_2014}{}
+12. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
+microbiome following colonization with human feces determines colonic
+tumor burden. Microbiome. 2014;2:20.
+
+\hypertarget{ref-flynn_metabolic_2016}{}
+13. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
+
+\hypertarget{ref-wang_structural_2012}{}
+14. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+segregation of gut microbiota between colorectal cancer patients and
+healthy volunteers. The ISME journal. 2012;6:320--9.
+
+\hypertarget{ref-chen_decreased_2013}{}
+15. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, et al.
+Decreased dietary fiber intake and structural alteration of gut
+microbiota in patients with advanced colorectal adenoma. The American
+Journal of Clinical Nutrition. 2013;97:1044--52.
+
+\hypertarget{ref-chen_human_2012}{}
+16. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+mucosa-associated microbiota in patients with colorectal cancer. PloS
+One. 2012;7:e39743.
+
+\hypertarget{ref-shen_molecular_2010}{}
+17. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, et al.
+Molecular characterization of mucosal adherent bacteria and associations
+with colorectal adenomas. Gut Microbes. 2010;1:138--47.
+
+\hypertarget{ref-kostic_genomic_2012}{}
+18. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+al. Genomic analysis identifies association of Fusobacterium with
+colorectal carcinoma. Genome Research. 2012;22:292--8.
+
+\hypertarget{ref-feng_gut_2015}{}
+19. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, et al. Gut
+microbiome development along the colorectal adenoma-carcinoma sequence.
+Nature Communications. 2015;6:6528.
+
 \hypertarget{ref-dejea_microbiota_2014}{}
-8. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+20. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 
+\hypertarget{ref-mima_fusobacterium_2015}{}
+21. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, et
+al. Fusobacterium nucleatum and T Cells in Colorectal Carcinoma. JAMA
+oncology. 2015;1:653--61.
+
+\hypertarget{ref-arthur_intestinal_2012}{}
+22. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+T-J, et al. Intestinal inflammation targets cancer-inducing activity of
+the microbiota. Science (New York, N.Y.). 2012;338:120--3.
+
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-9. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+23. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-zeller_potential_2014}{}
-10. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+24. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 
-\hypertarget{ref-flynn_metabolic_2016}{}
-11. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
-Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
-
-\hypertarget{ref-hassan_efficacy_2016}{}
-12. Hassan C, Repici A, Sharma P, Correale L, Zullo A, Bretthauer M, et
-al. Efficacy and safety of endoscopic resection of large colorectal
-polyps: A systematic review and meta-analysis. Gut. 2016;65:806--20.
+\hypertarget{ref-zackular_human_2014}{}
+25. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+microbiome as a screening tool for colorectal cancer. Cancer Prevention
+Research (Philadelphia, Pa.). 2014;7:1112--21.
 
 \hypertarget{ref-yu_metagenomic_2017}{}
-13. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
+26. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
 analysis of faecal microbiome as a tool towards targeted non-invasive
@@
 
-\hypertarget{ref-zackular_human_2014}{}
-14. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
-microbiome as a screening tool for colorectal cancer. Cancer Prevention
-Research (Philadelphia, Pa.). 2014;7:1112--21.
-
 \hypertarget{ref-warren_co-occurrence_2013}{}
-15. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
+27. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
 et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
@@
 \hypertarget{ref-louis_gut_2014}{}
-16. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
+28. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
 metabolites and colorectal cancer. Nature Reviews Microbiology
@@
 \hypertarget{ref-zhu_intake_2016}{}
-17. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
+29. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
 Proteins Substantially Increased the Relative Abundance of Genus
@@
 \hypertarget{ref-mu_colonic_2016}{}
-18. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
+30. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
 Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
@@
 \hypertarget{ref-ozdal_reciprocal_2016}{}
-19. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
+31. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
 Reciprocal Interactions between Polyphenols and Gut Microbiota and
@@
 \hypertarget{ref-obrien_impact_2013}{}
-20. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+32. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-21. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+33. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-22. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+34. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-23. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-04-17 submission/manuscript.Rmd
Editing

@@
 To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs.
+Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data.
 Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`).
@@
 The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2).
-The OTUs that were most important in the models for differentiating between patients with normal colons and those with adenomas or advanced adenomas were dominated by populations commonly considered as commensals (e.g.
-*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figure S3].
+The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 4A].
+Among the OTUs that were shared across the three models were those populations commonly considered as commensals (e.g.
+*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 4B].
 Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3].
@@
 ## Discussion
-This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differs and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
-Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide any additional change.
+
+This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [ADD REFERENCES FOR CAUSATION;  @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
+Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
+For all three groups of patients we observed changes in their microbiota.
+The microbiota of patients diagnosed with carcinoma changed significant more than the other groups.
+This change resulted in communities that more closely resembled those of patients with normal colons.
+
+* Does treatment affect the colonic microbiota in a predictable manner? For those with carcinoma
+
+* Does the treatment alter the community to more closely resemble that of individuals with normal colons? For those with carcinoma.
+
+* Heterogeneity of the population (common problem for all microbiome studies) made it impossible to identify speicfi microbiota that changed with treatment; however, RF made it possible to identify consortia of microbiota that were associated with treatment and recovery.
+
+* End.
+
+Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3].
+Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
+
+Here we show that the microbiome changes towards normal after treatment for carcinoma
+
 Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g.
@@
 These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].
-
-Unlike previous studies on the microbiome and CRC, ours focuses on comparing the microbiome during recovery from treatment in adenoma, advanced adenoma, and carcinoma groups.
-Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3].
-Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
 
@@
 Within this context the commensal community may be an important modifiable risk factor for monitoring and preventing CRC recurrence after treatment.
-
-A limitation in our study was that there was a significant difference in the time elapsed in the collection of the post-treatment sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).
-These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time.
-Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1].
-However, there are two reasons that this may not be the case.
-First, the advanced adenoma group had a higher dissimilarity than adenoma but lower dissimilarity then carcinoma and the collection time to their follow up samples was less than the adenoma group.
-Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].
-
-Another limitation of this study was that it drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.
-Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.
-We have done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.
 

### 2017-04-17 submission/supplemental.Rmd

@@
 
-\blandscape
+**Table S2: Summary of Common OTUs for All Diagnosis Groups**
 
-**Table S2: Summary of Common OTUs for All Diagnosis Groups**
+\footnotesize
 
@@
 kable(st2, format = "markdown", align = 'c', padding = 0)
-
 ```
-
+\normalsize
 \newpage
@@
 **Table S3: Summary of Comparisons based on Chemotherapy or Radiation Received**
-
+\footnotesize
 ```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
@@
   mutate(comparison = ifelse(comparison == "rads" | comparison == "rad", invisible("Radiation"), invisible("Chemotherapy")),
-         variable = c(rep("sobs", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4)))
+         variable = c(rep("Richness", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4)))
 
@@
 ```
-
-\elandscape
+\normalsize
 
@@
   filter(X != "prop_surg_adn_srn") %>%
-  mutate(variable = c("sobs", "Shannon", "Evenness", "Thetayc", "Adenoma Only\nPositive Probability", "SRN Only\nPositive Probablity"))
+  mutate(variable = c("Richness", "Shannon", "Evenness", "Thetayc", "Adenoma Only\nPositive Probability", "SRN Only\nPositive Probablity"))
 

### 2017-04-18 code/check_crc_probs_increase.R
add proability test oral/non-oral bacteria

@@ -86,5 +86,36 @@ combined_residents <- tbl_df(combined_data$probs_increase) %>% rename(probs_incr
 write.csv(combined_residents, "data/process/tables/inc_probs_crc_oral_residents_data.csv", row.names = F)
 
 
+## Test differences
 
 
+test <- combined_residents %>% group_by(probs_increase, sampleType) %>% 
+  select(-EDRN) %>% summarise_each(funs(median, IQR))
+
+write.csv(test, "data/process/tables/inc_probs_crc_oral_residents_summary.csv", row.names = F)
+
+tbl_df(c(
+  wilcox.test(
+  as.data.frame(filter(combined_residents, sampleType == "initial" & probs_increase == "No"))[, "oral_path"], 
+  as.data.frame(filter(combined_residents, sampleType == "followup" & probs_increase == "No"))[, "oral_path"], 
+  paired = TRUE)$p.value, 
+  
+  wilcox.test(
+  as.data.frame(filter(combined_residents, sampleType == "initial" & probs_increase == "Yes"))[, "oral_path"], 
+  as.data.frame(filter(combined_residents, sampleType == "followup" & probs_increase == "Yes"))[, "oral_path"], 
+  paired = TRUE)$p.value, 
+  
+  wilcox.test(
+  as.data.frame(filter(combined_residents, sampleType == "initial" & probs_increase == "No"))[, "residents"], 
+  as.data.frame(filter(combined_residents, sampleType == "followup" & probs_increase == "No"))[, "residents"], 
+  paired = TRUE)$p.value, 
+  
+  wilcox.test(
+  as.data.frame(filter(combined_residents, sampleType == "initial" & probs_increase == "Yes"))[, "residents"], 
+  as.data.frame(filter(combined_residents, sampleType == "followup" & probs_increase == "Yes"))[, "residents"], 
+  paired = TRUE)$p.value)) %>% rename(pvalue = value) %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH"), 
+         test = c("ivf_noinc_oralpath", "ivf_yesinc_oralpath", 
+                  "ivf_noinc_resident", "ivf_yesinc_resident"))
+
+write.csv(test, "data/process/tables/inc_probs_crc_oral_residents_pvalue_summary.csv", row.names = F)

### 2017-04-18 code/graph_check_crc_probs_increase.R

@@ -1,0 +1,59 @@
+### Graph increased versus decreased probs
+### create graphs to visualise differences (if there are any)
+## Marc Sze
+
+## Load in necessary libraries
+source('code/functions.R')
+loadLibs(c("dplyr", "tidyr", "ggplot2", "gridExtra"))
+
+
+## Load needed data files
+residents <- read.csv("data/process/tables/inc_probs_crc_oral_residents_data.csv", 
+                      header = T, stringsAsFactors = F)
+
+
+## Graph the data
+
+oral_p <- ggplot(residents, aes(factor(sampleType, 
+                             levels = c("initial", "followup"), 
+                             labels = c("Pre", "Post")), oral_path)) + 
+  geom_jitter(aes(color = probs_increase), width = 0.1) + 
+  scale_color_manual(name = "Carcinoma\nProbabiliy\nIncrease", 
+                     values = c('#8968CD', '#FF00FF')) + 
+  ggtitle("A") + theme_bw() +  xlab("") + ylab("% Relative Abundance") + 
+  stat_summary(aes(group = probs_increase), fun.y = median, 
+               colour = c('#8968CD', '#8968CD', '#FF00FF', '#FF00FF'), geom = "point", size = 4.5) + 
+  stat_summary(aes(group = probs_increase), fun.y = median, 
+               colour = c('#8968CD', '#8968CD', '#FF00FF', '#FF00FF'), geom = "line", size = 1) + 
+  coord_cartesian(ylim = c(0, 15)) + 
+  theme(plot.title = element_text(face = "bold", hjust = -0.08, size = 20), 
+        legend.position = c(0.9, 0.4)) + 
+  annotate("text", 
+          label = paste("Oral Pathogens"),x = 1.5, y = 15, size = 4)
+
+
+
+non_oral_p <- ggplot(residents, aes(factor(sampleType, 
+                             levels = c("initial", "followup"), 
+                             labels = c("Pre", "Post")), residents)) + 
+  geom_jitter(aes(color = probs_increase), width = 0.1) + 
+  scale_color_manual(name = "Carcinoma\nProbabiliy\nIncrease", 
+                     values = c('#8968CD', '#FF00FF')) + 
+  ggtitle("B") + theme_bw() +  xlab("") + ylab("% Relative Abundance") + 
+  stat_summary(aes(group = probs_increase), fun.y = median, 
[…]

### 2017-04-18 submission/supplemental.Rmd
discussion changes

@@
 alpha_table <- read.csv("../data/process/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
-supplemental_table1 <- separate(alpha_table, X, c("comparison", "alpha_metric"), sep="_")
+supplemental_table1 <- separate(alpha_table, X, c("comparison", "alpha_metric"), sep="_") %>% 
+  mutate(alpha_metric = gsub("sobs", "richness", alpha_metric), 
+         comparison = c(rep("Adenoma", 3), rep("Advanced Adenoma", 3), rep("Carcinoma", 3)))
 
@@
 
-**Table S3: Summary of Comparisons based on Chemotherapy or Radiation Received**
+**Table S3: Summary of Comparisons Based on Additive Treatment Received (Chemotherapy or Chemotherapy and Radiation)**
 \footnotesize
@@
 
-chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv',
-                              header = T, stringsAsFactors = F) %>% filter(X != "chemo_v_rads_IF") %>%
-  separate(X, c("comparison", "variable"), sep = "_") %>%
-  mutate(comparison = ifelse(comparison == "rads" | comparison == "rad", invisible("Radiation"), invisible("Chemotherapy")),
-         variable = c(rep("Richness", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4)))
+chemo_rad_summary <- read.csv('../data/process/tables/crc_probs_chemo_rad_pvalue_summary.csv',
+                              header = T, stringsAsFactors = F) %>% filter(X != "chemo_v_rads_red") %>% 
+  separate(X, c("comparison", "variable"), sep = "_") %>% mutate(comparison = c("Richness", "Shannon", "Evenness", "Thetayc", "Positive Probability")) %>% select(-variable)
 
-st3 <- cbind(chemo_rad_summary$variable, chemo_rad_summary$comparison, format(chemo_rad_summary$chemo_rad_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$removal_only_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$cr_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$ro_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$pvalue, scientific = TRUE, digits = 2), format(chemo_rad_summary$bh, scientific = TRUE, digits = 2))
+st3 <- cbind(chemo_rad_summary$comparison, format(chemo_rad_summary$chemo_rad_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$removal_only_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$cr_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$ro_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$pvalue, scientific = TRUE, digits = 2), format(chemo_rad_summary$bh, scientific = TRUE, digits = 2))
 
-colnames(st3) <- c("Variable", "Additive Treatment", "Additive Treatment Mean", "Removal Mean", "Additive Treatment SD", "Removal SD", "P-value", "BH Corrected")
+colnames(st3) <- c("Variable", "Additive Treatment Mean", "Removal Mean", "Additive Treatment SD", "Removal SD", "P-value", "BH Corrected")
 
@@
   filter(X != "prop_surg_adn_srn") %>%
-  mutate(variable = c("Richness", "Shannon", "Evenness", "Thetayc", "Adenoma Only\nPositive Probability", "SRN Only\nPositive Probablity"))
+  mutate(variable = c("Richness", "Shannon", "Evenness", "Thetayc", "Adenoma Only\nPositive Probability", "Advanced Adenoma Only\nPositive Probablity"))
 

### 2017-04-18 code/Run_Figure1.R
move A/B headings to left and mmake axis the same

@@ -40,7 +40,7 @@ differences_graph <- grid.arrange(
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
           legend.title = element_text(face="bold"), 
           legend.position = "none", 
-          title = element_text(face="bold", hjust = 0), 
+          plot.title = element_text(face="bold", hjust = -0.3, size = 20), 
           panel.grid.major = element_blank(), 
           panel.grid.minor = element_blank()) + coord_fixed(ratio = 1.8), 
   
@@ -53,7 +53,7 @@ differences_graph <- grid.arrange(
     theme_bw() + coord_equal() + ggtitle("B") + ylim(-0.7, 0.7) + xlim(-0.7, 0.7) + 
     scale_color_manual(values = c('#00EE76', '#228B22')) + scale_fill_manual(values = c('#00EE76', '#228B22')) + 
     scale_shape_manual(values = c(16, 1)) + 
-    theme(plot.title = element_text(face = "bold", hjust = 0), 
+    theme(plot.title = element_text(face="bold", hjust = -0.32, size = 20), 
           legend.position = c(0.15, 0.11), 
           legend.title = element_blank(), 
           legend.key = element_blank(), 
@@ -71,7 +71,7 @@ differences_graph <- grid.arrange(
     theme_bw() + coord_equal() + ggtitle("C") + ylim(-0.7, 0.7) + xlim(-0.7, 0.7) + 
     scale_color_manual(values = c('#CDAD00', '#8B7500')) + scale_fill_manual(values = c('#CDAD00', '#8B7500')) + 
     scale_shape_manual(values = c(16, 1)) + 
-    theme(plot.title = element_text(face = "bold", hjust = 0), 
+    theme(plot.title = element_text(face="bold", hjust = -0.32, size = 20), 
           legend.position = c(0.15, 0.11), 
           legend.title = element_blank(), 
           legend.key = element_blank(), 
@@ -90,7 +90,7 @@ differences_graph <- grid.arrange(
     coord_equal() + ggtitle("D") + ylim(-0.7, 0.7) + xlim(-0.7, 0.7) +
     scale_color_manual(values = c('#EEA2AD', '#DC143C')) + scale_fill_manual(values = c('#EEA2AD', '#DC143C')) + 
     scale_shape_manual(values = c(16, 1)) + 
-    theme(plot.title = element_text(face = "bold", hjust = 0), 
+    theme(plot.title = element_text(face="bold", hjust = -0.32, size = 20), 
[…]

### 2017-04-18 code/old/common_all_models.R

@@ -35,7 +35,7 @@ venn_graph <- as.data.frame(cbind(g1 = c(4.5, 5.4, 6.2), g2 = c(6.5, 5.6, 6.5)))
   coord_cartesian(xlim = c(1:10), ylim = c(4:8.5)) + xlab("") + ylab("") + 
   theme_bw() + ggtitle("A") + 
   theme(legend.position = "none", 
-        title = element_text(face = "bold"), 
+        plot.title = element_text(face = "bold", size = 20), 
         axis.text = element_blank(), 
         panel.grid = element_blank(), 
         panel.border = element_blank(), 

### 2017-04-18 submission/manuscript.Rmd

@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factor.
+microbiome; colorectal cancer; polyps; treatment; risk factor.
 
@@
 
-This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [ADD REFERENCES FOR CAUSATION;  @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
+This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
 Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
 For all three groups of patients we observed changes in their microbiota.
-The microbiota of patients diagnosed with carcinoma changed significant more than the other groups.
+The microbiota of patients diagnosed with carcinoma changed significantly more than the other groups.
 This change resulted in communities that more closely resembled those of patients with normal colons.
 
-* Does treatment affect the colonic microbiota in a predictable manner? For those with carcinoma
+For the carcinoma group there was a larger change between their initial and follow up samples then for either adenoma or advanced adenoma.
+This change was directionaly the same for each individual with carcinoma.
+This observed change has  important implications for future research into CRC recurrence.
+Knowing that the community changes towards normal after treatment means that it is reasonable to hypothesize that the microbiome could be a good biomarker for stratifying individuals between those most and least at risk for CRC recurrence.
+Interestingly, our `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability fits nicely within the range quoted for CRC recurrence (20-30%) [@hellinger_reoperation_2006; @ryuk_predictive_2014].
 
-* Does the treatment alter the community to more closely resemble that of individuals with normal colons? For those with carcinoma.
+A potential limitation, within our study, is that we do not know whether individuals who had an increased positive probability for carcinoma on follow up eventually had a subsequent CRC recurrence.
+This information would help to strengthen the case for the ability of our model to be used as a disease stratification tool for CRC recurrence.
+This would be an exciting area of future research that would be able to link treatment changes within the microbiome and CRC recurrence together.
 
-* Heterogeneity of the population (common problem for all microbiome studies) made it impossible to identify speicfi microbiota that changed with treatment; however, RF made it possible to identify consortia of microbiota that were associated with treatment and recovery.
+The microbiome changes associated with carcinoma were found to be towards those with normal colons.
+This would suggest that treatment for CRC is not only successful for removing the carcinoma but also at removing bacteria associated with them.
+This is important due to the numerous studies that have pointed towards a link between the community and tumorgenesis [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
+The importance of the community to this process is further highlighted in our data [Figure 2, 4, & S3].
+Interestingly, those who had an increase in their positive probability of carcinoma had no change in their median value for oral pathogens found within the carcinoma model but had a decrease in non-oral pathogens [Figure S4].
+Although this finding was not statistically significant, this trend would further highlight the importance of community to tumorgenesis.
+Future studies with larger power may be able to confirm these reported observations.
 
-* End.
+Although we were not able to identify specific OTUs due to the heterogeneity of the population.
+This is a common problem that occurs in other fields of microbiome study [@sze_looking_2016].
+This made it very difficult to identify individual OTUs associated with treatment and recovery towards a normal colon.
+To overcome this large individual heterogeneity we used RF models that made it easier to identify consortia of microbiota that were associated with treatment and recovery.
+A potential future direction would be to study whether specific gene signatures associated with these consortia of bacteria have a role to play in the recurrence of CRC.
 
-Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3].
-Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.
-
-Here we show that the microbiome changes towards normal after treatment for carcinoma
-
-Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g.
-*Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.
-The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models.
-This suggests that members within the commensal community may be the first that change during CRC pathogenesis.
-These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].
-
-Many of the common OTUs that we identified taxonomically classified to potential butyrate producers (e.g.
-*Clostridiales*, *Roseburia*.
-and *Anaerostipes*) [Table S2].
-Other OTUs classified to bacteria that are inhibited by polyphenols (e.g.
-*Bacteroides*).
-Both butyrate and polyphenols are thought to be protective against cancer, in part by reducing inflammation [@louis_gut_2014].
-These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.
-One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].
-High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].
-Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].
-Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.
-Within this context the commensal community may be an important modifiable risk factor for monitoring and preventing CRC recurrence after treatment.
-
-Building off of our results, an area for future research, is that within our study we do not know whether individuals who were still classified as positive by the carcinoma model eventually had a subsequent CRC recurrence.
-This information would help to strengthen the case for this model keeping numerous individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having carcinoma.
-It would also provide additional evidence that the microbiome could be used as a risk stratification tool in monitoring recurrence risk and whether different interventions could potentially change this community and lower the probability of future recurrence.
-
-Despite the stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment.
-Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of carcinoma after treatment provides evidence that there are specific signatures, attributable to both inflammatory and resident commensal organisms, associated with treatment.
-Our data provides evidence that commensal bacteria may be important in the development of polyps, potential transition of advanced adenoma to carcinoma, and recovery of the microbiome in CRC following treatment.
+Despite some of these stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable change in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment.
+Further, individuals with carcinoma had a larger change than adenoma or advanced adenoma indivdiuals and this change in community was towards those with normal colons.
+Our data provides additional evidence on the importance of the community  on recovery of the microbiome after CRC treatment and opens interesting avenues of research into how these changes may affect recurrence.
 

### 2017-04-18 submission/manuscript.tex

@@
 
-bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk
-factor.
+microbiome; colorectal cancer; polyps; treatment; risk factor.
 
@@
 groups. Because of the relatively small number of subject in each group,
-we restricted our models to only incorporate 10 OTUs. Despite this
+we restricted our models to only incorporate 10 OTUs to limit the
+likelihood that that models would overfit the data. Despite this
 restriction, the models performed well (adenoma AUC=0.69 - 0.92,
@@
 advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
-The OTUs that were most important in the models for differentiating
-between patients with normal colons and those with adenomas or advanced
-adenomas were dominated by populations commonly considered as commensals
-(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
-\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
-\emph{Roseburia}) {[}Figure S3{]}. Although many of these OTUs were also
-included in the model differentiating between patients with normal
-colons and those with carcinoma, this model also included OTUs
-affiliated with populations that have previously been associated with
-carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}) {[}Figure S3{]}. Finally, we applied these three
-models to the pre and post-treatment samples for each diagnosis group
-and quantified the change in the positive probability of the model. A
-decrease in the positive probability would indicate that the colon more
-closely resembled that of a patient with a normal colon. There was not a
+The OTUs that were incorporated into the adenoma and advanced adenoma
+models largely overlapped and those OTUs that were used to classify the
+carcinoma samples were largely distinct from those of the other two
+models {[}Figure 4A{]}. Among the OTUs that were shared across the three
+models were those populations commonly considered as commensals (e.g.
+\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
+\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
+4B{]}. Although many of these OTUs were also included in the model
+differentiating between patients with normal colons and those with
+carcinoma, this model also included OTUs affiliated with populations
+that have previously been associated with carcinoma
+(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
+{[}Figure S3{]}. Finally, we applied these three models to the pre and
+post-treatment samples for each diagnosis group and quantified the
+change in the positive probability of the model. A decrease in the
+positive probability would indicate that the colon more closely
+resembled that of a patient with a normal colon. There was not a
 significant change in the positive probability for the adenoma or
@@
 
-This study builds upon previous work from numerous labs that have
-considered both how the bacterial community between those with and
-without CRC differs and how it might be used as an early screening tool
-{[}23--27{]}. Here we show that the microbiome changes towards normal
-after treatment for carcinoma and that chemotherapy and radiation did
-not provide any additional change. Although some of the important OTUs
-classified to genera from bacteria considered the ``usual suspects''
-(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
-many did not. The majority of important OTUs had taxonomic
-classifications for resident gut microbes and were common for the
-adenoma, advanced adenoma, and carcinoma models. This suggests that
-members within the commensal community may be the first that change
-during CRC pathogenesis. These subtle changes, in turn, could be the
-first step in allowing more inflammatory bacteria to gain a foothold
-within the colon {[}13{]}.
+This study builds upon previous work that demonstrated a role for the
+microbiota in tumorigenesis and that the microbiome contains biomarkers
+to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
+comparing the microbiome of patients diagnosed with adenoma, advanced
+adenoma, and carcinoma before and after treatment. For all three groups
+of patients we observed changes in their microbiota. The microbiota of
+patients diagnosed with carcinoma changed significantly more than the
+other groups. This change resulted in communities that more closely
+resembled those of patients with normal colons.
 
-Unlike previous studies on the microbiome and CRC, ours focuses on
-comparing the microbiome during recovery from treatment in adenoma,
-advanced adenoma, and carcinoma groups. Although there were differences
-for genera associated with specific bacterium linked with CRC {[}Figure
-S3{]}, most of the important OTUs taxonomically classified to commensal
-bacteria {[}Figure S3{]}. Although these changes may be subtle, due to
-the lack of significant differences in the bacterial community pre- and
-post-treatment in adenoma and advanced adenoma {[}Figure 1{]}; they
-support the hypothesis that the first members of the community to change
-and potentially stay changed even after treatment are those that are
-commensal bacteria.
+For the carcinoma group there was a larger change between their initial
+and follow up samples then for either adenoma or advanced adenoma. This
+change was directionaly the same for each individual with carcinoma.
+This observed change has important implications for future research into
+CRC recurrence. Knowing that the community changes towards normal after
+treatment means that it is reasonable to hypothesize that the microbiome
+could be a good biomarker for stratifying individuals between those most
+and least at risk for CRC recurrence. Interestingly, our 23.08\%
+prevalence of increased carcinoma probability fits nicely within the
+range quoted for CRC recurrence (20-30\%) {[}3,4{]}.
 
-Many of the common OTUs that we identified taxonomically classified to
-potential butyrate producers (e.g. \emph{Clostridiales},
-\emph{Roseburia}. and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
-classified to bacteria that are inhibited by polyphenols (e.g.
-\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
-protective against cancer, in part by reducing inflammation {[}28{]}.
-These protective compounds are derived from the breakdown of fiber,
-fruits, and vegetables by resident gut microbes. One example of this
-potential diet-microbiome-inflammation-polyp axis is that
-\emph{Bacteroides}, which was highly prevalent in our models, are known
-to be increased in those with high non-meat based protein consumption
-{[}29{]}. High protein consumption in general has been linked with an
-increased CRC risk {[}30{]}. Conversely, \emph{Bacteroides} are
-inhibited by polyphenols which are derived from fruits and vegetables
-{[}31{]}. Our data fits with the hypothesis that the microbial
-metabolites from breakdown products within our own diet could not only
-help to shape the existing community but also have an effect on CRC risk
-and disease progression. Within this context the commensal community may
-be an important modifiable risk factor for monitoring and preventing CRC
-recurrence after treatment.
+A potential limitation, within our study, is that we do not know whether
+individuals who had an increased positive probability for carcinoma on
+follow up eventually had a subsequent CRC recurrence. This information
+would help to strengthen the case for the ability of our model to be
+used as a disease stratification tool for CRC recurrence. This would be
+an exciting area of future research that would be able to link treatment
+changes within the microbiome and CRC recurrence together.
 
-A limitation in our study was that there was a significant difference in
-the time elapsed in the collection of the post-treatment sample between
-adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
-with time passed being less for adenoma (255 ± 42 days) and advanced
-adenoma (250 ± 41) than carcinoma (351 ± 102). These results would
-indicate that some of the differences observed between the carcinoma and
-adenoma groups could be due to differences in collection time.
-Specifically, it could confound the observation that carcinomas changed
-more than adenomas {[}Figure 1{]}. However, there are two reasons that
-this may not be the case. First, the advanced adenoma group had a higher
-dissimilarity than adenoma but lower dissimilarity then carcinoma and
-the collection time to their follow up samples was less than the adenoma
-group. Second, this confounding would not affect the observations where
-models were used since they were built using a different cohort
-{[}Figures 2 \& S2-S3{]}.
+The microbiome changes associated with carcinoma were found to be
+towards those with normal colons. This would suggest that treatment for
+CRC is not only successful for removing the carcinoma but also at
+removing bacteria associated with them. This is important due to the
+numerous studies that have pointed towards a link between the community
+and tumorgenesis {[}10--12{]}. The importance of the community to this
+process is further highlighted in our data {[}Figure 2, 4, \& S3{]}.
+Interestingly, those who had an increase in their positive probability
+of carcinoma had no change in their median value for oral pathogens
+found within the carcinoma model but had a decrease in non-oral
+pathogens {[}Figure S4{]}. Although this finding was not statistically
+significant, this trend would further highlight the importance of
+community to tumorgenesis. Future studies with larger power may be able
+to confirm these reported observations.
 
-Another limitation of this study was that it drew heavily from those
-with Caucasian ancestry making it possible that the observations may not
-be representative of those with either Asian or African ancestry.
-Although our training and test set are relatively large we still run the
-risk of over-fitting or having a model that may not be representative of
-other populations. We have done our best to safeguard against this by
-not only running 10-fold cross validation but also having over 100
-different 80/20 splits to try and mimic the type of variation that might
-be expected to occur.
+Although we were not able to identify specific OTUs due to the
+heterogeneity of the population. This is a common problem that occurs in
+other fields of microbiome study {[}28{]}. This made it very difficult
+to identify individual OTUs associated with treatment and recovery
+towards a normal colon. To overcome this large individual heterogeneity
+we used RF models that made it easier to identify consortia of
+microbiota that were associated with treatment and recovery. A potential
+future direction would be to study whether specific gene signatures
+associated with these consortia of bacteria have a role to play in the
+recurrence of CRC.
 
-Building off of our results, an area for future research, is that within
-our study we do not know whether individuals who were still classified
-as positive by the carcinoma model eventually had a subsequent CRC
-recurrence. This information would help to strengthen the case for this
-model keeping numerous individuals above the cutoff threshold even
-though at follow up they were diagnosed as no longer having carcinoma.
-It would also provide additional evidence that the microbiome could be
-used as a risk stratification tool in monitoring recurrence risk and
-whether different interventions could potentially change this community
-and lower the probability of future recurrence.
-
-Despite the stated shortcomings our findings add to the existing
-scientific knowledge on CRC and the microbiome: That there is a
-measurable difference in the bacterial community after adenoma, advanced
-adenoma, or carcinoma treatment. Further, the ability for machine
-learning algorithms to take OTU data and successfully lower positive
-probability of carcinoma after treatment provides evidence that there
-are specific signatures, attributable to both inflammatory and resident
-commensal organisms, associated with treatment. Our data provides
-evidence that commensal bacteria may be important in the development of
-polyps, potential transition of advanced adenoma to carcinoma, and
-recovery of the microbiome in CRC following treatment.
+Despite some of these stated shortcomings our findings add to the
+existing scientific knowledge on CRC and the microbiome: That there is a
+measurable change in the bacterial community after adenoma, advanced
+adenoma, or carcinoma treatment. Further, individuals with carcinoma had
+a larger change than adenoma or advanced adenoma indivdiuals and this
+change in community was towards those with normal colons. Our data
+provides additional evidence on the importance of the community on
+recovery of the microbiome after CRC treatment and opens interesting
+avenues of research into how these changes may affect recurrence.
 
@@
 normalize back to a pre-colonoscopy community within this time period
-{[}32{]}. Our training cohort consisted of a total of 423 individuals
+{[}29{]}. Our training cohort consisted of a total of 423 individuals
 {[}Table 1{]}. Our study cohort consisted of 67 individuals with an
@@
 \textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
-described by Kozich, et al. {[}33{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}33{]}. The general workflow using mothur was:
+previously described {[}30{]}. The general workflow using mothur was:
 Paired-end reads were first merged into contigs, quality filtered,
@@
 
-\textbf{\emph{Model Building:}} The Random Forest {[}34{]} algorithm was
+\textbf{\emph{Model Building:}} The Random Forest {[}31{]} algorithm was
 used to create the model used to create the three models used. The
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}35{]} in the vegan package
+community structure utilized PERMANOVA {[}32{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}33{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 
-\hypertarget{ref-louis_gut_2014}{}
-28. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
-metabolites and colorectal cancer. Nature Reviews Microbiology
-{[}Internet{]}. 2014 {[}cited 2017 Feb 14{]};12:661--72. Available from:
-\url{http://www.nature.com/doifinder/10.1038/nrmicro3344}
-
-\hypertarget{ref-zhu_intake_2016}{}
-29. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
-Proteins Substantially Increased the Relative Abundance of Genus
-Lactobacillus in Rat Feces. PloS One. 2016;11:e0152678.
-
-\hypertarget{ref-mu_colonic_2016}{}
-30. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
-Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
-Compared with a Normal-Protein Diet. The Journal of Nutrition.
-2016;146:474--83.
-
-\hypertarget{ref-ozdal_reciprocal_2016}{}
-31. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
-Reciprocal Interactions between Polyphenols and Gut Microbiota and
-Effects on Bioaccessibility. Nutrients {[}Internet{]}. 2016 {[}cited
-2017 Feb 14{]};8:78. Available from:
-\url{http://www.mdpi.com/2072-6643/8/2/78}
+\hypertarget{ref-sze_looking_2016}{}
+28. Sze MA, Schloss PD. Looking for a Signal in the Noise: Revisiting
+Obesity and the Microbiome. mBio. 2016;7.
 
 \hypertarget{ref-obrien_impact_2013}{}
-32. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-33. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+30. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-34. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+31. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+32. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+33. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-04-19 submission/manuscript.tex
small extra corrections to discussion

@@
 of carcinoma had no change in their median value for oral pathogens
-found within the carcinoma model but had a decrease in non-oral
-pathogens {[}Figure S4{]}. Although this finding was not statistically
-significant, this trend would further highlight the importance of
-community to tumorgenesis. Future studies with larger power may be able
-to confirm these reported observations.
+within the carcinoma model but had a decrease in bacteria that were
+higher in carcinoma {[}Figure S4{]}. Although this finding was not
+statistically significant, this trend further highlights the importance
+of community to tumorgenesis. Future studies with larger power may be
+able to confirm these reported observations.
 
@@
 heterogeneity of the population. This is a common problem that occurs in
-other fields of microbiome study {[}28{]}. This made it very difficult
+other fields of microbiome study {[}28{]}. This makes it very difficult
 to identify individual OTUs associated with treatment and recovery
@@
 we used RF models that made it easier to identify consortia of
-microbiota that were associated with treatment and recovery. A potential
-future direction would be to study whether specific gene signatures
-associated with these consortia of bacteria have a role to play in the
-recurrence of CRC.
+microbiota that were associated with treatment and recovery. One
+potential future direction would be to study whether specific gene
+signatures associated with these consortia of bacteria are directly
+involved with the pathogenesis of CRC recurrence.
 
@@
 measurable change in the bacterial community after adenoma, advanced
-adenoma, or carcinoma treatment. Further, individuals with carcinoma had
-a larger change than adenoma or advanced adenoma indivdiuals and this
-change in community was towards those with normal colons. Our data
-provides additional evidence on the importance of the community on
-recovery of the microbiome after CRC treatment and opens interesting
-avenues of research into how these changes may affect recurrence.
+adenoma, or carcinoma treatment. Individuals with carcinoma had a larger
+change than adenoma or advanced adenoma individuals and this change in
+community was towards those with normal colons. Our data provides
+additional evidence on the importance of the community on recovery of
+the microbiome after CRC treatment and opens interesting avenues of
+research into how these changes may affect recurrence.
 

### 2017-04-19 submission/manuscript.Rmd

@@
 The importance of the community to this process is further highlighted in our data [Figure 2, 4, & S3].
-Interestingly, those who had an increase in their positive probability of carcinoma had no change in their median value for oral pathogens found within the carcinoma model but had a decrease in non-oral pathogens [Figure S4].
-Although this finding was not statistically significant, this trend would further highlight the importance of community to tumorgenesis.
+Interestingly, those who had an increase in their positive probability of carcinoma had no change in their median value for oral pathogens within the carcinoma model but had a decrease in bacteria that were higher in carcinoma [Figure S4].
+Although this finding was not statistically significant, this trend further highlights the importance of community to tumorgenesis.
 Future studies with larger power may be able to confirm these reported observations.
@@
 This is a common problem that occurs in other fields of microbiome study [@sze_looking_2016].
-This made it very difficult to identify individual OTUs associated with treatment and recovery towards a normal colon.
+This makes it very difficult to identify individual OTUs associated with treatment and recovery towards a normal colon.
 To overcome this large individual heterogeneity we used RF models that made it easier to identify consortia of microbiota that were associated with treatment and recovery.
-A potential future direction would be to study whether specific gene signatures associated with these consortia of bacteria have a role to play in the recurrence of CRC.
+One potential future direction would be to study whether specific gene signatures associated with these consortia of bacteria are directly involved with the pathogenesis of CRC recurrence.
 
 Despite some of these stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable change in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment.
-Further, individuals with carcinoma had a larger change than adenoma or advanced adenoma indivdiuals and this change in community was towards those with normal colons.
+Individuals with carcinoma had a larger change than adenoma or advanced adenoma individuals and this change in community was towards those with normal colons.
 Our data provides additional evidence on the importance of the community  on recovery of the microbiome after CRC treatment and opens interesting avenues of research into how these changes may affect recurrence.

### 2017-04-21 submission/manuscript.Rmd

--- 2017-04-18 submission/manuscript.Rmd
@@
 
-# The Fecal Microbiome Before and After Treatment for Colorectal Adenoma or Carcinoma
 ```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
-# initial microbiome changes
+# initial microbiota changes
 change_summary <- read.csv('../data/process/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
@@
 ```
+\begin{center}
+
+# The effect of treatment on the microbiota of patients diagnosed with colonic lesions
+
 \vspace{25mm}
-
-\begin{center}
-Running Title: Human Microbiome before and after Colorectal Cancer
 
@@
 
-$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, M.
+$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI
 
-$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, P.
+$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
 
@@
 **Background.** Colorectal cancer (CRC) is a worldwide health problem.
-Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiome after treatment for an adenoma or carcinoma.
+Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiota after treatment for an adenoma or carcinoma.
 This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with with disease to those associated with a normal colon.
@@
 
-**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiome will yield improvements in disease management.
+**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiota will yield improvements in disease management.
 
@@
 ***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1).
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1).
 The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A].
@@
 
-This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
-Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
+This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiota contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
+Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
 For all three groups of patients we observed changes in their microbiota.
 The microbiota of patients diagnosed with carcinoma changed significantly more than the other groups.
-This change resulted in communities that more closely resembled those of patients with normal colons.
+This change resulted in communities that more closely resembled those of patients with a normal colon.
+For the patients diagnosed with carcinomas there was a larger change between their initial and follow up samples than was observed for patients diagnosed with either adenoma or advanced adenoma.
+The changes in the community structure of those diagnosed with carcinoma were similar and resulted in communities that more closely resembled those found in patients with normal colons.
 
-For the carcinoma group there was a larger change between their initial and follow up samples then for either adenoma or advanced adenoma.
-This change was directionaly the same for each individual with carcinoma.
-This observed change has  important implications for future research into CRC recurrence.
-Knowing that the community changes towards normal after treatment means that it is reasonable to hypothesize that the microbiome could be a good biomarker for stratifying individuals between those most and least at risk for CRC recurrence.
-Interestingly, our `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability fits nicely within the range quoted for CRC recurrence (20-30%) [@hellinger_reoperation_2006; @ryuk_predictive_2014].
+Understanding the effect of treatment on the microbiota of those initially diagnosed with carcinomas may have important implications into understanding disease recurrence.
+The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted community structures that resembled patients with normal colons.
+This would suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at removing bacteria associated with them.
+Interestingly, 6 of the 26 patients diagnosed with carcinomas had an elevated probability of having carcinomas after treatment.
+This leads us to hypothesize that these individuals may be at a high risk of recurrence.
+Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30%), which would support our hypothesis [@hellinger_reoperation_2006; @ryuk_predictive_2014].
+Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence.
+Regardless, it is intriguing to think of using biomarkers from the microbiota to not only predict the presence of lesions, but to also assess the risk of developing them.
 
-A potential limitation, within our study, is that we do not know whether individuals who had an increased positive probability for carcinoma on follow up eventually had a subsequent CRC recurrence.
-This information would help to strengthen the case for the ability of our model to be used as a disease stratification tool for CRC recurrence.
-This would be an exciting area of future research that would be able to link treatment changes within the microbiome and CRC recurrence together.
+It is interesting that those patients diagnosed with adenoma and advanced adenoma did not experience a shift towards a community structure that resembled those with normal colons.
+This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma.
+Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter the structure of the community.
+In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
+This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas (Figure S2).
+Given the patchy distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
+A final hypothesis is that the specific type of treatment altered the structure of the microbiome.
+The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation.
+Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turn over of the colonic epithelium driving a more significant change in the structure of the microbiota.
+Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.
 
-The microbiome changes associated with carcinoma were found to be towards those with normal colons.
-This would suggest that treatment for CRC is not only successful for removing the carcinoma but also at removing bacteria associated with them.
-This is important due to the numerous studies that have pointed towards a link between the community and tumorgenesis [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
-The importance of the community to this process is further highlighted in our data [Figure 2, 4, & S3].
-Interestingly, those who had an increase in their positive probability of carcinoma had no change in their median value for oral pathogens found within the carcinoma model but had a decrease in non-oral pathogens [Figure S4].
-Although this finding was not statistically significant, this trend would further highlight the importance of community to tumorgenesis.
-Future studies with larger power may be able to confirm these reported observations.
+It was not clear why 6 of the 26 patients with carcinoma had a higher probability of having carcinomas after treatment.
+We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these 6 individuals (3 with Stage II and 3 with Stage III) was similar to the distribution observed among the other 20 patients.
+We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased or decreased (Wilcoxon Test; P-value=0.56).
+Finally, it is possible that these patients may not have responded to treatment as well as the other 20 patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
+Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.
 
-Although we were not able to identify specific OTUs due to the heterogeneity of the population.
-This is a common problem that occurs in other fields of microbiome study [@sze_looking_2016].
-This made it very difficult to identify individual OTUs associated with treatment and recovery towards a normal colon.
-To overcome this large individual heterogeneity we used RF models that made it easier to identify consortia of microbiota that were associated with treatment and recovery.
-A potential future direction would be to study whether specific gene signatures associated with these consortia of bacteria have a role to play in the recurrence of CRC.
-
-Despite some of these stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable change in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment.
-Further, individuals with carcinoma had a larger change than adenoma or advanced adenoma indivdiuals and this change in community was towards those with normal colons.
-Our data provides additional evidence on the importance of the community  on recovery of the microbiome after CRC treatment and opens interesting avenues of research into how these changes may affect recurrence.
+This study builds upon existing work that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
+The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those initially diagnosed with carcinoma.
+If such an approach is effective, it might be possible to even target the microbiota as part of combined therapies including currently standard approaches.
+Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.
 
@@
 
-***Study Design and Patient Sampling:***
+***Study Design and Patient Sampling.***
 Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016].
 Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
-Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.
+Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy.
 The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013].
-Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].
 Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 2].
+Patients were diagnosed by colonoscopic examination and histopathological review of biopsies.
+Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology.
 This study was approved by the University of Michigan Institutional Review Board.
@@
 
-***16S rRNA Gene Sequencing:***
+***16S rRNA Gene Sequencing.***
 Sequencing was completed as described by Kozich, et al.
@@
 
-***Sequence Processing:***
+***Sequence Processing.***
 The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013].
-The general workflow using mothur was: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.
-The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize uneven sampling.
+The general workflow using mothur included merging paired-end reads into contigs, filtering for low quality contigs, aligning to the SILVA database [reference], screening for chimeras using UCHIME [reference], classifying with a naive Bayesian classifier using the Ribosomal Database Project (RDP)[reference-Wang], and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm [reference-Westcott&SchlossAEM].
+The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize the impacts of uneven sampling.
 
-***Model Building:***
+***Model Building.***
+
+Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].
+
 The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used.
@@
 Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
-For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`
-
-The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.
+For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
 The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.
 
-***Statistical Analysis:***
+***Statistical Analysis.***
 The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.
 Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
-Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxson ranked sum test.
+Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxon ranked sum test.
 Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
 Unless otherwise stated the P-values reported are those that were BH corrected.
-Model rank importance was determined by obtining the median MDA from the 100 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.
+Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.
 
-***Analysis Overview:***
-We first tested whether there were any differences between pre- and post-treatment samples in alpha and beta diversity based on adenoma, SRN, or carcinoma.
-We then tested all OTUs for differences between pre- and post-treatment samples.
-We next used our specific models for adenoma, SRN, and carcinoma to test classification accuracy, response towards a normal microbiome, and common OTUs used across models.
-Finally, for the adenoma group differences between those that received surgery or not was tested while for the carcinoma group differences between those receiving chemotherapy and radiation was tested.
-
-***Reproducible Methods:***
+***Reproducible Methods.***
 A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
@@
 
-**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.*.
-A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.*.
-A) MDA of the most important variables in the adenoma model.
+**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.

### 2017-04-21 code/common_otu_graph.R
switch graphs

@@ -67,5 +67,5 @@ test_graph <- ggplot(combined_table,
 combined_graph <- grid.arrange(venn_graph, test_graph)
 
 
-ggsave(file = "results/figures/Figure4.pdf", combined_graph, 
+ggsave(file = "results/figures/Figure3.pdf", combined_graph, 
        width=6, height = 9, dpi = 300)

### 2017-04-21 submission/manuscript.Rmd
re-render good version

@@
 ```
-\begin{center}
 
@@
 
+\begin{center}
 \vspace{25mm}
-
-\vspace{10mm}
 
@@
 Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiota after treatment for an adenoma or carcinoma.
-This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with with disease to those associated with a normal colon.
+This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
 We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), and carcinoma (N=`r change_summary["thetayc", "crc_n"]`).
 
-**Results.** There were large changes the the bacterial communities associated with treatment across the three groups.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value>0.05).
+**Results.** There were large changes to the bacterial communities associated with treatment across the three groups.
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05).
 Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 Although we could clearly differentiate pre- and post-treatment communities from the three diagnosis groups, only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a carcinoma (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-Finally, tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with chemotherapy and those treated with chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
@@
 
-microbiome; colorectal cancer; polyps; treatment; risk factor.
+microbiota; colorectal cancer; polyps; treatment; risk factor.
 
@@
 Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].
-These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions.
+These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions (adenoma, advanced adenoma, and carcinoma).
 They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions.
@@
 Finally, we measured the predictive probability of diagnosis models to determine whether treatment would alter the probability of being diagnosed as having a normal colon or the colon of the original diagnosis group.
-Ultimately, we found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas.
+We found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas.
 Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence.
@@
 ***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1).
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples [Table 1].
 The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A].
-We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001).
-There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P>0.05).
-Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1BCD].
+We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`).
+There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
+Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
 We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`).
-Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value>0.05) [Table S1].
+Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].
 
@@
 To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data.
+Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data.
 Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`).
@@
 
-***Post-treatment samples from patients with carcinoma share***
+***Post-treatment samples from patients with carcinoma change towards a microbiota associated with normal***
 Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons.
-To test this, we used an expanded cohort of 423 individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
+To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2].
 We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon.
 The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2).
-The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 4A].
+The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
 Among the OTUs that were shared across the three models were those populations commonly considered as commensals (e.g.
-*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 4B].
+*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
 Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3].
@@
 A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon.
-There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 3].
+There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4].
 The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased.
@@
 Understanding the effect of treatment on the microbiota of those initially diagnosed with carcinomas may have important implications into understanding disease recurrence.
-The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted community structures that resembled patients with normal colons.
+The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted in community structures that resembled patients with normal colons.
 This would suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at removing bacteria associated with them.
@@
 This leads us to hypothesize that these individuals may be at a high risk of recurrence.
-Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30%), which would support our hypothesis [@hellinger_reoperation_2006; @ryuk_predictive_2014].
+Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]), which would support our hypothesis .
 Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence.
@@
 This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma.
-Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter the structure of the community.
+Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter said structure.
 In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
-This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas (Figure S2).
+This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S2].
 Given the patchy distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
 A final hypothesis is that the specific type of treatment altered the structure of the microbiome.
-The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation.
+The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma.
 Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turn over of the colonic epithelium driving a more significant change in the structure of the microbiota.
@@
 The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013].
-Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 2].
+Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1].
 Patients were diagnosed by colonoscopic examination and histopathological review of biopsies.
@@
 ***Model Building.***
-
-Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].
-
 The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used.
 The adenoma model classified normal versus adenoma, advanced adenoma was normal versus advanced adenoma, and carcinoma was normal versus carcinoma.
-The toal number of individuals in this data set was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.
+The toal number of individuals in this data set was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 1].
 There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` indivdiuals in the carcinoma model [Table 1].
@@
 
-**Figure 2: Treatment Response Based on Models Built for Adenoma, SRN, or Carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
-B) Positive probability change from initial to follow up sample in those with SRN.
+**Figure 2: The 10 OTUs used to classify treatment for Adenoma, Advanced Adenoma, and Carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs.
+
+**Figure 3: Treatment Response Based on Models Built for Adenoma, Advanced Adenoma, or Carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
+B) Positive probability change from initial to follow up sample in those with advanced adenoma.
 C) Positive probability change from initial to follow up sample in those with carcinoma..
 
-**Figure 3: Common OTUs Model Rank Based on Median MDA.** For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
+**Figure 4: Common OTUs to All Models.** A) Venn diagram showing the OTU overlap between each model.
+B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
 
@@
 
-**Table 1: Demographic Data of Training Cohort**
+**Table 1: Demographic Data of Pre and Post Treatment Cohort**
 
-**Table 2: Demographic Data of Pre and Post Treatment Cohort**
+**Table 2: Demographic Data of Training Cohort**
 
@@
 
-**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model.
+**Figure S3: Summary of Important OTUs for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.

### 2017-04-21 submission/manuscript.tex
re-render good version

--- 2017-04-18 submission/manuscript.tex
@@
 
-\section{The Fecal Microbiome Before and After Treatment for Colorectal
-Adenoma or
-Carcinoma}\label{the-fecal-microbiome-before-and-after-treatment-for-colorectal-adenoma-or-carcinoma}
-
-\vspace{25mm}
+\section{The effect of treatment on the microbiota of patients diagnosed
+with colonic
+lesions}\label{the-effect-of-treatment-on-the-microbiota-of-patients-diagnosed-with-colonic-lesions}
 
 \begin{center}
-Running Title: Human Microbiome before and after Colorectal Cancer
-
-\vspace{10mm}
+\vspace{25mm}
 
@@
 drive tumorigenesis, little is known about what happens to the
-microbiome after treatment for an adenoma or carcinoma. This study
+microbiota after treatment for an adenoma or carcinoma. This study
 tested the hypothesis that treatment for adenoma or carcinoma alters the
-abundance of bacterial populations associated with with disease to those
+abundance of bacterial populations associated with disease to those
 associated with a normal colon. We tested this hypothesis by sequencing
@@
 
-\textbf{Results.} There were large changes the the bacterial communities
+\textbf{Results.} There were large changes to the bacterial communities
 associated with treatment across the three groups. The communities from
@@
 significant difference between those with advanced adenoma and those
-with adenoma or carcinoma (P-value\textgreater{}0.05). Although
+with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
 treatment brought about large intrapersonal changes, the change in the
@@
 experienced a significant decrease in positive probability of having a
-carcinoma (P-value \textless{} 0.05). Finally, tested whether the type
-of treatment impacted the microbiota of those diagnosed with carcinomas
-and were unable to detect any significant differences in characteristics
-of these communities between individuals treated with chemotherapy and
-those treated with chemotherapy and radiation (P-value \textgreater{}
-0.05).
+carcinoma (P-value \textless{} 0.05). Finally, we tested whether the
+type of treatment impacted the microbiota of those diagnosed with
+carcinomas and were unable to detect any significant differences in
+characteristics of these communities between individuals treated with
+surgery alone and those treated with chemotherapy or chemotherapy and
+radiation (P-value \textgreater{} 0.05).
 
 \textbf{Conclusions.} Further exploration of the relationship between
-diagnosis, treatment, and the impact on the microbiome will yield
+diagnosis, treatment, and the impact on the microbiota will yield
 improvements in disease management.
@@
 have associated \emph{Fusobacterium nucleatum} and other oral pathogens
-with lesions. They have also noted that the loss of bacteria generally
-thought to produce short chain fatty acids, which can suppress
-inflammation, is associated with colonic lesions. Gut bacteria have a
-role in tumorigenesis and there is promise that these populations may be
-useful biomarkers for aiding in the early detection of disease.
+with lesions (adenoma, advanced adenoma, and carcinoma). They have also
+noted that the loss of bacteria generally thought to produce short chain
+fatty acids, which can suppress inflammation, is associated with colonic
+lesions. Gut bacteria have a role in tumorigenesis and there is promise
+that these populations may be useful biomarkers for aiding in the early
+detection of disease.
 
@@
 alter the probability of being diagnosed as having a normal colon or the
-colon of the original diagnosis group. Ultimately, we found that
-treatment alters the composition of the gut microbiota; however, we only
-observed a shift from diseased to normal colon for those initially
-diagnosed with carcinomas. Understanding how the community responds to
-these treatments could be a valuable tool for identifying biomarkers to
-quantify the risk of recurrence.
+colon of the original diagnosis group. We found that treatment alters
+the composition of the gut microbiota; however, we only observed a shift
+from diseased to normal colon for those initially diagnosed with
+carcinomas. Understanding how the community responds to these treatments
+could be a valuable tool for identifying biomarkers to quantify the risk
+of recurrence.
 
@@
 patients diagnosed with colonic lesions.}} Within our 67-person cohort
-we tested whether the microbiome of patients with adenoma (N=22),
+we tested whether the microbiota of patients with adenoma (N=22),
 advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
-between pre- and post-treatment samples (Table 1). The structure of the
-microbial communities of the pre- and post-treatment samples differed
-{[}Figure 1A{]}. We found that the communities obtained pre- and
-post-treament among the patients with carcinomas changed significantly
-more than those patients with adenoma (P-value \textless{} 0.001). There
-were no significant differences in the amount of change observed between
-the patients with adenoma and advanced adenoma or between the patients
-with advanced adenoma and carcinoma (P\textgreater{}0.05). Next, we
-tested whether there was a consistent direction in the change in the
-community structure between the pre and post-treatment samples for each
-of the diagnosis groups {[}Figure 1BCD{]}. We only observed a consistent
-shift in community structure for the patients with carcinoma when using
-a PERMANOVA test (adenoma P-value0.999, advanced adenoma P-value0.945,
-and carcinoma P-value0.005). Finally, we measured the number of observed
-OTUs, Shannon evenness, and Shannon diversity in the pre and
-post-treatment samples and did not observe a significant change for any
-of the diagnosis groups (P-value\textgreater{}0.05) {[}Table S1{]}.
+between pre- and post-treatment samples {[}Table 1{]}. The structure of
+the microbial communities of the pre- and post-treatment samples
+differed {[}Figure 1A{]}. We found that the communities obtained pre-
+and post-treament among the patients with carcinomas changed
+significantly more than those patients with adenoma (P-value = 5.4e-05).
+There were no significant differences in the amount of change observed
+between the patients with adenoma and advanced adenoma or between the
+patients with advanced adenoma and carcinoma (P-value \textgreater{}
+0.05). Next, we tested whether there was a consistent direction in the
+change in the community structure between the pre and post-treatment
+samples for each of the diagnosis groups {[}Figure 1B-D{]}. We only
+observed a consistent shift in community structure for the patients with
+carcinoma when using a PERMANOVA test (adenoma P-value0.999, advanced
+adenoma P-value0.945, and carcinoma P-value0.005). Finally, we measured
+the number of observed OTUs, Shannon evenness, and Shannon diversity in
+the pre and post-treatment samples and did not observe a significant
+change for any of the diagnosis groups (P-value \textgreater{} 0.05)
+{[}Table S1{]}.
 
@@
 between pre and post-treatment samples from each of the diagnosis
-groups. Because of the relatively small number of subject in each group,
-we restricted our models to only incorporate 10 OTUs to limit the
+groups. Because of the relatively small number of subjects in each
+group, we restricted our models to only incorporate 10 OTUs to limit the
 likelihood that that models would overfit the data. Despite this
@@
 
-\textbf{\emph{Post-treatment samples from patients with carcinoma
-share}} Next, we determined whether treatment changed the microbiota in
-a way that the post-treatment communities resembled that of patients
-with normal colons. To test this, we used an expanded cohort of 423
-individuals that were diagnosed under the same protocol as having normal
-colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
-We then constructed Random Forest models to classify the pre and
-post-treatment samples as having their original diagnosis or having a
-normal colon. The models performed well (adenoma AUC=0.62 - 0.72,
-advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
-The OTUs that were incorporated into the adenoma and advanced adenoma
-models largely overlapped and those OTUs that were used to classify the
-carcinoma samples were largely distinct from those of the other two
-models {[}Figure 4A{]}. Among the OTUs that were shared across the three
-models were those populations commonly considered as commensals (e.g.
+\textbf{\emph{Post-treatment samples from patients with carcinoma change
+towards a microbiota associated with normal}} Next, we determined
+whether treatment changed the microbiota in a way that the
+post-treatment communities resembled that of patients with normal
+colons. To test this, we used an expanded cohort of 423 individuals that
+were diagnosed under the same protocol as having normal colons or colons
+with adenoma, advanced adenoma, or carcinoma {[}Table 2{]}. We then
+constructed Random Forest models to classify the pre and post-treatment
+samples as having their original diagnosis or having a normal colon. The
+models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
+AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2). The OTUs that
+were incorporated into the adenoma and advanced adenoma models largely
+overlapped and those OTUs that were used to classify the carcinoma
+samples were largely distinct from those of the other two models
+{[}Figure 3A{]}. Among the OTUs that were shared across the three models
+were those populations commonly considered as commensals (e.g.
 \emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
 \emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
-4B{]}. Although many of these OTUs were also included in the model
+3B{]}. Although many of these OTUs were also included in the model
 differentiating between patients with normal colons and those with
@@
 positive probability would indicate that the colon more closely
-resembled that of a patient with a normal colon. There was not a
+resembled that of a patient with a normal colon. There was no
 significant change in the positive probability for the adenoma or
-advanced adenoma groups {[}Figure 3{]}. The positive probability for the
+advanced adenoma groups {[}Figure 4{]}. The positive probability for the
 pre and post-treatment samples from patients diagnosed with carcinoma
@@
 This study builds upon previous work that demonstrated a role for the
-microbiota in tumorigenesis and that the microbiome contains biomarkers
+microbiota in tumorigenesis and that the microbiota contains biomarkers
 to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
-comparing the microbiome of patients diagnosed with adenoma, advanced
+comparing the microbiota of patients diagnosed with adenoma, advanced
 adenoma, and carcinoma before and after treatment. For all three groups
@@
 other groups. This change resulted in communities that more closely
-resembled those of patients with normal colons.
+resembled those of patients with a normal colon. For the patients
+diagnosed with carcinomas there was a larger change between their
+initial and follow up samples than was observed for patients diagnosed
+with either adenoma or advanced adenoma. The changes in the community
+structure of those diagnosed with carcinoma were similar and resulted in
+communities that more closely resembled those found in patients with
+normal colons.
 
-For the carcinoma group there was a larger change between their initial
-and follow up samples then for either adenoma or advanced adenoma. This
-change was directionaly the same for each individual with carcinoma.
-This observed change has important implications for future research into
-CRC recurrence. Knowing that the community changes towards normal after
-treatment means that it is reasonable to hypothesize that the microbiome
-could be a good biomarker for stratifying individuals between those most
-and least at risk for CRC recurrence. Interestingly, our 23.08\%
-prevalence of increased carcinoma probability fits nicely within the
-range quoted for CRC recurrence (20-30\%) {[}3,4{]}.
+Understanding the effect of treatment on the microbiota of those
+initially diagnosed with carcinomas may have important implications into
+understanding disease recurrence. The microbiota changes associated with
+treatment among patients initially diagnosed with carcinomas resulted in
+community structures that resembled patients with normal colons. This
+would suggest that treatment for carcinoma is not only successful for
+removing the carcinoma but also at removing bacteria associated with
+them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had
+an elevated probability of having carcinomas after treatment. This leads
+us to hypothesize that these individuals may be at a high risk of
+recurrence. Interestingly, the 23.08\% prevalence of increased carcinoma
+probability from our study is within the expected rate of recurrence
+(20-30\% {[}3,4{]}), which would support our hypothesis . Unfortunately,
+because of the design of our study, it was not possible to determine
+whether individuals who had an increased positive probability for
+carcinoma experienced recurrence. Regardless, it is intriguing to think
+of using biomarkers from the microbiota to not only predict the presence
+of lesions, but to also assess the risk of developing them.
 
-A potential limitation, within our study, is that we do not know whether
-individuals who had an increased positive probability for carcinoma on
-follow up eventually had a subsequent CRC recurrence. This information
-would help to strengthen the case for the ability of our model to be
-used as a disease stratification tool for CRC recurrence. This would be
-an exciting area of future research that would be able to link treatment
-changes within the microbiome and CRC recurrence together.
+It is interesting that those patients diagnosed with adenoma and
+advanced adenoma did not experience a shift towards a community
+structure that resembled those with normal colons. This may be due to
+the fundamental differences between the features of adenomas and
+advanced adenomas and carcinoma. Specifically, carcinomas may create an
+inflammatory milieu that would impact the structure of the community and
+removal of that stimulus would alter said structure. In addition, it is
+possible that the difference between the microbiota of patients with
+adenoma and advanced adenoma and those with normal colons is subtle.
+This is supported by the reduced ability of our models to correctly
+classify patients with adenomas and advanced adenomas relative to those
+diagnosed with carcinomas {[}Figure S2{]}. Given the patchy distribution
+of microbiota across patients in the different diagnosis groups, it is
+possible that we lacked the statistical power to adequately characterize
+the change in the communities following treatment. A final hypothesis is
+that the specific type of treatment altered the structure of the
+microbiome. The treatment to remove adenomas and advanced adenomas was
+either polyp removal or surgical resection whereas it was surgical
+resection alone or in combination with chemotherapy or with chemotherapy
+and radiation for individuals with carcinoma. Because chemotherapy and
+radiation target rapidly growing cells, these treatments would be more
+likely to cause a turn over of the colonic epithelium driving a more
+significant change in the structure of the microbiota. Although, we were
+able to test for an effect across these specific types of treatment, the
+number of patients in each treatment group was relatively small.
 
-The microbiome changes associated with carcinoma were found to be
-towards those with normal colons. This would suggest that treatment for
-CRC is not only successful for removing the carcinoma but also at
-removing bacteria associated with them. This is important due to the
-numerous studies that have pointed towards a link between the community
-and tumorgenesis {[}10--12{]}. The importance of the community to this
-process is further highlighted in our data {[}Figure 2, 4, \& S3{]}.
-Interestingly, those who had an increase in their positive probability
-of carcinoma had no change in their median value for oral pathogens
-found within the carcinoma model but had a decrease in non-oral
-pathogens {[}Figure S4{]}. Although this finding was not statistically
-significant, this trend would further highlight the importance of
-community to tumorgenesis. Future studies with larger power may be able
-to confirm these reported observations.
+It was not clear why 6 of the 26 patients with carcinoma had a higher
+probability of having carcinomas after treatment. We hypothesized that
+these individuals may have had more severe tumors; however, the tumor
+severity of these 6 individuals (3 with Stage II and 3 with Stage III)
+was similar to the distribution observed among the other 20 patients. We
+also hypothesized that we may have sampled these patients later than the
+rest and their communities may have reverted to a carcinoma-associated
+state; however, there was not a statistically significant difference in
+the length of time between sample collection among those whose
+probabilities increased or decreased (Wilcoxon Test; P-value=0.56).
+Finally, it is possible that these patients may not have responded to
+treatment as well as the other 20 patients diagnosed with carcinoma and
+so the microbiota may not have been impacted the same way. Again,
+further studies looking at the role of the microbiota in recurrence are
+needed to understand the dynamics following treatment.
 
-Although we were not able to identify specific OTUs due to the
-heterogeneity of the population. This is a common problem that occurs in
-other fields of microbiome study {[}28{]}. This made it very difficult
-to identify individual OTUs associated with treatment and recovery
-towards a normal colon. To overcome this large individual heterogeneity
-we used RF models that made it easier to identify consortia of
-microbiota that were associated with treatment and recovery. A potential
-future direction would be to study whether specific gene signatures
-associated with these consortia of bacteria have a role to play in the
-recurrence of CRC.
-
-Despite some of these stated shortcomings our findings add to the
-existing scientific knowledge on CRC and the microbiome: That there is a
-measurable change in the bacterial community after adenoma, advanced
-adenoma, or carcinoma treatment. Further, individuals with carcinoma had
-a larger change than adenoma or advanced adenoma indivdiuals and this
-change in community was towards those with normal colons. Our data
-provides additional evidence on the importance of the community on
-recovery of the microbiome after CRC treatment and opens interesting
-avenues of research into how these changes may affect recurrence.
+This study builds upon existing work that has established a role for the
+microbiota in tumorigenesis and that demonstrated the utility of
+microbiome-based biomarkers to predict the presence of colonic lesions.
+The most exciting future direction from the current study is the
+possibility that markers within the microbiota could be used to evaluate
+the effect of treatment and predict recurrence for those initially
+diagnosed with carcinoma. If such an approach is effective, it might be
+possible to even target the microbiota as part of combined therapies
+including currently standard approaches. Our data provides additional
+evidence on the importance of the microbiota in tumorigenesis by
+addressing the recovery of the microbiota after treatment and opens
+interesting avenues of research into how these changes may affect
+recurrence.
 
@@
 
-\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
+\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
 have been previously reported in Baxter, et al {[}23{]}. Briefly, study
@@
 models for prediction were collected either prior to a colonoscopy or
-between 1 - 2 weeks after. The bacterial community has been shown to
-normalize back to a pre-colonoscopy community within this time period
-{[}29{]}. Our training cohort consisted of a total of 423 individuals
-{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
-initial sample as described and a follow up sample obtained between 188
-- 546 days after treatment of lesion {[}Table 2{]}. This study was
-approved by the University of Michigan Institutional Review Board. All
-study participants provided informed consent and the study itself
-conformed to the guidelines set out by the Helsinki Declaration.
+between one and two weeks after initial colonoscopy. The bacterial
+community has been shown to normalize back to a pre-colonoscopy
+community within this time period {[}28{]}. Our study cohort consisted
+of 67 individuals with an initial sample as described and a follow up
+sample obtained between 188 - 546 days after treatment of lesion
+{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
+histopathological review of biopsies. Patients were classified as having
+advanced adenoma if they had an adenoma greater than 1 cm, more than
+three adenomas of any size, or an adenoma with villous histology. This
+study was approved by the University of Michigan Institutional Review
+Board. All study participants provided informed consent and the study
+itself conformed to the guidelines set out by the Helsinki Declaration.
 
-\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
-described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
+\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
+described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 
-\textbf{\emph{Sequence Processing:}} The mothur software package
+\textbf{\emph{Sequence Processing.}} The mothur software package
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}30{]}. The general workflow using mothur was:
-Paired-end reads were first merged into contigs, quality filtered,
-aligned to the SILVA database, screened for chimeras, classified with a
-naive Bayesian classifier using the Ribosomal Database Project (RDP),
-and clustered into Operational Taxonomic Units (OTUs) using a 97\%
-similarity cutoff with an average neighbor clustering algorithm. The
-number of sequences for each sample was rarefied to 10523 to minimize
-uneven sampling.
+previously described {[}29{]}. The general workflow using mothur
+included merging paired-end reads into contigs, filtering for low
+quality contigs, aligning to the SILVA database {[}reference{]},
+screening for chimeras using UCHIME {[}reference{]}, classifying with a
+naive Bayesian classifier using the Ribosomal Database Project
+(RDP){[}reference-Wang{]}, and clustered into Operational Taxonomic
+Units (OTUs) using a 97\% similarity cutoff with an average neighbor
+clustering algorithm {[}reference-Westcott\&SchlossAEM{]}. The number of
+sequences for each sample was rarefied to 10523 to minimize the impacts
+of uneven sampling.
 
-\textbf{\emph{Model Building:}} The Random Forest {[}31{]} algorithm was
+\textbf{\emph{Model Building.}} The Random Forest {[}30{]} algorithm was
 used to create the model used to create the three models used. The
@@
 carcinoma. The toal number of individuals in this data set was 423
-individuals. There were a total of 239 individuals in the adenoma model,
-262 individuals in the advanced adenoma model, and 266 indivdiuals in
-the carcinoma model {[}Table 1{]}. Each model was then applied to our
-67-person cohort testing prediction of adenoma intial (adenoma n = 22)
-versus adenoma follow up (adenoma n = 0), advanced adenoma initial
-(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
-adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
-follow up (carcinoma n = 1).
+individuals {[}Table 1{]}. There were a total of 239 individuals in the
+adenoma model, 262 individuals in the advanced adenoma model, and 266
+indivdiuals in the carcinoma model {[}Table 1{]}. Each model was then
+applied to our 67-person cohort testing prediction of adenoma intial
+(adenoma n = 22) versus adenoma follow up (adenoma n = 0), advanced
+adenoma initial (advanced adenoma n = 19) versus advanced adenoma follow
+up (advanced adenoma n = 0), carcinoma initial (carinoma n = 26) versus
+carcinoma follow up (carcinoma n = 1).
 
@@
 the 67-person cohort was completed. For all three models the final
-optimized mtry was 2
+optimized mtry was 2. The hyper-parameter, mtry, defines the number of
+variables to investigate at each split before a new division of the data
+was created with the Random Forest model.
 
-The default cutoff of 0.5 was used as the threshold to classify
-individuals as positive or negative for lesion. The hyper-parameter,
-mtry, defines the number of variables to investigate at each split
-before a new division of the data was created with the Random Forest
-model.
-
-\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
+\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}32{]} in the vegan package
+community structure utilized PERMANOVA {[}31{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
 differences between pre- and post-treatment samples utilized a paired
-Wilcoxson ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}33{]}
+Wilcoxon ranked sum test. Where multiple comparison testing was
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}32{]}
 and a corrected P-value of less than 0.05 was considered significant.
 Unless otherwise stated the P-values reported are those that were BH
-corrected. Model rank importance was determined by obtining the median
-MDA from the 100 20 repeated 10-fold cross validation and then ranking
+corrected. Model rank importance was determined by obtaining the median
+MDA from the 100, 20 repeated 10-fold cross validation and then ranking
 from largest to smallest MDA.
 
-\textbf{\emph{Analysis Overview:}} We first tested whether there were
-any differences between pre- and post-treatment samples in alpha and
-beta diversity based on adenoma, SRN, or carcinoma. We then tested all
-OTUs for differences between pre- and post-treatment samples. We next
-used our specific models for adenoma, SRN, and carcinoma to test
-classification accuracy, response towards a normal microbiome, and
-common OTUs used across models. Finally, for the adenoma group
-differences between those that received surgery or not was tested while
-for the carcinoma group differences between those receiving chemotherapy
-and radiation was tested.
-
-\textbf{\emph{Reproducible Methods:}} A detailed and reproducible
+\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
 description of how the data were processed and analyzed can be found at
@@
 
-\textbf{Figure 2: Treatment Response Based on Models Built for Adenoma,
-SRN, or Carcinoma.} A) Positive probability change from initial to
-follow up sample in those with adenoma. B) Positive probability change
-from initial to follow up sample in those with SRN. C) Positive
+\textbf{Figure 2: The 10 OTUs used to classify treatment for Adenoma,
+Advanced Adenoma, and Carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
+OTUs. C) Carcinoma OTUs.
+
+\textbf{Figure 3: Treatment Response Based on Models Built for Adenoma,
+Advanced Adenoma, or Carcinoma.} A) Positive probability change from
+initial to follow up sample in those with adenoma. B) Positive
 probability change from initial to follow up sample in those with
-carcinoma..
+advanced adenoma. C) Positive probability change from initial to follow
+up sample in those with carcinoma..
 
-\textbf{Figure 3: Common OTUs Model Rank Based on Median MDA.} For each
-common OTU the lowest taxonomic identification and importance rank for
-each model run is shown.
+\textbf{Figure 4: Common OTUs to All Models.} A) Venn diagram showing
+the OTU overlap between each model. B) For each common OTU the lowest
+taxonomic identification and importance rank for each model run is
+shown.
 
@@
 
-\textbf{Table 1: Demographic Data of Training Cohort}
+\textbf{Table 1: Demographic Data of Pre and Post Treatment Cohort}
 
-\textbf{Table 2: Demographic Data of Pre and Post Treatment Cohort}
+\textbf{Table 2: Demographic Data of Training Cohort}
 
@@
 
-**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma
-Models.*. A) Adenoma ROC curve: The light green shaded areas represent
-the range of values of a 100 different 80/20 splits of the test set data
-and the dark green line represents the model using 100\% of the data set
-and what was used for subsequent classification. B) Advanced Adenoma ROC
-curve: The light yellow shaded areas represent the range of values of a
-100 different 80/20 splits of the test set data and the dark yellow line
+\textbf{Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and
+Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
+represent the range of values of a 100 different 80/20 splits of the
+test set data and the dark green line represents the model using 100\%
+of the data set and what was used for subsequent classification. B)
+Advanced Adenoma ROC curve: The light yellow shaded areas represent the
+range of values of a 100 different 80/20 splits of the test set data and
+the dark yellow line represents the model using 100\% of the data set
+and what was used for subsequent classification. C) Carcinoma ROC curve:
+The light red shaded areas represent the range of values of a 100
+different 80/20 splits of the test set data and the dark red line
 represents the model using 100\% of the data set and what was used for
-subsequent classification. C) Carcinoma ROC curve: The light red shaded
-areas represent the range of values of a 100 different 80/20 splits of
-the test set data and the dark red line represents the model using 100\%
-of the data set and what was used for subsequent classification.
+subsequent classification.
 
-**Figure S3: Summary of Important Variables for the Adenoma, Advanced
-Adenoma, and Carcinoma Models.*. A) MDA of the most important variables
+\textbf{Figure S3: Summary of Important OTUs for the Adenoma, Advanced
+Adenoma, and Carcinoma Models.} A) MDA of the most important variables
 in the adenoma model. The dark green point represents the mean and the
@@
 
-\hypertarget{ref-sze_looking_2016}{}
-28. Sze MA, Schloss PD. Looking for a Signal in the Noise: Revisiting
-Obesity and the Microbiome. mBio. 2016;7.
-
 \hypertarget{ref-obrien_impact_2013}{}
-29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+28. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-30. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+29. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-breiman_random_2001}{}
-31. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+30. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-32. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+31. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-33. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+32. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-04-21 submission/supplemental.Rmd
last minute changes

@@
 ---
-title: ""
-bibliography:
+title: ''
 csl: microbiome.csl
+fontsize: 12pt
 output:
+  word_document: default
   pdf_document:
-    keep_tex: true
     includes:
       in_header: header.tex
-fontsize: 12pt
+    keep_tex: yes
 geometry: margin=1.0in
+bibliography: null
 ---
 
-# Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment
+# The effect of treatment on the microbiota of patients diagnosed with colonic lesions
 

### 2017-04-21 submission/manuscript.Rmd

@@
 ---
-title: ""
-bibliography: reference.bib
+title: ''
+csl: microbiome.csl
+fontsize: 12pt
 output:
   pdf_document:
-    keep_tex: true
     includes:
       in_header: header.tex
-csl: microbiome.csl
-fontsize: 12pt
+    keep_tex: yes
+  word_document: default
 geometry: margin=1.0in
+bibliography: reference.bib
 ---
@@
 
-**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.*.
-A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
 Solid black points represent the median value for each diagnosis group.

### 2017-04-21 submission/manuscript.tex

@@
 
-**Figure 1: General Differences between Adenoma, Advanced Adenoma, and
-Carcinoma Groups After Treatment.*. A) A significant difference was
+\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
+and Carcinoma Groups After Treatment.} A) A significant difference was
 found between the adenoma and carcinoma group for thetayc (P-value =

### 2017-04-24 submission/manuscript.Rmd

@@
 **Results.** There were large changes to the bacterial communities associated with treatment across the three groups.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05).
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05).
 Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups.
-However, across the three models, there was little overlap between the OTUs that were indicative of treatment.
+Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups.
+Although the three models successfully differentiated between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment.
 Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent a treatment were more likely to have OTUs associated with normal colons.
 We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses.
-Although we could clearly differentiate pre- and post-treatment communities from the three diagnosis groups, only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a carcinoma (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a lesion (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 **Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiota will yield improvements in disease management.
-
-\newpage
 
@@
 Disease mortality has significantly decreased, thanks to improvements in screening [@jemal_cancer_2010].
-Despite this improvement, there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].
+Despite this improvement, there are still approximately 50,000 CRC-related deaths per year [@haggar_colorectal_2009].
 Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014].
@@
 Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
-Collectively, these studies hypothesize that the microbiota is altering the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
+Collectively, these studies support the hypothesis that the microbiota can alter the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
 
-Building on this evidence, several human-based studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015].
+Building upon this evidence, several human-based studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015].
 One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012].
 Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].
-These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions (adenoma, advanced adenoma, and carcinoma).
+These studies have associated *Fusobacterium nucleatum* and other oral pathogens with colonic lesions (adenoma, advanced adenoma, and carcinoma).
 They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions.
@@
 
-We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre- and post-treatment.
-We used both classical community analysis to compare the alpha and beta-diversity of communities pre- and post treatment.
+We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre and post-treatment.
+We used both classical community analysis to compare the alpha and beta-diversity of communities pre and post treatment.
 Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group.
@@
 ***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples [Table 1].
-The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A].
-We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`).
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
+The structure of the microbial communities of the pre and post-treatment samples differed [Figure 1A].
+We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`).
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
-We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`).
+We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value=`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value=`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value=`r beta_summary["Carcinoma", "PERMANOVA"]`).
 Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].
@@
 Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data.
-Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`).
+Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
@@
 We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon.
-The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2).
+The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
 The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
@@
 There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased.
-In fact, only `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment.
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment.
+In fact, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment.
 One of those was re-diagnosed with carcinoma on the follow up visit.
@@
 Finally, the change in the positive probability was not significantly different between the two treatment groups (P=`r chemo_rad_summary["red_crc", "bh"]`).
-Due to the relatively small number of samples in each treatment group, it is difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota
+Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota
 
@@
 This leads us to hypothesize that these individuals may be at a high risk of recurrence.
-Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]), which would support our hypothesis .
+Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]).
 Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence.
-Regardless, it is intriguing to think of using biomarkers from the microbiota to not only predict the presence of lesions, but to also assess the risk of developing them.
+Regardless, it is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions, but to also assess the risk of developing them.
 
@@
 Each group was randomly assigned to avoid biases based on sample collection location.
-The pre- and post-treatment samples were sequenced on the same run.
+The pre and post-treatment samples were sequenced on the same run.
 
@@
 Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
-Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxon ranked sum test.
+Comparisons between probabilities as well as overall OTU differences between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test.
 Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
@@
 Solid black points represent the median value for each diagnosis group.
-B) NMDS of the pre- and post-treatment samples for the adenoma group.
-C) NMDS of the pre- and post-treatment samples for the advanced adenoma group.
-D) NMDS of the pre- and post-treatment samples for the carcinoma group.
+B) NMDS of the pre and post-treatment samples for the adenoma group.
+C) NMDS of the pre and post-treatment samples for the advanced adenoma group.
+D) NMDS of the pre and post-treatment samples for the carcinoma group.
 
@@
 
-**Table 1: Demographic Data of Pre and Post Treatment Cohort**
+**Table 1: Demographic Data of pre and Post Treatment Cohort**
 

### 2017-04-24 submission/manuscript.tex
re-rendering as requested

@@
 patients with carcinomas changed significantly more than those with
-adenoma following treatment (P-value=5.4e-05); however, there was not a
-significant difference between those with advanced adenoma and those
-with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
+adenoma following treatment (P-value \textless{} 0.001); however, there
+was not a significant difference between those with advanced adenoma and
+those with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
 treatment brought about large intrapersonal changes, the change in the
@@
 Random Forest machine learning algorithm to identify groups of OTUs that
-allowed us to successfully distinguish between pre- and post-treatment
-samples for each of the diagnosis groups. However, across the three
-models, there was little overlap between the OTUs that were indicative
-of treatment. Next, we used a larger cohort that contained individuals
-with normal colons and those with adenomas, advanced adenomas, and
-carcinomas to determine whether individuals who underwent a treatment
-were more likely to have OTUs associated with normal colons. We again
-built Random Forest models and measured the change in the positive
-probability of having one of the three diagnoses. Although we could
-clearly differentiate pre- and post-treatment communities from the three
-diagnosis groups, only those patients that initially had carcinomas
-experienced a significant decrease in positive probability of having a
-carcinoma (P-value \textless{} 0.05). Finally, we tested whether the
-type of treatment impacted the microbiota of those diagnosed with
-carcinomas and were unable to detect any significant differences in
-characteristics of these communities between individuals treated with
-surgery alone and those treated with chemotherapy or chemotherapy and
-radiation (P-value \textgreater{} 0.05).
+allowed us to successfully distinguish between pre and post-treatment
+samples for each of the diagnosis groups. Although the three models
+successfully differentiated between the pre and post-treatment samples,
+there was little overlap between the OTUs that were indicative of
+treatment. Next, we used a larger cohort that contained individuals with
+normal colons and those with adenomas, advanced adenomas, and carcinomas
+to determine whether individuals who underwent a treatment were more
+likely to have OTUs associated with normal colons. We again built Random
+Forest models and measured the change in the positive probability of
+having one of the three diagnoses. Only those patients that initially
+had carcinomas experienced a significant decrease in positive
+probability of having a lesion (P-value \textless{} 0.05). Finally, we
+tested whether the type of treatment impacted the microbiota of those
+diagnosed with carcinomas and were unable to detect any significant
+differences in characteristics of these communities between individuals
+treated with surgery alone and those treated with chemotherapy or
+chemotherapy and radiation (P-value \textgreater{} 0.05).
 
@@
 improvements in disease management.
-
-\newpage
 
@@
 decreased, thanks to improvements in screening {[}1{]}. Despite this
-improvement, there are still approximately 50,000 deaths from the
-disease per year {[}2{]}. Current estimates indicate that 20-30\% of
-those who undergo treatment will experience recurrence and 30-50\% of
-those patients will die {[}3,4{]}. Identification of methods to assess
+improvement, there are still approximately 50,000 CRC-related deaths per
+year {[}2{]}. Current estimates indicate that 20-30\% of those who
+undergo treatment will experience recurrence and 30-50\% of those
+patients will die {[}3,4{]}. Identification of methods to assess
 patients' risk of recurrence is of great importance to reduce mortality
@@
 result in varied tumor burden {[}10--12{]}. Collectively, these studies
-hypothesize that the microbiota is altering the amount of inflammation
-in the colon and with it the rate of tumorigenesis {[}13{]}.
+support the hypothesis that the microbiota can alter the amount of
+inflammation in the colon and with it the rate of tumorigenesis
+{[}13{]}.
 
-Building on this evidence, several human-based studies have identified
+Building upon this evidence, several human-based studies have identified
 unique signatures of colonic lesions {[}14--19{]}. One line of research
@@
 have associated \emph{Fusobacterium nucleatum} and other oral pathogens
-with lesions (adenoma, advanced adenoma, and carcinoma). They have also
-noted that the loss of bacteria generally thought to produce short chain
-fatty acids, which can suppress inflammation, is associated with colonic
-lesions. Gut bacteria have a role in tumorigenesis and there is promise
-that these populations may be useful biomarkers for aiding in the early
-detection of disease.
+with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
+have also noted that the loss of bacteria generally thought to produce
+short chain fatty acids, which can suppress inflammation, is associated
+with colonic lesions. Gut bacteria have a role in tumorigenesis and
+there is promise that these populations may be useful biomarkers for
+aiding in the early detection of disease.
 
@@
 genes amplified from fecal samples of individuals with adenoma, advanced
-adenoma, and carcinomas pre- and post-treatment. We used both classical
+adenoma, and carcinomas pre and post-treatment. We used classical
 community analysis to compare the alpha and beta-diversity of
-communities pre- and post treatment. Next, we generated Random Forest
+communities pre and post treatment. Next, we generated Random Forest
 models to identify bacterial populations that were indicative of
@@
 advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
-between pre- and post-treatment samples {[}Table 1{]}. The structure of
-the microbial communities of the pre- and post-treatment samples
-differed {[}Figure 1A{]}. We found that the communities obtained pre-
-and post-treament among the patients with carcinomas changed
-significantly more than those patients with adenoma (P-value = 5.4e-05).
-There were no significant differences in the amount of change observed
-between the patients with adenoma and advanced adenoma or between the
-patients with advanced adenoma and carcinoma (P-value \textgreater{}
-0.05). Next, we tested whether there was a consistent direction in the
-change in the community structure between the pre and post-treatment
-samples for each of the diagnosis groups {[}Figure 1B-D{]}. We only
-observed a consistent shift in community structure for the patients with
-carcinoma when using a PERMANOVA test (adenoma P-value0.999, advanced
-adenoma P-value0.945, and carcinoma P-value0.005). Finally, we measured
-the number of observed OTUs, Shannon evenness, and Shannon diversity in
-the pre and post-treatment samples and did not observe a significant
-change for any of the diagnosis groups (P-value \textgreater{} 0.05)
-{[}Table S1{]}.
+between pre and post-treatment samples {[}Table 1{]}. The structure of
+the microbial communities of the pre and post-treatment samples differed
+{[}Figure 1A{]}. We found that the communities obtained pre and
+post-treament among the patients with carcinomas changed significantly
+more than those patients with adenoma (P-value = 5.4e-05). There were no
+significant differences in the amount of change observed between the
+patients with adenoma and advanced adenoma or between the patients with
+advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next, we
+tested whether there was a consistent direction in the change in the
+community structure between the pre and post-treatment samples for each
+of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
+shift in community structure for the patients with carcinoma when using
+a PERMANOVA test (adenoma P-value=0.999, advanced adenoma P-value=0.945,
+and carcinoma P-value=0.005). Finally, we measured the number of
+observed OTUs, Shannon evenness, and Shannon diversity in the pre and
+post-treatment samples and did not observe a significant change for any
+of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.
 
@@
 group, we restricted our models to only incorporate 10 OTUs to limit the
-likelihood that that models would overfit the data. Despite this
+likelihood that the models would overfit the data. Despite this
 restriction, the models performed well (adenoma AUC=0.69 - 0.92,
-advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
-the 10 OTUs that were used for each model had little overlap with each
-other {[}Figure 2{]}. These results support the earlier community-wide
-analysis where we observed that the treatment had an impact on the
-overall community structure; however, the effect of treatment was not
-consistent across patients and diagnosis groups.
+advanced adenoma AUC=0.80 - 1.00, carcinoma AUC=0.82 - 0.98).
+Interestingly, the 10 OTUs that were used for each model had little
+overlap with each other {[}Figure 2{]}. These results support the
+earlier community-wide analysis where we observed that the treatment had
+an impact on the overall community structure; however, the effect of
+treatment was not consistent across patients and diagnosis groups.
 
@@
 models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
-AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2). The OTUs that
+AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.90; Figure S2). The OTUs that
 were incorporated into the adenoma and advanced adenoma models largely
@@
 pre and post-treatment samples from patients diagnosed with carcinoma
-significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
-that were initially diagnosed with a carcinoma had a higher positive
-probability after treatment. One of those was re-diagnosed with
+significantly decreased with treatment. In fact, 6 of the 26 patients
+(23.08\%) that were initially diagnosed with a carcinoma had a higher
+positive probability after treatment. One of those was re-diagnosed with
 carcinoma on the follow up visit. These results indicate that although
@@
 where the change was directed towards what would be found in a normal
-colon
+colon.
 
@@
 significantly different between the two treatment groups (P=1). Due to
-the relatively small number of samples in each treatment group, it is
+the relatively small number of samples in each treatment group, it was
 difficult to make a definitive statement regarding the specific type of
-treatment on the amount of change in the structure of the microbiota
+treatment on the amount of change in the structure of the microbiota.
 
@@
 probability from our study is within the expected rate of recurrence
-(20-30\% {[}3,4{]}), which would support our hypothesis . Unfortunately,
-because of the design of our study, it was not possible to determine
-whether individuals who had an increased positive probability for
-carcinoma experienced recurrence. Regardless, it is intriguing to think
-of using biomarkers from the microbiota to not only predict the presence
-of lesions, but to also assess the risk of developing them.
+(20-30\% {[}3,4{]}). Unfortunately, because of the design of our study,
+it was not possible to determine whether individuals who had an
+increased positive probability for carcinoma experienced recurrence.
+Regardless, it is intriguing that it may be possible to use
+microbiome-based biomarkers to not only predict the presence of lesions,
+but to also assess the risk of developing them.
 
@@
 the length of time between sample collection among those whose
-probabilities increased or decreased (Wilcoxon Test; P-value=0.56).
+probabilities increased or decreased (Wilcoxon Test; P-value = 0.56).
 Finally, it is possible that these patients may not have responded to
@@
 Each group was randomly assigned to avoid biases based on sample
-collection location. The pre- and post-treatment samples were sequenced
+collection location. The pre and post-treatment samples were sequenced
 on the same run.
@@
 included merging paired-end reads into contigs, filtering for low
-quality contigs, aligning to the SILVA database {[}reference{]},
-screening for chimeras using UCHIME {[}reference{]}, classifying with a
-naive Bayesian classifier using the Ribosomal Database Project
-(RDP){[}reference-Wang{]}, and clustered into Operational Taxonomic
-Units (OTUs) using a 97\% similarity cutoff with an average neighbor
-clustering algorithm {[}reference-Westcott\&SchlossAEM{]}. The number of
-sequences for each sample was rarefied to 10523 to minimize the impacts
-of uneven sampling.
+quality contigs, aligning to the SILVA database {[}30{]}, screening for
+chimeras using UCHIME {[}31{]}, classifying with a naive Bayesian
+classifier using the Ribosomal Database Project (RDP){[}32{]}, and
+clustered into Operational Taxonomic Units (OTUs) using a 97\%
+similarity cutoff with an average neighbor clustering algorithm
+{[}33{]}. The number of sequences for each sample was rarefied to 10523
+to minimize the impacts of uneven sampling.
 
-\textbf{\emph{Model Building.}} The Random Forest {[}30{]} algorithm was
+\textbf{\emph{Model Building.}} The Random Forest {[}34{]} algorithm was
 used to create the model used to create the three models used. The
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}31{]} in the vegan package
+community structure utilized PERMANOVA {[}35{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
-differences between pre- and post-treatment samples utilized a paired
+differences between pre and post-treatment samples utilized a paired
 Wilcoxon ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}32{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 NULL). Solid black points represent the median value for each diagnosis
-group. B) NMDS of the pre- and post-treatment samples for the adenoma
-group. C) NMDS of the pre- and post-treatment samples for the advanced
-adenoma group. D) NMDS of the pre- and post-treatment samples for the
+group. B) NMDS of the pre and post-treatment samples for the adenoma
+group. C) NMDS of the pre and post-treatment samples for the advanced
+adenoma group. D) NMDS of the pre and post-treatment samples for the
 carcinoma group.
@@
 
-\textbf{Table 1: Demographic Data of Pre and Post Treatment Cohort}
+\textbf{Table 1: Demographic Data of pre and Post Treatment Cohort}
 
@@
 
+\hypertarget{ref-pruesse_silva_2007}{}
+30. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
+SILVA: A comprehensive online resource for quality checked and aligned
+ribosomal RNA sequence data compatible with ARB. Nucleic Acids Research.
+2007;35:7188--96.
+
+\hypertarget{ref-edgar_uchime_2011}{}
+31. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
+sensitivity and speed of chimera detection. Bioinformatics (Oxford,
+England). 2011;27:2194--200.
+
+\hypertarget{ref-wang_naive_2007}{}
+32. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
+for rapid assignment of rRNA sequences into the new bacterial taxonomy.
+Applied and Environmental Microbiology. 2007;73:5261--7.
+
+\hypertarget{ref-schloss_assessing_2011}{}
+33. Schloss PD, Westcott SL. Assessing and improving methods used in
+operational taxonomic unit-based approaches for 16S rRNA gene sequence
+analysis. Applied and Environmental Microbiology. 2011;77:3219--26.
+
 \hypertarget{ref-breiman_random_2001}{}
-30. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+34. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-31. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-32. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-04-24 submission/manuscript.Rmd

@@
 **Results.** There were large changes to the bacterial communities associated with treatment across the three groups.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05).
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05).
 Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre and post-treatment.
-We used both classical community analysis to compare the alpha and beta-diversity of communities pre and post treatment.
+We used classical community analysis to compare the alpha and beta-diversity of communities pre and post treatment.
 Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group.
@@
 To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data.
+Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data.
 Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
@@
 One of those was re-diagnosed with carcinoma on the follow up visit.
-These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon
+These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon.
 
@@
 Finally, the change in the positive probability was not significantly different between the two treatment groups (P=`r chemo_rad_summary["red_crc", "bh"]`).
-Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota
+Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
 
@@
 
-It was not clear why 6 of the 26 patients with carcinoma had a higher probability of having carcinomas after treatment.
-We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these 6 individuals (3 with Stage II and 3 with Stage III) was similar to the distribution observed among the other 20 patients.
-We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased or decreased (Wilcoxon Test; P-value=0.56).
-Finally, it is possible that these patients may not have responded to treatment as well as the other 20 patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
+It was not clear why `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` patients with carcinoma had a higher probability of having carcinomas after treatment.
+We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
+We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased or decreased (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`).
+Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
 Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.
@@
 The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013].
-The general workflow using mothur included merging paired-end reads into contigs, filtering for low quality contigs, aligning to the SILVA database [reference], screening for chimeras using UCHIME [reference], classifying with a naive Bayesian classifier using the Ribosomal Database Project (RDP)[reference-Wang], and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm [reference-Westcott&SchlossAEM].
+The general workflow using mothur included merging paired-end reads into contigs, filtering for low quality contigs, aligning to the SILVA database [@pruesse_silva_2007], screening for chimeras using UCHIME [@edgar_uchime_2011], classifying with a naive Bayesian classifier using the Ribosomal Database Project (RDP)[@wang_naive_2007], and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm [@schloss_assessing_2011].
 The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize the impacts of uneven sampling.

### 2017-04-28 submission/manuscript.Rmd

@@
 **Results.** There were large changes to the bacterial communities associated with treatment across the three groups.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05).
-Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups.
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
+There was no significant change in the microbiota between patients with adenoma and advanced adenoma, or between patients with advanced adenoma and carcinoma (P-value > 0.05).
+Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups.
 Although the three models successfully differentiated between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment.
-Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent a treatment were more likely to have OTUs associated with normal colons.
+Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons.
 We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses.
-Only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a lesion (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Only patients who had carcinomas experienced a significant decrease in positive probability of having a lesion after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05), indicating that the microbial milieu of the colon more closely resembled that of a normal colon.
 Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 
-Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009].
-Disease mortality has significantly decreased, thanks to improvements in screening [@jemal_cancer_2010].
-Despite this improvement, there are still approximately 50,000 CRC-related deaths per year [@haggar_colorectal_2009].
-Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014].
+Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@siegel_cancer_2016; @haggar_colorectal_2009].
+Disease mortality has significantly decreased, predominately due to improvements in screening [@haggar_colorectal_2009].
+Despite this improvement, there are still approximately 50,000 CRC-related deaths per year [@siegel_cancer_2016].
+Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 35% of all patients will die within five years [@hellinger_reoperation_2006; @ryuk_predictive_2014; @national_cancer_institute_seer_nodate].
 Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.
@@
 Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009].
-Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
+Furthermore, studies that shift the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces have shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
 Collectively, these studies support the hypothesis that the microbiota can alter the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
 
-Building upon this evidence, several human-based studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015].
+Building upon this evidence, several human studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015].
 One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012].
@@
 They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions.
-Gut bacteria have a role in tumorigenesis and there is promise that these populations may be useful biomarkers for aiding in the early detection of disease.
+This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease.
 
@@
 Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group.
-Finally, we measured the predictive probability of diagnosis models to determine whether treatment would alter the probability of being diagnosed as having a normal colon or the colon of the original diagnosis group.
-We found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas.
-Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence.
+Finally, we measured the predictive probabilities to assess whether treatment yielded bacterial communities similar to those individuals with normal colons.
+We found that treatment alters the composition of the gut microbiota and that, for those with carinomas, the gut microbiota shifted more towards that of a normal colon after treatment.
+In the individuals with carcinomas, no difference was found byt the type of treatment (surgery alone versus surgery with chemotherapy).
+Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence and the likelihood of survival.
 

### 2017-04-28 submission/manuscript.Rmd
complete Mary comments

@@
 However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1].
-It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test.
+It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
 To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
@@
 
-***Post-treatment samples from patients with carcinoma change towards a microbiota associated with normal***
+***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
 Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons.
 To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2].
-We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon.
+We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
 The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
@@
 The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment.
-In fact, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment.
-One of those was re-diagnosed with carcinoma on the follow up visit.
-These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon.
+Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
+These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by  patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.
 
@@
 We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P= `r chemo_rad_summary["dist", "bh"]`).
-Finally, the change in the positive probability was not significantly different between the two treatment groups (P=`r chemo_rad_summary["red_crc", "bh"]`).
+Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Finally, the change in the positive probability was not significantly different between the two treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
 Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
@@
 
-This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiota contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
+This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that microbiome-derived biomarkers can detect colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
 Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
-For all three groups of patients we observed changes in their microbiota.
-The microbiota of patients diagnosed with carcinoma changed significantly more than the other groups.
+For all three groups of patients, we observed changes in their microbiota.
+After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
 This change resulted in communities that more closely resembled those of patients with a normal colon.
-For the patients diagnosed with carcinomas there was a larger change between their initial and follow up samples than was observed for patients diagnosed with either adenoma or advanced adenoma.
-The changes in the community structure of those diagnosed with carcinoma were similar and resulted in communities that more closely resembled those found in patients with normal colons.
+This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.
+Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.
+It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to assess the risk of recurrence.
 
-Understanding the effect of treatment on the microbiota of those initially diagnosed with carcinomas may have important implications into understanding disease recurrence.
-The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted in community structures that resembled patients with normal colons.
-This would suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at removing bacteria associated with them.
-Interestingly, 6 of the 26 patients diagnosed with carcinomas had an elevated probability of having carcinomas after treatment.
-This leads us to hypothesize that these individuals may be at a high risk of recurrence.
-Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]).
-Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence.
-Regardless, it is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions, but to also assess the risk of developing them.
-
-It is interesting that those patients diagnosed with adenoma and advanced adenoma did not experience a shift towards a community structure that resembled those with normal colons.
+Patients diagnosed with adenoma and advanced adenoma, however, did not experience a shift towards a community structure that resembled those with normal colons.
 This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma.
@@
 This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S2].
-Given the patchy distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
-A final hypothesis is that the specific type of treatment altered the structure of the microbiome.
-The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma.
-Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turn over of the colonic epithelium driving a more significant change in the structure of the microbiota.
-Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.
+Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
 
-It was not clear why `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` patients with carcinoma had a higher probability of having carcinomas after treatment.
+There was a subset of patients (`r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` with carcinomas) who demonstrated an elevated probability of carcinoma after treatment.
+This may reflect an elevated risk of recurrence.
+The `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]).
 We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
@@
 
-This study builds upon existing work that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those initially diagnosed with carcinoma.
-If such an approach is effective, it might be possible to even target the microbiota as part of combined therapies including currently standard approaches.
+Our final hypothesis was that the specific type of treatment altered the structure of the microbiome.
+The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma.
+Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turnover of the colonic epithelium driving a more significant change in the structure of the microbiota.
+Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.
+
+This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
+The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
+If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy.
 Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.
@@
 The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used.
-The adenoma model classified normal versus adenoma, advanced adenoma was normal versus advanced adenoma, and carcinoma was normal versus carcinoma.
-The toal number of individuals in this data set was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 1].
-There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` indivdiuals in the carcinoma model [Table 1].
-Each model was then applied to our `r length(rownames(good_metaf))`-person cohort testing prediction of adenoma intial (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])`) versus adenoma follow up (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])`), advanced adenoma initial (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])`) versus advanced adenoma follow up (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])`), carcinoma initial (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])`) versus carcinoma follow up (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")[, "followUp"])`).
+The adenoma model classified normal versus adenoma, the advanced adenoma model classified normal versus advanced adenoma, and the carcinoma model classified normal versus carcinoma.
+The total number of individuals in the treatment models was `r length(rownames(good_metaf))` individuals.
+There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the treatment carcinoma model [Table 1].
+The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2].
+There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the carcinoma model [Table 2].
+Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] which assesed the prediction of pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])`), pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])`) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])`), pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))`).
 
@@
 All three models had an optimized mtry of `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
+The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.
 
@@
 For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
-The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.
+The only difference other than the sample set used between treatment models and normal versus diagnosis models was that only the top 10 OTUs were used to build each respective treatment model by diagnosis group to help avoid model overfitting.T
 
@@
 
-**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).
+**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
 Solid black points represent the median value for each diagnosis group.
@@
 B) Positive probability change from initial to follow up sample in those with advanced adenoma.
-C) Positive probability change from initial to follow up sample in those with carcinoma..
+C) Positive probability change from initial to follow up sample in those with carcinoma.
 
@@
 
+```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
+test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White)
+
+t2 <- as.data.frame(matrix(nrow = 5, ncol = 3, 
+                           dimnames = 
+                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
+                                  coln = c("Adenoma", "SRN", "Carcinoma"))))
+
+data_filter1 <- c("adenoma", "adenoma", "cancer")
+data_filter2 <- c("adenoma", "adv_adenoma", "cancer")
+
+# Generate the summary stats for each group
+for(i in 1:length(colnames(t2))){
+
+  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
+  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
+  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
+  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
+  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
+}
+
+# Generate SD for applicable metrics
+age_sd <- c()
+BMI_sd <- c()
+
+for(i in 1:length(colnames(t2))){
+
+  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
+  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
+}
+
+# Create combined table
+# "u\00B1" is to add a +/- symbol where applicable
+
+ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
+                                                    label = "\u00B1", 
+                                                    format(age_sd, digits = 2, nmsall = 2)), 
+                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
+                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
+                                                    label = "\u00B1", 
+                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
+                                        White = format(White, digits =2, nsmall = 2))
+
+ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
+             ttemptable$BMI, ttemptable$White)
+
+age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
+bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
+
+colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)")
+
+test <- t(ft1)
+colnames(test) <- test["Group", ]
+test <- test[-1, ]
+
+kable(test, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
 **Table 2: Demographic Data of Training Cohort**
+
+```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
+
+metadata <- filter(metaI, !sample %in% c(good_metaf$initial, good_metaf$followUp)) %>% 
+  select(Dx_Bin, dx, Age, Gender, BMI, White) %>% mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin), Dx_Bin = gsub("High Risk Normal", "Normal", Dx_Bin))
+
+t1 <- as.data.frame(matrix(nrow = 5, ncol = 4, 
+                           dimnames = 
+                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
+                                  coln = c("Normal", "Adenoma", "SRN", "Carcinoma"))))
+
+data_filter1 <- c("normal", "adenoma", "adenoma", "cancer")
+data_filter2 <- c("Normal", "Adenoma", "adv_adenoma", "Cancer")
+
+# Generate the summary stats for each group
+for(i in 1:length(colnames(t1))){
+
+  t1["Dx_Bin", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
+  t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
+  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
+  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
+  t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
+}
+
+# Generate SD for applicable metrics
+age_sd <- c()
+BMI_sd <- c()
+
+for(i in 1:length(colnames(t1))){
+
+  age_sd <- c(age_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
+  BMI_sd <- c(BMI_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
+}
+
+# Create combined table
+# "u\00B1" is to add a +/- symbol where applicable
+
+ttemptable <- as.data.frame(t(t1)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
+                                                    label = "\u00B1", 
+                                                    format(age_sd, digits = 2, nmsall = 2)), 
+                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
+                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
+                                                    label = "\u00B1", 
+                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
+                                        White = format(White, digits =2, nsmall = 2))
+
+ft1 <- cbind(Group = c("Normal", "Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
+             ttemptable$BMI, ttemptable$White)
+
+age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
+bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
+
+colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)")
+
+test <- t(ft1)
+colnames(test) <- test["Group", ]
+test <- test[-1, ]
+
+kable(test, format = "markdown", align = 'c', padding = 2)
+```
 

### 2017-04-28 submission/manuscript.tex
complete Mary comments

@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
 patients with carcinomas changed significantly more than those with
-adenoma following treatment (P-value \textless{} 0.001); however, there
-was not a significant difference between those with advanced adenoma and
-those with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
-treatment brought about large intrapersonal changes, the change in the
-abundance of individual OTUs to treatment was not consistent within
-diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
-of OTUs across patients and diagnosis groups was patchy, we used the
-Random Forest machine learning algorithm to identify groups of OTUs that
-allowed us to successfully distinguish between pre and post-treatment
-samples for each of the diagnosis groups. Although the three models
-successfully differentiated between the pre and post-treatment samples,
-there was little overlap between the OTUs that were indicative of
-treatment. Next, we used a larger cohort that contained individuals with
-normal colons and those with adenomas, advanced adenomas, and carcinomas
-to determine whether individuals who underwent a treatment were more
-likely to have OTUs associated with normal colons. We again built Random
-Forest models and measured the change in the positive probability of
-having one of the three diagnoses. Only those patients that initially
-had carcinomas experienced a significant decrease in positive
-probability of having a lesion (P-value \textless{} 0.05). Finally, we
-tested whether the type of treatment impacted the microbiota of those
-diagnosed with carcinomas and were unable to detect any significant
-differences in characteristics of these communities between individuals
-treated with surgery alone and those treated with chemotherapy or
-chemotherapy and radiation (P-value \textgreater{} 0.05).
+adenoma following treatment (P-value \textless{} 0.001). There was no
+significant change in the microbiota between patients with adenoma and
+advanced adenoma, or between patients with advanced adenoma and
+carcinoma (P-value \textgreater{} 0.05). Although treatment was
+associated with intrapersonal changes, the change in the abundance of
+individual OTUs to treatment was not consistent within diagnosis groups
+(P-value \textgreater{} 0.05). Because the distribution of OTUs across
+patients and diagnosis groups was irregular, we used the Random Forest
+machine learning algorithm to identify groups of OTUs that allowed us to
+successfully distinguish between pre and post-treatment samples for each
+of the diagnosis groups. Although the three models successfully
+differentiated between the pre and post-treatment samples, there was
+little overlap between the OTUs that were indicative of treatment. Next,
+we used a larger cohort that contained individuals with normal colons
+and those with adenomas, advanced adenomas, and carcinomas to determine
+whether individuals who underwent treatment were more likely to have
+OTUs associated with normal colons. We again built Random Forest models
+and measured the change in the positive probability of having one of the
+three diagnoses. Only patients who had carcinomas experienced a
+significant decrease in positive probability of having a lesion after
+treatment (P-value \textless{} 0.05), indicating that the microbial
+milieu of the colon more closely resembled that of a normal colon.
+Finally, we tested whether the type of treatment impacted the microbiota
+of those diagnosed with carcinomas and were unable to detect any
+significant differences in characteristics of these communities between
+individuals treated with surgery alone and those treated with
+chemotherapy or chemotherapy and radiation (P-value \textgreater{}
+0.05).
 
@@
 in the United States {[}1,2{]}. Disease mortality has significantly
-decreased, thanks to improvements in screening {[}1{]}. Despite this
-improvement, there are still approximately 50,000 CRC-related deaths per
-year {[}2{]}. Current estimates indicate that 20-30\% of those who
-undergo treatment will experience recurrence and 30-50\% of those
-patients will die {[}3,4{]}. Identification of methods to assess
-patients' risk of recurrence is of great importance to reduce mortality
-and healthcare costs.
+decreased, predominately due to improvements in screening {[}2{]}.
+Despite this improvement, there are still approximately 50,000
+CRC-related deaths per year {[}1{]}. Current estimates indicate that
+20-30\% of those who undergo treatment will experience recurrence and
+35\% of all patients will die within five years {[}3--5{]}.
+Identification of methods to assess patients' risk of recurrence is of
+great importance to reduce mortality and healthcare costs.
 
@@
 \emph{Fusobacterium nucleatum} that alter disease progression
-{[}5--9{]}. Furthermore, studies that alter the structure of the
+{[}6--10{]}. Furthermore, studies that shift the structure of the
 microbiota through the use of antibiotics or inoculation of germ free
-mice with human feces has shown that varying community compositions can
-result in varied tumor burden {[}10--12{]}. Collectively, these studies
+mice with human feces have shown that varying community compositions can
+result in varied tumor burden {[}11--13{]}. Collectively, these studies
 support the hypothesis that the microbiota can alter the amount of
 inflammation in the colon and with it the rate of tumorigenesis
-{[}13{]}.
+{[}14{]}.
 
-Building upon this evidence, several human-based studies have identified
-unique signatures of colonic lesions {[}14--19{]}. One line of research
+Building upon this evidence, several human studies have identified
+unique signatures of colonic lesions {[}15--20{]}. One line of research
 has identified community-level differences between those bacteria that
@@
 for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
-\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
+\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
 have proposed feces-based biomarkers that could be used to diagnose the
-presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
+presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
 have associated \emph{Fusobacterium nucleatum} and other oral pathogens
@@
 short chain fatty acids, which can suppress inflammation, is associated
-with colonic lesions. Gut bacteria have a role in tumorigenesis and
-there is promise that these populations may be useful biomarkers for
-aiding in the early detection of disease.
+with colonic lesions. This suggests that gut bacteria have a role in
+tumorigenesis with potential as useful biomarkers for aiding in the
+early detection of disease.
 
@@
 treatment for each diagnosis group. Finally, we measured the predictive
-probability of diagnosis models to determine whether treatment would
-alter the probability of being diagnosed as having a normal colon or the
-colon of the original diagnosis group. We found that treatment alters
-the composition of the gut microbiota; however, we only observed a shift
-from diseased to normal colon for those initially diagnosed with
-carcinomas. Understanding how the community responds to these treatments
-could be a valuable tool for identifying biomarkers to quantify the risk
-of recurrence.
+probabilities to assess whether treatment yielded bacterial communities
+similar to those individuals with normal colons. We found that treatment
+alters the composition of the gut microbiota and that, for those with
+carinomas, the gut microbiota shifted more towards that of a normal
+colon after treatment. In the individuals with carcinomas, no difference
+was found byt the type of treatment (surgery alone versus surgery with
+chemotherapy). Understanding how the community responds to these
+treatments could be a valuable tool for identifying biomarkers to
+quantify the risk of recurrence and the likelihood of survival.
 
@@
 significantly different in the pre and post-treatment groups {[}Figure
-S1{]}. It is likely that high inter-individual variation and the patchy
-distribution of OTUs across individuals limited the statistical power of
-the test. To overcome these problems we developed Random Forest models
-to identify collections of OTUs that would allow us to differentiate
-between pre and post-treatment samples from each of the diagnosis
-groups. Because of the relatively small number of subjects in each
-group, we restricted our models to only incorporate 10 OTUs to limit the
-likelihood that the models would overfit the data. Despite this
-restriction, the models performed well (adenoma AUC=0.69 - 0.92,
+S1{]}. It is likely that high inter-individual variation and the
+irregular distribution of OTUs across individuals limited the
+statistical power of the test. To overcome these problems we developed
+Random Forest models to identify collections of OTUs that would allow us
+to differentiate between pre and post-treatment samples from each of the
+diagnosis groups. Because of the relatively small number of subjects in
+each group, we restricted our models to only incorporate 10 OTUs to
+limit the likelihood that the models would overfit the data. Despite
+this restriction, the models performed well (adenoma AUC=0.69 - 0.92,
 advanced adenoma AUC=0.80 - 1.00, carcinoma AUC=0.82 - 0.98).
@@
 
-\textbf{\emph{Post-treatment samples from patients with carcinoma change
-towards a microbiota associated with normal}} Next, we determined
-whether treatment changed the microbiota in a way that the
-post-treatment communities resembled that of patients with normal
-colons. To test this, we used an expanded cohort of 423 individuals that
-were diagnosed under the same protocol as having normal colons or colons
-with adenoma, advanced adenoma, or carcinoma {[}Table 2{]}. We then
-constructed Random Forest models to classify the pre and post-treatment
-samples as having their original diagnosis or having a normal colon. The
+\textbf{\emph{Post-treatment samples from patients with carcinoma more
+closely resemble those of a normal colon.}} Next, we determined whether
+treatment changed the microbiota in a way that the post-treatment
+communities resembled that of patients with normal colons. To test this,
+we used an expanded cohort of 423 individuals that were diagnosed under
+the same protocol as having normal colons or colons with adenoma,
+advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
+Forest models to classify the study samples, with the 3 diagnosis groups
+(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
 models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
@@
 pre and post-treatment samples from patients diagnosed with carcinoma
-significantly decreased with treatment. In fact, 6 of the 26 patients
-(23.08\%) that were initially diagnosed with a carcinoma had a higher
-positive probability after treatment. One of those was re-diagnosed with
-carcinoma on the follow up visit. These results indicate that although
-there are changes in the microbiota associated with treatment, those
-experienced by the patients in the carcinoma group were the only ones
-where the change was directed towards what would be found in a normal
-colon.
+significantly decreased with treatment. Only, 6 of the 26 patients
+(23.08\%) who were diagnosed with a carcinoma had a higher positive
+probability after treatment; one of those was re-diagnosed with
+carcinoma on the follow up visit. These results indicate that, although
+there were changes in the microbiota associated with treatment, those
+experienced by patients with carcinoma after treatment yielded gut
+bacterial communities of greater similarity to that of a normal colon.
 
@@
 0.05). Furthermore, there was not a significant difference in the effect
-of the treatments on the amount of change in the community structure (P=
-1). Finally, the change in the positive probability was not
-significantly different between the two treatment groups (P=1). Due to
-the relatively small number of samples in each treatment group, it was
-difficult to make a definitive statement regarding the specific type of
-treatment on the amount of change in the structure of the microbiota.
+of the treatments on the amount of change in the community structure
+(P-value = 0.298). Finally, the change in the positive probability was
+not significantly different between the two treatment groups (P-value =
+0.999). Due to the relatively small number of samples in each treatment
+group, it was difficult to make a definitive statement regarding the
+specific type of treatment on the amount of change in the structure of
+the microbiota.
 
@@
 This study builds upon previous work that demonstrated a role for the
-microbiota in tumorigenesis and that the microbiota contains biomarkers
-to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
+microbiota in tumorigenesis and that microbiome-derived biomarkers can
+detect colonic lesions {[}6--10,24--28{]}. Our study focused on
 comparing the microbiota of patients diagnosed with adenoma, advanced
 adenoma, and carcinoma before and after treatment. For all three groups
-of patients we observed changes in their microbiota. The microbiota of
-patients diagnosed with carcinoma changed significantly more than the
-other groups. This change resulted in communities that more closely
-resembled those of patients with a normal colon. For the patients
-diagnosed with carcinomas there was a larger change between their
-initial and follow up samples than was observed for patients diagnosed
-with either adenoma or advanced adenoma. The changes in the community
-structure of those diagnosed with carcinoma were similar and resulted in
-communities that more closely resembled those found in patients with
-normal colons.
+of patients, we observed changes in their microbiota. After treatment,
+the microbiota of patients with carcinoma changed significantly more
+than the other groups. This change resulted in communities that more
+closely resembled those of patients with a normal colon. This may
+suggest that treatment for carcinoma is not only successful for removing
+the carcinoma but also at reducing the associated bacterial communities.
+Understanding the effect of treatment on the microbiota of those
+diagnosed with carcinomas may have important implications for reducing
+disease recurrence. It is intriguing that it may be possible to use
+microbiome-based biomarkers to not only predict the presence of lesions
+but to assess the risk of recurrence.
 
-Understanding the effect of treatment on the microbiota of those
-initially diagnosed with carcinomas may have important implications into
-understanding disease recurrence. The microbiota changes associated with
-treatment among patients initially diagnosed with carcinomas resulted in
-community structures that resembled patients with normal colons. This
-would suggest that treatment for carcinoma is not only successful for
-removing the carcinoma but also at removing bacteria associated with
-them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had
-an elevated probability of having carcinomas after treatment. This leads
-us to hypothesize that these individuals may be at a high risk of
-recurrence. Interestingly, the 23.08\% prevalence of increased carcinoma
-probability from our study is within the expected rate of recurrence
-(20-30\% {[}3,4{]}). Unfortunately, because of the design of our study,
-it was not possible to determine whether individuals who had an
-increased positive probability for carcinoma experienced recurrence.
-Regardless, it is intriguing that it may be possible to use
-microbiome-based biomarkers to not only predict the presence of lesions,
-but to also assess the risk of developing them.
+Patients diagnosed with adenoma and advanced adenoma, however, did not
+experience a shift towards a community structure that resembled those
+with normal colons. This may be due to the fundamental differences
+between the features of adenomas and advanced adenomas and carcinoma.
+Specifically, carcinomas may create an inflammatory milieu that would
+impact the structure of the community and removal of that stimulus would
+alter said structure. In addition, it is possible that the difference
+between the microbiota of patients with adenoma and advanced adenoma and
+those with normal colons is subtle. This is supported by the reduced
+ability of our models to correctly classify patients with adenomas and
+advanced adenomas relative to those diagnosed with carcinomas {[}Figure
+S2{]}. Given the irregular distribution of microbiota across patients in
+the different diagnosis groups, it is possible that we lacked the
+statistical power to adequately characterize the change in the
+communities following treatment.
 
-It is interesting that those patients diagnosed with adenoma and
-advanced adenoma did not experience a shift towards a community
-structure that resembled those with normal colons. This may be due to
-the fundamental differences between the features of adenomas and
-advanced adenomas and carcinoma. Specifically, carcinomas may create an
-inflammatory milieu that would impact the structure of the community and
-removal of that stimulus would alter said structure. In addition, it is
-possible that the difference between the microbiota of patients with
-adenoma and advanced adenoma and those with normal colons is subtle.
-This is supported by the reduced ability of our models to correctly
-classify patients with adenomas and advanced adenomas relative to those
-diagnosed with carcinomas {[}Figure S2{]}. Given the patchy distribution
-of microbiota across patients in the different diagnosis groups, it is
-possible that we lacked the statistical power to adequately characterize
-the change in the communities following treatment. A final hypothesis is
-that the specific type of treatment altered the structure of the
-microbiome. The treatment to remove adenomas and advanced adenomas was
-either polyp removal or surgical resection whereas it was surgical
-resection alone or in combination with chemotherapy or with chemotherapy
-and radiation for individuals with carcinoma. Because chemotherapy and
-radiation target rapidly growing cells, these treatments would be more
-likely to cause a turn over of the colonic epithelium driving a more
-significant change in the structure of the microbiota. Although, we were
-able to test for an effect across these specific types of treatment, the
-number of patients in each treatment group was relatively small.
-
-It was not clear why 6 of the 26 patients with carcinoma had a higher
-probability of having carcinomas after treatment. We hypothesized that
-these individuals may have had more severe tumors; however, the tumor
-severity of these 6 individuals (3 with Stage II and 3 with Stage III)
-was similar to the distribution observed among the other 20 patients. We
+There was a subset of patients (6 of the 26 with carcinomas) who
+demonstrated an elevated probability of carcinoma after treatment. This
+may reflect an elevated risk of recurrence. The 23.08\% prevalence of
+increased carcinoma probability from our study is within the expected
+rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
+individuals may have had more severe tumors; however, the tumor severity
+of these 6 individuals (3 with Stage II and 3 with Stage III) was
+similar to the distribution observed among the other 20 patients. We
 also hypothesized that we may have sampled these patients later than the
@@
 
-This study builds upon existing work that has established a role for the
-microbiota in tumorigenesis and that demonstrated the utility of
+Our final hypothesis was that the specific type of treatment altered the
+structure of the microbiome. The treatment to remove adenomas and
+advanced adenomas was either polyp removal or surgical resection whereas
+it was surgical resection alone or in combination with chemotherapy or
+with chemotherapy and radiation for individuals with carcinoma. Because
+chemotherapy and radiation target rapidly growing cells, these
+treatments would be more likely to cause a turnover of the colonic
+epithelium driving a more significant change in the structure of the
+microbiota. Although, we were able to test for an effect across these
+specific types of treatment, the number of patients in each treatment
+group was relatively small.
+
+This study expands upon existing research that has established a role
+for the microbiota in tumorigenesis and that demonstrated the utility of
 microbiome-based biomarkers to predict the presence of colonic lesions.
@@
 possibility that markers within the microbiota could be used to evaluate
-the effect of treatment and predict recurrence for those initially
-diagnosed with carcinoma. If such an approach is effective, it might be
-possible to even target the microbiota as part of combined therapies
-including currently standard approaches. Our data provides additional
-evidence on the importance of the microbiota in tumorigenesis by
-addressing the recovery of the microbiota after treatment and opens
+the effect of treatment and predict recurrence for those diagnosed with
+carcinoma. If such an approach is effective, it might be possible to
+target the microbiota as part of adjuvant therapy. Our data provides
+additional evidence on the importance of the microbiota in tumorigenesis
+by addressing the recovery of the microbiota after treatment and opens
 interesting avenues of research into how these changes may affect
@@
 \textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
-have been previously reported in Baxter, et al {[}23{]}. Briefly, study
+have been previously reported in Baxter, et al {[}24{]}. Briefly, study
 exclusion involved those who had already undergone surgery, radiation,
@@
 community has been shown to normalize back to a pre-colonoscopy
-community within this time period {[}28{]}. Our study cohort consisted
+community within this time period {[}29{]}. Our study cohort consisted
 of 67 individuals with an initial sample as described and a follow up
@@
 \textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
-described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}29{]}. The general workflow using mothur
+previously described {[}30{]}. The general workflow using mothur
 included merging paired-end reads into contigs, filtering for low
-quality contigs, aligning to the SILVA database {[}30{]}, screening for
-chimeras using UCHIME {[}31{]}, classifying with a naive Bayesian
-classifier using the Ribosomal Database Project (RDP){[}32{]}, and
+quality contigs, aligning to the SILVA database {[}31{]}, screening for
+chimeras using UCHIME {[}32{]}, classifying with a naive Bayesian
+classifier using the Ribosomal Database Project (RDP){[}33{]}, and
 clustered into Operational Taxonomic Units (OTUs) using a 97\%
 similarity cutoff with an average neighbor clustering algorithm
-{[}33{]}. The number of sequences for each sample was rarefied to 10523
+{[}34{]}. The number of sequences for each sample was rarefied to 10523
 to minimize the impacts of uneven sampling.
 
-\textbf{\emph{Model Building.}} The Random Forest {[}34{]} algorithm was
+\textbf{\emph{Model Building.}} The Random Forest {[}35{]} algorithm was
 used to create the model used to create the three models used. The
-adenoma model classified normal versus adenoma, advanced adenoma was
-normal versus advanced adenoma, and carcinoma was normal versus
-carcinoma. The toal number of individuals in this data set was 423
-individuals {[}Table 1{]}. There were a total of 239 individuals in the
-adenoma model, 262 individuals in the advanced adenoma model, and 266
-indivdiuals in the carcinoma model {[}Table 1{]}. Each model was then
-applied to our 67-person cohort testing prediction of adenoma intial
-(adenoma n = 22) versus adenoma follow up (adenoma n = 0), advanced
-adenoma initial (advanced adenoma n = 19) versus advanced adenoma follow
-up (advanced adenoma n = 0), carcinoma initial (carinoma n = 26) versus
-carcinoma follow up (carcinoma n = 1).
+adenoma model classified normal versus adenoma, the advanced adenoma
+model classified normal versus advanced adenoma, and the carcinoma model
+classified normal versus carcinoma. The total number of individuals in
+the treatment models was 67 individuals. There were a total of 22
+individuals in the treatment adenoma model, 19 individuals in the
+treatment advanced adenoma model, and 26 individuals in the treatment
+carcinoma model {[}Table 1{]}. The total number of individuals in the
+normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
+were a total of 239 individuals in the adenoma model, 262 individuals in
+the advanced adenoma model, and 266 individuals in the carcinoma model
+{[}Table 2{]}. Each model was then applied to our 67-person cohort
+{[}Table 1{]} which assesed the prediction of pre-treatment adenoma
+(adenoma n = 22) versus post-treatment adenoma (adenoma n = 0),
+pre-treatment advanced adenoma (advanced adenoma n = 19) versus
+post-treatment advanced adenoma (advanced adenoma n = 0), pre-treatment
+carcinoma (carinoma n = 26) versus post-treatment carcinoma (carcinoma n
+= 1).
 
@@
 then tested on the hold out data obtained from the 20\% component. All
-three models had an optimized mtry of 2.
+three models had an optimized mtry of 2. The hyper-parameter, mtry,
+defines the number of variables to investigate at each split before a
+new division of the data was created with the Random Forest model.
 
@@
 the 67-person cohort was completed. For all three models the final
-optimized mtry was 2. The hyper-parameter, mtry, defines the number of
-variables to investigate at each split before a new division of the data
-was created with the Random Forest model.
+optimized mtry was 2. The only difference other than the sample set used
+between treatment models and normal versus diagnosis models was that
+only the top 10 OTUs were used to build each respective treatment model
+by diagnosis group to help avoid model overfitting.T
 
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}35{]} in the vegan package
+community structure utilized PERMANOVA {[}36{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxon ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}37{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 found between the adenoma and carcinoma group for thetayc (P-value =
-NULL). Solid black points represent the median value for each diagnosis
-group. B) NMDS of the pre and post-treatment samples for the adenoma
-group. C) NMDS of the pre and post-treatment samples for the advanced
-adenoma group. D) NMDS of the pre and post-treatment samples for the
-carcinoma group.
+5.36e-05). Solid black points represent the median value for each
+diagnosis group. B) NMDS of the pre and post-treatment samples for the
+adenoma group. C) NMDS of the pre and post-treatment samples for the
+advanced adenoma group. D) NMDS of the pre and post-treatment samples
+for the carcinoma group.
 
@@
 advanced adenoma. C) Positive probability change from initial to follow
-up sample in those with carcinoma..
+up sample in those with carcinoma.
 
@@
 
+\begin{longtable}[]{@{}lccc@{}}
+\toprule
+& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
+\midrule
+\endhead
+n & 22 & 19 & 26\tabularnewline
+Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
+12.9\tabularnewline
+Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
+BMI (Mean ± SD) & 26.86 ± 3.9 & 25.80 ± 4.7 & 28.63 ± 7.2\tabularnewline
+Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
+\bottomrule
+\end{longtable}
+
+\newpage
+
 \textbf{Table 2: Demographic Data of Training Cohort}
+
+\begin{longtable}[]{@{}lcccc@{}}
+\toprule
+& Normal & Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
+\midrule
+\endhead
+n & 172 & 67 & 90 & 94\tabularnewline
+Age (Mean ± SD) & 54.29 ± 9.9 & 63.01 ± 13.1 & 64.07 ± 11.3 & 64.37 ±
+12.9\tabularnewline
+Sex (\%F) & 64.53 & 46.27 & 37.78 & 43.62\tabularnewline
+BMI (Mean ± SD) & 26.96 ± 5.3 & 25.68 ± 4.8 & 26.66 ± 4.9 & 29.27 ±
+6.7\tabularnewline
+Caucasian (\%) & 87.79 & 92.54 & 92.22 & 94.68\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-jemal_cancer_2010}{}
-1. Jemal A, Siegel R, Xu J, Ward E. Cancer statistics, 2010. CA: a
-cancer journal for clinicians. 2010;60:277--300.
+\hypertarget{ref-siegel_cancer_2016}{}
+1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
+journal for clinicians. 2016;66:7--30.
 
@@
 
+\hypertarget{ref-national_cancer_institute_seer_nodate}{}
+5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer
+{[}Internet{]}. {[}cited 2017 Apr 27{]}. Available from:
+\url{http://seer.cancer.gov/statfacts/html/colorect.html}
+
 \hypertarget{ref-goodwin_polyamine_2011}{}
-5. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
+6. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
 TR, et al. Polyamine catabolism contributes to enterotoxigenic
@@
 \hypertarget{ref-abed_fap2_2016}{}
-6. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
+7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
 Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
@@
 \hypertarget{ref-arthur_microbial_2014}{}
-7. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+8. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
 McCafferty J, et al. Microbial genomic analysis reveals the essential
@@
 \hypertarget{ref-kostic_fusobacterium_2013}{}
-8. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
+9. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
 et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
@@
 \hypertarget{ref-wu_human_2009}{}
-9. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
+10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
 human colonic commensal promotes colon tumorigenesis via activation of T
@@
 \hypertarget{ref-zackular_manipulation_2016}{}
-10. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+11. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
 Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
@@
 \hypertarget{ref-zackular_gut_2013}{}
-11. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
+12. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
 GY, et al. The gut microbiome modulates colon tumorigenesis. mBio.
@@
 \hypertarget{ref-baxter_structure_2014}{}
-12. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
+13. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
 microbiome following colonization with human feces determines colonic
@@
 \hypertarget{ref-flynn_metabolic_2016}{}
-13. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+14. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
 Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
@@
 \hypertarget{ref-wang_structural_2012}{}
-14. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
 segregation of gut microbiota between colorectal cancer patients and
@@
 \hypertarget{ref-chen_decreased_2013}{}
-15. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, et al.
+16. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, et al.
 Decreased dietary fiber intake and structural alteration of gut
@@
 \hypertarget{ref-chen_human_2012}{}
-16. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+17. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
 mucosa-associated microbiota in patients with colorectal cancer. PloS
@@
 \hypertarget{ref-shen_molecular_2010}{}
-17. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, et al.
+18. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, et al.
 Molecular characterization of mucosal adherent bacteria and associations
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-18. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+19. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
 al. Genomic analysis identifies association of Fusobacterium with
@@
 \hypertarget{ref-feng_gut_2015}{}
-19. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, et al. Gut
+20. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, et al. Gut
 microbiome development along the colorectal adenoma-carcinoma sequence.
@@
 \hypertarget{ref-dejea_microbiota_2014}{}
-20. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+21. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 \hypertarget{ref-mima_fusobacterium_2015}{}
-21. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, et
+22. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, et
 al. Fusobacterium nucleatum and T Cells in Colorectal Carcinoma. JAMA
@@
 \hypertarget{ref-arthur_intestinal_2012}{}
-22. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+23. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
 T-J, et al. Intestinal inflammation targets cancer-inducing activity of
@@
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-23. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-zeller_potential_2014}{}
-24. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+25. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 \hypertarget{ref-zackular_human_2014}{}
-25. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+26. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-yu_metagenomic_2017}{}
-26. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
+27. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
 analysis of faecal microbiome as a tool towards targeted non-invasive
@@
 \hypertarget{ref-warren_co-occurrence_2013}{}
-27. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
+28. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
 et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
@@
 \hypertarget{ref-obrien_impact_2013}{}
-28. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-29. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+30. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-pruesse_silva_2007}{}
-30. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
+31. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
 SILVA: A comprehensive online resource for quality checked and aligned
@@
 \hypertarget{ref-edgar_uchime_2011}{}
-31. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
+32. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
 sensitivity and speed of chimera detection. Bioinformatics (Oxford,
@@
 \hypertarget{ref-wang_naive_2007}{}
-32. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
+33. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
 for rapid assignment of rRNA sequences into the new bacterial taxonomy.
@@
 \hypertarget{ref-schloss_assessing_2011}{}
-33. Schloss PD, Westcott SL. Assessing and improving methods used in
+34. Schloss PD, Westcott SL. Assessing and improving methods used in
 operational taxonomic unit-based approaches for 16S rRNA gene sequence
@@
 \hypertarget{ref-breiman_random_2001}{}
-34. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+35. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+36. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+37. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-04-29 submission/manuscript.Rmd
niel edits up to results

@@
 Disease mortality has significantly decreased, predominately due to improvements in screening [@haggar_colorectal_2009].
-Despite this improvement, there are still approximately 50,000 CRC-related deaths per year [@siegel_cancer_2016].
+Despite these improvements, there are still approximately 50,000 CRC-related deaths per year in the United States [@siegel_cancer_2016].
 Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 35% of all patients will die within five years [@hellinger_reoperation_2006; @ryuk_predictive_2014; @national_cancer_institute_seer_nodate].
@@
 We found that treatment alters the composition of the gut microbiota and that, for those with carinomas, the gut microbiota shifted more towards that of a normal colon after treatment.
-In the individuals with carcinomas, no difference was found byt the type of treatment (surgery alone versus surgery with chemotherapy).
+In the individuals with carcinomas, no difference was found by the type of treatment (surgery alone versus surgery with chemotherapy).
 Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence and the likelihood of survival.
@@
 Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
-The structure of the microbial communities of the pre and post-treatment samples differed [Figure 1A].
+The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
 We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`).
@@
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
-We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value=`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value=`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value=`r beta_summary["Carcinoma", "PERMANOVA"]`).
+We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value =`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value =`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value =`r beta_summary["Carcinoma", "PERMANOVA"]`).
 Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].
@@
 Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data.
-Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
+Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
@@
 We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
+The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
 The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
-Among the OTUs that were shared across the three models were those populations commonly considered as commensals (e.g.
+Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g.
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing are marked decrease in relative abundance [Figure S4].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.

### 2017-04-29 submission/manuscript.tex
niel edits up to results

@@
 decreased, predominately due to improvements in screening {[}2{]}.
-Despite this improvement, there are still approximately 50,000
-CRC-related deaths per year {[}1{]}. Current estimates indicate that
-20-30\% of those who undergo treatment will experience recurrence and
-35\% of all patients will die within five years {[}3--5{]}.
-Identification of methods to assess patients' risk of recurrence is of
-great importance to reduce mortality and healthcare costs.
+Despite these improvements, there are still approximately 50,000
+CRC-related deaths per year in the United States {[}1{]}. Current
+estimates indicate that 20-30\% of those who undergo treatment will
+experience recurrence and 35\% of all patients will die within five
+years {[}3--5{]}. Identification of methods to assess patients' risk of
+recurrence is of great importance to reduce mortality and healthcare
+costs.
 
@@
 colon after treatment. In the individuals with carcinomas, no difference
-was found byt the type of treatment (surgery alone versus surgery with
+was found by the type of treatment (surgery alone versus surgery with
 chemotherapy). Understanding how the community responds to these
@@
 between pre and post-treatment samples {[}Table 1{]}. The structure of
-the microbial communities of the pre and post-treatment samples differed
-{[}Figure 1A{]}. We found that the communities obtained pre and
+the microbial communities of the pre and post-treatment samples
+differed, as measured by the \(\theta\)\textsubscript{YC} beta diversity
+metric {[}Figure 1A{]}. We found that the communities obtained pre and
 post-treament among the patients with carcinomas changed significantly
@@
 limit the likelihood that the models would overfit the data. Despite
-this restriction, the models performed well (adenoma AUC=0.69 - 0.92,
-advanced adenoma AUC=0.80 - 1.00, carcinoma AUC=0.82 - 0.98).
-Interestingly, the 10 OTUs that were used for each model had little
-overlap with each other {[}Figure 2{]}. These results support the
+this restriction, the models performed well (adenoma AUC range =0.69 -
+0.92, advanced adenoma AUC range =0.80 - 1.00, carcinoma AUC range =0.82
+- 0.98). Interestingly, the 10 OTUs that were used for each model had
+little overlap with each other {[}Figure 2{]}. These results support the
 earlier community-wide analysis where we observed that the treatment had
@@
 (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
-models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
-AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.90; Figure S2). The OTUs that
-were incorporated into the adenoma and advanced adenoma models largely
-overlapped and those OTUs that were used to classify the carcinoma
-samples were largely distinct from those of the other two models
-{[}Figure 3A{]}. Among the OTUs that were shared across the three models
-were those populations commonly considered as commensals (e.g.
-\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
-\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
-3B{]}. Although many of these OTUs were also included in the model
-differentiating between patients with normal colons and those with
-carcinoma, this model also included OTUs affiliated with populations
-that have previously been associated with carcinoma
-(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
-{[}Figure S3{]}. Finally, we applied these three models to the pre and
-post-treatment samples for each diagnosis group and quantified the
-change in the positive probability of the model. A decrease in the
-positive probability would indicate that the colon more closely
-resembled that of a patient with a normal colon. There was no
-significant change in the positive probability for the adenoma or
-advanced adenoma groups {[}Figure 4{]}. The positive probability for the
-pre and post-treatment samples from patients diagnosed with carcinoma
-significantly decreased with treatment. Only, 6 of the 26 patients
-(23.08\%) who were diagnosed with a carcinoma had a higher positive
-probability after treatment; one of those was re-diagnosed with
-carcinoma on the follow up visit. These results indicate that, although
-there were changes in the microbiota associated with treatment, those
-experienced by patients with carcinoma after treatment yielded gut
-bacterial communities of greater similarity to that of a normal colon.
+models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
+AUC range=0.68 - 0.77, carcinoma AUC range=0.84 - 0.90; Figure S2). The
+OTUs that were incorporated into the adenoma and advanced adenoma models
+largely overlapped and those OTUs that were used to classify the
+carcinoma samples were largely distinct from those of the other two
+models {[}Figure 3A{]}. Among the OTUs that were shared across the three
+models were those populations generally considered beneficial to their
+host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
+\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
+\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
+also included in the model differentiating between patients with normal
+colons and those with carcinoma, this model also included OTUs
+affiliated with populations that have previously been associated with
+carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
+\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]}. Finally, we applied
+these three models to the pre and post-treatment samples for each
+diagnosis group and quantified the change in the positive probability of
+the model. A decrease in the positive probability would indicate that
+the colon more closely resembled that of a patient with a normal colon.
+There was no significant change in the positive probability for the
+adenoma or advanced adenoma groups {[}Figure 4{]}. The positive
+probability for the pre and post-treatment samples from patients
+diagnosed with carcinoma significantly decreased with treatment. Only, 6
+of the 26 patients (23.08\%) who were diagnosed with a carcinoma had a
+higher positive probability after treatment; one of those was
+re-diagnosed with carcinoma on the follow up visit. These results
+indicate that, although there were changes in the microbiota associated
+with treatment, those experienced by patients with carcinoma after
+treatment yielded gut bacterial communities of greater similarity to
+that of a normal colon.
 

### 2017-04-29 code/Run_Figure1.R
niel changes

@@ -38,11 +38,12 @@ differences_graph <- grid.arrange(
                        labels = c("0.00", "0.25", "0.50", "0.75", "1.00"),limits = c(-0.06, 1.06)) + 
     xlab("") + theme_bw() + ggtitle("A") + 
     theme(axis.title = element_text(face="bold", hjust = 0.5), 
+          axis.text.x = element_text(size = 12), 
           legend.title = element_text(face="bold"), 
           legend.position = "none", 
-          plot.title = element_text(face="bold", hjust = -0.3, size = 20), 
+          plot.title = element_text(face="bold", hjust = -0.065, size = 20), 
           panel.grid.major = element_blank(), 
-          panel.grid.minor = element_blank()) + coord_fixed(ratio = 1.8), 
+          panel.grid.minor = element_blank()), 
   
   # Adenoma Only initial and follow up
   mutate(thetayc_adn_IF, samples = factor(samples, 
@@ -53,7 +54,7 @@ differences_graph <- grid.arrange(
     theme_bw() + coord_equal() + ggtitle("B") + ylim(-0.7, 0.7) + xlim(-0.7, 0.7) + 
     scale_color_manual(values = c('#00EE76', '#228B22')) + scale_fill_manual(values = c('#00EE76', '#228B22')) + 
     scale_shape_manual(values = c(16, 1)) + 
-    theme(plot.title = element_text(face="bold", hjust = -0.32, size = 20), 
+    theme(plot.title = element_text(face="bold", hjust = -0.28, size = 20), 
           legend.position = c(0.15, 0.11), 
           legend.title = element_blank(), 
           legend.key = element_blank(), 
@@ -101,39 +102,6 @@ differences_graph <- grid.arrange(
 
 
 ggsave(file = "results/figures/Figure1.pdf", differences_graph, 
-       width=11, height = 8.5, dpi = 300)
-
-
-
-
-
-
-
-# Difference from fit between adenoma and cancer
-#ggplot(difference_table_treatment, 
-#       aes(factor(dx, levels = c("adenoma", "cancer"), labels = c("Adenoma", "Carcinoma")), 
-#           fit_difference*-1, group = 1)) + 
-#  geom_jitter(aes(color=Dx_Bin), width = 0.3, size = 4, alpha = 0.5) + 
-#  stat_summary(fun.data = "mean_cl_boot", colour = "black", size = 1) + 
[…]

### 2017-04-29 code/Run_FigureS2.R

@@ -26,7 +26,7 @@ rocs_graph <- grid.arrange(
               size = 1.5, color = '#006400') + 
     scale_x_continuous(trans = "reverse") + theme_bw() + 
     ggtitle("A") + xlab("Sensitivity") + ylab("Specificity") + 
-    theme(plot.title = element_text(face= "bold", hjust = -0.11, size = 20), 
+    theme(plot.title = element_text(face= "bold", hjust = -0.19, size = 20), 
           axis.title = element_text(face = "bold")), 
   
   # SRN ROC curve information
@@ -39,7 +39,7 @@ rocs_graph <- grid.arrange(
               size = 1.5, color = '#EEC900') + 
     scale_x_continuous(trans = "reverse") + theme_bw() +  
     ggtitle("B") + xlab("Sensitivity") + ylab("Specificity") + 
-    theme(plot.title = element_text(face= "bold", hjust = -0.11, size = 20), 
+    theme(plot.title = element_text(face= "bold", hjust = -0.19, size = 20), 
           axis.title = element_text(face = "bold")), 
   
   # CRC ROC curve information
@@ -52,9 +52,9 @@ rocs_graph <- grid.arrange(
               size = 1.5, color = '#DC143C') + 
     scale_x_continuous(trans = "reverse") + theme_bw() +  
     ggtitle("C") + xlab("Sensitivity") + ylab("Specificity") + 
-    theme(plot.title = element_text(face= "bold", hjust = -0.11, size = 20), 
+    theme(plot.title = element_text(face= "bold", hjust = -0.19, size = 20), 
           axis.title = element_text(face = "bold")))
 
 
 ggsave(file = "results/figures/FigureS2.tiff", rocs_graph, 
-       width=6, height = 8, dpi = 300)
+       width=4, height = 8, dpi = 300)

### 2017-04-29 submission/manuscript.Rmd

@@
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
-A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon.
+A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon.
 There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment.
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals.
 Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
@@
 Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1].
-Patients were diagnosed by colonoscopic examination and histopathological review of biopsies.
+Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken.
 Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology.
@@
 DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
-The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.
+The V4 variable region was amplified and the resulting product was split between four sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.
 Each group was randomly assigned to avoid biases based on sample collection location.
@@
 ***Model Building.***
-The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used.
+The Random Forest [@breiman_random_2001] algorithm was used to create the three models used.
 The adenoma model classified normal versus adenoma, the advanced adenoma model classified normal versus advanced adenoma, and the carcinoma model classified normal versus carcinoma.
@@
 There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the carcinoma model [Table 2].
-Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] which assesed the prediction of pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])`), pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])`) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])`), pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))`).
+Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] which assesed the prediction of pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`), pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`).
 
@@
 For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
-The only difference other than the sample set used between treatment models and normal versus diagnosis models was that only the top 10 OTUs were used to build each respective treatment model by diagnosis group to help avoid model overfitting.T
+The only difference other than the sample set used between treatment models and normal versus diagnosis models was that only the top 10 OTUs were used to build each respective treatment model by diagnosis group to help avoid model overfitting.
 
@@
 
-**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
+**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) Thetayc distance from pre versus post sample within each individual.
+A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
 Solid black points represent the median value for each diagnosis group.
@@
 
-**Figure 3: Treatment Response Based on Models Built for Adenoma, Advanced Adenoma, or Carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
+**Figure 3: Common OTUs to All Models.** A) Venn diagram showing the OTU overlap between each model.
+B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
+
+**Figure 4: Treatment Response Based on Models Built for Adenoma, Advanced Adenoma, or Carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
 B) Positive probability change from initial to follow up sample in those with advanced adenoma.
 C) Positive probability change from initial to follow up sample in those with carcinoma.
-
-**Figure 4: Common OTUs to All Models.** A) Venn diagram showing the OTU overlap between each model.
-B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
 

### 2017-04-29 submission/manuscript.tex

@@
 shift in community structure for the patients with carcinoma when using
-a PERMANOVA test (adenoma P-value=0.999, advanced adenoma P-value=0.945,
-and carcinoma P-value=0.005). Finally, we measured the number of
-observed OTUs, Shannon evenness, and Shannon diversity in the pre and
+a PERMANOVA test (adenoma P-value =0.999, advanced adenoma P-value
+=0.945, and carcinoma P-value =0.005). Finally, we measured the number
+of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
 post-treatment samples and did not observe a significant change for any
@@
 limit the likelihood that the models would overfit the data. Despite
-this restriction, the models performed well (adenoma AUC range =0.69 -
-0.92, advanced adenoma AUC range =0.80 - 1.00, carcinoma AUC range =0.82
-- 0.98). Interestingly, the 10 OTUs that were used for each model had
-little overlap with each other {[}Figure 2{]}. These results support the
-earlier community-wide analysis where we observed that the treatment had
-an impact on the overall community structure; however, the effect of
+this restriction, the models performed well (adenoma AUC range = 0.69 -
+0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
+0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
+had little overlap with each other {[}Figure 2{]}. These results support
+the earlier community-wide analysis where we observed that the treatment
+had an impact on the overall community structure; however, the effect of
 treatment was not consistent across patients and diagnosis groups.
@@
 models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
-AUC range=0.68 - 0.77, carcinoma AUC range=0.84 - 0.90; Figure S2). The
-OTUs that were incorporated into the adenoma and advanced adenoma models
-largely overlapped and those OTUs that were used to classify the
+AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S2).
+The OTUs that were incorporated into the adenoma and advanced adenoma
+models largely overlapped and those OTUs that were used to classify the
 carcinoma samples were largely distinct from those of the other two
@@
 carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]}. Finally, we applied
-these three models to the pre and post-treatment samples for each
-diagnosis group and quantified the change in the positive probability of
-the model. A decrease in the positive probability would indicate that
-the colon more closely resembled that of a patient with a normal colon.
-There was no significant change in the positive probability for the
-adenoma or advanced adenoma groups {[}Figure 4{]}. The positive
-probability for the pre and post-treatment samples from patients
-diagnosed with carcinoma significantly decreased with treatment. Only, 6
-of the 26 patients (23.08\%) who were diagnosed with a carcinoma had a
-higher positive probability after treatment; one of those was
-re-diagnosed with carcinoma on the follow up visit. These results
-indicate that, although there were changes in the microbiota associated
-with treatment, those experienced by patients with carcinoma after
-treatment yielded gut bacterial communities of greater similarity to
-that of a normal colon.
+\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]} with some individuals
+showing are marked decrease in relative abundance {[}Figure S4{]}.
+Finally, we applied these three models to the pre and post-treatment
+samples for each diagnosis group and quantified the change in the
+positive probability of the model. A decrease in the positive
+probability would indicate that the microbiota more closely resembled
+that of a patient with a normal colon. There was no significant change
+in the positive probability for the adenoma or advanced adenoma groups
+{[}Figure 4{]}. The positive probability for the pre and post-treatment
+samples from patients diagnosed with carcinoma significantly decreased
+with treatment, suggesting a shift toward a normal microbiota for most
+individuals. Only, 6 of the 26 patients (23.08\%) who were diagnosed
+with a carcinoma had a higher positive probability after treatment; one
+of those was re-diagnosed with carcinoma on the follow up visit. These
+results indicate that, although there were changes in the microbiota
+associated with treatment, those experienced by patients with carcinoma
+after treatment yielded gut bacterial communities of greater similarity
+to that of a normal colon.
 
@@
 {[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
-histopathological review of biopsies. Patients were classified as having
-advanced adenoma if they had an adenoma greater than 1 cm, more than
-three adenomas of any size, or an adenoma with villous histology. This
-study was approved by the University of Michigan Institutional Review
-Board. All study participants provided informed consent and the study
-itself conformed to the guidelines set out by the Helsinki Declaration.
+histopathological review of any biopsies taken. Patients were classified
+as having advanced adenoma if they had an adenoma greater than 1 cm,
+more than three adenomas of any size, or an adenoma with villous
+histology. This study was approved by the University of Michigan
+Institutional Review Board. All study participants provided informed
+consent and the study itself conformed to the guidelines set out by the
+Helsinki Declaration.
 
@@
 automated pipetting system (Eppendorf). The V4 variable region was
-amplified and the resulting product was split between three sequencing
+amplified and the resulting product was split between four sequencing
 runs with normal, adenoma, and carcinoma evenly represented on each run.
@@
 \textbf{\emph{Model Building.}} The Random Forest {[}35{]} algorithm was
-used to create the model used to create the three models used. The
-adenoma model classified normal versus adenoma, the advanced adenoma
-model classified normal versus advanced adenoma, and the carcinoma model
-classified normal versus carcinoma. The total number of individuals in
-the treatment models was 67 individuals. There were a total of 22
-individuals in the treatment adenoma model, 19 individuals in the
-treatment advanced adenoma model, and 26 individuals in the treatment
-carcinoma model {[}Table 1{]}. The total number of individuals in the
-normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
-were a total of 239 individuals in the adenoma model, 262 individuals in
-the advanced adenoma model, and 266 individuals in the carcinoma model
-{[}Table 2{]}. Each model was then applied to our 67-person cohort
-{[}Table 1{]} which assesed the prediction of pre-treatment adenoma
-(adenoma n = 22) versus post-treatment adenoma (adenoma n = 0),
-pre-treatment advanced adenoma (advanced adenoma n = 19) versus
-post-treatment advanced adenoma (advanced adenoma n = 0), pre-treatment
-carcinoma (carinoma n = 26) versus post-treatment carcinoma (carcinoma n
-= 1).
+used to create the three models used. The adenoma model classified
+normal versus adenoma, the advanced adenoma model classified normal
+versus advanced adenoma, and the carcinoma model classified normal
+versus carcinoma. The total number of individuals in the treatment
+models was 67 individuals. There were a total of 22 individuals in the
+treatment adenoma model, 19 individuals in the treatment advanced
+adenoma model, and 26 individuals in the treatment carcinoma model
+{[}Table 1{]}. The total number of individuals in the normal versus
+diagnosis models was 423 individuals {[}Table 2{]}. There were a total
+of 239 individuals in the adenoma model, 262 individuals in the advanced
+adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
+Each model was then applied to our 67-person cohort {[}Table 1{]} which
+assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
+disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
+disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
+= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
+(advanced adenoma n = 0 and disease free n = 19), pre-treatment
+carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
+carcinoma (carcinoma n = 1 and disease free n = 25).
 
@@
 only the top 10 OTUs were used to build each respective treatment model
-by diagnosis group to help avoid model overfitting.T
+by diagnosis group to help avoid model overfitting.
 
@@
 \textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
-and Carcinoma Groups After Treatment.} A) A significant difference was
+and Carcinoma Groups After Treatment.} A) Thetayc distance from pre
+versus post sample within each individual. A significant difference was
 found between the adenoma and carcinoma group for thetayc (P-value =
@@
 
-\textbf{Figure 3: Treatment Response Based on Models Built for Adenoma,
+\textbf{Figure 3: Common OTUs to All Models.} A) Venn diagram showing
+the OTU overlap between each model. B) For each common OTU the lowest
+taxonomic identification and importance rank for each model run is
+shown.
+
+\textbf{Figure 4: Treatment Response Based on Models Built for Adenoma,
 Advanced Adenoma, or Carcinoma.} A) Positive probability change from
@@
 up sample in those with carcinoma.
-
-\textbf{Figure 4: Common OTUs to All Models.} A) Venn diagram showing
-the OTU overlap between each model. B) For each common OTU the lowest
-taxonomic identification and importance rank for each model run is
-shown.
 

### 2017-05-03 code/Run_FigureS1.R
remove histogram of pvalue graph

@@ -1,5 +1,5 @@
 ### Figure S1
-### Distribution of P-values from Paired Wilcoxson Analysis
+### ROC Curves summary for three models used
 ## Marc Sze
 
 #Load needed code and packages
@@ -8,17 +8,53 @@ source('code/functions.R')
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
 #Read data needed
-paired_table <- read.csv("data/process/tables/OTU_paired_wilcoxson_test.csv", header = T, stringsAsFactors = F)
+adn_model_roc <- read.csv("data/process/tables/adn_reduced_test_data_roc.csv", header = T)
+srn_model_roc <- read.csv("data/process/tables/srn_reduced_test_data_roc.csv", header = T)
+crc_model_roc <- read.csv("data/process/tables/crc_reduced_test_data_roc.csv", header = T)
 
-label_names <- c("Adenoma", "Advanced Adenoma", "Carcinoma")
-names(label_names) <- c("adn", "srn", "crc")
 
-pvalue_distribution_paired <- ggplot(paired_table, aes(x=BH_corrected)) + geom_histogram() + 
-  geom_vline(aes(xintercept = 0.05, color = "red"), linetype = 2, size = 1) + 
-  facet_wrap(~factor(analysis, levels = c("adn", "srn", "crc")), 
-             labeller = as_labeller(label_names)) + theme_bw() + 
-  xlab("Benjamini-Hochberg Corrected P-values") + ylab("Counts") + 
-  theme(legend.position = "none")
+# Create the graph
 
-ggsave(file = "results/figures/FigureS1.pdf", pvalue_distribution_paired, 
-       width=12, height = 6, dpi = 300)
+rocs_graph <- grid.arrange(
+  # Adenoma ROC curve information
+  filter(adn_model_roc, run != "middle_roc" & run != "full_roc") %>% 
+    ggplot(aes(x = sensitivities, y = specificities)) + 
+    geom_polygon(data = filter(adn_model_roc, run != "full_roc"), alpha = 0.5, fill = '#76EE00') + 
+    geom_line(data = filter(adn_model_roc, run != "full_roc"), 
+              aes(group = run), size = 1.25, color = '#76EE00') + 
+    geom_line(data = filter(adn_model_roc, run == "full_roc"), 
+              size = 1.5, color = '#006400') + 
[…]

### 2017-05-03 code/Run_FigureS2.R

@@ -1,5 +1,5 @@
 ### Figure S2
-### ROC Curves summary for three models used
+### Summary of Important OTUs within the model and their MDAs
 ## Marc Sze
 
 #Load needed code and packages
@@ -8,53 +8,154 @@ source('code/functions.R')
 loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
 
 #Read data needed
-adn_model_roc <- read.csv("data/process/tables/adn_reduced_test_data_roc.csv", header = T)
-srn_model_roc <- read.csv("data/process/tables/srn_reduced_test_data_roc.csv", header = T)
-crc_model_roc <- read.csv("data/process/tables/crc_reduced_test_data_roc.csv", header = T)
+adn_MDA_data_summary <- read.csv("data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv", 
+                                    header = T, stringsAsFactors = F)
+
+adn_MDA_full <- read.csv("data/process/tables/adn_reduced_lesion_model_top_vars_MDA.csv", 
+                            header = T, stringsAsFactors = F)
+
+srn_MDA_data_summary <- read.csv("data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv", 
+                                 header = T, stringsAsFactors = F)
+
+srn_MDA_full <- read.csv("data/process/tables/srn_reduced_model_top_vars_MDA.csv", 
+                         header = T, stringsAsFactors = F)
+
+crc_MDA_data_summary <- read.csv("data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv", 
+                                 header = T, stringsAsFactors = F)
+
+crc_MDA_full <- read.csv("data/process/tables/crc_reduced_lesion_model_top_vars_MDA.csv", 
+                         header = T, stringsAsFactors = F)
 
 
-# Create the graph
+tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
 
-rocs_graph <- grid.arrange(
-  # Adenoma ROC curve information
-  filter(adn_model_roc, run != "middle_roc" & run != "full_roc") %>% 
-    ggplot(aes(x = sensitivities, y = specificities)) + 
[…]

### 2017-05-03 code/Run_FigureS3.R

@@ -1,161 +1,88 @@
-### Figure S3
-### Summary of Important OTUs within the model and their MDAs
+### Create Figure S3 graph
+### Show results for initial and follow up samples based on 3 specific OTUs
 ## Marc Sze
 
-#Load needed code and packages
+
+#Load needed libraries
 source('code/functions.R')
 
-loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
-
-#Read data needed
-adn_MDA_data_summary <- read.csv("data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv", 
-                                    header = T, stringsAsFactors = F)
-
-adn_MDA_full <- read.csv("data/process/tables/adn_reduced_lesion_model_top_vars_MDA.csv", 
-                            header = T, stringsAsFactors = F)
-
-srn_MDA_data_summary <- read.csv("data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv", 
-                                 header = T, stringsAsFactors = F)
-
-srn_MDA_full <- read.csv("data/process/tables/srn_reduced_model_top_vars_MDA.csv", 
-                         header = T, stringsAsFactors = F)
-
-crc_MDA_data_summary <- read.csv("data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv", 
-                                 header = T, stringsAsFactors = F)
-
-crc_MDA_full <- read.csv("data/process/tables/crc_reduced_lesion_model_top_vars_MDA.csv", 
-                         header = T, stringsAsFactors = F)
+loadLibs(c("dplyr", "tidyr", "ggplot2"))
 
 
-tax <- read.delim('data/process/final.taxonomy', sep='\t', header=T, row.names=1)
+# Read in necessary data frames
+shared <- read.delim('data/process/final.shared', 
+                     header=T, sep='\t') %>% select(Group, contains("Otu0"))
+         
+         
+good_metaf <- read.csv('data/process/mod_metadata/good_metaf_final.csv', 
+                       header = T, stringsAsFactors = F)
 
-# Convert taxa table to a data frame with columns for each taxonomic division
[…]

### 2017-05-03 code/Run_FigureS4.R

@@ -1,88 +1,0 @@
-### Create Figure S4 graph
-### Show results for initial and follow up samples based on 3 specific OTUs
-## Marc Sze
-
-
-#Load needed libraries
-source('code/functions.R')
-
-loadLibs(c("dplyr", "tidyr", "ggplot2"))
-
-
-# Read in necessary data frames
-shared <- read.delim('data/process/final.shared', 
-                     header=T, sep='\t') %>% select(Group, contains("Otu0"))
-         
-         
-good_metaf <- read.csv('data/process/mod_metadata/good_metaf_final.csv', 
-                       header = T, stringsAsFactors = F)
-
-reduced_model <- read.csv('data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', 
-                          header = T, stringsAsFactors = F)
-
-crc_tax <- read.csv('data/process/tables/crc_rf_otu_tax.csv', 
-                    header = T, stringsAsFactors = F) %>% rename(otu = X)
-
-# select out only necessary groups
-test <- shared %>% slice(match(c(good_metaf$initial, good_metaf$followUp), Group))
-
-# Get Relative Abundance
-total_sub_Seqs <- test %>% select(contains("Otu0")) %>% rowSums()
-
-test2 <- as.data.frame(apply(select(test, -Group), 2, function(x) x/total_sub_Seqs)) %>% 
-  mutate(Group = test$Group)
-
-# Select out for only three specific OTUs 
-otus <- filter(crc_tax, Genus == "Porphyromonas" | Genus == "Parvimonas" | Genus == "Fusobacterium")[, "otu"]
-
-# Filter shared by these specific otus
-test2 <- test2 %>% select(Group, one_of(otus))
-
-# Create modified data table
-data_analysis <- data_frame(Group = c(good_metaf$initial, good_metaf$followUp), 
-                            EDRN = rep(good_metaf$EDRN, 2), 
-                            sampleType = c(rep("initial", length(good_metaf$EDRN)), rep("followup", length(good_metaf$EDRN))), 
-                            Dx_Bin = rep(good_metaf$Dx_Bin, 2), 
-                            dx = rep(good_metaf$dx, 2)) %>% 
-  inner_join(test2, by = "Group") %>% 
-  gather(key = otu, value = rel.abund, -EDRN, -Group, -sampleType, -Dx_Bin, -dx)
-
-
[…]

### 2017-05-03 code/old/Run_FigureS1.R

@@ -1,0 +1,24 @@
+### Figure S1
+### Distribution of P-values from Paired Wilcoxson Analysis
+## Marc Sze
+
+#Load needed code and packages
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))
+
+#Read data needed
+paired_table <- read.csv("data/process/tables/OTU_paired_wilcoxson_test.csv", header = T, stringsAsFactors = F)
+
+label_names <- c("Adenoma", "Advanced Adenoma", "Carcinoma")
+names(label_names) <- c("adn", "srn", "crc")
+
+pvalue_distribution_paired <- ggplot(paired_table, aes(x=BH_corrected)) + geom_histogram() + 
+  geom_vline(aes(xintercept = 0.05, color = "red"), linetype = 2, size = 1) + 
+  facet_wrap(~factor(analysis, levels = c("adn", "srn", "crc")), 
+             labeller = as_labeller(label_names)) + theme_bw() + 
+  xlab("Benjamini-Hochberg Corrected P-values") + ylab("Counts") + 
+  theme(legend.position = "none")
+
+ggsave(file = "results/figures/FigureS1.pdf", pvalue_distribution_paired, 
+       width=12, height = 6, dpi = 300)

### 2017-05-03 submission/manuscript.Rmd
update manuscirpt with new supplemental figs

@@
 First, we sought to identify individual OTUs that could account for the change in overall community structure.
-However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1].
+However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(all_OTU_pair_wilcox > 0.05, invisible(">"), invisible("<"))` 0.05).
 It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
@@
 We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
+The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1).
 The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
@@
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing are marked decrease in relative abundance [Figure S4].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing are marked decrease in relative abundance [Figure S3].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
@@
 In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
-This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S2].
+This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1].
 Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
@@
 
-**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of All OTUs Before and After Treatment**
-
-**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S3: Summary of Important OTUs for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model.
+**Figure S2: Summary of Important OTUs for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
@@
 The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.
+
+**Figure S3: Pre and Post Treatment Relative Abundance of CRC associated OTUs within the Carcinoma Model.**
 

### 2017-05-03 submission/manuscript.tex

@@
 Wilcoxon test we were unable to identify any OTUs that were
-significantly different in the pre and post-treatment groups {[}Figure
-S1{]}. It is likely that high inter-individual variation and the
-irregular distribution of OTUs across individuals limited the
-statistical power of the test. To overcome these problems we developed
-Random Forest models to identify collections of OTUs that would allow us
-to differentiate between pre and post-treatment samples from each of the
-diagnosis groups. Because of the relatively small number of subjects in
-each group, we restricted our models to only incorporate 10 OTUs to
-limit the likelihood that the models would overfit the data. Despite
-this restriction, the models performed well (adenoma AUC range = 0.69 -
-0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
-0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
-had little overlap with each other {[}Figure 2{]}. These results support
-the earlier community-wide analysis where we observed that the treatment
-had an impact on the overall community structure; however, the effect of
-treatment was not consistent across patients and diagnosis groups.
+significantly different in the pre and post-treatment groups (P-value
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, NA, NA, NA, NA, NA, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{}, NA, NA,
+NA, \textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textless{}, \textless{}, \textless{},
+\textless{}, \textless{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{}, NA, NA,
+NA, NA, NA, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{}, NA, NA,
+NA, NA, NA, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, NA, NA, NA, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, \textgreater{}, \textgreater{},
+\textgreater{}, \textgreater{}, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
+NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA 0.05). It is likely
+that high inter-individual variation and the irregular distribution of
+OTUs across individuals limited the statistical power of the test. To
+overcome these problems we developed Random Forest models to identify
+collections of OTUs that would allow us to differentiate between pre and
+post-treatment samples from each of the diagnosis groups. Because of the
+relatively small number of subjects in each group, we restricted our
+models to only incorporate 10 OTUs to limit the likelihood that the
+models would overfit the data. Despite this restriction, the models
+performed well (adenoma AUC range = 0.69 - 0.92, advanced adenoma AUC
+range = 0.80 - 1.00, carcinoma AUC range = 0.82 - 0.98). Interestingly,
+the 10 OTUs that were used for each model had little overlap with each
+other {[}Figure 2{]}. These results support the earlier community-wide
+analysis where we observed that the treatment had an impact on the
+overall community structure; however, the effect of treatment was not
+consistent across patients and diagnosis groups.
 
@@
 models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
-AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S2).
+AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
 The OTUs that were incorporated into the adenoma and advanced adenoma
@@
 carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]} with some individuals
-showing are marked decrease in relative abundance {[}Figure S4{]}.
+\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
+showing are marked decrease in relative abundance {[}Figure S3{]}.
 Finally, we applied these three models to the pre and post-treatment
@@
 advanced adenomas relative to those diagnosed with carcinomas {[}Figure
-S2{]}. Given the irregular distribution of microbiota across patients in
+S1{]}. Given the irregular distribution of microbiota across patients in
 the different diagnosis groups, it is possible that we lacked the
@@
 
-\textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
-Analysis of All OTUs Before and After Treatment}
-
-\textbf{Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and
+\textbf{Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and
 Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
@@
 
-\textbf{Figure S3: Summary of Important OTUs for the Adenoma, Advanced
+\textbf{Figure S2: Summary of Important OTUs for the Adenoma, Advanced
 Adenoma, and Carcinoma Models.} A) MDA of the most important variables
@@
 
+\textbf{Figure S3: Pre and Post Treatment Relative Abundance of CRC
+associated OTUs within the Carcinoma Model.}
+
 \newpage

### 2017-05-03 submission/manuscript.Rmd
put together Mack's changes

@@
 They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions.
-This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease.
+This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012;@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].
 
 Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota.
-If the community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence.
+If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence.
 To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner?
@@
 First, we sought to identify individual OTUs that could account for the change in overall community structure.
-However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(all_OTU_pair_wilcox > 0.05, invisible(">"), invisible("<"))` 0.05).
+However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected) > 0.05, invisible(">"), invisible("<"))` 0.05).
 It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
@@
 
-This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that microbiome-derived biomarkers can detect colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].
 Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
@@
 We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
-We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased or decreased (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`).
+We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` days) or decreased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`).
 Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
@@
 Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.
+Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.
+However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(mean(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` $\pm$ `r format(sd(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` days).
 
@@
 ```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
-test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White)
+test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time)
 
-t2 <- as.data.frame(matrix(nrow = 5, ncol = 3, 
+t2 <- as.data.frame(matrix(nrow = 6, ncol = 3, 
                            dimnames = 
-                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
+                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time"), 
                                   coln = c("Adenoma", "SRN", "Carcinoma"))))
@@
   t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
+
+  t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))
+
 }
@@
 BMI_sd <- c()
+post_time_sd <- c()
 
@@
   BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
+  post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])
+
 }
@@
                                                     format(BMI_sd, digits = 2, nmsall = 2)), 
-                                        White = format(White, digits =2, nsmall = 2))
+                                        White = format(White, digits =2, nsmall = 2), 
+                                        post_time = paste(format(post_time, digits = 2, nsmall = 2), 
+                                                    label = "\u00B1", 
+                                                    format(post_time_sd, digits = 2, nmsall = 2)))
 
 ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
-             ttemptable$BMI, ttemptable$White)
+             ttemptable$BMI, ttemptable$White, ttemptable$post_time)
 
@@
 bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
+post_time_label <- paste("Post-Treatment Sample (Mean ", label = "\u00B1", " SD)", sep = "")
 
-colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)")
+colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label)
 

### 2017-05-03 submission/manuscript.tex

--- 2017-04-29 submission/manuscript.tex
@@
 tumorigenesis with potential as useful biomarkers for aiding in the
-early detection of disease.
+early detection of disease {[}21--26{]}.
 
@@
 including resection, chemotherapy, and/or radiation affect the
-composition of the gut microbiota. If the community drives tumorigenesis
-then one would hypothesize that treatment to remove a lesion would
-affect the microbiota and risk of recurrence. To test this hypothesis,
-we addressed two related questions: Does treatment affect the colonic
-microbiota in a predictable manner? If so, does the treatment alter the
-community to more closely resemble that of individuals with normal
-colons?
+composition of the gut microbiota. If the microbial community drives
+tumorigenesis then one would hypothesize that treatment to remove a
+lesion would affect the microbiota and risk of recurrence. To test this
+hypothesis, we addressed two related questions: Does treatment affect
+the colonic microbiota in a predictable manner? If so, does the
+treatment alter the community to more closely resemble that of
+individuals with normal colons?
 
@@
 Wilcoxon test we were unable to identify any OTUs that were
-significantly different in the pre and post-treatment groups {[}Figure
-S1{]}. It is likely that high inter-individual variation and the
+significantly different in the pre and post-treatment groups (P-value NA
+0.05). It is likely that high inter-individual variation and the
 irregular distribution of OTUs across individuals limited the
@@
 
-This study builds upon previous work that demonstrated a role for the
-microbiota in tumorigenesis and that microbiome-derived biomarkers can
-detect colonic lesions {[}6--10,24--28{]}. Our study focused on
-comparing the microbiota of patients diagnosed with adenoma, advanced
-adenoma, and carcinoma before and after treatment. For all three groups
-of patients, we observed changes in their microbiota. After treatment,
-the microbiota of patients with carcinoma changed significantly more
-than the other groups. This change resulted in communities that more
-closely resembled those of patients with a normal colon. This may
-suggest that treatment for carcinoma is not only successful for removing
-the carcinoma but also at reducing the associated bacterial communities.
-Understanding the effect of treatment on the microbiota of those
-diagnosed with carcinomas may have important implications for reducing
-disease recurrence. It is intriguing that it may be possible to use
-microbiome-based biomarkers to not only predict the presence of lesions
-but to assess the risk of recurrence.
+Our study focused on comparing the microbiota of patients diagnosed with
+adenoma, advanced adenoma, and carcinoma before and after treatment. For
+all three groups of patients, we observed changes in their microbiota.
+After treatment, the microbiota of patients with carcinoma changed
+significantly more than the other groups. This change resulted in
+communities that more closely resembled those of patients with a normal
+colon. This may suggest that treatment for carcinoma is not only
+successful for removing the carcinoma but also at reducing the
+associated bacterial communities. Understanding the effect of treatment
+on the microbiota of those diagnosed with carcinomas may have important
+implications for reducing disease recurrence. It is intriguing that it
+may be possible to use microbiome-based biomarkers to not only predict
+the presence of lesions but to assess the risk of recurrence.
 
@@
 the length of time between sample collection among those whose
-probabilities increased or decreased (Wilcoxon Test; P-value = 0.56).
-Finally, it is possible that these patients may not have responded to
-treatment as well as the other 20 patients diagnosed with carcinoma and
-so the microbiota may not have been impacted the same way. Again,
-further studies looking at the role of the microbiota in recurrence are
-needed to understand the dynamics following treatment.
+probabilities increased (382 \(\pm\) 116 days) or decreased (341 \(\pm\)
+98 days) (Wilcoxon Test; P-value = 0.56). Finally, it is possible that
+these patients may not have responded to treatment as well as the other
+20 patients diagnosed with carcinoma and so the microbiota may not have
+been impacted the same way. Again, further studies looking at the role
+of the microbiota in recurrence are needed to understand the dynamics
+following treatment.
 
@@
 specific types of treatment, the number of patients in each treatment
-group was relatively small.
+group was relatively small. Finally, those undergoing surgery would have
+received antibiotics and this may be a potential confounder. However,
+our pre-treatment stool samples were obtained before the surgery and the
+post-treatment samples were obtained long after any effects due to
+antibiotic administration on the microbiome would be expected to occur
+(351 \(\pm\) 102 days).
 
@@
 community has been shown to normalize back to a pre-colonoscopy
-community within this time period {[}29{]}. Our study cohort consisted
+community within this time period {[}27{]}. Our study cohort consisted
 of 67 individuals with an initial sample as described and a follow up
@@
 \textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
-described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}30{]}. The general workflow using mothur
+previously described {[}28{]}. The general workflow using mothur
 included merging paired-end reads into contigs, filtering for low
-quality contigs, aligning to the SILVA database {[}31{]}, screening for
-chimeras using UCHIME {[}32{]}, classifying with a naive Bayesian
-classifier using the Ribosomal Database Project (RDP){[}33{]}, and
+quality contigs, aligning to the SILVA database {[}29{]}, screening for
+chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
+classifier using the Ribosomal Database Project (RDP){[}31{]}, and
 clustered into Operational Taxonomic Units (OTUs) using a 97\%
 similarity cutoff with an average neighbor clustering algorithm
-{[}34{]}. The number of sequences for each sample was rarefied to 10523
+{[}32{]}. The number of sequences for each sample was rarefied to 10523
 to minimize the impacts of uneven sampling.
 
-\textbf{\emph{Model Building.}} The Random Forest {[}35{]} algorithm was
+\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
 used to create the three models used. The adenoma model classified
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}36{]} in the vegan package
+community structure utilized PERMANOVA {[}34{]} in the vegan package
 (v2.4.1). Comparisons between probabilities as well as overall OTU
@@
 Wilcoxon ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}37{]}
+appropriate, a Benjamini-Hochberg (BH) correction was applied {[}35{]}
 and a corrected P-value of less than 0.05 was considered significant.
@@
 Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
+Post-Treatment Sample (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 & 350.85 ±
+102\tabularnewline
 \bottomrule
@@
 red points are the value of each of the 100 different runs.
+
+\textbf{Figure S3: Pre and Post Treatment Relative Abundance of CRC
+associated OTUs within the Carcinoma Model.}
 
@@
 
-\hypertarget{ref-yu_metagenomic_2017}{}
-27. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
-analysis of faecal microbiome as a tool towards targeted non-invasive
-biomarkers for colorectal cancer. Gut. 2017;66:70--8.
-
-\hypertarget{ref-warren_co-occurrence_2013}{}
-28. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
-et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
-Microbiome. 2013;1:16.
-
 \hypertarget{ref-obrien_impact_2013}{}
-29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+27. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-30. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-pruesse_silva_2007}{}
-31. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
+29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
 SILVA: A comprehensive online resource for quality checked and aligned
@@
 \hypertarget{ref-edgar_uchime_2011}{}
-32. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
+30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
 sensitivity and speed of chimera detection. Bioinformatics (Oxford,
@@
 \hypertarget{ref-wang_naive_2007}{}
-33. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
+31. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
 for rapid assignment of rRNA sequences into the new bacterial taxonomy.
@@
 \hypertarget{ref-schloss_assessing_2011}{}
-34. Schloss PD, Westcott SL. Assessing and improving methods used in
+32. Schloss PD, Westcott SL. Assessing and improving methods used in
 operational taxonomic unit-based approaches for 16S rRNA gene sequence
@@
 \hypertarget{ref-breiman_random_2001}{}
-35. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
+33. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
 {[}cited 2013 Feb 7{]};45:5--32. Available from:
@@
 \hypertarget{ref-anderson_permanova_2013}{}
-36. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+34. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-37. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+35. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-05-04 submission/manuscript.Rmd
final changes to table 1 with day info

@@
 bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
-post_time_label <- paste("Post-Treatment Sample (Mean ", label = "\u00B1", " SD)", sep = "")
+post_time_label <- paste("Days Between Colonoscopy (Mean ", label = "\u00B1", " SD)", sep = "")
 

### 2017-05-04 submission/manuscript.tex

@@
 Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
-Post-Treatment Sample (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 & 350.85 ±
-102\tabularnewline
+Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
+350.85 ± 102\tabularnewline
 \bottomrule

### 2017-05-13 submission/manuscript.Rmd
Schloss edits

@@
 
-Marc A Sze${^1}$, Nielson T Baxter${^2}$, Mack T Ruffin IV${^3}$, Mary AM Rogers${^2}$, and Patrick D Schloss${^1}$${^\dagger}$
+Marc A Sze${^1}$, Nielson T Baxter${^1,2}$, Mack T Ruffin IV${^3}$, Mary AM Rogers${^2}$, and Patrick D Schloss${^1}$${^\dagger}$
 
@@
 
-**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiota will yield improvements in disease management.
+**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
 
@@
 \newpage
-\linenumbers
 
@@
 We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre and post-treatment.
-We used classical community analysis to compare the alpha and beta-diversity of communities pre and post treatment.
+We used classical community analysis to compare the alpha and beta-diversity of communities pre and post-treatment.
 Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group.
@@
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
-We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`).
+We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`) [Comment: this should be formatted to be "P < 0.001"].
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
@@
 First, we sought to identify individual OTUs that could account for the change in overall community structure.
-However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected) > 0.05, invisible(">"), invisible("<"))` 0.05).
+However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected) > 0.05, invisible(">"), invisible("<"))` 0.05) [Comment: This is rendered weirdly in the PDF].
 It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
 To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data.
+To limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
 Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
@@
 Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter said structure.
-In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
+It is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
 This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1].
@@
 We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
-We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` days) or decreased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`).
+We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` days) or decreased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) [Comment: The value of what the plus/minus refers to is unclear.
+Because these are not normally distributed, you should make it clear that the numbers are the median and you should provide the IQR rather than the mean and standard deviation].
 Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
@@
 Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.
-However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(mean(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` $\pm$ `r format(sd(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` days).
+However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(mean(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` $\pm$ `r format(sd(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` days) [Comment: The value of what the plus/minus refers to is unclear.
+Because these are not normally distributed, you should make it clear that the numbers are the median and you should provide the IQR rather than the mean and standard deviation].
 
@@
 
-***Model Building.***
+***Model Building.*** [Comment: This entire section is pretty confusing. In the first paragraph it jumbles the treatment models and the diagnosis models]
 The Random Forest [@breiman_random_2001] algorithm was used to create the three models used.
@@
 
-The model included only OTU data obtained from 16S rRNA sequencing.
-Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.
-This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).
+The model included only OTU data obtained from 16S rRNA sequencing and was constructed using the R package caret (v`r paste(packageVersion("caret"))`).
 Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where normal and adenoma, normal and advanced adenoma, or normal and carcinoma were represented in the same proportion within both the whole data set and the 80/20 split.
@@
 Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
-Comparisons between probabilities as well as overall OTU differences between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test.
+Comparisons between probabilities as well as overall differences in the median relative abundance of each OTU between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test.
 Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
-Unless otherwise stated the P-values reported are those that were BH corrected.
+The P-values reported are those that were BH corrected.
 Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.
@@
 
-**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) Thetayc distance from pre versus post sample within each individual.
+**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) Thetayc distance from pre versus post sample within each individual.
 A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
@@
 
-**Figure 2: The 10 OTUs used to classify treatment for Adenoma, Advanced Adenoma, and Carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs.
+**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs.
 
-**Figure 3: Common OTUs to All Models.** A) Venn diagram showing the OTU overlap between each model.
+**Figure 3: OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
 B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
 
-**Figure 4: Treatment Response Based on Models Built for Adenoma, Advanced Adenoma, or Carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
+**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
 B) Positive probability change from initial to follow up sample in those with advanced adenoma.
@@
 
-**Table 1: Demographic Data of pre and Post Treatment Cohort**
+**Table 1: Demographic data of patients in the pre and post-treatment cohort**
 
@@
 
-**Table 2: Demographic Data of Training Cohort**
+**Table 2: Demographic data of training cohort**
 
@@
 
-**Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S1: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S2: Summary of Important OTUs for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model.
+**Figure S2: Summary of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
@@
 
-**Figure S3: Pre and Post Treatment Relative Abundance of CRC associated OTUs within the Carcinoma Model.**
+**Figure S3: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**
 

### 2017-05-13 submission/manuscript.Rmd
fix R code comments

@@
 This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
-We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), and carcinoma (N=`r change_summary["thetayc", "crc_n"]`).
+We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).
 
@@
 ***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
+Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
-We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`) [Comment: this should be formatted to be "P < 0.001"].
+We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001).
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
-We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value =`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value =`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value =`r beta_summary["Carcinoma", "PERMANOVA"]`).
+We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`).
 Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].
@@
 First, we sought to identify individual OTUs that could account for the change in overall community structure.
-However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected) > 0.05, invisible(">"), invisible("<"))` 0.05) [Comment: This is rendered weirdly in the PDF].
+However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05).
 It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
@@
 We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
-We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` days) or decreased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) [Comment: The value of what the plus/minus refers to is unclear.
-Because these are not normally distributed, you should make it clear that the numbers are the median and you should provide the IQR rather than the mean and standard deviation].
+We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` (`r format(IQR((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)`) days) or decreased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` (`r format(IQR((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)`) days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) (all days data displayed as median (IQR)).
 Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
@@
 Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.
-However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(mean(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` $\pm$ `r format(sd(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` days) [Comment: The value of what the plus/minus refers to is unclear.
-Because these are not normally distributed, you should make it clear that the numbers are the median and you should provide the IQR rather than the mean and standard deviation].
+However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(IQR(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)`) days).
 

### 2017-05-13 submission/manuscript.tex
fix R code comments

@@
 
-\textbf{Conclusions.} Further exploration of the relationship between
-diagnosis, treatment, and the impact on the microbiota will yield
-improvements in disease management.
+\textbf{Conclusions.} By better understanding the response of the
+microbiota to treatment for adenomas and carcinomas, it is likely that
+biomarkers will be validated that can be used to quantify the risk of
+recurrence and the likelihood of survival.
 
@@
 \newpage
-
-\linenumbers
 
@@
 community analysis to compare the alpha and beta-diversity of
-communities pre and post treatment. Next, we generated Random Forest
+communities pre and post-treatment. Next, we generated Random Forest
 models to identify bacterial populations that were indicative of
@@
 patients diagnosed with colonic lesions.}} Within our 67-person cohort
-we tested whether the microbiota of patients with adenoma (N=22),
-advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
-between pre and post-treatment samples {[}Table 1{]}. The structure of
-the microbial communities of the pre and post-treatment samples
-differed, as measured by the \(\theta\)\textsubscript{YC} beta diversity
-metric {[}Figure 1A{]}. We found that the communities obtained pre and
-post-treament among the patients with carcinomas changed significantly
-more than those patients with adenoma (P-value = 5.4e-05). There were no
-significant differences in the amount of change observed between the
-patients with adenoma and advanced adenoma or between the patients with
-advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next, we
-tested whether there was a consistent direction in the change in the
-community structure between the pre and post-treatment samples for each
-of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
-shift in community structure for the patients with carcinoma when using
-a PERMANOVA test (adenoma P-value =0.999, advanced adenoma P-value
-=0.945, and carcinoma P-value =0.005). Finally, we measured the number
-of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
-post-treatment samples and did not observe a significant change for any
-of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.
+we tested whether the microbiota of patients with adenoma (N = 22),
+advanced adenoma (N = 19), or carcinoma (N = 26) had any broad
+differences between pre and post-treatment samples {[}Table 1{]}. The
+structure of the microbial communities of the pre and post-treatment
+samples differed, as measured by the \(\theta\)\textsubscript{YC} beta
+diversity metric {[}Figure 1A{]}. We found that the communities obtained
+pre and post-treament among the patients with carcinomas changed
+significantly more than those patients with adenoma (P-value \textless{}
+0.001). There were no significant differences in the amount of change
+observed between the patients with adenoma and advanced adenoma or
+between the patients with advanced adenoma and carcinoma (P-value
+\textgreater{} 0.05). Next, we tested whether there was a consistent
+direction in the change in the community structure between the pre and
+post-treatment samples for each of the diagnosis groups {[}Figure
+1B-D{]}. We only observed a consistent shift in community structure for
+the patients with carcinoma when using a PERMANOVA test (adenoma P-value
+= 0.999, advanced adenoma P-value = 0.945, and carcinoma P-value =
+0.005). Finally, we measured the number of observed OTUs, Shannon
+evenness, and Shannon diversity in the pre and post-treatment samples
+and did not observe a significant change for any of the diagnosis groups
+(P-value \textgreater{} 0.05) {[}Table S1{]}.
 
@@
 Wilcoxon test we were unable to identify any OTUs that were
-significantly different in the pre and post-treatment groups (P-value NA
-0.05). It is likely that high inter-individual variation and the
-irregular distribution of OTUs across individuals limited the
+significantly different in the pre and post-treatment groups (P-value
+\textgreater{} 0.05). It is likely that high inter-individual variation
+and the irregular distribution of OTUs across individuals limited the
 statistical power of the test. To overcome these problems we developed
@@
 to differentiate between pre and post-treatment samples from each of the
-diagnosis groups. Because of the relatively small number of subjects in
-each group, we restricted our models to only incorporate 10 OTUs to
-limit the likelihood that the models would overfit the data. Despite
+diagnosis groups. To limit the likelihood that the models would overfit
+the data because of the relatively small number of subjects in each
+group, we restricted our models to only incorporate 10 OTUs. Despite
 this restriction, the models performed well (adenoma AUC range = 0.69 -
@@
 impact the structure of the community and removal of that stimulus would
-alter said structure. In addition, it is possible that the difference
-between the microbiota of patients with adenoma and advanced adenoma and
-those with normal colons is subtle. This is supported by the reduced
-ability of our models to correctly classify patients with adenomas and
-advanced adenomas relative to those diagnosed with carcinomas {[}Figure
-S1{]}. Given the irregular distribution of microbiota across patients in
-the different diagnosis groups, it is possible that we lacked the
+alter said structure. It is possible that the difference between the
+microbiota of patients with adenoma and advanced adenoma and those with
+normal colons is subtle. This is supported by the reduced ability of our
+models to correctly classify patients with adenomas and advanced
+adenomas relative to those diagnosed with carcinomas {[}Figure S1{]}.
+Given the irregular distribution of microbiota across patients in the
+different diagnosis groups, it is possible that we lacked the
 statistical power to adequately characterize the change in the
@@
 the length of time between sample collection among those whose
-probabilities increased (382 \(\pm\) 116 days) or decreased (341 \(\pm\)
-98 days) (Wilcoxon Test; P-value = 0.56). Finally, it is possible that
-these patients may not have responded to treatment as well as the other
-20 patients diagnosed with carcinoma and so the microbiota may not have
-been impacted the same way. Again, further studies looking at the role
-of the microbiota in recurrence are needed to understand the dynamics
-following treatment.
+probabilities increased (358 (121) days) or decreased (334 (143) days)
+(Wilcoxon Test; P-value = 0.56) (all days data displayed as median
+(IQR)). Finally, it is possible that these patients may not have
+responded to treatment as well as the other 20 patients diagnosed with
+carcinoma and so the microbiota may not have been impacted the same way.
+Again, further studies looking at the role of the microbiota in
+recurrence are needed to understand the dynamics following treatment.
 
@@
 antibiotic administration on the microbiome would be expected to occur
-(351 \(\pm\) 102 days).
+(344 (142) days).
 
@@
 
-\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
+\textbf{\emph{Model Building.}} {[}Comment: This entire section is
+pretty confusing. In the first paragraph it jumbles the treatment models
+and the diagnosis models{]} The Random Forest {[}33{]} algorithm was
 used to create the three models used. The adenoma model classified
@@
 
-The model included only OTU data obtained from 16S rRNA sequencing.
-Non-binary data was checked for near zero variance and OTUs that had
-near zero variance were removed. This pre-processing was performed with
-the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
-involved making 100 different 80/20 (train/test) splits of the data
-where normal and adenoma, normal and advanced adenoma, or normal and
-carcinoma were represented in the same proportion within both the whole
-data set and the 80/20 split. For each of the different splits, 20
-repeated 10-fold cross validation was performed on the 80\% component to
-optimize the mtry hyper-parameter by maximizing the AUC (Area Under the
-Curve of the Receiver Operator Characteristic). The resulting model was
-then tested on the hold out data obtained from the 20\% component. All
-three models had an optimized mtry of 2. The hyper-parameter, mtry,
-defines the number of variables to investigate at each split before a
-new division of the data was created with the Random Forest model.
+The model included only OTU data obtained from 16S rRNA sequencing and
+was constructed using the R package caret (v6.0.73). Optimization of the
+mtry hyper-parameter involved making 100 different 80/20 (train/test)
+splits of the data where normal and adenoma, normal and advanced
+adenoma, or normal and carcinoma were represented in the same proportion
+within both the whole data set and the 80/20 split. For each of the
+different splits, 20 repeated 10-fold cross validation was performed on
+the 80\% component to optimize the mtry hyper-parameter by maximizing
+the AUC (Area Under the Curve of the Receiver Operator Characteristic).
+The resulting model was then tested on the hold out data obtained from
+the 20\% component. All three models had an optimized mtry of 2. The
+hyper-parameter, mtry, defines the number of variables to investigate at
+each split before a new division of the data was created with the Random
+Forest model.
 
@@
 community structure utilized PERMANOVA {[}34{]} in the vegan package
-(v2.4.1). Comparisons between probabilities as well as overall OTU
-differences between pre and post-treatment samples utilized a paired
-Wilcoxon ranked sum test. Where multiple comparison testing was
-appropriate, a Benjamini-Hochberg (BH) correction was applied {[}35{]}
-and a corrected P-value of less than 0.05 was considered significant.
-Unless otherwise stated the P-values reported are those that were BH
-corrected. Model rank importance was determined by obtaining the median
-MDA from the 100, 20 repeated 10-fold cross validation and then ranking
-from largest to smallest MDA.
+(v2.4.1). Comparisons between probabilities as well as overall
+differences in the median relative abundance of each OTU between pre and
+post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
+multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
+correction was applied {[}35{]} and a corrected P-value of less than
+0.05 was considered significant. The P-values reported are those that
+were BH corrected. Model rank importance was determined by obtaining the
+median MDA from the 100, 20 repeated 10-fold cross validation and then
+ranking from largest to smallest MDA.
 
@@
 
-\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
-and Carcinoma Groups After Treatment.} A) Thetayc distance from pre
+\textbf{Figure 1: General differences between adenoma, advanced adenoma,
+and carcinoma groups after treatment.} A) Thetayc distance from pre
 versus post sample within each individual. A significant difference was
@@
 
-\textbf{Figure 2: The 10 OTUs used to classify treatment for Adenoma,
-Advanced Adenoma, and Carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
+\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
+advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
 OTUs. C) Carcinoma OTUs.
 
-\textbf{Figure 3: Common OTUs to All Models.} A) Venn diagram showing
-the OTU overlap between each model. B) For each common OTU the lowest
-taxonomic identification and importance rank for each model run is
-shown.
+\textbf{Figure 3: OTUs common to those models used to differentiate
+between patients with normal colons and those with adenoma, advanced
+adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
+each model. B) For each common OTU the lowest taxonomic identification
+and importance rank for each model run is shown.
 
-\textbf{Figure 4: Treatment Response Based on Models Built for Adenoma,
-Advanced Adenoma, or Carcinoma.} A) Positive probability change from
+\textbf{Figure 4: Treatment response based on models built for adenoma,
+advanced adenoma, or carcinoma.} A) Positive probability change from
 initial to follow up sample in those with adenoma. B) Positive
@@
 
-\textbf{Table 1: Demographic Data of pre and Post Treatment Cohort}
+\textbf{Table 1: Demographic data of patients in the pre and
+post-treatment cohort}
 
@@
 
-\textbf{Table 2: Demographic Data of Training Cohort}
+\textbf{Table 2: Demographic data of training cohort}
 
@@
 
-\textbf{Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and
-Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
+\textbf{Figure S1: ROC curves of the adenoma, advanced adenoma, and
+carcinoma models.} A) Adenoma ROC curve: The light green shaded areas
 represent the range of values of a 100 different 80/20 splits of the
@@
 
-\textbf{Figure S2: Summary of Important OTUs for the Adenoma, Advanced
-Adenoma, and Carcinoma Models.} A) MDA of the most important variables
+\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
+adenoma, and carcinoma models.} A) MDA of the most important variables
 in the adenoma model. The dark green point represents the mean and the
@@
 
-\textbf{Figure S3: Pre and Post Treatment Relative Abundance of CRC
-associated OTUs within the Carcinoma Model.}
+\textbf{Figure S3: Pre and post-treatment relative abundance of CRC
+associated OTUs within the carcinoma model.}
 

### 2017-05-13 submission/manuscript.Rmd
revised methods: model building

@@
 
-***Model Building.*** [Comment: This entire section is pretty confusing. In the first paragraph it jumbles the treatment models and the diagnosis models]
-The Random Forest [@breiman_random_2001] algorithm was used to create the three models used.
-The adenoma model classified normal versus adenoma, the advanced adenoma model classified normal versus advanced adenoma, and the carcinoma model classified normal versus carcinoma.
-The total number of individuals in the treatment models was `r length(rownames(good_metaf))` individuals.
-There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the treatment carcinoma model [Table 1].
+***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma).
+The total number of individuals in the pre versus post-treatment models was `r length(rownames(good_metaf))` individuals.
+There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the pre versus post-treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the pre versus post-treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the pre versus post-treatment carcinoma model [Table 1].
+
+Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis.
 The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2].
-There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the carcinoma model [Table 2].
-Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] which assesed the prediction of pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`), pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`).
+There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the normal versus carcinoma model [Table 2].
 
-The model included only OTU data obtained from 16S rRNA sequencing and was constructed using the R package caret (v`r paste(packageVersion("caret"))`).
-Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where normal and adenoma, normal and advanced adenoma, or normal and carcinoma were represented in the same proportion within both the whole data set and the 80/20 split.
+All models included only OTU data obtained from 16S rRNA sequencing and were processed and cleaned using the R package caret (v`r paste(packageVersion("caret"))`).
+Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split.
 For each of the different splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
 The resulting model was then tested on the hold out data obtained from the 20% component.
-All three models had an optimized mtry of `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
-The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.
+For all pre versus post-treatment models the optimized mtry was `r ifelse(as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")` and for all normal versus diagnosis models the optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
+The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].
 
-Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run.
+For each of the pre versus post-treamtent models assessment of the most important OTUs was then made by taking the top 10 OTUs by mean decrease in accuracy (MDA).
+These were then used to build each respective reduced OTU pre versus post-treatment model by diagnosis group to help avoid model overfitting.
+These reduced models were then put through the same process mentioned in the previous paragraph and were what was used for the final pre versus post-treatment models.
+For the normal versus diagnosis models the important OTUs were obtained by counting the number of times an OTU was present in the top 10% of MDA for each of the 100 different splits.
 This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs.
-The final collated list of variables was then run through the mtry optimization again.
-Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.
-For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
-The only difference other than the sample set used between treatment models and normal versus diagnosis models was that only the top 10 OTUs were used to build each respective treatment model by diagnosis group to help avoid model overfitting.
+These corresponding reduced OTU normal versus diagnosis models were then put through the same process mentioned in the previous paragraph and were what was used for the final normal versus diagnosis models.
+For the pre versus post-treatment models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")` and for the normal versus diagnosis models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
+
+Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)).
+The application of the pre versus post-treatment models generated the probabilites that the sample was a pre-treatment sample.
+The application of the normal versus diagnosis models generated the probabilities that the sample was that specific diagnosis (adenoma, advanced adenoma, or carcinoma).
 

### 2017-05-13 submission/manuscript.tex
revised methods: model building

@@
 the 16S rRNA genes in the feces of 67 individuals before and after
-treatment for adenoma (N=22), advanced adenoma (N=19), and carcinoma
-(N=26).
+treatment for adenoma (N = 22), advanced adenoma (N = 19), and carcinoma
+(N = 26).
 
@@
 
-\textbf{\emph{Model Building.}} {[}Comment: This entire section is
-pretty confusing. In the first paragraph it jumbles the treatment models
-and the diagnosis models{]} The Random Forest {[}33{]} algorithm was
-used to create the three models used. The adenoma model classified
-normal versus adenoma, the advanced adenoma model classified normal
-versus advanced adenoma, and the carcinoma model classified normal
-versus carcinoma. The total number of individuals in the treatment
-models was 67 individuals. There were a total of 22 individuals in the
-treatment adenoma model, 19 individuals in the treatment advanced
-adenoma model, and 26 individuals in the treatment carcinoma model
-{[}Table 1{]}. The total number of individuals in the normal versus
-diagnosis models was 423 individuals {[}Table 2{]}. There were a total
-of 239 individuals in the adenoma model, 262 individuals in the advanced
-adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
-Each model was then applied to our 67-person cohort {[}Table 1{]} which
-assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
-disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
-disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
-= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
-(advanced adenoma n = 0 and disease free n = 19), pre-treatment
-carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
-carcinoma (carcinoma n = 1 and disease free n = 25).
+\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
+used to create the three models used to classify pre and post-treatment
+samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
+total number of individuals in the pre versus post-treatment models was
+67 individuals. There were a total of 22 individuals in the pre versus
+post-treatment adenoma model, 19 individuals in the pre versus
+post-treatment advanced adenoma model, and 26 individuals in the pre
+versus post-treatment carcinoma model {[}Table 1{]}.
 
-The model included only OTU data obtained from 16S rRNA sequencing and
-was constructed using the R package caret (v6.0.73). Optimization of the
-mtry hyper-parameter involved making 100 different 80/20 (train/test)
-splits of the data where normal and adenoma, normal and advanced
-adenoma, or normal and carcinoma were represented in the same proportion
-within both the whole data set and the 80/20 split. For each of the
-different splits, 20 repeated 10-fold cross validation was performed on
-the 80\% component to optimize the mtry hyper-parameter by maximizing
+Similarily, the Random Forest {[}33{]} algorithm was also used to create
+the three models used to classify normal versus diagnosis. The total
+number of individuals in the normal versus diagnosis models was 423
+individuals {[}Table 2{]}. There were a total of 239 individuals in the
+normal versus adenoma model, 262 individuals in the normal versus
+advanced adenoma model, and 266 individuals in the normal versus
+carcinoma model {[}Table 2{]}.
+
+All models included only OTU data obtained from 16S rRNA sequencing and
+were processed and cleaned using the R package caret (v6.0.73).
+Optimization of the mtry hyper-parameter involved making 100 different
+80/20 (train/test) splits of the data where the same proportion was
+present within both the whole data set and the 80/20 split. For each of
+the different splits, 20 repeated 10-fold cross validation was performed
+on the 80\% component to optimize the mtry hyper-parameter by maximizing
 the AUC (Area Under the Curve of the Receiver Operator Characteristic).
 The resulting model was then tested on the hold out data obtained from
-the 20\% component. All three models had an optimized mtry of 2. The
-hyper-parameter, mtry, defines the number of variables to investigate at
-each split before a new division of the data was created with the Random
-Forest model.
+the 20\% component. For all pre versus post-treatment models the
+optimized mtry was 2 and for all normal versus diagnosis models the
+optimized mtry was 2. The hyper-parameter, mtry, defines the number of
+variables to investigate at each split before a new division of the data
+was created with the Random Forest model {[}33{]}.
 
-Assessment of the most important OTUs to the model involved counting the
-number of times an OTU was present in the top 10\% of mean decrease in
-accuracy (MDA) for each of the 100 different splits run. This was then
-followed with filtering of this list to variables that were only present
-in more than 50\% of these 100 runs. The final collated list of
-variables was then run through the mtry optimization again. Once the
-ideal mtry was found the entire sample set specific to normal versus
-adenoma, normal versus advanced adenoma, or normal versus carcinoma was
-used to create the final Random Forest model on which classifications on
-the 67-person cohort was completed. For all three models the final
-optimized mtry was 2. The only difference other than the sample set used
-between treatment models and normal versus diagnosis models was that
-only the top 10 OTUs were used to build each respective treatment model
-by diagnosis group to help avoid model overfitting.
+For each of the pre versus post-treamtent models assessment of the most
+important OTUs was then made by taking the top 10 OTUs by mean decrease
+in accuracy (MDA). These were then used to build each respective reduced
+OTU pre versus post-treatment model by diagnosis group to help avoid
+model overfitting. These reduced models were then put through the same
+process mentioned in the previous paragraph and were what was used for
+the final pre versus post-treatment models. For the normal versus
+diagnosis models the important OTUs were obtained by counting the number
+of times an OTU was present in the top 10\% of MDA for each of the 100
+different splits. This was then followed with filtering of this list to
+variables that were only present in more than 50\% of these 100 runs.
+These corresponding reduced OTU normal versus diagnosis models were then
+put through the same process mentioned in the previous paragraph and
+were what was used for the final normal versus diagnosis models. For the
+pre versus post-treatment models the final optimized mtry was 2 and for
+the normal versus diagnosis models the final optimized mtry was 2.
+
+Each model was then applied to our 67-person cohort {[}Table 1{]} based
+on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22 and disease
+free n = 0) versus post-treatment adenoma (adenoma n = 0 and disease
+free n = 22)), advanced adenoma pre-treatment advanced adenoma (advanced
+adenoma n = 19 and disease free n = 0 ) versus post-treatment advanced
+adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
+(pre-treatment carcinoma (carinoma n = 26 and disease free n = 0) versus
+post-treatment carcinoma (carcinoma n = 1 and disease free n = 25)). The
+application of the pre versus post-treatment models generated the
+probabilites that the sample was a pre-treatment sample. The application
+of the normal versus diagnosis models generated the probabilities that
+the sample was that specific diagnosis (adenoma, advanced adenoma, or
+carcinoma).
 

### 2017-05-14 submission/manuscript.tex
fix superscript and add actual ranges

@@
 
-Marc A Sze${^1}$, Nielson T Baxter${^1,2}$, Mack T Ruffin IV${^3}$, Mary AM Rogers${^2}$, and Patrick D Schloss${^1}$${^\dagger}$
+Marc A Sze${^1}$, Nielson T Baxter${^1}$${^,}$${^2}$, Mack T Ruffin IV${^3}$, Mary AM Rogers${^2}$, and Patrick D Schloss${^1}$${^\dagger}$
 
@@
 the length of time between sample collection among those whose
-probabilities increased (358 (121) days) or decreased (334 (143) days)
-(Wilcoxon Test; P-value = 0.56) (all days data displayed as median
-(IQR)). Finally, it is possible that these patients may not have
+probabilities increased (358 (336 - 458) days) or decreased (334 (256 -
+399) days) (Wilcoxon Test; P-value = 0.56) (all days data displayed as
+median (IQR)). Finally, it is possible that these patients may not have
 responded to treatment as well as the other 20 patients diagnosed with
@@
 antibiotic administration on the microbiome would be expected to occur
-(344 (142) days).
+(344 (266 - 408) days).
 

### 2017-05-14 submission/manuscript.Rmd

@@
 We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
-We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` (`r format(IQR((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)`) days) or decreased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` (`r format(IQR((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)`) days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) (all days data displayed as median (IQR)).
+We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["75%"], digits = 0)`) days) or decreased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["75%"], digits = 0)`) days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) (all days data displayed as median (IQR)).
 Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
@@
 Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.
-However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(IQR(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)`) days).
+However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["25%"], digits = 0)` - `r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["75%"], digits = 0)`) days).
 

### 2017-05-16 submission/manuscript.Rmd
formating changes

@@
 \end{center}
+
+Co-author e-mails:
+
+* marcsze@med.umich.med.edu
+* ntbaxter@umich.edu
+* mruffin@pennstatehealth.psu.edu
+* maryroge@med.umich.edu
 
@@
 
+## Conclusion
+
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
@@
   t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
-  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
+  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
   t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
@@
   t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
-  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
-  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
+  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
+  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(as.numeric(BMI), na.rm = TRUE))
   t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100

### 2017-05-16 submission/manuscript.tex
formating changes

@@
 \end{center}
+
+Co-author e-mails:
+
+\begin{itemize}
+\tightlist
+\item
+  \href{mailto:marcsze@med.umich.med.edu}{\nolinkurl{marcsze@med.umich.med.edu}}
+\item
+  \href{mailto:ntbaxter@umich.edu}{\nolinkurl{ntbaxter@umich.edu}}
+\item
+  \href{mailto:mruffin@pennstatehealth.psu.edu}{\nolinkurl{mruffin@pennstatehealth.psu.edu}}
+\item
+  \href{mailto:maryroge@med.umich.edu}{\nolinkurl{maryroge@med.umich.edu}}
+\end{itemize}
 
@@
 
+\subsection{Conclusion}\label{conclusion}
+
 This study expands upon existing research that has established a role
@@
 Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
-BMI (Mean ± SD) & 26.86 ± 3.9 & 25.80 ± 4.7 & 28.63 ± 7.2\tabularnewline
+BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
 Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
@@
 Sex (\%F) & 64.53 & 46.27 & 37.78 & 43.62\tabularnewline
-BMI (Mean ± SD) & 26.96 ± 5.3 & 25.68 ± 4.8 & 26.66 ± 4.9 & 29.27 ±
+BMI (Mean ± SD) & 26.97 ± 5.3 & 25.69 ± 4.8 & 26.66 ± 4.9 & 29.27 ±
 6.7\tabularnewline
@@
 \hypertarget{ref-breiman_random_2001}{}
-33. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
-{[}cited 2013 Feb 7{]};45:5--32. Available from:
-\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}
+33. Breiman L. Random Forests. Machine Learning. 2001;45:5--32.
 
@@
 the face of heterogeneous dispersions: What null hypothesis are you
-testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
-5{]};83:557--74. Available from:
-\url{http://doi.wiley.com/10.1890/12-2010.1}
+testing? Ecological Monographs. 2013;83:557--74.
 

### 2017-07-19 submission/manuscript.Rmd
r1 and r2 more difficult comments

@@
 # Extra data and meta data needed
-good_metaf <- read.csv('../data/process/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
+good_metaf <- read.csv('../data/process/mod_metadata/metaf_final.csv', header = T, stringsAsFactors = F)
 metaI <- read.csv('../data/process/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
@@
 Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota.
-If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence.
+If the microbial community has an affect on recurrence risk or tumorigenesis it would be reasonable to expect treatment to remove a lesion to affect the microbiota.
 To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner?
@@
 A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon.
-There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals.
+There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3)`).
 Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
@@
 For all three groups of patients, we observed changes in their microbiota.
+Some of these changes, specifically for adenoma, may be due to normal temporal variation but those with advanced adenoma and CRC clearly had large microbiota changes that would be greater than what could be expected from this type of variation [@schloss_dynamics_2014].
 After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
@@
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
-If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy.
+Although we cannot make a specific statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples and small effect sizes in the adenoma and advnaced adenoma groups, there are still a number of exciting new avenues to explore.
+One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
+If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy, if the biomarkers identified play a key role in the disease process.
 Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.
@@
 Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016].
-Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
+Briefly, samples were stored on ice for at least 24h before freezing.
+Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology.
+Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy.
@@
 All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
+
+***Treatment***
+The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1] but rather via colonoscopy.
+All CRC individuals were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
+The type of chemotherapy used for patients with CRC varied markably.
+The types used included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
+These were used individually or in combination with others depending on the patient [Table 1].
+If an individual was treated with radiation they were also always treated with chemotherapy.
+Radiation therapy generally used 18 mV photons for treatment.
 
@@
 Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis.
+These samples were obtained using the same methodology as described earlier in this section.
+All samples used for this component of model training were from pre-treatment samples.
 The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2].
@@
 
-**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs.
+**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
+B) Advanced Adenoma OTUs.
+C) Carcinoma OTUs.
+The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
 
@@
 ```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
-test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time)
+test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received, Abx)
 
-t2 <- as.data.frame(matrix(nrow = 6, ncol = 3,
+t2 <- as.data.frame(matrix(nrow = 10, ncol = 3, 
                            dimnames = 
-                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time"),
+                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time",  
+                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation", "antibiotics"), 
                                   coln = c("Adenoma", "SRN", "Carcinoma"))))
@@
   t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
-
   t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))
+  t2["Surgery_only", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Surgery == "Y" & chemo_received == "no") %>% select(Surgery) %>% count()
+  t2["Surg_Chemo", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & chemo_received == "yes" & radiation_received == "no") %>% select(chemo_received) %>% count()
+  t2["Surg_Chemo_Radiation", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & radiation_received == "yes") %>% select(radiation_received) %>% count()
+  t2["antibiotics", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Abx > 0) %>% select(Abx) %>% count()
 
@@
   post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])
-
 }
@@
                                                     label = "\u00B1",
-                                                    format(post_time_sd, digits = 2, nmsall = 2)))
+                                                    format(post_time_sd, digits = 2, nmsall = 2)), 
+                                        Surgery_only = format(Surgery_only, digits = 2, nsmall = 0), 
+                                        Surg_Chemo = format(Surg_Chemo, digits = 2, nsmall = 0), 
+                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0), 
+                                        antibiotics = format(antibiotics, digits = 2, nsmall = 0))
 
 ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
-             ttemptable$BMI, ttemptable$White, ttemptable$post_time)
+             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation, ttemptable$antibiotics)
 
@@
 
-colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label)
+colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label, 
+                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation", 
+                   "Recorded Antibiotic Use")
 

### 2017-07-19 code/antibiotic_comparisons.R
add antibiotic comparison first pass

@@ -1,0 +1,85 @@
+### Running Antibiotic comparisons 
+## Marc Sze
+
+
+###Load needed Libraries and functions
+source('code/functions.R')
+
+loadLibs(c("dplyr", "tidyr"))
+
+
+# Read in data tables
+good_metaf <- read.csv(
+  "data/process/mod_metadata/metaF_final.csv", 
+  stringsAsFactors = F, header = T) %>% 
+  mutate(lesion_follow = ifelse(Disease_Free == "n", 1, 0))
+
+# Load diversity data
+alpha_summary <- read.delim("data/process/final.groups.ave-std.summary", stringsAsFactors = F)
+thetaCompTotal <- read.dist('data/process/final.thetayc.0.03.lt.ave.dist')
+
+# Create variables that are to be used 
+initial_samples <- good_metaf$initial
+followup_samples <- good_metaf$followUp
+
+# Create alpha table for comparisons
+working_alpha <- alpha_summary %>% filter(method == "ave") %>% 
+  slice(match(c(initial_samples, followup_samples), group)) %>% 
+  select(group, sobs, shannon, shannoneven)
+
+# Modify meta data so that it is in a correct format
+mod_metaf <- good_metaf %>% slice(match(initial_samples, initial)) %>% 
+  bind_rows(slice(good_metaf, match(followup_samples, followUp))) %>% 
+  mutate(sampleType = c(rep("initial", 67), rep("followup", 67))) %>% 
+  select(-followUp) %>% rename(samples = initial)
+
+# Select specific beta function
+get_beta <- function(i, j, table_name){
+  
+  data_table <- get(table_name)
+  beta_values <- data_table[i, j]
+  
+  return(beta_values)
+}
+
+# create the beta data and join with original metadata
+working_beta <- as.data.frame(initial_samples) %>% 
+  mutate(distance = as.numeric(
+    mapply(get_beta, as.character(initial_samples), as.character(followup_samples), "thetaCompTotal"))) %>% 
+  rename(initial = initial_samples) %>% 
+  inner_join(good_metaf, by = "initial")
+
+
+
+## Make comparisons of surgery for adenoma and advanced adenoma within groups
+working_beta %>% filter(Dx_Bin == "adenoma") %>% group_by(Surgery) %>% 
+  summarise(median_value = median(distance), 
[…]

### 2017-07-20 code/antibiotic_comparisons.R
added table write out

@@ -80,6 +80,15 @@ surg_only_comparisons <- data_frame(
                 filter(working_beta, Dx_Bin == "adv_adenoma" & Surgery == "Y")[, "distance"])$p.value)) %>% 
   mutate(BH = p.adjust(pvalues, method = "BH"), comparison = c("adnVcrc", "srnVcrc", "adnVsrn"))
 
+# Write out needed tables
+write.csv(surg_only_comparisons, "data/process/tables/surgery_only_comparisons.csv", 
+          row.names = F)
+
+write.csv(surg_non_crc_summary, "data/process/tables/surgVnonsurg_noncrc.csv", 
+          row.names = F)
+
+write.csv(non_crc_surg_pvalues, "data/process/tables/surgVnonsurg_noncrc_pvalues.csv", 
+          row.names = F)
 
 
 # Want to graph both sections and add a supplemental figure 

### 2017-07-20 submission/manuscript.Rmd
properly incorporate all data tables to read in

@@
 
+# Antibiotic test data
+surg_only_comparisons <- read.csv("../data/process/tables/surgery_only_comparisons.csv", header = T)
+surgVnonsurg_noncrc <- read.csv("../data/process/tables/surgVnonsurg_noncrc.csv", header = T)
+surgVnonsurg_noncrc_pvalues <- read.csv("../data/process/tables/surgVnonsurg_noncrc_pvalues.csv", header = T)
+
 # Treatment model data
@@
 adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)
+adn_randomized_summary <- read.csv('../data/process/tables/reduced_adn_randomization_treatment_ROC_model_summary.csv', header = T)
 
@@
 srn_general_summary <- read.csv('../data/process/tables/srn_reduced_auc_summary.csv', header = T, row.names = 1)
+srn_randomized_summary <- read.csv('../data/process/tables/reduced_srn_randomization_treatment_ROC_model_summary.csv', header = T)
 
@@
 crc_probs <- read.csv('../data/process/tables/crc_reduced_follow_up_probability_summary.csv', header = T)
+crc_randomized_summary <- read.csv('../data/process/tables/reduced_crc_randomization_treatment_ROC_model_summary.csv', header = T)
 

### 2017-07-20 submission/manuscript.Rmd
move tables in front of fig legends
update supplemental figures

@@
 
-**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) Thetayc distance from pre versus post sample within each individual.
-A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
-Solid black points represent the median value for each diagnosis group.
-B) NMDS of the pre and post-treatment samples for the adenoma group.
-C) NMDS of the pre and post-treatment samples for the advanced adenoma group.
-D) NMDS of the pre and post-treatment samples for the carcinoma group.
-
-**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
-B) Advanced Adenoma OTUs.
-C) Carcinoma OTUs.
-The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
-
-**Figure 3: OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
-B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
-
-**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
-B) Positive probability change from initial to follow up sample in those with advanced adenoma.
-C) Positive probability change from initial to follow up sample in those with carcinoma.
-
-\newpage
-
 **Table 1: Demographic data of patients in the pre and post-treatment cohort**
@@
 
-**Figure S1: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) $\theta$~YC~ distance from pre versus post sample within each individual.
+A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
+Solid black points represent the median value for each diagnosis group.
+B) NMDS of the pre and post-treatment samples for the adenoma group.
+C) NMDS of the pre and post-treatment samples for the advanced adenoma group.
+D) NMDS of the pre and post-treatment samples for the carcinoma group.
+
+**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
+B) Advanced Adenoma OTUs.
+C) Carcinoma OTUs.
+The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
+
+**Figure 3: OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
+B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
+
+**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
+B) Positive probability change from initial to follow up sample in those with advanced adenoma.
+C) Positive probability change from initial to follow up sample in those with carcinoma.
+
+\newpage
+
+**Figure S1: Surgery comparisons by diagnosis using $\theta$~YC~ distance.** A) Differences by diagnosis group for pre versus post sample within each individual.
+No significant difference was found but the trend is similar to that found for the full data set.
+B) Difference between those receiving surgery and those that did not for the adenoma group.
+No significant difference was observed.
+C) Difference between those receiving surgery and those that did not for the advanced adenoma group.
+No significant difference was observed.
+
+**Figure S2: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S2: Summary of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
+**Figure S3: Summary of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
@@
 
-**Figure S3: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**
+**Figure S4: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**
 

### 2017-07-20 submission/manuscript.Rmd
update supplemental figure labels in text

@@
 We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1).
+The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
 The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
@@
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing are marked decrease in relative abundance [Figure S3].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing are marked decrease in relative abundance [Figure S4].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
@@
 It is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
-This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1].
+This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S2].
 Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.

### 2017-07-21 submission/manuscript.Rmd
complete incorporating all new analysis

@@
 Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
+None of the individuals in this study had any recorded antibiotic use before treatment for their adenoma, advanced adenoma, and carcinoma.
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
@@
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
+When analyzing only those who underwent surgery as a treatment these trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
+There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
@@
 It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
-To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-To limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
+We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
+To try and limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
 Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
+However, upon randomizing the pre and post treatment labels we found that Random Forest could work equally well (adenoma AUC range = `r format(min(adn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_randomized_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_randomized_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_randomized_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
-These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.
+Although treatment had an impact on the overall community structure; the effect of treatment was not consistent across patients and diagnosis groups.
 
@@
 Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`).
-We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power.
-We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
-Finally, the change in the positive probability was not significantly different between the two treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Furthermore, there was no significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Finally, the change in the positive probability was also not significantly different between the any treatment group (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
 Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
@@
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-Although we cannot make a specific statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples and small effect sizes in the adenoma and advnaced adenoma groups, there are still a number of exciting new avenues to explore.
+Although we cannot make a definitive statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, small effect sizes, potential changes in diet after treatment, changing medical care for chronic maladies, etc.
+in the adenoma and advnaced adenoma groups, there are still a number of exciting new potential avenues to explore.
 One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
@@
 Comparisons between probabilities as well as overall differences in the median relative abundance of each OTU between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test.
+Comparisons between different treatment for lesions utilized a Kruskal Wallis test.
 Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
@@
                    "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation", 
-                   "Recorded Antibiotic Use")
+                   "Pre-Treatment Antibiotic Use")
 

### 2017-07-21 submission/manuscript.Rmd
complete revision edits incorporation

@@
 
-# The effect of treatment on the microbiota of patients diagnosed with colonic lesions
+# Normalization of the microbiota in patients after treatment for colonic lesions
 
@@
 
-**Results.** There were large changes to the bacterial communities associated with treatment across the three groups.
+**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma groups and large changes associated with the carcinoma group.
 The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
@@
 Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups.
-Although the three models successfully differentiated between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment.
+Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to try and identify groups of OTUs that would allow us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups.
+Although the three models could successfully differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment and randomization of the diagnosis labels recapitulated the actual models AUCs.
 Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons.
-We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses.
-Only patients who had carcinomas experienced a significant decrease in positive probability of having a lesion after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05), indicating that the microbial milieu of the colon more closely resembled that of a normal colon.
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether these changes were still similar to their original diagnosis or not.
+Only patients who had carcinomas had a change that was towards a microbial milieu that resembles the normal colon (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
 Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
+**Conclusions.** Although it was difficult to identify significant differences between pre and post treatment samples for adenoma and advanced adenoma, this was not the case for carcinomas.
+Not only were there large changes in pre versus post treatment samples for those with carcinoma but also these changes were towards a more normal microbiota composition.
 
@@
 
-***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
+***Treatment for colonic lesions alters the bacterial community structure***
 Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
-None of the individuals in this study had any recorded antibiotic use before treatment for their adenoma, advanced adenoma, and carcinoma.
+None of the individuals in this study had any recorded antibiotic use before treatment of their respective lesion [Table 1].
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
@@
 
-***The effects of treatment are not consistent across treatment groups.***
+***The treatment of lesions are not consistent across diagnosis groups.***
 We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group.
@@
 Furthermore, there was no significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
-Finally, the change in the positive probability was also not significantly different between the any treatment group (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Finally, the change in the positive probability was also not significantly different between any treatment group (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
 Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
@@
 Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.
-It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to assess the risk of recurrence.
+It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to also assess the risk of recurrence.
 
@@
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-Although we cannot make a definitive statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, small effect sizes, potential changes in diet after treatment, changing medical care for chronic maladies, etc.
-in the adenoma and advnaced adenoma groups, there are still a number of exciting new potential avenues to explore.
+Although in the adenoma and advanced adenoma groups we cannot make a definitive statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, small effect sizes, potential changes in diet after treatment, changing medical care for chronic maladies, etc.
+There are still a number of exciting new potential avenues to explore based on the carcinoma observations.
 One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
@@
 ***Treatment***
+For this study treatment refers specifically to the removal of a lesion or with or without chemotherapy and radiation.
 The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1] but rather via colonoscopy.

### 2017-07-21 submission/manuscript.tex
complete revision edits incorporation

@@
 
-\section{The effect of treatment on the microbiota of patients diagnosed
-with colonic
-lesions}\label{the-effect-of-treatment-on-the-microbiota-of-patients-diagnosed-with-colonic-lesions}
+\section{Normalization of the microbiota in patients after treatment for
+colonic
+lesions}\label{normalization-of-the-microbiota-in-patients-after-treatment-for-colonic-lesions}
 
@@
 
-\textbf{Results.} There were large changes to the bacterial communities
-associated with treatment across the three groups. The communities from
-patients with carcinomas changed significantly more than those with
-adenoma following treatment (P-value \textless{} 0.001). There was no
-significant change in the microbiota between patients with adenoma and
-advanced adenoma, or between patients with advanced adenoma and
-carcinoma (P-value \textgreater{} 0.05). Although treatment was
-associated with intrapersonal changes, the change in the abundance of
-individual OTUs to treatment was not consistent within diagnosis groups
-(P-value \textgreater{} 0.05). Because the distribution of OTUs across
-patients and diagnosis groups was irregular, we used the Random Forest
-machine learning algorithm to identify groups of OTUs that allowed us to
-successfully distinguish between pre and post-treatment samples for each
-of the diagnosis groups. Although the three models successfully
-differentiated between the pre and post-treatment samples, there was
-little overlap between the OTUs that were indicative of treatment. Next,
-we used a larger cohort that contained individuals with normal colons
-and those with adenomas, advanced adenomas, and carcinomas to determine
-whether individuals who underwent treatment were more likely to have
-OTUs associated with normal colons. We again built Random Forest models
-and measured the change in the positive probability of having one of the
-three diagnoses. Only patients who had carcinomas experienced a
-significant decrease in positive probability of having a lesion after
-treatment (P-value \textless{} 0.05), indicating that the microbial
-milieu of the colon more closely resembled that of a normal colon.
-Finally, we tested whether the type of treatment impacted the microbiota
-of those diagnosed with carcinomas and were unable to detect any
-significant differences in characteristics of these communities between
-individuals treated with surgery alone and those treated with
+\textbf{Results.} There were small changes to the bacterial community
+associated with adenoma or advanced adenoma groups and large changes
+associated with the carcinoma group. The communities from patients with
+carcinomas changed significantly more than those with adenoma following
+treatment (P-value \textless{} 0.001). There was no significant change
+in the microbiota between patients with adenoma and advanced adenoma, or
+between patients with advanced adenoma and carcinoma (P-value
+\textgreater{} 0.05). Although treatment was associated with
+intrapersonal changes, the change in the abundance of individual OTUs to
+treatment was not consistent within diagnosis groups (P-value
+\textgreater{} 0.05). Because the distribution of OTUs across patients
+and diagnosis groups was irregular, we used the Random Forest machine
+learning algorithm to try and identify groups of OTUs that would allow
+us to successfully distinguish between pre and post-treatment samples
+for each of the diagnosis groups. Although the three models could
+successfully differentiate between the pre and post-treatment samples,
+there was little overlap between the OTUs that were indicative of
+treatment and randomization of the diagnosis labels recapitulated the
+actual models AUCs. Next, we used a larger cohort that contained
+individuals with normal colons and those with adenomas, advanced
+adenomas, and carcinomas to determine whether individuals who underwent
+treatment were more likely to have OTUs associated with normal colons.
+We again built Random Forest models and measured the change in the
+positive probability of having one of the three diagnoses to assess
+whether these changes were still similar to their original diagnosis or
+not. Only patients who had carcinomas had a change that was towards a
+microbial milieu that resembles the normal colon (P-value \textless{}
+0.05). Finally, we tested whether the type of treatment impacted the
+microbiota of those diagnosed with carcinomas and were unable to detect
+any significant differences in characteristics of these communities
+between individuals treated with surgery alone and those treated with
 chemotherapy or chemotherapy and radiation (P-value \textgreater{}
@@
 
-\textbf{Conclusions.} By better understanding the response of the
-microbiota to treatment for adenomas and carcinomas, it is likely that
-biomarkers will be validated that can be used to quantify the risk of
-recurrence and the likelihood of survival.
+\textbf{Conclusions.} Although it was difficult to identify significant
+differences between pre and post treatment samples for adenoma and
+advanced adenoma, this was not the case for carcinomas. Not only were
+there large changes in pre versus post treatment samples for those with
+carcinoma but also these changes were towards a more normal microbiota
+composition.
 
@@
 including resection, chemotherapy, and/or radiation affect the
-composition of the gut microbiota. If the microbial community drives
-tumorigenesis then one would hypothesize that treatment to remove a
-lesion would affect the microbiota and risk of recurrence. To test this
-hypothesis, we addressed two related questions: Does treatment affect
-the colonic microbiota in a predictable manner? If so, does the
+composition of the gut microbiota. If the microbial community has an
+affect on recurrence risk or tumorigenesis it would be reasonable to
+expect treatment to remove a lesion to affect the microbiota. To test
+this hypothesis, we addressed two related questions: Does treatment
+affect the colonic microbiota in a predictable manner? If so, does the
 treatment alter the community to more closely resemble that of
@@
 
-\textbf{\emph{Treatment alters the bacterial community structure of
-patients diagnosed with colonic lesions.}} Within our 67-person cohort
-we tested whether the microbiota of patients with adenoma (N = 22),
-advanced adenoma (N = 19), or carcinoma (N = 26) had any broad
-differences between pre and post-treatment samples {[}Table 1{]}. The
-structure of the microbial communities of the pre and post-treatment
-samples differed, as measured by the \(\theta\)\textsubscript{YC} beta
-diversity metric {[}Figure 1A{]}. We found that the communities obtained
-pre and post-treament among the patients with carcinomas changed
-significantly more than those patients with adenoma (P-value \textless{}
-0.001). There were no significant differences in the amount of change
-observed between the patients with adenoma and advanced adenoma or
-between the patients with advanced adenoma and carcinoma (P-value
-\textgreater{} 0.05). Next, we tested whether there was a consistent
-direction in the change in the community structure between the pre and
-post-treatment samples for each of the diagnosis groups {[}Figure
-1B-D{]}. We only observed a consistent shift in community structure for
-the patients with carcinoma when using a PERMANOVA test (adenoma P-value
-= 0.999, advanced adenoma P-value = 0.945, and carcinoma P-value =
-0.005). Finally, we measured the number of observed OTUs, Shannon
-evenness, and Shannon diversity in the pre and post-treatment samples
-and did not observe a significant change for any of the diagnosis groups
-(P-value \textgreater{} 0.05) {[}Table S1{]}.
+\textbf{\emph{Treatment for colonic lesions alters the bacterial
+community structure}} Within our 67-person cohort we tested whether the
+microbiota of patients with adenoma (N = 22), advanced adenoma (N = 19),
+or carcinoma (N = 26) had any broad differences between pre and
+post-treatment samples {[}Table 1{]}. None of the individuals in this
+study had any recorded antibiotic use before treatment of their
+respective lesion {[}Table 1{]}. The structure of the microbial
+communities of the pre and post-treatment samples differed, as measured
+by the \(\theta\)\textsubscript{YC} beta diversity metric {[}Figure
+1A{]}. We found that the communities obtained pre and post-treament
+among the patients with carcinomas changed significantly more than those
+patients with adenoma (P-value \textless{} 0.001). There were no
+significant differences in the amount of change observed between the
+patients with adenoma and advanced adenoma or between the patients with
+advanced adenoma and carcinoma (P-value \textgreater{} 0.05). When
+analyzing only those who underwent surgery as a treatment these trends
+persisted (P-value \textgreater{} 0.05) {[}Figure S1A{]}. There was also
+no difference in the change of \(\theta\)\textsubscript{YC} between pre
+and post treatment samples between those who had surgery versus those
+who did not for both the adenoma and advanced adenoma group (P-value
+\textgreater{} 0.05) {[}Figure S1B-C{]}. Next, we tested whether there
+was a consistent direction in the change in the community structure
+between the pre and post-treatment samples for each of the diagnosis
+groups {[}Figure 1B-D{]}. We only observed a consistent shift in
+community structure for the patients with carcinoma when using a
+PERMANOVA test (adenoma P-value = 0.999, advanced adenoma P-value =
+0.945, and carcinoma P-value = 0.005). Finally, we measured the number
+of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
+post-treatment samples and did not observe a significant change for any
+of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.
 
-\textbf{\emph{The effects of treatment are not consistent across
-treatment groups.}} We used two approaches to identify those bacterial
+\textbf{\emph{The treatment of lesions are not consistent across
+diagnosis groups.}} We used two approaches to identify those bacterial
 populations that change between the two samples for each diagnosis
@@
 and the irregular distribution of OTUs across individuals limited the
-statistical power of the test. To overcome these problems we developed
-Random Forest models to identify collections of OTUs that would allow us
-to differentiate between pre and post-treatment samples from each of the
-diagnosis groups. To limit the likelihood that the models would overfit
-the data because of the relatively small number of subjects in each
-group, we restricted our models to only incorporate 10 OTUs. Despite
-this restriction, the models performed well (adenoma AUC range = 0.69 -
-0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
-0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
-had little overlap with each other {[}Figure 2{]}. These results support
-the earlier community-wide analysis where we observed that the treatment
-had an impact on the overall community structure; however, the effect of
-treatment was not consistent across patients and diagnosis groups.
+statistical power of the test. We attempted to overcome these problems
+by using Random Forest models to identify collections of OTUs that would
+allow us to differentiate between pre and post-treatment samples from
+each of the diagnosis groups. To try and limit the likelihood that the
+models would overfit the data because of the relatively small number of
+subjects in each group, we restricted our models to only incorporate 10
+OTUs. Despite this restriction, the models performed well (adenoma AUC
+range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
+AUC range = 0.82 - 0.98). However, upon randomizing the pre and post
+treatment labels we found that Random Forest could work equally well
+(adenoma AUC range = 0.73 - 0.98, advanced adenoma AUC range = 0.70 -
+0.97, carcinoma AUC range = 0.89 - 0.99). Interestingly, the 10 OTUs
+that were used for each model had little overlap with each other
+{[}Figure 2{]}. Although treatment had an impact on the overall
+community structure; the effect of treatment was not consistent across
+patients and diagnosis groups.
 
@@
 models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
-AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
+AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S2).
 The OTUs that were incorporated into the adenoma and advanced adenoma
@@
 carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
-showing are marked decrease in relative abundance {[}Figure S3{]}.
+\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]} with some individuals
+showing are marked decrease in relative abundance {[}Figure S4{]}.
 Finally, we applied these three models to the pre and post-treatment
@@
 in the positive probability for the adenoma or advanced adenoma groups
-{[}Figure 4{]}. The positive probability for the pre and post-treatment
-samples from patients diagnosed with carcinoma significantly decreased
-with treatment, suggesting a shift toward a normal microbiota for most
-individuals. Only, 6 of the 26 patients (23.08\%) who were diagnosed
-with a carcinoma had a higher positive probability after treatment; one
-of those was re-diagnosed with carcinoma on the follow up visit. These
-results indicate that, although there were changes in the microbiota
-associated with treatment, those experienced by patients with carcinoma
-after treatment yielded gut bacterial communities of greater similarity
-to that of a normal colon.
+(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
+for the pre and post-treatment samples from patients diagnosed with
+carcinoma significantly decreased with treatment, suggesting a shift
+toward a normal microbiota for most individuals (P-value = 0.0156).
+Only, 6 of the 26 patients (23.08\%) who were diagnosed with a carcinoma
+had a higher positive probability after treatment; one of those was
+re-diagnosed with carcinoma on the follow up visit. These results
+indicate that, although there were changes in the microbiota associated
+with treatment, those experienced by patients with carcinoma after
+treatment yielded gut bacterial communities of greater similarity to
+that of a normal colon.
 
@@
 resection (N=12), surgical resection with chemotherapy (N=9), and
-surgical resection with chemotherapy and radiation (N=5). We focused on
-the patients with carcinoma and pooled those patients that received
-chemotherapy with those that received chemotherapy and radiation to
-improve our statistical power. We did not observe a significant
-difference in the effect of these treatments on the number of observed
-OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
-0.05). Furthermore, there was not a significant difference in the effect
-of the treatments on the amount of change in the community structure
-(P-value = 0.298). Finally, the change in the positive probability was
-not significantly different between the two treatment groups (P-value =
-0.999). Due to the relatively small number of samples in each treatment
-group, it was difficult to make a definitive statement regarding the
-specific type of treatment on the amount of change in the structure of
-the microbiota.
+surgical resection with chemotherapy and radiation (N=5). Regardless of
+treatment used there was no significant difference in the effect of
+these treatments on the number of observed OTUs, Shannon diversity, or
+Shannon evenness (P-value \textgreater{} 0.05). Furthermore, there was
+no significant difference in the effect of the treatments on the amount
+of change in the community structure (P-value = 0.532). Finally, the
+change in the positive probability was also not significantly different
+between any treatment group (P-value = 0.532). Due to the relatively
+small number of samples in each treatment group, it was difficult to
+make a definitive statement regarding the specific type of treatment on
+the amount of change in the structure of the microbiota.
 
@@
 all three groups of patients, we observed changes in their microbiota.
-After treatment, the microbiota of patients with carcinoma changed
-significantly more than the other groups. This change resulted in
-communities that more closely resembled those of patients with a normal
-colon. This may suggest that treatment for carcinoma is not only
-successful for removing the carcinoma but also at reducing the
-associated bacterial communities. Understanding the effect of treatment
-on the microbiota of those diagnosed with carcinomas may have important
-implications for reducing disease recurrence. It is intriguing that it
-may be possible to use microbiome-based biomarkers to not only predict
-the presence of lesions but to assess the risk of recurrence.
+Some of these changes, specifically for adenoma, may be due to normal
+temporal variation but those with advanced adenoma and CRC clearly had
+large microbiota changes that would be greater than what could be
+expected from this type of variation {[}27{]}. After treatment, the
+microbiota of patients with carcinoma changed significantly more than
+the other groups. This change resulted in communities that more closely
+resembled those of patients with a normal colon. This may suggest that
+treatment for carcinoma is not only successful for removing the
+carcinoma but also at reducing the associated bacterial communities.
+Understanding the effect of treatment on the microbiota of those
+diagnosed with carcinomas may have important implications for reducing
+disease recurrence. It is intriguing that it may be possible to use
+microbiome-based biomarkers to not only predict the presence of lesions
+but to also assess the risk of recurrence.
 
@@
 models to correctly classify patients with adenomas and advanced
-adenomas relative to those diagnosed with carcinomas {[}Figure S1{]}.
+adenomas relative to those diagnosed with carcinomas {[}Figure S2{]}.
 Given the irregular distribution of microbiota across patients in the
@@
 microbiome-based biomarkers to predict the presence of colonic lesions.
-The most exciting future direction from the current study is the
-possibility that markers within the microbiota could be used to evaluate
-the effect of treatment and predict recurrence for those diagnosed with
+Although in the adenoma and advanced adenoma groups we cannot make a
+definitive statement regarding the effect of a specific treatment on the
+gut microbiota, due to the low number of samples, small effect sizes,
+potential changes in diet after treatment, changing medical care for
+chronic maladies, etc. There are still a number of exciting new
+potential avenues to explore based on the carcinoma observations. One of
+the most exciting of these future directions is the possibility that
+markers within the microbiota could be used to potentially evaluate the
+effect of treatment and predict recurrence for those diagnosed with
 carcinoma. If such an approach is effective, it might be possible to
-target the microbiota as part of adjuvant therapy. Our data provides
+target the microbiota as part of adjuvant therapy, if the biomarkers
+identified play a key role in the disease process. Our data provides
 additional evidence on the importance of the microbiota in tumorigenesis
@@
 \textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
-have been previously reported in Baxter, et al {[}24{]}. Briefly, study
-exclusion involved those who had already undergone surgery, radiation,
-or chemotherapy, had colorectal cancer before a baseline fecal sample
-could be obtained, had IBD, a known hereditary non-polyposis colorectal
-cancer, or familial adenomatous polyposis. Samples used to build the
-models for prediction were collected either prior to a colonoscopy or
-between one and two weeks after initial colonoscopy. The bacterial
-community has been shown to normalize back to a pre-colonoscopy
-community within this time period {[}27{]}. Our study cohort consisted
-of 67 individuals with an initial sample as described and a follow up
-sample obtained between 188 - 546 days after treatment of lesion
-{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
-histopathological review of any biopsies taken. Patients were classified
-as having advanced adenoma if they had an adenoma greater than 1 cm,
-more than three adenomas of any size, or an adenoma with villous
-histology. This study was approved by the University of Michigan
-Institutional Review Board. All study participants provided informed
-consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+have been previously reported in Baxter, et al {[}24{]}. Briefly,
+samples were stored on ice for at least 24h before freezing. Although we
+cannot exclude that this sampling protocol may have impacted the gut
+microbiota composition all samples were subjected to the same
+methodology. Study exclusion involved those who had already undergone
+surgery, radiation, or chemotherapy, had colorectal cancer before a
+baseline fecal sample could be obtained, had IBD, a known hereditary
+non-polyposis colorectal cancer, or familial adenomatous polyposis.
+Samples used to build the models for prediction were collected either
+prior to a colonoscopy or between one and two weeks after initial
+colonoscopy. The bacterial community has been shown to normalize back to
+a pre-colonoscopy community within this time period {[}28{]}. Our study
+cohort consisted of 67 individuals with an initial sample as described
+and a follow up sample obtained between 188 - 546 days after treatment
+of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
+examination and histopathological review of any biopsies taken. Patients
+were classified as having advanced adenoma if they had an adenoma
+greater than 1 cm, more than three adenomas of any size, or an adenoma
+with villous histology. This study was approved by the University of
+Michigan Institutional Review Board. All study participants provided
+informed consent and the study itself conformed to the guidelines set
+out by the Helsinki Declaration.
+
+\textbf{\emph{Treatment}} For this study treatment refers specifically
+to the removal of a lesion or with or without chemotherapy and
+radiation. The majority of patients undergoing treatment for adenoma or
+advanced adenoma were not treated surigically {[}Table 1{]} but rather
+via colonoscopy. All CRC individuals were treated with at least surgery
+or a combination of surgery and chemotherapy or surgery, chermotherapy,
+and radiation. The type of chemotherapy used for patients with CRC
+varied markably. The types used included Oxaliplatin, Levicovorin,
+Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
+These were used individually or in combination with others depending on
+the patient {[}Table 1{]}. If an individual was treated with radiation
+they were also always treated with chemotherapy. Radiation therapy
+generally used 18 mV photons for treatment.
 
 \textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
-described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
+described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
 Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}28{]}. The general workflow using mothur
+previously described {[}29{]}. The general workflow using mothur
 included merging paired-end reads into contigs, filtering for low
-quality contigs, aligning to the SILVA database {[}29{]}, screening for
-chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
-classifier using the Ribosomal Database Project (RDP){[}31{]}, and
+quality contigs, aligning to the SILVA database {[}30{]}, screening for
+chimeras using UCHIME {[}31{]}, classifying with a naive Bayesian
+classifier using the Ribosomal Database Project (RDP){[}32{]}, and
 clustered into Operational Taxonomic Units (OTUs) using a 97\%
 similarity cutoff with an average neighbor clustering algorithm
-{[}32{]}. The number of sequences for each sample was rarefied to 10523
+{[}33{]}. The number of sequences for each sample was rarefied to 10523
 to minimize the impacts of uneven sampling.
 
-\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
+\textbf{\emph{Model Building.}} The Random Forest {[}34{]} algorithm was
 used to create the three models used to classify pre and post-treatment
@@
 
-Similarily, the Random Forest {[}33{]} algorithm was also used to create
-the three models used to classify normal versus diagnosis. The total
-number of individuals in the normal versus diagnosis models was 423
-individuals {[}Table 2{]}. There were a total of 239 individuals in the
-normal versus adenoma model, 262 individuals in the normal versus
-advanced adenoma model, and 266 individuals in the normal versus
-carcinoma model {[}Table 2{]}.
+Similarily, the Random Forest {[}34{]} algorithm was also used to create
+the three models used to classify normal versus diagnosis. These samples
+were obtained using the same methodology as described earlier in this
+section. All samples used for this component of model training were from
+pre-treatment samples. The total number of individuals in the normal
+versus diagnosis models was 423 individuals {[}Table 2{]}. There were a
+total of 239 individuals in the normal versus adenoma model, 262
+individuals in the normal versus advanced adenoma model, and 266
+individuals in the normal versus carcinoma model {[}Table 2{]}.
 
 All models included only OTU data obtained from 16S rRNA sequencing and
-were processed and cleaned using the R package caret (v6.0.73).
+were processed and cleaned using the R package caret (v6.0.76).
 Optimization of the mtry hyper-parameter involved making 100 different
@@
 variables to investigate at each split before a new division of the data
-was created with the Random Forest model {[}33{]}.
+was created with the Random Forest model {[}34{]}.
 
@@
 
-\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
+\textbf{\emph{Statistical Analysis.}} The R software package (v3.4.1)
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}34{]} in the vegan package
-(v2.4.1). Comparisons between probabilities as well as overall
+community structure utilized PERMANOVA {[}35{]} in the vegan package
+(v2.4.3). Comparisons between probabilities as well as overall
 differences in the median relative abundance of each OTU between pre and
-post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
-multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
-correction was applied {[}35{]} and a corrected P-value of less than
-0.05 was considered significant. The P-values reported are those that
-were BH corrected. Model rank importance was determined by obtaining the
-median MDA from the 100, 20 repeated 10-fold cross validation and then
-ranking from largest to smallest MDA.
+post-treatment samples utilized a paired Wilcoxon ranked sum test.
+Comparisons between different treatment for lesions utilized a Kruskal
+Wallis test. Where multiple comparison testing was appropriate, a
+Benjamini-Hochberg (BH) correction was applied {[}36{]} and a corrected
+P-value of less than 0.05 was considered significant. The P-values
+reported are those that were BH corrected. Model rank importance was
+determined by obtaining the median MDA from the 100, 20 repeated 10-fold
+cross validation and then ranking from largest to smallest MDA.
 
@@
 respective SRA study accession.
-
-\newpage
-
-\textbf{Figure 1: General differences between adenoma, advanced adenoma,
-and carcinoma groups after treatment.} A) Thetayc distance from pre
-versus post sample within each individual. A significant difference was
-found between the adenoma and carcinoma group for thetayc (P-value =
-5.36e-05). Solid black points represent the median value for each
-diagnosis group. B) NMDS of the pre and post-treatment samples for the
-adenoma group. C) NMDS of the pre and post-treatment samples for the
-advanced adenoma group. D) NMDS of the pre and post-treatment samples
-for the carcinoma group.
-
-\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
-advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
-OTUs. C) Carcinoma OTUs.
-
-\textbf{Figure 3: OTUs common to those models used to differentiate
-between patients with normal colons and those with adenoma, advanced
-adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
-each model. B) For each common OTU the lowest taxonomic identification
-and importance rank for each model run is shown.
-
-\textbf{Figure 4: Treatment response based on models built for adenoma,
-advanced adenoma, or carcinoma.} A) Positive probability change from
-initial to follow up sample in those with adenoma. B) Positive
-probability change from initial to follow up sample in those with
-advanced adenoma. C) Positive probability change from initial to follow
-up sample in those with carcinoma.
 
@@
 350.85 ± 102\tabularnewline
+Surgery Only & 4 & 4 & 12\tabularnewline
+Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
+Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
+Pre-Treatment Antibiotic Use & 0 & 0 & 0\tabularnewline
 \bottomrule
@@
 
-\textbf{Figure S1: ROC curves of the adenoma, advanced adenoma, and
+\textbf{Figure 1: General differences between adenoma, advanced adenoma,
+and carcinoma groups after treatment.} A) \(\theta\)\textsubscript{YC}
+distance from pre versus post sample within each individual. A
+significant difference was found between the adenoma and carcinoma group
+for thetayc (P-value = 5.36e-05). Solid black points represent the
+median value for each diagnosis group. B) NMDS of the pre and
+post-treatment samples for the adenoma group. C) NMDS of the pre and
+post-treatment samples for the advanced adenoma group. D) NMDS of the
+pre and post-treatment samples for the carcinoma group.
+
+\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
+advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
+OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
+MDA value obtained from 100 different 80/20 splits while the lighter
+colored circles represents the value obtained for a specific run.
+
+\textbf{Figure 3: OTUs common to those models used to differentiate
+between patients with normal colons and those with adenoma, advanced
+adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
+each model. B) For each common OTU the lowest taxonomic identification
+and importance rank for each model run is shown.
+
+\textbf{Figure 4: Treatment response based on models built for adenoma,
+advanced adenoma, or carcinoma.} A) Positive probability change from
+initial to follow up sample in those with adenoma. B) Positive
+probability change from initial to follow up sample in those with
+advanced adenoma. C) Positive probability change from initial to follow
+up sample in those with carcinoma.
+
+\newpage
+
+\textbf{Figure S1: Surgery comparisons by diagnosis using
+\(\theta\)\textsubscript{YC} distance.} A) Differences by diagnosis
+group for pre versus post sample within each individual. No significant
+difference was found but the trend is similar to that found for the full
+data set. B) Difference between those receiving surgery and those that
+did not for the adenoma group. No significant difference was observed.
+C) Difference between those receiving surgery and those that did not for
+the advanced adenoma group. No significant difference was observed.
+
+\textbf{Figure S2: ROC curves of the adenoma, advanced adenoma, and
 carcinoma models.} A) Adenoma ROC curve: The light green shaded areas
@@
 
-\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
+\textbf{Figure S3: Summary of important OTUs for the adenoma, advanced
 adenoma, and carcinoma models.} A) MDA of the most important variables
@@
 
-\textbf{Figure S3: Pre and post-treatment relative abundance of CRC
+\textbf{Figure S4: Pre and post-treatment relative abundance of CRC
 associated OTUs within the carcinoma model.}
@@
 
+\hypertarget{ref-schloss_dynamics_2014}{}
+27. Schloss PD, Iverson KD, Petrosino JF, Schloss SJ. The dynamics of a
+family's gut microbiota reveal variations on a theme. Microbiome.
+2014;2:25.
+
 \hypertarget{ref-obrien_impact_2013}{}
-27. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+28. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
 bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
@@
 \hypertarget{ref-kozich_development_2013}{}
-28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+29. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-pruesse_silva_2007}{}
-29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
+30. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
 SILVA: A comprehensive online resource for quality checked and aligned
@@
 \hypertarget{ref-edgar_uchime_2011}{}
-30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
+31. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
 sensitivity and speed of chimera detection. Bioinformatics (Oxford,
@@
 \hypertarget{ref-wang_naive_2007}{}
-31. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
+32. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
 for rapid assignment of rRNA sequences into the new bacterial taxonomy.
@@
 \hypertarget{ref-schloss_assessing_2011}{}
-32. Schloss PD, Westcott SL. Assessing and improving methods used in
+33. Schloss PD, Westcott SL. Assessing and improving methods used in
 operational taxonomic unit-based approaches for 16S rRNA gene sequence
@@
 \hypertarget{ref-breiman_random_2001}{}
-33. Breiman L. Random Forests. Machine Learning. 2001;45:5--32.
+34. Breiman L. Random Forests. Machine Learning. 2001;45:5--32.
 
 \hypertarget{ref-anderson_permanova_2013}{}
-34. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
 the face of heterogeneous dispersions: What null hypothesis are you
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-35. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2017-07-21 submission/manuscript.tex

@@
 \textbf{Results.} There were small changes to the bacterial community
-associated with adenoma or advanced adenoma groups and large changes
-associated with the carcinoma group. The communities from patients with
-carcinomas changed significantly more than those with adenoma following
-treatment (P-value \textless{} 0.001). There was no significant change
-in the microbiota between patients with adenoma and advanced adenoma, or
-between patients with advanced adenoma and carcinoma (P-value
-\textgreater{} 0.05). Although treatment was associated with
+associated with adenoma or advanced adenoma and large changes associated
+with the carcinoma. The communities from patients with carcinomas
+changed significantly more than those with adenoma following treatment
+(P-value \textless{} 0.001). Although treatment was associated with
 intrapersonal changes, the change in the abundance of individual OTUs to
@@
 learning algorithm to try and identify groups of OTUs that would allow
-us to successfully distinguish between pre and post-treatment samples
-for each of the diagnosis groups. Although the three models could
-successfully differentiate between the pre and post-treatment samples,
-there was little overlap between the OTUs that were indicative of
-treatment and randomization of the diagnosis labels recapitulated the
-actual models AUCs. Next, we used a larger cohort that contained
-individuals with normal colons and those with adenomas, advanced
-adenomas, and carcinomas to determine whether individuals who underwent
-treatment were more likely to have OTUs associated with normal colons.
-We again built Random Forest models and measured the change in the
-positive probability of having one of the three diagnoses to assess
-whether these changes were still similar to their original diagnosis or
-not. Only patients who had carcinomas had a change that was towards a
-microbial milieu that resembles the normal colon (P-value \textless{}
-0.05). Finally, we tested whether the type of treatment impacted the
-microbiota of those diagnosed with carcinomas and were unable to detect
-any significant differences in characteristics of these communities
-between individuals treated with surgery alone and those treated with
+us to distinguish between pre and post-treatment samples for each of the
+diagnosis groups. Although the three models could differentiate between
+the pre and post-treatment samples, there was little overlap between the
+OTUs that were indicative of treatment. Next, we used a larger cohort
+that contained individuals with normal colons and those with adenomas,
+advanced adenomas, and carcinomas to determine whether individuals who
+underwent treatment were more likely to have OTUs associated with normal
+colons. We again built Random Forest models and measured the change in
+the positive probability of having one of the three diagnoses to assess
+whether these changes were similar to their original diagnosis or not.
+Patients who had carcinomas changed towards a microbial milieu that
+resembles the normal colon (P-value \textless{} 0.05). Finally, we were
+unable to detect any significant differences in the microbiota between
+individuals treated with surgery alone and those treated with
 chemotherapy or chemotherapy and radiation (P-value \textgreater{}

### 2017-07-21 submission/manuscript.Rmd

@@
 
-**Background.** Colorectal cancer (CRC) is a worldwide health problem.
-Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiota after treatment for an adenoma or carcinoma.
+**Background.** Colorectal cancer is a worldwide health problem.
+Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
 This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
@@
 
-**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma groups and large changes associated with the carcinoma group.
+**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with the carcinoma.
 The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-There was no significant change in the microbiota between patients with adenoma and advanced adenoma, or between patients with advanced adenoma and carcinoma (P-value > 0.05).
 Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to try and identify groups of OTUs that would allow us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups.
-Although the three models could successfully differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment and randomization of the diagnosis labels recapitulated the actual models AUCs.
+Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to try and identify groups of OTUs that could distinguish between pre and post-treatment samples for each of the diagnosis groups.
+Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment.
 Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons.
-We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether these changes were still similar to their original diagnosis or not.
-Only patients who had carcinomas had a change that was towards a microbial milieu that resembles the normal colon (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether these changes were similar to their original diagnosis or not.
+Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Finally, we were unable to detect any significant differences in the microbiota between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 

### 2017-07-21 submission/manuscript_R1.Rmd
incorporate pat suggestion on Radomized labels

--- 2017-07-21 submission/manuscript.Rmd
@@
 Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
-None of the individuals in this study had any recorded antibiotic use before treatment of their respective lesion [Table 1].
+None of the individuals in this study had any recorded antibiotic use before treatment of their respective lesion.
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
@@
 When analyzing only those who underwent surgery as a treatment these trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
+Since surgery is another situation where antibiotics are used we tested our $\theta$~YC~ beta diversity in these individuals only.
 There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
@@
 ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
-             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation, ttemptable$antibiotics)
+             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation)
 
@@
 colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label, 
-                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation", 
-                   "Pre-Treatment Antibiotic Use")
+                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation")
 

### 2017-07-21 submission/manuscript.Rmd
saving original submission

--- 2017-07-21 submission/manuscript.Rmd
@@
 # Extra data and meta data needed
-good_metaf <- read.csv('../data/process/mod_metadata/metaf_final.csv', header = T, stringsAsFactors = F)
+good_metaf <- read.csv('../data/process/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
 metaI <- read.csv('../data/process/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
@@
 all_OTU_pair_wilcox <- read.csv("../data/process/tables/OTU_paired_wilcoxson_test.csv", header = T)
-
-# Antibiotic test data
-surg_only_comparisons <- read.csv("../data/process/tables/surgery_only_comparisons.csv", header = T)
-surgVnonsurg_noncrc <- read.csv("../data/process/tables/surgVnonsurg_noncrc.csv", header = T)
-surgVnonsurg_noncrc_pvalues <- read.csv("../data/process/tables/surgVnonsurg_noncrc_pvalues.csv", header = T)
 
@@
 adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)
-adn_randomized_summary <- read.csv('../data/process/tables/reduced_adn_randomization_treatment_ROC_model_summary.csv', header = T)
 
@@
 srn_general_summary <- read.csv('../data/process/tables/srn_reduced_auc_summary.csv', header = T, row.names = 1)
-srn_randomized_summary <- read.csv('../data/process/tables/reduced_srn_randomization_treatment_ROC_model_summary.csv', header = T)
 
@@
 crc_probs <- read.csv('../data/process/tables/crc_reduced_follow_up_probability_summary.csv', header = T)
-crc_randomized_summary <- read.csv('../data/process/tables/reduced_crc_randomization_treatment_ROC_model_summary.csv', header = T)
 
@@
 Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
-None of the individuals in this study had any recorded antibiotic use before treatment for their adenoma, advanced adenoma, and carcinoma.
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
@@
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
-When analyzing only those who underwent surgery as a treatment these trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
-There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
@@
 It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
-We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-To try and limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
+To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
+To limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
 Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
-However, upon randomizing the pre and post treatment labels we found that Random Forest could work equally well (adenoma AUC range = `r format(min(adn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_randomized_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_randomized_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_randomized_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
-Although treatment had an impact on the overall community structure; the effect of treatment was not consistent across patients and diagnosis groups.
+These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.
 
@@
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing are marked decrease in relative abundance [Figure S4].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing are marked decrease in relative abundance [Figure S3].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
 A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon.
-There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3)`).
+There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4].
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals.
 Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
@@
 Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`).
-Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Furthermore, there was no significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
-Finally, the change in the positive probability was also not significantly different between the any treatment group (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power.
+We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Finally, the change in the positive probability was not significantly different between the two treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
 Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
@@
 For all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma, may be due to normal temporal variation but those with advanced adenoma and CRC clearly had large microbiota changes that would be greater than what could be expected from this type of variation [@schloss_dynamics_2014].
+Some of these changes, specifically for adenoma and advanced adenoma, may be due to normal temporal variation but those with CRC clearly had large microbiota changes that would be greater than what could be expected from this type of variation.
 After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
@@
 It is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
-This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S2].
+This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1].
 Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
@@
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-Although we cannot make a definitive statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, small effect sizes, potential changes in diet after treatment, changing medical care for chronic maladies, etc.
-in the adenoma and advnaced adenoma groups, there are still a number of exciting new potential avenues to explore.
+Although we cannot make a specific statment regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, there are still a number of exciting new avenues to explore.
 One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
@@
 ***Treatment***
-The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1] but rather via colonoscopy.
-All CRC individuals were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
+The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1].
 The type of chemotherapy used for patients with CRC varied markably.
@@
 Comparisons between probabilities as well as overall differences in the median relative abundance of each OTU between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test.
-Comparisons between different treatment for lesions utilized a Kruskal Wallis test.
 Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
@@
 
+**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) Thetayc distance from pre versus post sample within each individual.
+A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
+Solid black points represent the median value for each diagnosis group.
+B) NMDS of the pre and post-treatment samples for the adenoma group.
+C) NMDS of the pre and post-treatment samples for the advanced adenoma group.
+D) NMDS of the pre and post-treatment samples for the carcinoma group.
+
+**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
+B) Advanced Adenoma OTUs.
+C) Carcinoma OTUs.
+The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
+
+**Figure 3: OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
+B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
+
+**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
+B) Positive probability change from initial to follow up sample in those with advanced adenoma.
+C) Positive probability change from initial to follow up sample in those with carcinoma.
+
+\newpage
+
 **Table 1: Demographic data of patients in the pre and post-treatment cohort**
@@
 ```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
-test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received, Abx)
+test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received)
 
-t2 <- as.data.frame(matrix(nrow = 10, ncol = 3, 
+t2 <- as.data.frame(matrix(nrow = 9, ncol = 3, 
                            dimnames = 
                              list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time",  
-                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation", "antibiotics"), 
+                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation"), 
                                   coln = c("Adenoma", "SRN", "Carcinoma"))))
@@
   t2["Surg_Chemo_Radiation", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & radiation_received == "yes") %>% select(radiation_received) %>% count()
-  t2["antibiotics", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Abx > 0) %>% select(Abx) %>% count()
-
 }
@@
                                         Surg_Chemo = format(Surg_Chemo, digits = 2, nsmall = 0), 
-                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0), 
-                                        antibiotics = format(antibiotics, digits = 2, nsmall = 0))
+                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0))
 
 ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
-             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation, ttemptable$antibiotics)
+             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation)
 
@@
 
-**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) $\theta$~YC~ distance from pre versus post sample within each individual.
-A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
-Solid black points represent the median value for each diagnosis group.
-B) NMDS of the pre and post-treatment samples for the adenoma group.
-C) NMDS of the pre and post-treatment samples for the advanced adenoma group.
-D) NMDS of the pre and post-treatment samples for the carcinoma group.
-
-**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
-B) Advanced Adenoma OTUs.
-C) Carcinoma OTUs.
-The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
-
-**Figure 3: OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
-B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
-
-**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
-B) Positive probability change from initial to follow up sample in those with advanced adenoma.
-C) Positive probability change from initial to follow up sample in those with carcinoma.
-
-\newpage
-
-**Figure S1: Surgery comparisons by diagnosis using $\theta$~YC~ distance.** A) Differences by diagnosis group for pre versus post sample within each individual.
-No significant difference was found but the trend is similar to that found for the full data set.
-B) Difference between those receiving surgery and those that did not for the adenoma group.
-No significant difference was observed.
-C) Difference between those receiving surgery and those that did not for the advanced adenoma group.
-No significant difference was observed.
-
-**Figure S2: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S1: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
@@
 
-**Figure S3: Summary of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
+**Figure S2: Summary of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
@@
 
-**Figure S4: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**
+**Figure S3: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**
 

### 2017-07-22 submission/manuscript.tex
add back old version

--- 2017-05-16 submission/manuscript.tex
@@
 \usepackage{lineno}
-
+\usepackage{microtype}
 \usepackage{pdfpages}
@@
 of the treatments on the amount of change in the community structure
-(P-value = 0.298). Finally, the change in the positive probability was
+(P-value = 0.532). Finally, the change in the positive probability was
 not significantly different between the two treatment groups (P-value =
-0.999). Due to the relatively small number of samples in each treatment
+0.532). Due to the relatively small number of samples in each treatment
 group, it was difficult to make a definitive statement regarding the
@@
 all three groups of patients, we observed changes in their microbiota.
-After treatment, the microbiota of patients with carcinoma changed
-significantly more than the other groups. This change resulted in
-communities that more closely resembled those of patients with a normal
-colon. This may suggest that treatment for carcinoma is not only
-successful for removing the carcinoma but also at reducing the
-associated bacterial communities. Understanding the effect of treatment
-on the microbiota of those diagnosed with carcinomas may have important
-implications for reducing disease recurrence. It is intriguing that it
-may be possible to use microbiome-based biomarkers to not only predict
-the presence of lesions but to assess the risk of recurrence.
+Some of these changes, specifically for adenoma and advanced adenoma,
+may be due to normal temporal variation but those with CRC clearly had
+large microbiota changes that would be greater than what could be
+expected from this type of variation. After treatment, the microbiota of
+patients with carcinoma changed significantly more than the other
+groups. This change resulted in communities that more closely resembled
+those of patients with a normal colon. This may suggest that treatment
+for carcinoma is not only successful for removing the carcinoma but also
+at reducing the associated bacterial communities. Understanding the
+effect of treatment on the microbiota of those diagnosed with carcinomas
+may have important implications for reducing disease recurrence. It is
+intriguing that it may be possible to use microbiome-based biomarkers to
+not only predict the presence of lesions but to assess the risk of
+recurrence.
 
@@
 microbiome-based biomarkers to predict the presence of colonic lesions.
-The most exciting future direction from the current study is the
-possibility that markers within the microbiota could be used to evaluate
+Although we cannot make a specific statment regarding the effect of a
+specific treatment on the gut microbiota, due to the low number of
+samples, there are still a number of exciting new avenues to explore.
+One of the most exciting of these future directions is the possibility
+that markers within the microbiota could be used to potentially evaluate
 the effect of treatment and predict recurrence for those diagnosed with
 carcinoma. If such an approach is effective, it might be possible to
-target the microbiota as part of adjuvant therapy. Our data provides
+target the microbiota as part of adjuvant therapy, if the biomarkers
+identified play a key role in the disease process. Our data provides
 additional evidence on the importance of the microbiota in tumorigenesis
@@
 \textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
-have been previously reported in Baxter, et al {[}24{]}. Briefly, study
-exclusion involved those who had already undergone surgery, radiation,
-or chemotherapy, had colorectal cancer before a baseline fecal sample
-could be obtained, had IBD, a known hereditary non-polyposis colorectal
-cancer, or familial adenomatous polyposis. Samples used to build the
-models for prediction were collected either prior to a colonoscopy or
-between one and two weeks after initial colonoscopy. The bacterial
-community has been shown to normalize back to a pre-colonoscopy
-community within this time period {[}27{]}. Our study cohort consisted
-of 67 individuals with an initial sample as described and a follow up
-sample obtained between 188 - 546 days after treatment of lesion
-{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
-histopathological review of any biopsies taken. Patients were classified
-as having advanced adenoma if they had an adenoma greater than 1 cm,
-more than three adenomas of any size, or an adenoma with villous
-histology. This study was approved by the University of Michigan
-Institutional Review Board. All study participants provided informed
-consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+have been previously reported in Baxter, et al {[}24{]}. Briefly,
+samples were stored on ice for at least 24h before freezing. Although we
+cannot exclude that this sampling protocol may have impacted the gut
+microbiota composition all samples were subjected to the same
+methodology. Study exclusion involved those who had already undergone
+surgery, radiation, or chemotherapy, had colorectal cancer before a
+baseline fecal sample could be obtained, had IBD, a known hereditary
+non-polyposis colorectal cancer, or familial adenomatous polyposis.
+Samples used to build the models for prediction were collected either
+prior to a colonoscopy or between one and two weeks after initial
+colonoscopy. The bacterial community has been shown to normalize back to
+a pre-colonoscopy community within this time period {[}27{]}. Our study
+cohort consisted of 67 individuals with an initial sample as described
+and a follow up sample obtained between 188 - 546 days after treatment
+of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
+examination and histopathological review of any biopsies taken. Patients
+were classified as having advanced adenoma if they had an adenoma
+greater than 1 cm, more than three adenomas of any size, or an adenoma
+with villous histology. This study was approved by the University of
+Michigan Institutional Review Board. All study participants provided
+informed consent and the study itself conformed to the guidelines set
+out by the Helsinki Declaration.
+
+\textbf{\emph{Treatment}} The majority of patients undergoing treatment
+for adenoma or advanced adenoma were not treated surigically {[}Table
+1{]}. The type of chemotherapy used for patients with CRC varied
+markably. The types used included Oxaliplatin, Levicovorin, Folfox,
+Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were
+used individually or in combination with others depending on the patient
+{[}Table 1{]}. If an individual was treated with radiation they were
+also always treated with chemotherapy. Radiation therapy generally used
+18 mV photons for treatment.
 
@@
 Similarily, the Random Forest {[}33{]} algorithm was also used to create
-the three models used to classify normal versus diagnosis. The total
-number of individuals in the normal versus diagnosis models was 423
-individuals {[}Table 2{]}. There were a total of 239 individuals in the
-normal versus adenoma model, 262 individuals in the normal versus
-advanced adenoma model, and 266 individuals in the normal versus
-carcinoma model {[}Table 2{]}.
+the three models used to classify normal versus diagnosis. These samples
+were obtained using the same methodology as described earlier in this
+section. All samples used for this component of model training were from
+pre-treatment samples. The total number of individuals in the normal
+versus diagnosis models was 423 individuals {[}Table 2{]}. There were a
+total of 239 individuals in the normal versus adenoma model, 262
+individuals in the normal versus advanced adenoma model, and 266
+individuals in the normal versus carcinoma model {[}Table 2{]}.
 
 All models included only OTU data obtained from 16S rRNA sequencing and
-were processed and cleaned using the R package caret (v6.0.73).
+were processed and cleaned using the R package caret (v6.0.76).
 Optimization of the mtry hyper-parameter involved making 100 different
@@
 
-\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
+\textbf{\emph{Statistical Analysis.}} The R software package (v3.4.1)
 was used for all statistical analysis. Comparisons between bacterial
 community structure utilized PERMANOVA {[}34{]} in the vegan package
-(v2.4.1). Comparisons between probabilities as well as overall
+(v2.4.3). Comparisons between probabilities as well as overall
 differences in the median relative abundance of each OTU between pre and
@@
 advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
-OTUs. C) Carcinoma OTUs.
+OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
+MDA value obtained from 100 different 80/20 splits while the lighter
+colored circles represents the value obtained for a specific run.
 
@@
 350.85 ± 102\tabularnewline
+Surgery Only & 4 & 4 & 12\tabularnewline
+Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
+Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
 \bottomrule

### 2017-07-22 submission/manuscript_R1.Rmd
add revision and marked up version

@@
 Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
-None of the individuals in this study had any recorded antibiotic use before treatment of their respective lesion.
+None of the individuals in this study had any recorded antibiotic usage that was not associated with surgical treatment of their respective lesion.
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
@@
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
-When analyzing only those who underwent surgery as a treatment these trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
-Since surgery is another situation where antibiotics are used we tested our $\theta$~YC~ beta diversity in these individuals only.
+Since surgery is a procedure where antibiotics are used, we tested our $\theta$~YC~ beta diversity in these individuals only.
+Although not significant the trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
 There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
@@
 Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
-However, upon randomizing the pre and post treatment labels we found that Random Forest could work equally well (adenoma AUC range = `r format(min(adn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_randomized_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_randomized_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_randomized_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].

### 2017-07-22 submission/manuscript.Rmd

@@
 Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota.
-If the microbial community has an affect on recurrence risk or tumorigenesis it would be reasonable to expect treatment to remove a lesion to affect the microbiota.
+If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence.
 To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner?
@@
 For all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma and advanced adenoma, may be due to normal temporal variation but those with CRC clearly had large microbiota changes that would be greater than what could be expected from this type of variation.
 After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
@@
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-Although we cannot make a specific statment regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, there are still a number of exciting new avenues to explore.
-One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
-If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy, if the biomarkers identified play a key role in the disease process.
+The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
+If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy.
 Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.
@@
 Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016].
-Briefly, samples were stored on ice for at least 24h before freezing.
-Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology.
-Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
+Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
 Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy.
@@
 All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
-
-***Treatment***
-The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1].
-The type of chemotherapy used for patients with CRC varied markably.
-The types used included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
-These were used individually or in combination with others depending on the patient [Table 1].
-If an individual was treated with radiation they were also always treated with chemotherapy.
-Radiation therapy generally used 18 mV photons for treatment.
 
@@
 Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis.
-These samples were obtained using the same methodology as described earlier in this section.
-All samples used for this component of model training were from pre-treatment samples.
 The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2].
@@
 
-**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
-B) Advanced Adenoma OTUs.
-C) Carcinoma OTUs.
-The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
+**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs.
 
@@
 ```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
-test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received)
+test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time)
 
-t2 <- as.data.frame(matrix(nrow = 9, ncol = 3, 
+t2 <- as.data.frame(matrix(nrow = 6, ncol = 3,
                            dimnames =
-                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time",  
-                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation"), 
+                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time"),
                                   coln = c("Adenoma", "SRN", "Carcinoma"))))
@@
   t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
+
   t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))
-  t2["Surgery_only", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Surgery == "Y" & chemo_received == "no") %>% select(Surgery) %>% count()
-  t2["Surg_Chemo", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & chemo_received == "yes" & radiation_received == "no") %>% select(chemo_received) %>% count()
-  t2["Surg_Chemo_Radiation", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & radiation_received == "yes") %>% select(radiation_received) %>% count()
+
 }
@@
   post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])
+
 }
@@
                                                     label = "\u00B1",
-                                                    format(post_time_sd, digits = 2, nmsall = 2)), 
-                                        Surgery_only = format(Surgery_only, digits = 2, nsmall = 0), 
-                                        Surg_Chemo = format(Surg_Chemo, digits = 2, nsmall = 0), 
-                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0))
+                                                    format(post_time_sd, digits = 2, nmsall = 2)))
 
 ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f,
-             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation)
+             ttemptable$BMI, ttemptable$White, ttemptable$post_time)
 
@@
 
-colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label, 
-                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation")
+colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label)
 

### 2017-08-04 submission/manuscript.tex

@@
 including resection, chemotherapy, and/or radiation affect the
-composition of the gut microbiota. If the microbial community has an
-affect on recurrence risk or tumorigenesis it would be reasonable to
-expect treatment to remove a lesion to affect the microbiota. To test
-this hypothesis, we addressed two related questions: Does treatment
-affect the colonic microbiota in a predictable manner? If so, does the
+composition of the gut microbiota. If the microbial community drives
+tumorigenesis then one would hypothesize that treatment to remove a
+lesion would affect the microbiota and risk of recurrence. To test this
+hypothesis, we addressed two related questions: Does treatment affect
+the colonic microbiota in a predictable manner? If so, does the
 treatment alter the community to more closely resemble that of
@@
 all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma and advanced adenoma,
-may be due to normal temporal variation but those with CRC clearly had
-large microbiota changes that would be greater than what could be
-expected from this type of variation. After treatment, the microbiota of
-patients with carcinoma changed significantly more than the other
-groups. This change resulted in communities that more closely resembled
-those of patients with a normal colon. This may suggest that treatment
-for carcinoma is not only successful for removing the carcinoma but also
-at reducing the associated bacterial communities. Understanding the
-effect of treatment on the microbiota of those diagnosed with carcinomas
-may have important implications for reducing disease recurrence. It is
-intriguing that it may be possible to use microbiome-based biomarkers to
-not only predict the presence of lesions but to assess the risk of
-recurrence.
+After treatment, the microbiota of patients with carcinoma changed
+significantly more than the other groups. This change resulted in
+communities that more closely resembled those of patients with a normal
+colon. This may suggest that treatment for carcinoma is not only
+successful for removing the carcinoma but also at reducing the
+associated bacterial communities. Understanding the effect of treatment
+on the microbiota of those diagnosed with carcinomas may have important
+implications for reducing disease recurrence. It is intriguing that it
+may be possible to use microbiome-based biomarkers to not only predict
+the presence of lesions but to assess the risk of recurrence.
 
@@
 microbiome-based biomarkers to predict the presence of colonic lesions.
-Although we cannot make a specific statment regarding the effect of a
-specific treatment on the gut microbiota, due to the low number of
-samples, there are still a number of exciting new avenues to explore.
-One of the most exciting of these future directions is the possibility
-that markers within the microbiota could be used to potentially evaluate
+The most exciting future direction from the current study is the
+possibility that markers within the microbiota could be used to evaluate
 the effect of treatment and predict recurrence for those diagnosed with
 carcinoma. If such an approach is effective, it might be possible to
-target the microbiota as part of adjuvant therapy, if the biomarkers
-identified play a key role in the disease process. Our data provides
+target the microbiota as part of adjuvant therapy. Our data provides
 additional evidence on the importance of the microbiota in tumorigenesis
@@
 \textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
-have been previously reported in Baxter, et al {[}24{]}. Briefly,
-samples were stored on ice for at least 24h before freezing. Although we
-cannot exclude that this sampling protocol may have impacted the gut
-microbiota composition all samples were subjected to the same
-methodology. Study exclusion involved those who had already undergone
-surgery, radiation, or chemotherapy, had colorectal cancer before a
-baseline fecal sample could be obtained, had IBD, a known hereditary
-non-polyposis colorectal cancer, or familial adenomatous polyposis.
-Samples used to build the models for prediction were collected either
-prior to a colonoscopy or between one and two weeks after initial
-colonoscopy. The bacterial community has been shown to normalize back to
-a pre-colonoscopy community within this time period {[}27{]}. Our study
-cohort consisted of 67 individuals with an initial sample as described
-and a follow up sample obtained between 188 - 546 days after treatment
-of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
-examination and histopathological review of any biopsies taken. Patients
-were classified as having advanced adenoma if they had an adenoma
-greater than 1 cm, more than three adenomas of any size, or an adenoma
-with villous histology. This study was approved by the University of
-Michigan Institutional Review Board. All study participants provided
-informed consent and the study itself conformed to the guidelines set
-out by the Helsinki Declaration.
-
-\textbf{\emph{Treatment}} The majority of patients undergoing treatment
-for adenoma or advanced adenoma were not treated surigically {[}Table
-1{]}. The type of chemotherapy used for patients with CRC varied
-markably. The types used included Oxaliplatin, Levicovorin, Folfox,
-Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were
-used individually or in combination with others depending on the patient
-{[}Table 1{]}. If an individual was treated with radiation they were
-also always treated with chemotherapy. Radiation therapy generally used
-18 mV photons for treatment.
+have been previously reported in Baxter, et al {[}24{]}. Briefly, study
+exclusion involved those who had already undergone surgery, radiation,
+or chemotherapy, had colorectal cancer before a baseline fecal sample
+could be obtained, had IBD, a known hereditary non-polyposis colorectal
+cancer, or familial adenomatous polyposis. Samples used to build the
+models for prediction were collected either prior to a colonoscopy or
+between one and two weeks after initial colonoscopy. The bacterial
+community has been shown to normalize back to a pre-colonoscopy
+community within this time period {[}27{]}. Our study cohort consisted
+of 67 individuals with an initial sample as described and a follow up
+sample obtained between 188 - 546 days after treatment of lesion
+{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
+histopathological review of any biopsies taken. Patients were classified
+as having advanced adenoma if they had an adenoma greater than 1 cm,
+more than three adenomas of any size, or an adenoma with villous
+histology. This study was approved by the University of Michigan
+Institutional Review Board. All study participants provided informed
+consent and the study itself conformed to the guidelines set out by the
+Helsinki Declaration.
 
@@
 Similarily, the Random Forest {[}33{]} algorithm was also used to create
-the three models used to classify normal versus diagnosis. These samples
-were obtained using the same methodology as described earlier in this
-section. All samples used for this component of model training were from
-pre-treatment samples. The total number of individuals in the normal
-versus diagnosis models was 423 individuals {[}Table 2{]}. There were a
-total of 239 individuals in the normal versus adenoma model, 262
-individuals in the normal versus advanced adenoma model, and 266
-individuals in the normal versus carcinoma model {[}Table 2{]}.
+the three models used to classify normal versus diagnosis. The total
+number of individuals in the normal versus diagnosis models was 423
+individuals {[}Table 2{]}. There were a total of 239 individuals in the
+normal versus adenoma model, 262 individuals in the normal versus
+advanced adenoma model, and 266 individuals in the normal versus
+carcinoma model {[}Table 2{]}.
 
@@
 advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
-OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
-MDA value obtained from 100 different 80/20 splits while the lighter
-colored circles represents the value obtained for a specific run.
+OTUs. C) Carcinoma OTUs.
 
@@
 350.85 ± 102\tabularnewline
-Surgery Only & 4 & 4 & 12\tabularnewline
-Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
-Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
 \bottomrule

### 2017-08-04 submission/manuscript_R1.Rmd

@@
 
-**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with the carcinoma.
+**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma.
 The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
@@
 We found that treatment alters the composition of the gut microbiota and that, for those with carinomas, the gut microbiota shifted more towards that of a normal colon after treatment.
-In the individuals with carcinomas, no difference was found by the type of treatment (surgery alone versus surgery with chemotherapy).
+In the individuals with carcinomas, no difference was found by the type of treatment (surgery alone, surgery with chemotherapy, surgery with chemotherapy and radiation).
 Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence and the likelihood of survival.
@@
 Although not significant the trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
-There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
+There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma groups (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
@@
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing are marked decrease in relative abundance [Figure S4].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing a marked decrease in relative abundance [Figure S4].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
@@
 For all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma, may be due to normal temporal variation but those with advanced adenoma and CRC clearly had large microbiota changes that would be greater than what could be expected from this type of variation [@schloss_dynamics_2014].
+Some of these changes, specifically for adenoma, may be due to normal temporal variation but those with advanced adenoma and carcinoma clearly had large microbiota changes that would be greater than what could be expected from this type of variation [@schloss_dynamics_2014].
 After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
@@
 Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.
-It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to also assess the risk of recurrence.
+It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to also assess the risk of recurrence due to these changes in the microbiota.
 
@@
 However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["25%"], digits = 0)` - `r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["75%"], digits = 0)`) days).
+We also found no difference $\theta$~YC~ in the adenoma or advanced adenoma group between those that received surgery and those that did not [Figure S1].
 
@@
 
-***Treatment***
-For this study treatment refers specifically to the removal of a lesion or with or without chemotherapy and radiation.
+***Treatment.***
+For this study treatment refers specifically to the removal of a lesion with or without chemotherapy and radiation.
 The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1] but rather via colonoscopy.
-All CRC individuals were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
+All carcinoma individuals were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
 The type of chemotherapy used for patients with CRC varied markably.
@@
 These samples were obtained using the same methodology as described earlier in this section.
-All samples used for this component of model training were from pre-treatment samples.
+All samples used for this component of model training were from pre-treatment samples only.
 The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2].

### 2017-08-12 submission/manuscript_R1.Rmd
update abstract with Pat's comments

@@
 The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to try and identify groups of OTUs that could distinguish between pre and post-treatment samples for each of the diagnosis groups.
-Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment.
-Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons.
-We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether these changes were similar to their original diagnosis or not.
-Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-Finally, we were unable to detect any significant differences in the microbiota between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
+Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment.
+To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treament samples.
+Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-**Conclusions.** Although it was difficult to identify significant differences between pre and post treatment samples for adenoma and advanced adenoma, this was not the case for carcinomas.
-Not only were there large changes in pre versus post treatment samples for those with carcinoma but also these changes were towards a more normal microbiota composition.
+**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
+Although it was difficult to identify significant differences between pre and post-treatment samples from patientes with adenoma and advanced adenoma, this was not the case for carcinomas.
+Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
 

### 2017-08-12 submission/manuscript_R1.Rmd
complete Pat edits upto Results

@@
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
-Since surgery is a procedure where antibiotics are used, we tested our $\theta$~YC~ beta diversity in these individuals only.
-Although not significant the trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A].
-There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma groups (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C].
 Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
@@
 We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-To try and limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
+To limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
 Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
-Although treatment had an impact on the overall community structure; the effect of treatment was not consistent across patients and diagnosis groups.
+Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups.
 
@@
 We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
+The models performed well (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
 The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
@@
 There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3)`).
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(round(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 2))`).
 Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
@@
 Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Furthermore, there was no significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
-Finally, the change in the positive probability was also not significantly different between any treatment group (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
+Finally, the change in the positive probability was not significantly different between any of the treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
 Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.

### 2017-08-12 submission/manuscript_R1.Rmd
add Pat's suggestions to discussion and conclusion

@@
 For all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma, may be due to normal temporal variation but those with advanced adenoma and carcinoma clearly had large microbiota changes that would be greater than what could be expected from this type of variation [@schloss_dynamics_2014].
+Some of these changes, specifically for adenoma, may be due to normal temporal variation, however, those with advanced adenoma and carcinoma clearly had large microbiota changes.
 After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
@@
 However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["25%"], digits = 0)` - `r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["75%"], digits = 0)`) days).
-We also found no difference $\theta$~YC~ in the adenoma or advanced adenoma group between those that received surgery and those that did not [Figure S1].
+We also found no difference in the community structure of those that received surgery and those that did not as a treatment for adenoma or advanced adenoma.
 
@@
 This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-Although in the adenoma and advanced adenoma groups we cannot make a definitive statement regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, small effect sizes, potential changes in diet after treatment, changing medical care for chronic maladies, etc.
-There are still a number of exciting new potential avenues to explore based on the carcinoma observations.
+We were surprised by the lack of a consistent signal that was associated with treatment of patients with adenomas or advanced adenomas.
+The lack of a large effect size may be due to differences in the role of bacteria in the formation of adnomas and carcinomas or it could be due to differences in the behaviors and medications within these classes of patients.
 One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.

### 2017-08-12 submission/manuscript_R1.Rmd
add Pat's method comments

@@
 The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1] but rather via colonoscopy.
-All carcinoma individuals were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
-The type of chemotherapy used for patients with CRC varied markably.
-The types used included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
+All patients diagnosed with carcinomas were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
+The type of chemotherapy used for patients with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
 These were used individually or in combination with others depending on the patient [Table 1].

### 2017-08-12 submission/manuscript_R1.Rmd
update figures with Pat's comments

@@
 
-**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) $\theta$~YC~ distance from pre versus post sample within each individual.
-A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
+**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) $\theta$~YC~ distances from pre versus post sample within each individual.
+A significant difference was found between the adenoma and carcinoma group (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
 Solid black points represent the median value for each diagnosis group.
@@
 
-**Figure S1: Surgery comparisons by diagnosis using $\theta$~YC~ distance.** A) Differences by diagnosis group for pre versus post sample within each individual.
-No significant difference was found but the trend is similar to that found for the full data set.
-B) Difference between those receiving surgery and those that did not for the adenoma group.
-No significant difference was observed.
-C) Difference between those receiving surgery and those that did not for the advanced adenoma group.
-No significant difference was observed.
-
-**Figure S2: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S1: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
 B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.

### 2017-08-12 submission/manuscript_R1.Rmd
update supplemental figure numbers

@@
 We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed well (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2).
+The models performed well (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1).
 The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
@@
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing a marked decrease in relative abundance [Figure S4].
+Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing a marked decrease in relative abundance [Figure S3].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.

### 2017-08-14 submission/manuscript_R1.tex
add tex versions

--- 2017-07-22 submission/manuscript.tex
@@
 
-\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
-problem. Despite growing evidence that members of the gut microbiota can
-drive tumorigenesis, little is known about what happens to the
-microbiota after treatment for an adenoma or carcinoma. This study
-tested the hypothesis that treatment for adenoma or carcinoma alters the
-abundance of bacterial populations associated with disease to those
-associated with a normal colon. We tested this hypothesis by sequencing
-the 16S rRNA genes in the feces of 67 individuals before and after
-treatment for adenoma (N = 22), advanced adenoma (N = 19), and carcinoma
-(N = 26).
+\textbf{Background.} Colorectal cancer is a worldwide health problem.
+Despite growing evidence that members of the gut microbiota can drive
+tumorigenesis, little is known about what happens to it after treatment
+for an adenoma or carcinoma. This study tested the hypothesis that
+treatment for adenoma or carcinoma alters the abundance of bacterial
+populations associated with disease to those associated with a normal
+colon. We tested this hypothesis by sequencing the 16S rRNA genes in the
+feces of 67 individuals before and after treatment for adenoma (N = 22),
+advanced adenoma (N = 19), and carcinoma (N = 26).
 
-\textbf{Results.} There were large changes to the bacterial communities
-associated with treatment across the three groups. The communities from
-patients with carcinomas changed significantly more than those with
-adenoma following treatment (P-value \textless{} 0.001). There was no
-significant change in the microbiota between patients with adenoma and
-advanced adenoma, or between patients with advanced adenoma and
-carcinoma (P-value \textgreater{} 0.05). Although treatment was
-associated with intrapersonal changes, the change in the abundance of
-individual OTUs to treatment was not consistent within diagnosis groups
-(P-value \textgreater{} 0.05). Because the distribution of OTUs across
-patients and diagnosis groups was irregular, we used the Random Forest
-machine learning algorithm to identify groups of OTUs that allowed us to
-successfully distinguish between pre and post-treatment samples for each
-of the diagnosis groups. Although the three models successfully
-differentiated between the pre and post-treatment samples, there was
-little overlap between the OTUs that were indicative of treatment. Next,
-we used a larger cohort that contained individuals with normal colons
-and those with adenomas, advanced adenomas, and carcinomas to determine
-whether individuals who underwent treatment were more likely to have
-OTUs associated with normal colons. We again built Random Forest models
-and measured the change in the positive probability of having one of the
-three diagnoses. Only patients who had carcinomas experienced a
-significant decrease in positive probability of having a lesion after
-treatment (P-value \textless{} 0.05), indicating that the microbial
-milieu of the colon more closely resembled that of a normal colon.
-Finally, we tested whether the type of treatment impacted the microbiota
-of those diagnosed with carcinomas and were unable to detect any
-significant differences in characteristics of these communities between
+\textbf{Results.} There were small changes to the bacterial community
+associated with adenoma or advanced adenoma and large changes associated
+with carcinoma. The communities from patients with carcinomas changed
+significantly more than those with adenoma following treatment (P-value
+\textless{} 0.001). Although treatment was associated with intrapersonal
+changes, the change in the abundance of individual OTUs in response to
+treatment was not consistent within diagnosis groups (P-value
+\textgreater{} 0.05). Because the distribution of OTUs across patients
+and diagnosis groups was irregular, we used the Random Forest machine
+learning algorithm to identify groups of OTUs that could be used to
+classify pre and post-treatment samples for each of the diagnosis
+groups. Although the three models could differentiate between the pre
+and post-treatment samples, there was little overlap between the OTUs
+that were indicative of each treatment. To determine whether individuals
+who underwent treatment were more likely to have OTUs associated with
+normal colons we used a larger cohort that contained individuals with
+normal colons and those with adenomas, advanced adenomas, and
+carcinomas. We again built Random Forest models and measured the change
+in the positive probability of having one of the three diagnoses to
+assess whether the post-treatment samples received the same
+classification as the pre-treament samples. Samples from patients who
+had carcinomas changed towards a microbial milieu that resembles the
+normal colon after treatment (P-value \textless{} 0.05). Finally, we
+were unable to detect any significant differences in the microbiota of
 individuals treated with surgery alone and those treated with
@@
 microbiota to treatment for adenomas and carcinomas, it is likely that
-biomarkers will be validated that can be used to quantify the risk of
-recurrence and the likelihood of survival.
+biomarkers will eventually be validated that can be used to quantify the
+risk of recurrence and the likelihood of survival. Although it was
+difficult to identify significant differences between pre and
+post-treatment samples from patientes with adenoma and advanced adenoma,
+this was not the case for carcinomas. Not only were there large changes
+in pre versus post-treatment samples for those with carcinoma, but these
+changes were towards a more normal microbiota.
 
@@
 colon after treatment. In the individuals with carcinomas, no difference
-was found by the type of treatment (surgery alone versus surgery with
-chemotherapy). Understanding how the community responds to these
-treatments could be a valuable tool for identifying biomarkers to
-quantify the risk of recurrence and the likelihood of survival.
+was found by the type of treatment (surgery alone, surgery with
+chemotherapy, surgery with chemotherapy and radiation). Understanding
+how the community responds to these treatments could be a valuable tool
+for identifying biomarkers to quantify the risk of recurrence and the
+likelihood of survival.
 
@@
 
-\textbf{\emph{Treatment alters the bacterial community structure of
-patients diagnosed with colonic lesions.}} Within our 67-person cohort
-we tested whether the microbiota of patients with adenoma (N = 22),
-advanced adenoma (N = 19), or carcinoma (N = 26) had any broad
-differences between pre and post-treatment samples {[}Table 1{]}. The
-structure of the microbial communities of the pre and post-treatment
-samples differed, as measured by the \(\theta\)\textsubscript{YC} beta
-diversity metric {[}Figure 1A{]}. We found that the communities obtained
-pre and post-treament among the patients with carcinomas changed
-significantly more than those patients with adenoma (P-value \textless{}
-0.001). There were no significant differences in the amount of change
-observed between the patients with adenoma and advanced adenoma or
-between the patients with advanced adenoma and carcinoma (P-value
-\textgreater{} 0.05). Next, we tested whether there was a consistent
-direction in the change in the community structure between the pre and
-post-treatment samples for each of the diagnosis groups {[}Figure
-1B-D{]}. We only observed a consistent shift in community structure for
-the patients with carcinoma when using a PERMANOVA test (adenoma P-value
-= 0.999, advanced adenoma P-value = 0.945, and carcinoma P-value =
-0.005). Finally, we measured the number of observed OTUs, Shannon
-evenness, and Shannon diversity in the pre and post-treatment samples
-and did not observe a significant change for any of the diagnosis groups
-(P-value \textgreater{} 0.05) {[}Table S1{]}.
+\textbf{\emph{Treatment for colonic lesions alters the bacterial
+community structure}} Within our 67-person cohort we tested whether the
+microbiota of patients with adenoma (N = 22), advanced adenoma (N = 19),
+or carcinoma (N = 26) had any broad differences between pre and
+post-treatment samples {[}Table 1{]}. None of the individuals in this
+study had any recorded antibiotic usage that was not associated with
+surgical treatment of their respective lesion. The structure of the
+microbial communities of the pre and post-treatment samples differed, as
+measured by the \(\theta\)\textsubscript{YC} beta diversity metric
+{[}Figure 1A{]}. We found that the communities obtained pre and
+post-treament among the patients with carcinomas changed significantly
+more than those patients with adenoma (P-value \textless{} 0.001). There
+were no significant differences in the amount of change observed between
+the patients with adenoma and advanced adenoma or between the patients
+with advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next,
+we tested whether there was a consistent direction in the change in the
+community structure between the pre and post-treatment samples for each
+of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
+shift in community structure for the patients with carcinoma when using
+a PERMANOVA test (adenoma P-value = 0.999, advanced adenoma P-value =
+0.945, and carcinoma P-value = 0.005). Finally, we measured the number
+of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
+post-treatment samples and did not observe a significant change for any
+of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.
 
-\textbf{\emph{The effects of treatment are not consistent across
-treatment groups.}} We used two approaches to identify those bacterial
+\textbf{\emph{The treatment of lesions are not consistent across
+diagnosis groups.}} We used two approaches to identify those bacterial
 populations that change between the two samples for each diagnosis
@@
 and the irregular distribution of OTUs across individuals limited the
-statistical power of the test. To overcome these problems we developed
-Random Forest models to identify collections of OTUs that would allow us
-to differentiate between pre and post-treatment samples from each of the
-diagnosis groups. To limit the likelihood that the models would overfit
-the data because of the relatively small number of subjects in each
-group, we restricted our models to only incorporate 10 OTUs. Despite
-this restriction, the models performed well (adenoma AUC range = 0.69 -
-0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
-0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
-had little overlap with each other {[}Figure 2{]}. These results support
-the earlier community-wide analysis where we observed that the treatment
-had an impact on the overall community structure; however, the effect of
-treatment was not consistent across patients and diagnosis groups.
+statistical power of the test. We attempted to overcome these problems
+by using Random Forest models to identify collections of OTUs that would
+allow us to differentiate between pre and post-treatment samples from
+each of the diagnosis groups. To limit the likelihood that the models
+would overfit the data because of the relatively small number of
+subjects in each group, we restricted our models to only incorporate 10
+OTUs. Despite this restriction, the models performed well (adenoma AUC
+range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
+AUC range = 0.82 - 0.98). Interestingly, the 10 OTUs that were used for
+each model had little overlap with each other {[}Figure 2{]}. Although
+treatment had an impact on the overall community structure, the effect
+of treatment was not consistent across patients and diagnosis groups.
 
@@
 (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
-models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
+models performed well (adenoma AUC range = 0.62 - 0.72, advanced adenoma
 AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
@@
 \emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
-showing are marked decrease in relative abundance {[}Figure S3{]}.
+showing a marked decrease in relative abundance {[}Figure S3{]}.
 Finally, we applied these three models to the pre and post-treatment
@@
 in the positive probability for the adenoma or advanced adenoma groups
-{[}Figure 4{]}. The positive probability for the pre and post-treatment
-samples from patients diagnosed with carcinoma significantly decreased
-with treatment, suggesting a shift toward a normal microbiota for most
-individuals. Only, 6 of the 26 patients (23.08\%) who were diagnosed
-with a carcinoma had a higher positive probability after treatment; one
-of those was re-diagnosed with carcinoma on the follow up visit. These
-results indicate that, although there were changes in the microbiota
-associated with treatment, those experienced by patients with carcinoma
-after treatment yielded gut bacterial communities of greater similarity
-to that of a normal colon.
+(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
+for the pre and post-treatment samples from patients diagnosed with
+carcinoma significantly decreased with treatment, suggesting a shift
+toward a normal microbiota for most individuals (P-value = 0.02). Only,
+6 of the 26 patients (23.08\%) who were diagnosed with a carcinoma had a
+higher positive probability after treatment; one of those was
+re-diagnosed with carcinoma on the follow up visit. These results
+indicate that, although there were changes in the microbiota associated
+with treatment, those experienced by patients with carcinoma after
+treatment yielded gut bacterial communities of greater similarity to
+that of a normal colon.
 
@@
 resection (N=12), surgical resection with chemotherapy (N=9), and
-surgical resection with chemotherapy and radiation (N=5). We focused on
-the patients with carcinoma and pooled those patients that received
-chemotherapy with those that received chemotherapy and radiation to
-improve our statistical power. We did not observe a significant
-difference in the effect of these treatments on the number of observed
-OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
-0.05). Furthermore, there was not a significant difference in the effect
-of the treatments on the amount of change in the community structure
-(P-value = 0.532). Finally, the change in the positive probability was
-not significantly different between the two treatment groups (P-value =
-0.532). Due to the relatively small number of samples in each treatment
-group, it was difficult to make a definitive statement regarding the
-specific type of treatment on the amount of change in the structure of
-the microbiota.
+surgical resection with chemotherapy and radiation (N=5). Regardless of
+treatment used there was no significant difference in the effect of
+these treatments on the number of observed OTUs, Shannon diversity, or
+Shannon evenness (P-value \textgreater{} 0.05). Furthermore, there was
+not a significant difference in the effect of the treatments on the
+amount of change in the community structure (P-value = 0.532). Finally,
+the change in the positive probability was not significantly different
+between any of the treatment groups (P-value = 0.532). Due to the
+relatively small number of samples in each treatment group, it was
+difficult to make a definitive statement regarding the specific type of
+treatment on the amount of change in the structure of the microbiota.
 
@@
 all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma and advanced adenoma,
-may be due to normal temporal variation but those with CRC clearly had
-large microbiota changes that would be greater than what could be
-expected from this type of variation. After treatment, the microbiota of
+Some of these changes, specifically for adenoma, may be due to normal
+temporal variation, however, those with advanced adenoma and carcinoma
+clearly had large microbiota changes. After treatment, the microbiota of
 patients with carcinoma changed significantly more than the other
@@
 intriguing that it may be possible to use microbiome-based biomarkers to
-not only predict the presence of lesions but to assess the risk of
-recurrence.
+not only predict the presence of lesions but to also assess the risk of
+recurrence due to these changes in the microbiota.
 
@@
 antibiotic administration on the microbiome would be expected to occur
-(344 (266 - 408) days).
+(344 (266 - 408) days). We also found no difference in the community
+structure of those that received surgery and those that did not as a
+treatment for adenoma or advanced adenoma.
 
@@
 microbiome-based biomarkers to predict the presence of colonic lesions.
-Although we cannot make a specific statment regarding the effect of a
-specific treatment on the gut microbiota, due to the low number of
-samples, there are still a number of exciting new avenues to explore.
-One of the most exciting of these future directions is the possibility
-that markers within the microbiota could be used to potentially evaluate
-the effect of treatment and predict recurrence for those diagnosed with
-carcinoma. If such an approach is effective, it might be possible to
-target the microbiota as part of adjuvant therapy, if the biomarkers
-identified play a key role in the disease process. Our data provides
-additional evidence on the importance of the microbiota in tumorigenesis
-by addressing the recovery of the microbiota after treatment and opens
-interesting avenues of research into how these changes may affect
-recurrence.
+We were surprised by the lack of a consistent signal that was associated
+with treatment of patients with adenomas or advanced adenomas. The lack
+of a large effect size may be due to differences in the role of bacteria
+in the formation of adnomas and carcinomas or it could be due to
+differences in the behaviors and medications within these classes of
+patients. One of the most exciting of these future directions is the
+possibility that markers within the microbiota could be used to
+potentially evaluate the effect of treatment and predict recurrence for
+those diagnosed with carcinoma. If such an approach is effective, it
+might be possible to target the microbiota as part of adjuvant therapy,
+if the biomarkers identified play a key role in the disease process. Our
+data provides additional evidence on the importance of the microbiota in
+tumorigenesis by addressing the recovery of the microbiota after
+treatment and opens interesting avenues of research into how these
+changes may affect recurrence.
 
@@
 
-\textbf{\emph{Treatment}} The majority of patients undergoing treatment
-for adenoma or advanced adenoma were not treated surigically {[}Table
-1{]}. The type of chemotherapy used for patients with CRC varied
-markably. The types used included Oxaliplatin, Levicovorin, Folfox,
-Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were
-used individually or in combination with others depending on the patient
+\textbf{\emph{Treatment.}} For this study treatment refers specifically
+to the removal of a lesion with or without chemotherapy and radiation.
+The majority of patients undergoing treatment for adenoma or advanced
+adenoma were not treated surigically {[}Table 1{]} but rather via
+colonoscopy. All patients diagnosed with carcinomas were treated with at
+least surgery or a combination of surgery and chemotherapy or surgery,
+chermotherapy, and radiation. The type of chemotherapy used for patients
+with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda,
+Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were used
+individually or in combination with others depending on the patient
 {[}Table 1{]}. If an individual was treated with radiation they were
@@
 section. All samples used for this component of model training were from
-pre-treatment samples. The total number of individuals in the normal
-versus diagnosis models was 423 individuals {[}Table 2{]}. There were a
-total of 239 individuals in the normal versus adenoma model, 262
+pre-treatment samples only. The total number of individuals in the
+normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
+were a total of 239 individuals in the normal versus adenoma model, 262
 individuals in the normal versus advanced adenoma model, and 266
@@
 differences in the median relative abundance of each OTU between pre and
-post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
-multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
-correction was applied {[}35{]} and a corrected P-value of less than
-0.05 was considered significant. The P-values reported are those that
-were BH corrected. Model rank importance was determined by obtaining the
-median MDA from the 100, 20 repeated 10-fold cross validation and then
-ranking from largest to smallest MDA.
+post-treatment samples utilized a paired Wilcoxon ranked sum test.
+Comparisons between different treatment for lesions utilized a Kruskal
+Wallis test. Where multiple comparison testing was appropriate, a
+Benjamini-Hochberg (BH) correction was applied {[}35{]} and a corrected
+P-value of less than 0.05 was considered significant. The P-values
+reported are those that were BH corrected. Model rank importance was
+determined by obtaining the median MDA from the 100, 20 repeated 10-fold
+cross validation and then ranking from largest to smallest MDA.
 
@@
 respective SRA study accession.
-
-\newpage
-
-\textbf{Figure 1: General differences between adenoma, advanced adenoma,
-and carcinoma groups after treatment.} A) Thetayc distance from pre
-versus post sample within each individual. A significant difference was
-found between the adenoma and carcinoma group for thetayc (P-value =
-5.36e-05). Solid black points represent the median value for each
-diagnosis group. B) NMDS of the pre and post-treatment samples for the
-adenoma group. C) NMDS of the pre and post-treatment samples for the
-advanced adenoma group. D) NMDS of the pre and post-treatment samples
-for the carcinoma group.
-
-\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
-advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
-OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
-MDA value obtained from 100 different 80/20 splits while the lighter
-colored circles represents the value obtained for a specific run.
-
-\textbf{Figure 3: OTUs common to those models used to differentiate
-between patients with normal colons and those with adenoma, advanced
-adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
-each model. B) For each common OTU the lowest taxonomic identification
-and importance rank for each model run is shown.
-
-\textbf{Figure 4: Treatment response based on models built for adenoma,
-advanced adenoma, or carcinoma.} A) Positive probability change from
-initial to follow up sample in those with adenoma. B) Positive
-probability change from initial to follow up sample in those with
-advanced adenoma. C) Positive probability change from initial to follow
-up sample in those with carcinoma.
 
@@
 \end{longtable}
+
+\newpage
+
+\textbf{Figure 1: General differences between adenoma, advanced adenoma,
+and carcinoma groups after treatment.} A) \(\theta\)\textsubscript{YC}
+distances from pre versus post sample within each individual. A
+significant difference was found between the adenoma and carcinoma group
+(P-value = 5.36e-05). Solid black points represent the median value for
+each diagnosis group. B) NMDS of the pre and post-treatment samples for
+the adenoma group. C) NMDS of the pre and post-treatment samples for the
+advanced adenoma group. D) NMDS of the pre and post-treatment samples
+for the carcinoma group.
+
+\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
+advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
+OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
+MDA value obtained from 100 different 80/20 splits while the lighter
+colored circles represents the value obtained for a specific run.
+
+\textbf{Figure 3: OTUs common to those models used to differentiate
+between patients with normal colons and those with adenoma, advanced
+adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
+each model. B) For each common OTU the lowest taxonomic identification
+and importance rank for each model run is shown.
+
+\textbf{Figure 4: Treatment response based on models built for adenoma,
+advanced adenoma, or carcinoma.} A) Positive probability change from
+initial to follow up sample in those with adenoma. B) Positive
+probability change from initial to follow up sample in those with
+advanced adenoma. C) Positive probability change from initial to follow
+up sample in those with carcinoma.
 

### 2017-08-14 submission/manuscript_R1.tex
move figs and tables to after references

@@
 assess whether the post-treatment samples received the same
-classification as the pre-treament samples. Samples from patients who
+classification as the pre-treatment samples. Samples from patients who
 had carcinomas changed towards a microbial milieu that resembles the
@@
 difficult to identify significant differences between pre and
-post-treatment samples from patientes with adenoma and advanced adenoma,
+post-treatment samples from patients with adenoma and advanced adenoma,
 this was not the case for carcinomas. Not only were there large changes
@@
 alters the composition of the gut microbiota and that, for those with
-carinomas, the gut microbiota shifted more towards that of a normal
+carcinomas, the gut microbiota shifted more towards that of a normal
 colon after treatment. In the individuals with carcinomas, no difference
@@
 {[}Figure 1A{]}. We found that the communities obtained pre and
-post-treament among the patients with carcinomas changed significantly
+post-treatment among the patients with carcinomas changed significantly
 more than those patients with adenoma (P-value \textless{} 0.001). There
@@
 each of the diagnosis groups. To limit the likelihood that the models
-would overfit the data because of the relatively small number of
+would over-fit the data because of the relatively small number of
 subjects in each group, we restricted our models to only incorporate 10
@@
 of a large effect size may be due to differences in the role of bacteria
-in the formation of adnomas and carcinomas or it could be due to
+in the formation of adenomas and carcinomas or it could be due to
 differences in the behaviors and medications within these classes of
@@
 The majority of patients undergoing treatment for adenoma or advanced
-adenoma were not treated surigically {[}Table 1{]} but rather via
+adenoma were not treated surgically {[}Table 1{]} but rather via
 colonoscopy. All patients diagnosed with carcinomas were treated with at
 least surgery or a combination of surgery and chemotherapy or surgery,
-chermotherapy, and radiation. The type of chemotherapy used for patients
+chemotherapy, and radiation. The type of chemotherapy used for patients
 with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda,
@@
 
-Similarily, the Random Forest {[}33{]} algorithm was also used to create
+Similarly, the Random Forest {[}33{]} algorithm was also used to create
 the three models used to classify normal versus diagnosis. These samples
@@
 
-For each of the pre versus post-treamtent models assessment of the most
+For each of the pre versus post-treatment models assessment of the most
 important OTUs was then made by taking the top 10 OTUs by mean decrease
@@
 OTU pre versus post-treatment model by diagnosis group to help avoid
-model overfitting. These reduced models were then put through the same
+model over-fitting. These reduced models were then put through the same
 process mentioned in the previous paragraph and were what was used for
@@
 adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
-(pre-treatment carcinoma (carinoma n = 26 and disease free n = 0) versus
-post-treatment carcinoma (carcinoma n = 1 and disease free n = 25)). The
-application of the pre versus post-treatment models generated the
-probabilites that the sample was a pre-treatment sample. The application
-of the normal versus diagnosis models generated the probabilities that
-the sample was that specific diagnosis (adenoma, advanced adenoma, or
-carcinoma).
+(pre-treatment carcinoma (carcinoma n = 26 and disease free n = 0)
+versus post-treatment carcinoma (carcinoma n = 1 and disease free n =
+25)). The application of the pre versus post-treatment models generated
+the probabilities that the sample was a pre-treatment sample. The
+application of the normal versus diagnosis models generated the
+probabilities that the sample was that specific diagnosis (adenoma,
+advanced adenoma, or carcinoma).
 
@@
 respective SRA study accession.
-
-\newpage
-
-\textbf{Table 1: Demographic data of patients in the pre and
-post-treatment cohort}
-
-\begin{longtable}[]{@{}lccc@{}}
-\toprule
-& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
-\midrule
-\endhead
-n & 22 & 19 & 26\tabularnewline
-Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
-12.9\tabularnewline
-Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
-BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
-Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
-Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
-350.85 ± 102\tabularnewline
-Surgery Only & 4 & 4 & 12\tabularnewline
-Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
-Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
-\bottomrule
-\end{longtable}
-
-\newpage
-
-\textbf{Table 2: Demographic data of training cohort}
-
-\begin{longtable}[]{@{}lcccc@{}}
-\toprule
-& Normal & Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
-\midrule
-\endhead
-n & 172 & 67 & 90 & 94\tabularnewline
-Age (Mean ± SD) & 54.29 ± 9.9 & 63.01 ± 13.1 & 64.07 ± 11.3 & 64.37 ±
-12.9\tabularnewline
-Sex (\%F) & 64.53 & 46.27 & 37.78 & 43.62\tabularnewline
-BMI (Mean ± SD) & 26.97 ± 5.3 & 25.69 ± 4.8 & 26.66 ± 4.9 & 29.27 ±
-6.7\tabularnewline
-Caucasian (\%) & 87.79 & 92.54 & 92.22 & 94.68\tabularnewline
-\bottomrule
-\end{longtable}
-
-\newpage
-
-\textbf{Figure 1: General differences between adenoma, advanced adenoma,
-and carcinoma groups after treatment.} A) \(\theta\)\textsubscript{YC}
-distances from pre versus post sample within each individual. A
-significant difference was found between the adenoma and carcinoma group
-(P-value = 5.36e-05). Solid black points represent the median value for
-each diagnosis group. B) NMDS of the pre and post-treatment samples for
-the adenoma group. C) NMDS of the pre and post-treatment samples for the
-advanced adenoma group. D) NMDS of the pre and post-treatment samples
-for the carcinoma group.
-
-\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
-advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
-OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
-MDA value obtained from 100 different 80/20 splits while the lighter
-colored circles represents the value obtained for a specific run.
-
-\textbf{Figure 3: OTUs common to those models used to differentiate
-between patients with normal colons and those with adenoma, advanced
-adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
-each model. B) For each common OTU the lowest taxonomic identification
-and importance rank for each model run is shown.
-
-\textbf{Figure 4: Treatment response based on models built for adenoma,
-advanced adenoma, or carcinoma.} A) Positive probability change from
-initial to follow up sample in those with adenoma. B) Positive
-probability change from initial to follow up sample in those with
-advanced adenoma. C) Positive probability change from initial to follow
-up sample in those with carcinoma.
-
-\newpage
-
-\textbf{Figure S1: ROC curves of the adenoma, advanced adenoma, and
-carcinoma models.} A) Adenoma ROC curve: The light green shaded areas
-represent the range of values of a 100 different 80/20 splits of the
-test set data and the dark green line represents the model using 100\%
-of the data set and what was used for subsequent classification. B)
-Advanced Adenoma ROC curve: The light yellow shaded areas represent the
-range of values of a 100 different 80/20 splits of the test set data and
-the dark yellow line represents the model using 100\% of the data set
-and what was used for subsequent classification. C) Carcinoma ROC curve:
-The light red shaded areas represent the range of values of a 100
-different 80/20 splits of the test set data and the dark red line
-represents the model using 100\% of the data set and what was used for
-subsequent classification.
-
-\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
-adenoma, and carcinoma models.} A) MDA of the most important variables
-in the adenoma model. The dark green point represents the mean and the
-lighter green points are the value of each of the 100 different runs. B)
-Summary of Important Variables in the advanced adenoma model. MDA of the
-most important variables in the SRN model. The dark yellow point
-represents the mean and the lighter yellow points are the value of each
-of the 100 different runs. C) MDA of the most important variables in the
-carcinoma model. The dark red point represents the mean and the lighter
-red points are the value of each of the 100 different runs.
-
-\textbf{Figure S3: Pre and post-treatment relative abundance of CRC
-associated OTUs within the carcinoma model.}
 
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\subsection{References}\label{references}
 
@@
 
+\newpage
+
+\textbf{Table 1: Demographic data of patients in the pre and
+post-treatment cohort}
+
+\begin{longtable}[]{@{}lccc@{}}
+\toprule
+& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
+\midrule
+\endhead
+n & 22 & 19 & 26\tabularnewline
+Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
+12.9\tabularnewline
+Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
+BMI (Mean ± SD) & 26.86 ± 3.9 & 25.81 ± 4.7 & 28.63 ± 7.2\tabularnewline
+Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
+Days Between Colonoscopy (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 &
+350.85 ± 102\tabularnewline
+Surgery Only & 4 & 4 & 12\tabularnewline
+Surgery \& Chemotherapy & 0 & 0 & 9\tabularnewline
+Surgery, Chemotherapy, \& Radiation & 0 & 0 & 5\tabularnewline
+\bottomrule
+\end{longtable}
+
+\newpage
+
+\textbf{Table 2: Demographic data of training cohort}
+
+\begin{longtable}[]{@{}lcccc@{}}
+\toprule
+& Normal & Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
+\midrule
+\endhead
+n & 172 & 67 & 90 & 94\tabularnewline
+Age (Mean ± SD) & 54.29 ± 9.9 & 63.01 ± 13.1 & 64.07 ± 11.3 & 64.37 ±
+12.9\tabularnewline
+Sex (\%F) & 64.53 & 46.27 & 37.78 & 43.62\tabularnewline
+BMI (Mean ± SD) & 26.97 ± 5.3 & 25.69 ± 4.8 & 26.66 ± 4.9 & 29.27 ±
+6.7\tabularnewline
+Caucasian (\%) & 87.79 & 92.54 & 92.22 & 94.68\tabularnewline
+\bottomrule
+\end{longtable}
+
+\newpage
+
+\textbf{Figure 1: General differences between adenoma, advanced adenoma,
+and carcinoma groups after treatment.} A) \(\theta\)\textsubscript{YC}
+distances from pre versus post sample within each individual. A
+significant difference was found between the adenoma and carcinoma group
+(P-value = 5.36e-05). Solid black points represent the median value for
+each diagnosis group. B) NMDS of the pre and post-treatment samples for
+the adenoma group. C) NMDS of the pre and post-treatment samples for the
+advanced adenoma group. D) NMDS of the pre and post-treatment samples
+for the carcinoma group.
+
+\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
+advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
+OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
+MDA value obtained from 100 different 80/20 splits while the lighter
+colored circles represents the value obtained for a specific run.
+
+\textbf{Figure 3: OTUs common to those models used to differentiate
+between patients with normal colons and those with adenoma, advanced
+adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
+each model. B) For each common OTU the lowest taxonomic identification
+and importance rank for each model run is shown.
+
+\textbf{Figure 4: Treatment response based on models built for adenoma,
+advanced adenoma, or carcinoma.} A) Positive probability change from
+initial to follow up sample in those with adenoma. B) Positive
+probability change from initial to follow up sample in those with
+advanced adenoma. C) Positive probability change from initial to follow
+up sample in those with carcinoma.
+
+\newpage
+
+\textbf{Figure S1: ROC curves of the adenoma, advanced adenoma, and
+carcinoma models.} A) Adenoma ROC curve: The light green shaded areas
+represent the range of values of a 100 different 80/20 splits of the
+test set data and the dark green line represents the model using 100\%
+of the data set and what was used for subsequent classification. B)
+Advanced Adenoma ROC curve: The light yellow shaded areas represent the
+range of values of a 100 different 80/20 splits of the test set data and
+the dark yellow line represents the model using 100\% of the data set
+and what was used for subsequent classification. C) Carcinoma ROC curve:
+The light red shaded areas represent the range of values of a 100
+different 80/20 splits of the test set data and the dark red line
+represents the model using 100\% of the data set and what was used for
+subsequent classification.
+
+\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
+adenoma, and carcinoma models.} A) MDA of the most important variables
+in the adenoma model. The dark green point represents the mean and the
+lighter green points are the value of each of the 100 different runs. B)
+Summary of Important Variables in the advanced adenoma model. MDA of the
+most important variables in the SRN model. The dark yellow point
+represents the mean and the lighter yellow points are the value of each
+of the 100 different runs. C) MDA of the most important variables in the
+carcinoma model. The dark red point represents the mean and the lighter
+red points are the value of each of the 100 different runs.
+
+\textbf{Figure S3: Pre and post-treatment relative abundance of CRC
+associated OTUs within the carcinoma model.}
+
+\newpage
+
 \end{document}

### 2017-09-12 submission/manuscript_R2.Rmd
first commit of second revision

--- 2017-08-12 submission/manuscript_R1.Rmd
@@
 To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
-We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treament samples.
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
 Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 **Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
-Although it was difficult to identify significant differences between pre and post-treatment samples from patientes with adenoma and advanced adenoma, this was not the case for carcinomas.
+Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas.
 Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
@@
 Finally, we measured the predictive probabilities to assess whether treatment yielded bacterial communities similar to those individuals with normal colons.
-We found that treatment alters the composition of the gut microbiota and that, for those with carinomas, the gut microbiota shifted more towards that of a normal colon after treatment.
+We found that treatment alters the composition of the gut microbiota and that, for those with carcinomas, the gut microbiota shifted more towards that of a normal colon after treatment.
 In the individuals with carcinomas, no difference was found by the type of treatment (surgery alone, surgery with chemotherapy, surgery with chemotherapy and radiation).
@@
 The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
-We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001).
+We found that the communities obtained pre and post-treatment among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001).
 There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
@@
 We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-To limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
+To limit the likelihood that the models would over-fit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
 Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
@@
 We were surprised by the lack of a consistent signal that was associated with treatment of patients with adenomas or advanced adenomas.
-The lack of a large effect size may be due to differences in the role of bacteria in the formation of adnomas and carcinomas or it could be due to differences in the behaviors and medications within these classes of patients.
+The lack of a large effect size may be due to differences in the role of bacteria in the formation of adenomas and carcinomas or it could be due to differences in the behaviors and medications within these classes of patients.
 One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
@@
 For this study treatment refers specifically to the removal of a lesion with or without chemotherapy and radiation.
-The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1] but rather via colonoscopy.
-All patients diagnosed with carcinomas were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chermotherapy, and radiation.
+The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surgically [Table 1] but rather via colonoscopy.
+All patients diagnosed with carcinomas were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chemotherapy, and radiation.
 The type of chemotherapy used for patients with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
@@
 
-Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis.
+Similarly, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis.
 These samples were obtained using the same methodology as described earlier in this section.
@@
 
-For each of the pre versus post-treamtent models assessment of the most important OTUs was then made by taking the top 10 OTUs by mean decrease in accuracy (MDA).
-These were then used to build each respective reduced OTU pre versus post-treatment model by diagnosis group to help avoid model overfitting.
+For each of the pre versus post-treatment models assessment of the most important OTUs was then made by taking the top 10 OTUs by mean decrease in accuracy (MDA).
+These were then used to build each respective reduced OTU pre versus post-treatment model by diagnosis group to help avoid model over-fitting.
 These reduced models were then put through the same process mentioned in the previous paragraph and were what was used for the final pre versus post-treatment models.
@@
 
-Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)).
-The application of the pre versus post-treatment models generated the probabilites that the sample was a pre-treatment sample.
+Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)).
+The application of the pre versus post-treatment models generated the probabilities that the sample was a pre-treatment sample.
 The application of the normal versus diagnosis models generated the probabilities that the sample was that specific diagnosis (adenoma, advanced adenoma, or carcinoma).
@@
 Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
+
+\newpage
+
+## Declarations
+
+### Ethics approval and consent to participate
+
+The University of Michigan Institutional Review Board approved this study, and all subjects provided informed consent. This study conformed to the guidelines of the Helsinki Declaration.
+
+### Consent for publication
+
+Not applicable.
+
+### Availability of data and material
+
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
+Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
+
+### Competing Interests
+
+All authors declare that they do not have any relevant competing interests to report.
+
+### Funding
+
+This study was supported by funding from the National Institutes of Health to P. Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).
+
+### Authors' contributions
+
+All authors were involved in the conception and design of the study.
+MAS analyzed the data.
+NTB processed samples and analyzed the data.
+All authors interpreted the data.
+MAS and PDS wrote the manuscript.
+All authors reviewed and revised the manuscript.
+All authors read and approved the final manuscript.
+
+### Acknowledgements
+
+The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+We would also like to thank Amanda Elmore for reviewing and correcting code error and providing feedback on manuscript drafts.
+We would also like to thank Nicholas Lesniak for providing feedback on manuscript drafts.
+
+\newpage
+
+## References
+
+<div id="refs"></div>
 
@@
 \newpage
-
-## Declarations
-
-### Ethics approval and consent to participate
-
-The University of Michigan Institutional Review Board approved this study, and all subjects provided informed consent. This study conformed to the guidelines of the Helsinki Declaration.
-
-### Consent for publication
-
-Not applicable.
-
-### Availability of data and material
-
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
-Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
-
-### Competing Interests
-
-All authors declare that they do not have any relevant competing interests to report.
-
-### Funding
-
-This study was supported by funding from the National Institutes of Health to P. Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).
-
-### Authors' contributions
-
-All authors were involved in the conception and design of the study.
-MAS analyzed the data.
-NTB processed samples and analyzed the data.
-All authors interpreted the data.
-MAS and PDS wrote the manuscript.
-All authors reviewed and revised the manuscript.
-All authors read and approved the final manuscript.
-
-### Acknowledgements
-
-The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-We would also like to thank Amanda Elmore for reviewing and correcting code error and providing feedback on manuscript drafts.
-We would also like to thank Nicholas Lesniak for providing feedback on manuscript drafts.
-
-\newpage
-
-## References

### 2017-09-12 code/treatment_otu_graph.R
update figure to include top 10 from full model

@@ -6,68 +6,75 @@ source('code/functions.R')
 
 loadLibs(c("tidyr", "dplyr", "scales", "ggplot2", "gridExtra"))
 
-# Load needed data
-adn_treat_varsMDA <- read.csv('data/process/tables/reduced_adn_treatment_top_vars_MDA_Summary.csv', 
-                        header = T, stringsAsFactors = F) %>% 
-  mutate(rank = seq(1:10), tax_id = gsub("_unclassified", "", tax_id))
+# Load needed data and take only top 10
+adn_treat_varsMDA <- read.csv('data/process/tables/adn_treatment_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F) %>% slice(1:10)
 
-srn_treat_varsMDA <- read.csv('data/process/tables/reduced_srn_treatment_top_vars_MDA_Summary.csv', 
-                        header = T, stringsAsFactors = F) %>% 
-  mutate(rank = seq(1:10), tax_id = gsub("_unclassified", "", tax_id))
+srn_treat_varsMDA <- read.csv('data/process/tables/srn_treatment_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F) %>% slice(1:10)
 
-crc_treat_varsMDA <- read.csv('data/process/tables/reduced_crc_treatment_top_vars_MDA_Summary.csv', 
-                        header = T, stringsAsFactors = F) %>% 
-  mutate(rank = seq(1:10), tax_id = gsub("_unclassified", "", tax_id))
+crc_treat_varsMDA <- read.csv('data/process/tables/crc_treatment_MDA_Summary.csv', 
+                        header = T, stringsAsFactors = F) %>% slice(1:10)
 
 
-adn_data <- read.csv('data/process/tables/reduced_adn_treatment_top_vars_MDA_full_data.csv', 
-                     header = T, stringsAsFactors = F)
+adn_data <- read.csv('data/process/tables/adn_treatment_raw_mda_values.csv', 
+                     header = T, stringsAsFactors = F) %>% 
+  filter(Variable %in% adn_treat_varsMDA$Variable)
 
-srn_data <- read.csv('data/process/tables/reduced_srn_treatment_top_vars_MDA_full_data.csv', 
-                     header = T, stringsAsFactors = F)
[…]

### 2017-09-12 code/treatment_otu_graph.R
update to include appropriate files

@@ -102,7 +102,7 @@ srn_treat_MDA_graph <- srn_data %>% filter(Overall > 0) %>%
                     levels = rev(unique(srn_data$Variable)), 
                     labels = rev(srn_axis)), log10(Overall))) + 
   geom_point(color = '#F0E68C') + stat_summary(fun.y = "median", colour = '#EEC900', geom = "point", size = 2.5) + 
-  coord_flip(ylim = c(-0.3, 0.8)) + theme_bw() + ylab("Log10 MDA") + xlab("") +  ggtitle("B") + 
+  coord_flip(ylim = c(-3, 0.8)) + theme_bw() + ylab("Log10 MDA") + xlab("") +  ggtitle("B") + 
   scale_x_discrete(labels = rev(srn_labels)) + 
   theme(plot.title = element_text(face = "bold", hjust = -0.60, size = 20), 
         legend.position = "none", 
@@ -110,11 +110,12 @@ srn_treat_MDA_graph <- srn_data %>% filter(Overall > 0) %>%
         axis.text.y = element_text(size = 10))
 
 
-crc_treat_MDA_graph <- ggplot(crc_data, aes(factor(otu, 
-                                                 levels = rev(unique(crc_data$otu)), 
-                                                 labels = rev(crc_axis)), log10(value))) + 
+crc_treat_MDA_graph <- crc_data %>% filter(Overall > 0) %>% 
+  ggplot(aes(factor(Variable, 
+                              levels = rev(unique(crc_data$Variable)), 
+                              labels = rev(crc_axis)), log10(Overall))) + 
   geom_point(color = '#FFB6C1') + stat_summary(fun.y = "median", colour = '#DC143C', geom = "point", size = 2.5) + 
-  coord_flip(ylim = c(-0.3, 0.8)) + theme_bw() + ylab("Log10 MDA") + xlab("") +  ggtitle("C") + 
+  coord_flip(ylim = c(-3, 0.8)) + theme_bw() + ylab("Log10 MDA") + xlab("") +  ggtitle("C") + 
   scale_x_discrete(labels = rev(crc_labels)) + 
   theme(plot.title = element_text(face = "bold", hjust = -0.60, size = 20), 
         legend.position = "none", 

### 2017-09-12 submission/manuscript_R2.Rmd
fix hperlink

@@
 ***Reproducible Methods.***
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_FollowUps_Microbiome_2017.
 Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

### 2017-09-13 submission/manuscript_R2.Rmd
update table loading

@@
 # Extra data and meta data needed
-good_metaf <- read.csv('../data/process/mod_metadata/metaf_final.csv', header = T, stringsAsFactors = F)
+good_metaf <- read.csv('../data/process/mod_metadata/metaF_final.csv', header = T, stringsAsFactors = F)
 metaI <- read.csv('../data/process/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
@@
 
-# Antibiotic test data
-surg_only_comparisons <- read.csv("../data/process/tables/surgery_only_comparisons.csv", header = T)
-surgVnonsurg_noncrc <- read.csv("../data/process/tables/surgVnonsurg_noncrc.csv", header = T)
-surgVnonsurg_noncrc_pvalues <- read.csv("../data/process/tables/surgVnonsurg_noncrc_pvalues.csv", header = T)
-
 # Treatment model data
-adn_treat_AUC_summary <- read.csv('../data/process/tables/reduced_adn_treatment_ROC_model_summary.csv',
+adn_treat_AUC_summary <- read.csv('../data/process/tables/adn_treatment_ROC_model_summary.csv',
                                   stringsAsFactors = F, header = T)
-srn_treat_AUC_summary <- read.csv('../data/process/tables/reduced_srn_treatment_ROC_model_summary.csv',
+srn_treat_AUC_summary <- read.csv('../data/process/tables/srn_treatment_ROC_model_summary.csv',
                                   stringsAsFactors = F, header = T)
-crc_treat_AUC_summary <- read.csv('../data/process/tables/reduced_crc_treatment_ROC_model_summary.csv',
+crc_treat_AUC_summary <- read.csv('../data/process/tables/crc_treatment_ROC_model_summary.csv',
                                   stringsAsFactors = F, header = T)
 
-adn_treat_imp_vars <- read.csv('../data/process/tables/reduced_adn_treatment_top_vars_MDA_Summary.csv',
+adn_treat_imp_vars <- read.csv('../data/process/tables/adn_treatment_MDA_Summary.csv',
                                   stringsAsFactors = F, header = T)
-srn_treat_imp_vars <- read.csv('../data/process/tables/reduced_srn_treatment_top_vars_MDA_Summary.csv',
+srn_treat_imp_vars <- read.csv('../data/process/tables/srn_treatment_MDA_Summary.csv',
                                   stringsAsFactors = F, header = T)
-crc_treat_imp_vars <- read.csv('../data/process/tables/reduced_crc_treatment_top_vars_MDA_Summary.csv',
+crc_treat_imp_vars <- read.csv('../data/process/tables/crc_treatment_MDA_Summary.csv',
                                   stringsAsFactors = F, header = T)
@@
 # Adenoma Model
-adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv', header = T)
-adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
-adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)
-adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)
+adn_varsMDA <- read.csv('../data/process/tables/adn_MDA_Summary.csv', header = T)
+adn_AUC_run_summary <- read.csv('../data/process/tables/adn_ROC_model_summary.csv', header = T)
+adn_general_summary <- read.csv('../data/process/tables/adn_auc_summary.csv', header = T, row.names = 1)
 
 # SRN Model
-srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T)
-srn_imp_vars <- read.csv('../data/process/tables/srn_rf_otu_tax.csv', header = T, row.names = 1)
-srn_AUC_run_summary <- read.csv('../data/process/tables/srn_Reduced_ROC_model_summary.csv', header = T)
-srn_general_summary <- read.csv('../data/process/tables/srn_reduced_auc_summary.csv', header = T, row.names = 1)
+srn_varsMDA <- read.csv('../data/process/tables/srn_MDA_Summary.csv', header = T)
+srn_AUC_run_summary <- read.csv('../data/process/tables/srn_ROC_model_summary.csv', header = T)
+srn_general_summary <- read.csv('../data/process/tables/srn_auc_summary.csv', header = T, row.names = 1)
 
 # CRC Model
-crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
-crc_imp_vars <- read.csv('../data/process/tables/crc_rf_otu_tax.csv', header = T, row.names = 1)
-crc_AUC_run_summary <- read.csv('../data/process/tables/crc_Reduced_ROC_model_summary.csv', header = T)
-crc_general_summary <- read.csv('../data/process/tables/crc_reduced_auc_summary.csv', header = T, row.names = 1)
-crc_probs <- read.csv('../data/process/tables/crc_reduced_follow_up_probability_summary.csv', header = T)
+crc_varsMDA <- read.csv('../data/process/tables/crc_MDA_Summary.csv', header = T)
+crc_AUC_run_summary <- read.csv('../data/process/tables/crc_ROC_model_summary.csv', header = T)
+crc_general_summary <- read.csv('../data/process/tables/crc_auc_summary.csv', header = T, row.names = 1)
+crc_probs <- read.csv('../data/process/tables/crc_follow_up_probability_summary.csv', header = T)
 
@@
 all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F) %>% rename(comparison = X)
-all_confu_matrix <- read.csv('../data/process/tables/all_crc_srn_adn_models_confusion_summary.csv', header = T, stringsAsFactors = F)
 

### 2017-09-13 submission/manuscript_R2.Rmd
complete modifications to align with back end

@@
 comb_adn_surg_common_vars_summary <- read.csv('../data/process/tables/all_adn_surg_pvalue_summary.csv', header = T, stringsAsFactors = F)
+
+# Permutation testing of treatment model
+treat_random_pvalues <- read.csv('../data/process/tables/treatment_comparison_AUC_summary.csv', header = T, stringsAsFactors = F)
+treat_auc_act <- read.csv('../data/process/tables/actual_treatment_AUC_summary.csv', header = T, stringsAsFactors = F)
+treat_auc_rand <- read.csv('../data/process/tables/randomization_treatment_AUC_summary.csv', header = T, stringsAsFactors = F)
 
@@
 We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-To limit the likelihood that the models would over-fit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs.
-Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
-Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2].
+The models overall performed moderately, with the carcinoma model being the best (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
+Interestingly, the top 10 most important OTUs by MDA that were used for each model had little overlap with each other [Figure 2].
 Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups.
+Both the adenoma and carcioma treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.0001).
 
@@
 We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed well (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1).
-The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
+The models performed moderately with CRC being the best (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1).
+The OTUs that were in the top 10% of importance for the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
 Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g.
 *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing a marked decrease in relative abundance [Figure S3].
+Although many of important OTUs in the top 10% were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing a marked decrease in relative abundance [Figure S3].
 Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
@@
 There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(round(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 2))`).
+The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(round(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3))`).
 Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
@@
 The `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]).
-We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
+We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage I, `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage II, and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[3]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
 We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["75%"], digits = 0)`) days) or decreased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["75%"], digits = 0)`) days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) (all days data displayed as median (IQR)).
@@
 The resulting model was then tested on the hold out data obtained from the 20% component.
-For all pre versus post-treatment models the optimized mtry was `r ifelse(as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")` and for all normal versus diagnosis models the optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
+For adenoma pre versus post-treatment models the optimized mtrys were `r names(table(adn_treat_AUC_summary$best_mtry))[1]` (n = `r table(adn_treat_AUC_summary$best_mtry)[1]`), `r names(table(adn_treat_AUC_summary$best_mtry))[2]` (n = `r table(adn_treat_AUC_summary$best_mtry)[2]`), and `r names(table(adn_treat_AUC_summary$best_mtry))[3]` (n = `r table(adn_treat_AUC_summary$best_mtry)[3]`) for adenoma.
+For advanced adenoma the mtrys were  `r names(table(srn_treat_AUC_summary$best_mtry))[1]` (n = `r table(srn_treat_AUC_summary$best_mtry)[1]`), `r names(table(srn_treat_AUC_summary$best_mtry))[2]` (n = `r table(srn_treat_AUC_summary$best_mtry)[2]`), and `r names(table(srn_treat_AUC_summary$best_mtry))[3]` (n = `r table(srn_treat_AUC_summary$best_mtry)[3]`).
+For carcinoma the mtrys were  `r names(table(crc_treat_AUC_summary$best_mtry))[1]` (n = `r table(crc_treat_AUC_summary$best_mtry)[1]`), `r names(table(crc_treat_AUC_summary$best_mtry))[2]` (n = `r table(crc_treat_AUC_summary$best_mtry)[2]`), and `r names(table(crc_treat_AUC_summary$best_mtry))[3]` (n = `r table(crc_treat_AUC_summary$best_mtry)[3]`).
+For the normal versus diagnosis models the optimized mtrys were `r names(table(adn_AUC_run_summary$best_mtry))[1]` (n = `r table(adn_AUC_run_summary$best_mtry)[1]`), `r names(table(adn_AUC_run_summary$best_mtry))[2]` (n = `r table(adn_AUC_run_summary$best_mtry)[2]`), and `r names(table(adn_AUC_run_summary$best_mtry))[3]` (n = `r table(adn_AUC_run_summary$best_mtry)[3]`) for adenoma.
+For advanced adenoma `r names(table(srn_AUC_run_summary$best_mtry))[1]` (n = `r table(srn_AUC_run_summary$best_mtry)[1]`), `r names(table(srn_AUC_run_summary$best_mtry))[2]` (n = `r table(srn_AUC_run_summary$best_mtry)[2]`), and `r names(table(srn_AUC_run_summary$best_mtry))[3]` (n = `r table(srn_AUC_run_summary$best_mtry)[3]`).
+For carcinoma `r names(table(crc_AUC_run_summary$best_mtry))[1]` (n = `r table(crc_AUC_run_summary$best_mtry)[1]`), `r names(table(crc_AUC_run_summary$best_mtry))[2]` (n = `r table(crc_AUC_run_summary$best_mtry)[2]`), and `r names(table(crc_AUC_run_summary$best_mtry))[3]` (n = `r table(crc_AUC_run_summary$best_mtry)[3]`).
 The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].
 
-For each of the pre versus post-treatment models assessment of the most important OTUs was then made by taking the top 10 OTUs by mean decrease in accuracy (MDA).
-These were then used to build each respective reduced OTU pre versus post-treatment model by diagnosis group to help avoid model over-fitting.
-These reduced models were then put through the same process mentioned in the previous paragraph and were what was used for the final pre versus post-treatment models.
-For the normal versus diagnosis models the important OTUs were obtained by counting the number of times an OTU was present in the top 10% of MDA for each of the 100 different splits.
-This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs.
-These corresponding reduced OTU normal versus diagnosis models were then put through the same process mentioned in the previous paragraph and were what was used for the final normal versus diagnosis models.
-For the pre versus post-treatment models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")` and for the normal versus diagnosis models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.
+The treatment models were then assessed for their strength to classify whether individuals had obtained treatment.
+We also assessed whether the models performed better than random by creating 100 permutations of a data set in which the labels were randomly assigned.
+To assess similarity of the treatment models we looked for overlap between their top 10 most important OTUs.
 
-Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)).
+For the diagnosis models that were trained on a completely independent sample set, each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)).
 The application of the pre versus post-treatment models generated the probabilities that the sample was a pre-treatment sample.
 The application of the normal versus diagnosis models generated the probabilities that the sample was that specific diagnosis (adenoma, advanced adenoma, or carcinoma).
+We also looked at similarities between the models by looking for overlap in the top 10% of important OTUs to each respective model.
 
@@
 
-**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
+**Figure 2: The top 10 most important OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
 B) Advanced Adenoma OTUs.
@@
 
-**Figure 3: OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
+**Figure 3: Top 10% most important OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
 B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
@@
 
-**Figure S2: Summary of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
+**Figure S2: Summary of top 10% of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
 The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.

### 2017-09-13 submission/manuscript_R2.tex
complete modifications to align with back end

--- 2017-08-14 submission/manuscript_R1.tex
@@
 allow us to differentiate between pre and post-treatment samples from
-each of the diagnosis groups. To limit the likelihood that the models
-would over-fit the data because of the relatively small number of
-subjects in each group, we restricted our models to only incorporate 10
-OTUs. Despite this restriction, the models performed well (adenoma AUC
-range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
-AUC range = 0.82 - 0.98). Interestingly, the 10 OTUs that were used for
-each model had little overlap with each other {[}Figure 2{]}. Although
-treatment had an impact on the overall community structure, the effect
-of treatment was not consistent across patients and diagnosis groups.
+each of the diagnosis groups. The models overall performed moderately,
+with the carcinoma model being the best (adenoma AUC range = 0.54 -
+0.83, advanced adenoma AUC range = 0.34 - 0.65, carcinoma AUC range =
+0.82 - 0.98). Interestingly, the top 10 most important OTUs by MDA that
+were used for each model had little overlap with each other {[}Figure
+2{]}. Although treatment had an impact on the overall community
+structure, the effect of treatment was not consistent across patients
+and diagnosis groups. Both the adenoma and carcioma treatment models had
+AUCs that were significantly higher than a random model permutation
+(P-value \textless{} 0.0001).
 
@@
 (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
-models performed well (adenoma AUC range = 0.62 - 0.72, advanced adenoma
-AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
-The OTUs that were incorporated into the adenoma and advanced adenoma
-models largely overlapped and those OTUs that were used to classify the
-carcinoma samples were largely distinct from those of the other two
-models {[}Figure 3A{]}. Among the OTUs that were shared across the three
-models were those populations generally considered beneficial to their
-host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
+models performed moderately with CRC being the best (adenoma AUC range =
+0.50 - 0.62, advanced adenoma AUC range = 0.53 - 0.67, carcinoma AUC
+range = 0.71 - 0.82; Figure S1). The OTUs that were in the top 10\% of
+importance for the adenoma and advanced adenoma models largely
+overlapped and those OTUs that were used to classify the carcinoma
+samples were largely distinct from those of the other two models
+{[}Figure 3A{]}. Among the OTUs that were shared across the three models
+were those populations generally considered beneficial to their host
+(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
 \emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
-\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
-also included in the model differentiating between patients with normal
-colons and those with carcinoma, this model also included OTUs
-affiliated with populations that have previously been associated with
-carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
+\emph{Roseburia}) {[}Figures 3B{]}. Although many of important OTUs in
+the top 10\% were also included in the model differentiating between
+patients with normal colons and those with carcinoma, this model also
+included OTUs affiliated with populations that have previously been
+associated with carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
 \emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
@@
 carcinoma significantly decreased with treatment, suggesting a shift
-toward a normal microbiota for most individuals (P-value = 0.02). Only,
-6 of the 26 patients (23.08\%) who were diagnosed with a carcinoma had a
+toward a normal microbiota for most individuals (P-value = 0.001). Only,
+7 of the 26 patients (26.92\%) who were diagnosed with a carcinoma had a
 higher positive probability after treatment; one of those was
@@
 not a significant difference in the effect of the treatments on the
-amount of change in the community structure (P-value = 0.532). Finally,
+amount of change in the community structure (P-value = 0.375). Finally,
 the change in the positive probability was not significantly different
-between any of the treatment groups (P-value = 0.532). Due to the
+between any of the treatment groups (P-value = 0.375). Due to the
 relatively small number of samples in each treatment group, it was
@@
 
-There was a subset of patients (6 of the 26 with carcinomas) who
+There was a subset of patients (7 of the 26 with carcinomas) who
 demonstrated an elevated probability of carcinoma after treatment. This
-may reflect an elevated risk of recurrence. The 23.08\% prevalence of
+may reflect an elevated risk of recurrence. The 26.92\% prevalence of
 increased carcinoma probability from our study is within the expected
@@
 individuals may have had more severe tumors; however, the tumor severity
-of these 6 individuals (3 with Stage II and 3 with Stage III) was
-similar to the distribution observed among the other 20 patients. We
-also hypothesized that we may have sampled these patients later than the
-rest and their communities may have reverted to a carcinoma-associated
-state; however, there was not a statistically significant difference in
-the length of time between sample collection among those whose
-probabilities increased (358 (336 - 458) days) or decreased (334 (256 -
-399) days) (Wilcoxon Test; P-value = 0.56) (all days data displayed as
-median (IQR)). Finally, it is possible that these patients may not have
-responded to treatment as well as the other 20 patients diagnosed with
-carcinoma and so the microbiota may not have been impacted the same way.
-Again, further studies looking at the role of the microbiota in
-recurrence are needed to understand the dynamics following treatment.
+of these 7 individuals (1 with Stage I, 3 with Stage II, and 3 with
+Stage III) was similar to the distribution observed among the other 19
+patients. We also hypothesized that we may have sampled these patients
+later than the rest and their communities may have reverted to a
+carcinoma-associated state; however, there was not a statistically
+significant difference in the length of time between sample collection
+among those whose probabilities increased (331 (246 - 358) days) or
+decreased (364 (301 - 434) days) (Wilcoxon Test; P-value = 0.39) (all
+days data displayed as median (IQR)). Finally, it is possible that these
+patients may not have responded to treatment as well as the other 19
+patients diagnosed with carcinoma and so the microbiota may not have
+been impacted the same way. Again, further studies looking at the role
+of the microbiota in recurrence are needed to understand the dynamics
+following treatment.
 
@@
 The resulting model was then tested on the hold out data obtained from
-the 20\% component. For all pre versus post-treatment models the
-optimized mtry was 2 and for all normal versus diagnosis models the
-optimized mtry was 2. The hyper-parameter, mtry, defines the number of
+the 20\% component. For adenoma pre versus post-treatment models the
+optimized mtrys were 2 (n = 42), 42 (n = 26), and 914 (n = 32) for
+adenoma. For advanced adenoma the mtrys were 2 (n = 26), 52 (n = 14),
+and 1353 (n = 60). For carcinoma the mtrys were 2 (n = 36), 41 (n = 54),
+and 863 (n = 10). For the normal versus diagnosis models the optimized
+mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
+adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
+NA), and (n = NA). The hyper-parameter, mtry, defines the number of
 variables to investigate at each split before a new division of the data
@@
 
-For each of the pre versus post-treatment models assessment of the most
-important OTUs was then made by taking the top 10 OTUs by mean decrease
-in accuracy (MDA). These were then used to build each respective reduced
-OTU pre versus post-treatment model by diagnosis group to help avoid
-model over-fitting. These reduced models were then put through the same
-process mentioned in the previous paragraph and were what was used for
-the final pre versus post-treatment models. For the normal versus
-diagnosis models the important OTUs were obtained by counting the number
-of times an OTU was present in the top 10\% of MDA for each of the 100
-different splits. This was then followed with filtering of this list to
-variables that were only present in more than 50\% of these 100 runs.
-These corresponding reduced OTU normal versus diagnosis models were then
-put through the same process mentioned in the previous paragraph and
-were what was used for the final normal versus diagnosis models. For the
-pre versus post-treatment models the final optimized mtry was 2 and for
-the normal versus diagnosis models the final optimized mtry was 2.
+The treatment models were then assessed for their strength to classify
+whether individuals had obtained treatment. We also assessed whether the
+models performed better than random by creating 100 permutations of a
+data set in which the labels were randomly assigned. To assess
+similarity of the treatment models we looked for overlap between their
+top 10 most important OTUs.
 
-Each model was then applied to our 67-person cohort {[}Table 1{]} based
-on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22 and disease
-free n = 0) versus post-treatment adenoma (adenoma n = 0 and disease
-free n = 22)), advanced adenoma pre-treatment advanced adenoma (advanced
-adenoma n = 19 and disease free n = 0 ) versus post-treatment advanced
-adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
-(pre-treatment carcinoma (carcinoma n = 26 and disease free n = 0)
-versus post-treatment carcinoma (carcinoma n = 1 and disease free n =
-25)). The application of the pre versus post-treatment models generated
-the probabilities that the sample was a pre-treatment sample. The
-application of the normal versus diagnosis models generated the
+For the diagnosis models that were trained on a completely independent
+sample set, each model was then applied to our 67-person cohort {[}Table
+1{]} based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22
+and disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
+disease free n = 22)), advanced adenoma pre-treatment advanced adenoma
+(advanced adenoma n = 19 and disease free n = 0 ) versus post-treatment
+advanced adenoma (advanced adenoma n = 0 and disease free n = 19), and
+carcinoma (pre-treatment carcinoma (carcinoma n = 26 and disease free n
+= 0) versus post-treatment carcinoma (carcinoma n = 1 and disease free n
+= 25)). The application of the pre versus post-treatment models
+generated the probabilities that the sample was a pre-treatment sample.
+The application of the normal versus diagnosis models generated the
 probabilities that the sample was that specific diagnosis (adenoma,
-advanced adenoma, or carcinoma).
+advanced adenoma, or carcinoma). We also looked at similarities between
+the models by looking for overlap in the top 10\% of important OTUs to
+each respective model.
 
@@
 description of how the data were processed and analyzed can be found at
-\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
-have been deposited into the NCBI Sequence Read Archive (SRP062005 and
-SRP096978) and the necessary metadata can be found at
+\url{https://github.com/SchlossLab/Sze_FollowUps_Microbiome_2017}. Raw
+sequences have been deposited into the NCBI Sequence Read Archive
+(SRP062005 and SRP096978) and the necessary metadata can be found at
 \url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
@@
 
-\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
-advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
-OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
-MDA value obtained from 100 different 80/20 splits while the lighter
-colored circles represents the value obtained for a specific run.
+\textbf{Figure 2: The top 10 most important OTUs used to classify
+treatment for adenoma, advanced adenoma, and carcinoma.} A) Adenoma
+OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs. The darker circle
+highlights the median log10 MDA value obtained from 100 different 80/20
+splits while the lighter colored circles represents the value obtained
+for a specific run.
 
-\textbf{Figure 3: OTUs common to those models used to differentiate
-between patients with normal colons and those with adenoma, advanced
-adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
-each model. B) For each common OTU the lowest taxonomic identification
-and importance rank for each model run is shown.
+\textbf{Figure 3: Top 10\% most important OTUs common to those models
+used to differentiate between patients with normal colons and those with
+adenoma, advanced adenoma, and carcinoma.} A) Venn diagram showing the
+OTU overlap between each model. B) For each common OTU the lowest
+taxonomic identification and importance rank for each model run is
+shown.
 
@@
 
-\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
-adenoma, and carcinoma models.} A) MDA of the most important variables
-in the adenoma model. The dark green point represents the mean and the
-lighter green points are the value of each of the 100 different runs. B)
-Summary of Important Variables in the advanced adenoma model. MDA of the
-most important variables in the SRN model. The dark yellow point
-represents the mean and the lighter yellow points are the value of each
-of the 100 different runs. C) MDA of the most important variables in the
-carcinoma model. The dark red point represents the mean and the lighter
-red points are the value of each of the 100 different runs.
+\textbf{Figure S2: Summary of top 10\% of important OTUs for the
+adenoma, advanced adenoma, and carcinoma models.} A) MDA of the most
+important variables in the adenoma model. The dark green point
+represents the mean and the lighter green points are the value of each
+of the 100 different runs. B) Summary of Important Variables in the
+advanced adenoma model. MDA of the most important variables in the SRN
+model. The dark yellow point represents the mean and the lighter yellow
+points are the value of each of the 100 different runs. C) MDA of the
+most important variables in the carcinoma model. The dark red point
+represents the mean and the lighter red points are the value of each of
+the 100 different runs.
 

### 2017-09-13 submission/manuscript_R2.Rmd
bring abstrct into word count for msphere

@@
 
-**Background.** Colorectal cancer is a worldwide health problem.
-Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
-This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
-We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).
+**Background.** Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
+This study tests whether treatment alters the abundance of bacterial populations associated with disease versus those with a normal colon by sequencing the 16S rRNA genes in feces of `r length(rownames(good_metaf))` individuals before and after treatment (adenoma n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma n = `r change_summary["thetayc", "srn_n"]`), and carcinoma n = `r change_summary["thetayc", "crc_n"]`).
 
-**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
-Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment.
-To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
-We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
-Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+**Results.** The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
+Although Random Forest models classified pre and post-treatment samples for each diagnosis group, there
+was little overlap between the OTUs that were indicative of each treatment.
+Training Random Forest models on a larger cohort that contained individuals with normal colons, adenomas, advanced adenomas, and carcinomas we assessed the positive probability of having one of three diagnoses and whether the pre and post-treatment samples were classified the same.
+Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+There was no significant difference in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
-Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas.
-Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
-
-### Keywords
-
-microbiota; colorectal cancer; polyps; treatment; risk factor.
+**Conclusions.** It was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma.
+However, for carcinomas, not only were there large changes in pre versus post-treatment samples, but these changes were towards a more normal microbiota.
 

### 2017-09-13 submission/manuscript_R2.tex
bring abstrct into word count for msphere

@@
 
-\textbf{Background.} Colorectal cancer is a worldwide health problem.
-Despite growing evidence that members of the gut microbiota can drive
-tumorigenesis, little is known about what happens to it after treatment
-for an adenoma or carcinoma. This study tested the hypothesis that
-treatment for adenoma or carcinoma alters the abundance of bacterial
-populations associated with disease to those associated with a normal
-colon. We tested this hypothesis by sequencing the 16S rRNA genes in the
-feces of 67 individuals before and after treatment for adenoma (N = 22),
-advanced adenoma (N = 19), and carcinoma (N = 26).
+\textbf{Background.} Despite growing evidence that members of the gut
+microbiota can drive tumorigenesis, little is known about what happens
+to it after treatment for an adenoma or carcinoma. This study tests
+whether treatment alters the abundance of bacterial populations
+associated with disease versus those with a normal colon by sequencing
+the 16S rRNA genes in feces of 67 individuals before and after treatment
+(adenoma n = 22), advanced adenoma n = 19), and carcinoma n = 26).
 
-\textbf{Results.} There were small changes to the bacterial community
-associated with adenoma or advanced adenoma and large changes associated
-with carcinoma. The communities from patients with carcinomas changed
+\textbf{Results.} The communities from patients with carcinomas changed
 significantly more than those with adenoma following treatment (P-value
-\textless{} 0.001). Although treatment was associated with intrapersonal
-changes, the change in the abundance of individual OTUs in response to
-treatment was not consistent within diagnosis groups (P-value
-\textgreater{} 0.05). Because the distribution of OTUs across patients
-and diagnosis groups was irregular, we used the Random Forest machine
-learning algorithm to identify groups of OTUs that could be used to
-classify pre and post-treatment samples for each of the diagnosis
-groups. Although the three models could differentiate between the pre
-and post-treatment samples, there was little overlap between the OTUs
-that were indicative of each treatment. To determine whether individuals
-who underwent treatment were more likely to have OTUs associated with
-normal colons we used a larger cohort that contained individuals with
-normal colons and those with adenomas, advanced adenomas, and
-carcinomas. We again built Random Forest models and measured the change
-in the positive probability of having one of the three diagnoses to
-assess whether the post-treatment samples received the same
-classification as the pre-treatment samples. Samples from patients who
-had carcinomas changed towards a microbial milieu that resembles the
-normal colon after treatment (P-value \textless{} 0.05). Finally, we
-were unable to detect any significant differences in the microbiota of
+\textless{} 0.001). Although Random Forest models classified pre and
+post-treatment samples for each diagnosis group, there was little
+overlap between the OTUs that were indicative of each treatment.
+Training Random Forest models on a larger cohort that contained
+individuals with normal colons, adenomas, advanced adenomas, and
+carcinomas we assessed the positive probability of having one of three
+diagnoses and whether the pre and post-treatment samples were classified
+the same. Patients who had carcinomas changed towards a microbial milieu
+that resembles the normal colon after treatment (P-value \textless{}
+0.05). There was no significant difference in the microbiota of
 individuals treated with surgery alone and those treated with
@@
 
-\textbf{Conclusions.} By better understanding the response of the
-microbiota to treatment for adenomas and carcinomas, it is likely that
-biomarkers will eventually be validated that can be used to quantify the
-risk of recurrence and the likelihood of survival. Although it was
-difficult to identify significant differences between pre and
-post-treatment samples from patients with adenoma and advanced adenoma,
-this was not the case for carcinomas. Not only were there large changes
-in pre versus post-treatment samples for those with carcinoma, but these
+\textbf{Conclusions.} It was difficult to identify significant
+differences between pre and post-treatment samples from patients with
+adenoma and advanced adenoma. However, for carcinomas, not only were
+there large changes in pre versus post-treatment samples, but these
 changes were towards a more normal microbiota.
-
-\subsubsection{Keywords}\label{keywords}
-
-microbiota; colorectal cancer; polyps; treatment; risk factor.
 

### 2017-09-13 submission/supplemental.Rmd
update abstract word count and supplemental to msphere

--- 2017-04-18 submission/supplemental.Rmd
@@
 ---
-title: ""
-bibliography:
+title: ''
 csl: microbiome.csl
+fontsize: 12pt
 output:
   pdf_document:
-    keep_tex: true
     includes:
       in_header: header.tex
-fontsize: 12pt
+    keep_tex: yes
 geometry: margin=1.0in
+bibliography: null
 ---
 
-# Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment
+# Normalization of the microbiota in patients after treatment for colonic lesions
 
@@
 
-adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+adn_varsMDA <- read.csv('../data/process/tables/adn_MDA_Summary.csv', header = T, stringsAsFactors = F)
 
-srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+srn_varsMDA <- read.csv('../data/process/tables/srn_MDA_Summary.csv', header = T, stringsAsFactors = F)
 
-crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
+crc_varsMDA <- read.csv('../data/process/tables/crc_MDA_Summary.csv', header = T, stringsAsFactors = F)
 
 # Create selection vector
-common_vars <- common_vars_summary$otu
-common_num <- as.numeric(gsub("Otu", "", common_vars_summary$otu))
+common_vars <- common_vars_summary$Variable
+common_num <- as.numeric(gsub("Otu", "", common_vars_summary$Variable))
 
 # Create MDA vectors
-adn_MDA <- filter(adn_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)) %>% select(median_MDA)
-srn_MDA <- filter(srn_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)) %>% select(median_MDA)
-crc_MDA <- filter(crc_varsMDA, otu %in% common_vars) %>% slice(match(common_vars, otu)) %>% select(median_MDA)
+adn_MDA <- filter(adn_varsMDA, Variable %in% common_vars) %>% slice(match(common_vars, Variable)) %>% select(median_mda)
+srn_MDA <- filter(srn_varsMDA, Variable %in% common_vars) %>% slice(match(common_vars, Variable)) %>% select(median_mda)
+crc_MDA <- filter(crc_varsMDA, Variable %in% common_vars) %>% slice(match(common_vars, Variable)) %>% select(median_mda)
 
-st2 <- select(common_vars_summary, otu) %>%
-  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>%
+st2 <- select(common_vars_summary, Variable) %>%
+  mutate(Tax = gsub("_unclassified", "", common_vars_summary$tax_ID)) %>%
   mutate(otu = paste("OTU", common_num, sep = "")) %>%
-  mutate(adn_MDA = round(adn_MDA$median_MDA, digits = 2), srn_MDA = round(srn_MDA$median_MDA, digits = 2), crc_MDA = round(crc_MDA$median_MDA, digits = 2)) %>%
+  mutate(adn_MDA = round(adn_MDA$median_mda, digits = 2), srn_MDA = round(srn_MDA$median_mda, digits = 2), crc_MDA = round(crc_MDA$median_mda, digits = 2)) %>%
   arrange(desc(adn_MDA), desc(srn_MDA), desc(crc_MDA))

### 2017-09-13 submission/manuscript_R2.Rmd
add to make everything inline with msystems.csl

@@
 title: ''
-csl: microbiome.csl
+csl: msystems.csl
 fontsize: 12pt

### 2017-09-13 submission/manuscript_R2.tex
add to make everything inline with msystems.csl

@@
 Colorectal cancer (CRC) is the third most common cause of cancer deaths
-in the United States {[}1,2{]}. Disease mortality has significantly
-decreased, predominately due to improvements in screening {[}2{]}.
-Despite these improvements, there are still approximately 50,000
-CRC-related deaths per year in the United States {[}1{]}. Current
-estimates indicate that 20-30\% of those who undergo treatment will
-experience recurrence and 35\% of all patients will die within five
-years {[}3--5{]}. Identification of methods to assess patients' risk of
-recurrence is of great importance to reduce mortality and healthcare
-costs.
+in the United States (1, 2). Disease mortality has significantly
+decreased, predominately due to improvements in screening (2). Despite
+these improvements, there are still approximately 50,000 CRC-related
+deaths per year in the United States (1). Current estimates indicate
+that 20-30\% of those who undergo treatment will experience recurrence
+and 35\% of all patients will die within five years (3--5).
+Identification of methods to assess patients' risk of recurrence is of
+great importance to reduce mortality and healthcare costs.
 
@@
 \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
-\emph{Fusobacterium nucleatum} that alter disease progression
-{[}6--10{]}. Furthermore, studies that shift the structure of the
-microbiota through the use of antibiotics or inoculation of germ free
-mice with human feces have shown that varying community compositions can
-result in varied tumor burden {[}11--13{]}. Collectively, these studies
-support the hypothesis that the microbiota can alter the amount of
-inflammation in the colon and with it the rate of tumorigenesis
-{[}14{]}.
+\emph{Fusobacterium nucleatum} that alter disease progression (6--10).
+Furthermore, studies that shift the structure of the microbiota through
+the use of antibiotics or inoculation of germ free mice with human feces
+have shown that varying community compositions can result in varied
+tumor burden (11--13). Collectively, these studies support the
+hypothesis that the microbiota can alter the amount of inflammation in
+the colon and with it the rate of tumorigenesis (14).
 
 Building upon this evidence, several human studies have identified
-unique signatures of colonic lesions {[}15--20{]}. One line of research
-has identified community-level differences between those bacteria that
-are found on and adjacent to colonic lesions and have supported a role
-for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
-\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
-have proposed feces-based biomarkers that could be used to diagnose the
-presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
-have associated \emph{Fusobacterium nucleatum} and other oral pathogens
-with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
-have also noted that the loss of bacteria generally thought to produce
-short chain fatty acids, which can suppress inflammation, is associated
-with colonic lesions. This suggests that gut bacteria have a role in
+unique signatures of colonic lesions (15--20). One line of research has
+identified community-level differences between those bacteria that are
+found on and adjacent to colonic lesions and have supported a role for
+\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
+\emph{Fusobacterium nucleatum} in tumorigenesis (21--23). Others have
+proposed feces-based biomarkers that could be used to diagnose the
+presence of colonic adenomas and carcinomas (24--26). These studies have
+associated \emph{Fusobacterium nucleatum} and other oral pathogens with
+colonic lesions (adenoma, advanced adenoma, and carcinoma). They have
+also noted that the loss of bacteria generally thought to produce short
+chain fatty acids, which can suppress inflammation, is associated with
+colonic lesions. This suggests that gut bacteria have a role in
 tumorigenesis with potential as useful biomarkers for aiding in the
-early detection of disease {[}21--26{]}.
+early detection of disease (21--26).
 
@@
 associated with carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
+\emph{Parvimonas}) (24--26) {[}Figure S2{]} with some individuals
 showing a marked decrease in relative abundance {[}Figure S3{]}.
@@
 increased carcinoma probability from our study is within the expected
-rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
+rate of recurrence (20-30\% (3, 4)). We hypothesized that these
 individuals may have had more severe tumors; however, the tumor severity
@@
 \textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
-have been previously reported in Baxter, et al {[}24{]}. Briefly,
-samples were stored on ice for at least 24h before freezing. Although we
-cannot exclude that this sampling protocol may have impacted the gut
-microbiota composition all samples were subjected to the same
-methodology. Study exclusion involved those who had already undergone
-surgery, radiation, or chemotherapy, had colorectal cancer before a
-baseline fecal sample could be obtained, had IBD, a known hereditary
-non-polyposis colorectal cancer, or familial adenomatous polyposis.
-Samples used to build the models for prediction were collected either
-prior to a colonoscopy or between one and two weeks after initial
-colonoscopy. The bacterial community has been shown to normalize back to
-a pre-colonoscopy community within this time period {[}27{]}. Our study
-cohort consisted of 67 individuals with an initial sample as described
-and a follow up sample obtained between 188 - 546 days after treatment
-of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
-examination and histopathological review of any biopsies taken. Patients
-were classified as having advanced adenoma if they had an adenoma
-greater than 1 cm, more than three adenomas of any size, or an adenoma
-with villous histology. This study was approved by the University of
-Michigan Institutional Review Board. All study participants provided
-informed consent and the study itself conformed to the guidelines set
-out by the Helsinki Declaration.
+have been previously reported in Baxter, et al (24). Briefly, samples
+were stored on ice for at least 24h before freezing. Although we cannot
+exclude that this sampling protocol may have impacted the gut microbiota
+composition all samples were subjected to the same methodology. Study
+exclusion involved those who had already undergone surgery, radiation,
+or chemotherapy, had colorectal cancer before a baseline fecal sample
+could be obtained, had IBD, a known hereditary non-polyposis colorectal
+cancer, or familial adenomatous polyposis. Samples used to build the
+models for prediction were collected either prior to a colonoscopy or
+between one and two weeks after initial colonoscopy. The bacterial
+community has been shown to normalize back to a pre-colonoscopy
+community within this time period (27). Our study cohort consisted of 67
+individuals with an initial sample as described and a follow up sample
+obtained between 188 - 546 days after treatment of lesion {[}Table 1{]}.
+Patients were diagnosed by colonoscopic examination and
+histopathological review of any biopsies taken. Patients were classified
+as having advanced adenoma if they had an adenoma greater than 1 cm,
+more than three adenomas of any size, or an adenoma with villous
+histology. This study was approved by the University of Michigan
+Institutional Review Board. All study participants provided informed
+consent and the study itself conformed to the guidelines set out by the
+Helsinki Declaration.
 
@@
 \textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
-described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
-Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
-automated pipetting system (Eppendorf). The V4 variable region was
-amplified and the resulting product was split between four sequencing
-runs with normal, adenoma, and carcinoma evenly represented on each run.
-Each group was randomly assigned to avoid biases based on sample
-collection location. The pre and post-treatment samples were sequenced
-on the same run.
+described by Kozich, et al. (28). DNA extraction used the 96-well Soil
+DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated
+pipetting system (Eppendorf). The V4 variable region was amplified and
+the resulting product was split between four sequencing runs with
+normal, adenoma, and carcinoma evenly represented on each run. Each
+group was randomly assigned to avoid biases based on sample collection
+location. The pre and post-treatment samples were sequenced on the same
+run.
 
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described {[}28{]}. The general workflow using mothur
-included merging paired-end reads into contigs, filtering for low
-quality contigs, aligning to the SILVA database {[}29{]}, screening for
-chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
-classifier using the Ribosomal Database Project (RDP){[}31{]}, and
-clustered into Operational Taxonomic Units (OTUs) using a 97\%
-similarity cutoff with an average neighbor clustering algorithm
-{[}32{]}. The number of sequences for each sample was rarefied to 10523
-to minimize the impacts of uneven sampling.
+previously described (28). The general workflow using mothur included
+merging paired-end reads into contigs, filtering for low quality
+contigs, aligning to the SILVA database (29), screening for chimeras
+using UCHIME (30), classifying with a naive Bayesian classifier using
+the Ribosomal Database Project (RDP)(31), and clustered into Operational
+Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
+neighbor clustering algorithm (32). The number of sequences for each
+sample was rarefied to 10523 to minimize the impacts of uneven sampling.
 
-\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
+\textbf{\emph{Model Building.}} The Random Forest (33) algorithm was
 used to create the three models used to classify pre and post-treatment
@@
 
-Similarly, the Random Forest {[}33{]} algorithm was also used to create
-the three models used to classify normal versus diagnosis. These samples
+Similarly, the Random Forest (33) algorithm was also used to create the
+three models used to classify normal versus diagnosis. These samples
 were obtained using the same methodology as described earlier in this
@@
 variables to investigate at each split before a new division of the data
-was created with the Random Forest model {[}33{]}.
+was created with the Random Forest model (33).
 
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA {[}34{]} in the vegan package
+community structure utilized PERMANOVA (34) in the vegan package
 (v2.4.3). Comparisons between probabilities as well as overall
@@
 Wallis test. Where multiple comparison testing was appropriate, a
-Benjamini-Hochberg (BH) correction was applied {[}35{]} and a corrected
+Benjamini-Hochberg (BH) correction was applied (35) and a corrected
 P-value of less than 0.05 was considered significant. The P-values
@@
 \hypertarget{ref-siegel_cancer_2016}{}
-1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
-journal for clinicians. 2016;66:7--30.
+1. Siegel RL, Miller KD, Jemal A. 2016. Cancer statistics, 2016. CA: a
+cancer journal for clinicians 66:7--30.
 
 \hypertarget{ref-haggar_colorectal_2009}{}
-2. Haggar FA, Boushey RP. Colorectal cancer epidemiology: Incidence,
-mortality, survival, and risk factors. Clinics in Colon and Rectal
-Surgery. 2009;22:191--7.
+2. Haggar FA, Boushey RP. 2009. Colorectal cancer epidemiology:
+Incidence, mortality, survival, and risk factors. Clinics in Colon and
+Rectal Surgery 22:191--197.
 
 \hypertarget{ref-hellinger_reoperation_2006}{}
-3. Hellinger MD, Santiago CA. Reoperation for recurrent colorectal
-cancer. Clinics in Colon and Rectal Surgery. 2006;19:228--36.
+3. Hellinger MD, Santiago CA. 2006. Reoperation for recurrent colorectal
+cancer. Clinics in Colon and Rectal Surgery 19:228--236.
 
 \hypertarget{ref-ryuk_predictive_2014}{}
-4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, et al.
-Predictive factors and the prognosis of recurrence of colorectal cancer
-within 2 years after curative resection. Annals of Surgical Treatment
-and Research. 2014;86:143--51.
+4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, Jun SH, Kwon
+YC. 2014. Predictive factors and the prognosis of recurrence of
+colorectal cancer within 2 years after curative resection. Annals of
+Surgical Treatment and Research 86:143--151.
 
 \hypertarget{ref-national_cancer_institute_seer_nodate}{}
-5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer
-{[}Internet{]}. {[}cited 2017 Apr 27{]}. Available from:
-\url{http://seer.cancer.gov/statfacts/html/colorect.html}
+5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer.
 
@@
 6. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
-TR, et al. Polyamine catabolism contributes to enterotoxigenic
-Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
-National Academy of Sciences of the United States of America.
-2011;108:15354--9.
+TR, Hacker-Prietz A, Rabizadeh S, Woster PM, Sears CL, Casero RA. 2011.
+Polyamine catabolism contributes to enterotoxigenic Bacteroides
+fragilis-induced colon tumorigenesis. Proceedings of the National
+Academy of Sciences of the United States of America 108:15354--15359.
 
 \hypertarget{ref-abed_fap2_2016}{}
-7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
-Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
-Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
-Microbe. 2016;20:215--25.
+7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, Sol A,
+Naor R, Pikarsky E, Atlan KA, Mellul A, Chaushu S, Manson AL, Earl AM,
+Ou N, Brennan CA, Garrett WS, Bachrach G. 2016. Fap2 Mediates
+Fusobacterium nucleatum Colorectal Adenocarcinoma Enrichment by Binding
+to Tumor-Expressed Gal-GalNAc. Cell Host \& Microbe 20:215--225.
 
@@
 8. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
-McCafferty J, et al. Microbial genomic analysis reveals the essential
-role of inflammation in bacteria-induced colorectal cancer. Nature
-Communications. 2014;5:4724.
+McCafferty J, Fodor AA, Jobin C. 2014. Microbial genomic analysis
+reveals the essential role of inflammation in bacteria-induced
+colorectal cancer. Nature Communications 5:4724.
 
@@
 9. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
-et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
-modulates the tumor-immune microenvironment. Cell Host \& Microbe.
-2013;14:207--15.
+Clancy TE, Chung DC, Lochhead P, Hold GL, El-Omar EM, Brenner D, Fuchs
+CS, Meyerson M, Garrett WS. 2013. Fusobacterium nucleatum potentiates
+intestinal tumorigenesis and modulates the tumor-immune
+microenvironment. Cell Host \& Microbe 14:207--215.
 
 \hypertarget{ref-wu_human_2009}{}
-10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
-human colonic commensal promotes colon tumorigenesis via activation of T
-helper type 17 T cell responses. Nature Medicine. 2009;15:1016--22.
+10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, Huso DL,
+Brancati FL, Wick E, McAllister F, Housseau F, Pardoll DM, Sears CL.
+2009. A human colonic commensal promotes colon tumorigenesis via
+activation of T helper type 17 T cell responses. Nature Medicine
+15:1016--1022.
 
 \hypertarget{ref-zackular_manipulation_2016}{}
-11. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
-Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
+11. Zackular JP, Baxter NT, Chen GY, Schloss PD. 2016. Manipulation of
+the Gut Microbiota Reveals Role in Colon Tumorigenesis. mSphere 1.
 
@@
 12. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
-GY, et al. The gut microbiome modulates colon tumorigenesis. mBio.
-2013;4:e00692--00613.
+GY, Schloss PD. 2013. The gut microbiome modulates colon tumorigenesis.
+mBio 4:e00692--00613.
 
 \hypertarget{ref-baxter_structure_2014}{}
-13. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
-microbiome following colonization with human feces determines colonic
-tumor burden. Microbiome. 2014;2:20.
+13. Baxter NT, Zackular JP, Chen GY, Schloss PD. 2014. Structure of the
+gut microbiome following colonization with human feces determines
+colonic tumor burden. Microbiome 2:20.
 
 \hypertarget{ref-flynn_metabolic_2016}{}
-14. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
-Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
+14. Flynn KJ, Baxter NT, Schloss PD. 2016. Metabolic and Community
+Synergy of Oral Bacteria in Colorectal Cancer. mSphere 1.
 
 \hypertarget{ref-wang_structural_2012}{}
-15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
-segregation of gut microbiota between colorectal cancer patients and
-healthy volunteers. The ISME journal. 2012;6:320--9.
+15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, Jia W, Cai S, Zhao L.
+2012. Structural segregation of gut microbiota between colorectal cancer
+patients and healthy volunteers. The ISME journal 6:320--329.
 
 \hypertarget{ref-chen_decreased_2013}{}
-16. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, et al.
+16. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, Yang L, Liu
+Z-J, Yuan Y-Z, Liu F, Wu J-X, Zhong L, Fang D-C, Zou W, Fang J-Y. 2013.
 Decreased dietary fiber intake and structural alteration of gut
 microbiota in patients with advanced colorectal adenoma. The American
-Journal of Clinical Nutrition. 2013;97:1044--52.
+Journal of Clinical Nutrition 97:1044--1052.
 
 \hypertarget{ref-chen_human_2012}{}
-17. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
-mucosa-associated microbiota in patients with colorectal cancer. PloS
-One. 2012;7:e39743.
+17. Chen W, Liu F, Ling Z, Tong X, Xiang C. 2012. Human intestinal lumen
+and mucosa-associated microbiota in patients with colorectal cancer.
+PloS One 7:e39743.
 
 \hypertarget{ref-shen_molecular_2010}{}
-18. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, et al.
-Molecular characterization of mucosal adherent bacteria and associations
-with colorectal adenomas. Gut Microbes. 2010;1:138--47.
+18. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, Jovov
+B, Abdo Z, Sandler RS, Keku TO. 2010. Molecular characterization of
+mucosal adherent bacteria and associations with colorectal adenomas. Gut
+Microbes 1:138--147.
 
 \hypertarget{ref-kostic_genomic_2012}{}
-19. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
-al. Genomic analysis identifies association of Fusobacterium with
-colorectal carcinoma. Genome Research. 2012;22:292--8.
+19. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM,
+Ojesina AI, Jung J, Bass AJ, Tabernero J, Baselga J, Liu C, Shivdasani
+RA, Ogino S, Birren BW, Huttenhower C, Garrett WS, Meyerson M. 2012.
+Genomic analysis identifies association of Fusobacterium with colorectal
+carcinoma. Genome Research 22:292--298.
 
 \hypertarget{ref-feng_gut_2015}{}
-20. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, et al. Gut
-microbiome development along the colorectal adenoma-carcinoma sequence.
-Nature Communications. 2015;6:6528.
+20. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, Zhang D, Xia H,
+Xu X, Jie Z, Su L, Li X, Li X, Li J, Xiao L, Huber-Schönauer U,
+Niederseer D, Xu X, Al-Aama JY, Yang H, Wang J, Kristiansen K, Arumugam
+M, Tilg H, Datz C, Wang J. 2015. Gut microbiome development along the
+colorectal adenoma-carcinoma sequence. Nature Communications 6:6528.
 
@@
 21. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, et al. Microbiota organization is a distinct feature of
-proximal colorectal cancers. Proceedings of the National Academy of
-Sciences of the United States of America. 2014;111:18321--6.
+Rossetti BJ, Peterson SN, Snesrud EC, Borisy GG, Lazarev M, Stein E,
+Vadivelu J, Roslani AC, Malik AA, Wanyiri JW, Goh KL, Thevambiga I, Fu
+K, Wan F, Llosa N, Housseau F, Romans K, Wu X, McAllister FM, Wu S,
+Vogelstein B, Kinzler KW, Pardoll DM, Sears CL. 2014. Microbiota
+organization is a distinct feature of proximal colorectal cancers.
+Proceedings of the National Academy of Sciences of the United States of
+America 111:18321--18326.
 
 \hypertarget{ref-mima_fusobacterium_2015}{}
-22. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, et
-al. Fusobacterium nucleatum and T Cells in Colorectal Carcinoma. JAMA
-oncology. 2015;1:653--61.
+22. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, Kim
+SA, Masuda A, Nowak JA, Nosho K, Kostic AD, Giannakis M, Watanabe H,
+Bullman S, Milner DA, Harris CC, Giovannucci E, Garraway LA, Freeman GJ,
+Dranoff G, Chan AT, Garrett WS, Huttenhower C, Fuchs CS, Ogino S. 2015.
+Fusobacterium nucleatum and T Cells in Colorectal Carcinoma. JAMA
+oncology 1:653--661.
 
@@
 23. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
-T-J, et al. Intestinal inflammation targets cancer-inducing activity of
-the microbiota. Science (New York, N.Y.). 2012;338:120--3.
+T-J, Campbell BJ, Abujamel T, Dogan B, Rogers AB, Rhodes JM, Stintzi A,
+Simpson KW, Hansen JJ, Keku TO, Fodor AA, Jobin C. 2012. Intestinal
+inflammation targets cancer-inducing activity of the microbiota. Science
+(New York, NY) 338:120--123.
 
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
-improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine. 2016;8:37.
+24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. 2016. Microbiota-based
+model improves the sensitivity of fecal immunochemical test for
+detecting colonic lesions. Genome Medicine 8:37.
 
 \hypertarget{ref-zeller_potential_2014}{}
-25. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
-Potential of fecal microbiota for early-stage detection of colorectal
-cancer. Molecular Systems Biology. 2014;10:766.
+25. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, Amiot
+A, Böhm J, Brunetti F, Habermann N, Hercog R, Koch M, Luciani A, Mende
+DR, Schneider MA, Schrotz-King P, Tournigand C, Tran Van Nhieu J, Yamada
+T, Zimmermann J, Benes V, Kloor M, Ulrich CM, Knebel Doeberitz M von,
+Sobhani I, Bork P. 2014. Potential of fecal microbiota for early-stage
+detection of colorectal cancer. Molecular Systems Biology 10:766.
 
 \hypertarget{ref-zackular_human_2014}{}
-26. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+26. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. 2014. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
-Research (Philadelphia, Pa.). 2014;7:1112--21.
+Research (Philadelphia, Pa) 7:1112--1121.
 
 \hypertarget{ref-obrien_impact_2013}{}
-27. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
-bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
+27. O'Brien CL, Allison GE, Grimpen F, Pavli P. 2013. Impact of
+colonoscopy bowel preparation on intestinal microbiota. PloS One
+8:e62815.
 
 \hypertarget{ref-kozich_development_2013}{}
-28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
 Development of a dual-index sequencing strategy and curation pipeline
 for analyzing amplicon sequence data on the MiSeq Illumina sequencing
-platform. Applied and Environmental Microbiology. 2013;79:5112--20.
+platform. Applied and Environmental Microbiology 79:5112--5120.
 
 \hypertarget{ref-pruesse_silva_2007}{}
-29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
-SILVA: A comprehensive online resource for quality checked and aligned
-ribosomal RNA sequence data compatible with ARB. Nucleic Acids Research.
-2007;35:7188--96.
+29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J,
+Glöckner FO. 2007. SILVA: A comprehensive online resource for quality
+checked and aligned ribosomal RNA sequence data compatible with ARB.
+Nucleic Acids Research 35:7188--7196.
 
 \hypertarget{ref-edgar_uchime_2011}{}
-30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
-sensitivity and speed of chimera detection. Bioinformatics (Oxford,
-England). 2011;27:2194--200.
+30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. 2011. UCHIME
+improves sensitivity and speed of chimera detection. Bioinformatics
+(Oxford, England) 27:2194--2200.
 
 \hypertarget{ref-wang_naive_2007}{}
-31. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
-for rapid assignment of rRNA sequences into the new bacterial taxonomy.
-Applied and Environmental Microbiology. 2007;73:5261--7.
+31. Wang Q, Garrity GM, Tiedje JM, Cole JR. 2007. Naive Bayesian
+classifier for rapid assignment of rRNA sequences into the new bacterial
+taxonomy. Applied and Environmental Microbiology 73:5261--5267.
 
 \hypertarget{ref-schloss_assessing_2011}{}
-32. Schloss PD, Westcott SL. Assessing and improving methods used in
-operational taxonomic unit-based approaches for 16S rRNA gene sequence
-analysis. Applied and Environmental Microbiology. 2011;77:3219--26.
+32. Schloss PD, Westcott SL. 2011. Assessing and improving methods used
+in operational taxonomic unit-based approaches for 16S rRNA gene
+sequence analysis. Applied and Environmental Microbiology 77:3219--3226.
 
 \hypertarget{ref-breiman_random_2001}{}
-33. Breiman L. Random Forests. Machine Learning. 2001;45:5--32.
+33. Breiman L. 2001. Random Forests. Machine Learning 45:5--32.
 
 \hypertarget{ref-anderson_permanova_2013}{}
-34. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
-the face of heterogeneous dispersions: What null hypothesis are you
-testing? Ecological Monographs. 2013;83:557--74.
+34. Anderson MJ, Walsh DCI. 2013. PERMANOVA, ANOSIM, and the Mantel test
+in the face of heterogeneous dispersions: What null hypothesis are you
+testing? Ecological Monographs 83:557--574.
 
 \hypertarget{ref-benjamini_controlling_1995}{}
-35. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
-practical and powerful approach to multiple testing. Journal of the
-Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
+35. Benjamini Y, Hochberg Y. 1995. Controlling the false discovery rate:
+A practical and powerful approach to multiple testing. Journal of the
+Royal Statistical Society Series B (Methodological) 57:289--300.
 

### 2017-09-13 submission/manuscript_R2.Rmd
add importance statement and R2 to makefile

@@
 However, for carcinomas, not only were there large changes in pre versus post-treatment samples, but these changes were towards a more normal microbiota.
+
+\newpage
+
+## Importance
+
+This study helps to clarify what happens to the microbiota after generalized treatment for an adenoma or carcinoma.
+It is a necessary first step to ultimately investigating how the microbiota changes in those with carcinoma recurrence.
+This is important since a large number of those treated for carcinoma will have a recurrence.
+Finding better ways to stratify those most and least likely to recur using the microbiota could provide a cheaper and more accurate alternative to the current status quo.
+Our results suggest that the microbiota could be used to stratify individuals for risk of recurrence and provides a starting point for future studies into this area.
 

### 2017-09-13 submission/manuscript_R2.tex
add importance statement and R2 to makefile

@@
 changes were towards a more normal microbiota.
+
+\newpage
+
+\subsection{Importance}\label{importance}
+
+This study helps to clarify what happens to the microbiota after
+generalized treatment for an adenoma or carcinoma. It is a necessary
+first step to ultimately investigating how the microbiota changes in
+those with carcinoma recurrence. This is important since a large number
+of those treated for carcinoma will have a recurrence. Finding better
+ways to stratify those most and least likely to recur using the
+microbiota could provide a cheaper and more accurate alternative to the
+current status quo. Our results suggest that the microbiota could be
+used to stratify individuals for risk of recurrence and provides a
+starting point for future studies into this area.
 

### 2017-09-13 code/Run_render_revision2_paper.R

@@ -1,0 +1,11 @@
+### List of commands to render Revision 1 pdf
+### renders both main text and supplemental
+
+# Load needed functions
+source('code/functions.R')
+
+# Load needed library
+loadLibs(c("knitr", "rmarkdown"))
+
+# Render the final pdfs
+render('submission/manuscript_R2.Rmd', clean = FALSE)

### 2017-09-13 submission/manuscript_R2.Rmd
fix NAs in text

@@
 For carcinoma the mtrys were  `r names(table(crc_treat_AUC_summary$best_mtry))[1]` (n = `r table(crc_treat_AUC_summary$best_mtry)[1]`), `r names(table(crc_treat_AUC_summary$best_mtry))[2]` (n = `r table(crc_treat_AUC_summary$best_mtry)[2]`), and `r names(table(crc_treat_AUC_summary$best_mtry))[3]` (n = `r table(crc_treat_AUC_summary$best_mtry)[3]`).
-For the normal versus diagnosis models the optimized mtrys were `r names(table(adn_AUC_run_summary$best_mtry))[1]` (n = `r table(adn_AUC_run_summary$best_mtry)[1]`), `r names(table(adn_AUC_run_summary$best_mtry))[2]` (n = `r table(adn_AUC_run_summary$best_mtry)[2]`), and `r names(table(adn_AUC_run_summary$best_mtry))[3]` (n = `r table(adn_AUC_run_summary$best_mtry)[3]`) for adenoma.
-For advanced adenoma `r names(table(srn_AUC_run_summary$best_mtry))[1]` (n = `r table(srn_AUC_run_summary$best_mtry)[1]`), `r names(table(srn_AUC_run_summary$best_mtry))[2]` (n = `r table(srn_AUC_run_summary$best_mtry)[2]`), and `r names(table(srn_AUC_run_summary$best_mtry))[3]` (n = `r table(srn_AUC_run_summary$best_mtry)[3]`).
-For carcinoma `r names(table(crc_AUC_run_summary$best_mtry))[1]` (n = `r table(crc_AUC_run_summary$best_mtry)[1]`), `r names(table(crc_AUC_run_summary$best_mtry))[2]` (n = `r table(crc_AUC_run_summary$best_mtry)[2]`), and `r names(table(crc_AUC_run_summary$best_mtry))[3]` (n = `r table(crc_AUC_run_summary$best_mtry)[3]`).
+For the normal versus diagnosis models the optimized mtrys were `r names(table(adn_AUC_run_summary$mtry))[1]` (n = `r table(adn_AUC_run_summary$mtry)[1]`), `r names(table(adn_AUC_run_summary$mtry))[2]` (n = `r table(adn_AUC_run_summary$mtry)[2]`), and `r names(table(adn_AUC_run_summary$mtry))[3]` (n = `r table(adn_AUC_run_summary$mtry)[3]`) for adenoma.
+For advanced adenoma `r names(table(srn_AUC_run_summary$mtry))[1]` (n = `r table(srn_AUC_run_summary$mtry)[1]`), `r names(table(srn_AUC_run_summary$mtry))[2]` (n = `r table(srn_AUC_run_summary$mtry)[2]`), and `r names(table(srn_AUC_run_summary$mtry))[3]` (n = `r table(srn_AUC_run_summary$mtry)[3]`).
+For carcinoma `r names(table(crc_AUC_run_summary$mtry))[1]` (n = `r table(crc_AUC_run_summary$mtry)[1]`) and `r names(table(crc_AUC_run_summary$mtry))[2]` (n = `r table(crc_AUC_run_summary$mtry)[2]`).
 The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].

### 2017-09-13 submission/manuscript_R2.tex

@@
 and 863 (n = 10). For the normal versus diagnosis models the optimized
-mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
-adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
-NA), and (n = NA). The hyper-parameter, mtry, defines the number of
-variables to investigate at each split before a new division of the data
-was created with the Random Forest model (33).
+mtrys were 2 (n = 16), 37 (n = 44), and 707 (n = 40) for adenoma. For
+advanced adenoma 2 (n = 33), 37 (n = 50), and 696 (n = 17). For
+carcinoma 36 (n = 96) and 649 (n = 4). The hyper-parameter, mtry,
+defines the number of variables to investigate at each split before a
+new division of the data was created with the Random Forest model (33).
 

### 2017-09-15 submission/manuscript_R2.Rmd
Pats edits

--- 2017-09-13 submission/manuscript_R2.Rmd
@@
 
-**Background.** Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
-This study tests whether treatment alters the abundance of bacterial populations associated with disease versus those with a normal colon by sequencing the 16S rRNA genes in feces of `r length(rownames(good_metaf))` individuals before and after treatment (adenoma n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma n = `r change_summary["thetayc", "srn_n"]`), and carcinoma n = `r change_summary["thetayc", "crc_n"]`).
+Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about how treatment impacts the colonic community.
+This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
+We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).
+The structure of the microbiota in all three groups changed following treatment (P`r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"<0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
+To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons (N=XXX) and untreated individuals with adenomas (N=XXX), advanced adenomas (N=XXX), and carcinomas (N=XXX).
+Using these data, we built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
+The microbiota of patients diagnosed with carcinomas changed towards that of a normal colon following treatment (P<0.XXX).
+Those were initially diagnosed as having adenomas or advanced adenomas were no more likely to have a microbiota resembling that of a normal colon.
+These results further illuminate our understanding of the role of the gut microbiota in tumorigenesis and demonstrate the potential for using the gut microbiota as biomarkers for assessing risk and the effectiveness of treatment.
 
-**Results.** The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-Although Random Forest models classified pre and post-treatment samples for each diagnosis group, there
-was little overlap between the OTUs that were indicative of each treatment.
-Training Random Forest models on a larger cohort that contained individuals with normal colons, adenomas, advanced adenomas, and carcinomas we assessed the positive probability of having one of three diagnoses and whether the pre and post-treatment samples were classified the same.
-Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was no significant difference in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+\newpage
 
-**Conclusions.** It was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma.
-However, for carcinomas, not only were there large changes in pre versus post-treatment samples, but these changes were towards a more normal microbiota.
+## Importance
+
+Colorectal cancer is a worldwide health problem.
+Previous work in animal models and human populations has demonstrated that the gut microbiota has an important role in promoting tumorigenesis through a variety of mechanisms.
+Less well understood is how the gut microbiota changes following treatment for varying stages of tumorigenesis.
+We observed that following treatment of patients diagnosed with carcinoma, their gut microbiota is likely revert to a state associated with individuals who had normal colons whereas those diagnosed with adenomas and advanced adenomas were not.
+This study was a necessary step to ultimately investigating how the microbiota changes in those with carcinoma recurrence and predicting which patients are more likely to recur.
+Finding better strategies to stratify patients based on recurrence risk could provide a more effective means of planning therapies and monitoring patients after treatment.
 
@@
 We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-The models overall performed moderately, with the carcinoma model being the best (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
+The adenoma and carcinoma models performed well (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` and carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`); however, the model for patients treated for advanced adenomas was not able to reliably differentiate between the pre and post-treatment samples (advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
 Interestingly, the top 10 most important OTUs by MDA that were used for each model had little overlap with each other [Figure 2].
@@
 
-***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma).
-The total number of individuals in the pre versus post-treatment models was `r length(rownames(good_metaf))` individuals.
-There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the pre versus post-treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the pre versus post-treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the pre versus post-treatment carcinoma model [Table 1].
+***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma) as well as to assess the probability that a sample was more similar to the patient's original diagnosis or that of a disease-free patient.
+All models included only OTU data obtained from 16S rRNA sequencing and were processed using the caret (v`r paste(packageVersion("caret"))`) R package.
+For each model we optimized the mtry hyper-parameter, which defines the number of OTUs to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].
+To insure that our optimization did not result in over-fitting of the data, we made 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split.
+For each of the 100 splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
+The resulting model was then tested on the 20% of the data that were held out.
+A summary of the mtry hyperparameter values that were tried is available in Table SX.
+The reported P-values for each model relative to a random labeling was assessed by comparing the distribution of the 100 80/20 splits for the correctly labeled data to the distribution of randomly labeled data.
 
-Similarly, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis.
-These samples were obtained using the same methodology as described earlier in this section.
-All samples used for this component of model training were from pre-treatment samples only.
-The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2].
-There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the normal versus carcinoma model [Table 2].
-
-All models included only OTU data obtained from 16S rRNA sequencing and were processed and cleaned using the R package caret (v`r paste(packageVersion("caret"))`).
-Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split.
-For each of the different splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
-The resulting model was then tested on the hold out data obtained from the 20% component.
-For adenoma pre versus post-treatment models the optimized mtrys were `r names(table(adn_treat_AUC_summary$best_mtry))[1]` (n = `r table(adn_treat_AUC_summary$best_mtry)[1]`), `r names(table(adn_treat_AUC_summary$best_mtry))[2]` (n = `r table(adn_treat_AUC_summary$best_mtry)[2]`), and `r names(table(adn_treat_AUC_summary$best_mtry))[3]` (n = `r table(adn_treat_AUC_summary$best_mtry)[3]`) for adenoma.
-For advanced adenoma the mtrys were  `r names(table(srn_treat_AUC_summary$best_mtry))[1]` (n = `r table(srn_treat_AUC_summary$best_mtry)[1]`), `r names(table(srn_treat_AUC_summary$best_mtry))[2]` (n = `r table(srn_treat_AUC_summary$best_mtry)[2]`), and `r names(table(srn_treat_AUC_summary$best_mtry))[3]` (n = `r table(srn_treat_AUC_summary$best_mtry)[3]`).
-For carcinoma the mtrys were  `r names(table(crc_treat_AUC_summary$best_mtry))[1]` (n = `r table(crc_treat_AUC_summary$best_mtry)[1]`), `r names(table(crc_treat_AUC_summary$best_mtry))[2]` (n = `r table(crc_treat_AUC_summary$best_mtry)[2]`), and `r names(table(crc_treat_AUC_summary$best_mtry))[3]` (n = `r table(crc_treat_AUC_summary$best_mtry)[3]`).
-For the normal versus diagnosis models the optimized mtrys were `r names(table(adn_AUC_run_summary$best_mtry))[1]` (n = `r table(adn_AUC_run_summary$best_mtry)[1]`), `r names(table(adn_AUC_run_summary$best_mtry))[2]` (n = `r table(adn_AUC_run_summary$best_mtry)[2]`), and `r names(table(adn_AUC_run_summary$best_mtry))[3]` (n = `r table(adn_AUC_run_summary$best_mtry)[3]`) for adenoma.
-For advanced adenoma `r names(table(srn_AUC_run_summary$best_mtry))[1]` (n = `r table(srn_AUC_run_summary$best_mtry)[1]`), `r names(table(srn_AUC_run_summary$best_mtry))[2]` (n = `r table(srn_AUC_run_summary$best_mtry)[2]`), and `r names(table(srn_AUC_run_summary$best_mtry))[3]` (n = `r table(srn_AUC_run_summary$best_mtry)[3]`).
-For carcinoma `r names(table(crc_AUC_run_summary$best_mtry))[1]` (n = `r table(crc_AUC_run_summary$best_mtry)[1]`), `r names(table(crc_AUC_run_summary$best_mtry))[2]` (n = `r table(crc_AUC_run_summary$best_mtry)[2]`), and `r names(table(crc_AUC_run_summary$best_mtry))[3]` (n = `r table(crc_AUC_run_summary$best_mtry)[3]`).
-The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].
-
-The treatment models were then assessed for their strength to classify whether individuals had obtained treatment.
-We also assessed whether the models performed better than random by creating 100 permutations of a data set in which the labels were randomly assigned.
-To assess similarity of the treatment models we looked for overlap between their top 10 most important OTUs.
-
-For the diagnosis models that were trained on a completely independent sample set, each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)).
-The application of the pre versus post-treatment models generated the probabilities that the sample was a pre-treatment sample.
-The application of the normal versus diagnosis models generated the probabilities that the sample was that specific diagnosis (adenoma, advanced adenoma, or carcinoma).
-We also looked at similarities between the models by looking for overlap in the top 10% of important OTUs to each respective model.
+The three diagnosis models were constructed by using the data from Baxter et al.
+[@baxter_microbiota-based_2016], which was censored for the pre-treatment samples of the patients that we had post-treatment samples.
+The treatment models were then used to quantify the model probability that a patient with an initial diagnosis retained that diagnosis or a disease-free diagnosis.
 

### 2017-09-15 submission/manuscript_R2.Rmd
Response to reviewers

@@
 Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota.
-If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence.
+If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would not only remove the lesion, but also the microbiota that promoted the tumorigenesis and hence the risk of recurrence.
 To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner?

### 2017-09-15 submission/manuscript_R2.Rmd
revert back to microbiome format

--- 2017-09-13 submission/manuscript_R2.Rmd
@@
 title: ''
-csl: msystems.csl
+csl: microbiome.csl
 fontsize: 12pt
@@
 
-**Background.** Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
-This study tests whether treatment alters the abundance of bacterial populations associated with disease versus those with a normal colon by sequencing the 16S rRNA genes in feces of `r length(rownames(good_metaf))` individuals before and after treatment (adenoma n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma n = `r change_summary["thetayc", "srn_n"]`), and carcinoma n = `r change_summary["thetayc", "crc_n"]`).
+**Background.** Colorectal cancer is a worldwide health problem.
+Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
+This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
+We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).
 
-**Results.** The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-Although Random Forest models classified pre and post-treatment samples for each diagnosis group, there
-was little overlap between the OTUs that were indicative of each treatment.
-Training Random Forest models on a larger cohort that contained individuals with normal colons, adenomas, advanced adenomas, and carcinomas we assessed the positive probability of having one of three diagnoses and whether the pre and post-treatment samples were classified the same.
-Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
-There was no significant difference in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma.
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
+Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
+Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment.
+To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
+Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-**Conclusions.** It was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma.
-However, for carcinomas, not only were there large changes in pre versus post-treatment samples, but these changes were towards a more normal microbiota.
+**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
+Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas.
+Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
 
-\newpage
+### Keywords
 
-## Importance
-
-This study helps to clarify what happens to the microbiota after generalized treatment for an adenoma or carcinoma.
-It is a necessary first step to ultimately investigating how the microbiota changes in those with carcinoma recurrence.
-This is important since a large number of those treated for carcinoma will have a recurrence.
-Finding better ways to stratify those most and least likely to recur using the microbiota could provide a cheaper and more accurate alternative to the current status quo.
-Our results suggest that the microbiota could be used to stratify individuals for risk of recurrence and provides a starting point for future studies into this area.
+microbiota; colorectal cancer; polyps; treatment; risk factor.
 

### 2017-09-15 submission/manuscript_R2.tex
revert back to microbiome format

--- 2017-09-13 submission/manuscript_R2.tex
@@
 
-\textbf{Background.} Despite growing evidence that members of the gut
-microbiota can drive tumorigenesis, little is known about what happens
-to it after treatment for an adenoma or carcinoma. This study tests
-whether treatment alters the abundance of bacterial populations
-associated with disease versus those with a normal colon by sequencing
-the 16S rRNA genes in feces of 67 individuals before and after treatment
-(adenoma n = 22), advanced adenoma n = 19), and carcinoma n = 26).
+\textbf{Background.} Colorectal cancer is a worldwide health problem.
+Despite growing evidence that members of the gut microbiota can drive
+tumorigenesis, little is known about what happens to it after treatment
+for an adenoma or carcinoma. This study tested the hypothesis that
+treatment for adenoma or carcinoma alters the abundance of bacterial
+populations associated with disease to those associated with a normal
+colon. We tested this hypothesis by sequencing the 16S rRNA genes in the
+feces of 67 individuals before and after treatment for adenoma (N = 22),
+advanced adenoma (N = 19), and carcinoma (N = 26).
 
-\textbf{Results.} The communities from patients with carcinomas changed
+\textbf{Results.} There were small changes to the bacterial community
+associated with adenoma or advanced adenoma and large changes associated
+with carcinoma. The communities from patients with carcinomas changed
 significantly more than those with adenoma following treatment (P-value
-\textless{} 0.001). Although Random Forest models classified pre and
-post-treatment samples for each diagnosis group, there was little
-overlap between the OTUs that were indicative of each treatment.
-Training Random Forest models on a larger cohort that contained
-individuals with normal colons, adenomas, advanced adenomas, and
-carcinomas we assessed the positive probability of having one of three
-diagnoses and whether the pre and post-treatment samples were classified
-the same. Patients who had carcinomas changed towards a microbial milieu
-that resembles the normal colon after treatment (P-value \textless{}
-0.05). There was no significant difference in the microbiota of
+\textless{} 0.001). Although treatment was associated with intrapersonal
+changes, the change in the abundance of individual OTUs in response to
+treatment was not consistent within diagnosis groups (P-value
+\textgreater{} 0.05). Because the distribution of OTUs across patients
+and diagnosis groups was irregular, we used the Random Forest machine
+learning algorithm to identify groups of OTUs that could be used to
+classify pre and post-treatment samples for each of the diagnosis
+groups. Although the three models could differentiate between the pre
+and post-treatment samples, there was little overlap between the OTUs
+that were indicative of each treatment. To determine whether individuals
+who underwent treatment were more likely to have OTUs associated with
+normal colons we used a larger cohort that contained individuals with
+normal colons and those with adenomas, advanced adenomas, and
+carcinomas. We again built Random Forest models and measured the change
+in the positive probability of having one of the three diagnoses to
+assess whether the post-treatment samples received the same
+classification as the pre-treatment samples. Samples from patients who
+had carcinomas changed towards a microbial milieu that resembles the
+normal colon after treatment (P-value \textless{} 0.05). Finally, we
+were unable to detect any significant differences in the microbiota of
 individuals treated with surgery alone and those treated with
@@
 
-\textbf{Conclusions.} It was difficult to identify significant
-differences between pre and post-treatment samples from patients with
-adenoma and advanced adenoma. However, for carcinomas, not only were
-there large changes in pre versus post-treatment samples, but these
+\textbf{Conclusions.} By better understanding the response of the
+microbiota to treatment for adenomas and carcinomas, it is likely that
+biomarkers will eventually be validated that can be used to quantify the
+risk of recurrence and the likelihood of survival. Although it was
+difficult to identify significant differences between pre and
+post-treatment samples from patients with adenoma and advanced adenoma,
+this was not the case for carcinomas. Not only were there large changes
+in pre versus post-treatment samples for those with carcinoma, but these
 changes were towards a more normal microbiota.
+
+\subsubsection{Keywords}\label{keywords}
+
+microbiota; colorectal cancer; polyps; treatment; risk factor.
 
@@
 Colorectal cancer (CRC) is the third most common cause of cancer deaths
-in the United States (1, 2). Disease mortality has significantly
-decreased, predominately due to improvements in screening (2). Despite
-these improvements, there are still approximately 50,000 CRC-related
-deaths per year in the United States (1). Current estimates indicate
-that 20-30\% of those who undergo treatment will experience recurrence
-and 35\% of all patients will die within five years (3--5).
-Identification of methods to assess patients' risk of recurrence is of
-great importance to reduce mortality and healthcare costs.
+in the United States {[}1,2{]}. Disease mortality has significantly
+decreased, predominately due to improvements in screening {[}2{]}.
+Despite these improvements, there are still approximately 50,000
+CRC-related deaths per year in the United States {[}1{]}. Current
+estimates indicate that 20-30\% of those who undergo treatment will
+experience recurrence and 35\% of all patients will die within five
+years {[}3--5{]}. Identification of methods to assess patients' risk of
+recurrence is of great importance to reduce mortality and healthcare
+costs.
 
@@
 \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
-\emph{Fusobacterium nucleatum} that alter disease progression (6--10).
-Furthermore, studies that shift the structure of the microbiota through
-the use of antibiotics or inoculation of germ free mice with human feces
-have shown that varying community compositions can result in varied
-tumor burden (11--13). Collectively, these studies support the
-hypothesis that the microbiota can alter the amount of inflammation in
-the colon and with it the rate of tumorigenesis (14).
+\emph{Fusobacterium nucleatum} that alter disease progression
+{[}6--10{]}. Furthermore, studies that shift the structure of the
+microbiota through the use of antibiotics or inoculation of germ free
+mice with human feces have shown that varying community compositions can
+result in varied tumor burden {[}11--13{]}. Collectively, these studies
+support the hypothesis that the microbiota can alter the amount of
+inflammation in the colon and with it the rate of tumorigenesis
+{[}14{]}.
 
 Building upon this evidence, several human studies have identified
-unique signatures of colonic lesions (15--20). One line of research has
-identified community-level differences between those bacteria that are
-found on and adjacent to colonic lesions and have supported a role for
-\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
-\emph{Fusobacterium nucleatum} in tumorigenesis (21--23). Others have
-proposed feces-based biomarkers that could be used to diagnose the
-presence of colonic adenomas and carcinomas (24--26). These studies have
-associated \emph{Fusobacterium nucleatum} and other oral pathogens with
-colonic lesions (adenoma, advanced adenoma, and carcinoma). They have
-also noted that the loss of bacteria generally thought to produce short
-chain fatty acids, which can suppress inflammation, is associated with
-colonic lesions. This suggests that gut bacteria have a role in
+unique signatures of colonic lesions {[}15--20{]}. One line of research
+has identified community-level differences between those bacteria that
+are found on and adjacent to colonic lesions and have supported a role
+for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
+\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
+have proposed feces-based biomarkers that could be used to diagnose the
+presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
+have associated \emph{Fusobacterium nucleatum} and other oral pathogens
+with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
+have also noted that the loss of bacteria generally thought to produce
+short chain fatty acids, which can suppress inflammation, is associated
+with colonic lesions. This suggests that gut bacteria have a role in
 tumorigenesis with potential as useful biomarkers for aiding in the
-early detection of disease (21--26).
+early detection of disease {[}21--26{]}.
 
@@
 associated with carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
-\emph{Parvimonas}) (24--26) {[}Figure S2{]} with some individuals
+\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
 showing a marked decrease in relative abundance {[}Figure S3{]}.
@@
 increased carcinoma probability from our study is within the expected
-rate of recurrence (20-30\% (3, 4)). We hypothesized that these
+rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
 individuals may have had more severe tumors; however, the tumor severity
@@
 \textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
-have been previously reported in Baxter, et al (24). Briefly, samples
-were stored on ice for at least 24h before freezing. Although we cannot
-exclude that this sampling protocol may have impacted the gut microbiota
-composition all samples were subjected to the same methodology. Study
-exclusion involved those who had already undergone surgery, radiation,
-or chemotherapy, had colorectal cancer before a baseline fecal sample
-could be obtained, had IBD, a known hereditary non-polyposis colorectal
-cancer, or familial adenomatous polyposis. Samples used to build the
-models for prediction were collected either prior to a colonoscopy or
-between one and two weeks after initial colonoscopy. The bacterial
-community has been shown to normalize back to a pre-colonoscopy
-community within this time period (27). Our study cohort consisted of 67
-individuals with an initial sample as described and a follow up sample
-obtained between 188 - 546 days after treatment of lesion {[}Table 1{]}.
-Patients were diagnosed by colonoscopic examination and
-histopathological review of any biopsies taken. Patients were classified
-as having advanced adenoma if they had an adenoma greater than 1 cm,
-more than three adenomas of any size, or an adenoma with villous
-histology. This study was approved by the University of Michigan
-Institutional Review Board. All study participants provided informed
-consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+have been previously reported in Baxter, et al {[}24{]}. Briefly,
+samples were stored on ice for at least 24h before freezing. Although we
+cannot exclude that this sampling protocol may have impacted the gut
+microbiota composition all samples were subjected to the same
+methodology. Study exclusion involved those who had already undergone
+surgery, radiation, or chemotherapy, had colorectal cancer before a
+baseline fecal sample could be obtained, had IBD, a known hereditary
+non-polyposis colorectal cancer, or familial adenomatous polyposis.
+Samples used to build the models for prediction were collected either
+prior to a colonoscopy or between one and two weeks after initial
+colonoscopy. The bacterial community has been shown to normalize back to
+a pre-colonoscopy community within this time period {[}27{]}. Our study
+cohort consisted of 67 individuals with an initial sample as described
+and a follow up sample obtained between 188 - 546 days after treatment
+of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
+examination and histopathological review of any biopsies taken. Patients
+were classified as having advanced adenoma if they had an adenoma
+greater than 1 cm, more than three adenomas of any size, or an adenoma
+with villous histology. This study was approved by the University of
+Michigan Institutional Review Board. All study participants provided
+informed consent and the study itself conformed to the guidelines set
+out by the Helsinki Declaration.
 
@@
 \textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
-described by Kozich, et al. (28). DNA extraction used the 96-well Soil
-DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated
-pipetting system (Eppendorf). The V4 variable region was amplified and
-the resulting product was split between four sequencing runs with
-normal, adenoma, and carcinoma evenly represented on each run. Each
-group was randomly assigned to avoid biases based on sample collection
-location. The pre and post-treatment samples were sequenced on the same
-run.
+described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
+Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
+automated pipetting system (Eppendorf). The V4 variable region was
+amplified and the resulting product was split between four sequencing
+runs with normal, adenoma, and carcinoma evenly represented on each run.
+Each group was randomly assigned to avoid biases based on sample
+collection location. The pre and post-treatment samples were sequenced
+on the same run.
 
@@
 (v1.37.5) was used to process the 16S rRNA gene sequences and has been
-previously described (28). The general workflow using mothur included
-merging paired-end reads into contigs, filtering for low quality
-contigs, aligning to the SILVA database (29), screening for chimeras
-using UCHIME (30), classifying with a naive Bayesian classifier using
-the Ribosomal Database Project (RDP)(31), and clustered into Operational
-Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
-neighbor clustering algorithm (32). The number of sequences for each
-sample was rarefied to 10523 to minimize the impacts of uneven sampling.
+previously described {[}28{]}. The general workflow using mothur
+included merging paired-end reads into contigs, filtering for low
+quality contigs, aligning to the SILVA database {[}29{]}, screening for
+chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
+classifier using the Ribosomal Database Project (RDP){[}31{]}, and
+clustered into Operational Taxonomic Units (OTUs) using a 97\%
+similarity cutoff with an average neighbor clustering algorithm
+{[}32{]}. The number of sequences for each sample was rarefied to 10523
+to minimize the impacts of uneven sampling.
 
-\textbf{\emph{Model Building.}} The Random Forest (33) algorithm was
+\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
 used to create the three models used to classify pre and post-treatment
@@
 
-Similarly, the Random Forest (33) algorithm was also used to create the
-three models used to classify normal versus diagnosis. These samples
+Similarly, the Random Forest {[}33{]} algorithm was also used to create
+the three models used to classify normal versus diagnosis. These samples
 were obtained using the same methodology as described earlier in this
@@
 and 863 (n = 10). For the normal versus diagnosis models the optimized
-mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
-adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
-NA), and (n = NA). The hyper-parameter, mtry, defines the number of
-variables to investigate at each split before a new division of the data
-was created with the Random Forest model (33).
+mtrys were 2 (n = 16), 37 (n = 44), and 707 (n = 40) for adenoma. For
+advanced adenoma 2 (n = 33), 37 (n = 50), and 696 (n = 17). For
+carcinoma 36 (n = 96) and 649 (n = 4). The hyper-parameter, mtry,
+defines the number of variables to investigate at each split before a
+new division of the data was created with the Random Forest model
+{[}33{]}.
 
@@
 was used for all statistical analysis. Comparisons between bacterial
-community structure utilized PERMANOVA (34) in the vegan package
+community structure utilized PERMANOVA {[}34{]} in the vegan package
 (v2.4.3). Comparisons between probabilities as well as overall
@@
 Wallis test. Where multiple comparison testing was appropriate, a
-Benjamini-Hochberg (BH) correction was applied (35) and a corrected
+Benjamini-Hochberg (BH) correction was applied {[}35{]} and a corrected
 P-value of less than 0.05 was considered significant. The P-values
@@
 \hypertarget{ref-siegel_cancer_2016}{}
-1. Siegel RL, Miller KD, Jemal A. 2016. Cancer statistics, 2016. CA: a
-cancer journal for clinicians 66:7--30.
+1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
+journal for clinicians. 2016;66:7--30.
 
 \hypertarget{ref-haggar_colorectal_2009}{}
-2. Haggar FA, Boushey RP. 2009. Colorectal cancer epidemiology:
-Incidence, mortality, survival, and risk factors. Clinics in Colon and
-Rectal Surgery 22:191--197.
+2. Haggar FA, Boushey RP. Colorectal cancer epidemiology: Incidence,
+mortality, survival, and risk factors. Clinics in Colon and Rectal
+Surgery. 2009;22:191--7.
 
 \hypertarget{ref-hellinger_reoperation_2006}{}
-3. Hellinger MD, Santiago CA. 2006. Reoperation for recurrent colorectal
-cancer. Clinics in Colon and Rectal Surgery 19:228--236.
+3. Hellinger MD, Santiago CA. Reoperation for recurrent colorectal
+cancer. Clinics in Colon and Rectal Surgery. 2006;19:228--36.
 
 \hypertarget{ref-ryuk_predictive_2014}{}
-4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, Jun SH, Kwon
-YC. 2014. Predictive factors and the prognosis of recurrence of
-colorectal cancer within 2 years after curative resection. Annals of
-Surgical Treatment and Research 86:143--151.
+4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, et al.
+Predictive factors and the prognosis of recurrence of colorectal cancer
+within 2 years after curative resection. Annals of Surgical Treatment
+and Research. 2014;86:143--51.
 
 \hypertarget{ref-national_cancer_institute_seer_nodate}{}
-5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer.
+5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer
+{[}Internet{]}. {[}cited 2017 Apr 27{]}. Available from:
+\url{http://seer.cancer.gov/statfacts/html/colorect.html}
 
@@
 6. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
-TR, Hacker-Prietz A, Rabizadeh S, Woster PM, Sears CL, Casero RA. 2011.
-Polyamine catabolism contributes to enterotoxigenic Bacteroides
-fragilis-induced colon tumorigenesis. Proceedings of the National
-Academy of Sciences of the United States of America 108:15354--15359.
+TR, et al. Polyamine catabolism contributes to enterotoxigenic
+Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
+National Academy of Sciences of the United States of America.
+2011;108:15354--9.
 
 \hypertarget{ref-abed_fap2_2016}{}
-7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, Sol A,
-Naor R, Pikarsky E, Atlan KA, Mellul A, Chaushu S, Manson AL, Earl AM,
-Ou N, Brennan CA, Garrett WS, Bachrach G. 2016. Fap2 Mediates
-Fusobacterium nucleatum Colorectal Adenocarcinoma Enrichment by Binding
-to Tumor-Expressed Gal-GalNAc. Cell Host \& Microbe 20:215--225.
+7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
+Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
+Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
+Microbe. 2016;20:215--25.
 
@@
 8. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
-McCafferty J, Fodor AA, Jobin C. 2014. Microbial genomic analysis
-reveals the essential role of inflammation in bacteria-induced
-colorectal cancer. Nature Communications 5:4724.
+McCafferty J, et al. Microbial genomic analysis reveals the essential
+role of inflammation in bacteria-induced colorectal cancer. Nature
+Communications. 2014;5:4724.
 
@@
 9. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
-Clancy TE, Chung DC, Lochhead P, Hold GL, El-Omar EM, Brenner D, Fuchs
-CS, Meyerson M, Garrett WS. 2013. Fusobacterium nucleatum potentiates
-intestinal tumorigenesis and modulates the tumor-immune
-microenvironment. Cell Host \& Microbe 14:207--215.
+et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
+modulates the tumor-immune microenvironment. Cell Host \& Microbe.
+2013;14:207--15.
 
 \hypertarget{ref-wu_human_2009}{}
-10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, Huso DL,
-Brancati FL, Wick E, McAllister F, Housseau F, Pardoll DM, Sears CL.
-2009. A human colonic commensal promotes colon tumorigenesis via
-activation of T helper type 17 T cell responses. Nature Medicine
-15:1016--1022.
+10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
+human colonic commensal promotes colon tumorigenesis via activation of T
+helper type 17 T cell responses. Nature Medicine. 2009;15:1016--22.
 
 \hypertarget{ref-zackular_manipulation_2016}{}
-11. Zackular JP, Baxter NT, Chen GY, Schloss PD. 2016. Manipulation of
-the Gut Microbiota Reveals Role in Colon Tumorigenesis. mSphere 1.
+11. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
 
@@
 12. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
-GY, Schloss PD. 2013. The gut microbiome modulates colon tumorigenesis.
-mBio 4:e00692--00613.
+GY, et al. The gut microbiome modulates colon tumorigenesis. mBio.
+2013;4:e00692--00613.
 
 \hypertarget{ref-baxter_structure_2014}{}
-13. Baxter NT, Zackular JP, Chen GY, Schloss PD. 2014. Structure of the
-gut microbiome following colonization with human feces determines
-colonic tumor burden. Microbiome 2:20.
+13. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
+microbiome following colonization with human feces determines colonic
+tumor burden. Microbiome. 2014;2:20.
 
 \hypertarget{ref-flynn_metabolic_2016}{}
-14. Flynn KJ, Baxter NT, Schloss PD. 2016. Metabolic and Community
-Synergy of Oral Bacteria in Colorectal Cancer. mSphere 1.
+14. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
 
 \hypertarget{ref-wang_structural_2012}{}
-15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, Jia W, Cai S, Zhao L.
-2012. Structural segregation of gut microbiota between colorectal cancer
-patients and healthy volunteers. The ISME journal 6:320--329.
+15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+segregation of gut microbiota between colorectal cancer patients and
+healthy volunteers. The ISME journal. 2012;6:320--9.
 
 \hypertarget{ref-chen_decreased_2013}{}
-16. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, Yang L, Liu
-Z-J, Yuan Y-Z, Liu F, Wu J-X, Zhong L, Fang D-C, Zou W, Fang J-Y. 2013.
+16. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, et al.
 Decreased dietary fiber intake and structural alteration of gut
 microbiota in patients with advanced colorectal adenoma. The American
-Journal of Clinical Nutrition 97:1044--1052.
+Journal of Clinical Nutrition. 2013;97:1044--52.
 
 \hypertarget{ref-chen_human_2012}{}
-17. Chen W, Liu F, Ling Z, Tong X, Xiang C. 2012. Human intestinal lumen
-and mucosa-associated microbiota in patients with colorectal cancer.
-PloS One 7:e39743.
+17. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+mucosa-associated microbiota in patients with colorectal cancer. PloS
+One. 2012;7:e39743.
 
 \hypertarget{ref-shen_molecular_2010}{}
-18. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, Jovov
-B, Abdo Z, Sandler RS, Keku TO. 2010. Molecular characterization of
-mucosal adherent bacteria and associations with colorectal adenomas. Gut
-Microbes 1:138--147.
+18. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, et al.
+Molecular characterization of mucosal adherent bacteria and associations
+with colorectal adenomas. Gut Microbes. 2010;1:138--47.
 
 \hypertarget{ref-kostic_genomic_2012}{}
-19. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM,
-Ojesina AI, Jung J, Bass AJ, Tabernero J, Baselga J, Liu C, Shivdasani
-RA, Ogino S, Birren BW, Huttenhower C, Garrett WS, Meyerson M. 2012.
-Genomic analysis identifies association of Fusobacterium with colorectal
-carcinoma. Genome Research 22:292--298.
+19. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+al. Genomic analysis identifies association of Fusobacterium with
+colorectal carcinoma. Genome Research. 2012;22:292--8.
 
 \hypertarget{ref-feng_gut_2015}{}
-20. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, Zhang D, Xia H,
-Xu X, Jie Z, Su L, Li X, Li X, Li J, Xiao L, Huber-Schönauer U,
-Niederseer D, Xu X, Al-Aama JY, Yang H, Wang J, Kristiansen K, Arumugam
-M, Tilg H, Datz C, Wang J. 2015. Gut microbiome development along the
-colorectal adenoma-carcinoma sequence. Nature Communications 6:6528.
+20. Feng Q, Liang S, Jia H, Stadlmayr A, Tang L, Lan Z, et al. Gut
+microbiome development along the colorectal adenoma-carcinoma sequence.
+Nature Communications. 2015;6:6528.
 
@@
 21. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, Peterson SN, Snesrud EC, Borisy GG, Lazarev M, Stein E,
-Vadivelu J, Roslani AC, Malik AA, Wanyiri JW, Goh KL, Thevambiga I, Fu
-K, Wan F, Llosa N, Housseau F, Romans K, Wu X, McAllister FM, Wu S,
-Vogelstein B, Kinzler KW, Pardoll DM, Sears CL. 2014. Microbiota
-organization is a distinct feature of proximal colorectal cancers.
-Proceedings of the National Academy of Sciences of the United States of
-America 111:18321--18326.
+Rossetti BJ, et al. Microbiota organization is a distinct feature of
+proximal colorectal cancers. Proceedings of the National Academy of
+Sciences of the United States of America. 2014;111:18321--6.
 
 \hypertarget{ref-mima_fusobacterium_2015}{}
-22. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, Kim
-SA, Masuda A, Nowak JA, Nosho K, Kostic AD, Giannakis M, Watanabe H,
-Bullman S, Milner DA, Harris CC, Giovannucci E, Garraway LA, Freeman GJ,
-Dranoff G, Chan AT, Garrett WS, Huttenhower C, Fuchs CS, Ogino S. 2015.
-Fusobacterium nucleatum and T Cells in Colorectal Carcinoma. JAMA
-oncology 1:653--661.
+22. Mima K, Sukawa Y, Nishihara R, Qian ZR, Yamauchi M, Inamura K, et
+al. Fusobacterium nucleatum and T Cells in Colorectal Carcinoma. JAMA
+oncology. 2015;1:653--61.
 
@@
 23. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
-T-J, Campbell BJ, Abujamel T, Dogan B, Rogers AB, Rhodes JM, Stintzi A,
-Simpson KW, Hansen JJ, Keku TO, Fodor AA, Jobin C. 2012. Intestinal
-inflammation targets cancer-inducing activity of the microbiota. Science
-(New York, NY) 338:120--123.
+T-J, et al. Intestinal inflammation targets cancer-inducing activity of
+the microbiota. Science (New York, N.Y.). 2012;338:120--3.
 
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. 2016. Microbiota-based
-model improves the sensitivity of fecal immunochemical test for
-detecting colonic lesions. Genome Medicine 8:37.
+24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. 2016;8:37.
 
 \hypertarget{ref-zeller_potential_2014}{}
-25. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, Amiot
-A, Böhm J, Brunetti F, Habermann N, Hercog R, Koch M, Luciani A, Mende
-DR, Schneider MA, Schrotz-King P, Tournigand C, Tran Van Nhieu J, Yamada
-T, Zimmermann J, Benes V, Kloor M, Ulrich CM, Knebel Doeberitz M von,
-Sobhani I, Bork P. 2014. Potential of fecal microbiota for early-stage
-detection of colorectal cancer. Molecular Systems Biology 10:766.
+25. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+Potential of fecal microbiota for early-stage detection of colorectal
+cancer. Molecular Systems Biology. 2014;10:766.
 
 \hypertarget{ref-zackular_human_2014}{}
-26. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. 2014. The human gut
+26. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
-Research (Philadelphia, Pa) 7:1112--1121.
+Research (Philadelphia, Pa.). 2014;7:1112--21.
 
 \hypertarget{ref-obrien_impact_2013}{}
-27. O'Brien CL, Allison GE, Grimpen F, Pavli P. 2013. Impact of
-colonoscopy bowel preparation on intestinal microbiota. PloS One
-8:e62815.
+27. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
+bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
 
 \hypertarget{ref-kozich_development_2013}{}
-28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline
 for analyzing amplicon sequence data on the MiSeq Illumina sequencing
-platform. Applied and Environmental Microbiology 79:5112--5120.
+platform. Applied and Environmental Microbiology. 2013;79:5112--20.
 
 \hypertarget{ref-pruesse_silva_2007}{}
-29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J,
-Glöckner FO. 2007. SILVA: A comprehensive online resource for quality
-checked and aligned ribosomal RNA sequence data compatible with ARB.
-Nucleic Acids Research 35:7188--7196.
+29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
+SILVA: A comprehensive online resource for quality checked and aligned
+ribosomal RNA sequence data compatible with ARB. Nucleic Acids Research.
+2007;35:7188--96.
 
 \hypertarget{ref-edgar_uchime_2011}{}
-30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. 2011. UCHIME
-improves sensitivity and speed of chimera detection. Bioinformatics
-(Oxford, England) 27:2194--2200.
+30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
+sensitivity and speed of chimera detection. Bioinformatics (Oxford,
+England). 2011;27:2194--200.
 
 \hypertarget{ref-wang_naive_2007}{}
-31. Wang Q, Garrity GM, Tiedje JM, Cole JR. 2007. Naive Bayesian
-classifier for rapid assignment of rRNA sequences into the new bacterial
-taxonomy. Applied and Environmental Microbiology 73:5261--5267.
+31. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
+for rapid assignment of rRNA sequences into the new bacterial taxonomy.
+Applied and Environmental Microbiology. 2007;73:5261--7.
 
 \hypertarget{ref-schloss_assessing_2011}{}
-32. Schloss PD, Westcott SL. 2011. Assessing and improving methods used
-in operational taxonomic unit-based approaches for 16S rRNA gene
-sequence analysis. Applied and Environmental Microbiology 77:3219--3226.
+32. Schloss PD, Westcott SL. Assessing and improving methods used in
+operational taxonomic unit-based approaches for 16S rRNA gene sequence
+analysis. Applied and Environmental Microbiology. 2011;77:3219--26.
 
 \hypertarget{ref-breiman_random_2001}{}
-33. Breiman L. 2001. Random Forests. Machine Learning 45:5--32.
+33. Breiman L. Random Forests. Machine Learning. 2001;45:5--32.
 
 \hypertarget{ref-anderson_permanova_2013}{}
-34. Anderson MJ, Walsh DCI. 2013. PERMANOVA, ANOSIM, and the Mantel test
-in the face of heterogeneous dispersions: What null hypothesis are you
-testing? Ecological Monographs 83:557--574.
+34. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
+the face of heterogeneous dispersions: What null hypothesis are you
+testing? Ecological Monographs. 2013;83:557--74.
 
 \hypertarget{ref-benjamini_controlling_1995}{}
-35. Benjamini Y, Hochberg Y. 1995. Controlling the false discovery rate:
-A practical and powerful approach to multiple testing. Journal of the
-Royal Statistical Society Series B (Methodological) 57:289--300.
+35. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+practical and powerful approach to multiple testing. Journal of the
+Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
 

### 2017-09-15 submission/manuscript_R2.Rmd

--- 2017-09-15 submission/manuscript_R2.Rmd
@@
 
-Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about how treatment impacts the colonic community.
+**Background.** Colorectal cancer is a worldwide health problem.
+Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
 This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
 We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).
-The structure of the microbiota in all three groups changed following treatment (P`r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"<0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons (N=XXX) and untreated individuals with adenomas (N=XXX), advanced adenomas (N=XXX), and carcinomas (N=XXX).
-Using these data, we built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
-The microbiota of patients diagnosed with carcinomas changed towards that of a normal colon following treatment (P<0.XXX).
-Those were initially diagnosed as having adenomas or advanced adenomas were no more likely to have a microbiota resembling that of a normal colon.
-These results further illuminate our understanding of the role of the gut microbiota in tumorigenesis and demonstrate the potential for using the gut microbiota as biomarkers for assessing risk and the effectiveness of treatment.
 
-\newpage
+**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma.
+The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
+Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
+Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment.
+To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
+We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
+Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
 
-## Importance
-
-Colorectal cancer is a worldwide health problem.
-Previous work in animal models and human populations has demonstrated that the gut microbiota has an important role in promoting tumorigenesis through a variety of mechanisms.
-Less well understood is how the gut microbiota changes following treatment for varying stages of tumorigenesis.
-We observed that following treatment of patients diagnosed with carcinoma, their gut microbiota is likely revert to a state associated with individuals who had normal colons whereas those diagnosed with adenomas and advanced adenomas were not.
-This study was a necessary step to ultimately investigating how the microbiota changes in those with carcinoma recurrence and predicting which patients are more likely to recur.
-Finding better strategies to stratify patients based on recurrence risk could provide a more effective means of planning therapies and monitoring patients after treatment.
+**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
+Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas.
+Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
 

### 2017-09-15 submission/supplemental.Rmd
update with Table S5

@@
 ```
+
+\newpage
+
+**Table S5: Summary of mtry ranges used for the treatment and diagnosis models**
+
+\footnotesize
+
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+
+treat_rand <- sapply(c("adn", "SRN", "CRC"), 
+                     function(x) read.csv(paste("../data/process/tables/", x, 
+                                                "_randomization_treatment_ROC_model_summary.csv", sep = ""), 
+                                          header = T, stringsAsFactors = F), simplify = F)
+
+treat_actual <- sapply(c("adn", "srn", "crc"), 
+                       function(x) read.csv(paste("../data/process/tables/", x, 
+                                                  "_treatment_ROC_model_summary.csv", sep = ""), 
+                                            header = T, stringsAsFactors = F), simplify = F)
+
+diag_actual <- sapply(c("adn", "srn", "crc"), 
+                       function(x) read.csv(paste("../data/process/tables/", x, 
+                                                  "_ROC_model_summary.csv", sep = ""), 
+                                            header = T, stringsAsFactors = F), simplify = F)
+
+treat_rand_adn <- as.data.frame(table(treat_rand$adn$best_mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("random", length(mtry)), disease = rep("adenoma", length(mtry)))
+treat_rand_srn <- as.data.frame(table(treat_rand$SRN$best_mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("random", length(mtry)), disease = rep("advanced adenoma", length(mtry)))
+treat_rand_crc <- as.data.frame(table(treat_rand$CRC$best_mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("random", length(mtry)), disease = rep("carcinoma", length(mtry)))
+
+treat_act_adn <- as.data.frame(table(treat_actual$adn$best_mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("treatment", length(mtry)), disease = rep("adenoma", length(mtry)))
+treat_act_srn <- as.data.frame(table(treat_actual$srn$best_mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("treatment", length(mtry)), disease = rep("advanced adenoma", length(mtry)))
+treat_act_crc <- as.data.frame(table(treat_actual$crc$best_mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("treatment", length(mtry)), disease = rep("carcinoma", length(mtry)))
+
+diag_act_adn <- as.data.frame(table(diag_actual$adn$mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("diagnosis", length(mtry)), disease = rep("adenoma", length(mtry)))
+diag_act_srn <- as.data.frame(table(diag_actual$srn$mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("diagnosis", length(mtry)), disease = rep("advanced adenoma", length(mtry)))
+diag_act_crc <- as.data.frame(table(diag_actual$crc$mtry)) %>% 
+  rename(mtry = Var1, occurance = Freq) %>% 
+  mutate(model = rep("diagnosis", length(mtry)), disease = rep("carcinoma", length(mtry)))
+
+full_table <- rbind(treat_rand_adn, treat_rand_srn, treat_rand_crc, 
+                    treat_act_adn, treat_act_srn, treat_act_crc, 
+                    diag_act_adn, diag_act_srn, diag_act_crc)
+
+colnames(full_table) <- c("mtry", "Occurence", "Model", "Disease")
+
+kable(full_table, format = "markdown", align = 'c', padding = 2)
+
+```
+\normalsize

### 2017-09-15 submission/manuscript_R2.tex

--- 2017-09-13 submission/manuscript_R2.tex
@@
 
-\textbf{Conclusions.} It was difficult to identify significant
-differences between pre and post-treatment samples from patients with
-adenoma and advanced adenoma. However, for carcinomas, not only were
-there large changes in pre versus post-treatment samples, but these
+\textbf{Conclusions.} By better understanding the response of the
+microbiota to treatment for adenomas and carcinomas, it is likely that
+biomarkers will eventually be validated that can be used to quantify the
+risk of recurrence and the likelihood of survival. Although it was
+difficult to identify significant differences between pre and
+post-treatment samples from patients with adenoma and advanced adenoma,
+this was not the case for carcinomas. Not only were there large changes
+in pre versus post-treatment samples for those with carcinoma, but these
 changes were towards a more normal microbiota.
 
-\newpage
+\subsubsection{Keywords}\label{keywords}
 
-\subsection{Importance}\label{importance}
-
-This study helps to clarify what happens to the microbiota after
-generalized treatment for an adenoma or carcinoma. It is a necessary
-first step to ultimately investigating how the microbiota changes in
-those with carcinoma recurrence. This is important since a large number
-of those treated for carcinoma will have a recurrence. Finding better
-ways to stratify those most and least likely to recur using the
-microbiota could provide a cheaper and more accurate alternative to the
-current status quo. Our results suggest that the microbiota could be
-used to stratify individuals for risk of recurrence and provides a
-starting point for future studies into this area.
+microbiota; colorectal cancer; polyps; treatment; risk factor.
 
@@
 tumorigenesis then one would hypothesize that treatment to remove a
-lesion would affect the microbiota and risk of recurrence. To test this
-hypothesis, we addressed two related questions: Does treatment affect
-the colonic microbiota in a predictable manner? If so, does the
+lesion would not only remove the lesion, but also the microbiota that
+promoted the tumorigenesis and hence the risk of recurrence. To test
+this hypothesis, we addressed two related questions: Does treatment
+affect the colonic microbiota in a predictable manner? If so, does the
 treatment alter the community to more closely resemble that of
@@
 allow us to differentiate between pre and post-treatment samples from
-each of the diagnosis groups. The models overall performed moderately,
-with the carcinoma model being the best (adenoma AUC range = 0.54 -
-0.83, advanced adenoma AUC range = 0.34 - 0.65, carcinoma AUC range =
-0.82 - 0.98). Interestingly, the top 10 most important OTUs by MDA that
-were used for each model had little overlap with each other {[}Figure
-2{]}. Although treatment had an impact on the overall community
-structure, the effect of treatment was not consistent across patients
-and diagnosis groups. Both the adenoma and carcioma treatment models had
-AUCs that were significantly higher than a random model permutation
-(P-value \textless{} 0.0001).
+each of the diagnosis groups. The adenoma and carcinoma models performed
+well (adenoma AUC range = 0.54 - 0.83 and carcinoma AUC range = 0.82 -
+0.98); however, the model for patients treated for advanced adenomas was
+not able to reliably differentiate between the pre and post-treatment
+samples (advanced adenoma AUC range = 0.34 - 0.65). Interestingly, the
+top 10 most important OTUs by MDA that were used for each model had
+little overlap with each other {[}Figure 2{]}. Although treatment had an
+impact on the overall community structure, the effect of treatment was
+not consistent across patients and diagnosis groups. Both the adenoma
+and carcioma treatment models had AUCs that were significantly higher
+than a random model permutation (P-value \textless{} 0.0001).
 
@@
 used to create the three models used to classify pre and post-treatment
-samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
-total number of individuals in the pre versus post-treatment models was
-67 individuals. There were a total of 22 individuals in the pre versus
-post-treatment adenoma model, 19 individuals in the pre versus
-post-treatment advanced adenoma model, and 26 individuals in the pre
-versus post-treatment carcinoma model {[}Table 1{]}.
+samples by diagnosis (adenoma, advanced adenoma, or carcinoma) as well
+as to assess the probability that a sample was more similar to the
+patient's original diagnosis or that of a disease-free patient. All
+models included only OTU data obtained from 16S rRNA sequencing and were
+processed using the caret (v6.0.76) R package. For each model we
+optimized the mtry hyper-parameter, which defines the number of OTUs to
+investigate at each split before a new division of the data was created
+with the Random Forest model (33). To insure that our optimization did
+not result in over-fitting of the data, we made 100 different 80/20
+(train/test) splits of the data where the same proportion was present
+within both the whole data set and the 80/20 split. For each of the 100
+splits, 20 repeated 10-fold cross validation was performed on the 80\%
+component to optimize the mtry hyper-parameter by maximizing the AUC
+(Area Under the Curve of the Receiver Operator Characteristic). The
+resulting model was then tested on the 20\% of the data that were held
+out. A summary of the mtry hyperparameter values that were tried is
+available in Table S5. The reported P-values for each model relative to
+a random labeling was assessed by comparing the distribution of the 100
+80/20 splits for the correctly labeled data to the distribution of
+randomly labeled data.
 
-Similarly, the Random Forest (33) algorithm was also used to create the
-three models used to classify normal versus diagnosis. These samples
-were obtained using the same methodology as described earlier in this
-section. All samples used for this component of model training were from
-pre-treatment samples only. The total number of individuals in the
-normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
-were a total of 239 individuals in the normal versus adenoma model, 262
-individuals in the normal versus advanced adenoma model, and 266
-individuals in the normal versus carcinoma model {[}Table 2{]}.
-
-All models included only OTU data obtained from 16S rRNA sequencing and
-were processed and cleaned using the R package caret (v6.0.76).
-Optimization of the mtry hyper-parameter involved making 100 different
-80/20 (train/test) splits of the data where the same proportion was
-present within both the whole data set and the 80/20 split. For each of
-the different splits, 20 repeated 10-fold cross validation was performed
-on the 80\% component to optimize the mtry hyper-parameter by maximizing
-the AUC (Area Under the Curve of the Receiver Operator Characteristic).
-The resulting model was then tested on the hold out data obtained from
-the 20\% component. For adenoma pre versus post-treatment models the
-optimized mtrys were 2 (n = 42), 42 (n = 26), and 914 (n = 32) for
-adenoma. For advanced adenoma the mtrys were 2 (n = 26), 52 (n = 14),
-and 1353 (n = 60). For carcinoma the mtrys were 2 (n = 36), 41 (n = 54),
-and 863 (n = 10). For the normal versus diagnosis models the optimized
-mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
-adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
-NA), and (n = NA). The hyper-parameter, mtry, defines the number of
-variables to investigate at each split before a new division of the data
-was created with the Random Forest model (33).
-
-The treatment models were then assessed for their strength to classify
-whether individuals had obtained treatment. We also assessed whether the
-models performed better than random by creating 100 permutations of a
-data set in which the labels were randomly assigned. To assess
-similarity of the treatment models we looked for overlap between their
-top 10 most important OTUs.
-
-For the diagnosis models that were trained on a completely independent
-sample set, each model was then applied to our 67-person cohort {[}Table
-1{]} based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22
-and disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
-disease free n = 22)), advanced adenoma pre-treatment advanced adenoma
-(advanced adenoma n = 19 and disease free n = 0 ) versus post-treatment
-advanced adenoma (advanced adenoma n = 0 and disease free n = 19), and
-carcinoma (pre-treatment carcinoma (carcinoma n = 26 and disease free n
-= 0) versus post-treatment carcinoma (carcinoma n = 1 and disease free n
-= 25)). The application of the pre versus post-treatment models
-generated the probabilities that the sample was a pre-treatment sample.
-The application of the normal versus diagnosis models generated the
-probabilities that the sample was that specific diagnosis (adenoma,
-advanced adenoma, or carcinoma). We also looked at similarities between
-the models by looking for overlap in the top 10\% of important OTUs to
-each respective model.
+The three diagnosis models were constructed by using the data from
+Baxter et al. (24), which was censored for the pre-treatment samples of
+the patients that we had post-treatment samples. The treatment models
+were then used to quantify the model probability that a patient with an
+initial diagnosis retained that diagnosis or a disease-free diagnosis.
 

### 2017-09-15 submission/manuscript_R2.Rmd
update the pvalue for the abstract

@@
 We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
-Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05).
+Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.001, invisible("<"), invisible(">"))` 0.001).
 Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
@@
 Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
+
+### Keywords
+
+microbiota; colorectal cancer; polyps; treatment; risk factor.
 
@@
 The resulting model was then tested on the 20% of the data that were held out.
-A summary of the mtry hyperparameter values that were tried is available in Table SX.
+A summary of the mtry hyperparameter values that were tried is available in Table S5.
 The reported P-values for each model relative to a random labeling was assessed by comparing the distribution of the 100 80/20 splits for the correctly labeled data to the distribution of randomly labeled data.

### 2017-09-15 submission/manuscript_R2.tex

@@
 had carcinomas changed towards a microbial milieu that resembles the
-normal colon after treatment (P-value \textless{} 0.05). Finally, we
+normal colon after treatment (P-value \textless{} 0.001). Finally, we
 were unable to detect any significant differences in the microbiota of

### 2017-09-15 submission/manuscript_R2.Rmd

@@
 Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
-Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment.
+Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value  treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`).
+Furthermore, there was little overlap between the OTUs that were indicative of each treatment.
 To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
@@
 All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
+The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.”
 

### 2017-09-15 submission/manuscript_R2.tex

@@
 classify pre and post-treatment samples for each of the diagnosis
-groups. Although the three models could differentiate between the pre
-and post-treatment samples, there was little overlap between the OTUs
-that were indicative of each treatment. To determine whether individuals
-who underwent treatment were more likely to have OTUs associated with
-normal colons we used a larger cohort that contained individuals with
-normal colons and those with adenomas, advanced adenomas, and
-carcinomas. We again built Random Forest models and measured the change
-in the positive probability of having one of the three diagnoses to
-assess whether the post-treatment samples received the same
-classification as the pre-treatment samples. Samples from patients who
-had carcinomas changed towards a microbial milieu that resembles the
-normal colon after treatment (P-value \textless{} 0.001). Finally, we
-were unable to detect any significant differences in the microbiota of
-individuals treated with surgery alone and those treated with
-chemotherapy or chemotherapy and radiation (P-value \textgreater{}
+groups. Although the adenoma and carcinoma models could reliably
+differentiate between the pre and post-treatment samples (P-value
+treatment models had AUCs that were significantly higher than a random
+model permutation (P-value \textless{} 0.001), the advanced-adenoma
+model could not (P-value = 0.61). Furthermore, there was little overlap
+between the OTUs that were indicative of each treatment. To determine
+whether individuals who underwent treatment were more likely to have
+OTUs associated with normal colons we used a larger cohort that
+contained individuals with normal colons and those with adenomas,
+advanced adenomas, and carcinomas. We again built Random Forest models
+and measured the change in the positive probability of having one of the
+three diagnoses to assess whether the post-treatment samples received
+the same classification as the pre-treatment samples. Samples from
+patients who had carcinomas changed towards a microbial milieu that
+resembles the normal colon after treatment (P-value \textless{} 0.001).
+Finally, we were unable to detect any significant differences in the
+microbiota of individuals treated with surgery alone and those treated
+with chemotherapy or chemotherapy and radiation (P-value \textgreater{}
 0.05).
@@
 consent and the study itself conformed to the guidelines set out by the
-Helsinki Declaration.
+Helsinki Declaration. The original protocol for the study did not
+provide for tracking patients after the follow up samples and so it was
+not possible for us to ascertain their diagnosis after the completion of
+the study.''
 

### 2017-09-15 submission/manuscript_R2.Rmd
remove stray quotation

@@
 All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
-The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.”
+The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.
 

### 2017-09-15 submission/manuscript_R2.tex
remove stray quotation

@@
 not possible for us to ascertain their diagnosis after the completion of
-the study.''
+the study.
 

### 2017-09-15 submission/manuscript_R2.Rmd
add missing parenthesis

--- 2017-09-15 submission/manuscript_R2.Rmd
@@
 Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
-Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment.
+Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001)), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`).
+Furthermore, there was little overlap between the OTUs that were indicative of each treatment.
 To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
@@
 All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
+The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.
 

### 2017-09-15 submission/manuscript_R2.tex
add missing parenthesis

@@
 treatment models had AUCs that were significantly higher than a random
-model permutation (P-value \textless{} 0.001), the advanced-adenoma
+model permutation (P-value \textless{} 0.001)), the advanced-adenoma
 model could not (P-value = 0.61). Furthermore, there was little overlap

### 2017-09-16 submission/manuscript_R2.Rmd
incorporate Pat sentence change

--- 2017-09-15 submission/manuscript_R2.Rmd
@@
 Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
-Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value  treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`).
+Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`).
 Furthermore, there was little overlap between the OTUs that were indicative of each treatment.

### 2017-09-16 submission/manuscript_R2.tex

--- 2017-09-15 submission/manuscript_R2.tex
@@
 differentiate between the pre and post-treatment samples (P-value
-treatment models had AUCs that were significantly higher than a random
-model permutation (P-value \textless{} 0.001), the advanced-adenoma
-model could not (P-value = 0.61). Furthermore, there was little overlap
-between the OTUs that were indicative of each treatment. To determine
-whether individuals who underwent treatment were more likely to have
-OTUs associated with normal colons we used a larger cohort that
-contained individuals with normal colons and those with adenomas,
-advanced adenomas, and carcinomas. We again built Random Forest models
-and measured the change in the positive probability of having one of the
-three diagnoses to assess whether the post-treatment samples received
-the same classification as the pre-treatment samples. Samples from
-patients who had carcinomas changed towards a microbial milieu that
-resembles the normal colon after treatment (P-value \textless{} 0.001).
-Finally, we were unable to detect any significant differences in the
-microbiota of individuals treated with surgery alone and those treated
-with chemotherapy or chemotherapy and radiation (P-value \textgreater{}
-0.05).
+\textless{} 0.001), the advanced-adenoma model could not (P-value =
+0.61). Furthermore, there was little overlap between the OTUs that were
+indicative of each treatment. To determine whether individuals who
+underwent treatment were more likely to have OTUs associated with normal
+colons we used a larger cohort that contained individuals with normal
+colons and those with adenomas, advanced adenomas, and carcinomas. We
+again built Random Forest models and measured the change in the positive
+probability of having one of the three diagnoses to assess whether the
+post-treatment samples received the same classification as the
+pre-treatment samples. Samples from patients who had carcinomas changed
+towards a microbial milieu that resembles the normal colon after
+treatment (P-value \textless{} 0.001). Finally, we were unable to detect
+any significant differences in the microbiota of individuals treated
+with surgery alone and those treated with chemotherapy or chemotherapy
+and radiation (P-value \textgreater{} 0.05).
 
